### 2013-03-17 writeup/conclusion.tex
Intro writing

\section{Conclusion}

VIII. Conclusion (.5pg)

HATS represent a new class of database system:
* Can't get serializability, but can get other properties
* Key challenge: designing distributed algorithms and architectures for partial failure, asynchrony
* Rearchitecture of DBMSs and better understanding of applications are key

References (1pg)

### 2013-03-17 writeup/evaluation.tex
Intro writing

\section{evaluation}

V. Evaluation (1.5pg)

The goal in this section is twofold:
1.) Look at existing algorithms and their coordination cost in the abstract
2.) Perform limited real-world deployments to validate these abstract costs

1.) There are a bunch of existing systems that aren't HATS but provide stronger semantics. What's their cost? (N.B. this is almost a related work survey here, but I think, if done right, it can be very instructive.)
Not HA, Serializable:
* Two-phase locking: for transaction size T, T+1 round-trips (directly to each database)
* Deterministic transaction scheduling: 1 round-trip per partition (to scheduler)
* Optimistic concurrency control: 1 round-trip per partition (to database cycle checker)
* MVCC: min 1 round-trip per partition (to partition timestamp manager)

Not HA, not serializable, but "transactional":
* MDCC: need to check, 1-2 round-trip times because uses Paxos variant
* Walter: 1 rount-trip time to each preferred site
* Eiger: Sticky-HA, assumes linearizable clusters, reads/writes stall in presence of coordinator failures
* Chan and Gray read-only transaction: reads stall in the presence of coordinator failure
* Bolt-on causal consistency: sticky-HA, could be used to implement TA+RR, uses client cache
* COPS: Sticky-HA, assumes linearizable clusters
* Bayou: not HA since client doesn't cache (pretty sure)

2.) We ran YCSB on EC2 (and maybe TPC-C if we have time) with a few models:
	- Eventual Consistency
	- Eventual consistency with a "master" for updates--simulates the lower bound on a non-HAT system.
	- Transactional Atomicity and Read Committed

	-This is going to look a lot like the ping tests: much higher latency for "non-HATs."
	-Our Naive 2PL implementation bottlenecks extremely quickly.
	-On, say, 5 servers, can get hundreds of thousands of TPS

	-Not meant as an exhaustive study, but validates our intuitions. Coordination costs of non-HAT systems are unaccounted for, will only go up, and we haven't even talked about availability.

### 2013-03-17 writeup/relatedwork.tex
Intro writing

\section{Related Work}

VI. Related Work (.75pg)

* Probably do a good job with this one, in addition to the first part of Section V. We can talk at length about the relationship between distributed systems and databases.

1.) High Availability and CAP (giving the DB folks due credit here!)
2.) Highly Available Distributed Systems
3.) Weak Isolation (there's a lot here, pay plenty of lip service to Adya)
4.) Available/unavailable systems
	--Stress that, with the advent of Spanner, transactions are coming back; HATS represents new design space. (Can even say "Spanner borrowed traditional database techniques; would benefit from HAT approach (c.f., 100ms geo-replicated latencies)"; also talk about optimization for read-only transactions).

### 2013-03-17 thebes-code/scripts/setup_hosts.py
checkpoing scripts

@@ -8,6 +8,7 @@ from common_funcs import sed
 from common_funcs import upload_file
 from common_funcs import run_script
 from common_funcs import fetch_file_single
+from common_funcs import fetch_file_single_compressed
 from threading import Thread, Lock
 import os
 from datetime import datetime
@@ -82,10 +83,10 @@ class Host:
 
 # UTILITIES
 def run_cmd_in_thebes(hosts, cmd, user='root'):
-    run_cmd(hosts, "cd /home/ubuntu/thebes/thebes-code; %s" % cmd, user)
+    run_cmd(hosts, "cd /home/ubuntu/thebes/thebes-code; ulimit -u unlimited; %s" % cmd, user)
 
 def run_cmd_in_ycsb(hosts, cmd, user='root'):
-    run_cmd(hosts, "cd /home/ubuntu/thebes/ycsb-0.1.4; %s" % cmd, user)
+    run_cmd(hosts, "cd /home/ubuntu/thebes/ycsb-0.1.4; ulimit -u unlimited; %s" % cmd, user)
 
 
 def get_instances(regionName):
@@ -149,6 +150,7 @@ def get_num_nonterminated_instances(regionName):
 
     for line in open("/tmp/running.txt"):
         line = line.split()
+        print line
         if line[0] == "INSTANCE" and line[3] != "terminated":
             num_nonterminated = num_nonterminated + 1
         elif line[0] == "TAG" and line[3] in tag_blacklist:
@@ -156,6 +158,7 @@ def get_num_nonterminated_instances(regionName):
                 blacklisted_hosts.append(line[2])
 
     system("rm /tmp/running.txt")
+    print blacklisted_hosts
     return num_nonterminated-len(blacklisted_hosts)
 
 def make_instancefile(name, hosts):
@@ -203,7 +206,7 @@ def provision_clusters(regions, use_spot, anti_slow):
 
 def provision_graphite(region):
     global AMIs
-    if region == None:
+    if region == None or region=="":
         pprint('Graphite not enabled.')
 
     provision_instance(region.name, 1)
@@ -211,12 +214,12 @@ def provision_graphite(region):
 
 def provision_spot(regionName, num):
     global AMIs
-    system("ec2-request-spot-instances %s --region %s -t t1.micro -price 0.02 " \
+    system("ec2-request-spot-instances %s --region %s -t m1.xlarge -price 0.50 " \
[…]

### 2013-03-17 writeup/futurework.tex

\section{Future Work}

VII. Future Work (.75pg)

* What properties are available?
	* We've focused on ACID Isolation, but there are lots of other models to consider?
	* Which consistency criteria? Which models? There are infinite, so how do we choose?
	* Set ourselves up for APEs

* Full exploration of HAT space
	* Many parameters; $2^10$ models at the least
	* Need to mitigate the cost of tracking causality with high availability
		* The cost moving from S-HA (like Eiger) to truly HA (R-HA) seems to be high
		* e.g., Eiger relies on metadata GC to keep writes small; is O(|keys|) in limit
			* Not meant as a fault--it's optimized for a particular scenario--but curious to consider

* Revisit semantics-based concurrency control
	* Could taxonomize properties like escrow
	* Can also consider models with SLAs and bounded messaging asynchrony
	* Our focus is on "classic" distributed systems trade-off
		* However, lots of interesting hybrids (maybe cite Doug Terry's recent stuff)

### 2013-03-17 writeup/hat-analysis.tex

IV. HATS and ACID (4pg)

We will show what semantics are available in HATS as well as which aren't. Our strategy will be to build up a set of properties until we reach a point of unavailability. We will start with isolation and atomicity since they are well-defined in the literature and most meaty.

A. *What's Available in Isolation and Atomicity?*

* We can prevent Dirty Write phenomena by consistently ordering transactions
	* This is basically a cross-data item convergency property, as are found in eventually consistent systems!
* Read Committed
	* Don't read dirty data by never exposing dirty data to readers!
* ANSI Repeatable Read => rename to cut isolation
	* Repeatable Read is a tricky guarantee. In ANSI SQL Spec, rather weak, but in follow-on work, it's pretty strong.
	* The property that the ANSI SQL spec talks about is about reading a "snapshot" of data items, along with your own writes.
		* A "snapshot" describes a "consistent cut" across data items; we'll call it "cut isolation"--when paired with below guarantees, makes sense.
	* This doesn't provide any constraints on writes.

* There are several properties that aren't discussed in the formal literature on isolation guarantees but which are really useful.

* Transactional atomicity: "A" in ACID
	* Once a transaction reads one of another transactions's writes, its subsequent reads will return the transactions's other writes (or a suitable "later" write).
	* Be really pedagogical here: "transactional atomicity" versus "distributed linearizability" both called "atomicity"; we actually explore the differences later, in Discussion section.
	* Talk about implementation here: it's basically uniform eager reliable delivery from distributed systems literature!
	* Note that this doesn't make recency guarantees (forward reference to later section on "impossibility")

* Session guarantees give useful guarantees within and across transactions
	* Without them, cut isolation is sort of meaningless: we can always just return NULL!
	* Several properties: monotonic writes, monotonic reads, writes follow reads, read your writes
	* We can define them within a transaction only (e.g., transactional monotonic writes) or across subsequent transactions by the same client (e.g., client monotonic writes).

* Session guarantees are tricky
	* You can implement all but read-your-writes in a R-HA system
		* Never show a write until all replicas in the system have seen it!
		* This means that, if a client switches replicas, then the replica will have a satisfying set of writes.
                * Take S3 as an example of a system where stickiness coudl have helped, but no one really talked about it!
		* Downside: when do writes become visible? Only when all replicas have seen it!
		* Side note: assuming that, if replicas can enter and leave, they can only become active once they "catch up" to global lower bound
			* We leave dynamic replica selection as Future Work (N.B. in APEs!)
	* Read-your-writes requires S-HA
		* Example: partition a client from all but one replica, have the client issue a write, which has to commit. Next, lift the partition between the client and the other replicas and partition the client and its previous replica. It can't read it's own write!
		* This also means that causality, which is MR+MW+WFR+RYW is unavailable!
		* S-HA is *doable* but it does result in substantially lowered unavailability unless one caches (e.g., bolt-on causal consistency)

* What do we have?
	* In Adya, we get up to Causal+Cut Isolation for S-HA, and PL-2 for R-HA.
	* In distributed systems terminology, in S-HA we have causal consistency but each transactions' updates are a cycle in the happens-before graph!
		* Stress that this is a big unification of the existing models.
	* Recap, pedagogically, why this is highly available and what this means: no locks, no central coordination, no need for RTTs; bring back to Section II.

B. *What's not available in Isolation?*

Now, at least in terms of the existing literature, we hit the limits.

Let's sketch out a few conditions, then we'll discuss them later:
* Lost Update
	* Simple proof: T1 R(x) W(x=x+1) T2 R(x) W(x=x+1); execute on opposite sides 
* Read Skew
	* Similar easy proof
* Write Skew
	* Similar easy proof
* Anti-dependency cycles
	* Serializability
* Recency guarantees on data items
	* CAP

Now refer to Adya's chart. No PL-SI, PL-SS, PL-2.99, etc.

C. *What's available in consistency?*

* In general, we can't maintain correctness conditions over arbitrary data items.
	* This is due to serializability problems.
	* Uniqueness, for example, is out the window

* We can evaluate locally checkable correctness criteria (e.g., check for null)
	(QUESTION: should we save this for APEs?)

* However, with semantic knowledge, we can do a little bit better
	* Commutative updates and logical monotonicity are fine; e.g., write-write conflicts don't matter

Example: In TPC-C New Order, the new order id assignment isn't commutative, tough. but the inventory checking *is*!

D. Durability

* If you want data to survive *F* failures, you'll have to replicate to *F* replicas.
	* This is pretty easy.

E. Summary

* Left with a bunch of binary properties: 
R-HA: Prevent Dirty Write, Read Committed, Cut Isolation, Transactional Atomicity, Transactional Monotonic Reads, Transactional Monotonic Writes, Transactional Writes-Follow-Reads, Client Monotonic Reads, Client Monotonic Writes, Client Writes-Follow-Reads
S-HA: Client Read Your Writes, Transactional Read Your Writes

* Combining all of the above results in the strongest models we've yet seen. No

*Discussion*

* Composition of Properties can be tricky:
	* Consider ensuring RR and TA:
		* If you do this naively, you may end up returning lots of nulls
		* Give example: T1 w(x1) w(y1) T2 w(x2) w(y2) T3 r(x) r(y) => T3 better read x1, y1 or x2, y2!
	* Some are more expensive; not likely to achieve causality with less than O(clients) metadata required
		* Note that making stronger assumption about clusters, like linearizability *can* help (but then not HA!)

* Transactional Atomicity versus Linearizability
	* What's the difference between our TA and distributed systems linearizability? Is there some sort of mapping?
		* TA: writes to multiple keys are indivisible to readers
		* Linearizability: writes to single key on multiple servers are indivisible to readers
	* The differences are two-fold:
		* In linearizability (and safe and regular registers), writes are visible to clients immediately after they finish
			* HATS have no recency requirement
		* In linearizability, all clients see all writes at the same time!
			* In TA, clients see writes at different times depending on what replicas they contact
			* No such analog in traditional distributed systems literature on replica consistency

* S-HA Monotonic Reads and Monotonic Writes
	* In the model we discussed, you can get monotonic reads in a R-HA system by waiting until all replicas see a write.
	* This leads to poor visibility--writes take a long time to show up. Even if you didn't want to read your writes, you might want to read other peoples' writes!
	* Accordingly, there are valid reasons for S-HA session guarantees, namely, visibility.

### 2013-03-17 writeup/hat-vldb.tex

% THIS IS AN EXAMPLE DOCUMENT FOR VLDB 2012
% based on ACM SIGPROC-SP.TEX VERSION 2.7
% Modified by  Gerald Weber <gerald@cs.auckland.ac.nz>
% Removed the requirement to include *bbl file in here. (AhmetSacan, Sep2012)
% Fixed the equation on page 3 to prevent line overflow. (AhmetSacan, Sep2012)

\documentclass{vldb}
\usepackage{graphicx, url}
\usepackage{balance}  % for  \balance command ON LAST PAGE  (only there!)

\begin{document}

% ****************** TITLE ****************************************

\title{Highly Available Transactions: Virtues and Limitations}

\maketitle

\begin{abstract}
foo bar baz
\end{abstract}

\input{intro.tex}

\input{motivation.tex}

\input{model.tex}

\input{hat-analysis.tex}

\input{evaluation.tex}

\input{relatedwork.tex}

\input{futurework.tex}

\input{conclusion.tex}

VIII. Conclusion (.5pg)

HATS represent a new class of database system:
* Can't get serializability, but can get other properties
* Key challenge: designing distributed algorithms and architectures for partial failure, asynchrony
* Rearchitecture of DBMSs and better understanding of applications are key

References (1pg)

foo~\cite{adya}

% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{abbrv}
% vldb_sample.bib is the name of the Bibliography in this case
\bibliography{hat-vldb}  
% You must have a proper ".bib" file
%  and remember to run:
% latex bibtex latex latex
% to resolve all references

\subsection{References}

Generated by bibtex from your ~.bib file. Run latex, then bibtex, then
latex twice (to resolve references).

%APPENDIX is optional.
% ****************** APPENDIX **************************************
% Example of an appendix; typically would start on a new page
%pagebreak

\begin{appendix}

\end{appendix}

\end{document}

### 2013-03-17 writeup/intro.tex

\section{Introduction}

Many recent systems have eschewed ACID guarantees.

CAP Theorem says you can't have consistency and availability and also talks about partitions.

What does CAP have to say about ACID? Recent workshop result on Highly Available Transactions briefly demonstrated that some ACID guarantees are indeed achievable in a highly available manner.

Many ACID systems were designed either i.) for single systems or ii.)
to provide serializability. The combination of weak isolation and
distributed systems necessitates further study and
design. \textit{What is the benefit of high availability and what is
  the cost?}

Highly Available Transactions are the class of multi-object operations
and their respective semantics that can be provided with high
availability.

Weak isolation has a cost: in the words of one database pioneer, you
``fall off a cliff'' when it comes to reasoning about your
program. However, it also has a benefit: availability and latency.

Our goal in this paper is to objectively map the space of trade-offs and

 This paper fills in the remaining gaps.

What does high availability even mean in a transaction environment?

What models are achievable, and which aren't?

In this paper, we make the following contributions:

\begin{itemize}
\item We present a model for high availability in a transactional environment, including Replica-agnostic High Availability (R-HA) and Sticky High Availability (S-HA).
\item We analyze the existing literature on ACID properties (particularly, ANSI SQL Spec, Adya, Gray) to show which properties* with a focus on isolation* are achievable in HATS. Achievable properties include transactional atomicity, variants of repeatable read isolation, and session guarantees. We also describe unavailable properties, like preventing Lost Update and Read and Write Skew. We believe that this is the first such characterization in the literature.
\item We further motivate the study of HATS via the analysis of existing concurrency control algorithms in a highly available environment as well as through deployment of an experimental database prototype on public cloud infrastructure.
\end{itemize}

### 2013-03-17 writeup/model.tex

\section{What is High Availability?}

* System model
	* Consider transactions that issue a series of reads and writes, then commit or abort
	* Each data item is located on one or more servers

* HA Definitions
	* Straw-man definition: every transaction will recieve a response
		* Problem: system that always aborts is trivially available!

* Replica-HA
	* (Handwaving in the outline) If a client can contact at least one replica for each data item in a transaction, then the transaction will eventually commit, even in the presence of infinite network partitions.

* Sticky-HA
	* If a client continues to contact the same set of replicas across transactions and the client can contact at least one replica for each data item in a transaciton, then the transaction will eventually commit, even in the presence of infinite network partitions.
	* We will use this definition when it comes to causality and read your writes guarantees!
	* Can be achieved by either "sticking" clients to the same set of servers or by having the client act as a replica itself, via caching

### 2013-03-17 writeup/motivation.tex

\section{The Case for HATS}

Why High Availability?
* Highly available system operation is often desirable; if you can contact one server, you can get a response!
* As a corrollary, an "AP" system is faster than a "CP" system; see "PACELC" by Abadi.
	* Look, here are some datapoints we recently measured on EC2. Network times are at least 10x faster, often much more for "AP"

Availability examples

Within DC at Microsoft: 40.8 mean link failures per day, 136 at 95th
percentile, with high variability, median time to repair around 5
minutes but tail up to one week. Probability of device failing and
impacting network (per year) is near 10\% for several core components
like core routers and primary-backup connectors. Network redundancy
only reduces median impact of failures by up to
40\%~\cite{sigcomm-dc}.

HP Labs: WAN, LAN, and connectivity problems account for 28.1\% of all
customer support tickets, 39\% is network hardware, only 6\% for
configuration and 15\% for software. Median duration of highest
priority is on the order of several hours (1:54 for WAN, 2:45 for LAN,
2:13 for Hardware), going to a full day for all
incidents.~\cite{turner2012failure}.

Network partitions happen at Google~\cite{dean-keynote}.

Many anecdotes online: Amazon's EC2 and RDS down due to net
partition~\cite{amazon-netpartition}, as a result of AZ failures full
site outages like Reddit, Foursquare, and Heroku~\cite{ec2-downsites},
accounts of network partitions~\cite{netpartition-anecdote1}

Despite speculation over partition
frequency~\cite{stonebraker2010errors}, many practitioners indicate
that partitions do indeed occur: in the words of James Hamilton, Vice
President and Distinguished Engineer on the Amazon Web Services team,
``network partitions should be rare but net gear continues to cause
more issues than it should''~\cite{hamilton-partitions}

Many of our discussions with practitioners---especially those
operating on public cloud infrastructure---confirm that partitions are
a reality.

Area of active research in networking
community~\cite{surviving-failures-bodik}

Over WAN:

From 2002 study of Sprint's networks, median time to repair is between 2 and < 1000 seconds, approximately 2000 seconds between router failures and around 8,000 seconds between optical failures. Median failure rate is one every 3000 seconds.~\cite{ip-backbone-failures}.

From 2002 study of VoIP, 50 minute failure caused 100\% packet loss between east and west coasts~\cite{voip-partitions}.

* As an example of a real system going from "CP" to "AP", [PNUTS dropped per-record mastering](\url{http://developer.yahoo.com/blogs/ydn/posts/2010/06/sherpa_update/#4})

What about existing systems?
* Database technology developed for single-node systems; gold standard: serializability
	* Serializability is not actually highly available; give an example!
	* Much of the database literature presumes serializability and does not consider high availability.
	* For the literature that doesn't presume serializability, it's not presented in a HA context?
* Consequence: traditional systems are not optimized for high availability
	* Consider two-phase locking in a distributed environment
	* As we will see (forward reference to Evaluation), many existing transactional systems encounter similar difficulties
* Key question in this paper: What transactional semantics are highly available, and which aren't?
	* Side note: there are infinite incomparable consistency models (e.g., always return 2, always return 3, …)
	* Our goal is to unify distributed systems literature with ACID transaction model.
* Argument: Highly Available Transactional Systems (HATS) require rethinking existing models. This work is a first step.

In this sense, HATs are similar to RAID: optimizing for graceful
handling of a worst-case failure scenario improves average-case
performance.

### 2013-03-18 writeup/hat-vldb.tex
text

@@
 \documentclass{vldb}
-\usepackage{graphicx, url}
+\usepackage{times, graphicx, url, xcolor, multicol, minipage, subfig}
 \usepackage{balance}  % for  \balance command ON LAST PAGE  (only there!)
@@
 
-VIII. Conclusion (.5pg)
-
-HATS represent a new class of database system:
-* Can't get serializability, but can get other properties
-* Key challenge: designing distributed algorithms and architectures for partial failure, asynchrony
-* Rearchitecture of DBMSs and better understanding of applications are key
-
-References (1pg)
-
 foo~\cite{adya}

### 2013-03-18 writeup/model.tex
text

@@
-\section{What is High Availability?}
+\section{High Availability}
+\label{sec:availability}
+
+highly available algorithms guaranteeing
+``always on'' operation are applicable in modern distributed
+contexts. This is because, a high level, a highly available system
+(which we formally define in Section~\ref{sec:availabilty}) guarantees
+two useful properties: during network partitions, the system responds
+to requests without blocking, and, during normal, partition-free
+operation, it provides guaranteed one round-trip-time (RTT) latency.
+
+In this section, we survey evidence regarding the prevalence of
+network partitions as well as quantify the cost of round-trip times
+over public cloud infrastructure. These results are independent of
+algorithms employed and, while they reflect state-of-the-art
+technology, they are
+
+Why High Availability?
+* Highly available system operation is often desirable; if you can contact one server, you can get a response!
+* As a corrollary, an "AP" system is faster than a "CP" system; see "PACELC" by Abadi.
+	* Look, here are some datapoints we recently measured on EC2. Network times are at least 10x faster, often much more for "AP"
 
@@
 	* Can be achieved by either "sticking" clients to the same set of servers or by having the client act as a replica itself, via caching
+
+* As an example of a real system going from "CP" to "AP", [PNUTS dropped per-record mastering](\url{http://developer.yahoo.com/blogs/ydn/posts/2010/06/sherpa_update/#4})
+
+What about existing systems?
+* Database technology developed for single-node systems; gold standard: serializability
+	* Serializability is not actually highly available; give an example!
+	* Much of the database literature presumes serializability and does not consider high availability.
+	* For the literature that doesn't presume serializability, it's not presented in a HA context?
+* Consequence: traditional systems are not optimized for high availability
+	* Consider two-phase locking in a distributed environment
+	* As we will see (forward reference to Evaluation), many existing transactional systems encounter similar difficulties
+* Key question in this paper: What transactional semantics are highly available, and which aren't?
+	* Side note: there are infinite incomparable consistency models (e.g., always return 2, always return 3, …)
+	* Our goal is to unify distributed systems literature with ACID transaction model.
+* Argument: Highly Available Transactional Systems (HATS) require rethinking existing models. This work is a first step.
+
+In this sense, HATs are similar to RAID: optimizing for graceful
+handling of a worst-case failure scenario improves average-case
+performance.

### 2013-03-18 writeup/motivation.tex
text

@@
-\section{The Case for HATS}
+\section{Partitions and Latency}
 
-Why High Availability?
-* Highly available system operation is often desirable; if you can contact one server, you can get a response!
-* As a corrollary, an "AP" system is faster than a "CP" system; see "PACELC" by Abadi.
-	* Look, here are some datapoints we recently measured on EC2. Network times are at least 10x faster, often much more for "AP"
+Peter Deutsch begins his list of ``Fallacies of Distributed
+Computing'' with two concerns: ``\textit{1.)}  The network is
+reliable. \textit{2.)} Latency is zero.'' On a single-node system,
+communication systems are robust and relatively fast, and most systems
+do not need to prepare for partial component failure. However, as
+Deutsch points out, these assumptions are no longer valid in a
+distributed setting. To extend Jim Gray's storage latency analogy, if
+memory is around 1.5 hours from San Francisco and disk is as ``far
+away'' as Pluto (2 years), in a distributed context, another computer
+can be even farther---nearing or even worse than Gray's location of
+tape access near Andromeda (2,000 years)---with the added concern that
+messages can be lost in transit, destroyed by ``black holes'' in the
+form of network partitions. Accordingly, accounting for partial
+failure and communication expense is core to distributed systems and
+algorithm design; ignoring them ``cause[s] big trouble and painful
+learning experiences.''
 
-Availability examples
+While prominent members of the database community have publicly
+speculated about the unimportance of network partitions in distributed
+database design~\cite{stonebraker2010errors}, there is mounting
+evidence---both anecdotal and experimental---that partitions occur and
+latency is non-negligible. In this section, we briefly outline this
+evidence and attempt to rigorously quantify the costs of ignoring
+network unreliability and latency.
 
-Within DC at Microsoft: 40.8 mean link failures per day, 136 at 95th
-percentile, with high variability, median time to repair around 5
-minutes but tail up to one week. Probability of device failing and
-impacting network (per year) is near 10\% for several core components
-like core routers and primary-backup connectors. Network redundancy
-only reduces median impact of failures by up to
-40\%~\cite{sigcomm-dc}.
+\subsection{Partitions}
 
-HP Labs: WAN, LAN, and connectivity problems account for 28.1\% of all
-customer support tickets, 39\% is network hardware, only 6\% for
-configuration and 15\% for software. Median duration of highest
-priority is on the order of several hours (1:54 for WAN, 2:45 for LAN,
-2:13 for Hardware), going to a full day for all
-incidents.~\cite{turner2012failure}.
+Network partitions---or, informally, communication link failures
+between participants in a distributed system---are challenging. A
+system facing network partitions faces the impossibility of
+simultaneously maintaining a correct ``global'' view of system state
+while simultaneously providing operation to all of its clients. The
+rate at which a network experiences partitions and the severity and
+duration of each partition depends on the given network deployment,
+hardware, topology, administration, and a host of other
+factors. However, given the \textit{possibility} of partitions, a
+system must be prepared to either lose availability of operations or
+its ability to reason about global state.
 
-Network partitions happen at Google~\cite{dean-keynote}.
+For networks that do not partition, maintaining semantic guarantees
+and available operation is
+simple~\cite{stonebraker2010errors}. However, this is often not the
+case: according to James Hamilton, Vice President and Distinguished
+Engineer on the Amazon Web Service team, ``network partitions should
+be rare but net gear continues to cause more issues than it
+should''~\cite{hamilton-partitions}. Anecdotal evidence from companies
+such as Google~\cite{dean-keynote} confirm Hamilton's statement, while
+a host of recent and highly public cloud computing incidents highlight
+partition behavior. In April 2011, a network misconfiguration led to a
+twelve-hour long series of outages across the Amazon EC2 and RDS
+services~\cite{amazon-netpartition}. Subsequent misconfigurations and
+partial failures such as another EC2 failure in October 2012 have led
+to full site outages for web services like Reddit, Foursquare, and
+Heroku~\cite{ec2-downsites}. At a global scale, hardware
+failures---like the 2011 outages in Level3's backbones in North
+America and Europe due a 2011 bug in Juniper's
+routers~\cite{juniper-partition}---and misconfigurations---such as BGP
+faults introduced by Pakistan in 2008~\cite{pakistan-youtube} and
+academics in 2010~\cite{research-experiment-partition}---can cause
+widespread---even Internet-wide---partitions. Many of our discussions
+with practitioners---especially those operating on public cloud
+infrastructure---confirm that partitions are a reality.
 
-Many anecdotes online: Amazon's EC2 and RDS down due to net
-partition~\cite{amazon-netpartition}, as a result of AZ failures full
-site outages like Reddit, Foursquare, and Heroku~\cite{ec2-downsites},
-accounts of network partitions~\cite{netpartition-anecdote1}
+Several recent studies quantify partition behavior. A 2011 study of
+several Microsoft datacenters found a mean of 40.8 network link
+failures per day (95th percentile: 136), with a median time to repair
+of around five minutes, and up to one week. The annual probability of
+a device failing and impacting the network is near 10\% for several
+critical components like core routers and primary-backup connectors,
+while network redundancy only reduces the median impact of failures by
+up to 40\%.~\cite{sigcomm-dc}. A 2010 study of five years of over 200
+wide-area routers found an average of 16.2--302.0 failures per link
+per year with an average annual downtime of 24--497 minutes (95th
+percentile at least 34 hours) per link per year~\cite{sigcomm-wan}. In
+HP's managed enterprise networks, WAN, LAN, and connectivity problems
+account for 28.1\% of all customer support tickets while 39\% of
+tickets relate to network hardware.  The median incident duration for
+highest priority WAN, LAN, and configuration-related tickets is ranges
+from 114---165 minutes for WAN and LAN issues and up to a day for all
+incidents.~\cite{turner2012failure}. The incidence of network failures
+is further confirmed by several prior
+studies~\cite{labovitz-failures}, showing for example, on Sprint's
+WANs, a median time to repair between 2 and 1000 seconds, and a median
+time between failures of approximately 3000
+seconds.~\cite{ip-backbone-failures}. Indeed, isolating, quantifying,
+and designing for network failures is an area of active research in
+networking community~\cite{surviving-failures-bodik,
+  uw-failure-networks}.
 
-Despite speculation over partition
-frequency~\cite{stonebraker2010errors}, many practitioners indicate
-that partitions do indeed occur: in the words of James Hamilton, Vice
-President and Distinguished Engineer on the Amazon Web Services team,
-``network partitions should be rare but net gear continues to cause
-more issues than it should''~\cite{hamilton-partitions}
+These results---which do not account for server failures (another form
+of system partition that has also been quantified at
+scale~\cite{google-availability})---indicate that partitions
+\textit{do} occur within and across modern datacenters. We do not
+intend to fixate on the particular statistical behavior but instead
+observe that, as a general trend, partitions exist and must
+accordingly be met with either unavailability at some sites or, as we
+will discuss, relaxed semantic guarantees.
 
-Many of our discussions with practitioners---especially those
-operating on public cloud infrastructure---confirm that partitions are
-a reality.
+\subsection{Latency}
 
-Area of active research in networking
-community~\cite{surviving-failures-bodik}
+Even with fault-free networks, distributed systems face the challenge
+of communication latency, Deutsch's second ``Fallacy.'' Fundamentally,
+the speed at which two servers can communicate is (according to modern
+physics) bounded by the speed of light. Two servers located on
+opposite sides of the Earth face, in the best case---communicating via
+a hypothetical link crossing through the center of the
+planet---require a minimum 85.1 ms RTT (133.7 ms if sent at ground
+level). Of course, not every operation requires contacting another
+server around the world, but, as services are replicated to multiple,
+geographically distinct sites, the cost of communication grows.
 
-Over WAN:
+In real deployments, latencies are higher than the speed of light due
+to routing, congestion, and processing times; if not, servers a
+kilometer apart would enjoy a 6.7 $\mu$s RTT. To illustrate the
+difference between intra-datacenter, inter-datacenter, and
+inter-planetary networks, we performed a measurement study of network
+behavior on Amazon's EC2, a widely-used public compute cloud. We
+measured one week of ping times between all seven EC2 regions, across
+three ``availability zones'', or closely co-located datacenters, in
+the Eastern United States, and within a single ``availabilty zone,''
+at a granularity of 1s (dataset will be released).
 
-From 2002 study of Sprint's networks, median time to repair is between 2 and < 1000 seconds, approximately 2000 seconds between router failures and around 8,000 seconds between optical failures.
-Median failure rate is one every 3000 seconds.~\cite{ip-backbone-failures}.
+We summarize the results of our network measurement study in
+Table~\ref{table:rtt}. On average, inter-datacenter communication is
+between 1.82 and 6.38 times faster than across geographically
+co-located datacenters and 40 and 647 times faster than across
+geographically distributed datacenters. The cost of wide-area
+communication exceeds the speed of light: for example, communicating
+from S\~{a}o Paulo to Singapore is lower-bounded by speed-of-light
+transit RTT of 106.7 ms but, on average, incurs a 362.8 ms RTT (95th
+percentile: 649ms). This is perhaps unsurprising due to modern
+Internet topologies, but this three-fold overhead is
+non-negligible. As shown in Figure~\ref{fig:rtt}, the distribution of
+latencies varies between links, but the trend is clear: remote
+communication has a substantial cost.
 
-From 2002 study of VoIP, 50 minute failure caused 100\% packet loss between east and west coasts~\cite{voip-partitions}.
+\definecolor{min-lat-color}{HTML}{B2FF99}
+\definecolor{max-lat-color}{HTML}{FF7F7F}
+
+\begin{table}
+\subfloat[Cross-region (OR:~Oregon, VA:~Virginia, TO:~Tokyo, IR:~Ireland, SY:~Sydney, SP:~S\~{a}o Paulo, SI:~Singapore)] {
+  \begin{tabular}{|c|c|c|c|c|c|c|c|c|}
+\hline
+& \multicolumn{1}{c}{OR} & \multicolumn{1}{c}{VA} & \multicolumn{1}{c}{TO} & \multicolumn{1}{c}{IR} & \multicolumn{1}{c}{SY} & \multicolumn{1}{c}{SP} & \multicolumn{1}{c|}{SI} \\\hline
+CA & \colorbox{min-lat-color}{22.5}   & 84.5   & 143.7   & 169.8   & 179.1   & 185.9   & 186.9  \\
+OR &  & 82.9   & 135.1   & 170.6   & 200.6   & 207.8   & 234.4  \\
+VA & &  & 202.4   & 107.9   & 265.6   & 163.4   & 253.5  \\
+TO & & &  & 278.3   & 144.2   & 301.4   & 90.6  \\
+IR & & & &  & 346.2   & 239.8   & 234.1  \\
+SY & & & & &  & 333.6   & 243.1  \\
+SP & & & & & &  & \colorbox{max-lat-color}{362.8}  \\
+\hline
+  \end{tabular}
+}\vspace{.5em}
+
+\subfloat[Across \texttt{us-east} AZs]{
+  \makebox[.2\textwidth]{
+    \begin{tabular}{|c|c|c|}\hline
+ & \multicolumn{1}{c}{C} & \multicolumn{1}{c|}{D}\\\hline
+B & \colorbox{min-lat-color}{1.08} & 3.12 \\
+C & & \colorbox{max-lat-color}{3.57}  \\
+\hline
+  \end{tabular}}
+ }
+\subfloat[Within \texttt{us-east-a} AZ] {
+  \makebox[.25\textwidth]{
+  \begin{tabular}{|c|c|c|}\hline
+ & \multicolumn{1}{c}{H2} & \multicolumn{1}{c|}{H3}\\\hline
+H1  & 0.55   & \colorbox{max-lat-color}{0.56} \\
+H2 &  & \colorbox{min-lat-color}{0.50}  \\
+\hline
+  \end{tabular}
+ }}
+\caption{Mean RTT times on EC2 (min and max highlighted)}
+\label{table:rtt}
+\end{table}
+
+\begin{figure}
+\includegraphics[width=\columnwidth]{graphs/ping-plot.pdf}
+\caption{CDF of round-trip times for slowest inter- and intra-
+  availability zone links compared to cross-region links.}
+\label{fig:rtt}
+\end{figure}
 

### 2013-03-18 thebes-code/common/src/main/java/edu/berkeley/thebes/common/data/DataItem.java
Dependency resolution overhaul

@@ -4,11 +4,12 @@ import java.nio.ByteBuffer;
 import java.util.List;
 
 import com.google.common.base.Objects;
+import com.google.common.collect.ComparisonChain;
 import com.google.common.collect.Lists;
 
 import edu.berkeley.thebes.common.thrift.ThriftDataItem;
 
-public class DataItem {
+public class DataItem implements Comparable<DataItem> {
 	private final ByteBuffer data;
 	private Version version;
 	private List<String> transactionKeys;
@@ -86,4 +87,12 @@ public class DataItem {
 				Objects.equal(getVersion(), di.getVersion()) &&
 				Objects.equal(getTransactionKeys(), di.getTransactionKeys());
 	}
+
+    @Override
+    public int compareTo(DataItem o) {
+        return ComparisonChain.start()
+                .compare(version, version)
+                .compare(data, data)
+                .result();
+    }
 }

### 2013-03-18 thebes-code/common/src/main/java/edu/berkeley/thebes/common/data/Version.java

@@ -1,6 +1,7 @@
 package edu.berkeley.thebes.common.data;
 
 import com.google.common.base.Objects;
+import com.google.common.collect.ComparisonChain;
 import com.google.common.primitives.Longs;
 import com.google.common.primitives.Shorts;
 
@@ -9,6 +10,7 @@ import edu.berkeley.thebes.common.thrift.ThriftVersion;
 public class Version implements Comparable<Version> {
     public static final Version NULL_VERSION = new Version((short) -1, -1);
     
+    // TODO: Add logical clock so we can send multiple puts at the same physical timestamp
 	private final short clientID;
 	private final long timestamp;
 	
@@ -38,9 +40,10 @@ public class Version implements Comparable<Version> {
 
 	@Override
 	public int compareTo(Version other) {
-		int timestampCompare = Longs.compare(timestamp, other.getTimestamp());
-		int clientIdCompare = Shorts.compare(clientID, other.getClientID());
-		return timestampCompare != 0 ? timestampCompare : clientIdCompare; 
+	    return ComparisonChain.start()
+	            .compare(timestamp, other.getTimestamp())
+	            .compare(clientID, other.getClientID())
+	            .result();
 	}
 	
 	@Override

### 2013-03-18 thebes-code/hat/common/src/main/java/edu/berkeley/thebes/hat/common/thrift/AntiEntropyService.java

@@ -36,9 +36,7 @@ public class AntiEntropyService {
 
     public void put(String key, edu.berkeley.thebes.common.thrift.ThriftDataItem value) throws org.apache.thrift.TException;
 
-    public void waitForTransactionalDependencies(List<DataDependencyRequest> dependencyRequests) throws org.apache.thrift.TException;
-
-    public void receiveTransactionalDependencies(List<Long> fulfilledRequestIds) throws org.apache.thrift.TException;
+    public void ackDependentWriteInPending(String myKey, String ackedKey) throws org.apache.thrift.TException;
 
   }
 
@@ -46,9 +44,7 @@ public class AntiEntropyService {
 
     public void put(String key, edu.berkeley.thebes.common.thrift.ThriftDataItem value, org.apache.thrift.async.AsyncMethodCallback<AsyncClient.put_call> resultHandler) throws org.apache.thrift.TException;
 
-    public void waitForTransactionalDependencies(List<DataDependencyRequest> dependencyRequests, org.apache.thrift.async.AsyncMethodCallback<AsyncClient.waitForTransactionalDependencies_call> resultHandler) throws org.apache.thrift.TException;
-
-    public void receiveTransactionalDependencies(List<Long> fulfilledRequestIds, org.apache.thrift.async.AsyncMethodCallback<AsyncClient.receiveTransactionalDependencies_call> resultHandler) throws org.apache.thrift.TException;
+    public void ackDependentWriteInPending(String myKey, String ackedKey, org.apache.thrift.async.AsyncMethodCallback<AsyncClient.ackDependentWriteInPending_call> resultHandler) throws org.apache.thrift.TException;
 
   }
 
@@ -85,28 +81,17 @@ public class AntiEntropyService {
       sendBase("put", args);
     }
 
-    public void waitForTransactionalDependencies(List<DataDependencyRequest> dependencyRequests) throws org.apache.thrift.TException
+    public void ackDependentWriteInPending(String myKey, String ackedKey) throws org.apache.thrift.TException
     {
-      send_waitForTransactionalDependencies(dependencyRequests);
+      send_ackDependentWriteInPending(myKey, ackedKey);
     }
 
-    public void send_waitForTransactionalDependencies(List<DataDependencyRequest> dependencyRequests) throws org.apache.thrift.TException
+    public void send_ackDependentWriteInPending(String myKey, String ackedKey) throws org.apache.thrift.TException
     {
-      waitForTransactionalDependencies_args args = new waitForTransactionalDependencies_args();
[…]

### 2013-03-18 thebes-code/hat/server/src/main/java/edu/berkeley/thebes/hat/server/antientropy/AntiEntropyServiceHandler.java

@@ -9,7 +9,9 @@ import org.apache.thrift.TException;
 import org.slf4j.Logger;
 import org.slf4j.LoggerFactory;
 
+import edu.berkeley.thebes.common.config.ConfigParameterTypes.PersistenceEngine;
 import edu.berkeley.thebes.common.data.DataItem;
+import edu.berkeley.thebes.common.persistence.IPersistenceEngine;
 import edu.berkeley.thebes.common.thrift.ThriftDataItem;
 import edu.berkeley.thebes.hat.common.data.DataDependency;
 import edu.berkeley.thebes.hat.common.thrift.AntiEntropyService;
@@ -20,52 +22,30 @@ public class AntiEntropyServiceHandler implements AntiEntropyService.Iface {
     
     DependencyResolver dependencyResolver;
     AntiEntropyServiceRouter router;
+    IPersistenceEngine persistenceEngine;
 
-    public AntiEntropyServiceHandler(AntiEntropyServiceRouter router, DependencyResolver dependencyResolver) {
+    public AntiEntropyServiceHandler(AntiEntropyServiceRouter router,
+            DependencyResolver dependencyResolver, IPersistenceEngine persistenceEngine) {
         this.dependencyResolver = dependencyResolver;
         this.router = router;
-        startFulfillRequestThread();
+        this.persistenceEngine = persistenceEngine;
     }
 
     @Override
     public void put(String key,
-                    ThriftDataItem value) throws TException{
-    	logger.trace("Received anti-entropy put for key!");
-        dependencyResolver.asyncApplyNewWrite(key,
-                                              DataItem.fromThrift(value));
-    }
-
-    private LinkedBlockingQueue<DataDependencyRequest> requestsToFulfill = new LinkedBlockingQueue<DataDependencyRequest>();
-
-    // todo: make this multi-threaded/not stupid
-    private void startFulfillRequestThread() {
-        (new Thread(new Runnable() {
-            @Override
-            public void run() {
-                while(true) {
-                    try {
-                        DataDependencyRequest request = requestsToFulfill.take();
-//                        logger.debug("Waiting for atomic dependency...");
-                        dependencyResolver.blockForAtomicDependency(DataDependency.fromThrift(request.getDependency()));
-//                        logger.debug("Waiting FULFILLED for atomic dependency...");
-                        router.sendDependencyResolved(request);
-                    } catch(Exception e) {
-                        logger.error("error: ", e);
-                    }
-                }
[…]

### 2013-03-18 thebes-code/hat/server/src/main/java/edu/berkeley/thebes/hat/server/antientropy/clustering/AntiEntropyServiceRouter.java

@@ -1,251 +1,148 @@
 package edu.berkeley.thebes.hat.server.antientropy.clustering;
 
 
-import com.google.common.collect.Maps;
-import edu.berkeley.thebes.hat.common.data.DataDependency;
-import edu.berkeley.thebes.hat.common.thrift.DataDependencyRequest;
-import edu.berkeley.thebes.hat.server.dependencies.DependencyResolver;
+import org.apache.thrift.TException;
 import org.apache.thrift.transport.TTransportException;
 import org.slf4j.Logger;
 import org.slf4j.LoggerFactory;
 
 import com.google.common.collect.Lists;
+import com.google.common.collect.Queues;
+import com.google.common.util.concurrent.Uninterruptibles;
 
 import edu.berkeley.thebes.common.clustering.RoutingHash;
 import edu.berkeley.thebes.common.config.Config;
 import edu.berkeley.thebes.common.thrift.ServerAddress;
+import edu.berkeley.thebes.common.thrift.ThriftDataItem;
 import edu.berkeley.thebes.hat.common.thrift.AntiEntropyService;
 import edu.berkeley.thebes.hat.common.thrift.ThriftUtil;
+import edu.berkeley.thebes.hat.server.dependencies.PendingWrite;
 
-import java.util.ArrayList;
 import java.util.List;
-import java.util.Map;
 import java.util.concurrent.LinkedBlockingQueue;
-import java.util.concurrent.Semaphore;
+import java.util.concurrent.TimeUnit;
 
 public class AntiEntropyServiceRouter {
     private static Logger logger = LoggerFactory.getLogger(AntiEntropyServiceRouter.class);
 
+    /** Siblings replicate the same data. */
+    private static List<AntiEntropyService.Client> siblingClients = Lists.newArrayList();
+    /** Neighbors live in the same cluster. Includes self! */
     private static List<AntiEntropyService.Client> neighborClients = Lists.newArrayList();
-    private static List<AntiEntropyService.Client> waitClusterClients = Lists.newArrayList();
-    private static List<AntiEntropyService.Client> notifyClusterClients = Lists.newArrayList();
     private static int numServersInCluster;
 
     public void bootstrapAntiEntropyRouting() throws TTransportException {
         if (Config.isStandaloneServer()) {
             logger.debug("Server marked as standalone; not starting anti-entropy!");
-            return;
+            // TODO: Fix this.
[…]

### 2013-03-18 thebes-code/hat/server/src/main/java/edu/berkeley/thebes/hat/server/dependencies/DependencyResolver.java

@@ -1,7 +1,12 @@
 package edu.berkeley.thebes.hat.server.dependencies;
 
+import java.util.Iterator;
 import java.util.List;
 import java.util.Map;
+import java.util.Set;
+import java.util.concurrent.ConcurrentHashMap;
+import java.util.concurrent.ConcurrentMap;
+import java.util.concurrent.ConcurrentSkipListSet;
 import java.util.concurrent.Semaphore;
 import java.util.concurrent.TimeUnit;
 import java.util.concurrent.atomic.AtomicInteger;
@@ -13,12 +18,15 @@ import org.apache.thrift.async.AsyncMethodCallback;
 import org.slf4j.Logger;
 import org.slf4j.LoggerFactory;
 
+import com.google.common.collect.ComparisonChain;
 import com.google.common.collect.Lists;
 import com.google.common.collect.Maps;
+import com.google.common.collect.Sets;
 import com.yammer.metrics.Metrics;
 import com.yammer.metrics.core.Timer;
 import com.yammer.metrics.core.TimerContext;
 
+import edu.berkeley.thebes.common.config.ConfigParameterTypes.PersistenceEngine;
 import edu.berkeley.thebes.common.data.DataItem;
 import edu.berkeley.thebes.common.data.Version;
 import edu.berkeley.thebes.common.persistence.IPersistenceEngine;
@@ -56,188 +64,77 @@ waitInQueueForDependency() need to be notified and so
 notifyNewLocalWrite() is called.
 */
 
-public class DependencyResolver {
-
-    private final Timer waitingDependencyTimerMetric = Metrics.newTimer(DependencyResolver.class,
-                                                                        "hat-dependencies-check-waiting",
-                                                                        TimeUnit.MILLISECONDS,
-                                                                        TimeUnit.SECONDS);
-
-    private final Timer resolvingDependencyTimerMetric = Metrics.newTimer(DependencyResolver.class,
-                                                                          "hat-dependencies-resolving",
-                                                                          TimeUnit.MILLISECONDS,
-                                                                          TimeUnit.SECONDS);
-
-    private final Timer resolvingAtomicDependencyTimerMetric = Metrics.newTimer(DependencyResolver.class,
-                                                                          "hat-dependencies-resolving-atomic",
-                                                                          TimeUnit.MILLISECONDS,
-                                                                          TimeUnit.SECONDS);
-
-    private class DependencyWaitingQueue {
[…]

### 2013-03-18 thebes-code/hat/server/src/main/java/edu/berkeley/thebes/hat/server/replica/ReplicaServiceHandler.java

@@ -1,8 +1,8 @@
 package edu.berkeley.thebes.hat.server.replica;
 
-import java.util.List;
-
 import org.apache.thrift.TException;
+import org.slf4j.Logger;
+import org.slf4j.LoggerFactory;
 
 import edu.berkeley.thebes.common.data.DataItem;
 import edu.berkeley.thebes.common.data.Version;
@@ -10,58 +10,59 @@ import edu.berkeley.thebes.common.persistence.IPersistenceEngine;
 import edu.berkeley.thebes.common.thrift.ThriftDataItem;
 import edu.berkeley.thebes.common.thrift.ThriftVersion;
 import edu.berkeley.thebes.hat.common.thrift.ReplicaService;
-
-import edu.berkeley.thebes.hat.server.antientropy.AntiEntropyServer;
+import edu.berkeley.thebes.hat.server.antientropy.clustering.AntiEntropyServiceRouter;
 import edu.berkeley.thebes.hat.server.dependencies.DependencyResolver;
-import edu.berkeley.thebes.hat.server.dependencies.PendingWrites;
-import org.slf4j.Logger;
-import org.slf4j.LoggerFactory;
 
 public class ReplicaServiceHandler implements ReplicaService.Iface {
     private IPersistenceEngine persistenceEngine;
-    private AntiEntropyServer antiEntropyServer;
+    private AntiEntropyServiceRouter antiEntropyRouter;
     private DependencyResolver dependencyResolver;
-    private PendingWrites pendingWrites;
     private static Logger logger = LoggerFactory.getLogger(ReplicaServiceHandler.class);
 
 
     public ReplicaServiceHandler(IPersistenceEngine persistenceEngine,
-                                 PendingWrites pendingWrites,
-                                 AntiEntropyServer antiEntropyServer,
+                                 AntiEntropyServiceRouter antiEntropyRouter,
                                  DependencyResolver dependencyResolver) {
         this.persistenceEngine = persistenceEngine;
-        this.pendingWrites = pendingWrites;
-        this.antiEntropyServer = antiEntropyServer;
+        this.antiEntropyRouter = antiEntropyRouter;
         this.dependencyResolver = dependencyResolver;
     }
 
     @Override
     public boolean put(String key,
-                       ThriftDataItem value) throws TException {
+                       ThriftDataItem valueThrift) throws TException {
+        DataItem value = DataItem.fromThrift(valueThrift);
         if(logger.isTraceEnabled())
             logger.trace("received PUT request for key: '"+key+
                          "' value: '"+value+
[…]

### 2013-03-18 writeup/hat-analysis.tex
text
checkpoint writing

@@
+\section{Highly Available Transactions}
+\label{sec:hats}
+
+\begin{quote}
+``...one should not throw out the C so quickly, since there are real
+  error scenarios where CAP does not apply and it seems like a bad
+  tradeoff in many of the other situations.''---Michael
+  Stonebraker~\cite{stonebraker2010errors}
+\end{quote}
+
+While $1$-available and $1$-sticky available systems provide several
+benefits, they come with a cost: the semantic guarantees that they are
+able to offer are reduced as a result of their coordination penalty.
+
+What about existing systems?
+* Database technology developed for single-node systems; gold standard: serializability
+	* Serializability is not actually highly available; give an example!
+	* Much of the database literature presumes serializability and does not consider high availability.
+	* For the literature that doesn't presume serializability, it's not presented in a HA context?
+* Consequence: traditional systems are not optimized for high availability
+	* Consider two-phase locking in a distributed environment
+	* As we will see (forward reference to Evaluation), many existing transactional systems encounter similar difficulties
+* Key question in this paper: What transactional semantics are highly available, and which aren't?
+	* Side note: there are infinite incomparable consistency models (e.g., always return 2, always return 3, …)
+	* Our goal is to unify distributed systems literature with ACID transaction model.
+* Argument: Highly Available Transactional Systems (HATS) require rethinking existing models. This work is a first step.
+
+In this sense, HATs are similar to RAID: optimizing for graceful
+handling of a worst-case failure scenario improves average-case
+performance.
+
 IV. HATS and ACID (4pg)
@@
 
+Stickiness and PNUTS
+* As an example of a real system going from "CP" to "AP", [PNUTS dropped per-record mastering](\url{http://developer.yahoo.com/blogs/ydn/posts/2010/06/sherpa_update/#4})
+
 * Composition of Properties can be tricky:

### 2013-03-18 writeup/model.tex
checkpoint writing

--- 2013-03-17 writeup/model.tex
@@
-\section{What is High Availability?}
+\section{High Availability}
+\label{sec:availability}
 
-* System model
-	* Consider transactions that issue a series of reads and writes, then commit or abort
-	* Each data item is located on one or more servers
+Given the presence of network partitions and latency, systems
+designers have a choice: pay the cost of possible unavailability and
+potentially expensive network round-trips or seek alternative
+algorithms that avoid them. Before we address the \textit{cost} of
+avoiding unavailability and latency, we first need a way to classify
+whether an algorithm actually incurs these penalties.
 
-* HA Definitions
-	* Straw-man definition: every transaction will recieve a response
-		* Problem: system that always aborts is trivially available!
+Highly available algorithms ensure ``always on'' operation and promise
+a guaranteed response. If users of a highly available system contact a
+(set of) replicas, then they will be \textit{available}, or able to
+execute operations without having to block when replicas cannot
+communicate. A useful corollary to providing high availability is
+that, in the absence of failures, the system can provide a response
+without having to contact other replicas and incur high
+latency~\cite{abadi-pacelc}. In this section, we develop a model and
+taxonomy for high availability in its various forms. This is
+challenging for several reasons, including handling fault-tolerance
+and, in a transactional context, handling aborts.
 
-* Replica-HA
-	* (Handwaving in the outline) If a client can contact at least one replica for each data item in a transaction, then the transaction will eventually commit, even in the presence of infinite network partitions.
+\subsection{Replica Availability}
 
-* Sticky-HA
-	* If a client continues to contact the same set of replicas across transactions and the client can contact at least one replica for each data item in a transaciton, then the transaction will eventually commit, even in the presence of infinite network partitions.
-	* We will use this definition when it comes to causality and read your writes guarantees!
-	* Can be achieved by either "sticking" clients to the same set of servers or by having the client act as a replica itself, via caching
+Traditionally, a system provides high availability if every user that
+can contact a server eventually receives a response from that server,
+even in the presence of arbitrary, indefinite network partitions
+between servers~\cite{gilbert-cap}. This is a useful definition: if we
+can contact at least one server, then we can receive a
+response. However, there are several other cases we should consider.
+
+First, a client may wish to ensure that the effects of its operations
+will persist in the face of server failure. If a client wants to
+tolerate one server fault, then it should ensure that its operations
+reach at least two replicas before it considers the operation
+completed. According to our above definition, this $1$-fault tolerance
+is not highly available; in fact, according to the above definition,
+high availability is at odds with fault-tolerance! Instead, we propose
+\textit{$K$-availability}: if a client can contact at least $K$
+servers, then it eventually receives a response from all of
+them. $1$-fault tolerance requires $2$-availability, while the
+traditional ``high availability'' is equivalent to
+$1$-availability. $2$-availability will require more messaging and
+possibly, depending on server layout, higher latency and more
+unsuccessful operations in the presence of network and server
+failures.
+
+Second, clients may wish to ensure available operation in the presence
+of a minority of servers failing. This requirement is relatively
+common in standard consensus algorithms like two-phase commit and
+Paxos and is, by traditional standards, hardly ``highly available.''
+However, it is worth acknowledging that, in an $N$ server cluster,
+this \textit{majority availability} (a special case of
+$K$-availability with $K=\lceil \frac{N}{2} \rceil$) provides more
+frequent responses and lower latency than $N$-availability. A key
+differentiation between generic $K$-availability and majority
+availability is that $K$ availability is independent of the number of
+servers under consideration. Fault-tolerance effectively has
+``constant'' scaling cost, whereas majority availability based
+algorithms are not as fortunate.
+
+Third, clients may wish to contact the same server (or set of servers)
+across subsequent operations. As we will discuss in
+Section~\ref{sec:hat}, clients can ensure continuity between
+operations (e.g., reading their prior updates to a data item) by
+maintaining affinity or ``stickiness'' with a server or set of
+servers~\cite{vogels-defs}. Ultimately, maintaining availability
+during stickiness requires fate-sharing between the client and its set
+of servers: if the servers a client continues to contact fail, or,
+equivalently, if the client is partitioned from its designated
+servers, then it may not be able to achieve its desired semantic
+guarantees. We say that a system provides \textit{sticky availability}
+for a user if a client's subsequent operations are executed against
+the same server and, whenever the client can contact the server, it
+eventually receives a response. As above, it may be useful for a
+client to achieve stickiness \textit{and} fault tolerance, so we can
+generalize sticky availability to \textit{$K$-sticky availability},
+which provides sticky availability with $K$ distinct servers. In
+certain cases, a client may wish to become $1$-sticky available by
+acting as a server itself; for example, in a data store, a client
+might cache its reads and writes. However, a single client cannot
+achieve $K$-sticky availability with $K>1$ on its own. ``Sticking''
+all clients to a designated set of servers (i.e., $K$-master
+availability) is a special case of sticky availability and is less
+available.
+
+\begin{figure}
+\centering
+\begin{tikzpicture}[scale=0.8]
+  \tikzstyle{every node}=[font=\small]
+ \node[draw=none,fill=none] (1) at (0,0) {$1$-availability}; 
+ \node[draw=none,fill=none] (2) at (0,1) {$2$-availability}; 
+ \node[draw=none,fill=none] (m-1) at (0, 2) {\ldots}; 
+ \node[draw=none,fill=none] (m) at (0, 3) {majority availability}; 
+ \node[draw=none,fill=none] (m+1) at (0, 4) {\ldots}; 
+ \node[draw=none,fill=none] (n) at (0, 5) {$N$-availability}; 
+ \node[draw=none,fill=none] (1s) at (4, 1) {$1$-sticky availability};
+ \node[draw=none,fill=none] (2s) at (4, 2) {$2$-sticky availability};
+ \node[draw=none,fill=none] (n2s) at (4, 3) {\ldots};
+ \node[draw=none,fill=none] (n1s) at (4, 4) {$(N$$-$$1)$-sticky availability};
+
+ \draw [->] (1) -- (2);
+ \draw [->] (2) -- (m-1);
+ \draw [->] (m-1) -- (m);
+ \draw [->] (m) -- (m+1);
+ \draw [->] (m+1) -- (n);
+ \draw [->] (1) -- (1s);
+ \draw [->] (1s) -- (2s);
+ \draw [->] (2) -- (2s);
+ \draw [->] (2s) -- (n2s);
+ \draw [->] (n2s) -- (n1s);
+ \draw [->] (n1s) -- (n);
+
+\end{tikzpicture}
+\label{fig:availability-order}
+\caption{Hierarchy of replica availability levels for $N>3$ servers.}
+\end{figure}
+
+We show a hierarchy of replica availabilty levels in
+Figure~\ref{fig:availability-order}. $K$-availability is ``more
+available'' than $K$-sticky available, but, in general, $K$-sticky
+availability is incomparable with $(K+1)$-sticky
+availability. However, $N$-sticky availability is effectively the same
+as $N$-availability--if operations must be performed on all replicas,
+then subsequent requests will trivially access the same set of
+replicas.
+
+There are three further aspects of replica availability that deserve
+mention. First, availability may be decomposed into operation-specific
+levels. For instance, many quorum-based replication systems allow for
+tuning based on workload and desired performance; a read-one,
+write-all (ROWA) quorum-replicated data store would experience
+$N$-availability for writes but $1$-availability for reads. Second, we
+have assumed that server membership in the system is static. In
+practice, servers may join and leave a deployment over time. Aside
+from setting $N=\infty$, systems may wish to reevaluate their
+algorithm choice, particularly for $N$ and majority available
+algorithms. Third, in many deployments, such as a data store, a given
+set of operations may be logically restricted to a subset of the
+servers in the cluster. In a partitioned database, the only servers
+who perform read and write operations are typically those that are
+designed as replicas for a given item. Accordingly, $N$ is determined
+by the replication factor of the data store (e.g., three to five) as
+opposed to the number of servers comprising the data store (e.g., tens
+to thousands). In this paper, we do not focus on further
+taxonomization of replica availability (e.g., within versus across
+datacenters) but believe there is worthwhile future work in doing so.
+
+\subsection{Transactional Availability}
+
+Thus far, we have focused on single-operation availability. This is
+standard in distributed systems literature (e.g., atomic, regular, and
+safe register models all concern single objects), yet the database
+literature largely focuses on transactions: groups of multiple
+operations over multiple objects. Replica availability alone is
+insufficient to describe availability guarantees for
+transactions. Additionally, given the choice of \textit{commit} and
+\textit{abort} responses---which signal transaction success or failure
+to a client---we must be careful in defining availability.
+
+First, if a client wishes to execute a transaction that performs
+operations on $D$ distinct data items that are spread across servers,
+then it must be able to contact servers for those data items. This is
+different from, say, $1$-replica availability or $D$-replica
+availability; instead, the client requires a degree of replica
+availability for \textit{each} respective data item. This this
+\textit{transactional requirement for replica availability} may result
+in lower availability than a non-transactional requirement (e.g.,
+requesting from two servers instead of one for a two-item transaction)
+but not necessarily (e.g., both items in a two-item transaction are
+located on the same server).
+
+Second, given the possiblity of aborts, we need to ensure useful
+forward progress. A system can trivially guarantee clients a response
+by always aborting their
+transactions~\cite{transaction-liveness}. However, this is an
+unsatisfactory system because nothing good (transaction commit) ever
+happens; instead, we require a \textit{liveness} property. We cannot
+guarantee that every transaction will commit---transactions may choose
+to abort themselves---but we need to make sure that the system will
+not indefinitely abort transactions on its own volition. We call a
+transaction abort due to a transaction's own choosing (e.g., as part
+of the transaction itself or due to a would-be integrity constraint
+violation) an \textit{internal abort} and an abort due to system
+implementation an \textit{external abort}.
+
+We say that a system provides \textit{transactional availability} if,
+given replica availability for every data item in a transaction, the
+transaction eventually commits or internally aborts. For example,
+system will violate transactional $2$-availability if client that can
+contact two servers for each of four data items in its transaction
+cannot eventually commit in the absence of internal aborts.
+
+\subsection{Discussion}
+
+In order to provide a framework in which to evaluate partition and
+latency sensitivity, we have introduced a taxonomy of replica and
+transactional availability requirements. The many variants of this
+taxonomy are interesting on their own, but, given our goal, two models
+stand out as particularly useful.
+
+$1$-availability is the standard model for ``high availability'' in a
+distributed environment. Any system purporting to be ``highly
+available'' today is---if it is
+honest---$1$-available. $1$-availability is particularly useful as it
+means that \textit{any} server (or, in the case of a database system,
+replica) can handle a request. This is particularly strong:
+availability is guaranteed unless a client is partitioned from
+\textit{all} servers and, aside from latency associated with
+contacting the server to begin with, requests can complete effectively
+locally. This mitigates many of the costs of the distributed
+model---albeit at the expense of semantic guarantees
+(Section~\ref{sec:hats}). $1$-sticky availability is also fairly
+common and is often cited as a technique for cheaply providing several
+useful guarantees, like the aforementioned ``read your writes''
+guarantees.
+
+In this work, we will focus on developing the theory of $1$-available
+and $1$-sticky available systems. While it is possible to provide low
+latency in addition to higher $K$-availability and $K$-sticky
+availabililty (e.g., for fault tolerance via increased server count
+within a given datacenter), we reserve this exploration for future work.

### 2013-03-18 thebes-code/hat/server/src/main/java/edu/berkeley/thebes/hat/server/antientropy/AntiEntropyServiceHandler.java

@@ -45,7 +45,9 @@ public class AntiEntropyServiceHandler implements AntiEntropyService.Iface {
                 while(true) {
                     try {
                         DataDependencyRequest request = requestsToFulfill.take();
+//                        logger.debug("Waiting for atomic dependency...");
                         dependencyResolver.blockForAtomicDependency(DataDependency.fromThrift(request.getDependency()));
+//                        logger.debug("Waiting FULFILLED for atomic dependency...");
                         router.sendDependencyResolved(request);
                     } catch(Exception e) {
                         logger.error("error: ", e);

### 2013-03-18 thebes-code/hat/server/src/main/java/edu/berkeley/thebes/hat/server/dependencies/DependencyResolver.java

@@ -180,6 +180,7 @@ public class DependencyResolver {
         }
 
         public void notifyResolved() {
+            logger.debug("Resolved!: " + waitingCount.get());
             if(this.waitingCount.decrementAndGet() == 0) {
                 persistenceEngine.put(key, write);
                 notifyNewLocalWrite(key, write);
@@ -193,6 +194,9 @@ public class DependencyResolver {
             AtomicInteger waitCount = new AtomicInteger(write.getTransactionKeys().size());
 
             for(String atomicKey : write.getTransactionKeys()) {
+                if (atomicKey.equals(key)) {
+                    logger.debug("Waiting on self");
+                }
                 router.waitForDependencyRemote(key,
                                                new DataDependency(atomicKey, write.getVersion()),
                                                new WaitingResolvedDependency(key,
@@ -206,6 +210,7 @@ public class DependencyResolver {
                                    final DataItem write) throws TException {
 
         if(write.getTransactionKeys() == null || write.getTransactionKeys().isEmpty()) {
+            logger.debug("Autoresolve!");
             persistenceEngine.put(key, write);
             notifyNewLocalWrite(key, write);
             return;

### 2013-03-18 thebes-code/scripts/common_funcs.py

@@ -6,11 +6,11 @@ from time import sleep
 
 def run_cmd(hosts, cmd, user="root"):
     print("parallel-ssh -t 1000 -O StrictHostKeyChecking=no -l %s -h hosts/%s.txt \"%s\"" % (user, hosts, cmd))
-    system("parallel-ssh -t 1000 -O StrictHostKeyChecking=no -l %s -h hosts/%s.txt \"%s\"" % (user, hosts, cmd))
+    system("parallel-ssh -t 1000 -O StrictHostKeyChecking=no -l %s -h hosts/%s.txt \"ulimit -u unlimited; %s\"" % (user, hosts, cmd))
 
 def run_cmd_single(host, cmd, user="root"):
     print("ssh -o StrictHostKeyChecking=no %s@%s \"%s\"" % (user, host, cmd))
-    system("ssh -o StrictHostKeyChecking=no %s@%s \"%s\"" % (user, host, cmd))
+    system("ssh -o StrictHostKeyChecking=no %s@%s \"ulimit -u unlimited; %s\"" % (user, host, cmd))
 
 def run_process_single(host, cmd, user="root", stdout=None, stderr=None):
     subprocess.call("ssh %s@%s \"%s\"" % (user, host, cmd),
@@ -26,6 +26,9 @@ def run_script(hosts, script, user="root"):
     run_cmd(hosts, "bash /tmp/%s" % (script.split("/")[-1]), user)
 
 def fetch_file_single(host, remote, local, user="root"):
+    system("scp %s@%s:%s '%s'" % (user, host, remote, local))
+
+def fetch_file_single_compressed(host, remote, local, user="root"):
     system("scp %s@%s:%s '%s'" % (user, host, remote, local))
 
 def get_host_ips(hosts):

### 2013-03-18 thebes-code/scripts/setup_hosts.py

@@ -8,6 +8,7 @@ from common_funcs import sed
 from common_funcs import upload_file
 from common_funcs import run_script
 from common_funcs import fetch_file_single
+from common_funcs import fetch_file_single_compressed
 from threading import Thread, Lock
 import os
 from datetime import datetime
@@ -82,10 +83,10 @@ class Host:
 
 # UTILITIES
 def run_cmd_in_thebes(hosts, cmd, user='root'):
-    run_cmd(hosts, "cd /home/ubuntu/thebes/thebes-code; %s" % cmd, user)
+    run_cmd(hosts, "cd /home/ubuntu/thebes/thebes-code; ulimit -u unlimited; %s" % cmd, user)
 
 def run_cmd_in_ycsb(hosts, cmd, user='root'):
-    run_cmd(hosts, "cd /home/ubuntu/thebes/ycsb-0.1.4; %s" % cmd, user)
+    run_cmd(hosts, "cd /home/ubuntu/thebes/ycsb-0.1.4; ulimit -u unlimited; %s" % cmd, user)
 
 
 def get_instances(regionName):
@@ -149,6 +150,7 @@ def get_num_nonterminated_instances(regionName):
 
     for line in open("/tmp/running.txt"):
         line = line.split()
+        print line
         if line[0] == "INSTANCE" and line[3] != "terminated":
             num_nonterminated = num_nonterminated + 1
         elif line[0] == "TAG" and line[3] in tag_blacklist:
@@ -156,6 +158,7 @@ def get_num_nonterminated_instances(regionName):
                 blacklisted_hosts.append(line[2])
 
     system("rm /tmp/running.txt")
+    print blacklisted_hosts
     return num_nonterminated-len(blacklisted_hosts)
 
 def make_instancefile(name, hosts):
@@ -203,7 +206,7 @@ def provision_clusters(regions, use_spot, anti_slow):
 
 def provision_graphite(region):
     global AMIs
-    if region == None:
+    if region == None or region=="":
         pprint('Graphite not enabled.')
 
     provision_instance(region.name, 1)
@@ -211,12 +214,12 @@ def provision_graphite(region):
 
 def provision_spot(regionName, num):
     global AMIs
-    system("ec2-request-spot-instances %s --region %s -t t1.micro -price 0.02 " \
+    system("ec2-request-spot-instances %s --region %s -t m1.xlarge -price 0.50 " \
[…]

### 2013-03-19 writeup/hat-analysis.tex
more writing

@@
 
-While $1$-available and $1$-sticky available systems provide several
-benefits, they come with a cost: the semantic guarantees that they are
-able to offer are reduced as a result of their coordination penalty.
+HAT systems offer substantial latency and availability benefits, yet
+they come with a cost. In the distributed systems community, there is
+a host of results that taxonomize the relationships between various
+distributed consistency models. Perhaps most famously, Brewer's CAP
+Theorem---as formally proven by Gilbert and
+Lynch~\cite{needed}---states that it is impossible to maintain
+linearizability (``C,'' or ``consistency''---albeit poorly named), or
+the ability to read the last completed write, and replica availability
+in the presence of failures~\cite{needed}. While the CAP Theorem has
+has roots in decades of distributed system designs~\cite{needed}, it
+has had large scale impact on the design and development of
+distributed systems, particularly for Internet
+services---unsurprisingly, the context in which Brewer formulated his
+finding.
+
+\subsection{ACID and Modern Databases}
+
+While linearizability is a useful property to consider, there are
+several other guarantees that are perhaps more relevant to a database
+system. In particular, the database tradition of providing ACID
+guarantees: atomicity, consistency, isolation, and durability are,
+with few exceptions, not addressed by the distributed systems
+literature. There are several factors behind this occurrence, the
+least of which is that distributed systems rarely consider
+multi-object guarantees, whereas ACID transactions are frequently used
+to provide semantic guarantees across multiple operations on multiple
+objects. While linearizability (unfortunately often called
+``atomicity'') ``composes'' across multiple objects, a linearizable
+system does not provide ACID ``atomicity'' (which we will call
+``transactional atomicity'' for clarity): just because two writes are
+visible to all writers immediately after they complete does not, on
+its own, provide any guarantees about the mutual visibility of the two
+writes.
+
+Database designers and researchers have long realized that
+serializability---the gold standard of ACID isolation, which
+guarantees application-level consistency---is not achievable in a
+highly available system. Indeed, decades of research on
+semantics-based concurrency control and coordination-reducing
+techniques such as escrow transactions and sagas allow serializability
+in the presence of partial system knowledge or otherwise reduce
+coordination cost. However, the majority of these algorithms are
+intended to preserve application-level consistency, which, for
+arbitrary applications and arbitrary sequence of operations, cannot be
+ensured with high availability in the presence of partitions.
+
+Of course, database systems offer a range of ACID properties besides
+serializability. Even on a single-node system, the coordination
+penalties associated with ensuring equivalence with a serial execution
+can be severe and manifest themselves in the form of concurrency
+control contention (and, subsequently, performance degradation,
+scalability limitations, and, often, external aborts like
+deadlocks). Instead, databases offer a host of so-called \textit{weak
+  isolation} models that allow varying restrictions on the space of
+schedules that are allowable by the system. None of these weak
+isolation models guarantees serializability, but the benefits of these
+models can outweigh the costs of possible application-level
+consistency anomalies that arise from their use.
+
+To understand the prevalence of these weak isolation models, we
+surveyed the default and maximum isolation models provided by 18
+relational databases, often claiming to provide ``ACID'' or ``NewSQL''
+functionality. As shown in Table~\ref{table:existing}, only three out
+of 18 databases provides serializability by default, and at least
+eight do not provide serializability as an option at all. This is
+particularly surprising when we consider the widespread deployment of
+many of these databases, like Oracle 11g, which are known to power
+major businesses and product functionality. We might view the use of
+weak consistency is somewhat of an ``arms race'': anecdotally, one
+major RDBMS vendor was forced to lower their default isolation level
+in order to compare favorably in out-of-the-box performance
+comparisons with its major competitor, which had already lowered its
+default setting. Applications can guard against inconsistency in a
+weakly isolated environment by performing their own locking external
+to the database or by performing ad-hoc concurrency control within the
+database. However, this appears error-prone and, without further
+evidence, it remains somewhat puzzling why this is indeed favorable to
+providing serializability with in the database itself.
+
+\begin{table}
+\begin{small}
+\begin{tabular}{|l|c|c|}
+\hline
+Database & Default & Maximum\\\hline
+Actian Ingres 10.0/10S~\cite{actian-docs} & S & S\\
+Aerospike~\cite{aerospike-docs} & RC & RC\\
+Akiban Persistit~\cite{akiban-docs} & SI & SI\\
+Clustrix CLX 4100~\cite{clustrix-docs} & RR & ?\\
+Greenplum 4.1~\cite{greenplum-docs} & RC & S \\
+IBM DB2 10 for z/OS~\cite{db2-docs} & CS & S\\
+IBM Informix 11.50~\cite{informix-docs} & Depends & RR\\
+MySQL 5.6~\cite{innodb-docs} & RR & S \\
+MemSQL 1b~\cite{memsql-docs} & RC & RC\\
+MS SQL Server 2012~\cite{ms-sql-docs} & RC & S \\
+NuoDB~\cite{nuodb-docs} & CR & CR\\
+Oracle 11g~\cite{oracle-docs} & RC & SI\\
+Oracle Berkeley DB~\cite{bdb-reg-docs} & S & S\\
+Oracle Berkeley DB JE~\cite{bdb-je-docs} & RR & S\\
+Postgres 9.2.2~\cite{postgres-docs} & RC & S\\
+SAP HANA~\cite{hana-docs} & RC & SI\\
+ScaleDB 1.02~\cite{scaledb-docs} & RC & RC\\
+VoltDB~\cite{voltdb-docs} & S & S\\
+\hline
+\multicolumn{3}{|p{7cm}|}{{\footnotesize RC: read committed, RR: repeatable read, SI: snapshot isolation, S: serializability, CS: cursor stability, CR: consistent read}}\\\hline
+
+\end{tabular}
+\caption{Default and maximum isolation levels for ACID and NewSQL
+  databases as of January 2013 (from
+  \protect\cite{needed-hotos}).}\vspace{-1.5em}
+\label{table:existing}
+\end{small}
+\end{table}
+
+Given that these transactional models are frequently used, our
+inability to provide serializability to provide general-purpose HATs
+appears not to be a show-stopper. If applications cope with
+inconsistency or, on a single node server, have already decided that
+the benefits of weak isolation outweigh potential application
+inconsistencies, then perhaps, in a highly available environment, some
+may make a similar decision given the appropriate choice of semantic
+guarantees.
+
+The primary challenge in providing weak isolation in a HAT context is
+that, unfortunately, it is unclear \textit{which} isolation guarantees
+can be provided with high availability. Standard discussions of weak
+isolation are presented in a single-node context---a possible
+consequence of their original context: a lock-based (and therefore
+unavailable, even in a distributed context) single-server
+deployment~\cite{needed}. We are not aware of any prior literature
+that provides guidance as to the relationship between weak isolation
+and high availability. Prior work has examined the relationship
+between serializability and high availability and weak isolation on a
+single-server but never weak isolation and high availability taken
+together. This apparent gap in the database literature is a possible
+contributing factor to the often-apparent attitude that
+general-purpose transactional semantics are either too expensive or
+too lacking in availability guarantees to provide in modern
+distributed databases---except under special circumstances, such as
+operations over groups of co-located data items.
+
+\subsection{Achievable HAT Semantics}
+
+In this paper, we will describe which ACID semantics are achievable in
+HATs as well as which are not. Our strategy is to explore successively
+more powerful models until we reach a frontier of semantic guarantees
+which are unavailable to HATs. To begin, we present achievable
+semantics, offering proof-of-concept algorithms to demonstrate
+feasibility. When possible, we draw on existing properties and
+definitions from the database and distributed systems literature,
+providing a brief, informal explanation of each guarantee and
+illustrate it with an example. For our ACID guarantees, we draw
+largely on Adya's thesis work on defining weak isolation
+models~\cite{needed} and, where appropriate, supplement with
+definitions and discussion from Berenson et al.'s SIGMOD 1995 critique
+of the ANSI SQL Specification~\cite{needed} (hereafter, Berenson) as
+well as the ANSI SQL specification~\cite{needed} (hereafter, ANSI)
+itself. For other guarantees, such as session properties, we draw on
+both the database and distributed systems literature. We provide a set
+of formal definitions and semantics in the Appendix.
+
+\subsubsection{ACID Isolation Guarantees}
+
+To begin, Adya's \textit{Read Uncommitted} isolation requires that
+transactions are totally ordered and that writes within transactions
+are ordered consistently with this order (prohibiting ``Dirty
+Writes,'' or $G0$)~\cite{adya}. If two transactions write to the same
+set of data items, then the final database state cannot contain the
+``earlier'' transaction's writes to the data items. For example, in
+the below example, $T_3$ should eventually only read $a=b=1$ or
+$a=b=2$ but not $a=2, b=1$:
+\begin{align*}
+\small
+T_1 &: w_x(1)~w_y(1)
+\\T_2 &: w_x(2)~w_y(2)
+\\T_3 &: r_x(a)~r_y(b)\\[-2em]
+\end{align*}
+The difference between this total order on transactions and the total
+order required by serializability is that the Read Uncommitted total
+order is completely arbitrary (i.e., an order must exist), whereas a
+serializable total order must be equivalent to a serial
+execution. Interestingly, Read Uncommitted is a close analog with
+eventually consistent systems's concept of replica convergence, in
+which all replicas for a data item must end up with the same data
+item. This convergence is usually achieved by picking a ``winning''
+update---which requires ordering---or by maintaining a growing a set
+of updates---whose cardinality implicitly defines an order (with
+convergence achieved when all updates are in the set); we can see Read
+Uncommitted as a cross-item convergence property. It is easily
+achieved via applying the same logical timestamp to each update in a
+transaction.
+
+\textit{Read Committed} isolation requires that transactions do not
+read uncommitted versions of data items (prohibiting both ``Dirty
+Writes''---as above---and ``Dirty Reads'' phenomena; captured by
+Adya's $G1\{a,b,c\}$, ANSI's $P1$, and Berenson's ``broad'' $P1$
+(2.2)). For instance, in the example below, $T_3$ should never see
+$a=1$, and, if $T_2$ aborts, $T_3$ should never read $a=3$:
+\vspace{-.5em}
+\begin{align*}
+\small
+T_1 &: w_x(1)~w_x(2)
+\\T_2 &: w_x(3)\\
+T_3 &: r_x(a)\\[-2em]
+\end{align*}
+It is fairly easy to prvent ``Dirty Reads'': if transactions never
+write uncommitted data to the database, then transactions will never
+read dirty data. Clients can buffer their writes until they commit,
+or, alternatively, can send them to servers, who will not serve writes
+until clients notify that they have been committed.
+
+\textit{Repeatable Read} isolation is a contentious property. As
+Berenson et al. point out, Gray's original Repeatable Read lock-based
+implementation provides substantially richer guarantees than those
+that are promised by the ANSI SQL specification. Gray, Berenson, and
+Adya effectively preclude all serializability anomalies but the
+Phantom Problem (which we have explicitly not yet considered and will
+do so in Section~\ref{sec:unachievable-hat}). Nevertheless, the ANSI
+SQL specification provides a useful property which, although is not
+true to Gray's spirit of Repeatable Read, is used in distributed
+consistency models. The ANSI Repeatable Read guarantee requires that,
+along with observing Read Committed isolation, each transaction can
+only read only version of each data item that it did not itself
+produce. In the example below, $T_3$'s must see $a=1$:
+\begin{align*}
+\small
+T_1 &: w_x(1)
+\\T_2 &: w_x(2)
+\\T_3 &: r_x(1)~r_x(a)
+\end{align*}
+This isolation property is much weaker than serializability, but it is
+a good faith literal interpretation of the phrase ``Repeatable Read'':
+transactions see their own writes. By itself, this property can be
+implemented trivially---always read \texttt{null} for each
+key. However, when coupled with additional properties, like
+transactional atomicity, it becomes stronger: the repeated reads must
+also obey some additional ordering properties. Distributed systems
+describe a ``consistent snapshot'' across a set of related events as a
+\textit{cut} across a set of participants or, in our case, data
+items. Accordingly, to capture the notion that the transaction should
+read from a consistent cut, and to disambiguate from the
+aforementioned stronger Repeatable Read properties, we call ANSI SQL
+Repeatable Read \textit{cut isolation}. In the absence of additional
+guarantees, it is possible to satisfy cut isolation with high
+availability by caching each transaction's reads or, alternatively,
+using multi-versioning on the server and ensuring that each
+transaction's successive reads return the same version of each data
+item.
+
+\subsubsection{ACID Atomicity Guarantees}
+
+Transactional atomicity is core to ACID guarantees. Although, at least
+by the acronym, not an ``isolation'' property, transactional atomicity
+restricts transactions' ability to view the effects of partially
+completed transactions: within a transactions, either all effects of
+another transaction are observed, or none are (equivalently, once some
+of the effects of a transaction are observed, all effects are
+observed). This is a strictly stronger guarantee than Read Committed
+isolation because, in Read Committed, we can read a subset of
+committed writes whereas, with Transactional Atomicity, we must read
+all of them. For example, if $T_1$ commits, then, given Transactional
+Atomicity, $T_2$ must observe $b=c=1$ (or later versions):
+\vspace{-.5em}
+\begin{align*}
+\small
+T_1 &: w_x(1)~w_y(1)~w_z(1)
+\\T_2 &: r_x(a)~r_y(1)~r_x(b)~r_z(c)~\\[-1.5em]
+\end{align*}
+$T_2$ can also observe $a=\bot$ or $a=1$.
+
+Perhaps perplexingly, discussions of Transactional Atomicity are
+absent from existing discussions of weak isolation levels. This is
+perhaps again due to the single-node context in which prior work was
+developed: on a single server, atomicity is near-trivial to achieve
+via lightweight locking and/or local concurrency control over data
+items. (In this spirit, multi-item atomicity is common in so-called
+``entity groups,'' which logically represent groups of co-located
+items.) In contrast, in a distributed environment, atomicity over
+arbitrary groups of items is considerably more difficult to achieve
+with high availability. The key challenge is that different servers
+responsible for subsets of a transaction's updates must coordinate
+regarding when to reveal their writes to a transaction---without
+violating availability due to, say, locking, or a missing
+transactional write.
+
+(MORE HERE FROM MY NOTES)
+
+Transactional atomicity is achievable in a HAT environment. To ensure
+transactional atomicity, servers only need to know when all of a
+transaction's updates are present on their respective servers but do
+not need to know when all respective servers know when all of a
+transaction's updates are present on their respective servers. This
+subtle distinction corresponds to the difference between the
+distributed systems concepts of consensus and uniform reliable
+broadcast. The former requires explicit agreement across servers,
+while the latter simply requires that, if one server reveals a write,
+then all servers reveal the write.
+
+(TODO: fix me--HotOS copy!)
+
+Nonetheless, Coordinating transactional atomicity across replicas is
+challenging. For example, if, as is common~\cite{readonly, eiger}, we
+use a master to determine the correct set of updates to read, then the
+system will not be available when clients are partitioned from the
+master. Instead, we rely on decentralized \textit{background}
+agreement to decide when to show new values, which can safely stall in
+the presence of partitions. Each server keeps a known good value for
+each data item $d$ it is assigned, $d_{good}$ and a set of pending
+updates to $d$, $d_{pending}$. We use asynchronous agreement to move
+updates from $d_{pending}$ to $d_{good}$: all of $d_{good}$'s
+transactional siblings (or later values) are guaranteed present on all
+replicas. Without aborts, this requires only one synchronous round
+trip time (RTT).
+
+\noindent\textit{Writes:} First, consider the case where transactions
+do not internally abort. On commit, for each written data item $d$, we
+send the last written value of $d$ ($d_v$) with the transaction ID and
+$L_v$, a list of all data items and values written in the transaction,
+to available replicas for $d$. The replicas immediately reply with a
+response to \textit{commit}, place $d_v$ it in their respective
+$d_{pending}$, and send a receipt to all replicas for items in
+$L_v$. Once a server receives receipts from all replicas for all items
+in $L_v$, the server sets $d_{good} = d_v$ if $d_v$'s ID is greater
+than $d_{good}$'s ID. Accordingly, a new value will only appear in
+$d_{good}$ when the other values written with $d_{good}$ ($L_v$) are
+resident on every respective replica (though possibly in their
+$d_{pending}$).\vspace{.5em}
+
+\noindent\textit{Reads:} Each client maintains a vector of transaction
+IDs $V$, with one entry per data item ($V(d)$). When a client reads a
+new data item $d_i$, the server returns both $d_i$ and $L_v$, and, for
+all data items $k \in L_v$, the client sets $V(k)$ to $d_i$'s ID. When
+a client sends a read request for item $x$, it also sends $V(x)$, and
+the replica returns \textit{either} $x_{good}$ if it matches $V(x)$ or
+occurs after $V(x)$ in the transaction order or, if $x_{good}$ is
+unsatisfactory, $V(x)$ from $x_{pending}$. If $V(x)$ is empty, the
+server returns $x_{good}$.\vspace{.5em}
+
+\noindent\textit{Aborts:} If we allow internal aborts, writes require
+2 RTTs: one to check constraints and one to execute the above. Write
+propagation and update visibility are achieved asynchronously, as in
+the no-abort case.\vspace{.5em}
+
+This implementation is entirely masterless and both reads and writes
+do not block for coordination. One key property is that we do not, in general,
+dictate when writes become visible (see: session guarantees,
+\S\ref{sec:impossible}). Accordingly, it is highly available.\vspace{.5em}
+
+\subsubsection{Session Guarantees}
+
+So far, our models have not considered guarantees across transactions,
+but, in the distributed systems context, many useful \textit{safety}
+guarantees span multiple operations. There is no guarantee of forward
+progression between transactions or any guarantees on the ordering
+between transactions (other than that there exists some arbitrary
+ordering).
+
+\textit{Session guarantees} are used to describe continuity between
+groups of transactions. Informally, a \textit{session} is a form of
+context that consists of logical state that should persist between
+operations. Users accessing a web site may want to experience a notion
+of ``forward progress'' for the duration that they are logged in, and
+the duration of the ``log in'' to ``log out'' period forms a
+session. We can define session guarantees at several granularities
+(e.g., per-transaction, per-client, per-hour) but we choose to operate
+under a client-centric model.
+
+There are several guarantees that we can make with high availability:
+
+\vspace{.25em}\noindent\textit{\textbf{Monotonic reads}} stipulates
+that a client's subsequent reads to a given object ``never return any
+previous values''; reads from each item progress forward in a total
+order. Under Read Uncommitted, the ordering of reads should respect the
+total ordering on transactions.
+
+\vspace{.25em}\noindent\textit{\textbf{Monotonic writes}} requires
+that transactions from each individual client be serialized. Under
+Read Uncommitted, the transaction order should also respect the order
+in which each client submitted transactions.
+
+\vspace{.25em}\noindent\textit{\textbf{Writes Follow Reads}} requires
+that, if a client reads a write originating from transaction $T_1$
+then performs transaction $T_2$, then a client can only read $T_2$'s
+writes if it can also read $T_1$'s prior write (or overwritten values
+for $T_1$'s write). This requirement forms the basis for potential
+causality: writes follow reads obeys the ``reads-from'' relation that
+captures all events that influenced each
+transaction~\cite{causalmemory}. Under Read Uncommitted, the
+transaction order should respect the reads-from order (note that the
+total order on transactions precludes reads-from cycles).
+
+We can achieve the above guarantees by forcing servers to wait to
+reveal new writes until their respect dependencies are fulfilled on
+all replicas. In the case of Monotonic Reads, servers should only
+reveal new writes to a data item when all other servers responsible
+for the item have also seen them. Otherwise, a client might read a
+newer version, subsequently become partitioned from its prior server,
+and contact a comparatively out-of-date server. In the case of
+Monotonic Writes, each new write should become visible only when each
+client's prior write is visible everywhere, along with Monotonic Reads
+guarantees. For Writes Follow Reads, a servers should similarly wait
+to reveal items until their dependencies are present
+everywhere.
+
+Effectively, this mechanism requires that clients only read from a
+globally agreed upon lower bound on the writes in the system. This is
+indeed highly available as a client will never block due to inability
+to find a server with a sufficiently up-to-date version of a data
+item. However, it does not imply that transactions will read their own
+writes or provide much confidence in making forward progress through
+the version history in the presence of partitions. The problem is
+that, if a server is partitioned, under the highly available model, we
+cannot discount the possibility that an unfortunate client will be
+forced to issue her next requests against the partitioned server!
+
+The solution to this conundrum is to give up high availability in
+favor of sticky availability. Sticky availability permits two
+additional models, which we first define and then prove are
+unachievable in a generic highly available system:
+
+\noindent\textit{\textbf{Read your writes}} requires
+that whenever a client reads a given data item after updating it, the
+read returns the updated value (or a value with a higher ID).
+
+\vspace{.25em}\noindent\textit{\textbf{Causal consistency}} is the
+combination of all of the prior session
+guarantees~\cite{daudjee-session} and is also referred to as PL-2L
+isolation~\cite{adya}). We can also consider arbitrary
+application-defined partial orders, as in explicit
+causality~\cite{explicit-socc}, if dependencies are specified at
+commit time.
+
+TRASH
+
+We can accomplish these three requirements in two ways. First, clients
+can locally cache their reads and writes, updating caches when they
+read admissible replacements from
+replicas~\cite{sessionguarantees}. Second, clients can ensure
+\textit{stickiness} with groups~\cite{vogels-defs} (subject to the
+same groups caveat as before).  For monotonic writes, clients should
+generate increasing IDs for each transaction they perform. Caching and
+ID generation are both available.
+
+To provide writes follow reads,
+severs (or clients) can buffer transaction writes until the
+transactions the writes depend on have been written to all
+replicas~\cite{cops}. If we consider a group model, each group can
+independently apply transactions in order via a (possibly sharded)
+log-shipping approach~\cite{causalmemory, cops, eiger, swift}.
+
+\subsection{Unachievable HAT Semantics}
+\label{sec:unachievable-hat}
 
@@
 		* S-HA is *doable* but it does result in substantially lowered unavailability unless one caches (e.g., bolt-on causal consistency)
+
+\noindent\textit{Groups:} We can optimize write visibility delay by
+partitioning servers and clients into groups, such that each group
+contains a fully replicated set of data items and all clients within a
+group contact only the servers in the group. Once all transaction
+values are present within a group, values can be installed as
+$d_{good}$ (as opposed to once they are present on all replicas). For
+example, in a multi-datacenter setting, all clients and servers within
+a datacenter may form a group. However, group-based atomicity is not
+highly available according to our definition unless clients and
+servers in a group fail together, as is often assumed in a
+geo-replicated context~\cite{cops, eiger}.
 

### 2013-03-19 writeup/hat-analysis.tex
more writing

@@
 
+Read your writes is unavailable in a highly available system. Consider
+a client that executes the following two transactions in succession:
+\vspace{-.5em}
+\begin{align*}
+\small
+T_1 &: w_x(1)
+\\T_2 &: r_x(a)
+\end{align*}
+If the client executes $T_1$ against a server that is partitioned from
+the rest of the other servers, the server must allow $T_1$ to
+commit. If the client executes $T_2$ against the same (partitioned)
+server, then it will be able to read its writes. However, if the
+network topology shifts and the client can only contact a different
+server which is partitioned from the rest of the cluster, then the
+client will not be able to read its own writes and the system will
+have to either stall indefinitely to allow the client to read her
+writes (violating transactional availability) or will have to
+sacrifice read your writes. Accordingly, read your writes requires
+stickiness, and, because causal consistency requires read your writes,
+causal consistency also requires stickiness.
+
+Read-your-writes can be accomplished via client-side caching and
+alternatively comes ``for free'' when requests are routed to a sticky
+set of servers. Achieving causal consistency requires maintaining
+read-your-writes along with the prior session guarantees. Both are
+reasonable strategies often employed in real-world
+systems~\cite{vogels-defs}, albeit at the cost of true high
+availability or client-side storage.
+
+We discuss additional uses for stickiness in Section~\ref{sec:discussion}.
+
+\subsubsection{Additional Guarantees}
+
+A HAT system can make \textbf{limited consistency} guarantees. It can
+execute commutative and logically monotonic~\cite{needed} operations
+without the risk of invalidating (also monotonic) application-level
+integrity constraints. Our goal in this paper is not to sketch the
+entire space of consistency models that are achievable (see
+Section~\ref{sec:futurework}, however we specifically evaluate TPC-C
+transaction semantics under HAT consistency guarantees in
+Section~\ref{sec:Evaluation}.
+
+As we briefly discussed in Section~\ref{sec:availabile}, a client
+requiring \textbf{durability} of surviving $F$ server faults requires
+at least $F+1$-availability.
+
+Finally, to require that servers propagate writes between replicas, we
+can require \textbf{convergence}, or eventual consistency for each
+data item: in the absence of new mutations to a data item, in the
+absence of partitions, all servers eventually agree on the value for
+each item. This is typically accomplished by any number of
+anti-entropy protocols, which periodically update neighboring servers
+with the latest value for each data item.
+
+\subsection{Unachievable HAT Semantics}
+\label{sec:unachievable-hat}
+
+While there are infinitely many HAT models
+(Section~\ref{sec:futurework}), at this point, we have exhausted our
+range of achievable, previously defined (useful) HAT semantics. Before
+summarizing our possibility results, we will present impossibility
+results for HATs, also defined in terms of previously identified isolation and consistency anomalies.
+
+\subsubsection{Unachievable ACID Isolation}
+
+* Lost Update
+	* Simple proof: T1 R(x) W(x=x+1) T2 R(x) W(x=x+1); execute on opposite sides 
+* Read Skew
+	* Similar easy proof
+* Write Skew
+	* Similar easy proof
+* Anti-dependency cycles
+	* Serializability
+* Recency guarantees on data items
+	* CAP
+
+\subsection{Summary}
+\label{sec:hat-summary}
+
+ \newcommand{\lostupdate}{$^\dagger$}
+ \newcommand{\rwskew}{$^\ddagger$}
+ \newcommand{\mphantom}{$^\star$}
+ \newcommand{\linearizable}{$^\oplus$}
+
+\begin{table}
+\begin{tabular}{| c | p{6cm} | }\hline
+HA & Transactional Atomicity (TA), Read Uncommitted (RU), Read
+Committed (RC), Cut Isolation (CI), Monotonic Reads (MR), Monotonic
+Writes (MW), Writes Follow Reads (WFR)\\\hline Sticky-HA & Read Your
+Writes (RYW), Causal Consistency\\\hline Unavailable & Cursor
+Stability (CS)\lostupdate, Read Consistency (CR)\lostupdate, Snapshot
+Isolation (SI)\lostupdate, Repeatable Read (RR)\lostupdate\rwskew,
+Serializability (S)\lostupdate\rwskew\mphantom, Causal
+Serializability\lostupdate\rwskew\mphantom, Strict
+Serializability\lostupdate\rwskew\mphantom\linearizable \\\hline
+\end{tabular}
+\caption{Summary of highly available, sticky highly available, and
+  unavailable models considered in this paper. Unavailable models are
+  labeled by cause of unavailability: preventing lost
+  update\lostupdate, preventing read/write skew\rwskew, preventing
+  phantoms\mphantom, and requiring linearizability\linearizable.}
+\label{table:hatcompared}
+\end{table}
+
+\begin{figure}
+\centering
+\begin{tikzpicture}[scale=0.8]
+  \tikzstyle{sticky}=[rectangle,draw=blue!50,fill=blue!20,thick]
+  \tikzstyle{noha}=[ellipse,draw=red!50,fill=red!20,thick, inner sep=0pt,minimum size=12pt]
+
+  \tikzstyle{every node}=[font=\small]
+
+ \node[draw=none,fill=none] (ci) at (1.2, 1.6) {\underline{CI}};
+ \node[draw=none,fill=none] (rc) at (1.2, .8) {\underline{RC}};
+ \node[draw=none,fill=none] (ru) at (1.2, 0) {\underline{RU}};
+
+ \node[draw=none,fill=none] (ta) at (0, 0) {\underline{TA}};
+ \node[draw=none,fill=none] (mr) at (2.4, 0) {\underline{MR}};
+ \node[draw=none,fill=none] (mw) at (3.6, 0) {\underline{MW}};
+ \node[draw=none,fill=none] (wfr) at (4.8,0) {\underline{WFR}};
+ \node at (6,0) [sticky] (ryw) {RYW};
+
+ \node at (3.6, 2) [sticky] (causal) {Causal};
+ \node[noha] (cs) at (1.7, 2.4) {CS};
+ \node[noha] (rr) at (.7, 2.6) {RR};
+ \node[noha] (si) at (-.4, 2.2) {SI};
+ \node[noha] (cr) at (-.4, 1.2) {CR};
+ \node[noha] (s) at (.5, 3.5) {S};
+ \node[noha] (causal-s) at (3.6, 3.5) {Causal S};
+ \node[noha] (strict-s) at (6, 3.5) {Strict S};
+
+ \draw [->] (ru) -- (rc);
+ \draw [->] (rc) -- (ci);
+
+ \draw [->, blue] (mr) -- (causal);
+ \draw [->, blue] (mw) -- (causal);
+ \draw [->, blue] (wfr) -- (causal);
+ \draw [->, blue] (ryw) -- (causal);
+
+ \draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (ru) -- (mr);
+ \draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (rc) -- (ta);
+ \draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (ci) -- (ta);
+ \draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (rc) -- (mr);
+ \draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (ci) -- (mr);
+ \draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (ta) -- (ru);
+ \draw[snake=coil, blue, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (ru) -- (causal);
+ \draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (mr) -- (mw);
+ \draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (wfr) -- (mw);
+ \draw[snake=coil, blue, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (wfr) -- (ryw);
+
+ \draw [->, red] (ci) -- (cs);
+ \draw [->, red] (cs) -- (rr);
+ \draw [->, red] (ci) -- (si);
+ \draw [->, red] (rr) -- (s);
+ \draw [->, red] (si) -- (s);
+ \draw [->, red] (ta) -- (cr);
+ \draw [->, red] (rc) -- (cr);
+ \draw [->, red] (cr) -- (si);
+ \draw [->, red] (ta) -- (rr);
+ \draw [->, red] (s) -- (causal-s);
+ \draw [->, red] (causal) -- (causal-s);
+ \draw [->, red] (causal-s) -- (strict-s);
+
+\end{tikzpicture}
+\label{fig:hat-order}
+\caption{HAT, sticky HAT (in boxes), and unavailable models (circled)
+  from Table~\protect\ref{table:hatcompared} compared graphically. Directed
+  edges represent ordering by model strength, squiggle edges represent
+  (transitive) compatible models, while absence of edges represents
+  incomparable models. The combination of all underlined models
+  represents maximum ACID HAT semantics, while the combination of
+  underlined and boxed models represents maximum ACID sticky HAT
+  models.}
+\label{fig:hatcompared}
+\end{figure}
+
+\subsection{Discussion}
+\label{sec:discussion}
+
 TRASH
@@
 log-shipping approach~\cite{causalmemory, cops, eiger, swift}.
-
-\subsection{Unachievable HAT Semantics}
-\label{sec:unachievable-hat}
 
@@
 Let's sketch out a few conditions, then we'll discuss them later:
-* Lost Update
-	* Simple proof: T1 R(x) W(x=x+1) T2 R(x) W(x=x+1); execute on opposite sides 
-* Read Skew
-	* Similar easy proof
-* Write Skew
-	* Similar easy proof
-* Anti-dependency cycles
-	* Serializability
-* Recency guarantees on data items
-	* CAP
 

### 2013-03-19 thebes-code/common/src/main/java/edu/berkeley/thebes/common/data/Version.java
Merge?

@@ -18,7 +18,14 @@ public class Version implements Comparable<Version> {
 	}
 	
 	public static Version fromThrift(ThriftVersion version) {
+<<<<<<< Updated upstream
 		return new Version(version.getClientID(), version.getTimestamp());
+=======
+	    if (version == null)
+	        return null;
+		return new Version(version.getClientID(), version.getLogicalTime(),
+		        version.getTimestamp());
+>>>>>>> Stashed changes
 	}
 	
 	public static ThriftVersion toThrift(Version version) {

### 2013-03-19 writeup/model.tex
more writing

@@
   \tikzstyle{every node}=[font=\small]
- \node[draw=none,fill=none] (1) at (0,0) {$1$-availability}; 
+ \node[draw=none,fill=none] (1) at (0,0) {$1$-availability (high availability)}; 
  \node[draw=none,fill=none] (2) at (0,1) {$2$-availability}; 
@@
  \node[draw=none,fill=none] (n) at (0, 5) {$N$-availability}; 
- \node[draw=none,fill=none] (1s) at (4, 1) {$1$-sticky availability};
- \node[draw=none,fill=none] (2s) at (4, 2) {$2$-sticky availability};
- \node[draw=none,fill=none] (n2s) at (4, 3) {\ldots};
- \node[draw=none,fill=none] (n1s) at (4, 4) {$(N$$-$$1)$-sticky availability};
+ \node[draw=none,fill=none] (1s) at (4, 1.5) {$1$-sticky availability};
+ \node[draw=none,fill=none] (2s) at (4, 2.5) {$2$-sticky availability};
+ \node[draw=none,fill=none] (n2s) at (4, 3.5) {\ldots};
+ \node[draw=none,fill=none] (n1s) at (4, 4.5) {$(N$$-$$1)$-sticky availability};
 
@@
  \draw [->] (1) -- (1s);
+ \draw [->] (1s) -- (m);
+ \draw [->] (2s) -- (m+1);
  \draw [->] (1s) -- (2s);
@@
 then subsequent requests will trivially access the same set of
-replicas.
+replicas. Similarly, with majority availability, at least one replica
+will be present in all requests, so $1$-sticky availability is weaker
+than majority availability, and, more generally, $K$-sticky
+availability is weaker than $\lceil \frac{N}{2} \rceil$$+$$K$$-$$1$
+availability.
 

### 2013-03-19 writeup/motivation.tex

@@
 Computing'' with two concerns: ``\textit{1.)}  The network is
-reliable. \textit{2.)} Latency is zero.'' On a single-node system,
-communication systems are robust and relatively fast, and most systems
-do not need to prepare for partial component failure. However, as
-Deutsch points out, these assumptions are no longer valid in a
-distributed setting. To extend Jim Gray's storage latency analogy, if
-memory is around 1.5 hours from San Francisco and disk is as ``far
-away'' as Pluto (2 years), in a distributed context, another computer
-can be even farther---nearing or even worse than Gray's location of
-tape access near Andromeda (2,000 years)---with the added concern that
-messages can be lost in transit, destroyed by ``black holes'' in the
-form of network partitions. Accordingly, accounting for partial
-failure and communication expense is core to distributed systems and
-algorithm design; ignoring them ``cause[s] big trouble and painful
-learning experiences.''
+reliable. \textit{2.)} Latency is zero''~\cite{fallacies-deutsch}. On
+a single-node system, communication systems are robust and relatively
+fast, and most systems do not need to prepare for partial component
+failure. However, as Deutsch points out, these assumptions are no
+longer valid in a distributed setting. To extend Jim Gray's storage
+latency analogy, if memory is the distance from San Diego to Olympia
+(1.5 hours), and disk is as ``far away'' as Pluto (2 years), in a
+distributed context, another computer can be even farther---nearing or
+even worse than Gray's location of tape access near Andromeda (2,000
+years)---with the added concern that messages can be lost in transit,
+destroyed by ``black holes'' in the form of network
+partitions. Accordingly, accounting for partial failure and
+communication expense is core to distributed systems and algorithm
+design; ignoring them ``cause[s] big trouble and painful learning
+experiences''~\cite{fallacies-deutsch}.
 
@@
 non-negligible. As shown in Figure~\ref{fig:rtt}, the distribution of
-latencies varies between links, but the trend is clear: remote
-communication has a substantial cost.
+latencies varies between links, but the trend---in line with other
+recent studies~\cite{redblue, mdcc}---is clear: remote communication
+has a substantial cost.
 
@@
 \end{figure}
-
-* As an example of a real system going from "CP" to "AP", [PNUTS dropped per-record mastering](\url{http://developer.yahoo.com/blogs/ydn/posts/2010/06/sherpa_update/#4})
-
-What about existing systems?
-* Database technology developed for single-node systems; gold standard: serializability
-	* Serializability is not actually highly available; give an example!
-	* Much of the database literature presumes serializability and does not consider high availability.
-	* For the literature that doesn't presume serializability, it's not presented in a HA context?
-* Consequence: traditional systems are not optimized for high availability
-	* Consider two-phase locking in a distributed environment
-	* As we will see (forward reference to Evaluation), many existing transactional systems encounter similar difficulties
-* Key question in this paper: What transactional semantics are highly available, and which aren't?
-	* Side note: there are infinite incomparable consistency models (e.g., always return 2, always return 3, …)
-	* Our goal is to unify distributed systems literature with ACID transaction model.
-* Argument: Highly Available Transactional Systems (HATS) require rethinking existing models. This work is a first step.
-
-In this sense, HATs are similar to RAID: optimizing for graceful
-handling of a worst-case failure scenario improves average-case
-performance.

### 2013-03-19 writeup/hat-analysis.tex
More writing

@@
 only read only version of each data item that it did not itself
-produce. In the example below, $T_3$'s must see $a=1$:
+produce (Berenson et al. call this preventing ``Fuzzy Read,'' or
+P2). In the example below, $T_3$'s must see $a=1$:
 \begin{align*}
@@
 aforementioned stronger Repeatable Read properties, we call ANSI SQL
-Repeatable Read \textit{cut isolation}. In the absence of additional
+Repeatable Read \textbf{Cut Isolation}. In the absence of additional
 guarantees, it is possible to satisfy cut isolation with high
@@
 item.
+
+We can consider two variants of Cut Isolation. The first is relatively
+straightforward, and reasons about cuts over discrete data items. We
+will call this \textit{Item Cut Isolation}. The second is known as
+preventing the \textit{Phantom Problem}, whereby two successive
+predicate-based reads return different data. We call predicate-based
+cut isolation, which prevents Phansoms, \textbf{Phantom Cut
+  Isolation}.
 
@@
 transactional write.
+
+Together with Cut Isolation, Transactional Atomicity prevents Read
+Skew (Berenson et al.'s A5A).
 
@@
 anti-entropy protocols, which periodically update neighboring servers
-with the latest value for each data item.
+with the latest value for each data item. This convergence property
+should be considered mandatory given the spirit of existing literature
+on isolation protocols: in all models we have encountered, the
+database should behave as if there is some total order over
+transactions. Indeed, non-convergent systems are remarkably difficult
+to reason about.
 
@@
 summarizing our possibility results, we will present impossibility
-results for HATs, also defined in terms of previously identified isolation and consistency anomalies.
+results for HATs, also defined in terms of previously identified
+isolation and consistency anomalies.
 
@@
 
-* Lost Update
-	* Simple proof: T1 R(x) W(x=x+1) T2 R(x) W(x=x+1); execute on opposite sides 
-* Read Skew
-	* Similar easy proof
-* Write Skew
-	* Similar easy proof
-* Anti-dependency cycles
-	* Serializability
+The simplest way to describe which isolation properties are achievable
+with high availability is in terms of preventing different isolation
+anomalies. We have already shown that it is possible prevent several
+anomalies, like Dirty Read, Dirty Write, Fuzzy Read, and atomicity
+violations. In this section, we demonstrate that preventing Lost
+Update, Write Skew, and Phantoms requires unavailability.
+
+In the words of Berenson et al., \textbf{Lost Update} occurs when one
+transaction $T1$ reads a given data item, a second transaction $T2$
+updates the same data item, then $T1$ modifies the data item based on
+its original read of the data item, ``missing'' or ``losing'' $T2$'s
+newer update. Consider the following transactions:
+\begin{align*}
+\small
+T_1 &: r_x(a) w_x(a+1)
+\\T_2 &: w_x(2)
+\end{align*}
+If $T_1$'s read from $x$ precedes $T_2$'s write to $x$ but $T_2$'s
+write to $x$ precedes $T_1$'s write to $x$, then a database containing
+only these two transactions will end in a state that could not have
+resulted in a serial execution.
+
+It is impossible to prevent Lost Update in a highly available
+environment. Consider two clients who submit the following $T_1$ and
+$T_2$ on opposite sides of a network partition:
+\begin{align*}
+\small
+T_1 &: r_x(100)~w_x(100+20=120)
+\\T_2 &: r_x(100)~w_x(100+30=130)
+\end{align*}
+The final state of the database will converge to an execution that
+could not have resulted from a serial excecution. To prevent this from
+happening, either $T_1$ or $T_2$ should not have committed. Each
+client's respective server might try to detect that another write
+occurred, but this requires knowing the version of the latest write to
+$x$. In this example, this reduces to a requirement for
+linearizability, which is, via Gilbert and Lynch's proof of the CAP
+Theorem, provably unachievable with high availabilty.
+
+\textbf{Write Skew} is a generalization of Lost Update to multiple
+keys. It occurs when one transaction $T1$ reads a given data item $x$,
+a second transaction $T2$ reads a different data item $y$, then $T1$
+writes to $y$ and commits and $T2$ writes to $x$ and commits. As an
+example of Write Skew, consider the following two transactions:
+\begin{align*}
+\small
+T_1 &: r_y(0)~w_x(1)
+\\T_2 &: r_x(0)~w_y(1)
+\end{align*}
+As Berenson et al. describe, if there was an integrity constraint
+between $x$ and $y$ such that only one of $x$ or $y$ should be one at
+any given time, then this write skew would lead to a non-serializable
+execution. Write skew in particular is a somewhat esoteric
+anomaly---for example, it does not appear in TPC-C---but, as a
+generalization of Lost Update, is also unavailable.
+
+\textbf{}
+
 * Recency guarantees on data items
@@
  \newcommand{\rwskew}{$^\ddagger$}
- \newcommand{\mphantom}{$^\star$}
  \newcommand{\linearizable}{$^\oplus$}
@@
 HA & Transactional Atomicity (TA), Read Uncommitted (RU), Read
-Committed (RC), Cut Isolation (CI), Monotonic Reads (MR), Monotonic
-Writes (MW), Writes Follow Reads (WFR)\\\hline Sticky-HA & Read Your
-Writes (RYW), Causal Consistency\\\hline Unavailable & Cursor
-Stability (CS)\lostupdate, Read Consistency (CR)\lostupdate, Snapshot
-Isolation (SI)\lostupdate, Repeatable Read (RR)\lostupdate\rwskew,
-Serializability (S)\lostupdate\rwskew\mphantom, Causal
-Serializability\lostupdate\rwskew\mphantom, Strict
-Serializability\lostupdate\rwskew\mphantom\linearizable \\\hline
+Committed (RC), Item Cut Isolation (P-CI), Phantom Cut Isolation
+(P-CI), Monotonic Reads (MR), Monotonic Writes (MW), Writes Follow
+Reads (WFR)\\\hline Sticky-HA & Read Your Writes (RYW), Causal
+Consistency\\\hline Unavailable & Cursor Stability (CS)\lostupdate,
+Read Consistency (CR)\lostupdate, Snapshot Isolation (SI)\lostupdate,
+Repeatable Read (RR)\lostupdate\rwskew, Serializability
+(S)\lostupdate\rwskew, Causal
+Serializability\lostupdate\rwskew, Strict
+Serializability\lostupdate\rwskew\linearizable \\\hline
 \end{tabular}
@@
   labeled by cause of unavailability: preventing lost
-  update\lostupdate, preventing read/write skew\rwskew, preventing
-  phantoms\mphantom, and requiring linearizability\linearizable.}
+  update\lostupdate, preventing write skew\rwskew, and requiring
+  linearizability\linearizable.}
 \label{table:hatcompared}
@@
 
- \node[draw=none,fill=none] (ci) at (1.2, 1.6) {\underline{CI}};
+ \node[draw=none,fill=none] (ici) at (1.2, 1.6) {\underline{I-CI}};
+ \node[draw=none,fill=none] (pci) at (1.2, 2.4) {\underline{P-CI}};
  \node[draw=none,fill=none] (rc) at (1.2, .8) {\underline{RC}};
@@
  \node[draw=none,fill=none] (mw) at (3.6, 0) {\underline{MW}};
- \node[draw=none,fill=none] (wfr) at (4.8,0) {\underline{WFR}};
- \node at (6,0) [sticky] (ryw) {RYW};
+ \node[draw=none,fill=none] (wfr) at (4.9,0) {\underline{WFR}};
+ \node at (6.3,0) [sticky] (ryw) {\underline{RYW}};
 
- \node at (3.6, 2) [sticky] (causal) {Causal};
- \node[noha] (cs) at (1.7, 2.4) {CS};
- \node[noha] (rr) at (.7, 2.6) {RR};
- \node[noha] (si) at (-.4, 2.2) {SI};
+ \node at (3.6, 2) [sticky] (causal) {\underline{Causal}};
+ \node[noha] (cs) at (2, 2.8) {CS};
+ \node[noha] (rr) at (.6, 3.0) {RR};
+ \node[noha] (si) at (-.4, 2.4) {SI};
  \node[noha] (cr) at (-.4, 1.2) {CR};
- \node[noha] (s) at (.5, 3.5) {S};
+ \node[noha] (s) at (-.4, 3.5) {S};
  \node[noha] (causal-s) at (3.6, 3.5) {Causal S};
@@
  \draw [->] (ru) -- (rc);
- \draw [->] (rc) -- (ci);
+ \draw [->] (rc) -- (ici);
+ \draw [->] (ici) -- (pci);
 
@@
 
- \draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (ru) -- (mr);
- \draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (rc) -- (ta);
- \draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (ci) -- (ta);
- \draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (rc) -- (mr);
- \draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (ci) -- (mr);
- \draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (ta) -- (ru);
- \draw[snake=coil, blue, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (ru) -- (causal);
- \draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (mr) -- (mw);
- \draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (wfr) -- (mw);
- \draw[snake=coil, blue, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (wfr) -- (ryw);
+ %\draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (ru) -- (mr);
+ %\draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (rc) -- (ta);
+ %\draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (ici) -- (ta);
+ %\draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (pci) -- (ta);
+ %\draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (rc) -- (mr);
+ %\draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (pci) -- (mr);
+ %\draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (ici) -- (mr);
+ %\draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (ta) -- (ru);
+ %\draw[snake=coil, blue, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (ru) -- (causal);
+ %\draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (mr) -- (mw);
+ %\draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (wfr) -- (mw);
+ %\draw[snake=coil, blue, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (wfr) -- (ryw);
 
- \draw [->, red] (ci) -- (cs);
+ \draw [->, red] (ici) -- (cs);
  \draw [->, red] (cs) -- (rr);
- \draw [->, red] (ci) -- (si);
+ \draw [->, red] (pci) -- (si);
  \draw [->, red] (rr) -- (s);
@@
 \caption{HAT, sticky HAT (in boxes), and unavailable models (circled)
-  from Table~\protect\ref{table:hatcompared} compared graphically. Directed
-  edges represent ordering by model strength, squiggle edges represent
-  (transitive) compatible models, while absence of edges represents
-  incomparable models. The combination of all underlined models
-  represents maximum ACID HAT semantics, while the combination of
-  underlined and boxed models represents maximum ACID sticky HAT
-  models.}
+  from Table~\protect\ref{table:hatcompared} compared
+  graphically. Directed edges represent ordering by model strength,
+  while all underlined models can be combined (e.g., the powerset of
+  the set of underlined models are all valid models). The combination
+  of all underlined, non-boxed models represents maximum ACID HAT
+  semantics, while the combination of underlined and boxed models
+  represents maximum ACID sticky HAT models.}
 \label{fig:hatcompared}

### 2013-03-20 writeup/evaluation.tex
more writing
More writing

@@
-\section{evaluation}
+\section{Analysis}
+\label{sec:evaluation}
 
-V. Evaluation (1.5pg)
+With a firm understanding of which of several ACID and distributed
+consistency properties are available and not, we perform analysis of
+their implications for existing algorithms, applications, and
+deployments. Specifically:
 
-The goal in this section is twofold:
-1.) Look at existing algorithms and their coordination cost in the abstract
-2.) Perform limited real-world deployments to validate these abstract costs
+\begin{enumerate}
+\item We revisit traditional database concurrency control with a focus
+  on coordination costs and on high availability.
+\item We examine the properties required by a realistic OLTP
+  application patterned on TPC-C.
+\item We perform a brief experimental evaluation of HAT versus non-HAT
+  properties on public cloud infrastructure.
+\end{enumerate}
+
+\subsection{Existing Algorithms}
+
+As we have discussed, many existing database transaction and
+concurrency control algorithms are not designed for high
+availability. These algorithms often presume a single-server
+deployment or the requirement for serializability. As a consequence,
+traditional transaction processing systems are not well-optimized for
+a HAT context.
+
+ designed for either single-node deployments or
+to achieve the are designed for
 
@@
 
+\subsection{Application Requirements}
+
 2.) We ran YCSB on EC2 (and maybe TPC-C if we have time) with a few models:
@@
 	-Not meant as an exhaustive study, but validates our intuitions. Coordination costs of non-HAT systems are unaccounted for, will only go up, and we haven't even talked about availability.
+
+\subsection{Experimental Costs}

### 2013-03-20 writeup/hat-vldb.tex
checkpoint writing
more writing
More writing

@@
 \documentclass{vldb}
-\usepackage{times, graphicx, url, xcolor, multicol, minipage, subfig}
+\usepackage{times, graphicx, url, xcolor, multicol, minipage, subfig, tikz}
 \usepackage{balance}  % for  \balance command ON LAST PAGE  (only there!)
+
+\usetikzlibrary{positioning, snakes, shapes}
 
@@
 \begin{abstract}
-foo bar baz
+Given the presence of network latency and the possibility of partial
+failure and network partitions, many systems eschew transactional
+functionality providing strong semantic guarantees on groups of
+multiple operations over multiple data items. In this work, we
+consider the problem of providing transactional guarantees that are
+highly available, or do not suffer from network latency or system
+partitions.  We introduce a new taxonomy of highly available
+algorithms and analyze a combination of ACID and distributed
+consistency guarantees to identify which can and cannot be achieved by
+Highly Available Transaction (HAT) systems. In doing so, we
+analytically and experimentally quantify the cost of violating high
+availability as well as the cost of providing necessarily relaxed
+semantics.
 \end{abstract}
@@
 
-\subsection{References}
-
-Generated by bibtex from your ~.bib file. Run latex, then bibtex, then
-latex twice (to resolve references).
-
 %APPENDIX is optional.

### 2013-03-20 writeup/evaluation.tex
More eval

@@
 
- designed for either single-node deployments or
-to achieve the are designed for
+We begin by considering several existing algorithms that provide
+serializability but are not highly available. In summary, all
+algorithms in this class require at least one round trip time (RTT)
+before committing.
 
-1.) There are a bunch of existing systems that aren't HATS but provide stronger semantics.
-What's their cost?
-(N.B.
-this is almost a related work survey here, but I think, if done right, it can be very instructive.)
-Not HA, Serializable:
-* Two-phase locking: for transaction size T, T+1 round-trips (directly to each database)
-* Deterministic transaction scheduling: 1 round-trip per partition (to scheduler)
-* Optimistic concurrency control: 1 round-trip per partition (to database cycle checker)
-* MVCC: min 1 round-trip per partition (to partition timestamp manager)
+\begin{itemize}
+\item \textbf{Two-phase locking} (ignoring deadlocks) requires $T$
+  lock operations and 1 unlock operation for a transaction of length
+  $T$. If all items are located on a single master server, this
+  requires one RTT (for requests originating from servers other than
+  the master). If items are located on $S$ ($S \leq T$) different
+  servers, two-phase locking requires, at minimum, $2(S-1)$ RTTs---one
+  for lock and another for unlock.
 
-Not HA, not serializable, but "transactional":
-* MDCC: need to check, 1-2 round-trip times because uses Paxos variant
-* Walter: 1 rount-trip time to each preferred site
-* Eiger: Sticky-HA, assumes linearizable clusters, reads/writes stall in presence of coordinator failures
-* Chan and Gray read-only transaction: reads stall in the presence of coordinator failure
-* Bolt-on causal consistency: sticky-HA, could be used to implement TA+RR, uses client cache
-* COPS: Sticky-HA, assumes linearizable clusters
-* Bayou: not HA since client doesn't cache (pretty sure)
+\item \textbf{Optimistic concurrency control} allows reads from any
+  server but transactions must be checked by a scheduler before
+  commit. With one scheduler per database, this requires one RTT for
+  concurrency control. If schedulers are sharded by database
+  partition, a multi-partition transaction may require contacting
+  several different schedulers and, possibly, an atomic commit
+  protocol, which becomes substantially more expensive.
+
+\item \textbf{Deterministic transaction scheduling} is effecitvely the
+  inverse of optimistic concurrency control: a scheduler first orders
+  the transactions, then sends them to database replicas for in-order
+  execution. These techniques, pioneered in the context of totally
+  ordered atomic broadcast systems, have recently seen increased
+  attention and effectively require communication overhead similar to
+  optimistic concurrency control.
+
+\item\textbf{Multiversion concurrency control} allows readers to read
+  older versions of data items but, like OCC, must eventually be
+  checked for write-write conflicts, requiring a scheduler.
+
+\end{itemize}
+
+There are several mechanisms that do not provide serializability (and
+often provide HAT semantics) but which are not available:
+
+\begin{itemize}
+\item \textbf{Lock-based protocols} such as those used to provide weak
+  isolation in Gray's original paper require at least one RTT to the a
+  lock server (or, alternatively, a majority quorum). In the presence
+  of failure, distributed locking can be challenging.
+\item The \textbf{MDCC} commit protocol uses a Paxos variant and escrow
+  transactions to provide Repeatable Read isolation with 1-2 RTTs. (TODO: CHECK)
+\item \textbf{Parallel Snapshot Isolation} as developed by the Walter
+  system provides (unavailable) snapshot isolation variant and, even
+  when writes commute, requires 1 RTT to enforce read
+  recency. However, PSI without recency requirements could be made
+  highly available.
+\item \textbf{Eiger} provides read-only and write-only transactions
+  with item cut isolation and causal consistency. It relies on sticky
+  high availability and groups within a datacenter with linearizable
+  clusters. In the presence of server failure during write
+  transactions, reads and writes may become unavailable. \textbf{COPS}
+  provides similar semantics for read-only transactions but does not
+  suffer from unavailability as it does not have a coordinator.
+\item \textbf{Chan and Gray read-only transactions} use a similar
+  algorithm to Eiger to provide read-only transactions with item cut
+  isolation and causal consistency but coordinate on reads instead of
+  writes.
+\item \textbf{Bolt-on Causal Consistency} and \textbf{Swift} can
+  provide read-only and write-only transactions with item cut
+  isolation and causal consistency with sticky high availability and
+  both rely on client-side caching.
+\item \textbf{Bayou} provides several session guarantees but, without
+  client-side caching, actually faces unavailability in the presence
+  of failure. (TODO: CHECK)
+\item \textbf{Brantner's S3 Database} uses caching to provide several
+  guarantees such as XX and YY but is unavailable during failures
+  (TODO: CHECK).
+\end{itemize}
 

### 2013-03-20 thebes-code/common/src/main/java/edu/berkeley/thebes/common/data/Version.java
Add unit test!

@@ -20,9 +20,9 @@ public class Version implements Comparable<Version> {
 	private final long timestamp;
 	
 	public Version(short clientID, long logicalTime, long timestamp) {
-		this.clientID = clientID;
-		this.logicalTime = logicalTime;
-		this.timestamp = timestamp;
+		this.clientID = (short) (pow2Less1(numBitsClientID) & clientID);
+		this.logicalTime = pow2Less1(numBitsLogicalTime) & logicalTime;
+		this.timestamp = pow2Less1(numBitsTimestamp) & timestamp;
 	}
 	
 	public static Version fromThrift(ThriftVersion thriftVersion) {

### 2013-03-20 thebes-code/hat/common/src/main/java/edu/berkeley/thebes/hat/common/thrift/AntiEntropyService.java
Reduce relay overhead even further (sadly)

@@ -36,7 +36,7 @@ public class AntiEntropyService {
 
     public void put(String key, edu.berkeley.thebes.common.thrift.ThriftDataItem value) throws org.apache.thrift.TException;
 
-    public void ackDependentWriteInPending(String myKey, String ackedKey, edu.berkeley.thebes.common.thrift.ThriftVersion version) throws org.apache.thrift.TException;
+    public void ackDependentWriteInPending(String myKey, edu.berkeley.thebes.common.thrift.ThriftVersion version) throws org.apache.thrift.TException;
 
   }
 
@@ -44,7 +44,7 @@ public class AntiEntropyService {
 
     public void put(String key, edu.berkeley.thebes.common.thrift.ThriftDataItem value, org.apache.thrift.async.AsyncMethodCallback<AsyncClient.put_call> resultHandler) throws org.apache.thrift.TException;
 
-    public void ackDependentWriteInPending(String myKey, String ackedKey, edu.berkeley.thebes.common.thrift.ThriftVersion version, org.apache.thrift.async.AsyncMethodCallback<AsyncClient.ackDependentWriteInPending_call> resultHandler) throws org.apache.thrift.TException;
+    public void ackDependentWriteInPending(String myKey, edu.berkeley.thebes.common.thrift.ThriftVersion version, org.apache.thrift.async.AsyncMethodCallback<AsyncClient.ackDependentWriteInPending_call> resultHandler) throws org.apache.thrift.TException;
 
   }
 
@@ -81,16 +81,15 @@ public class AntiEntropyService {
       sendBase("put", args);
     }
 
-    public void ackDependentWriteInPending(String myKey, String ackedKey, edu.berkeley.thebes.common.thrift.ThriftVersion version) throws org.apache.thrift.TException
+    public void ackDependentWriteInPending(String myKey, edu.berkeley.thebes.common.thrift.ThriftVersion version) throws org.apache.thrift.TException
     {
-      send_ackDependentWriteInPending(myKey, ackedKey, version);
+      send_ackDependentWriteInPending(myKey, version);
     }
 
-    public void send_ackDependentWriteInPending(String myKey, String ackedKey, edu.berkeley.thebes.common.thrift.ThriftVersion version) throws org.apache.thrift.TException
+    public void send_ackDependentWriteInPending(String myKey, edu.berkeley.thebes.common.thrift.ThriftVersion version) throws org.apache.thrift.TException
     {
[…]

### 2013-03-20 thebes-code/hat/server/src/main/java/edu/berkeley/thebes/hat/server/antientropy/AntiEntropyServiceHandler.java

@@ -46,9 +46,9 @@ public class AntiEntropyServiceHandler implements AntiEntropyService.Iface {
     }
 
     @Override
-    public void ackDependentWriteInPending(String myKey, String ackedKey,
+    public void ackDependentWriteInPending(String myKey,// String ackedKey,
             ThriftVersion version) throws TException {
-        dependencyResolver.dependentWriteAcked(myKey, ackedKey, 
+        dependencyResolver.dependentWriteAcked(myKey, null /* ackedKey */,
                 Version.fromThrift(version));
     }
 }

### 2013-03-20 thebes-code/hat/server/src/main/java/edu/berkeley/thebes/hat/server/antientropy/clustering/AntiEntropyServiceRouter.java

@@ -114,7 +114,7 @@ public class AntiEntropyServiceRouter {
             for (String dependentKey : writeToAnnounce.getValue().getTransactionKeys()) {
                 AntiEntropyService.Client neighborClient = neighborClients.get(
                         RoutingHash.hashKey(dependentKey, numServersInCluster));
-                neighborClient.ackDependentWriteInPending(dependentKey, writeKey,
+                neighborClient.ackDependentWriteInPending(dependentKey,
                         Version.toThrift(writeVersion));
             }
         } catch (TException e) {

### 2013-03-20 thebes-code/hat/server/src/main/java/edu/berkeley/thebes/hat/server/dependencies/DependencyResolver.java

@@ -110,6 +110,7 @@ public class DependencyResolver implements PendingWrite.WriteReadyCallback {
         pendingWritesMap.get(write.getKey()).remove(write);
     }
     
+    // NB: null is a valid ackedKey... only used for efficieny reasons
     public void dependentWriteAcked(String myKey, String ackedKey, Version version) {
         Ack ack = new Ack(ackedKey, version);
         

### 2013-03-20 thebes-code/hat/server/src/main/java/edu/berkeley/thebes/hat/server/dependencies/PendingWrite.java

@@ -18,23 +18,36 @@ public class PendingWrite implements Comparable<PendingWrite> {
     private final DataItem value;
     private final Version transactionVersion;
     private final Set<String> pendingAckedKeys;
+    private int numUnnamedAcks; // TODO: Remove either this code back or the one above.
     private final WriteReadyCallback callback;
     
     public PendingWrite(String key, DataItem value, WriteReadyCallback callback) {
+        if (value.getTransactionKeys().isEmpty()) {
+            throw new IllegalStateException("Pending writes must be waiting on at least one other key.");
+        }
+        
         this.key = key;
         this.value = value;
         this.transactionVersion = value.getVersion();
         this.pendingAckedKeys = Sets.newHashSet(value.getTransactionKeys());
         this.callback = callback;
+        this.numUnnamedAcks = 0;
+        
     }
 
     public boolean isWaitingFor(String ackedKey, Version version) {
-        return version.equals(transactionVersion) && pendingAckedKeys.contains(ackedKey);
+        return version.equals(transactionVersion) &&
+                (ackedKey == null || pendingAckedKeys.contains(ackedKey));
     }
 
     public void keyAcked(String ackedKey) {
-        pendingAckedKeys.remove(ackedKey);
-        if (pendingAckedKeys.isEmpty()) {
+        if (ackedKey != null) {
+            pendingAckedKeys.remove(ackedKey);
+        } else {
+            numUnnamedAcks ++;
+        }
+        
+        if (numUnnamedAcks == pendingAckedKeys.size() || pendingAckedKeys.isEmpty()) {
             callback.writeReady(this);
         }
     }

### 2013-03-20 thebes-code/hat/server/src/main/java/edu/berkeley/thebes/hat/server/dependencies/DependencyResolver.java
Fix compareTo on Ack

@@ -149,10 +149,14 @@ public class DependencyResolver implements PendingWrite.WriteReadyCallback {
 
         @Override
         public int compareTo(Ack o) {
-            return ComparisonChain.start()
-                    .compare(key, o.key)
-                    .compare(version, o.version)
-                    .result();
+            if (key != null && o.key != null) {
+                return ComparisonChain.start()
+                        .compare(key, o.key)
+                        .compare(version, o.version)
+                        .result();
+            } else {
+                return version.compareTo(o.version);
+            }
         } 
     }
 }

### 2013-03-20 writeup/evaluation.tex
eval

@@
 There are several mechanisms that do not provide serializability (and
-often provide HAT semantics) but which are not available:
+often provide HAT semantics) but which are not
+available. Non-serializability can but does not necessarily imply high
+availability and low latency:
 

### 2013-03-20 thebes-code/common/src/main/java/edu/berkeley/thebes/common/data/Version.java
Check for NULL_VERSION on toThrift

@@ -40,7 +40,7 @@ public class Version implements Comparable<Version> {
 	}
 	
 	public static ThriftVersion toThrift(Version version) {
-        if(version == null)
+        if (version == null || version == NULL_VERSION)
             return null;
         
         long l = 0;

### 2013-03-20 thebes-code/common/src/main/java/edu/berkeley/thebes/common/config/Config.java
LevelDB and null fixes

@@ -293,6 +293,14 @@ public class Config {
         return getOption(ConfigParameters.LOGGER_LEVEL);
     }
 
+    public static String getDiskDatabaseFile() {
+        return getOption(ConfigParameters.DISK_DATABASE_FILE);
+    }
+
+    public static boolean doCleanDatabaseFile() {
+        return getOption(ConfigParameters.DO_CLEAN_DATABASE_FILE);
+    }
+
     /** Returns true if this server is the Master of a 2PL replica set. */
     public static Boolean isMaster() {
         return txnMode == TransactionMode.TWOPL &&

### 2013-03-20 thebes-code/common/src/main/java/edu/berkeley/thebes/common/config/ConfigParameterTypes.java

@@ -7,7 +7,8 @@ package edu.berkeley.thebes.common.config;
  */
 public interface ConfigParameterTypes {
     public enum PersistenceEngine implements ConfigParameterTypes {
-        MEMORY;
+        MEMORY,
+        LEVELDB;
     }
     
     public enum TransactionMode implements ConfigParameterTypes {

### 2013-03-20 thebes-code/common/src/main/java/edu/berkeley/thebes/common/data/DataItem.java

@@ -15,7 +15,10 @@ public class DataItem implements Comparable<DataItem> {
 	private List<String> transactionKeys;
 
 	public DataItem(byte[] data, Version version, List<String> transactionKeys) {
-		this.data = ByteBuffer.wrap(data);
+        if(data != null)
+		    this.data = ByteBuffer.wrap(data);
+        else
+            this.data = null;
 		this.version = version;
 		this.transactionKeys = transactionKeys;
 	}

### 2013-03-20 thebes-code/common/src/main/java/edu/berkeley/thebes/common/data/Version.java

@@ -27,7 +27,7 @@ public class Version implements Comparable<Version> {
 	
 	public static Version fromThrift(ThriftVersion thriftVersion) {
 	    if (thriftVersion == null)
-	        return NULL_VERSION;
+	        return null;
 	    
 	    long version = thriftVersion.getVersion();
 	    

### 2013-03-20 thebes-code/common/src/main/java/edu/berkeley/thebes/common/persistence/IPersistenceEngine.java

@@ -2,12 +2,14 @@ package edu.berkeley.thebes.common.persistence;
 
 import edu.berkeley.thebes.common.data.DataItem;
 
+import java.io.IOException;
+
 public interface IPersistenceEngine {
-    public void open();
+    public void open() throws IOException;
 
     public boolean put(String key, DataItem value);
 
     public DataItem get(String key);
 
-    public void close();
+    public void close() throws IOException;
 }

### 2013-03-20 thebes-code/common/src/main/java/edu/berkeley/thebes/common/persistence/disk/LevelDBPersistenceEngine.java

@@ -1,0 +1,47 @@
+package edu.berkeley.thebes.common.persistence.disk;
+
+import edu.berkeley.thebes.common.config.Config;
+import edu.berkeley.thebes.common.data.DataItem;
+import edu.berkeley.thebes.common.persistence.IPersistenceEngine;
+import edu.berkeley.thebes.common.thrift.ThriftDataItem;
+import org.iq80.leveldb.*;
+import static org.fusesource.leveldbjni.JniDBFactory.*;
+import java.io.*;
+
+public class LevelDBPersistenceEngine implements IPersistenceEngine {
+    DB db;
+
+    public void open() throws IOException {
+        Options options = new Options();
+        options.createIfMissing(true);
+
+        if(Config.doCleanDatabaseFile()) {
+            try {
+                //not proud, but damn Guava for removing removeRecursively
+                Runtime.getRuntime().exec("rm -rf "+Config.getDiskDatabaseFile());
+            } catch(Exception e) {}
+        }
+
+
+        db = factory.open(new File(Config.getDiskDatabaseFile()), options);
+    }
+
+    public boolean put(String key, DataItem value) {
+        db.put(key.getBytes(), DataItem.toThrift(value).getData());
+        return true;
+    }
+
+    public DataItem get(String key) {
+        byte[] byteRet = db.get(key.getBytes());
+
+        if(byteRet == null)
+            return null;
+
+        ThriftDataItem tdrRet = new ThriftDataItem();
+        return DataItem.fromThrift(tdrRet.setData(byteRet));
+    }
+
+    public void close() throws IOException {
+        db.close();
+    }
+}

### 2013-03-20 thebes-code/hat/client/src/main/java/edu/berkeley/thebes/hat/client/ThebesHATClient.java

@@ -250,7 +250,7 @@ public class ThebesHATClient implements IThebesClient {
         DataItem ret = doGet(key);
 
         if(isolationLevel == IsolationLevel.REPEATABLE_READ) {
-            if(ret != null) {
+            if(ret != null && ret.getTransactionKeys() != null) {
                 /*
                   If the value we just read was part of a transaction that i.) wrote to a key
                   we've already read and ii.) is ordered after the transaction that wrote to

### 2013-03-20 thebes-code/hat/server/src/main/java/edu/berkeley/thebes/hat/server/ThebesHATServer.java

@@ -1,5 +1,6 @@
 package edu.berkeley.thebes.hat.server;
 
+import edu.berkeley.thebes.common.persistence.disk.LevelDBPersistenceEngine;
 import org.slf4j.LoggerFactory;
 
 import javax.naming.ConfigurationException;
@@ -32,6 +33,9 @@ public class ThebesHATServer {
             switch (engineType) {
             case MEMORY:
                 engine = new MemoryPersistenceEngine();
+                break;
+            case LEVELDB:
+                engine = new LevelDBPersistenceEngine();
                 break;
             default:
                 throw new ConfigurationException("unexpected persistency type: " + engineType);

### 2013-03-20 thebes-code/hat/server/src/main/java/edu/berkeley/thebes/hat/server/antientropy/clustering/AntiEntropyServiceRouter.java

@@ -36,7 +36,7 @@ public class AntiEntropyServiceRouter {
         if (Config.isStandaloneServer()) {
             logger.debug("Server marked as standalone; not starting anti-entropy!");
             // TODO: Fix this.
-            throw new IllegalArgumentException("Standalone mode disabled: Need anti-entropy for TA");
+            return;
         }
 
         Uninterruptibles.sleepUninterruptibly(Config.getAntiEntropyBootstrapTime(),

### 2013-03-20 thebes-code/hat/server/src/main/java/edu/berkeley/thebes/hat/server/replica/ReplicaServiceHandler.java

@@ -58,15 +58,19 @@ public class ReplicaServiceHandler implements ReplicaService.Iface {
         if(logger.isTraceEnabled())
             logger.trace("received GET request for key: '"+key+
                          "' requiredVersion: "+ requiredVersion+
-                         ", found verison: " + (ret == null ? null : ret.getVersion()));
+                         ", found version: " + (ret == null ? null : ret.getVersion()));
 
-        if (requiredVersion != null &&
+        if (requiredVersion != null && requiredVersion != Version.NULL_VERSION &&
                 (ret == null || requiredVersion.compareTo(ret.getVersion()) > 0)) {
             ret = dependencyResolver.retrievePendingItem(key, requiredVersion);
 
             if (ret == null)
                 throw new TException(String.format("suitable version was not found! time: %d clientID: %d",
                                                    requiredVersion.getTimestamp(), requiredVersion.getClientID()));
+        }
+
+        if(ret == null) {
+            return new ThriftDataItem().setVersion(Version.toThrift(Version.NULL_VERSION));
         }
 
         return DataItem.toThrift(ret);

### 2013-03-20 writeup/futurework.tex

@@
 \section{Future Work}
+\label{sec:futurework}
 
-VII. Future Work (.75pg)
+\textbf{What is the space of consistency and isolation properties?} In
+this work, we have taxonomized several well-defined and
+well-documented consistency properties. This is useful as it
+contextualizes decades of existing database design and algorithmic
+techniques but does not provide guidance as to what consistency models
+are necessarily the ``strongest''---or admit the fewest number of
+executions---within each availability class. While a recent report
+shows that no model stronger than causal consistency can be achieved
+in a sticky highly available, one-way convergent system~\cite{cac},
+there are infinite highly available and sticky highly available models
+to consider. For example, always returning the value $1$ in response
+to reads to any data item is incomparable to (neither stronger nor
+weaker than) causal consistency: returning $1$ does not respect
+causality, and causal consistency returns values other than
+$1$. Always returning the value $2$ is similarly incomparable, and, as
+an example, we can enumerate models that return all integers and real
+numbers--an uncountably infinite set--along with models that respond
+to every tenth reply and so on.
 
-* What properties are available?
-	* We've focused on ACID Isolation, but there are lots of other models to consider?
-	* Which consistency criteria? Which models? There are infinite, so how do we choose?
-	* Set ourselves up for APEs
+The current state of the art places several useful models within
+various categories in the availability space. This is still a
+practical result---take, for instance, CAP's impact on distributed
+systems of the last decade. But what property defines the boundaries
+between classes of highly available systems? Is there a meta-property
+that exactly characterizes what properties are achievable with high
+availability or sticky high availability? This is an area of ongoing
+research within the HAT project.
 
-* Full exploration of HAT space
-	* Many parameters; $2^10$ models at the least
-	* Need to mitigate the cost of tracking causality with high availability
-		* The cost moving from S-HA (like Eiger) to truly HA (R-HA) seems to be high
-		* e.g., Eiger relies on metadata GC to keep writes small; is O(|keys|) in limit
-			* Not meant as a fault--it's optimized for a particular scenario--but curious to consider
+\textbf{What are the costs and benefits within HAT designs?} There are
+many possible combinations of HAT guarantees, and the performance and
+ease of implementation of a HAT system will likely vary depending on
+which guarantees it attempts to provide. As we have alluded to, the
+cost in moving from a sticky high available system to a truly highly
+available system seems expensive. As a concrete example, the Eiger and
+COPS systems use metadata garbage collection and assume linearizable
+clients within each datacenter in order to keep the overhead of
+providing causal consistency small. Without performing extensive
+client-side caching, these techniques do not seem applicable to a
+highly available implementation. This is not intended as a fault of
+the prior systems---they perform well for the task at hand---but hint
+at the complexity of moving between sticky and truly highly available
+systems.
 
-* Revisit semantics-based concurrency control
-	* Could taxonomize properties like escrow
-	* Can also consider models with SLAs and bounded messaging asynchrony
-	* Our focus is on "classic" distributed systems trade-off
-		* However, lots of interesting hybrids (maybe cite Doug Terry's recent stuff)
+\textbf{Where do semantics-based concurrency control lie in the HAT
+  landscape?} The vast amount of literature on semantics-based
+concurrency control provides hints as to when coordination can be
+avoided but still providing ``strongly consistent'' outcomes. We have
+performed a narrow case study in this work, but a general
+taxonomization of techniques such as escrow transactions and Sagas
+could help answer the challenge of programming in a weakly consistent
+environment. These techniques might be aided by considering models
+with service-level agreements or bounded messaging asynchrony. While
+our focus is on ``classic'' distributed systems trade-offs, there are
+several interesting hybrids to consider.

### 2013-03-20 writeup/hat-analysis.tex

@@
 cut isolation, which prevents Phansoms, \textbf{Phantom Cut
-  Isolation}.
+  Isolation}.\footnote{Oracle provides an isolation level called
+  ``Statement Level Read Consistency''; this is analogous to Phantom
+  Cut Isolation at the level of a single statement. We do not consider
+  this isolation model here as it is subsumed by standard Snapshot
+  Isolation but we believe it is easily adaptable.}
 
@@
 systems~\cite{vogels-defs}, albeit at the cost of true high
-availability or client-side storage.
+availability or client-side storage. Perhaps interestingly,
 
@@
 
-\textbf{}
+The inability to prevent Lost Update means that Consistent Read,
+Snapshot Isolation, and Cursor Stability are all unavailable. The
+inability to prevent Lost Update or Write Skew means that Repeatable
+Read, Conflict Serializability, and One-Copy Serializability are
+unavailable.
 
-* Recency guarantees on data items
-	* CAP
+\subsubsection{Unavailable Recency Guarantees}
+
+Data storage engines make various recency guarantees on reads of data
+items. One of the most famous is linearizability, which states that
+reads will return the last completed write to a data item, and there
+are several other (weaker) variants such as safe and regular register
+semantics. When applied to transactional semantics, the combination of
+one-copy serializability and linearizability is called \textit{strong
+  (or strict) one-copy serializability} or \textit{externally
+  consistent transactions} (e.g., Spanner). It is also common,
+particularly in systems that allow reading from masters and slaves, to
+provide a guarantee such as ``read a version that is no more than five
+seconds out of date'' or similar. Unfortunately, as the CAP Theorem
+demonstrates, linearizability is unavailabile, and any indefinitely
+long partition can cause violation of any recency bound.
 
@@
 \label{sec:hat-summary}
+
+We summarize our results in Table~\ref{table:hatcompared}. A wide
+range of isolation levels are achievable in a highly available system,
+including transactional atomicity, cut isolation, and several session
+guarantees. With sticky availability, a system can achieve PRAM
+consistency, read your writes guarantees, and causal
+consistency. However, many other prominent models, such as Snapshot
+Isolation, One-Copy Serializability, and Strict Serializability cannot
+be achieved due to the inability to prevent Lost Update and Write Skew phenomena.
+
+We illustrate the hierarchy of available, sticky available, and
+unavailble consistency models we have discussed in
+Table~\ref{fig:hatcompared}. Many models are simultaneously
+achievable: for example, a HAT system can provide monotonic reads and
+transactional atomicity or transactional atomicity and read your
+writes and, for maximum semantic richness, can provide all of
+Transactional Atomicity, Predicate Cut Isolation, Monotonic Reads,
+Monotonic Writes, and Writes Follow Reads. An unavailable system might
+provide PRAM and Serializability, as advocated by Salem and
+Daudjee~\cite{needed}.
+
+To the best of our knowledge, this is the first unification of
+database ACID, distributed consistency, and session guarantee
+models. Perhaps satisfactorily, strong one-copy serializability
+subsumes all other models, while considering the (large) power set of
+all compatible models (e.g., only considering client-centric session
+consistency guarantees, the diagram depicts 32 possible HAT
+combinations) hints at the vast expanse of consistency models found in
+the literature. This taxonimization is \textit{not} exhaustive, but we
+believe it lends substantial clarity into the relationships between a
+large subset of the prominent ACID and distributed consistency
+moels. Additional models of read/write transactions that we have
+omitted should be easily classifiable based on the available
+primitives and unavailable anomalies we have already discussed.
 
@@
 HA & Transactional Atomicity (TA), Read Uncommitted (RU), Read
-Committed (RC), Item Cut Isolation (P-CI), Phantom Cut Isolation
+Committed (RC), Item Cut Isolation (P-CI), Predicate Cut Isolation
 (P-CI), Monotonic Reads (MR), Monotonic Writes (MW), Writes Follow
-Reads (WFR)\\\hline Sticky-HA & Read Your Writes (RYW), Causal
-Consistency\\\hline Unavailable & Cursor Stability (CS)\lostupdate,
-Read Consistency (CR)\lostupdate, Snapshot Isolation (SI)\lostupdate,
-Repeatable Read (RR)\lostupdate\rwskew, Serializability
-(S)\lostupdate\rwskew, Causal
-Serializability\lostupdate\rwskew, Strict
-Serializability\lostupdate\rwskew\linearizable \\\hline
+Reads (WFR)\\\hline Sticky-HA & Read Your Writes (RYW), PRAM,
+Causal\\\hline Unavailable & Cursor Stability (CS)\lostupdate,
+Snapshot Isolation (SI)\lostupdate, Repeatable Read
+(RR)\lostupdate\rwskew, Conflict Serializability
+(CSR)\lostupdate\rwskew, One-Copy Serializability
+(1SR)\lostupdate\rwskew, Recency\linearizable, Safe\linearizable,
+Regular\linearizable, Strong 1SR\lostupdate\rwskew\linearizable
+\\\hline
 \end{tabular}
@@
   update\lostupdate, preventing write skew\rwskew, and requiring
-  linearizability\linearizable.}
+  recency guarantees\linearizable.}
 \label{table:hatcompared}
@@
 
- \node[draw=none,fill=none] (ici) at (1.2, 1.6) {\underline{I-CI}};
- \node[draw=none,fill=none] (pci) at (1.2, 2.4) {\underline{P-CI}};
- \node[draw=none,fill=none] (rc) at (1.2, .8) {\underline{RC}};
- \node[draw=none,fill=none] (ru) at (1.2, 0) {\underline{RU}};
+ \node[draw=none,fill=none] (ici) at (1.2, 1.6) {I-CI};
+ \node[draw=none,fill=none] (pci) at (1.2, 2.4) {P-CI};
+ \node[draw=none,fill=none] (rc) at (1.2, .8) {RC};
+ \node[draw=none,fill=none] (ru) at (1.2, 0) {RU};
 
- \node[draw=none,fill=none] (ta) at (0, 0) {\underline{TA}};
- \node[draw=none,fill=none] (mr) at (2.4, 0) {\underline{MR}};
- \node[draw=none,fill=none] (mw) at (3.6, 0) {\underline{MW}};
- \node[draw=none,fill=none] (wfr) at (4.9,0) {\underline{WFR}};
- \node at (6.3,0) [sticky] (ryw) {\underline{RYW}};
+ \node[draw=none,fill=none] (ta) at (0, 0) {TA};
+ \node[draw=none,fill=none] (mr) at (2.4, 0) {MR};
+ \node[draw=none,fill=none] (mw) at (3.6, 0) {MW};
+ \node[draw=none,fill=none] (wfr) at (6.3,0) {WFR};
+ \node at (5,0) [sticky] (ryw) {RYW};
 
- \node at (3.6, 2) [sticky] (causal) {\underline{Causal}};
+ \node[noha](recency) at (7.7, 0) {recency};
+ \node[noha](safe) at (7.7, 1) {safe};
+ \node[noha](regular) at (7.7, 2) {regular};
+ \node[noha](linearizable) at (7.7, 3) {linearizable};
+ \node at (3.6, 2) [sticky] (causal) {causal};
+ \node at (3.6, 1) [sticky] (pram) {PRAM};
  \node[noha] (cs) at (2, 2.8) {CS};
@@
  \node[noha] (si) at (-.4, 2.4) {SI};
- \node[noha] (cr) at (-.4, 1.2) {CR};
- \node[noha] (s) at (-.4, 3.5) {S};
- \node[noha] (causal-s) at (3.6, 3.5) {Causal S};
- \node[noha] (strict-s) at (6, 3.5) {Strict S};
+ \node[noha] (s) at (-.4, 3.5) {CSR};
+ \node[noha] (sr) at (2.2, 3.6) {1SR};
+ \node[noha] (ssr) at (4.95, 3.7) {Strong-1SR};
+
+ \draw [->, red] (recency) -- (safe);
+ \draw [->, red] (safe) -- (regular);
+ \draw [->, red] (regular) -- (linearizable);
+ \draw [->, red] (linearizable) -- (ssr);
+ \draw [->, red] (sr) -- (ssr);
 
@@
 
- \draw [->, blue] (mr) -- (causal);
- \draw [->, blue] (mw) -- (causal);
+ \draw [->, blue] (mr) -- (pram);
+ \draw [->, blue] (mw) -- (pram);
  \draw [->, blue] (wfr) -- (causal);
- \draw [->, blue] (ryw) -- (causal);
+ \draw [->, blue] (ryw) -- (pram);
+ \draw [->, blue] (pram) -- (causal);
 
@@
  \draw [->, red] (si) -- (s);
- \draw [->, red] (ta) -- (cr);
- \draw [->, red] (rc) -- (cr);
- \draw [->, red] (cr) -- (si);
+ \draw [->, red] (ta) -- (si);
  \draw [->, red] (ta) -- (rr);
- \draw [->, red] (s) -- (causal-s);
- \draw [->, red] (causal) -- (causal-s);
- \draw [->, red] (causal-s) -- (strict-s);
+ \draw [->, red] (s) -- (sr);
+ \draw [->, red] (causal) -- (sr);
 
@@
   from Table~\protect\ref{table:hatcompared} compared
-  graphically. Directed edges represent ordering by model strength,
-  while all underlined models can be combined (e.g., the powerset of
-  the set of underlined models are all valid models). The combination
-  of all underlined, non-boxed models represents maximum ACID HAT
-  semantics, while the combination of underlined and boxed models
-  represents maximum ACID sticky HAT models.}
+  graphically. Directed edges represent ordering by model
+  strength. Models that do not share a common ancestor can be
+  simultaneously achieved, and the resulting availability is that of
+  the weakest available model in the combination.}
 \label{fig:hatcompared}
@@
 
-TRASH
+In this section, we discuss several subleties we have mostly elided up
+to this point, specifically addressing model composition,
+transactional atomicity versus linearizability, and stickiness
+requirements.
 
-We can accomplish these three requirements in two ways. First, clients
-can locally cache their reads and writes, updating caches when they
-read admissible replacements from
-replicas~\cite{sessionguarantees}. Second, clients can ensure
-\textit{stickiness} with groups~\cite{vogels-defs} (subject to the
-same groups caveat as before).  For monotonic writes, clients should
-generate increasing IDs for each transaction they perform. Caching and
-ID generation are both available.
+\noindent\textbf{Model Composition} may be non-trivial. Consider the following transactions:
+\begin{align*}
+\small
+T_1 &: w_x(1)~w_y(1)
+\\T_2 &: w_x(2)~w_y(1)
+\\T_3 &: r_x(a)~r_y(b)
+\end{align*}
+If we want to guarantee both cut isolation and transactional atomicity
+and these are the only transactions that are executed by the system,
+then $T_3$ needs to read $a=b=\bot$, $a=b=1$, or $a=b=2$. This means
+that either the implementation should frequently return null
+(definitely undesirable and possibly non-convergent), keep multiple
+versions of each data item (necessitating potentially complicated
+distributed garbage collection), or use pre-declared read sets to
+fetch a consistent cut of keys before the transaction begins to
+execute. Using client-side caching can alleviate some of these
+challenges, but then the system becomes sticky high available.
 
-To provide writes follow reads,
-severs (or clients) can buffer transaction writes until the
-transactions the writes depend on have been written to all
-replicas~\cite{cops}. If we consider a group model, each group can
-independently apply transactions in order via a (possibly sharded)
-log-shipping approach~\cite{causalmemory, cops, eiger, swift}.
+Composition cost varies by property. For instance, Charron-Bost has
+proven that, to capture causality between $N$ communicating processes,
+standard vector-based approaches face an upper bound of $O(N)$. This
+means that, with $100,000$ clients, each write might be accompanied by
+$100,000$ vector timestamps. This is difficult to scale. By
+compromising on availability (e.g., treating a datacenter as a
+linearizable cluster), this overhead can be reduced, but it is much
+cheaper to provide, say, read your writes, compared to causality.
 
-What about existing systems?
-* Database technology developed for single-node systems; gold standard: serializability
-	* Serializability is not actually highly available; give an example!
-	* Much of the database literature presumes serializability and does not consider high availability.
-	* For the literature that doesn't presume serializability, it's not presented in a HA context?
-* Consequence: traditional systems are not optimized for high availability
-	* Consider two-phase locking in a distributed environment
-	* As we will see (forward reference to Evaluation), many existing transactional systems encounter similar difficulties
-* Key question in this paper: What transactional semantics are highly available, and which aren't?
-	* Side note: there are infinite incomparable consistency models (e.g., always return 2, always return 3, …)
-	* Our goal is to unify distributed systems literature with ACID transaction model.
-* Argument: Highly Available Transactional Systems (HATS) require rethinking existing models. This work is a first step.
+\noindent The relationship between \textbf{Transactional Atomicity and
+  Linearizability} is not immediately clear. One might attempt to find
+a mapping between the two: Transactional Atomicity dictates that
+writes to multiple keys across multiple servers are made visible to
+readers all at once, while Linearizability dictates taht writes to a
+single key on multiple servers are made visible to all readers at
+once. The difference is two-old. First, in linearizable (and safe and
+regular) systems, writes are made visible to all clients immediately
+after they finish. With Transactional Atomicity, there is no recency
+requirement. Second, in linearizabile systems, all clients see all
+writes at the same time. With Transactional Atommicity, clients see
+writes at different times depending on which relicas they contact. We
+are not aware of a comparable analog in the distributed systems
+literature. Accordingly, despite apparent similarities, Transactional
+Atomicity is much weaker than Linearizability.
 
-In this sense, HATs are similar to RAID: optimizing for graceful
-handling of a worst-case failure scenario improves average-case
-performance.
+\noindent There is a substantial \textbf{visibility benefit to sticky
+  availability}: writes can safely become visible to readers much more
+quickly in a sticky available system. In the model we discussed, it is
+possible to achieve several properties like monotonic reads in a
+highly available system by waiting until all servers have seen a
+write. However, this leads to horrible visibility---new writes will
+not become visible to clients in the presence of partitions. A client
+that does not want to guarantee read-your-writes (due to the sticky
+availability requirement) may still wish to read other clients' writes
+with timeliness. Accordingly, a system may forgo high availability in
+order to provide better visibility. Alternatively, a system may
+enforce stickiness in the absence of partitions, then ``lose'' session
+guarantees by having clients reconnect to an available replica. This
+latter strategy provides session guarantees between partitions, and
+each partition event may cause a ``regression'' in the desired order
+of partitions, albeit at the benefit of availablilty.
 
-IV. HATS and ACID (4pg)
-
-We will show what semantics are available in HATS as well as which aren't.
-Our strategy will be to build up a set of properties until we reach a point of unavailability.
-We will start with isolation and atomicity since they are well-defined in the literature and most meaty.
-
-A. *What's Available in Isolation and Atomicity?*
-
-* We can prevent Dirty Write phenomena by consistently ordering transactions
-	* This is basically a cross-data item convergency property, as are found in eventually consistent systems!
-* Read Committed
-	* Don't read dirty data by never exposing dirty data to readers!
-* ANSI Repeatable Read => rename to cut isolation
-	* Repeatable Read is a tricky guarantee. In ANSI SQL Spec, rather weak, but in follow-on work, it's pretty strong.
-	* The property that the ANSI SQL spec talks about is about reading a "snapshot" of data items, along with your own writes.
-		* A "snapshot" describes a "consistent cut" across data items; we'll call it "cut isolation"--when paired with below guarantees, makes sense.
-	* This doesn't provide any constraints on writes.
-
-* There are several properties that aren't discussed in the formal literature on isolation guarantees but which are really useful.
-
-* Transactional atomicity: "A" in ACID
-	* Once a transaction reads one of another transactions's writes, its subsequent reads will return the transactions's other writes (or a suitable "later" write).
-	* Be really pedagogical here: "transactional atomicity" versus "distributed linearizability" both called "atomicity"; we actually explore the differences later, in Discussion section.
-	* Talk about implementation here: it's basically uniform eager reliable delivery from distributed systems literature!
-	* Note that this doesn't make recency guarantees (forward reference to later section on "impossibility")
-
-* Session guarantees give useful guarantees within and across transactions
-	* Without them, cut isolation is sort of meaningless: we can always just return NULL!
-	* Several properties: monotonic writes, monotonic reads, writes follow reads, read your writes
-	* We can define them within a transaction only (e.g., transactional monotonic writes) or across subsequent transactions by the same client (e.g., client monotonic writes).
-
-* Session guarantees are tricky
-	* You can implement all but read-your-writes in a R-HA system
-		* Never show a write until all replicas in the system have seen it!
-		* This means that, if a client switches replicas, then the replica will have a satisfying set of writes.
-                * Take S3 as an example of a system where stickiness coudl have helped, but no one really talked about it!
-		* Downside: when do writes become visible? Only when all replicas have seen it!
-		* Side note: assuming that, if replicas can enter and leave, they can only become active once they "catch up" to global lower bound
-			* We leave dynamic replica selection as Future Work (N.B. in APEs!)
-	* Read-your-writes requires S-HA
-		* Example: partition a client from all but one replica, have the client issue a write, which has to commit.
-Next, lift the partition between the client and the other replicas and partition the client and its previous replica.
-It can't read it's own write!
-		* This also means that causality, which is MR+MW+WFR+RYW is unavailable!
-		* S-HA is *doable* but it does result in substantially lowered unavailability unless one caches (e.g., bolt-on causal consistency)
-
-\noindent\textit{Groups:} We can optimize write visibility delay by
-partitioning servers and clients into groups, such that each group
-contains a fully replicated set of data items and all clients within a
-group contact only the servers in the group. Once all transaction
-values are present within a group, values can be installed as
-$d_{good}$ (as opposed to once they are present on all replicas). For
-example, in a multi-datacenter setting, all clients and servers within
-a datacenter may form a group. However, group-based atomicity is not
-highly available according to our definition unless clients and
-servers in a group fail together, as is often assumed in a
-geo-replicated context~\cite{cops, eiger}.
-
-* What do we have?
-	* In Adya, we get up to Causal+Cut Isolation for S-HA, and PL-2 for R-HA.
-	* In distributed systems terminology, in S-HA we have causal consistency but each transactions' updates are a cycle in the happens-before graph!
-		* Stress that this is a big unification of the existing models.
-	* Recap, pedagogically, why this is highly available and what this means: no locks, no central coordination, no need for RTTs; bring back to Section II.
-
-B. *What's not available in Isolation?*
-
-Now, at least in terms of the existing literature, we hit the limits.
-
-Let's sketch out a few conditions, then we'll discuss them later:
-
-Now refer to Adya's chart. No PL-SI, PL-SS, PL-2.99, etc.
-
-C. *What's available in consistency?*
-
-* In general, we can't maintain correctness conditions over arbitrary data items.
-	* This is due to serializability problems.
-	* Uniqueness, for example, is out the window
-
-* We can evaluate locally checkable correctness criteria (e.g., check for null)
-	(QUESTION: should we save this for APEs?)
-
-* However, with semantic knowledge, we can do a little bit better
-	* Commutative updates and logical monotonicity are fine; e.g., write-write conflicts don't matter
-
-Example: In TPC-C New Order, the new order id assignment isn't commutative, tough. but the inventory checking *is*!
-
-D. Durability
-
-* If you want data to survive *F* failures, you'll have to replicate to *F* replicas.
-	* This is pretty easy.
-
-E. Summary
-
-* Left with a bunch of binary properties: 
-R-HA: Prevent Dirty Write, Read Committed, Cut Isolation, Transactional Atomicity, Transactional Monotonic Reads, Transactional Monotonic Writes, Transactional Writes-Follow-Reads, Client Monotonic Reads, Client Monotonic Writes, Client Writes-Follow-Reads
-S-HA: Client Read Your Writes, Transactional Read Your Writes
-
-* Combining all of the above results in the strongest models we've yet seen. No
-
-*Discussion*
-
-Stickiness and PNUTS
-* As an example of a real system going from "CP" to "AP", [PNUTS dropped per-record mastering](\url{http://developer.yahoo.com/blogs/ydn/posts/2010/06/sherpa_update/#4})
-
-* Composition of Properties can be tricky:
-	* Consider ensuring RR and TA:
-		* If you do this naively, you may end up returning lots of nulls
-		* Give example: T1 w(x1) w(y1) T2 w(x2) w(y2) T3 r(x) r(y) => T3 better read x1, y1 or x2, y2!
-	* Some are more expensive; not likely to achieve causality with less than O(clients) metadata required
-		* Note that making stronger assumption about clusters, like linearizability *can* help (but then not HA!)
-
-* Transactional Atomicity versus Linearizability
-	* What's the difference between our TA and distributed systems linearizability? Is there some sort of mapping?
-		* TA: writes to multiple keys are indivisible to readers
-		* Linearizability: writes to single key on multiple servers are indivisible to readers
-	* The differences are two-fold:
-		* In linearizability (and safe and regular registers), writes are visible to clients immediately after they finish
-			* HATS have no recency requirement
-		* In linearizability, all clients see all writes at the same time!
-			* In TA, clients see writes at different times depending on what replicas they contact
-			* No such analog in traditional distributed systems literature on replica consistency
-
-* S-HA Monotonic Reads and Monotonic Writes
-	* In the model we discussed, you can get monotonic reads in a R-HA system by waiting until all replicas see a write.
-	* This leads to poor visibility--writes take a long time to show up. Even if you didn't want to read your writes, you might want to read other peoples' writes!
-	* Accordingly, there are valid reasons for S-HA session guarantees, namely, visibility.
+It is increasingly common in the practically-oriented distributed
+systems literature to assume that clients will be able to remain
+sticky with servers within its datacenter. More generally, if
+client-server sticky pairs are partitioned into disjoint sets, or
+groups, then additional guarantees like transactional atomicity can be
+accomplished with better visibility: writes can be made visible once
+they are present on all other servers within a group (as opposed to
+all servers in the system).

### 2013-03-20 writeup/intro.tex
More writing

@@
 
-Many recent systems have eschewed ACID guarantees.
+Requirements for high availability, predictable low latency, and
+simplicity of coordination protocols have led many recent distributed
+database designs to renounce many guarantees provided by traditional
+database systems. In particular, many systems forgo providing ACID
+guarantees, which often provide single-system image (SSI)
+programmability---at the provable expense of increased coordination
+and, by proxy, unavailability. The formal introduction of the CAP
+Theorem, which proves the incompatiblity of ``always on'' operation
+with guarantees on data item recency in the presence of unreliable
+networks has led to a large number of systems prioritizing high
+availability, which brings with it guaranteed low latency, which is especially
+important in geo-replicated contexts. Accordingly, we observe a dearth
+of systems providing general-purpose multi-operation
+\textit{transactions} over multiple data items with useful semantics.
 
-CAP Theorem says you can't have consistency and availability and also talks about partitions.
+The relationship of the CAP Theorem with ACID semantics is poorly
+understood. While serializability---the gold standard of SSI
+transactions---is unavailable, databases have a long tradition of
+providing weaker isolation and consistency guarantees in order to
+provide greater concurrency and higher performance. While these weaker
+guarantees do not provide SSI for arbitrary general-purpose read/write
+transactions, they are widely employed (Section~\ref{needed}),
+overwhelmingly as the default setting in today's traditional ``ACID''
+and up-and-coming ``NewSQL'' databases and often as the maximum
+guarantees offered. While the prevalence of these weak isolation
+configurations is encouraging, the current literature does not provide
+guidance as to the operation of weakly consistent isolation guarantees
+in a distributed context. Rather, we have a strong understanding of
+weak isolation in the single-server context from which they originated
+as well as several techniques for providing distributed
+serializability but little to no guidance as to which ACID guarantees
+are compatible with Highly Available Transactions (HATS), or can be
+provided with high availability and low latency.
 
-What does CAP have to say about ACID? Recent workshop result on Highly Available Transactions briefly demonstrated that some ACID guarantees are indeed achievable in a highly available manner.
+This work attempts to bridge this gap between ACID properties and
+highly available systems. We analyze a range of existing ACID and
+distributed consistency models in order to provide a formal
+taxonomization of which models are achievalbe in a HAT system.  While
+several guarantees, like serializability and Snapshot Isolation, are
+unachievable due to reliance on preventing Lost Update and Write Skew
+phenomena, many other ACID guarantees such as providing Read Committed
+isolation, preventing Fuzzy Read phenomena, and ensuring atomicity of
+transactional updates are achievable. We also attempt to bridge the
+often distant worlds of distributed system consistency and ACID
+database consistency by analyzing the relationship between so-called
+session guarantees and ``register'' semantics. In performing this
+taxonomization, we also introduce a hierarchy of highly available
+system guarantees as, for example, some guarantees require
+``stickiness,'' or fate-sharing between clients and database
+servers. We believe that this analysis is the first that has been
+performed over such a broad range of ACID and distributed consistency
+models.
 
-Many ACID systems were designed either i.) for single systems or ii.)
-to provide serializability. The combination of weak isolation and
-distributed systems necessitates further study and
-design. \textit{What is the benefit of high availability and what is
-  the cost?}
-
-Highly Available Transactions are the class of multi-object operations
-and their respective semantics that can be provided with high
-availability.
-
-Weak isolation has a cost: in the words of one database pioneer, you
-``fall off a cliff'' when it comes to reasoning about your
-program. However, it also has a benefit: availability and latency.
-
-Our goal in this paper is to objectively map the space of trade-offs and
-
- This paper fills in the remaining gaps.
-
-What does high availability even mean in a transaction environment?
-
-What models are achievable, and which aren't?
+While this model classification is useful, we seek to understand the
+implications of choosing to employ high availability. HATs provide
+high availability and low latency but suffer from weakened semantic
+guarantees. Traditional ACID properties provide strong semantic
+guarantees but do not provide high availability or low latency. We
+attempt to objectively study these virtues and limitations by
+performing an extensive survey of practitioner accounts and research
+literature, performing experimental analyses on modern cloud
+infrastructure, and analyzing representative applications for their
+semantic requirements. While our results are not definitive, they
+suggest that HATs offer a one to three order of magnitude latency
+decrease compared to traditional protocols, while providing acceptable
+semantics for programs with monotonic and commutative
+updates. Similarly, HAT guarantees can be used to ensure that
+arbitrary foreign key constraints are satisfied but not necessarily
+uniqueness constraints. For programs with non-monotonic transactional
+requirements, HATs may fall short, necessitating greater coordination
+strategies. Speed and availability are not without a price, but we
+attempt to estimate the size of the tag. Succintly, \textit{what is
+  benefits do highly available transactions offer and what is their
+  cost?}
 
@@
 Achievable properties include transactional atomicity, variants of repeatable read isolation, and session guarantees.
-We also describe unavailable properties, like preventing Lost Update and Read and Write Skew.
+We also describe unavailable properties, like preventing Lost Update and Write Skew.
 We believe that this is the first such characterization in the literature.

### 2013-03-20 writeup/motivation.tex

@@
 a host of recent and highly public cloud computing incidents highlight
-partition behavior. In April 2011, a network misconfiguration led to a
-twelve-hour long series of outages across the Amazon EC2 and RDS
+partition behavior. As an example of an actual system rearchitecture,
+less than one year after its announcement, Yahoo!'s PNUTS developers
+explicitly added a support for weaker, highly available operation,
+explaining that ``strict adherence [to strong consistency] leads to
+difficult situations under network partitioning or server
+failures...in many circumstances, applications need a relaxed
+approach''~\cite{pnuts-update}.
+
+Provider outages are often publicly observable. In April 2011, a
+network misconfiguration led to a twelve-hour long series of outages
+across the Amazon EC2 and RDS
 services~\cite{amazon-netpartition}. Subsequent misconfigurations and

### 2013-03-20 thebes-code/common/src/main/java/edu/berkeley/thebes/common/clustering/RoutingHash.java
Decrease TA messaging a lot

@@ -2,6 +2,12 @@ package edu.berkeley.thebes.common.clustering;
 
 public class RoutingHash {
     public static int hashKey(String key, int numServers) {
-        return Math.abs(key.hashCode()) % numServers;
+        try {
+            return Integer.parseInt(""+key.charAt(key.length()-1));
+        } catch (Exception e) {
+            return 0;
+        }
+        // TODODODO
+//        return Math.abs(key.hashCode()) % numServers;
     }
 }

### 2013-03-20 thebes-code/common/src/main/java/edu/berkeley/thebes/common/config/Config.java

@@ -254,7 +254,9 @@ public class Config {
     }
 
     public static List<ServerAddress> getServersInCluster() {
-        return clusterServers;
+        return Lists.<ServerAddress>newArrayList(new ServerAddress(1, 1, "host1", 123),new ServerAddress(1, 1, "host2", 123),new ServerAddress(1, 1, "host3", 123));
+        // TODODODO
+//        return clusterServers;
     }
 
     public static List<ServerAddress> getSiblingServers() {

### 2013-03-20 thebes-code/common/src/main/java/edu/berkeley/thebes/common/thrift/ServerAddress.java

@@ -1,5 +1,7 @@
 package edu.berkeley.thebes.common.thrift;
 
+import com.google.common.base.Objects;
+
 /** Simply stores all the attributes of some external server. */
 public class ServerAddress {
     private final int clusterID;
@@ -30,5 +32,21 @@ public class ServerAddress {
     @Override
     public String toString() {
         return ip + ":" + port + " [" + clusterID + ", " + serverID + "]";
+    }
+    
+    public boolean equals(Object o) {
+        if (o == null || !(o instanceof ServerAddress)) {
+            return false;
+        }
+        
+        ServerAddress other = (ServerAddress) o;
+        return Objects.equal(clusterID, other.getClusterID())
+                && Objects.equal(serverID, other.getServerID())
+                && Objects.equal(ip, other.getIP())
+                && Objects.equal(port, other.getPort());
+    }
+    
+    public int hashCode() {
+        return Objects.hashCode(clusterID, serverID, ip, port);
     }
 }

### 2013-03-20 thebes-code/hat/common/src/main/java/edu/berkeley/thebes/hat/common/thrift/AntiEntropyService.java

@@ -36,7 +36,7 @@ public class AntiEntropyService {
 
     public void put(String key, edu.berkeley.thebes.common.thrift.ThriftDataItem value) throws org.apache.thrift.TException;
 
-    public void ackDependentWriteInPending(String myKey, edu.berkeley.thebes.common.thrift.ThriftVersion version) throws org.apache.thrift.TException;
+    public void ackTransactionPending(edu.berkeley.thebes.common.thrift.ThriftVersion transactionId) throws org.apache.thrift.TException;
 
   }
 
@@ -44,7 +44,7 @@ public class AntiEntropyService {
 
     public void put(String key, edu.berkeley.thebes.common.thrift.ThriftDataItem value, org.apache.thrift.async.AsyncMethodCallback<AsyncClient.put_call> resultHandler) throws org.apache.thrift.TException;
 
-    public void ackDependentWriteInPending(String myKey, edu.berkeley.thebes.common.thrift.ThriftVersion version, org.apache.thrift.async.AsyncMethodCallback<AsyncClient.ackDependentWriteInPending_call> resultHandler) throws org.apache.thrift.TException;
+    public void ackTransactionPending(edu.berkeley.thebes.common.thrift.ThriftVersion transactionId, org.apache.thrift.async.AsyncMethodCallback<AsyncClient.ackTransactionPending_call> resultHandler) throws org.apache.thrift.TException;
 
   }
 
@@ -81,17 +81,16 @@ public class AntiEntropyService {
       sendBase("put", args);
     }
 
-    public void ackDependentWriteInPending(String myKey, edu.berkeley.thebes.common.thrift.ThriftVersion version) throws org.apache.thrift.TException
+    public void ackTransactionPending(edu.berkeley.thebes.common.thrift.ThriftVersion transactionId) throws org.apache.thrift.TException
     {
-      send_ackDependentWriteInPending(myKey, version);
+      send_ackTransactionPending(transactionId);
     }
 
-    public void send_ackDependentWriteInPending(String myKey, edu.berkeley.thebes.common.thrift.ThriftVersion version) throws org.apache.thrift.TException
+    public void send_ackTransactionPending(edu.berkeley.thebes.common.thrift.ThriftVersion transactionId) throws org.apache.thrift.TException
     {
-      ackDependentWriteInPending_args args = new ackDependentWriteInPending_args();
-      args.setMyKey(myKey);
-      args.setVersion(version);
-      sendBase("ackDependentWriteInPending", args);
[…]

### 2013-03-20 thebes-code/hat/server/src/main/java/edu/berkeley/thebes/hat/server/antientropy/AntiEntropyServiceHandler.java

@@ -46,9 +46,7 @@ public class AntiEntropyServiceHandler implements AntiEntropyService.Iface {
     }
 
     @Override
-    public void ackDependentWriteInPending(String myKey,// String ackedKey,
-            ThriftVersion version) throws TException {
-        dependencyResolver.dependentWriteAcked(myKey, null /* ackedKey */,
-                Version.fromThrift(version));
+    public void ackTransactionPending(ThriftVersion transactionId) throws TException {
+        dependencyResolver.ackTransactionPending(Version.fromThrift(transactionId));
     }
 }

### 2013-03-20 thebes-code/hat/server/src/main/java/edu/berkeley/thebes/hat/server/antientropy/clustering/AntiEntropyServiceRouter.java

@@ -20,6 +20,7 @@ import edu.berkeley.thebes.hat.common.thrift.ThriftUtil;
 import edu.berkeley.thebes.hat.server.dependencies.PendingWrite;
 
 import java.util.List;
+import java.util.Set;
 import java.util.concurrent.LinkedBlockingQueue;
 import java.util.concurrent.TimeUnit;
 
@@ -72,11 +73,11 @@ public class AntiEntropyServiceRouter {
     /** Stores the writes we receive and need to forward to all siblings */
     private final LinkedBlockingQueue<QueuedWrite> writesToForwardSiblings;
     /** Stores the writes we've put into pending, and need to notify all dependent neighbors. */
-    private final LinkedBlockingQueue<PendingWrite> pendingWritesToAnnounce;
+    private final LinkedBlockingQueue<QueuedTransactionAnnouncement> pendingTransactionAnnouncements;
     
     public AntiEntropyServiceRouter() {
         this.writesToForwardSiblings = Queues.newLinkedBlockingQueue();
-        this.pendingWritesToAnnounce = Queues.newLinkedBlockingQueue();
+        this.pendingTransactionAnnouncements = Queues.newLinkedBlockingQueue();
     }
     
     /** Our cluster got a new write, forward to the replicas in other clusters. */
@@ -98,24 +99,20 @@ public class AntiEntropyServiceRouter {
         }
     }
 
-    /** Our replica got a write, announce it to others in this cluster who depend on it. */
-    public void announcePendingWrite(PendingWrite write) {
-        pendingWritesToAnnounce.add(write);
+    /** Announce that a transaction is ready to some set of servers. */
+    public void announceTransactionReady(Version transactionID, Set<Integer> servers) {
+        pendingTransactionAnnouncements.add(
+                new QueuedTransactionAnnouncement(transactionID, servers));
     }
     
     /** Actually does the announcement! Called in its own thread. */
     private void announceNextQueuedPendingWrite() {
         try {
-            PendingWrite writeToAnnounce = pendingWritesToAnnounce.take();
-            String writeKey = writeToAnnounce.getKey();
-            Version writeVersion = writeToAnnounce.getVersion();
+            QueuedTransactionAnnouncement announcement = pendingTransactionAnnouncements.take();
             
-            // TODO: Efficiency! Group keys, take multiple writes at a time.
-            for (String dependentKey : writeToAnnounce.getValue().getTransactionKeys()) {
-                AntiEntropyService.Client neighborClient = neighborClients.get(
[…]

### 2013-03-20 thebes-code/hat/server/src/main/java/edu/berkeley/thebes/hat/server/dependencies/DependencyResolver.java

@@ -3,7 +3,6 @@ package edu.berkeley.thebes.hat.server.dependencies;
 import org.slf4j.Logger;
 import org.slf4j.LoggerFactory;
 
-import com.google.common.collect.ComparisonChain;
 import com.google.common.collect.Maps;
 
 import edu.berkeley.thebes.common.data.DataItem;
@@ -11,10 +10,11 @@ import edu.berkeley.thebes.common.data.Version;
 import edu.berkeley.thebes.common.persistence.IPersistenceEngine;
 import edu.berkeley.thebes.hat.server.antientropy.clustering.AntiEntropyServiceRouter;
 
-import java.util.Iterator;
 import java.util.Set;
 import java.util.concurrent.ConcurrentMap;
 import java.util.concurrent.ConcurrentSkipListSet;
+import java.util.concurrent.atomic.AtomicBoolean;
+import java.util.concurrent.atomic.AtomicInteger;
 
 /*
 The following class does most of the heavy lifting for the HAT
@@ -46,39 +46,49 @@ waitInQueueForDependency() need to be notified and so
 notifyNewLocalWrite() is called.
 */
 
-public class DependencyResolver implements PendingWrite.WriteReadyCallback {
+public class DependencyResolver {
     private static Logger logger = LoggerFactory.getLogger(DependencyResolver.class);
     
     private final AntiEntropyServiceRouter router;
     private final IPersistenceEngine persistenceEngine;
     private final ConcurrentMap<String, Set<PendingWrite>> pendingWritesMap;
-    private final ConcurrentMap<String, Set<Ack>> unresolvedAcksMap;
+    private final ConcurrentMap<Version, TransactionQueue> pendingTransactionsMap;
+    private final ConcurrentMap<Version, AtomicInteger> unresolvedAcksMap;
     
     public DependencyResolver(AntiEntropyServiceRouter router,
             IPersistenceEngine persistenceEngine) {
         this.persistenceEngine = persistenceEngine;
         this.router = router;
         this.pendingWritesMap = Maps.newConcurrentMap();
+        this.pendingTransactionsMap = Maps.newConcurrentMap();
         this.unresolvedAcksMap = Maps.newConcurrentMap();
     }
 
     public void addPendingWrite(String key, DataItem value) {
-        pendingWritesMap.putIfAbsent(key, new ConcurrentSkipListSet<PendingWrite>());
+        Version version = value.getVersion();
         
[…]

### 2013-03-20 thebes-code/hat/server/src/main/java/edu/berkeley/thebes/hat/server/dependencies/DependencyResolverTest.java

@@ -11,8 +11,10 @@ import edu.berkeley.thebes.hat.server.antientropy.clustering.AntiEntropyServiceR
 
 import java.nio.ByteBuffer;
 import java.util.Map;
+import java.util.Set;
 import java.util.concurrent.atomic.AtomicLong;
 
+import junit.framework.AssertionFailedError;
 import junit.framework.TestCase;
 
 public class DependencyResolverTest extends TestCase {
@@ -33,40 +35,74 @@ public class DependencyResolverTest extends TestCase {
     
     public void testBasic() {
         Version xact1 = getTransactionId();
-        router.expect("hello");
+        router.expect(xact1);
         resolver.addPendingWrite("hello",
                 makeDataItem(xact1, "World!", "depKey1", "depKey2"));
+        router.expect(null);
         assertPending("hello", "World!", xact1);
         
-        resolver.dependentWriteAcked("hello", "depKey1", xact1);
+        resolver.ackTransactionPending(xact1);
         assertPending("hello", "World!", xact1);
         
-        resolver.dependentWriteAcked("hello", "depKey2", xact1);
+        resolver.ackTransactionPending(xact1);
         assertGood("hello", "World!", xact1);
+    }
+
+    public void testWaitForAllSelf() {
+        Version xact1 = getTransactionId();
+        router.expect(xact1);
+        resolver.addPendingWrite("hello",
+                makeDataItem(xact1, "World!", "hello", "depKey0", "depKey1", "depKey2"));
+        
+        try {
+            router.expect(null);
+            fail();
+        } catch (AssertionFailedError e) { /* ok */ }
+        
+        assertPending("hello", "World!", xact1);
+        
+        resolver.ackTransactionPending(xact1);
+        assertPending("hello", "World!", xact1);
+
+        resolver.addPendingWrite("depKey0",
+                makeDataItem(xact1, "Worldz!", "hello", "depKey0", "depKey1", "depKey2"));
+        router.expect(null);
+
+        assertPending("hello", "World!", xact1);
+        assertPending("depKey0", "Worldz!", xact1);
+        
+        resolver.ackTransactionPending(xact1);
+        assertPending("hello", "World!", xact1);
+        assertPending("depKey0", "Worldz!", xact1);
+        
+        resolver.ackTransactionPending(xact1);
+        assertGood("hello", "World!", xact1);
+        assertGood("depKey0", "Worldz!", xact1);
+
     }
     
     public void testPrematureAck() {
         Version xact1 = getTransactionId();
         
[…]

### 2013-03-21 thebes-code/common/src/main/java/edu/berkeley/thebes/common/clustering/RoutingHash.java
Oops, get rid of the testing stuff

@@ -2,12 +2,6 @@ package edu.berkeley.thebes.common.clustering;
 
 public class RoutingHash {
     public static int hashKey(String key, int numServers) {
-        try {
-            return Integer.parseInt(""+key.charAt(key.length()-1));
-        } catch (Exception e) {
-            return 0;
-        }
-        // TODODODO
-//        return Math.abs(key.hashCode()) % numServers;
+        return Math.abs(key.hashCode()) % numServers;
     }
 }

### 2013-03-21 thebes-code/common/src/main/java/edu/berkeley/thebes/common/config/Config.java

@@ -254,9 +254,7 @@ public class Config {
     }
 
     public static List<ServerAddress> getServersInCluster() {
-        return Lists.<ServerAddress>newArrayList(new ServerAddress(1, 1, "host1", 123),new ServerAddress(1, 1, "host2", 123),new ServerAddress(1, 1, "host3", 123));
-        // TODODODO
-//        return clusterServers;
+        return clusterServers;
     }
 
     public static List<ServerAddress> getSiblingServers() {

### 2013-03-21 thebes-code/hat/server/src/main/java/edu/berkeley/thebes/hat/server/dependencies/DependencyResolver.java
Fix exception business`

@@ -1,5 +1,6 @@
 package edu.berkeley.thebes.hat.server.dependencies;
 
+import org.apache.thrift.TException;
 import org.slf4j.Logger;
 import org.slf4j.LoggerFactory;
 
@@ -106,15 +107,11 @@ public class DependencyResolver {
             return;
         }
         
-        try {
-            for (PendingWrite write : queue.pendingWrites) {
-                persistenceEngine.put(write.getKey(), write.getValue());
-                pendingWritesMap.get(write.getKey()).remove(write);
-            }
-            pendingTransactionsMap.remove(queue.version);
-        } catch (TException e) {
-            logger.error("error in writeReady: ", e);
+        for (PendingWrite write : queue.pendingWrites) {
+            persistenceEngine.put(write.getKey(), write.getValue());
+            pendingWritesMap.get(write.getKey()).remove(write);
         }
+        pendingTransactionsMap.remove(queue.version);
     }
     
     public DataItem retrievePendingItem(String key, Version version) {

### 2013-03-21 thebes-code/hat/server/src/main/java/edu/berkeley/thebes/hat/server/dependencies/DependencyResolver.java
Weird null case

@@ -138,6 +138,7 @@ public class DependencyResolver {
         // No currently known PendingWrites wanted our ack!
         // Hopefully we'll soon have one that does, so keep it around.
         unresolvedAcksMap.putIfAbsent(transactionId, new AtomicInteger(0));
+        logger.debug("Hi: " + unresolvedAcksMap.get(transactionId));
         unresolvedAcksMap.get(transactionId).incrementAndGet();
     }
     

### 2013-03-21 writeup/evaluation.tex

@@
 
+Thus far, we have largely ignored the question of which applications
+actually require HAT semantics versus traditional ACID guarantees. As
+we summarized in Section~\ref{needed}, the main cost of providing HAT
+updates is the inability to prevent Lost Update, Write Skew, and
+provide recency bounds. In this Section, we attempt to understand when
+these guarantees matter both abstractly and by the use of a
+representative customer order transaction application patterned off of
+TPC-C.
+
+Recent work on the CALM Theorem, Lattice-based state update, and
+Commutative and Replicated Data Types demonstrates that, if updates
+logically commute, then they can be performed in different orders at
+different locations in a distributed system. Accordingly, as long as
+all writes are delivered to all replicas, then a system executing
+monotonic logic with commutative operators should not suffer from
+application-level consistency anomalies as a result of Lost Update or
+Write Skew anomalies. However, applications with non-monotonic state
+mutation will not, in general, be able to maintain application-level
+consistency constraints. Similarly, applications requiring
+write visibility between clients should opt for unavailability instead
+of possibly missing a write visibility deadline.
+
+Absent a necessary and sufficient result describing exactly which
+application logics can be achieved with high availability and to make
+these requirements more concrete, we consider an example application:
+the TPC-C benchmark. Our goal is \textit{not} to provide a performance
+analysis, as any HAT configuration will be non-compliant. Rather, we
+examine the five queries contained in the benchmarks---ostensibly
+representative of a standard transaction processing application---to
+understand which are achievable in a HAT context.
+
+Integrity constraints:
+--Mostly SUM, AVG, COUNT
+--This is okay and foreign key constraints are provided by TA
+
+New Order:
+--Lookup of warehouse codes is fine, are not mutated
+--Assigning a version order is not okay, but we can still get uniqueness
+--Stock adjustment is monotonic and can't go below zero due to replenish. However, might overstock!
+
+Payment:
+--Monotonic, simply increase the number of each respective field
+
+Delivery:
+--Problematic; is non-monotonic, removing from the order table
+--Without additional semantics, have to block
+--CALM analysis would catch this, enforce coordination
+--Does \textit{not} block new New Order transactions
+
+Order Status:
+--Non-monotonic
+--Read your writes may be problematic
+
+While TPC-C isolation tests are for lock-based models (and HATs are,
+by requirement, non-locking), consider TPC-E tests. Modifying the
+description and image together is fine, etc.
+
+\subsection{Experimental Costs}
+
 2.) We ran YCSB on EC2 (and maybe TPC-C if we have time) with a few models:
@@
 	-Not meant as an exhaustive study, but validates our intuitions. Coordination costs of non-HAT systems are unaccounted for, will only go up, and we haven't even talked about availability.
-
-\subsection{Experimental Costs}

### 2013-03-22 thebes-code/hat/server/src/main/java/edu/berkeley/thebes/hat/server/replica/ReplicaServiceHandler.java
npe

@@ -40,7 +40,7 @@ public class ReplicaServiceHandler implements ReplicaService.Iface {
         antiEntropyRouter.sendWriteToSiblings(key, valueThrift);
 
         // TODO: Hmm, if siblings included us, we wouldn't even need to do this...
-        if (value.getTransactionKeys().isEmpty()) {
+        if (value.getTransactionKeys() == null || value.getTransactionKeys().isEmpty()) {
             persistenceEngine.put(key, value);
         } else {
             dependencyResolver.addPendingWrite(key, value);

### 2013-03-22 thebes-code/hat/server/src/main/java/edu/berkeley/thebes/hat/server/antientropy/AntiEntropyServiceHandler.java
NPE2

@@ -38,7 +38,7 @@ public class AntiEntropyServiceHandler implements AntiEntropyService.Iface {
                     ThriftDataItem valueThrift) throws TException{
     	logger.trace("Received anti-entropy put for key " + key);
         DataItem value = new DataItem(valueThrift);
-        if (value.getTransactionKeys().isEmpty()) {
+        if (value.getTransactionKeys() == null || value.getTransactionKeys().isEmpty()) {
             persistenceEngine.put(key, value);
         } else {
             dependencyResolver.addPendingWrite(key, value);

### 2013-03-22 writeup/evaluation.tex
meh; sort of crappy TPC-C writeup

@@
 
-Integrity constraints:
---Mostly SUM, AVG, COUNT
---This is okay and foreign key constraints are provided by TA
+TPC-C consists of five transactions, capturing operation of a
+real-world warehouse, including sales, payments, and deliveries. We
+analyze each transaction and its interactions below.
 
-New Order:
---Lookup of warehouse codes is fine, are not mutated
---Assigning a version order is not okay, but we can still get uniqueness
---Stock adjustment is monotonic and can't go below zero due to replenish. However, might overstock!
+\begin{itemize}
+\item \textbf{New-Order} places an order for a variable quantity of
+  data items, updating warehouse stock as needed. The New-Order
+  transaction selects a sales district, assigns the new order an order
+  ID number, adjusts the remaining warehouse stock, and writes a
+  placeholder entry for the pending order. There are two mutations in
+  this transaction which we must be careful of.
 
-Payment:
---Monotonic, simply increase the number of each respective field
+  First, assigning an order ID number is challenging. We can easily
+  assign a \textit{unique} order ID, by, say, concatenating the user
+  ID with a timestamp and server ID. However, the TPC-C specification
+  requires that order numbers be sequentially assigned, which we
+  cannot guarantee in HATs due to Lost Update. Accordingly, while HATs
+  cannot provide compliant TPC-C execution due to a non-monotonic
+  counter increase, they can allow placement of new orders, which is a
+  commutative and monotonic operation
 
-Delivery:
---Problematic; is non-monotonic, removing from the order table
---Without additional semantics, have to block
---CALM analysis would catch this, enforce coordination
---Does \textit{not} block new New Order transactions
+  Second, decrementing inventory counts is a potentially risky
+  operation as, without coordination, inventory counts may become
+  negative. However, TPC-C automatically ``refills'' inventory counts
+  (by 91 items) when there are 10 or fewer items
+  remaining. Accordingly, even in the presence of concurrent new
+  orders to the same item, the stock will never fall below
+  zero. However, if the system propagates ``refill'' operations
+  between replicas, then an $R$ replica system might end up with
+  $91(R-1)$ more stock than a single-site implementation.
 
-Order Status:
---Non-monotonic
---Read your writes may be problematic
+\item \textbf{Payment} updates running balances for warehouses,
+  districts, and customer records as well as providing an audit
+  trail. All balance operations can be made commutative, so the
+  transaction can execute under HATs.
 
-While TPC-C isolation tests are for lock-based models (and HATs are,
-by requirement, non-locking), consider TPC-E tests. Modifying the
-description and image together is fine, etc.
+\item \textbf{Order-Status} is a read-only transaction and can be
+  safely executed under HATs. The TPC-C specification does not require
+  recency bounds but, in practice, they may be useful. If a client is
+  sticky with her replica set, she will read her own writes, which may
+  be sufficient.
+
+\item \textbf{Delivery} represents the fulfillment of a pending order
+  created by a New Order transaction: it deletes the order from the
+  pending list, updates the customer's balance, updates the order's
+  carrier ID and delivery time, and updates the customer balance. The
+  challenge in delivery comes in deletion, which is non-monotonic: an
+  item might be removed twice from the pending table. If delivery can
+  be made idempotent---that is, if multiple Delivery transactions have
+  the same effect as one---then multiple clients can execute the same
+  Deliveries without fatal consequences.
+
+  In practice, given that a physical shipment is likely only delivered
+  once, it is unlikely that multiple agents will attempt a Delivery
+  transaction for the same order simultaneously. However, in this
+  case, we have simply moved the non-monotonicity to the real world:
+  the carrier that picks up the package for an order must ensure that
+  she is the only carrier who has done so. While we can program around
+  the other operations in the order, HATs cannot--without the help of
+  business-specific processes--serve Delivery transactions. That said,
+  New Orders may be placed.
+
+\item \textbf{Stock-Level} is another read-only transaction and can be
+  executed without any non-HAT guarantees, with the same caveats
+  regarding recency.
+\end{itemize}
+
+Throughout execution, TPC-C requires the maintainence of several
+integrity constraints. For example, Consistency Condition 1 (3.3.2.1)
+requires that each warehouse's sales count must reflect the sum of its
+subordinate sales districts. This integrity constraint spans two
+tables but, given the ability to update both tables atomically via
+transactional atomicity, can be easily maintained. Consistency
+Conditions 4 through 12 (3.3.2.4-12) can similarly be satisfied by
+applying updates atomically across tables. Consistency Conditions 2
+and 3 (3.3.2.2-3) concern order ID assignment and are more
+problematic: effectively, each New Order placed by a customer in a
+district must have an ID that is one the prior order in that
+district. This requires preventing Lost Update within each district's
+table. However, as we have seen, the required transactions can execute
+without these two conditions.
+
+In summary, many---but not all---TPC-C transactions are well-served by
+HATs. The two problematic transactions---New Order and Payment---rely
+on non-monotonic state updates. The former transaction can be modified
+to ensure ID uniqueness at the expense of dense ID ordering, while the
+latter is an inherently non-monotonic action requiring external
+compensation or stronger consistency protocols. We expect that,
+especially for read-dominated workloads found in many online services,
+HAT guarantees may suffice for many applications. Finally, while TPC-C
+is not subject to write skew anomalies, we note that many TPC-E
+isolation tests, such as one that requires simultaneously modifying a
+product description and its image, are satisfied by HATs.
 

### 2013-03-23 writeup/hat-vldb.tex
woo; intro pass--sort of crappy
text editing pass

--- 2013-03-18 writeup/hat-vldb.tex
@@
 \begin{abstract}
-foo bar baz
+Given the presence of network latency and the possibility of partial
+failure and network partitions, many modern distributed systems eschew
+transactional functionality, which provides strong semantic guarantees
+on groups of multiple operations over multiple data items. In this
+work, we consider the problem of providing transactional guarantees
+that are highly available, or do not suffer from unavailability during
+system partitions or typically incur high network latency.  We
+introduce a new taxonomy of highly available algorithms and analyze a
+combination of ACID and distributed consistency guarantees to identify
+which can and cannot be achieved by Highly Available Transaction (HAT)
+systems. In doing so, we analytically and experimentally quantify the
+benefit of providing high availability as well as the associated
+semantic costs.
 \end{abstract}
@@
 % vldb_sample.bib is the name of the Bibliography in this case
+\footnotesize
 \bibliography{hat-vldb}  
@@
 
-\begin{appendix}
+%\begin{appendix}
 
-\end{appendix}
+%\end{appendix}
 

### 2013-03-23 writeup/motivation.tex
text editing pass

--- 2013-03-19 writeup/motivation.tex
@@
-\section{Partitions and Latency}
+\section{Motivation}
+\label{sec:motivation}
 
-Peter Deutsch begins his list of ``Fallacies of Distributed
-Computing'' with two concerns: ``\textit{1.)}  The network is
-reliable. \textit{2.)} Latency is zero''~\cite{fallacies-deutsch}. On
-a single-node system, communication systems are robust and relatively
-fast, and most systems do not need to prepare for partial component
-failure. However, as Deutsch points out, these assumptions are no
-longer valid in a distributed setting. To extend Jim Gray's storage
-latency analogy, if memory is the distance from San Diego to Olympia
-(1.5 hours), and disk is as ``far away'' as Pluto (2 years), in a
-distributed context, another computer can be even farther---nearing or
-even worse than Gray's location of tape access near Andromeda (2,000
-years)---with the added concern that messages can be lost in transit,
-destroyed by ``black holes'' in the form of network
-partitions. Accordingly, accounting for partial failure and
-communication expense is core to distributed systems and algorithm
-design; ignoring them ``cause[s] big trouble and painful learning
+Peter Deutsch begins his classic list of ``Fallacies of Distributed
+Computing'' with two concerns fundamental to distributed database
+systems: ``\textit{1.)}  The network is reliable. \textit{2.)} Latency
+is zero''~\cite{fallacies-deutsch}. On a single server, communication
+systems are robust and relatively fast, and most single-node systems
+do not need to prepare for partial component failure. However, these
+assumptions are no longer valid in a distributed setting. To extend
+Jim Gray's storage latency analogy, if memory is 1.5 hours from the
+CPU, disk is 2 years away, especially across wide-area networks,
+remote server accesses can exceed 30 years, with the added concern
+that messages can be lost in transit or delayed indefinitely by
+network partitions~\cite{gray-rules}. Accordingly, ignoring network
+failures and latency ``cause[s] big trouble and painful learning
 experiences''~\cite{fallacies-deutsch}.
 
-While prominent members of the database community have publicly
-speculated about the unimportance of network partitions in distributed
-database design~\cite{stonebraker2010errors}, there is mounting
+In this section, we attempt to quantify the impact of network
+partitions and latency in real-world deployments. While the importance
+of network behavior is a topic of considerable speculation within the
+database community~\cite{stonebraker2010errors}, there is mounting
 evidence---both anecdotal and experimental---that partitions occur and
-latency is non-negligible. In this section, we briefly outline this
-evidence and attempt to rigorously quantify the costs of ignoring
-network unreliability and latency.
+latency is non-negligible.
 
@@
 simultaneously maintaining a correct ``global'' view of system state
-while simultaneously providing operation to all of its clients. The
-rate at which a network experiences partitions and the severity and
-duration of each partition depends on the given network deployment,
-hardware, topology, administration, and a host of other
-factors. However, given the \textit{possibility} of partitions, a
-system must be prepared to either lose availability of operations or
-its ability to reason about global state.
+while simultaneously providing operation to all of its
+clients~\cite{davidson-survey}. For networks that do not partition,
+maintaining semantic guarantees and available operation is
+simple~\cite{stonebraker2010errors}. However, given the
+\textit{possibility} of partitions, a system must be prepared to
+either lose availability of operations or its ability to reason about
+global state.
 
-For networks that do not partition, maintaining semantic guarantees
-and available operation is
-simple~\cite{stonebraker2010errors}. However, this is often not the
-case: according to James Hamilton, Vice President and Distinguished
-Engineer on the Amazon Web Service team, ``network partitions should
-be rare but net gear continues to cause more issues than it
-should''~\cite{hamilton-partitions}. Anecdotal evidence from companies
-such as Google~\cite{dean-keynote} confirm Hamilton's statement, while
-a host of recent and highly public cloud computing incidents highlight
-partition behavior. In April 2011, a network misconfiguration led to a
-twelve-hour long series of outages across the Amazon EC2 and RDS
+According to James Hamilton, Vice President and Distinguished Engineer
+on the Amazon Web Service team, ``network partitions should be rare
+but net gear continues to cause more issues than it
+should''~\cite{hamilton-partitions}. Anecdotal evidence confirms
+Hamilton's assertion. In April 2011, a network misconfiguration led to
+a twelve-hour long series of outages across the Amazon EC2 and RDS
 services~\cite{amazon-netpartition}. Subsequent misconfigurations and
@@
 Heroku~\cite{ec2-downsites}. At a global scale, hardware
-failures---like the 2011 outages in Level3's backbones in North
-America and Europe due a 2011 bug in Juniper's
-routers~\cite{juniper-partition}---and misconfigurations---such as BGP
-faults introduced by Pakistan in 2008~\cite{pakistan-youtube} and
-academics in 2010~\cite{research-experiment-partition}---can cause
-widespread---even Internet-wide---partitions. Many of our discussions
-with practitioners---especially those operating on public cloud
-infrastructure---confirm that partitions are a reality.
+failures---like the 2011 outages in Internet backbones in North
+America and Europe due a router bug~\cite{juniper-partition}---and
+misconfigurations---such as BGP faults in 2008~\cite{pakistan-youtube}
+and 2010~\cite{research-experiment-partition}---can cause widespread
+partitions. Many of our discussions with practitioners---especially
+those operating on public cloud infrastructure---as well as reports
+from large-scale operators like Google~\cite{dean-keynote} confirm
+that partitions are a reality.
 
-Several recent studies quantify partition behavior. A 2011 study of
-several Microsoft datacenters found a mean of 40.8 network link
-failures per day (95th percentile: 136), with a median time to repair
-of around five minutes, and up to one week. The annual probability of
-a device failing and impacting the network is near 10\% for several
-critical components like core routers and primary-backup connectors,
-while network redundancy only reduces the median impact of failures by
-up to 40\%.~\cite{sigcomm-dc}. A 2010 study of five years of over 200
+Several recent studies rigorously quantify partition behavior. A 2011
+study of several Microsoft datacenters found a mean of 40.8 network
+link failures per day (95th percentile: 136), with a median time to
+repair of around five minutes, and up to one week. Perhaps
+surprisingly, network redundancy only reduces the median impact of
+failures by up to 40\%, meaning providers cannot easily reduce the
+incidence of partitions~\cite{sigcomm-dc}. A 2010 study of over 200
 wide-area routers found an average of 16.2--302.0 failures per link
-per year with an average annual downtime of 24--497 minutes (95th
-percentile at least 34 hours) per link per year~\cite{sigcomm-wan}. In
+per year with an average annual downtime of 24--497 minutes per link
+per year (95th percentile at least 34 hours)~\cite{sigcomm-wan}. In
 HP's managed enterprise networks, WAN, LAN, and connectivity problems
@@
 tickets relate to network hardware.  The median incident duration for
-highest priority WAN, LAN, and configuration-related tickets is ranges
-from 114---165 minutes for WAN and LAN issues and up to a day for all
-incidents.~\cite{turner2012failure}. The incidence of network failures
-is further confirmed by several prior
-studies~\cite{labovitz-failures}, showing for example, on Sprint's
-WANs, a median time to repair between 2 and 1000 seconds, and a median
-time between failures of approximately 3000
-seconds.~\cite{ip-backbone-failures}. Indeed, isolating, quantifying,
-and designing for network failures is an area of active research in
-networking community~\cite{surviving-failures-bodik,
+highest priority tickets ranges from 114--188 minutes and up to a full
+day for all tickets.~\cite{turner2012failure}. The incidence of
+network failures is further confirmed by several prior studies showing
+for example, on Sprint's WANs, a median time to repair between 2 and
+1000 seconds, and a median time between failures of approximately 3000
+seconds.~\cite{ip-backbone-failures} and frequent path routing
+failures on the Internet~\cite{labovitz-failures}. Indeed, isolating,
+quantifying, and designing for network failures is an area of active
+research in networking community~\cite{surviving-failures-bodik,
   uw-failure-networks}.
 
-These results---which do not account for server failures (another form
-of system partition that has also been quantified at
+These results---which do not consider server-level failures (another
+form of system partition that has also been studied at
 scale~\cite{google-availability})---indicate that partitions
 \textit{do} occur within and across modern datacenters. We do not
-intend to fixate on the particular statistical behavior but instead
+intend to fixate on any particular statistical behavior but instead
 observe that, as a general trend, partitions exist and must
-accordingly be met with either unavailability at some sites or, as we
-will discuss, relaxed semantic guarantees.
+accordingly be met with either unavailability at some servers or, as
+we will discuss, relaxed semantic guarantees.
 
@@
 planet---require a minimum 85.1 ms RTT (133.7 ms if sent at ground
-level). Of course, not every operation requires contacting another
-server around the world, but, as services are replicated to multiple,
-geographically distinct sites, the cost of communication grows.
+level). As services are replicated to multiple, geographically
+distinct sites, the cost of communication increases this upper bound.
 
@@
 the Eastern United States, and within a single ``availabilty zone,''
-at a granularity of 1s (dataset will be released).
+at a granularity of 1s (dataset to be released).
 
 We summarize the results of our network measurement study in
-Table~\ref{table:rtt}. On average, inter-datacenter communication is
-between 1.82 and 6.38 times faster than across geographically
-co-located datacenters and 40 and 647 times faster than across
-geographically distributed datacenters. The cost of wide-area
+Table~\ref{table:rtt}. On average, inter-datacenter communication (1c)
+is between 1.82 and 6.38 times faster than across geographically
+co-located datacenters (1b) and 40 and 647 times faster than across
+geographically distributed datacenters (1a). The cost of wide-area
 communication exceeds the speed of light: for example, communicating
@@
 transit RTT of 106.7 ms but, on average, incurs a 362.8 ms RTT (95th
-percentile: 649ms). This is perhaps unsurprising due to modern
-Internet topologies, but this three-fold overhead is
-non-negligible. As shown in Figure~\ref{fig:rtt}, the distribution of
-latencies varies between links, but the trend---in line with other
-recent studies~\cite{redblue, mdcc}---is clear: remote communication
+percentile: 649ms). As shown in Figure~\ref{fig:rtt}, the distribution
+of latencies varies between links, but the trend---in line with other
+recent studies~\cite{mdcc, redblue}---is clear: remote communication
 has a substantial cost.
@@
 \end{figure}
+
+\subsection{Modern Systems}
+\label{sec:modernacid}
+
+Mitigating the effects of partitions and network latency is a
+difficult task, requiring system desigers to make hard trade-offs. In
+this section, we briefly discuss known semantic limits for systems
+wishing to maintain high availability and their impact on modern
+database systems. While the gold standard of ACID databases is
+unavailable, we show that many database systems operate under much
+weaker models, motivating the study of these models in a highly
+available context in the remainder of this paper.
+
+Most notable among existing impossibility results for distributed
+systems is Brewer's CAP Theorem, which, as formally proven by Gilbert
+and Lynch, states that it is impossible to maintain linearizability
+(``C,'' or ``consistency''---albeit poorly named), or the ability to
+read the last completed write, and replica availability (``A'') in the
+presence of network partitions (``P'')~\cite{gilbert-cap}. Brewer's
+Theorem has roots in decades of distributed database
+designs~\cite{davidson-survey} and has had a substantial impact on
+recent large-scale distributed systems design. Yet, while CAP is
+narrowly scoped to the impossibility linearizability, it is often
+interpreted to preclude a larger set of guarantees, including ACID, or
+transactional atomicity, consistency, isolation and durability
+guarantees.
+
+Database designers and researchers have long realized that
+serializability---the gold standard of ACID isolation, which
+guarantees application-level consistency---is not achievable in a
+highly available system~\cite{davidson-survey}.  However, database
+systems offer a range of ACID properties besides
+serializability. Within a single-node database, the coordination
+penalties associated with ensuring equivalence with a serial execution
+can be severe and manifest themselves in the form of concurrency
+control contention (and, subsequently, performance degradation,
+scalability limitations, and, often, external aborts like
+deadlocks)~\cite{gray-isolation}. Instead, databases offer a host of
+so-called \textit{weak isolation} models that allow varying
+restrictions on the space of schedules that are allowable by the
+system~\cite{adya}. None of these weak isolation models guarantees
+serializability, but the benefits of these models can outweigh the
+costs of possible application-level consistency anomalies that arise
+from their use.
+
+To understand the prevalence of these weak isolation models, in an
+earlier workshop paper~\cite{hat-hotos}, we surveyed the default and
+maximum isolation models provided by 18 relational databases, often
+claiming to provide ``ACID'' or ``NewSQL'' functionality. As shown in
+Table~\ref{table:existing}, only three out of 18 databases provides
+serializability by default, and at least eight do not provide
+serializability as an option at all. This is particularly surprising
+when we consider the widespread deployment of many of these databases,
+like Oracle 11g, which are known to power major businesses and product
+functionality. Given that these transactional models are frequently
+used, our inability to provide serializability in arbitrary HATs is
+non-fatal to practical applications. If application writers and
+database vendors have already decided that the benefits of weak
+isolation outweigh potential application inconsistencies, then, as we
+will see, in a highly available environment, similar decisions may be
+appropriate.
+
+\begin{table}
+\begin{small}
+\begin{tabular}{|l|c|c|}
+\hline
+Database & Default & Maximum\\\hline
+Actian Ingres 10.0/10S~\cite{actian-docs} & S & S\\
+Aerospike~\cite{aerospike-docs} & RC & RC\\
+Akiban Persistit~\cite{akiban-docs} & SI & SI\\
+Clustrix CLX 4100~\cite{clustrix-docs} & RR & ?\\
+Greenplum 4.1~\cite{greenplum-docs} & RC & S \\
+IBM DB2 10 for z/OS~\cite{db2-docs} & CS & S\\
+IBM Informix 11.50~\cite{informix-docs} & Depends & RR\\
+MySQL 5.6~\cite{innodb-docs} & RR & S \\
+MemSQL 1b~\cite{memsql-docs} & RC & RC\\
+MS SQL Server 2012~\cite{ms-sql-docs} & RC & S \\
+NuoDB~\cite{nuodb-docs} & CR & CR\\
+Oracle 11g~\cite{oracle-docs} & RC & SI\\
+Oracle Berkeley DB~\cite{bdb-reg-docs} & S & S\\
+Oracle Berkeley DB JE~\cite{bdb-je-docs} & RR & S\\
+Postgres 9.2.2~\cite{postgres-docs} & RC & S\\
+SAP HANA~\cite{hana-docs} & RC & SI\\
+ScaleDB 1.02~\cite{scaledb-docs} & RC & RC\\
+VoltDB~\cite{voltdb-docs} & S & S\\
+\hline
+\multicolumn{3}{|p{7cm}|}{{\footnotesize RC: read committed, RR: repeatable read, SI: snapshot isolation, S: serializability, CS: cursor stability, CR: consistent read}}\\\hline
+
+\end{tabular}
+\caption{Default and maximum isolation levels for ACID and NewSQL
+  databases as of January 2013 (from
+  \protect\cite{needed-hotos}).}\vspace{-1.5em}
+\label{table:existing}
+\end{small}
+\end{table}
+
+The primary challenge in providing weak isolation in a HAT context is
+that, unfortunately, it is unclear \textit{which} of these isolation
+guarantees can be provided with high availability. The existing
+algorithms for providing weak isolation are typically presented for a
+single-node context and are therefore unavailable. We are not aware of
+any prior literature that provides guidance as to the relationship
+between weak isolation and high availability. Prior work has examined
+the relationship between serializability and high
+availability~\cite{davidson-survey} and weak isolation on a
+single-server~\cite{adya, ansicritique} but never weak isolation and
+high availability taken together. We believe that this apparent gap in
+the database literature is a possible contributing factor to the
+often-apparent attitude that general-purpose transactional semantics
+are either too expensive to provide in many modern distributed
+databases---except under special circumstances, such as providing
+operations over groups of co-located data items.

### 2013-03-25 thebes-code/scripts/setup_hosts.py
add backend tests and some gc tuning

@@ -215,12 +215,12 @@ def provision_graphite(region):
 def provision_spot(regionName, num):
     global AMIs
     system("ec2-request-spot-instances %s --region %s -t m1.xlarge -price 0.50 " \
-           "-b '/dev/sdb=ephemeral0' -b '/dev/sdc=ephemeral1 -k thebes -g thebes -n %d" % (AMIs[regionName], regionName, num));
+           "-b '/dev/sdb=ephemeral0' -b '/dev/sdc=ephemeral1' -k thebes -g thebes -n %d" % (AMIs[regionName], regionName, num));
 
 def provision_instance(regionName, num):
     global AMIs
     system("ec2-run-instances %s --region %s -t m1.xlarge " \
-           "-b '/dev/sdb=ephemeral0' -b '/dev/sdc=ephemeral1 -k thebes -g thebes -n %d > /tmp/instances" % (AMIs[regionName], regionName, num));
+           "-b '/dev/sdb=ephemeral0' -b '/dev/sdc=ephemeral1' -k thebes -g thebes -n %d > /tmp/instances" % (AMIs[regionName], regionName, num));
     #system("ec2-run-instances %s -n %d -g 'cassandra' --t m1.large -k " \
     #   "'lenovo-pub' -b '/dev/sdb=ephemeral0' -b '/dev/sdc=ephemeral1'" %
     #   (AMIs[region], n))
@@ -442,11 +442,11 @@ def getNextClientID():
 def start_servers(clusters, use2PL, thebesArgString):
     baseCmd = "ulimit -u unlimited; cd /home/ubuntu/thebes/thebes-code; rm *.log; screen -d -m "
     if not use2PL:
-        runServerCmd = baseCmd + "java -ea -Dclusterid=%d -Dserverid=%d %s -jar hat/server/target/hat-server-1.0-SNAPSHOT.jar 1>server.log 2>&1"
+        runServerCmd = baseCmd + "java -XX:+UseParallelGC -Xms7G -Xmx15G -ea -Dclusterid=%d -Dserverid=%d %s -jar hat/server/target/hat-server-1.0-SNAPSHOT.jar 1>server.log 2>&1"
     else:
[…]

### 2013-03-25 thebes-code/scripts/setup_hosts.py
xms is xmx

@@ -442,11 +442,11 @@ def getNextClientID():
 def start_servers(clusters, use2PL, thebesArgString):
     baseCmd = "ulimit -u unlimited; cd /home/ubuntu/thebes/thebes-code; rm *.log; screen -d -m "
     if not use2PL:
-        runServerCmd = baseCmd + "java -XX:+UseParallelGC -Xms7G -Xmx15G -ea -Dclusterid=%d -Dserverid=%d %s -jar hat/server/target/hat-server-1.0-SNAPSHOT.jar 1>server.log 2>&1"
+        runServerCmd = baseCmd + "java -XX:+UseParallelGC -Xms15G -Xmx15G -ea -Dclusterid=%d -Dserverid=%d %s -jar hat/server/target/hat-server-1.0-SNAPSHOT.jar 1>server.log 2>&1"
     else:
-        runServerCmd = baseCmd + "java -XX:+UseParallelGC -Xms7G Xmx15G -ea -Dclusterid=%d -Dserverid=%d %s -jar twopl/server/target/twopl-server-1.0-SNAPSHOT.jar 1>server.log 2>&1"
+        runServerCmd = baseCmd + "java -XX:+UseParallelGC -Xms15G Xmx15G -ea -Dclusterid=%d -Dserverid=%d %s -jar twopl/server/target/twopl-server-1.0-SNAPSHOT.jar 1>server.log 2>&1"
 
-    runTMCmd = baseCmd + "java -XX:+UseParallelGC -Xms7G -Xmx15G -ea -Dclusterid=%d -Dclientid=%d %s -jar twopl/tm/target/twopl-tm-1.0-SNAPSHOT.jar 1>tm.log 2>&1"
+    runTMCmd = baseCmd + "java -XX:+UseParallelGC -Xms15G -Xmx15G -ea -Dclusterid=%d -Dclientid=%d %s -jar twopl/tm/target/twopl-tm-1.0-SNAPSHOT.jar 1>tm.log 2>&1"
 
 
     pprint('Starting servers...')

### 2013-03-25 writeup/evaluation.tex

@@
 
-As we have discussed, many existing database transaction and
-concurrency control algorithms are not designed for high
-availability. These algorithms often presume a single-server
-deployment or the requirement for serializability. As a consequence,
-traditional transaction processing systems are not well-optimized for
-a HAT context.
+Many existing database transaction and concurrency control algorithms
+are not designed for high availability. These algorithms often presume
+a single-server deployment or the requirement for serializability. As
+a consequence, traditional transaction processing systems are not
+well-optimized for a HAT context. In this section, we briefly discuss
+design decisions and algorithmic details that preclude high
+availability.
 
@@
   server but transactions must be checked by a scheduler before
-  commit. With one scheduler per database, this requires one RTT for
-  concurrency control. If schedulers are sharded by database
-  partition, a multi-partition transaction may require contacting
-  several different schedulers and, possibly, an atomic commit
-  protocol, which becomes substantially more expensive.
+  commit. With one scheduler per database, this requires one RTT
+  specifically for concurrency control. If schedulers are sharded by
+  database partition, a multi-partition transaction may require
+  contacting several different schedulers and, possibly, an atomic
+  commit protocol, which becomes substantially more expensive.
 
-\item \textbf{Deterministic transaction scheduling} is effecitvely the
+\item \textbf{Deterministic transaction scheduling} is ffectively the
   inverse of optimistic concurrency control: a scheduler first orders
@@
 
-There are several mechanisms that do not provide serializability (and
-often provide HAT semantics) but which are not
-available. Non-serializability can but does not necessarily imply high
-availability and low latency:
+Perhaps unsurprisingly, the reliance on a glocal total order
+necessitates a minimum of one round-trip to a designated master or
+coordination service for each of these classic algorithms. This is
+simply a cost of providing serializability, and the cost of the
+(minimum one) round trip will be determined by the deployment
+environment, as we saw in Section~\ref{sec:motivation}.
+
+There also are several mechanisms that do not provide serializability
+but provide transactional semantics with varying availability
+properties:
 
@@
   lock server (or, alternatively, a majority quorum). In the presence
-  of failure, distributed locking can be challenging.
+  of failure, distributed locking is challenging.
 \item The \textbf{MDCC} commit protocol uses a Paxos variant and escrow
@@
   suffer from unavailability as it does not have a coordinator.
-\item \textbf{Chan and Gray read-only transactions} use a similar
+\item \textbf{Chan and Gray's Read-only transactions} use a similar
   algorithm to Eiger to provide read-only transactions with item cut
@@
 Thus far, we have largely ignored the question of which applications
-actually require HAT semantics versus traditional ACID guarantees. As
-we summarized in Section~\ref{needed}, the main cost of providing HAT
-updates is the inability to prevent Lost Update, Write Skew, and
-provide recency bounds. In this Section, we attempt to understand when
-these guarantees matter both abstractly and by the use of a
-representative customer order transaction application patterned off of
-TPC-C.
+require semantics which are unavailable to HATs. As we showed in
+Section~\ref{sec:hats}, the main cost of choosing HATs comes in the
+inability to prevent Lost Update, Write Skew, and provide recency
+bounds. In this Section, we attempt to understand when these
+guarantees matter both abstractly and by the use of a representative
+customer order transaction application patterned off of TPC-C.
 
@@
 
+As an example of an actual system rearchitecture,
+less than one year after its announcement, Yahoo!'s PNUTS developers
+explicitly added a support for weaker, highly available operation,
+explaining that ``strict adherence [to strong consistency] leads to
+difficult situations under network partitioning or server
+failures...in many circumstances, applications need a relaxed
+approach''~\cite{pnuts-update}.
+
 Absent a necessary and sufficient result describing exactly which
@@
 representative of a standard transaction processing application---to
-understand which are achievable in a HAT context.
-
-TPC-C consists of five transactions, capturing operation of a
-real-world warehouse, including sales, payments, and deliveries. We
-analyze each transaction and its interactions below.
+understand which are achievable in a HAT context. TPC-C consists of
+five transactions, capturing operation of a real-world warehouse,
+including sales, payments, and deliveries. We analyze each transaction
+and its interactions below.
 
@@
   ID number, adjusts the remaining warehouse stock, and writes a
-  placeholder entry for the pending order. There are two mutations in
-  this transaction which we must be careful of.
+  placeholder entry for the pending order. There are two mutations we
+  must be consider in detail.
 
   First, assigning an order ID number is challenging. We can easily
-  assign a \textit{unique} order ID, by, say, concatenating the user
-  ID with a timestamp and server ID. However, the TPC-C specification
-  requires that order numbers be sequentially assigned, which we
-  cannot guarantee in HATs due to Lost Update. Accordingly, while HATs
-  cannot provide compliant TPC-C execution due to a non-monotonic
-  counter increase, they can allow placement of new orders, which is a
-  commutative and monotonic operation
+  assign a \textit{unique} order ID, by, say, concatenating the
+  customer ID with a timestamp and server ID. However, the TPC-C
+  specification requires that order numbers are assigned sequentially,
+  which we cannot guarantee in HATs due to Lost Update. Accordingly,
+  HATs can allow placement of new orders with unique identifiers but
+  cannot provide compliant TPC-C execution due to the non-monotonic
+  counter increase.
 
@@
   operation as, without coordination, inventory counts may become
-  negative. However, TPC-C automatically ``refills'' inventory counts
+  negative. However, TPC-C automatically ``restocks'' inventory counts
   (by 91 items) when there are 10 or fewer items
@@
   orders to the same item, the stock will never fall below
-  zero. However, if the system propagates ``refill'' operations
-  between replicas, then an $R$ replica system might end up with
-  $91(R-1)$ more stock than a single-site implementation.
+  zero. However, an $R$ replica HAT system might end up with $91(R-1)$
+  more stock than a single-site implementation.
 

### 2013-03-25 writeup/hat-analysis.tex

@@
 
-HAT systems offer substantial latency and availability benefits, yet
-they come with a cost. In the distributed systems community, there is
-a host of results that taxonomize the relationships between various
-distributed consistency models. Perhaps most famously, Brewer's CAP
-Theorem---as formally proven by Gilbert and
-Lynch~\cite{needed}---states that it is impossible to maintain
-linearizability (``C,'' or ``consistency''---albeit poorly named), or
-the ability to read the last completed write, and replica availability
-in the presence of failures~\cite{needed}. While the CAP Theorem has
-has roots in decades of distributed system designs~\cite{needed}, it
-has had large scale impact on the design and development of
-distributed systems, particularly for Internet
-services---unsurprisingly, the context in which Brewer formulated his
-finding.
+HAT systems---providing transactions with transactional availability
+with high availablity ($1$-available) or sticky high availability
+($1$-sticky-available)---offer substantial latency and availability
+benefits, yet they come with a cost to achievable semantics. In this
+section, we delineate which of several ACID, distributed consistency,
+and session consistency levels can be achieved with high availability
+(e.g., transactional atomicity, variants of Repeatable Read isolation,
+and many session guarantees) and sticky high availability (e.g.,
+read-your-writes, PRAM, and causal consistency) and which cannot
+(e.g., preventing Lost Update and Write Skew).
 
-\subsection{ACID and Modern Databases}
-
-While linearizability is a useful property to consider, there are
-several other guarantees that are perhaps more relevant to a database
-system. In particular, the database tradition of providing ACID
-guarantees: atomicity, consistency, isolation, and durability are,
-with few exceptions, not addressed by the distributed systems
-literature. There are several factors behind this occurrence, the
-least of which is that distributed systems rarely consider
-multi-object guarantees, whereas ACID transactions are frequently used
-to provide semantic guarantees across multiple operations on multiple
-objects. While linearizability (unfortunately often called
-``atomicity'') ``composes'' across multiple objects, a linearizable
-system does not provide ACID ``atomicity'' (which we will call
-``transactional atomicity'' for clarity): just because two writes are
-visible to all writers immediately after they complete does not, on
-its own, provide any guarantees about the mutual visibility of the two
-writes.
-
-Database designers and researchers have long realized that
-serializability---the gold standard of ACID isolation, which
-guarantees application-level consistency---is not achievable in a
-highly available system. Indeed, decades of research on
-semantics-based concurrency control and coordination-reducing
-techniques such as escrow transactions and sagas allow serializability
-in the presence of partial system knowledge or otherwise reduce
-coordination cost. However, the majority of these algorithms are
-intended to preserve application-level consistency, which, for
-arbitrary applications and arbitrary sequence of operations, cannot be
-ensured with high availability in the presence of partitions.
-
-Of course, database systems offer a range of ACID properties besides
-serializability. Even on a single-node system, the coordination
-penalties associated with ensuring equivalence with a serial execution
-can be severe and manifest themselves in the form of concurrency
-control contention (and, subsequently, performance degradation,
-scalability limitations, and, often, external aborts like
-deadlocks). Instead, databases offer a host of so-called \textit{weak
-  isolation} models that allow varying restrictions on the space of
-schedules that are allowable by the system. None of these weak
-isolation models guarantees serializability, but the benefits of these
-models can outweigh the costs of possible application-level
-consistency anomalies that arise from their use.
-
-To understand the prevalence of these weak isolation models, we
-surveyed the default and maximum isolation models provided by 18
-relational databases, often claiming to provide ``ACID'' or ``NewSQL''
-functionality. As shown in Table~\ref{table:existing}, only three out
-of 18 databases provides serializability by default, and at least
-eight do not provide serializability as an option at all. This is
-particularly surprising when we consider the widespread deployment of
-many of these databases, like Oracle 11g, which are known to power
-major businesses and product functionality. We might view the use of
-weak consistency is somewhat of an ``arms race'': anecdotally, one
-major RDBMS vendor was forced to lower their default isolation level
-in order to compare favorably in out-of-the-box performance
-comparisons with its major competitor, which had already lowered its
-default setting. Applications can guard against inconsistency in a
-weakly isolated environment by performing their own locking external
-to the database or by performing ad-hoc concurrency control within the
-database. However, this appears error-prone and, without further
-evidence, it remains somewhat puzzling why this is indeed favorable to
-providing serializability with in the database itself.
-
-\begin{table}
-\begin{small}
-\begin{tabular}{|l|c|c|}
-\hline
-Database & Default & Maximum\\\hline
-Actian Ingres 10.0/10S~\cite{actian-docs} & S & S\\
-Aerospike~\cite{aerospike-docs} & RC & RC\\
-Akiban Persistit~\cite{akiban-docs} & SI & SI\\
-Clustrix CLX 4100~\cite{clustrix-docs} & RR & ?\\
-Greenplum 4.1~\cite{greenplum-docs} & RC & S \\
-IBM DB2 10 for z/OS~\cite{db2-docs} & CS & S\\
-IBM Informix 11.50~\cite{informix-docs} & Depends & RR\\
-MySQL 5.6~\cite{innodb-docs} & RR & S \\
-MemSQL 1b~\cite{memsql-docs} & RC & RC\\
-MS SQL Server 2012~\cite{ms-sql-docs} & RC & S \\
-NuoDB~\cite{nuodb-docs} & CR & CR\\
-Oracle 11g~\cite{oracle-docs} & RC & SI\\
-Oracle Berkeley DB~\cite{bdb-reg-docs} & S & S\\
-Oracle Berkeley DB JE~\cite{bdb-je-docs} & RR & S\\
-Postgres 9.2.2~\cite{postgres-docs} & RC & S\\
-SAP HANA~\cite{hana-docs} & RC & SI\\
-ScaleDB 1.02~\cite{scaledb-docs} & RC & RC\\
-VoltDB~\cite{voltdb-docs} & S & S\\
-\hline
-\multicolumn{3}{|p{7cm}|}{{\footnotesize RC: read committed, RR: repeatable read, SI: snapshot isolation, S: serializability, CS: cursor stability, CR: consistent read}}\\\hline
-
-\end{tabular}
-\caption{Default and maximum isolation levels for ACID and NewSQL
-  databases as of January 2013 (from
-  \protect\cite{needed-hotos}).}\vspace{-1.5em}
-\label{table:existing}
-\end{small}
-\end{table}
-
-Given that these transactional models are frequently used, our
-inability to provide serializability to provide general-purpose HATs
-appears not to be a show-stopper. If applications cope with
-inconsistency or, on a single node server, have already decided that
-the benefits of weak isolation outweigh potential application
-inconsistencies, then perhaps, in a highly available environment, some
-may make a similar decision given the appropriate choice of semantic
-guarantees.
-
-The primary challenge in providing weak isolation in a HAT context is
-that, unfortunately, it is unclear \textit{which} isolation guarantees
-can be provided with high availability. Standard discussions of weak
-isolation are presented in a single-node context---a possible
-consequence of their original context: a lock-based (and therefore
-unavailable, even in a distributed context) single-server
-deployment~\cite{needed}. We are not aware of any prior literature
-that provides guidance as to the relationship between weak isolation
-and high availability. Prior work has examined the relationship
-between serializability and high availability and weak isolation on a
-single-server but never weak isolation and high availability taken
-together. This apparent gap in the database literature is a possible
-contributing factor to the often-apparent attitude that
-general-purpose transactional semantics are either too expensive or
-too lacking in availability guarantees to provide in modern
-distributed databases---except under special circumstances, such as
-operations over groups of co-located data items.
+When possible, we draw on existing properties and definitions from the
+database and distributed systems literature, providing a brief,
+informal explanation of each guarantee and illustrate it with an
+example. In particular, for our ACID guarantees, we draw largely on
+Adya's dissertation~\cite{adya}, Berenson et al.'s 1995 critique of
+the ANSI SQL Specification~\cite{ansicritique}, and the ANSI SQL
+specification~\cite{ansi-sql}. We provide a set of formal definitions
+and semantics in the extended version of this work.
 
@@
 
-In this paper, we will describe which ACID semantics are achievable in
-HATs as well as which are not. Our strategy is to explore successively
-more powerful models until we reach a frontier of semantic guarantees
-which are unavailable to HATs. To begin, we present achievable
-semantics, offering proof-of-concept algorithms to demonstrate
-feasibility. When possible, we draw on existing properties and
-definitions from the database and distributed systems literature,
-providing a brief, informal explanation of each guarantee and
-illustrate it with an example. For our ACID guarantees, we draw
-largely on Adya's thesis work on defining weak isolation
-models~\cite{needed} and, where appropriate, supplement with
-definitions and discussion from Berenson et al.'s SIGMOD 1995 critique
-of the ANSI SQL Specification~\cite{needed} (hereafter, Berenson) as
-well as the ANSI SQL specification~\cite{needed} (hereafter, ANSI)
-itself. For other guarantees, such as session properties, we draw on
-both the database and distributed systems literature. We provide a set
-of formal definitions and semantics in the Appendix.
+To begin, we present achievable semantics, offering proof-of-concept
+algorithms to demonstrate feasibility.
 
@@
 the below example, $T_3$ should eventually only read $a=b=1$ or
-$a=b=2$ but not $a=2, b=1$:
+$a=b=2$ but not $a=2, b=1$ or $a=1, b=2$:
 \begin{align*}
@@
 serializable total order must be equivalent to a serial
-execution. Interestingly, Read Uncommitted is a close analog with
-eventually consistent systems's concept of replica convergence, in
-which all replicas for a data item must end up with the same data
-item. This convergence is usually achieved by picking a ``winning''
-update---which requires ordering---or by maintaining a growing a set
-of updates---whose cardinality implicitly defines an order (with
-convergence achieved when all updates are in the set); we can see Read
-Uncommitted as a cross-item convergence property. It is easily
-achieved via applying the same logical timestamp to each update in a
-transaction.
+execution. Read Uncommitted is similar to the distributed systems
+concept of replica convergence, in which all replicas for a data item
+must end up with the same data item. This convergence is usually
+achieved by picking a ``winning'' update, requiring a means of totally
+ordering updates to each item; one can view Read Uncommitted as a
+cross-item convergence property. It is easily achieved via applying
+the same logical timestamp to each update in a transaction.
 
@@
 Writes''---as above---and ``Dirty Reads'' phenomena; captured by
-Adya's $G1\{a,b,c\}$, ANSI's $P1$, and Berenson's ``broad'' $P1$
+Adya's $G1a-c$, ANSI's $P1$, and Berenson's ``broad'' $P1$
 (2.2)). For instance, in the example below, $T_3$ should never see
@@
 \end{align*}
-It is fairly easy to prvent ``Dirty Reads'': if transactions never
-write uncommitted data to the database, then transactions will never
-read dirty data. Clients can buffer their writes until they commit,
-or, alternatively, can send them to servers, who will not serve writes
-until clients notify that they have been committed.
+It is fairly easy to prvent ``Dirty Reads'': if each transaction never
+writes uncommitted data to the database, then transactions will never
+read each others' dirty data. Clients can buffer their writes until
+they commit, or, alternatively, can send them to servers, who will not
+serve new writes until clients notify them that the writes have been
+committed.
 
 \textit{Repeatable Read} isolation is a contentious property. As
-Berenson et al. point out, Gray's original Repeatable Read lock-based
-implementation provides substantially richer guarantees than those
-that are promised by the ANSI SQL specification. Gray, Berenson, and
-Adya effectively preclude all serializability anomalies but the
-Phantom Problem (which we have explicitly not yet considered and will
-do so in Section~\ref{sec:unachievable-hat}). Nevertheless, the ANSI
-SQL specification provides a useful property which, although is not
-true to Gray's spirit of Repeatable Read, is used in distributed
-consistency models. The ANSI Repeatable Read guarantee requires that,
-along with observing Read Committed isolation, each transaction can
-only read only version of each data item that it did not itself
-produce (Berenson et al. call this preventing ``Fuzzy Read,'' or
-P2). In the example below, $T_3$'s must see $a=1$:
+Berenson et al. discuss~\cite{ansicritique}, Gray's original
+Repeatable Read lock-based implementation provides substantially
+richer guarantees than those that are promised by the ANSI SQL
+specification~\cite{gray-isolation}. Gray~\cite{gray-isolation},
+Berenson et al.~\cite{ansicritique}, and Adya~\cite{adya} all
+interpret Repeatable Read as providing serializability for all
+operations except predicate-based reads. However, in its description
+of Repeatabable Read, the ANSI SQL specification provides a useful
+property which, although is not true to Gray's spirit of Repeatable
+Read, is also commonly found in distributed consistency models: the
+ANSI Repeatable Read guarantee requires that, along with observing
+Read Committed isolation, each transaction will only read one version
+of each data item that it did not itself produce (preventing ``Fuzzy
+Read,'' or P2). In the example below, $T_3$'s must see $a=1$:
 \begin{align*}
@@
 \end{align*}
-This isolation property is much weaker than serializability, but it is
-a good faith literal interpretation of the phrase ``Repeatable Read'':
-transactions see their own writes. By itself, this property can be
-implemented trivially---always read \texttt{null} for each
-key. However, when coupled with additional properties, like
-transactional atomicity, it becomes stronger: the repeated reads must
-also obey some additional ordering properties. Distributed systems
-describe a ``consistent snapshot'' across a set of related events as a
-\textit{cut} across a set of participants or, in our case, data
-items. Accordingly, to capture the notion that the transaction should
-read from a consistent cut, and to disambiguate from the
-aforementioned stronger Repeatable Read properties, we call ANSI SQL
-Repeatable Read \textbf{Cut Isolation}. In the absence of additional
-guarantees, it is possible to satisfy cut isolation with high
-availability by caching each transaction's reads or, alternatively,
-using multi-versioning on the server and ensuring that each
-transaction's successive reads return the same version of each data
-item.
+This isolation property is a literal interpretation of the phrase
+``Repeatable Read'': unless a transaction modifies a given data item,
+the value of the data item should not change during the
+transaction. By itself, this property is rather weak---we can always
+read \texttt{null} for each data item. However, when coupled with
+additional properties, like transactional atomicity, it is stronger:
+the repeated reads must obey additional ordering constraints. In fact,
+distributed systems describe a ``consistent snapshot'' across a set of
+related events as a \textit{cut} across a set of participants or, in
+our case, data items. Accordingly, to capture the notion that the
+transaction should read from a (non-changing) consistent cut over data
+items (and to disambiguate from the aforementioned stronger Repeatable
+Read properties), we call ANSI SQL Repeatable Read (preventing P2)
+\textbf{Cut Isolation}.It is possible to satisfy cut isolation with
+high availability by caching appropriate versions for a transaction's
+reads or, alternatively, using multi-versioning on the server and
+ensuring that each transaction's successive reads return the same
+version of each data item.
 
-We can consider two variants of Cut Isolation. The first is relatively
-straightforward, and reasons about cuts over discrete data items. We
-will call this \textit{Item Cut Isolation}. The second is known as
-preventing the \textit{Phantom Problem}, whereby two successive
-predicate-based reads return different data. We call predicate-based
-cut isolation, which prevents Phansoms, \textbf{Phantom Cut
-  Isolation}.\footnote{Oracle provides an isolation level called
-  ``Statement Level Read Consistency''; this is analogous to Phantom
-  Cut Isolation at the level of a single statement. We do not consider
-  this isolation model here as it is subsumed by standard Snapshot
-  Isolation but we believe it is easily adaptable.}
+We can further consider two variants of Cut Isolation. The first is
+relatively straightforward and provides a cut across individual data
+items. We will call this \textit{Item Cut Isolation}. The second
+provides a cut over individual data items plus logical ranges of data
+items, or predicate-based reads. This prevents the \textit{Phantom
+  Problem}, whereby two successive predicate-based reads return
+different data~\cite{gray-isolation}. We call predicate-based cut isolation, which prevents
+Phantoms, \textit{Phantom Cut Isolation}.\footnote{Oracle provides an
+  isolation level called ``Statement Level Read Consistency''; this is
+  analogous to Phantom Cut Isolation at the level of a single
+  statement. We do not consider this isolation model here as it is
+  subsumed by standard Snapshot Isolation and we believe our
+  discussion is easily extended to incorporate it.}
 
@@
 Transactional atomicity is core to ACID guarantees. Although, at least
-by the acronym, not an ``isolation'' property, transactional atomicity
-restricts transactions' ability to view the effects of partially
-completed transactions: within a transactions, either all effects of
-another transaction are observed, or none are (equivalently, once some
-of the effects of a transaction are observed, all effects are
-observed). This is a strictly stronger guarantee than Read Committed
-isolation because, in Read Committed, we can read a subset of
-committed writes whereas, with Transactional Atomicity, we must read
-all of them. For example, if $T_1$ commits, then, given Transactional
-Atomicity, $T_2$ must observe $b=c=1$ (or later versions):
+by the acronym, it is not an ``isolation'' property, transactional
+atomicity restricts transactions' ability to view the effects of
+partially completed transactions. Under transactionational atomicity,
+within each transaction, either all effects of another transaction are
+observed, or none are (equivalently, once some of the effects of a
+transaction are observed, all effects are observed). This is a
+strictly stronger guarantee than Read Committed isolation: in Read
+Committed, we can read a subset of committed writes whereas, with
+transactional atomicity, we must read all of them. For example, if
+$T_1$ commits, then, given transactional atomicity, $T_2$ must observe
+$b=c=1$ (or later versions):
 \vspace{-.5em}
@@
 
-Perhaps perplexingly, discussions of Transactional Atomicity are
+Perhaps perplexingly, discussions of transactional atomicity are
 absent from existing discussions of weak isolation levels. This is
@@
 violating availability due to, say, locking, or a missing
-transactional write.
-
-Together with Cut Isolation, Transactional Atomicity prevents Read
-Skew (Berenson et al.'s A5A).
-
-(MORE HERE FROM MY NOTES)
+transactional write. Together with Cut Isolation, Transactional
+Atomicity prevents Read Skew anomalies (Berenson et al.'s
+A5A~\cite{ansicritique}).
 
@@
 subtle distinction corresponds to the difference between the
-distributed systems concepts of consensus and uniform reliable
-broadcast. The former requires explicit agreement across servers,
-while the latter simply requires that, if one server reveals a write,
-then all servers reveal the write.
-
-(TODO: fix me--HotOS copy!)
-
-Nonetheless, Coordinating transactional atomicity across replicas is
-challenging. For example, if, as is common~\cite{readonly, eiger}, we
-use a master to determine the correct set of updates to read, then the
-system will not be available when clients are partitioned from the
-master. Instead, we rely on decentralized \textit{background}
-agreement to decide when to show new values, which can safely stall in
-the presence of partitions. Each server keeps a known good value for
-each data item $d$ it is assigned, $d_{good}$ and a set of pending
-updates to $d$, $d_{pending}$. We use asynchronous agreement to move
-updates from $d_{pending}$ to $d_{good}$: all of $d_{good}$'s
-transactional siblings (or later values) are guaranteed present on all
-replicas. Without aborts, this requires only one synchronous round
-trip time (RTT).
-
-\noindent\textit{Writes:} First, consider the case where transactions
-do not internally abort. On commit, for each written data item $d$, we
-send the last written value of $d$ ($d_v$) with the transaction ID and
-$L_v$, a list of all data items and values written in the transaction,
-to available replicas for $d$. The replicas immediately reply with a
-response to \textit{commit}, place $d_v$ it in their respective
-$d_{pending}$, and send a receipt to all replicas for items in
-$L_v$. Once a server receives receipts from all replicas for all items
-in $L_v$, the server sets $d_{good} = d_v$ if $d_v$'s ID is greater
-than $d_{good}$'s ID. Accordingly, a new value will only appear in
-$d_{good}$ when the other values written with $d_{good}$ ($L_v$) are
-resident on every respective replica (though possibly in their
-$d_{pending}$).\vspace{.5em}
-
-\noindent\textit{Reads:} Each client maintains a vector of transaction
-IDs $V$, with one entry per data item ($V(d)$). When a client reads a
-new data item $d_i$, the server returns both $d_i$ and $L_v$, and, for
-all data items $k \in L_v$, the client sets $V(k)$ to $d_i$'s ID. When
-a client sends a read request for item $x$, it also sends $V(x)$, and
-the replica returns \textit{either} $x_{good}$ if it matches $V(x)$ or
-occurs after $V(x)$ in the transaction order or, if $x_{good}$ is
-unsatisfactory, $V(x)$ from $x_{pending}$. If $V(x)$ is empty, the
-server returns $x_{good}$.\vspace{.5em}
-
-\noindent\textit{Aborts:} If we allow internal aborts, writes require
-2 RTTs: one to check constraints and one to execute the above. Write
-propagation and update visibility are achieved asynchronously, as in
-the no-abort case.\vspace{.5em}
-
-This implementation is entirely masterless and both reads and writes
-do not block for coordination. One key property is that we do not, in general,
-dictate when writes become visible (see: session guarantees,
-\S\ref{sec:impossible}). Accordingly, it is highly available.\vspace{.5em}
+distributed systems concepts of consensus and reliable broadcast. The
+former requires explicit agreement across servers, while the latter
+simply requires that, if one server reveals a write, then all servers
+reveal the write. We omit a full discussion of the algorithm and
+instead refer the reader to the extended version of this paper.
 
@@
 between transactions (other than that there exists some arbitrary
-ordering).
-
-\textit{Session guarantees} are used to describe continuity between
-groups of transactions. Informally, a \textit{session} is a form of
-context that consists of logical state that should persist between
+ordering). \textit{Session guarantees} are used to describe continuity
+between groups of transactions issued by the same logical
+user. Informally, a \textit{session} is a form of context that
+consists of logical state that should persist between
 operations. Users accessing a web site may want to experience a notion
-of ``forward progress'' for the duration that they are logged in, and
-the duration of the ``log in'' to ``log out'' period forms a
-session. We can define session guarantees at several granularities
-(e.g., per-transaction, per-client, per-hour) but we choose to operate
-under a client-centric model.
+of ``forward progress'' for the duration that they are logged in, so
+all transactions submitted between ``log in'' and ``log out'' might
+form a session. We can define sessions at several granularities (e.g.,
+per-transaction, per-client, per-hour) but, for simplicity, will refer
+to client-level guarantees in the discussion.
 
@@
 previous values''; reads from each item progress forward in a total
-order. Under Read Uncommitted, the ordering of reads should respect the
-total ordering on transactions.
+order. The ordering of reads should respect any total ordering on
+transactions.
 
 \vspace{.25em}\noindent\textit{\textbf{Monotonic writes}} requires
-that transactions from each individual client be serialized. Under
-Read Uncommitted, the transaction order should also respect the order
-in which each client submitted transactions.
+that transactions from each individual client be serialized. Any order
+on transactions should also respect the order in which each session
+submitted transactions.
 
 \vspace{.25em}\noindent\textit{\textbf{Writes Follow Reads}} requires
-that, if a client reads a write originating from transaction $T_1$
-then performs transaction $T_2$, then a client can only read $T_2$'s
+that, if a session reads a write originating from transaction $T_1$
+then performs transaction $T_2$, then a session can only read $T_2$'s
 writes if it can also read $T_1$'s prior write (or overwritten values
@@
 captures all events that influenced each
-transaction~\cite{causalmemory}. Under Read Uncommitted, the
-transaction order should respect the reads-from order (note that the
-total order on transactions precludes reads-from cycles).
+transaction~\cite{causalmemory}. Any transaction order should respect
+the reads-from order (note that the total order on transactions
+precludes reads-from cycles).
 
 We can achieve the above guarantees by forcing servers to wait to
-reveal new writes until their respect dependencies are fulfilled on
+reveal new writes until their respective dependencies are fulfilled on
 all replicas. In the case of Monotonic Reads, servers should only
@@
 guarantees. For Writes Follow Reads, a servers should similarly wait
-to reveal items until their dependencies are present
-everywhere.
+to reveal items until their dependencies are present everywhere.
 
 Effectively, this mechanism requires that clients only read from a
-globally agreed upon lower bound on the writes in the system. This is
-indeed highly available as a client will never block due to inability
-to find a server with a sufficiently up-to-date version of a data
-item. However, it does not imply that transactions will read their own
-writes or provide much confidence in making forward progress through
-the version history in the presence of partitions. The problem is
-that, if a server is partitioned, under the highly available model, we
-cannot discount the possibility that an unfortunate client will be
-forced to issue her next requests against the partitioned server!
+globally agreed upon lower bound on the versions written to the
+system. This is highly available as a client will never block
+due to inability to find a server with a sufficiently up-to-date
+version of a data item. However, it does not imply that transactions
+will read their own writes or provide much confidence in making
+forward progress through the version history in the presence of
+partitions. The problem is that, if a server is partitioned, under the
+highly available model, we cannot discount the possibility that an
+unfortunate client will be forced to issue her next requests against
+the partitioned server!
 
@@
 
-\vspace{.25em}\noindent\textit{\textbf{Causal consistency}} is the
-combination of all of the prior session
-guarantees~\cite{daudjee-session} and is also referred to as PL-2L
-isolation~\cite{adya}). We can also consider arbitrary
-application-defined partial orders, as in explicit
-causality~\cite{explicit-socc}, if dependencies are specified at
-commit time.
+\noindent\textit{\textbf{PRAM}} provides the illusion of serializing
+each session's operations and is the combination of monotonic reads,
+monotonic writes, and read your writes.
+
+\noindent\textit{\textbf{Causal consistency}} is the combination of
+all of the session guarantees~\cite{sessiontocausal} (alternatively,
+PRAM and writes-follow-reads) and is also referred to as PL-2L
+isolation~\cite{adya}).
 
@@
 
-Read-your-writes can be accomplished via client-side caching and
-alternatively comes ``for free'' when requests are routed to a sticky
-set of servers. Achieving causal consistency requires maintaining
-read-your-writes along with the prior session guarantees. Both are
-reasonable strategies often employed in real-world
-systems~\cite{vogels-defs}, albeit at the cost of true high
-availability or client-side storage. Perhaps interestingly,
-
-We discuss additional uses for stickiness in Section~\ref{sec:discussion}.
+However, read-your-writes can be accomplished via client-side caching
+and alternatively comes ``for free'' when requests are routed to a
+sticky set of servers, while causality and PRAM can be accomplished
+via sticky versions of the prior algorithms.
 
@@
 transaction semantics under HAT consistency guarantees in
-Section~\ref{sec:Evaluation}.
+Section~\ref{sec:evaluation}.
 
-As we briefly discussed in Section~\ref{sec:availabile}, a client
+As we briefly discussed in Section~\ref{sec:availability}, a client
 requiring \textbf{durability} of surviving $F$ server faults requires
-at least $F+1$-availability.
+at least ($F+1$)-availability.
 
@@
 While there are infinitely many HAT models
-(Section~\ref{sec:futurework}), at this point, we have exhausted our
-range of achievable, previously defined (useful) HAT semantics. Before
-summarizing our possibility results, we will present impossibility
-results for HATs, also defined in terms of previously identified
-isolation and consistency anomalies.
+(Section~\ref{sec:futurework}), at this point, we have largely
+exhausted the range of achievable, previously defined (useful) HAT
+semantics. Before summarizing our possibility results, we will present
+impossibility results for HATs, also defined in terms of previously
+identified isolation and consistency anomalies.
 
@@
 
-The simplest way to describe which isolation properties are achievable
-with high availability is in terms of preventing different isolation
-anomalies. We have already shown that it is possible prevent several
-anomalies, like Dirty Read, Dirty Write, Fuzzy Read, and atomicity
-violations. In this section, we demonstrate that preventing Lost
-Update, Write Skew, and Phantoms requires unavailability.
+In this section, we demonstrate that preventing Lost Update and Write
+Skew---and therefore providing Snapshot Isolation, Repeatable Read,
+and Serializability---requires unavailability.
 
@@
 its original read of the data item, ``missing'' or ``losing'' $T2$'s
-newer update. Consider the following transactions:
+newer update. Consider a database containing only the following
+transactions:
 \begin{align*}
 \small
-T_1 &: r_x(a) w_x(a+1)
+T_1 &: r_x(a) w_x(a+2)
 \\T_2 &: w_x(2)
 \end{align*}
-If $T_1$'s read from $x$ precedes $T_2$'s write to $x$ but $T_2$'s
-write to $x$ precedes $T_1$'s write to $x$, then a database containing
-only these two transactions will end in a state that could not have
-resulted in a serial execution.
+If $T_1$ reads $a=1$ but $T_2$'s write to $x$ precedes $T_1$'s write
+operation, then the database will end up with $a=3$---a state that
+could not have resulted in a serial execution, arising from $T_2$'s
+``Lost Update.''
 
@@
 As Berenson et al. describe, if there was an integrity constraint
-between $x$ and $y$ such that only one of $x$ or $y$ should be one at
-any given time, then this write skew would lead to a non-serializable
-execution. Write skew in particular is a somewhat esoteric
-anomaly---for example, it does not appear in TPC-C---but, as a
-generalization of Lost Update, is also unavailable.
+between $x$ and $y$ such that only one of $x$ or $y$ should have value
+$1$ at any given time, then this write skew would lead to a
+non-serializable execution. Write skew in particular is a somewhat
+esoteric anomaly---for example, it does not appear in
+TPC-C~\cite{snapshot-serializable}---but, as a generalization of Lost
+Update, is also unavailable.
 
 The inability to prevent Lost Update means that Consistent Read,
-Snapshot Isolation, and Cursor Stability are all unavailable. The
-inability to prevent Lost Update or Write Skew means that Repeatable
-Read, Conflict Serializability, and One-Copy Serializability are
-unavailable.
+Snapshot Isolation, and Cursor Stability guarantees are all
+unavailable. The inability to prevent Lost Update or Write Skew means
+that Repeatable Read, Conflict Serializability, and One-Copy
+Serializability guarantees are unavailable.
 
@@
 Data storage engines make various recency guarantees on reads of data
-items. One of the most famous is linearizability, which states that
-reads will return the last completed write to a data item, and there
-are several other (weaker) variants such as safe and regular register
-semantics. When applied to transactional semantics, the combination of
-one-copy serializability and linearizability is called \textit{strong
-  (or strict) one-copy serializability} or \textit{externally
-  consistent transactions} (e.g., Spanner). It is also common,
+items. As we have discussed, one of the most famous is
+linearizability, which states that reads will return the last
+completed write to a data item, and there are several other (weaker)
+variants such as safe and regular register semantics. When applied to
+transactional semantics, the combination of one-copy serializability
+and linearizability is called \textit{strong (or strict) one-copy
+  serializability} or \textit{externally consistent transactions}
+(e.g., Google's Spanner~\cite{spanner}). It is also common,
 particularly in systems that allow reading from masters and slaves, to
 provide a guarantee such as ``read a version that is no more than five
-seconds out of date'' or similar. Unfortunately, as the CAP Theorem
-demonstrates, linearizability is unavailabile, and any indefinitely
-long partition can cause violation of any recency bound.
+seconds out of date'' or similar. Unfortunately, an indefinitely long
+partition can cause violation of any recency bound.
 
@@
 unavailble consistency models we have discussed in
-Table~\ref{fig:hatcompared}. Many models are simultaneously
+Figure~\ref{fig:hatcompared}. Many models are simultaneously
 achievable: for example, a HAT system can provide monotonic reads and
@@
 
- \node[draw=none,fill=none] (ici) at (1.2, 1.6) {I-CI};
- \node[draw=none,fill=none] (pci) at (1.2, 2.4) {P-CI};
- \node[draw=none,fill=none] (rc) at (1.2, .8) {RC};
- \node[draw=none,fill=none] (ru) at (1.2, 0) {RU};
+ \node[draw=none,fill=none] (ici) at (0, 0) {I-CI};
+ \node[draw=none,fill=none] (pci) at (1.1, 1.3) {P-CI};
+ \node[draw=none,fill=none] (rc) at (-1.2, .8) {RC};
+ \node[draw=none,fill=none] (ru) at (-1.2, 0) {RU};
 
- \node[draw=none,fill=none] (ta) at (0, 0) {TA};
+ \node[draw=none,fill=none] (ta) at (1.2, 0) {TA};
  \node[draw=none,fill=none] (mr) at (2.4, 0) {MR};
@@
  \node at (3.6, 1) [sticky] (pram) {PRAM};
- \node[noha] (cs) at (2, 2.8) {CS};
- \node[noha] (rr) at (.6, 3.0) {RR};
- \node[noha] (si) at (-.4, 2.4) {SI};
- \node[noha] (s) at (-.4, 3.5) {CSR};
+ \node[noha] (cs) at (-1.2, 1.8) {CS};
+ \node[noha] (rr) at (-.8, 2.7) {RR};
+ \node[noha] (si) at (2.2, 2.4) {SI};
+ \node[noha] (s) at (0, 3.5) {CSR};
  \node[noha] (sr) at (2.2, 3.6) {1SR};
@@
  \draw [->] (ru) -- (rc);
- \draw [->] (rc) -- (ici);
  \draw [->] (ici) -- (pci);
@@
 
- \draw [->, red] (ici) -- (cs);
+ \draw [->, red] (rc) -- (cs);
  \draw [->, red] (cs) -- (rr);
  \draw [->, red] (pci) -- (si);
+ \draw [->, red] (ici) -- (rr);
  \draw [->, red] (rr) -- (s);
@@
 T_1 &: w_x(1)~w_y(1)
-\\T_2 &: w_x(2)~w_y(1)
+\\T_2 &: w_x(2)~w_y(2)
 \\T_3 &: r_x(a)~r_y(b)
@@
 
-Composition cost varies by property. For instance, Charron-Bost has
-proven that, to capture causality between $N$ communicating processes,
-standard vector-based approaches face an upper bound of $O(N)$. This
-means that, with $100,000$ clients, each write might be accompanied by
-$100,000$ vector timestamps. This is difficult to scale. By
-compromising on availability (e.g., treating a datacenter as a
-linearizable cluster), this overhead can be reduced, but it is much
+Composition cost also varies by property. For instance, Charron-Bost
+has proven that, to capture causality between $N$ communicating
+processes, standard vector-based approaches face an upper bound of
+$O(N)$~\cite{charron-bost} storage per write. This means that, with
+$100,000$ clients, each write might be accompanied by $100,000$
+timestamps per vector. This is difficult to scale. By compromising on
+availability (e.g., treating a datacenter as a linearizable cluster),
+this overhead can be reduced~\cite{cops, eiger}, but it is much
 cheaper to provide, say, read your writes, compared to causality.
@@
 \noindent The relationship between \textbf{Transactional Atomicity and
-  Linearizability} is not immediately clear. One might attempt to find
-a mapping between the two: Transactional Atomicity dictates that
-writes to multiple keys across multiple servers are made visible to
-readers all at once, while Linearizability dictates taht writes to a
-single key on multiple servers are made visible to all readers at
-once. The difference is two-old. First, in linearizable (and safe and
-regular) systems, writes are made visible to all clients immediately
+  Linearizability} is non-obvious. One might attempt to find a mapping
+between the two: Transactional Atomicity dictates that writes to
+multiple keys across multiple servers are made visible to readers all
+at once, while Linearizability dictates that writes to a single key on
+multiple servers are made visible to all readers at once. The
+difference is two-old. First, in linearizable (and safe and regular)
+systems, writes are made visible to all clients \textit{immediately}
 after they finish. With Transactional Atomicity, there is no recency
 requirement. Second, in linearizabile systems, all clients see all
-writes at the same time. With Transactional Atommicity, clients see
-writes at different times depending on which relicas they contact. We
-are not aware of a comparable analog in the distributed systems
+writes at the same time. With Transactional Atomicity, clients see
+writes at different times depending on which replicas they contact. We
+are not aware of an analogous model in the distributed systems
 literature. Accordingly, despite apparent similarities, Transactional
@@
 highly available system by waiting until all servers have seen a
-write. However, this leads to horrible visibility---new writes will
-not become visible to clients in the presence of partitions. A client
-that does not want to guarantee read-your-writes (due to the sticky
-availability requirement) may still wish to read other clients' writes
-with timeliness. Accordingly, a system may forgo high availability in
-order to provide better visibility. Alternatively, a system may
-enforce stickiness in the absence of partitions, then ``lose'' session
-guarantees by having clients reconnect to an available replica. This
-latter strategy provides session guarantees between partitions, and
-each partition event may cause a ``regression'' in the desired order
-of partitions, albeit at the benefit of availablilty.
-
-It is increasingly common in the practically-oriented distributed
-systems literature to assume that clients will be able to remain
-sticky with servers within its datacenter. More generally, if
-client-server sticky pairs are partitioned into disjoint sets, or
-groups, then additional guarantees like transactional atomicity can be
-accomplished with better visibility: writes can be made visible once
-they are present on all other servers within a group (as opposed to
-all servers in the system).
+write. However, this incurs severe visibility penalties---new writes
+will not become visible to clients in the presence of partitions. A
+client that does not want to guarantee read-your-writes (due to the
+sticky availability requirement) may still wish to read other clients'
+writes with timeliness. Accordingly, a system may forgo high
+availability in order to provide better visibility. Alternatively, a
+system may provide stickiness in the absence of partitions, then
+``lose'' session guarantees by having clients reconnect to an
+available replica. This latter strategy provides session guarantees in
+the absence of partitions but, to provide availability, sacrifices
+them in the event of partitions.

### 2013-03-25 writeup/model.tex

--- 2013-03-18 writeup/model.tex
@@
 
-highly available algorithms guaranteeing
-``always on'' operation are applicable in modern distributed
-contexts. This is because, a high level, a highly available system
-(which we formally define in Section~\ref{sec:availabilty}) guarantees
-two useful properties: during network partitions, the system responds
-to requests without blocking, and, during normal, partition-free
-operation, it provides guaranteed one round-trip-time (RTT) latency.
+To understand which guarantees can be provided with high availabilty,
+we must first understand what properties define high availability. In
+this section, we will formulate a model that captures a range of
+availability models, including ``true'' high availability,
+availablility with stickiness, and several other useful variants.
 
-In this section, we survey evidence regarding the prevalence of
-network partitions as well as quantify the cost of round-trip times
-over public cloud infrastructure. These results are independent of
-algorithms employed and, while they reflect state-of-the-art
-technology, they are
+Informally, highly available algorithms ensure ``always on'' operation
+and promise a guaranteed response. If users of a highly available
+system are able to contact a (set of) server(s) in a system, they are
+guaranteed a response as the (set of) server(s) does not need to
+communicate with any others. This lack of fast-path coordination also
+means that a highly available system can also offer low latency
+operation~\cite{abadi-pacelc}. To describe whether a transactional
+system is highly available, we need both a way to describe what
+servers a client must contact as well as what kinds of responses a
+server can provide, especially given the possibility of aborts.
 
-Why High Availability?
-* Highly available system operation is often desirable; if you can contact one server, you can get a response!
-* As a corrollary, an "AP" system is faster than a "CP" system; see "PACELC" by Abadi.
-	* Look, here are some datapoints we recently measured on EC2. Network times are at least 10x faster, often much more for "AP"
+\subsection{Replica Availability}
 
-* System model
-	* Consider transactions that issue a series of reads and writes, then commit or abort
-	* Each data item is located on one or more servers
+Traditionally, a system provides high availability if every user that
+can contact a server eventually receives a response from that server,
+even in the presence of arbitrary, indefinitely long network
+partitions between servers~\cite{gilbert-cap}. This is a useful
+definition: if we can contact at least one server, then we can receive
+a response. However, there are several other cases we should consider.
 
-* HA Definitions
-	* Straw-man definition: every transaction will recieve a response
-		* Problem: system that always aborts is trivially available!
+First, clients may wish to contact multiple servers. For example, to
+ensure that the effects of its operations will persist in the face of
+server failures, clients may wish to wait until operations reach more
+than one server before they considers operations completed. However,
+according to our definition of high availability, server-fault
+tolerance is unavailable. To allow flexibility in the number of
+servers required for an operation, we propose
+\textit{$K$-availability}: if a client can contact $K$ servers, then
+it eventually receives a response from all of them, even in the
+presence of indefinitely long network partitions. Compared to a system
+providing high availability ($1$-availability), a system with
+$2$-availability and will require more messaging, possibly higher
+latency and more unsuccessful operations in the presence of network
+and server failures.  Additionally, as is commmon in standard
+consensus algorithms like Paxos, clients may want to contact a
+majority of $N$ servers in a system. This \textit{majority
+  availability} (a special case of $K$-availability with $K=\lceil
+\frac{N}{2} \rceil$) is more available than $N$-availability.
 
-* Replica-HA
-	* (Handwaving in the outline) If a client can contact at least one replica for each data item in a transaction, then the transaction will eventually commit, even in the presence of infinite network partitions.
+Clients may also wish to contact the same server (or set of servers)
+across subsequent operations. As we will discuss in
+Section~\ref{sec:hats}, clients can ensure continuity between
+operations (e.g., reading their prior updates to a data item) by
+maintaining affinity or ``stickiness'' with a server or set of
+servers~\cite{vogels-defs}. Simultaneously aintaining availability and
+stickiness requires fate-sharing between the client and its set of
+servers: if a client's sticky servers fail, or, equivalently, if the
+client is partitioned from its sticky servers, then it may face
+unavailability. We say that a system provides \textit{sticky
+  availability} if a client's subsequent operations are executed
+against a server that has observed all of its prior operations, then
+it eventually receives a response, even in the presence of
+indefinitely long partitions. As above, it may be useful for a client
+to achieve stickiness \textit{and} contact many replicas (i.e., for
+fault tolerance), so we generalize sticky availability to
+\textit{$K$-sticky availability}, which provides sticky availability
+with $K$ distinct servers. A client may choose to become $1$-sticky
+available by acting as a server itself; for example, a client might
+cache its reads and writes. However, a single client cannot achieve
+$K$-sticky availability with $K>1$ on its own. Moreover, ``sticking''
+all clients to a designated set of servers (i.e., master availability)
+is a special case of sticky availability and is less available than
+the equivalent sticky availability.
 
-* Sticky-HA
-	* If a client continues to contact the same set of replicas across transactions and the client can contact at least one replica for each data item in a transaciton, then the transaction will eventually commit, even in the presence of infinite network partitions.
-	* We will use this definition when it comes to causality and read your writes guarantees!
-	* Can be achieved by either "sticking" clients to the same set of servers or by having the client act as a replica itself, via caching
+\begin{figure}
+\centering
+\begin{tikzpicture}[scale=0.8]
+  \tikzstyle{every node}=[font=\small]
+ \node[draw=none,fill=none] (1) at (0,0) {$1$-availability (high availability)}; 
+ \node[draw=none,fill=none] (2) at (0,1) {$2$-availability}; 
+ \node[draw=none,fill=none] (m-1) at (0, 2) {\ldots}; 
+ \node[draw=none,fill=none] (m) at (0, 3) {majority availability}; 
+ \node[draw=none,fill=none] (m+1) at (0, 4) {\ldots}; 
+ \node[draw=none,fill=none] (n) at (0, 5) {$N$-availability}; 
+ \node[draw=none,fill=none] (1s) at (4, 1.5) {$1$-sticky availability};
+ \node[draw=none,fill=none] (2s) at (4, 2.5) {$2$-sticky availability};
+ \node[draw=none,fill=none] (n2s) at (4, 3.5) {\ldots};
+ \node[draw=none,fill=none] (n1s) at (4, 4.5) {$(N$$-$$1)$-sticky availability};
 
-* As an example of a real system going from "CP" to "AP", [PNUTS dropped per-record mastering](\url{http://developer.yahoo.com/blogs/ydn/posts/2010/06/sherpa_update/#4})
+ \draw [->] (1) -- (2);
+ \draw [->] (2) -- (m-1);
+ \draw [->] (m-1) -- (m);
+ \draw [->] (m) -- (m+1);
+ \draw [->] (m+1) -- (n);
+ \draw [->] (1) -- (1s);
+ \draw [->] (1s) -- (m);
+ \draw [->] (2s) -- (m+1);
+ \draw [->] (1s) -- (2s);
+ \draw [->] (2) -- (2s);
+ \draw [->] (2s) -- (n2s);
+ \draw [->] (n2s) -- (n1s);
+ \draw [->] (n1s) -- (n);
 
-What about existing systems?
-* Database technology developed for single-node systems; gold standard: serializability
-	* Serializability is not actually highly available; give an example!
-	* Much of the database literature presumes serializability and does not consider high availability.
-	* For the literature that doesn't presume serializability, it's not presented in a HA context?
-* Consequence: traditional systems are not optimized for high availability
-	* Consider two-phase locking in a distributed environment
-	* As we will see (forward reference to Evaluation), many existing transactional systems encounter similar difficulties
-* Key question in this paper: What transactional semantics are highly available, and which aren't?
-	* Side note: there are infinite incomparable consistency models (e.g., always return 2, always return 3, …)
-	* Our goal is to unify distributed systems literature with ACID transaction model.
-* Argument: Highly Available Transactional Systems (HATS) require rethinking existing models. This work is a first step.
+\end{tikzpicture}
+\label{fig:availability-order}
+\caption{Hierarchy of replica availability levels for $N>3$ servers.}
+\end{figure}
 
-In this sense, HATs are similar to RAID: optimizing for graceful
-handling of a worst-case failure scenario improves average-case
-performance.
+We show the hierarchy of replica availabilty levels in
+Figure~\ref{fig:availability-order}. $K$-sticky availability subsumes
+$K$-availability, but $K$-sticky availability is incomparable with
+$(K+1)$-availability. $N$-sticky availability is equivalent to
+$N$-availability, while majority availability subsumes $1$-sticky
+availability More generally, $\lceil \frac{N}{2} \rceil$$+$$K$$-$$1$
+availability subsumes $K$-sticky availability. We have omitted
+discussion of operation-specific availability levels (e.g.,
+$N$-availability for writes and $1$-availability for reads in a
+write-all, read-one data store), system membership changes, or
+heterogeneous replicas (e.g., servers in a local and remote
+datacenters) but believe there are several avenues for further
+taxnomization.
+
+\subsection{Transactional Availability}
+
+Thus far, we have focused on single-operation availability. This is
+standard in distributed systems literature (e.g., atomic, regular, and
+safe register models all concern single objects~\cite{herlihy-art}),
+yet the database literature largely focuses on transactions: groups of
+multiple operations over multiple objects. Accordingly, by itself,
+replica availability is insufficient to describe availability
+guarantees for transactions. Additionally, given the choice of
+\textit{commit} and \textit{abort} responses---which signal
+transaction success or failure to a client---we must be careful in
+defining availability.
+
+If a client wishes to execute a transaction that performs operations
+on multiple data items, then the client must be able to contact and
+receive a response from at least one server responsible for each data
+item. This may result in lower availability than a non-transactional
+requirement. Additionally, given the possiblity of aborts, we need to
+ensure useful forward progress. A system can trivially guarantee
+clients a response by always aborting their
+transactions~\cite{transaction-liveness}. However, this is an
+unsatisfactory system because nothing good (transaction commit) ever
+happens; instead, we require a \textit{liveness} property. We cannot
+guarantee that every transaction will commit---transactions may choose
+to abort themselves---but we need to make sure that the system will
+not indefinitely abort transactions on its own volition. We call a
+transaction abort due to a transaction's own choosing (e.g., as part
+of the transaction itself or due to a would-be integrity constraint
+violation) an \textit{internal abort} and an abort due to system
+implementation an \textit{external abort}.
+
+We say that a system provides \textit{transactional availability} if,
+given replica availability for every data item in a transaction, the
+transaction eventually commits or internally aborts. For example,
+system will violate transactional $2$-availability if client that can
+contact two servers for each of data items in its transaction does not
+eventually commit in the absence of internal aborts.

### 2013-03-25 ycsb-0.1.4/core/src/main/java/com/yahoo/ycsb/TerminatorThread.java
lol, thread.stop()

@@ -60,6 +60,7 @@ public class TerminatorThread extends Thread {
           if (t.isAlive()) {
             System.err.println("Still waiting for thread " + t.getName() + " to complete. " +
                 "Workload status: " + workload.isStopRequested());
+            t.stop();
           }
         } catch (InterruptedException e) {
           // Do nothing. Don't know why I was interrupted.

### 2013-03-25 ycsb-0.1.4/core/src/main/java/com/yahoo/ycsb/TerminatorThread.java
don't wait

@@ -61,6 +61,7 @@ public class TerminatorThread extends Thread {
             System.err.println("Still waiting for thread " + t.getName() + " to complete. " +
                 "Workload status: " + workload.isStopRequested());
             t.stop();
+            break;
           }
         } catch (InterruptedException e) {
           // Do nothing. Don't know why I was interrupted.

### 2013-03-25 thebes-code/hat/server/src/main/java/edu/berkeley/thebes/hat/server/dependencies/DependencyResolver.java
Reduce time we own lock & reduce memory overhead

@@ -54,7 +54,6 @@ public class DependencyResolver {
     
     private final AntiEntropyServiceRouter router;
     private final IPersistenceEngine persistenceEngine;
-    private final ConcurrentMap<String, Set<PendingWrite>> pendingWritesMap;
     private final ConcurrentMap<Version, TransactionQueue> pendingTransactionsMap;
     private final ConcurrentMap<Version, AtomicInteger> unresolvedAcksMap;
     
@@ -64,7 +63,6 @@ public class DependencyResolver {
             IPersistenceEngine persistenceEngine) {
         this.persistenceEngine = persistenceEngine;
         this.router = router;
-        this.pendingWritesMap = Maps.newConcurrentMap();
         this.pendingTransactionsMap = Maps.newConcurrentMap();
         this.unresolvedAcksMap = Maps.newConcurrentMap();
         this.unresolvedAcksLock = new ReentrantLock();
@@ -73,11 +71,9 @@ public class DependencyResolver {
     public void addPendingWrite(String key, DataItem value) {
         Version version = value.getVersion();
         
-        pendingWritesMap.putIfAbsent(key, new ConcurrentSkipListSet<PendingWrite>());
         pendingTransactionsMap.putIfAbsent(version, new TransactionQueue(version));
         
         PendingWrite newPendingWrite = new PendingWrite(key, value);
-        pendingWritesMap.get(key).add(newPendingWrite);
         
         TransactionQueue transQueue = pendingTransactionsMap.get(version);
         transQueue.add(newPendingWrite);
@@ -90,58 +86,57 @@ public class DependencyResolver {
         // Check any unresolved acks associated with this key
         // TODO: Examine the implications of this!
         unresolvedAcksLock.lock();
+        boolean shouldCommit = false;
         try {
             if (unresolvedAcksMap.containsKey(version)) {
                 AtomicInteger numAcksForTransaction = unresolvedAcksMap.get(version);
-                for (int i = 0; i < numAcksForTransaction.get(); i ++) {
-                    ackAndMaybeCommit(transQueue);
+                for (int i = 0; i < numAcksForTransaction.get() && !shouldCommit; i ++) {
+                    shouldCommit = transQueue.serverAcked();
                 }
                 unresolvedAcksMap.remove(version);
             }
         } finally {
             unresolvedAcksLock.unlock();
-        }            
-        
-        if (Math.random() < .001) {
-            int numPendingWrites = 0;
-            for (Set<PendingWrite> pendingWrites : pendingWritesMap.values()) {
[…]

### 2013-03-25 writeup/intro.tex

@@
 networks has led to a large number of systems prioritizing high
-availability, which brings with it guaranteed low latency, which is especially
-important in geo-replicated contexts. Accordingly, we observe a dearth
-of systems providing general-purpose multi-operation
-\textit{transactions} over multiple data items with useful semantics.
+availability and guaranteed low latency, especially important in
+geo-replicated contexts. We observe a dearth of systems providing
+general-purpose multi-operation \textit{transactions} over multiple
+data items with useful semantics.
 
@@
 and up-and-coming ``NewSQL'' databases and often as the maximum
-guarantees offered. While the prevalence of these weak isolation
-configurations is encouraging, the current literature does not provide
-guidance as to the operation of weakly consistent isolation guarantees
-in a distributed context. Rather, we have a strong understanding of
-weak isolation in the single-server context from which they originated
-as well as several techniques for providing distributed
-serializability but little to no guidance as to which ACID guarantees
-are compatible with Highly Available Transactions (HATS), or can be
-provided with high availability and low latency.
+guarantees offered (Section~\ref{sec:modernacid}). While the
+prevalence of these weak isolation configurations is encouraging, the
+current literature does not provide guidance as to the operation of
+weakly consistent isolation guarantees in a distributed
+context. Rather, we have a strong understanding of weak isolation in
+the single-server context from which it originated as well as several
+techniques for providing distributed serializability but little to no
+guidance as to which ACID guarantees are compatible with Highly
+Available Transactions (HATs), or can be provided with high
+availability and low latency.
 
@@
 database consistency by analyzing the relationship between so-called
-session guarantees and ``register'' semantics. In performing this
-taxonomization, we also introduce a hierarchy of highly available
-system guarantees as, for example, some guarantees require
-``stickiness,'' or fate-sharing between clients and database
+session guarantees, ``register'' semantics, and ACID models. In
+performing this taxonomization, we also introduce a hierarchy of
+highly available system guarantees as, for example, some guarantees
+require ``stickiness,'' or fate-sharing between clients and database
 servers. We believe that this analysis is the first that has been
@@
 
-While this model classification is useful, we seek to understand the
-implications of choosing to employ high availability. HATs provide
-high availability and low latency but suffer from weakened semantic
-guarantees. Traditional ACID properties provide strong semantic
-guarantees but do not provide high availability or low latency. We
-attempt to objectively study these virtues and limitations by
-performing an extensive survey of practitioner accounts and research
-literature, performing experimental analyses on modern cloud
-infrastructure, and analyzing representative applications for their
-semantic requirements. While our results are not definitive, they
-suggest that HATs offer a one to three order of magnitude latency
-decrease compared to traditional protocols, while providing acceptable
-semantics for programs with monotonic and commutative
-updates. Similarly, HAT guarantees can be used to ensure that
-arbitrary foreign key constraints are satisfied but not necessarily
-uniqueness constraints. For programs with non-monotonic transactional
-requirements, HATs may fall short, necessitating greater coordination
-strategies. Speed and availability are not without a price, but we
-attempt to estimate the size of the tag. Succintly, \textit{what is
-  benefits do highly available transactions offer and what is their
-  cost?}
+While this model classification is useful, we also seek to understand
+the implications of choosing to employ highly available systems. HATs
+provide high availability and low latency but cannot offer all of the
+semantic guarantees of unavailable systems. Traditional ACID
+properties provide strong semantic guarantees but do not provide high
+availability or low latency. We attempt to objectively study these
+virtues and limitations by performing an extensive survey of
+practitioner accounts and research literature, performing experimental
+analyses on modern cloud infrastructure, and analyzing representative
+applications for their semantic requirements. While our results are
+not definitive, they suggest that HATs offer a one to three order of
+magnitude latency decrease compared to traditional protocols while
+providing acceptable semantics for many programs, particularly those
+with purely monotonic logic and commutative updates. HAT guarantees
+can also ensure arbitrary foreign key constraints on updates and
+sometimes provide uniqeness guarantees. However, for programs with
+non-monotonic logic, HATs may fall short, requiring more
+coordination-intensive mechanisms. Concisely, we attempt to understand
+\textit{what benefits do highly available transactions offer and what
+  is their cost?}
 
@@
 \begin{itemize}
-\item We present a model for high availability in a transactional environment, including Replica-agnostic High Availability (R-HA) and Sticky High Availability (S-HA).
-\item We analyze the existing literature on ACID properties (particularly, ANSI SQL Spec, Adya, Gray) to show which properties* with a focus on isolation* are achievable in HATS.
-Achievable properties include transactional atomicity, variants of repeatable read isolation, and session guarantees.
-We also describe unavailable properties, like preventing Lost Update and Write Skew.
-We believe that this is the first such characterization in the literature.
-\item We further motivate the study of HATS via the analysis of existing concurrency control algorithms in a highly available environment as well as through deployment of an experimental database prototype on public cloud infrastructure.
+\item We present a model for high availability in a transactional
+  environment, replica-based High Availability, Sticky High
+  Availability, and Transactional High Availability.
+
+\item We analyze the existing literature on ACID properties and
+  replicated data semantics to show which properties are achievable in
+  HAT systems. Achievable properties include transactional atomicity,
+  variants of repeatable read isolation, and session guarantees. We
+  also describe unavailable properties, like preventing Lost Update
+  and Write Skew. We believe that this is the first such
+  characterization in the literature.
+
+\item We further motivate the study of HATs via the analysis of
+  existing concurrency control algorithms, a case-study of running a
+  transactional application with HATs, and via deployment of a HAT
+  database prototype on public cloud infrastructure.
 \end{itemize}

### 2013-03-25 thebes-code/hat/server/src/main/java/edu/berkeley/thebes/hat/server/dependencies/DependencyResolver.java
Throw TException as appropriate

@@ -112,7 +112,7 @@ public class DependencyResolver {
 //        }
     }
     
-    private void commit(TransactionQueue queue) {
+    private void commit(TransactionQueue queue) throws TException {
         for (PendingWrite write : queue.pendingWrites) {
             persistenceEngine.put(write.getKey(), write.getValue());
         }

### 2013-03-26 thebes-code/common/src/main/java/edu/berkeley/thebes/common/persistence/util/LockManager.java
Fix lockmanager bug; now we actually GC entries and no one waits on an unlocked lock

@@ -1,30 +1,61 @@
 package edu.berkeley.thebes.common.persistence.util;
 
-import java.util.WeakHashMap;
+import com.google.common.collect.Maps;
+
+import java.util.Map;
 import java.util.concurrent.locks.Condition;
 import java.util.concurrent.locks.Lock;
 import java.util.concurrent.locks.ReentrantLock;
 
 public class LockManager {
-    private WeakHashMap<String, Condition> lockTable;
+
+    private class LockWaiting {
+        private long numWaiters;
+        private Condition condition;
+
+        public LockWaiting(Condition condition) {
+            numWaiters = 0;
+            this.condition = condition;
+        }
+
+        public boolean isEmpty() {
+            return numWaiters == 0;
+        }
+
+        public void signalWaiter() {
+            assert(numWaiters > 0);
+            condition.signal();
+        }
+
+        public void waitInterruptibly() {
+            numWaiters++;
+            condition.awaitUninterruptibly();
+            numWaiters--;
+        }
+    }
+
+    private Map<String, LockWaiting> lockTable;
     private Lock tableLock = new ReentrantLock();
 
     public LockManager() {
-        lockTable = new WeakHashMap<String, Condition>();
+        lockTable = Maps.newHashMap();
     }
 
     public void lock(String key) {
         tableLock.lock();
         if(!lockTable.containsKey(key))
-            lockTable.put(key, tableLock.newCondition());
+            lockTable.put(key, new LockWaiting(tableLock.newCondition()));
         else
-            lockTable.get(key).awaitUninterruptibly();
+            lockTable.get(key).waitInterruptibly();
         tableLock.unlock();
     }
 
     public synchronized void unlock(String key) {
         tableLock.lock();
-        lockTable.get(key).signal();
+        if(lockTable.get(key).isEmpty())
+            lockTable.remove(key);
+        else
+            lockTable.get(key).signalWaiter();
         tableLock.unlock();
     }
 }

### 2013-03-26 writeup/conclusion.tex
Another pass at writing

@@
 
-VIII. Conclusion (.5pg)
+In this paper, we have begun to understand the relationship between
+multi-operation, multi-object transactions and high availability. We
+have outlined requirements for high availability in a transactional
+context, or Highly Available Transactions (HATs) and taxonomized
+guarantees spanning both distributed systems and ACID database
+literature accordingly. Many transactional guarantees are achievable
+with high availability, while a few, like preventing Lost Update and
+Write Skew, are not. To understand the benefits of high availability
+and low latency, we classified the prevalence and cost of partitions
+and network latency both in the literature and via experiment, while
+analyzing existing transactions to determine whether HATs were
+sufficient to ensure application consistency.
 
-HATS represent a new class of database system:
-* Can't get serializability, but can get other properties
-* Key challenge: designing distributed algorithms and architectures for partial failure, asynchrony
-* Rearchitecture of DBMSs and better understanding of applications are key
-
-References (1pg)
+Fundamentally, high availability represents a new constraint for
+database designs. While the CAP Theorem has had a large impact on
+distributed data storage design, our knowledge of database systems has
+not kept pace. The key distributed systems challenge of designing
+distributed algorithms and architectures for partial failure and
+network asynchrony is largely unaddressed in the literature on
+transaction processing systems. We believe that, in order to fully
+realize the potential of HAT systems, further rearchitecture and study
+of existing database designs as well as better understanding of
+application semantics and their relationship to distributed
+consistency will be necessary.

### 2013-03-26 writeup/evaluation.tex
Another pass at writing

@@
 With a firm understanding of which of several ACID and distributed
-consistency properties are available and not, we perform analysis of
-their implications for existing algorithms, applications, and
-deployments. Specifically:
+consistency properties are available and which are not, we begin to
+understand their implications for existing algorithms, applications,
+and deployments. Specifically:
 
@@
 
-We begin by considering several existing algorithms that provide
-serializability but are not highly available. In summary, all
-algorithms in this class require at least one round trip time (RTT)
+\vspace{.5em}\noindent\textbf{Serializablity} We begin by considering several
+existing algorithms that provide serializability but are not highly
+available. In summary, all algorithms in this
+class~\cite{bernstein-book} require at least one round trip time (RTT)
 before committing.
 
-\begin{itemize}
-\item \textbf{Two-phase locking} (ignoring deadlocks) requires $T$
-  lock operations and 1 unlock operation for a transaction of length
-  $T$. If all items are located on a single master server, this
-  requires one RTT (for requests originating from servers other than
-  the master). If items are located on $S$ ($S \leq T$) different
-  servers, two-phase locking requires, at minimum, $2(S-1)$ RTTs---one
-  for lock and another for unlock.
+\vspace{.5em}\noindent\textit{Two-phase locking} (ignoring deadlocks) requires $T$ lock
+operations and 1 unlock operation for a transaction of length $T$. If
+all items are located on a single master server, this requires one RTT
+(for requests originating from servers other than the master). If
+items are located on $S$ ($S \leq T$) different servers, two-phase
+locking requires, at minimum, $2(S-1)$ RTTs---one for lock and another
+for unlock.
 
-\item \textbf{Optimistic concurrency control} allows reads from any
-  server but transactions must be checked by a scheduler before
-  commit. With one scheduler per database, this requires one RTT
+\vspace{.5em}\noindent\textit{Optimistic concurrency control} allows
+reads from any server but transactions must be checked by a scheduler
+before commit. With one scheduler per database, this requires one RTT
 specifically for concurrency control. If schedulers are sharded by
@@
 contacting several different schedulers and, possibly, an atomic
-  commit protocol, which becomes substantially more expensive.
+commit protocol, which becomes substantially more
+expensive.
 
-\item \textbf{Deterministic transaction scheduling} is ffectively the
-  inverse of optimistic concurrency control: a scheduler first orders
-  the transactions, then sends them to database replicas for in-order
-  execution. These techniques, pioneered in the context of totally
-  ordered atomic broadcast systems, have recently seen increased
-  attention and effectively require communication overhead similar to
-  optimistic concurrency control.
+\vspace{.5em}\noindent\textit{Deterministic transaction scheduling} is
+effectively the inverse of optimistic concurrency control: a scheduler
+first orders the transactions, then sends them to database replicas
+for in-order execution. These techniques, pioneered in the context of
+totally ordered atomic broadcast
+systems~\cite{deterministic-scheduling}, have recently seen increased
+attention~\cite{calvin} and effectively require communication overhead
+similar to optimistic concurrency control. Finally,
 
-\item\textbf{Multiversion concurrency control} allows readers to read
-  older versions of data items but, like OCC, must eventually be
-  checked for write-write conflicts, requiring a scheduler.
+\vspace{.5em}\noindent\textit{Multiversion concurrency control} allows readers to read older
+versions of data items but, like OCC, must eventually be checked for
+write-write conflicts, requiring a scheduler.
 
-\end{itemize}
-
-Perhaps unsurprisingly, the reliance on a glocal total order
-necessitates a minimum of one round-trip to a designated master or
-coordination service for each of these classic algorithms. This is
+\vspace{.5em} Perhaps unsurprisingly, the reliance on a global total
+order necessitates a minimum of one round-trip to a designated master
+or coordination service for each of these classic algorithms. This is
 simply a cost of providing serializability, and the cost of the
@@
 
-There also are several mechanisms that do not provide serializability
-but provide transactional semantics with varying availability
-properties:
+\vspace{.5em}\noindent\textbf{Non-serializablity} While it is possible to use
+several of the prior mechanisms for HAT systems---for example, a
+multi-version storage system is a reasonable candidate for the basis
+of a HAT database, albeit with write conflict checking
+disabled---several techniques, like locking, are not suitable for
+HATs. In fact, there are several mechanisms that do not provide
+serializability but provide transactional semantics with varying
+availability properties:
 
-\begin{itemize}
-\item \textbf{Lock-based protocols} such as those used to provide weak
-  isolation in Gray's original paper require at least one RTT to the a
-  lock server (or, alternatively, a majority quorum). In the presence
-  of failure, distributed locking is challenging.
-\item The \textbf{MDCC} commit protocol uses a Paxos variant and escrow
-  transactions to provide Repeatable Read isolation with 1-2 RTTs. (TODO: CHECK)
-\item \textbf{Parallel Snapshot Isolation} as developed by the Walter
-  system provides (unavailable) snapshot isolation variant and, even
-  when writes commute, requires 1 RTT to enforce read
-  recency. However, PSI without recency requirements could be made
-  highly available.
-\item \textbf{Eiger} provides read-only and write-only transactions
-  with item cut isolation and causal consistency. It relies on sticky
-  high availability and groups within a datacenter with linearizable
-  clusters. In the presence of server failure during write
-  transactions, reads and writes may become unavailable. \textbf{COPS}
-  provides similar semantics for read-only transactions but does not
-  suffer from unavailability as it does not have a coordinator.
-\item \textbf{Chan and Gray's Read-only transactions} use a similar
-  algorithm to Eiger to provide read-only transactions with item cut
-  isolation and causal consistency but coordinate on reads instead of
-  writes.
-\item \textbf{Bolt-on Causal Consistency} and \textbf{Swift} can
-  provide read-only and write-only transactions with item cut
-  isolation and causal consistency with sticky high availability and
-  both rely on client-side caching.
-\item \textbf{Bayou} provides several session guarantees but, without
-  client-side caching, actually faces unavailability in the presence
-  of failure. (TODO: CHECK)
-\item \textbf{Brantner's S3 Database} uses caching to provide several
-  guarantees such as XX and YY but is unavailable during failures
-  (TODO: CHECK).
-\end{itemize}
+\vspace{.5em}\noindent\textit{Lock-based protocols} such as those used to provide weak
+  isolation in Gray's original proposal~\cite{gray-isolation} require
+  at least one RTT to the a lock server (or, alternatively, a majority
+  quorum). In the presence of failure, distributed locking is
+  challenging.
+
+\vspace{.5em}\noindent The \textit{MDCC} commit protocol uses a Paxos
+variant and escrow transactions to provide read committed (without
+Lost Update) and Non-monotonic Snapshot Isolation with 1-2 RTTs but is
+unavailable due to Lost Update~\cite{mdcc}.
+
+\vspace{.5em}\noindent\textit{Parallel Snapshot Isolation} as
+developed by the Walter system provides (unavailable) snapshot
+isolation variant and, even when writes commute, requires 1 RTT to
+enforce read recency~\cite{walter}. However, like other Snapshot
+Isolation variants, PSI could be made highly available by giving up
+recency requirements and Lost Update prevention.
+
+\vspace{.5em}\noindent\textit\textbf{Eiger} provides read-only and
+write-only transactions with item cut isolation and causal
+consistency. It relies on sticky high availability with a linearizable
+set of replicas within a datacenter. This provides low latency across
+data centers. In the presence of server failure during write
+transactions, reads and writes may become unavailable. Eiger's
+predecessor \textit{COPS} provides similar semantics for read-only
+transactions but does not suffer from unavailability as it does not
+rely have a coordinator.
+
+\vspace{.5em}\noindent\textit{Chan and Gray's Read-only transactions}
+use a similar algorithm to Eiger to provide read-only transactions
+with item cut isolation and causal consistency but client version
+selection coordination is performed on reads instead of
+writes~\cite{readonly}.
+
+\vspace{.5em}\noindent\textit{Bolt-on Causal
+  Consistency}~\cite{bolton} and \textit{Swift}~\cite{swift} can
+provide read-only and write-only transactions with item cut isolation
+and causal consistency with sticky high availability and both rely on
+client-side caching.
+
+\vspace{.5em}\noindent\textit{Bayou} provides several session
+guarantees but, in its original implementation, may face unavailabilty
+in the presence of failure~\cite{sessionguarantees}.
+
+\vspace{.5em}\noindent\textit{Brantner's S3 Database} uses caching to
+provide several session and transactional guarantees on top of cloud
+infrastructure, often relying on linearizable queues and, accordingly,
+are often unavailable~\cite{kraska-s3}
+
+\vspace{.5em}
+
+As we have seen, it is possible to implement many guarantees weaker
+than serializability---including guarantees that are achievable with
+high availability---and not achieve high availability. These
+designs---though often unavailable---offer benefits unavailable to
+HATs: for example, lock-based read committed, provide recency
+guarantees in addition to read committed isolation! We expect that
+further understanding the trade-offs between semantic strength,
+performance, and storage and metadata requirements will remain an area
+of active research for the foreseeable future.
 
@@
 
-Recent work on the CALM Theorem, Lattice-based state update, and
-Commutative and Replicated Data Types demonstrates that, if updates
-logically commute, then they can be performed in different orders at
-different locations in a distributed system. Accordingly, as long as
-all writes are delivered to all replicas, then a system executing
-monotonic logic with commutative operators should not suffer from
-application-level consistency anomalies as a result of Lost Update or
-Write Skew anomalies. However, applications with non-monotonic state
-mutation will not, in general, be able to maintain application-level
-consistency constraints. Similarly, applications requiring
-write visibility between clients should opt for unavailability instead
-of possibly missing a write visibility deadline.
+\vspace{.5em}\noindent\textbf{Commutativity and Monotonicity} Recent work on
+the CALM Theorem~\cite{calm}, Lattice-based state
+update~\cite{blooml}, and Commutative and Replicated Data
+Types~\cite{crdt} demonstrates that, if updates logically commute,
+then they can be performed in different orders at different locations
+in a distributed system. Accordingly, as long as all writes are
+delivered to all replicas, then a system executing monotonic logic
+with commutative operators should not suffer from application-level
+consistency anomalies as a result of Lost Update or Write Skew
+anomalies. However, applications with non-monotonic state mutation
+will not, in general, be able to maintain application-level
+consistency constraints. Similarly, applications requiring write
+visibility between clients should opt for unavailability instead of
+possibly missing a write visibility deadline.
 
-As an example of an actual system rearchitecture,
-less than one year after its announcement, Yahoo!'s PNUTS developers
-explicitly added a support for weaker, highly available operation,
-explaining that ``strict adherence [to strong consistency] leads to
-difficult situations under network partitioning or server
-failures...in many circumstances, applications need a relaxed
-approach''~\cite{pnuts-update}.
-
-Absent a necessary and sufficient result describing exactly which
+\vspace{.5em}\noindent\textbf{TPC-C} Absent a necessary and sufficient result describing exactly which
 application logics can be achieved with high availability and to make
@@
 five transactions, capturing operation of a real-world warehouse,
-including sales, payments, and deliveries. We analyze each transaction
-and its interactions below.
+including sales, payments, and deliveries. We analyze each kind of
+transaction and its interactions below.
 
-\begin{itemize}
-\item \textbf{New-Order} places an order for a variable quantity of
-  data items, updating warehouse stock as needed. The New-Order
-  transaction selects a sales district, assigns the new order an order
+\vspace{.5em}\noindent\textit{New-Order} places an order for a
+variable quantity of data items, updating warehouse stock as
+needed. The transaction selects a sales district, assigns the order an
 ID number, adjusts the remaining warehouse stock, and writes a
@@
 First, assigning an order ID number is challenging. We can easily
-  assign a \textit{unique} order ID, by, say, concatenating the
-  customer ID with a timestamp and server ID. However, the TPC-C
-  specification requires that order numbers are assigned sequentially,
-  which we cannot guarantee in HATs due to Lost Update. Accordingly,
-  HATs can allow placement of new orders with unique identifiers but
-  cannot provide compliant TPC-C execution due to the non-monotonic
-  counter increase.
+assign a \textit{unique} order ID, by, say, concatenating the customer
+ID with a timestamp and server ID. However, the TPC-C specification
+requires that order numbers are assigned sequentially, which we cannot
+guarantee in HATs because assigning sequential IDs requires preventing
+Lost Update. Accordingly, HATs can allow placement of new orders with
+unique identifiers but cannot provide compliant TPC-C execution due to
+the requirement for a non-monotonic counter increase.
 
-  Second, decrementing inventory counts is a potentially risky
-  operation as, without coordination, inventory counts may become
-  negative. However, TPC-C automatically ``restocks'' inventory counts
-  (by 91 items) when there are 10 or fewer items
-  remaining. Accordingly, even in the presence of concurrent new
-  orders to the same item, the stock will never fall below
-  zero. However, an $R$ replica HAT system might end up with $91(R-1)$
-  more stock than a single-site implementation.
+Second, decrementing inventory counts is, in general, a potentially
+risky operation because, without coordination, inventory counts may
+become negative. However, TPC-C automatically ``restocks'' inventory
+counts (by 91 items) when there are 10 or fewer items
+remaining. Accordingly, even in the presence of concurrent new orders
+to the same item, the stock will never fall below zero. However, an
+$R$ replica HAT system might end up with $91(R-1)$ more stock than in
+a single-site implementation.
 
-\item \textbf{Payment} updates running balances for warehouses,
-  districts, and customer records as well as providing an audit
-  trail. All balance operations can be made commutative, so the
-  transaction can execute under HATs.
+\vspace{.5em}\noindent\textit{Payment} updates running balances for
+warehouses, districts, and customer records as well as providing an
+audit trail. All balance increase operations can be made commutative,
+so the transaction can execute under HATs.
 
-\item \textbf{Order-Status} is a read-only transaction and can be
-  safely executed under HATs. The TPC-C specification does not require
-  recency bounds but, in practice, they may be useful. If a client is
-  sticky with her replica set, she will read her own writes, which may
-  be sufficient.
+\vspace{.5em}\noindent\textit{Order-Status} is a read-only transaction
+and can be safely executed under HATs. The TPC-C specification does
+not require recency bounds but, in practice, they may be desirable. If
+a client is sticky with a set of replicas, she will read her own
+writes, which may be sufficient.
 
-\item \textbf{Delivery} represents the fulfillment of a pending order
-  created by a New Order transaction: it deletes the order from the
-  pending list, updates the customer's balance, updates the order's
-  carrier ID and delivery time, and updates the customer balance. The
-  challenge in delivery comes in deletion, which is non-monotonic: an
-  item might be removed twice from the pending table. If delivery can
-  be made idempotent---that is, if multiple Delivery transactions have
-  the same effect as one---then multiple clients can execute the same
-  Deliveries without fatal consequences.
+\vspace{.5em}\noindent\textit{Delivery} represents the fulfillment of
+a pending order created by a New Order transaction: it deletes the
+order from the pending list, updates the customer's balance, updates
+the order's carrier ID and delivery time, and updates the customer
+balance. The challenge in delivery comes in deletion, which is
+non-monotonic: an item might be removed twice from the pending
+table. If delivery can be made idempotent---that is, if multiple
+Delivery transactions have the same effect as one---then multiple
+clients can execute the same Deliveries without fatal consequences.
 
@@
 once, it is unlikely that multiple agents will attempt a Delivery
-  transaction for the same order simultaneously. However, in this
-  case, we have simply moved the non-monotonicity to the real world:
-  the carrier that picks up the package for an order must ensure that
-  she is the only carrier who has done so. While we can program around
-  the other operations in the order, HATs cannot--without the help of
+transaction for the same order simultaneously. However, in this case,
+we have simply moved the non-monotonicity to the real world: the
+carrier that picks up the package for an order must ensure that she is
+the only carrier who has done so. While we can program around the
+other operations in the order, HATs cannot--without the help of
 business-specific processes--serve Delivery transactions. That said,
-  New Orders may be placed.
+New-Orders may be placed with high availability.
 
-\item \textbf{Stock-Level} is another read-only transaction and can be
-  executed without any non-HAT guarantees, with the same caveats
-  regarding recency.
-\end{itemize}
+\vspace{.5em}\noindent\textit{Stock-Level} is another read-only
+transaction and can be executed without any non-HAT guarantees, with
+the same caveats regarding recency.
 
-Throughout execution, TPC-C requires the maintainence of several
-integrity constraints. For example, Consistency Condition 1 (3.3.2.1)
-requires that each warehouse's sales count must reflect the sum of its
-subordinate sales districts. This integrity constraint spans two
-tables but, given the ability to update both tables atomically via
-transactional atomicity, can be easily maintained. Consistency
-Conditions 4 through 12 (3.3.2.4-12) can similarly be satisfied by
-applying updates atomically across tables. Consistency Conditions 2
-and 3 (3.3.2.2-3) concern order ID assignment and are more
-problematic: effectively, each New Order placed by a customer in a
-district must have an ID that is one the prior order in that
-district. This requires preventing Lost Update within each district's
-table. However, as we have seen, the required transactions can execute
-without these two conditions.
+\vspace{.5em} Throughout execution, TPC-C also requires the
+maintainence of several integrity constraints. For example,
+Consistency Condition 1 (3.3.2.1) requires that each warehouse's sales
+count must reflect the sum of its subordinate sales districts. This
+integrity constraint spans two tables but, given the ability to update
+both tables atomically via transactional atomicity, can be easily
+maintained. Consistency Conditions 4 through 12 (3.3.2.4-12) can
+similarly be satisfied by applying updates atomically across
+tables. Consistency Conditions 2 and 3 (3.3.2.2-3) concern order ID
+assignment and are more problematic, as we have discussed.
 
 In summary, many---but not all---TPC-C transactions are well-served by
-HATs. The two problematic transactions---New Order and Payment---rely
+HATs. The two problematic transactions---New-Order and Payment---rely
 on non-monotonic state updates. The former transaction can be modified
-to ensure ID uniqueness at the expense of dense ID ordering, while the
-latter is an inherently non-monotonic action requiring external
-compensation or stronger consistency protocols. We expect that,
-especially for read-dominated workloads found in many online services,
-HAT guarantees may suffice for many applications. Finally, while TPC-C
-is not subject to write skew anomalies, we note that many TPC-E
-isolation tests, such as one that requires simultaneously modifying a
-product description and its image, are satisfied by HATs.
+to ensure ID uniqueness at the expense of sequential per-district ID
+ordering, while the latter is an inherently non-monotonic action
+requiring external compensation or stronger consistency protocols. We
+expect that, especially for read-dominated workloads found in many
+online services, HAT guarantees may suffice for many
+applications. Finally, while TPC-C is not subject to write skew
+anomalies, we note that many TPC-E isolation tests, such as one that
+requires simultaneously modifying a product description and its image,
+are satisfied by HATs.
 

### 2013-03-26 writeup/futurework.tex
Another pass at writing

@@
 \label{sec:futurework}
+
+We believe that there are several promising avenues for future work on
+Highly Available Transaction systems:
 
@@
 
-The current state of the art places several useful models within
-various categories in the availability space. This is still a
-practical result---take, for instance, CAP's impact on distributed
-systems of the last decade. But what property defines the boundaries
-between classes of highly available systems? Is there a meta-property
-that exactly characterizes what properties are achievable with high
-availability or sticky high availability? This is an area of ongoing
-research within the HAT project.
+This paper places several useful models within various categories in
+the availability space. This is still a practical result---take, for
+instance, CAP's impact on distributed systems of the last decade. But
+what property defines the boundaries between classes of highly
+available systems? Is there a meta-property that exactly characterizes
+what properties are achievable with high availability or sticky high
+availability? This is an area of ongoing research within the Berkeley
+HAT project.
 
@@
 ease of implementation of a HAT system will likely vary depending on
-which guarantees it attempts to provide. As we have alluded to, the
+which guarantees it attempts to provide. As we have discussed, the
 cost in moving from a sticky high available system to a truly highly
 available system seems expensive. As a concrete example, the Eiger and
-COPS systems use metadata garbage collection and assume linearizable
-clients within each datacenter in order to keep the overhead of
-providing causal consistency small. Without performing extensive
-client-side caching, these techniques do not seem applicable to a
-highly available implementation. This is not intended as a fault of
-the prior systems---they perform well for the task at hand---but hint
-at the complexity of moving between sticky and truly highly available
-systems.
+COPS systems~\cite{cops, eiger} use metadata garbage collection and
+assume linearizable clients within each datacenter in order to keep
+the overhead of providing causal consistency small. Without performing
+extensive client-side caching, these techniques do not seem applicable
+to a highly available implementation. This is not intended as a fault
+of the prior systems---they perform well for the task at hand---but
+hint at the complexity of moving between sticky and truly highly
+available systems.
 
@@
 performed a narrow case study in this work, but a general
-taxonomization of techniques such as escrow transactions and Sagas
-could help answer the challenge of programming in a weakly consistent
-environment. These techniques might be aided by considering models
-with service-level agreements or bounded messaging asynchrony. While
-our focus is on ``classic'' distributed systems trade-offs, there are
-several interesting hybrids to consider.
+taxonomization of techniques such as escrow transactions~\cite{escrow}
+and Sagas~\cite{sagas} could help answer the challenge of programming
+in a weakly consistent environment. These techniques might be aided by
+considering models with service-level agreements or bounded messaging
+asynchrony. While our focus is on ``classic'' distributed systems
+trade-offs, there are several interesting hybrids to consider.

### 2013-03-26 writeup/hat-analysis.tex
Another pass at writing

@@
 database and distributed systems literature, providing a brief,
-informal explanation of each guarantee and illustrate it with an
-example. In particular, for our ACID guarantees, we draw largely on
-Adya's dissertation~\cite{adya}, Berenson et al.'s 1995 critique of
-the ANSI SQL Specification~\cite{ansicritique}, and the ANSI SQL
+informal explanation and example for each guarantee. In particular,
+for our ACID guarantees, we draw largely on Atul Adya's
+dissertation~\cite{adya}, Berenson et al.'s 1995 critique of the ANSI
+SQL Specification~\cite{ansicritique}, and the ANSI SQL
 specification~\cite{ansi-sql}. We provide a set of formal definitions
-and semantics in the extended version of this work.
+and semantics in our extended Technical Report~\cite{hat-tr}.
 
@@
 To begin, we present achievable semantics, offering proof-of-concept
-algorithms to demonstrate feasibility.
+algorithms to demonstrate feasibility. These guarantees are fairly
+straightforward and have been introduced---albeit with greater brevity
+and discussion elided---in preliminary prior work~\cite{hat-hotos}. In
+our examples, we exclusively consider read and write operations. We
+denote a write of value $v$ to data item $d$ as $w_d(v)$ and a read
+from data item $d$ returning $v$ as $r_d(v)$ and assume that all data
+items have the null value, $\bot$, at database initialization. Unless
+otherwise specified, all example transactions commit.
 
@@
 Writes''---as above---and ``Dirty Reads'' phenomena; captured by
-Adya's $G1a-c$, ANSI's $P1$, and Berenson's ``broad'' $P1$
-(2.2)). For instance, in the example below, $T_3$ should never see
+Adya's $G1a-c$, ANSI's $P1$, and ``broad'' $P1$ (2.2) from Berenson et
+al.). For instance, in the example below, $T_3$ should never see
 $a=1$, and, if $T_2$ aborts, $T_3$ should never read $a=3$:
@@
 transaction. By itself, this property is rather weak---we can always
-read \texttt{null} for each data item. However, when coupled with
+read $\bot$ for each data item. However, when coupled with
 additional properties, like transactional atomicity, it is stronger:
@@
   Problem}, whereby two successive predicate-based reads return
-different data~\cite{gray-isolation}. We call predicate-based cut isolation, which prevents
-Phantoms, \textit{Phantom Cut Isolation}.\footnote{Oracle provides an
-  isolation level called ``Statement Level Read Consistency''; this is
-  analogous to Phantom Cut Isolation at the level of a single
-  statement. We do not consider this isolation model here as it is
-  subsumed by standard Snapshot Isolation and we believe our
-  discussion is easily extended to incorporate it.}
+different data~\cite{gray-isolation}. We call predicate-based cut
+isolation, which prevents Phantoms, \textit{Phantom Cut
+  Isolation}.\footnote{Oracle provides an isolation level called
+  ``Statement Level Read Consistency''; this is analogous to Phantom
+  Cut Isolation at the level of a single operation within a
+  multi-operation transaction. We do not consider this isolation model
+  here as it is subsumed by standard Snapshot Isolation and we believe
+  our discussion is easily extended to incorporate it.}
 
@@
 atomicity restricts transactions' ability to view the effects of
-partially completed transactions. Under transactionational atomicity,
+partially completed transactions. Under transactional atomicity,
 within each transaction, either all effects of another transaction are
@@
 Committed, we can read a subset of committed writes whereas, with
-transactional atomicity, we must read all of them. For example, if
-$T_1$ commits, then, given transactional atomicity, $T_2$ must observe
-$b=c=1$ (or later versions):
+transactional atomicity, we must read all of them. For example, $T_2$
+must observe $b=c=1$ (or later versions for each key):
 \vspace{-.5em}
@@
 perhaps again due to the single-node context in which prior work was
-developed: on a single server, atomicity is near-trivial to achieve
-via lightweight locking and/or local concurrency control over data
-items. (In this spirit, multi-item atomicity is common in so-called
-``entity groups,'' which logically represent groups of co-located
-items.) In contrast, in a distributed environment, atomicity over
-arbitrary groups of items is considerably more difficult to achieve
-with high availability. The key challenge is that different servers
-responsible for subsets of a transaction's updates must coordinate
-regarding when to reveal their writes to a transaction---without
-violating availability due to, say, locking, or a missing
-transactional write. Together with Cut Isolation, Transactional
-Atomicity prevents Read Skew anomalies (Berenson et al.'s
-A5A~\cite{ansicritique}).
+developed: on a single server (or a fully replicated database),
+atomicity is near-trivial to achieve via lightweight locking and/or
+local concurrency control over data items. (In this spirit, multi-item
+atomicity is common in so-called ``entity groups,'' which logically
+represent groups of co-located items~\cite{entitygroup}.) In contrast,
+in a distributed environment, atomicity over arbitrary groups of items
+is considerably more difficult to achieve with high availability. The
+key challenge is that different servers responsible for subsets of a
+transaction's updates must coordinate regarding when to reveal their
+writes to a transaction---without violating availability due to, say,
+locking, or a missing transactional write. Together with cut
+isolation, transactional atomicity prevents Read Skew anomalies
+(Berenson et al.'s A5A~\cite{ansicritique}).
 
 Transactional atomicity is achievable in a HAT environment. To ensure
-transactional atomicity, servers only need to know when all of a
-transaction's updates are present on their respective servers but do
-not need to know when all respective servers know when all of a
-transaction's updates are present on their respective servers. This
-subtle distinction corresponds to the difference between the
-distributed systems concepts of consensus and reliable broadcast. The
-former requires explicit agreement across servers, while the latter
-simply requires that, if one server reveals a write, then all servers
-reveal the write. We omit a full discussion of the algorithm and
-instead refer the reader to the extended version of this paper.
+transactional atomicity, servers need to know when all of a
+transaction's updates are present on their respective
+replicas. However, servers do \textit{not} need to agree exactly when
+all replicas are aware that all of a transaction's updates are present
+on their respective replicas. This subtle distinction corresponds to
+the difference between the distributed systems concepts of reliable
+broadcast and consensus. The former requires that, if one server
+reveals a write, then all servers reveal a write, while the latter
+requires explicit agreement across servers (and is unavailable). We
+omit a full discussion of the algorithm and instead refer the reader
+to the extended version of this paper~\cite{hat-tr}.
 
@@
 
-So far, our models have not considered guarantees across transactions,
-but, in the distributed systems context, many useful \textit{safety}
-guarantees span multiple operations. There is no guarantee of forward
-progression between transactions or any guarantees on the ordering
+So far, our models have not considered what happens across
+transactions: we have not guaranteed any ordering or continuity
 between transactions (other than that there exists some arbitrary
-ordering). \textit{Session guarantees} are used to describe continuity
-between groups of transactions issued by the same logical
-user. Informally, a \textit{session} is a form of context that
-consists of logical state that should persist between
-operations. Users accessing a web site may want to experience a notion
-of ``forward progress'' for the duration that they are logged in, so
-all transactions submitted between ``log in'' and ``log out'' might
-form a session. We can define sessions at several granularities (e.g.,
-per-transaction, per-client, per-hour) but, for simplicity, will refer
-to client-level guarantees in the discussion.
+ordering). However, in the distributed systems context, many useful
+\textit{safety} guarantees span multiple application actions.
+\textit{Session guarantees} are used to describe guarantees across
+groups of operations (here, transactions) issued by the same logical
+user~\cite{sessionguarantees}. Informally, a \textit{session} is an
+application context that consists of logical state that should persist
+between operations. For example, users accessing a web site may want
+to experience a notion of ``forward progress'' for the duration that
+they are logged in; all transactions submitted between ``log in'' and
+``log out'' operations might form a session. We can define sessions at
+several granularities (e.g., per-transaction, per-client, per-time unit)
+but, for simplicity, will refer to client-level guarantees in the
+discussion.
 
@@
 
-\vspace{.25em}\noindent\textit{\textbf{Monotonic reads}} stipulates
+\vspace{.5em}\noindent\textit{\textbf{Monotonic reads}} stipulates
 that a client's subsequent reads to a given object ``never return any
@@
 
-\vspace{.25em}\noindent\textit{\textbf{Monotonic writes}} requires
+\vspace{.5em}\noindent\textit{\textbf{Monotonic writes}} requires
 that transactions from each individual client be serialized. Any order
@@
 
-\vspace{.25em}\noindent\textit{\textbf{Writes Follow Reads}} requires
+\vspace{.5em}\noindent\textit{\textbf{Writes Follow Reads}} requires
 that, if a session reads a write originating from transaction $T_1$
@@
 the reads-from order (note that the total order on transactions
-precludes reads-from cycles).
+precludes reads-from cycles).\vspace{.5em}
 
@@
 
-\noindent\textit{\textbf{Read your writes}} requires
+\vspace{.5em}\noindent\textit{\textbf{Read your writes}} requires
 that whenever a client reads a given data item after updating it, the
@@
 
-\noindent\textit{\textbf{PRAM}} provides the illusion of serializing
-each session's operations and is the combination of monotonic reads,
-monotonic writes, and read your writes.
+\vspace{.5em}\noindent\textit{\textbf{PRAM}} (Pipelined Random Access
+Memory) provides the illusion of serializing each session's operations
+and is the combination of monotonic reads, monotonic writes, and read
+your writes~\cite{herlihy-art}.
 
-\noindent\textit{\textbf{Causal consistency}} is the combination of
+\vspace{.5em}\noindent\textit{\textbf{Causal consistency}} is the combination of
 all of the session guarantees~\cite{sessiontocausal} (alternatively,
 PRAM and writes-follow-reads) and is also referred to as PL-2L
-isolation~\cite{adya}).
+isolation~\cite{adya}).\vspace{.5em}
 
@@
 
-A HAT system can make \textbf{limited consistency} guarantees. It can
-execute commutative and logically monotonic~\cite{needed} operations
-without the risk of invalidating (also monotonic) application-level
-integrity constraints. Our goal in this paper is not to sketch the
-entire space of consistency models that are achievable (see
-Section~\ref{sec:futurework}, however we specifically evaluate TPC-C
-transaction semantics under HAT consistency guarantees in
-Section~\ref{sec:evaluation}.
+\noindent{\textbf{Consistency}} A HAT system can make limited
+consistency guarantees. It can execute commutative and logically
+monotonic~\cite{calm} operations without the risk of invalidating
+(also monotonic) application-level integrity constraints. Our goal in
+this paper is not to sketch the entire space of consistency models
+that are achievable (see Section~\ref{sec:futurework}, however we
+specifically evaluate TPC-C transaction semantics under HAT
+consistency guarantees in Section~\ref{sec:evaluation}.
 
-As we briefly discussed in Section~\ref{sec:availability}, a client
-requiring \textbf{durability} of surviving $F$ server faults requires
-at least ($F+1$)-availability.
+\vspace{.5em}\noindent{\textbf{Durability}} As we briefly discussed in
+Section~\ref{sec:availability}, a client requiring durability
+of surviving $F$ server faults requires at least ($F+1$)-availability.
 
-Finally, to require that servers propagate writes between replicas, we
-can require \textbf{convergence}, or eventual consistency for each
-data item: in the absence of new mutations to a data item, in the
-absence of partitions, all servers eventually agree on the value for
-each item. This is typically accomplished by any number of
-anti-entropy protocols, which periodically update neighboring servers
-with the latest value for each data item. This convergence property
-should be considered mandatory given the spirit of existing literature
-on isolation protocols: in all models we have encountered, the
-database should behave as if there is some total order over
+\vspace{.5em}\noindent{\textbf{Convergence}} Finally, to
+require that the system propagates writes between replicas, we can
+require convergence, or eventual consistency for each data item: in
+the absence of new mutations to a data item, in the absence of
+partitions, all servers eventually agree on the value for each
+item. This is typically accomplished by any number of anti-entropy
+protocols, which periodically update neighboring servers with the
+latest value for each data item. This convergence property should be
+considered mandatory given the spirit of existing literature on
+isolation protocols: in all models we have encountered, the database
+should behave as if there is some total order over
 transactions. Indeed, non-convergent systems are remarkably difficult
@@
 impossibility results for HATs, also defined in terms of previously
-identified isolation and consistency anomalies.
+identified isolation and consistency anomalies. Perhaps most notably,
+it is impossible to prevent Lost Update or Write Skew in a HAT system.
 
@@
 Skew---and therefore providing Snapshot Isolation, Repeatable Read,
-and Serializability---requires unavailability.
+and serializability---requires unavailability.
 
@@
 linearizability, which is, via Gilbert and Lynch's proof of the CAP
-Theorem, provably unachievable with high availabilty.
+Theorem, provably unachievable with high
+availabilty~\cite{gilbert-cap}.
 
@@
 items. As we have discussed, one of the most famous is
-linearizability, which states that reads will return the last
-completed write to a data item, and there are several other (weaker)
-variants such as safe and regular register semantics. When applied to
-transactional semantics, the combination of one-copy serializability
-and linearizability is called \textit{strong (or strict) one-copy
-  serializability} or \textit{externally consistent transactions}
-(e.g., Google's Spanner~\cite{spanner}). It is also common,
-particularly in systems that allow reading from masters and slaves, to
-provide a guarantee such as ``read a version that is no more than five
-seconds out of date'' or similar. Unfortunately, an indefinitely long
-partition can cause violation of any recency bound.
+linearizability~\cite{herlihy-art}, which states that reads will
+return the last completed write to a data item, and there are several
+other (weaker) variants such as safe and regular register
+semantics. When applied to transactional semantics, the combination of
+one-copy serializability and linearizability is called \textit{strong
+  (or strict) one-copy serializability} or \textit{externally
+  consistent transactions} (e.g., Google's
+Spanner~\cite{spanner})~\cite{adya}. It is also common, particularly
+in systems that allow reading from masters and slaves, to provide a
+guarantee such as ``read a version that is no more than five seconds
+out of date'' or similar. Unfortunately, an indefinitely long
+partition can cause violation of any recency bound, so recency bounds
+are unavailable to HAT systems.
 
@@
 writes and, for maximum semantic richness, can provide all of
-Transactional Atomicity, Predicate Cut Isolation, Monotonic Reads,
-Monotonic Writes, and Writes Follow Reads. An unavailable system might
+transactional atomicity, predicate cut isolation, monotonic reads,
+monotonic writes, and writes follow reads. An unavailable system might
 provide PRAM and Serializability, as advocated by Salem and
-Daudjee~\cite{needed}.
+Daudjee~\cite{daudjee-session}.
 
@@
 subsumes all other models, while considering the (large) power set of
-all compatible models (e.g., only considering client-centric session
-consistency guarantees, the diagram depicts 32 possible HAT
+all compatible models (e.g., the diagram depicts 144 possible highly available
 combinations) hints at the vast expanse of consistency models found in
@@
 
-\begin{table}
+\begin{table}[t!]
 \begin{tabular}{| c | p{6cm} | }\hline
@@
 
-\begin{figure}
+\begin{figure}[t!]
 \centering
@@
 
-\noindent\textbf{Model Composition} may be non-trivial. Consider the following transactions:
+\vspace{.5em}\noindent\textbf{Model Composition} Choosing between
+combinations of compatible guarantees requires care. Consider the
+following transactions:
 \begin{align*}
@@
 If we want to guarantee both cut isolation and transactional atomicity
-and these are the only transactions that are executed by the system,
-then $T_3$ needs to read $a=b=\bot$, $a=b=1$, or $a=b=2$. This means
-that either the implementation should frequently return null
-(definitely undesirable and possibly non-convergent), keep multiple
-versions of each data item (necessitating potentially complicated
-distributed garbage collection), or use pre-declared read sets to
-fetch a consistent cut of keys before the transaction begins to
-execute. Using client-side caching can alleviate some of these
-challenges, but then the system becomes sticky high available.
+and the system only executes $T_1$, $T_2$, and $T_3$, then $T_3$ needs
+to read $a=b=\bot$, $a=b=1$, or $a=b=2$. This means that either the
+implementation should frequently return $\bot$ (definitely undesirable
+and possibly non-convergent), keep multiple versions of each data item
+(necessitating potentially complicated distributed garbage
+collection), or use pre-declared read sets to fetch a consistent cut
+of keys before each transaction begins to execute. Using client-side
+caching can alleviate some of these challenges, but then the system
+becomes sticky high available.
 
@@
 processes, standard vector-based approaches face an upper bound of
-$O(N)$~\cite{charron-bost} storage per write. This means that, with
-$100,000$ clients, each write might be accompanied by $100,000$
+$O(N)$ storage per write~\cite{charron-bost}. This means that, with
+$100K$ clients, each write might be accompanied by $100K$
 timestamps per vector. This is difficult to scale. By compromising on
@@
 
-\noindent The relationship between \textbf{Transactional Atomicity and
-  Linearizability} is non-obvious. One might attempt to find a mapping
-between the two: Transactional Atomicity dictates that writes to
-multiple keys across multiple servers are made visible to readers all
-at once, while Linearizability dictates that writes to a single key on
-multiple servers are made visible to all readers at once. The
+\vspace{.5em}\noindent\textbf{Linearizability and Transactional
+  Atomicity} The relationship between linearizability and
+transactional atomicity is non-obvious. One might attempt to find a
+mapping between the two: transactional atomicity dictates that writes
+to multiple keys across multiple servers are made visible to readers
+all at once, while linearizability dictates that writes to a single
+key on multiple servers are made visible to all readers at once. The
 difference is two-old. First, in linearizable (and safe and regular)
 systems, writes are made visible to all clients \textit{immediately}
-after they finish. With Transactional Atomicity, there is no recency
-requirement. Second, in linearizabile systems, all clients see all
-writes at the same time. With Transactional Atomicity, clients see
+after they finish. With transactional atomicity, there is no recency
+guarantee. Second, in linearizabile systems, all clients see all
+writes at the same time. With transactional atomicity, clients see
 writes at different times depending on which replicas they contact. We
 are not aware of an analogous model in the distributed systems
-literature. Accordingly, despite apparent similarities, Transactional
-Atomicity is much weaker than Linearizability.
+literature. Accordingly, despite apparent similarities, transactional
+atomicity is much weaker than linearizability.
 
-\noindent There is a substantial \textbf{visibility benefit to sticky
-  availability}: writes can safely become visible to readers much more
-quickly in a sticky available system. In the model we discussed, it is
-possible to achieve several properties like monotonic reads in a
-highly available system by waiting until all servers have seen a
-write. However, this incurs severe visibility penalties---new writes
-will not become visible to clients in the presence of partitions. A
-client that does not want to guarantee read-your-writes (due to the
-sticky availability requirement) may still wish to read other clients'
-writes with timeliness. Accordingly, a system may forgo high
-availability in order to provide better visibility. Alternatively, a
-system may provide stickiness in the absence of partitions, then
-``lose'' session guarantees by having clients reconnect to an
-available replica. This latter strategy provides session guarantees in
-the absence of partitions but, to provide availability, sacrifices
-them in the event of partitions.
+\vspace{.5em}\noindent\textbf{Visibility and Stickiness} Sticky
+availability can result in much better write \textit{visibility}:
+clients will be able to safely read writes more quickly in a sticky
+available system. In the model we discussed, it is possible to achieve
+several properties like monotonic reads in a highly available system
+by waiting until all servers have seen a write. However, this incurs
+severe visibility penalties---new writes will not become visible to
+clients in the presence of partitions. A client that does not want to
+guarantee read-your-writes (due to the sticky availability
+requirement) may still wish to read other clients' writes with
+timeliness. Accordingly, a system may forgo high availability in order
+to provide better visibility. We also note that a system may wish to
+provide stickiness in the absence of partitions, then, in the presence
+of partitions, ``lose'' session guarantees by having clients reconnect
+to an available replica. This latter strategy provides session
+guarantees in the absence of partitions but, to provide availability,
+sacrifices them in the event of partitions.

### 2013-03-26 writeup/intro.tex
Another pass at writing

--- 2013-03-17 writeup/intro.tex
@@
 
-Many recent systems have eschewed ACID guarantees.
+Requirements for high availability, predictable low latency, and
+simplicity of coordination protocols have led many recent distributed
+database designers to renounce several guarantees provided by
+traditional database systems. In particular, many modern systems forgo
+providing ACID guarantees---which often provide single-system image
+(SSI) programmability but suffer from unavailability and often
+expensive coordination costs. The CAP Theorem, demonstrating the
+incompatibility of ``always on'' operation with guarantees on data
+item recency in the presence of unreliable networks, has led a large
+number of systems to prioritize high availability and guaranteed low
+latency, which are especially important in geo-replicated contexts. We
+observe a dearth of systems providing general-purpose multi-operation
+\textit{transactions} over multiple data items with useful semantics.
 
-CAP Theorem says you can't have consistency and availability and also talks about partitions.
+The relationship of the CAP Theorem and ACID semantics is poorly
+understood. While serializability---the gold standard of SSI
+transactions---is unavailable, databases have a long tradition of
+providing weaker isolation and consistency guarantees. While these
+weaker guarantees do not provide SSI for arbitrary general-purpose
+read/write transactions, they often admit greater concurrency and
+higher performance than comparable SSI algorithms. Accordingly, weak
+isolation is widely employed (Section~\ref{needed}), overwhelmingly as
+the default setting in today's traditional ``ACID'' and up-and-coming
+``NewSQL'' databases and often as the maximum guarantees offered
+(Section~\ref{sec:modernacid}). While the prevalence of these
+configurations is encouraging, the current literature does not provide
+guidance as to the operation of weakly consistent isolation guarantees
+in a distributed context. Rather, we have a strong understanding of
+weak isolation in the single-server context from which it originated
+as well as several techniques for providing distributed
+serializability but we have little to no insight into which ACID
+guarantees are compatible with Highly Available Transactions (HATs),
+or can be provided with high availability and low latency.
 
-What does CAP have to say about ACID? Recent workshop result on Highly Available Transactions briefly demonstrated that some ACID guarantees are indeed achievable in a highly available manner.
+This work attempts to bridge the gap between ACID properties and
+highly available systems. We analyze a range of existing ACID and
+distributed consistency models to provide a rigorous taxonomization of
+which models are achievable in a HAT system.  While several
+guarantees, like Serializability and Snapshot Isolation, are
+unachievable due to reliance on preventing Lost Update and Write Skew
+phenomena, many other ACID guarantees are HAT compliant, such as
+providing Read Committed isolation, preventing Fuzzy Read phenomena,
+and ensuring atomicity of transactional updates. We also attempt to
+bridge the often distant worlds of distributed system consistency and
+ACID database consistency by analyzing the relationship between
+so-called session guarantees, ``register'' semantics, and ACID
+properties. In doing so, we introduce a hierarchy of highly available
+properties as, for example, some guarantees require ``stickiness,'' or
+fate-sharing between clients and database servers. We believe that
+this analysis is the first that has been performed over such a broad
+range of ACID and distributed consistency models.
 
-Many ACID systems were designed either i.) for single systems or ii.)
-to provide serializability. The combination of weak isolation and
-distributed systems necessitates further study and
-design. \textit{What is the benefit of high availability and what is
-  the cost?}
-
-Highly Available Transactions are the class of multi-object operations
-and their respective semantics that can be provided with high
-availability.
-
-Weak isolation has a cost: in the words of one database pioneer, you
-``fall off a cliff'' when it comes to reasoning about your
-program. However, it also has a benefit: availability and latency.
-
-Our goal in this paper is to objectively map the space of trade-offs and
-
- This paper fills in the remaining gaps.
-
-What does high availability even mean in a transaction environment?
-
-What models are achievable, and which aren't?
+While this model classification is useful, we also seek to understand
+the implications for end-users of HAT systems. HATs provide high
+availability and low latency but cannot offer all of the semantic
+guarantees of unavailable systems---when is this trade-off worthwhile?
+We attempt to objectively study these virtues and limitations of both
+classic and HAT systems by performing an extensive survey of
+practitioner accounts and research literature, performing experimental
+analyses on modern cloud infrastructure, and analyzing representative
+applications for their semantic requirements. While our results are
+not definitive for all applications, they suggest that HATs offer a
+one to three order of magnitude latency decrease compared to
+traditional protocols and provide acceptable semantics for a wide
+range of programs, particularly those with monotonic logic and
+commutative updates. HAT systems can also provide arbitrary foreign
+key constraints on multi-item updates and sometimes provide uniqeness
+guarantees. However, particularly for programs with non-monotonic
+logic, HATs may fall short, requiring more coordination-intensive
+mechanisms. Concisely, we attempt to understand \textit{what benefits
+  do Highly Available Transactions offer and what is their cost?}
 
@@
 \begin{itemize}
-\item We present a model for high availability in a transactional environment, including Replica-agnostic High Availability (R-HA) and Sticky High Availability (S-HA).
-\item We analyze the existing literature on ACID properties (particularly, ANSI SQL Spec, Adya, Gray) to show which properties* with a focus on isolation* are achievable in HATS.
-Achievable properties include transactional atomicity, variants of repeatable read isolation, and session guarantees.
-We also describe unavailable properties, like preventing Lost Update and Read and Write Skew.
-We believe that this is the first such characterization in the literature.
-\item We further motivate the study of HATS via the analysis of existing concurrency control algorithms in a highly available environment as well as through deployment of an experimental database prototype on public cloud infrastructure.
+\item We present a model for high availability in a transactional
+  environment, replica-based High Availability, Sticky High
+  Availability, and Transactional High Availability.
+
+\item We analyze the existing literature on ACID properties and
+  replicated data semantics to show which properties are achievable in
+  HAT systems. Achievable properties include transactional atomicity,
+  variants of repeatable read isolation, and session guarantees. We
+  also describe unavailable properties, like preventing Lost Update
+  and Write Skew. We believe that this is the first such
+  characterization in the literature.
+
+\item We further motivate the study of HATs via the analysis of
+  existing concurrency control algorithms, a case-study of running a
+  transactional application with HATs, and via deployment of a HAT
+  database prototype on public cloud infrastructure.
 \end{itemize}

### 2013-03-26 writeup/model.tex
Another pass at writing

@@
 To understand which guarantees can be provided with high availabilty,
-we must first understand what properties define high availability. In
-this section, we will formulate a model that captures a range of
-availability models, including ``true'' high availability,
-availablility with stickiness, and several other useful variants.
+we must first define what high availability means. In this section, we
+will formulate a model that captures a range of availability models,
+including ``true'' high availability, availablility with stickiness,
+and several other useful variants.
 
 Informally, highly available algorithms ensure ``always on'' operation
-and promise a guaranteed response. If users of a highly available
-system are able to contact a (set of) server(s) in a system, they are
-guaranteed a response as the (set of) server(s) does not need to
-communicate with any others. This lack of fast-path coordination also
-means that a highly available system can also offer low latency
-operation~\cite{abadi-pacelc}. To describe whether a transactional
-system is highly available, we need both a way to describe what
-servers a client must contact as well as what kinds of responses a
-server can provide, especially given the possibility of aborts.
+and guaranteed low latency. If users of a highly available system are
+able to contact a (set of) server(s) in a system, they are guaranteed
+a response as the (set of) server(s) does not need to communicate with
+any others. If servers are partitioned from one another, they do not
+need to stall in order to provide a ``safe'' query response. This lack
+of fast-path coordination also means that a highly available system
+also provides low latency~\cite{abadi-pacelc}. However, to properly
+describe whether a \textit{transactional} system is highly available,
+we need both a way to describe what servers a client must contact as
+well as what kinds of responses a server can provide, especially given
+the possibility of aborts.
 
@@
 even in the presence of arbitrary, indefinitely long network
-partitions between servers~\cite{gilbert-cap}. This is a useful
-definition: if we can contact at least one server, then we can receive
-a response. However, there are several other cases we should consider.
+partitions between servers~\cite{gilbert-cap}. As in a standard
+distributed database, designated servers might handle operations on
+different data items;\footnote{There is a further distinction between a
+  fully replicated system, in which all servers are replicas for all
+  data items, and a partially replicated system, in which at least one
+  server acts as a replica for only a subset of all data items. For
+  generality, and, for applicability to many modern ``sharded'' or
+  ``partitioned'' data storage systems like Dynamo, we consider
+  partially replicated systems.} a server that can handle an
+operation for a given data item is called a \textit{replica} for that
+item. This is a useful definition: if a client can contact at least
+one replica for an item on which it wishes to perform an operation,
+then the replica will provide a response to the client. However, there
+are several other cases we should consider.
 
 First, clients may wish to contact multiple servers. For example, to
-ensure that the effects of its operations will persist in the face of
+ensure that the effects of operations will persist in the face of
 server failures, clients may wish to wait until operations reach more
-than one server before they considers operations completed. However,
+than one server before they consider operations completed. However,
 according to our definition of high availability, server-fault
@@
 providing high availability ($1$-availability), a system with
-$2$-availability and will require more messaging, possibly higher
-latency and more unsuccessful operations in the presence of network
-and server failures.  Additionally, as is commmon in standard
-consensus algorithms like Paxos, clients may want to contact a
-majority of $N$ servers in a system. This \textit{majority
+$2$-availability will require more messaging and possibly higher
+latency and may experience more unsuccessful operations in the
+presence of network and server failures.  Additionally, as is commmon
+in standard consensus algorithms like Paxos, clients may want to
+contact a majority of $N$ servers in a system. This \textit{majority
   availability} (a special case of $K$-availability with $K=\lceil
-\frac{N}{2} \rceil$) is more available than $N$-availability.
+\frac{N}{2} \rceil$) is more available than full $N$-availability.
 
@@
 maintaining affinity or ``stickiness'' with a server or set of
-servers~\cite{vogels-defs}. Simultaneously aintaining availability and
-stickiness requires fate-sharing between the client and its set of
+servers~\cite{vogels-defs}. Simultaneously maintaining availability
+and stickiness requires fate-sharing between the client and its set of
 servers: if a client's sticky servers fail, or, equivalently, if the
@@
   availability} if a client's subsequent operations are executed
-against a server that has observed all of its prior operations, then
-it eventually receives a response, even in the presence of
-indefinitely long partitions. As above, it may be useful for a client
-to achieve stickiness \textit{and} contact many replicas (i.e., for
-fault tolerance), so we generalize sticky availability to
+against a server that has observed all effects of its prior
+operations, then it eventually receives a response, even in the
+presence of indefinitely long partitions. As above, it may be useful
+for a client to achieve stickiness \textit{and} contact many replicas
+(i.e., for fault tolerance), so we generalize sticky availability to
 \textit{$K$-sticky availability}, which provides sticky availability
@@
 is a special case of sticky availability and is less available than
-the equivalent sticky availability.
+the equivalent non-mastered sticky availability.
 
@@
 \end{tikzpicture}
+\caption{Hierarchy of replica availability levels for $N>3$ servers.}
 \label{fig:availability-order}
-\caption{Hierarchy of replica availability levels for $N>3$ servers.}
 \end{figure}
 
-We show the hierarchy of replica availabilty levels in
+We show a hierarchy of replica availability levels in
 Figure~\ref{fig:availability-order}. $K$-sticky availability subsumes
@@
 $N$-availability, while majority availability subsumes $1$-sticky
-availability More generally, $\lceil \frac{N}{2} \rceil$$+$$K$$-$$1$
+availability. More generally, $\lceil \frac{N}{2} \rceil$$+$$K$$-$$1$
 availability subsumes $K$-sticky availability. We have omitted
@@
 
-Thus far, we have focused on single-operation availability. This is
-standard in distributed systems literature (e.g., atomic, regular, and
-safe register models all concern single objects~\cite{herlihy-art}),
-yet the database literature largely focuses on transactions: groups of
-multiple operations over multiple objects. Accordingly, by itself,
-replica availability is insufficient to describe availability
-guarantees for transactions. Additionally, given the choice of
-\textit{commit} and \textit{abort} responses---which signal
-transaction success or failure to a client---we must be careful in
-defining availability.
+So far, we have focused on single-object, single-operation
+availability. This is standard in distributed systems literature
+(e.g., atomic, regular, and safe register models all concern single
+objects~\cite{herlihy-art}), yet the database literature largely
+focuses on transactions: groups of multiple operations over multiple
+objects. Accordingly, by itself, replica availability is insufficient
+to describe availability guarantees for transactions. Additionally,
+given the choice of \textit{commit} and \textit{abort}
+responses---which signal transaction success or failure to a
+client---we must be careful in how we define availability.
 
@@
 on multiple data items, then the client must be able to contact and
-receive a response from at least one server responsible for each data
-item. This may result in lower availability than a non-transactional
+receive a response from at least one replica for each data item. This
+may result in lower availability than a non-transactional
 requirement. Additionally, given the possiblity of aborts, we need to
@@
 given replica availability for every data item in a transaction, the
-transaction eventually commits or internally aborts. For example,
+transaction eventually commits or internally aborts~\cite{hat-hotos}. For example,
 system will violate transactional $2$-availability if client that can

### 2013-03-26 writeup/motivation.tex
Another pass at writing

@@
 is zero''~\cite{fallacies-deutsch}. On a single server, communication
-systems are robust and relatively fast, and most single-node systems
-do not need to prepare for partial component failure. However, these
-assumptions are no longer valid in a distributed setting. To extend
-Jim Gray's storage latency analogy, if memory is 1.5 hours from the
-CPU, disk is 2 years away, especially across wide-area networks,
-remote server accesses can exceed 30 years, with the added concern
-that messages can be lost in transit or delayed indefinitely by
-network partitions~\cite{gray-rules}. Accordingly, ignoring network
-failures and latency ``cause[s] big trouble and painful learning
-experiences''~\cite{fallacies-deutsch}.
+channels are robust and relatively fast, and few single-server systems
+need to account for a possibility of partial system failure. However,
+both of these assumptions are no longer valid in a distributed
+setting. To extend Jim Gray's storage latency analogy, if memory is
+1.5 hours from the CPU and disk is 2 years away, then, especially
+across wide-area networks (WANs), remote server accesses can exceed 30
+years, with the added concern that messages can be lost in transit or
+delayed indefinitely by network
+partitions~\cite{gray-rules}. Accordingly, ignoring network failures
+and latency can ``cause big trouble and painful learning
+experiences''~\cite{fallacies-deutsch}. As an example of an actual system rearchitecture,
+less than one year after its announcement, Yahoo!'s PNUTS developers
+explicitly added a support for weaker, highly available operation,
+explaining that ``strict adherence [to strong consistency] leads to
+difficult situations under network partitioning or server
+failures...in many circumstances, applications need a relaxed
+approach''~\cite{pnuts-update}.
 
-In this section, we attempt to quantify the impact of network
-partitions and latency in real-world deployments. While the importance
-of network behavior is a topic of considerable speculation within the
+In this section, we quantify the impact of network partitions and
+latency in several real-world deployments. While the importance of
+network behavior is a topic of considerable speculation within the
 database community~\cite{stonebraker2010errors}, there is mounting
 evidence---both anecdotal and experimental---that partitions occur and
-latency is non-negligible.
+network latencies are non-negligible.
 
-\subsection{Partitions}
+\subsection{Partitions in the Real World}
 
@@
 either lose availability of operations or its ability to reason about
-global state.
+global state. In this section, we present evidence for the frequency
+and non-triviality of WAN and LAN partitions.
 
 According to James Hamilton, Vice President and Distinguished Engineer
-on the Amazon Web Service team, ``network partitions should be rare
+on the Amazon Web Services team, ``network partitions should be rare
 but net gear continues to cause more issues than it
@@
 services~\cite{amazon-netpartition}. Subsequent misconfigurations and
-partial failures such as another EC2 failure in October 2012 have led
-to full site outages for web services like Reddit, Foursquare, and
-Heroku~\cite{ec2-downsites}. At a global scale, hardware
-failures---like the 2011 outages in Internet backbones in North
-America and Europe due a router bug~\cite{juniper-partition}---and
-misconfigurations---such as BGP faults in 2008~\cite{pakistan-youtube}
-and 2010~\cite{research-experiment-partition}---can cause widespread
+partial failures such as another EC2 outage in October 2012 have led
+to full site disruptions for popular web services like Reddit,
+Foursquare, and Heroku~\cite{ec2-downsites}. At a global scale,
+hardware failures---like the 2011 outages in Internet backbones in
+North America and Europe due a router
+bug~\cite{juniper-partition}---and misconfigurations---such as BGP
+faults in 2008~\cite{pakistan-youtube} and
+2010~\cite{research-experiment-partition}---can cause widespread
 partitions. Many of our discussions with practitioners---especially
@@
 from large-scale operators like Google~\cite{dean-keynote} confirm
-that partitions are a reality.
+that partitions are a reality for service operators today.
 
-Several recent studies rigorously quantify partition behavior. A 2011
-study of several Microsoft datacenters found a mean of 40.8 network
-link failures per day (95th percentile: 136), with a median time to
-repair of around five minutes, and up to one week. Perhaps
-surprisingly, network redundancy only reduces the median impact of
-failures by up to 40\%, meaning providers cannot easily reduce the
-incidence of partitions~\cite{sigcomm-dc}. A 2010 study of over 200
+Several recent studies quantify partition behavior more rigorously. A
+2011 study of several Microsoft datacenters found a mean of 40.8
+network link failures per day (95th percentile: 136), with a median
+time to repair of around five minutes (and up to one week). Perhaps
+surprisingly, provisioning redundant networks only reduces impact of
+failures by up to 40\%, meaning network providers cannot easily
+curtail partition behavior~\cite{sigcomm-dc}. A 2010 study of over 200
 wide-area routers found an average of 16.2--302.0 failures per link
@@
 highest priority tickets ranges from 114--188 minutes and up to a full
-day for all tickets.~\cite{turner2012failure}. The incidence of
-network failures is further confirmed by several prior studies showing
-for example, on Sprint's WANs, a median time to repair between 2 and
-1000 seconds, and a median time between failures of approximately 3000
-seconds.~\cite{ip-backbone-failures} and frequent path routing
-failures on the Internet~\cite{labovitz-failures}. Indeed, isolating,
-quantifying, and designing for network failures is an area of active
+day for all tickets~\cite{turner2012failure}. The incidence of network
+failures is further confirmed by several prior studies showing for
+example, on Sprint's WANs, a median time to repair between 2 and 1000
+seconds, and a median time between failures of approximately 3000
+seconds~\cite{ip-backbone-failures}, as well as frequent path routing
+failures on the Internet~\cite{labovitz-failures}. Isolating,
+quantifying, and designing for these network failures is an area of active
 research in networking community~\cite{surviving-failures-bodik,
@@
 
-\subsection{Latency}
+\subsection{Latency: An Omnipresent Concern}
 
 Even with fault-free networks, distributed systems face the challenge
-of communication latency, Deutsch's second ``Fallacy.'' Fundamentally,
-the speed at which two servers can communicate is (according to modern
-physics) bounded by the speed of light. Two servers located on
-opposite sides of the Earth face, in the best case---communicating via
-a hypothetical link crossing through the center of the
-planet---require a minimum 85.1 ms RTT (133.7 ms if sent at ground
-level). As services are replicated to multiple, geographically
-distinct sites, the cost of communication increases this upper bound.
+of communication latency, Deutsch's second ``Fallacy.'' In this
+section, we quantify expected latencies, which are often
+huge---hundreds of milliseconds---in a geo-replicated context.
 
-In real deployments, latencies are higher than the speed of light due
-to routing, congestion, and processing times; if not, servers a
-kilometer apart would enjoy a 6.7 $\mu$s RTT. To illustrate the
-difference between intra-datacenter, inter-datacenter, and
-inter-planetary networks, we performed a measurement study of network
-behavior on Amazon's EC2, a widely-used public compute cloud. We
-measured one week of ping times between all seven EC2 regions, across
-three ``availability zones'', or closely co-located datacenters, in
-the Eastern United States, and within a single ``availabilty zone,''
-at a granularity of 1s (dataset to be released).
+Fundamentally, the speed at which two servers can communicate is
+(according to modern physics) bounded by the speed of light. Two
+servers located on opposite sides of the Earth face, in the best
+case---communicating via a hypothetical link crossing through the
+center of the planet---require a minimum 85.1 ms RTT (133.7 ms if sent
+at ground level). As services are replicated to multiple,
+geographically distinct sites, the cost of communication increases to
+this upper bound.
+
+In real deployments, latencies are actually higher than the speed of
+light dictates due to routing, congestion, and processing times; if
+not, servers a kilometer apart would enjoy a 6.7 $\mu$s RTT. To
+illustrate the difference between intra-datacenter, inter-datacenter,
+and inter-planetary networks, we performed a measurement study of
+network behavior on Amazon's EC2, a widely-used public compute
+cloud. We measured one week of ping times between all seven EC2
+regions, across three ``availability zones'', or closely co-located
+datacenters, and within a single ``availabilty zone,'' at a
+granularity of 1s (dataset to be released).
 
@@
 is between 1.82 and 6.38 times faster than across geographically
-co-located datacenters (1b) and 40 and 647 times faster than across
-geographically distributed datacenters (1a). The cost of wide-area
-communication exceeds the speed of light: for example, communicating
-from S\~{a}o Paulo to Singapore is lower-bounded by speed-of-light
-transit RTT of 106.7 ms but, on average, incurs a 362.8 ms RTT (95th
-percentile: 649ms). As shown in Figure~\ref{fig:rtt}, the distribution
-of latencies varies between links, but the trend---in line with other
-recent studies~\cite{mdcc, redblue}---is clear: remote communication
-has a substantial cost.
+co-located datacenters (1b) and between 40 and 647 times faster than
+across geographically distributed datacenters (1a). The cost of
+wide-area communication exceeds the speed of light: for example,
+communicating from S\~{a}o Paulo to Singapore is lower-bounded by
+speed-of-light transit RTT of 106.7 ms but, on average, incurs a 362.8
+ms RTT (95th percentile: 649ms). As shown in Figure~\ref{fig:rtt}, the
+distribution of latencies varies between links, but the trend---in
+line with other recent studies~\cite{mdcc, redblue}---is clear: remote
+communication has a substantial cost.
 
@@
 
-\subsection{Modern Systems}
+\subsection{CAP, Isolation, and Modern Systems}
 \label{sec:modernacid}
@@
 Mitigating the effects of partitions and network latency is a
-difficult task, requiring system desigers to make hard trade-offs. In
-this section, we briefly discuss known semantic limits for systems
-wishing to maintain high availability and their impact on modern
-database systems. While the gold standard of ACID databases is
+difficult task, requiring systems designers to make fundamental
+trade-offs. In this section, we briefly discuss known semantic limits
+for systems wishing to maintain high availability and their impact on
+modern database systems. While the gold standard of ACID databases is
 unavailable, we show that many database systems operate under much
@@
 
-Most notable among existing impossibility results for distributed
-systems is Brewer's CAP Theorem, which, as formally proven by Gilbert
-and Lynch, states that it is impossible to maintain linearizability
-(``C,'' or ``consistency''---albeit poorly named), or the ability to
-read the last completed write, and replica availability (``A'') in the
-presence of network partitions (``P'')~\cite{gilbert-cap}. Brewer's
-Theorem has roots in decades of distributed database
-designs~\cite{davidson-survey} and has had a substantial impact on
-recent large-scale distributed systems design. Yet, while CAP is
-narrowly scoped to the impossibility linearizability, it is often
+One of the most popularly celebrated impossibility results for
+distributed systems is Brewer's CAP Theorem, which, as formally proven
+by Gilbert and Lynch, states that it is impossible to maintain
+linearizability (``C,'' or ``consistency''---albeit poorly named), or
+the ability to read the last completed write, and replica availability
+(``A'') in the presence of network partitions
+(``P'')~\cite{gilbert-cap}. Brewer's Theorem has roots in decades of
+distributed database designs~\cite{davidson-survey} and has had a
+substantial impact on recent large-scale distributed systems design:
+while CAP is narrowly scoped to the impossibility of building highly
+available linearizable systems linearizability, it is often
 interpreted to preclude a larger set of guarantees, including ACID, or
 transactional atomicity, consistency, isolation and durability
-guarantees.
+guarantees, as evidenced by the lack of highly available systems
+providing them.  As an example of an production database redesign
+explicitly attributable to partition behavior, less than one year
+after its announcement, Yahoo!'s PNUTS system developers explicitly
+added a support for weaker, highly available operation, explaining
+that ``strict adherence [to strong consistency] leads to difficult
+situations under network partitioning or server failures...in many
+circumstances, applications need a relaxed
+approach''~\cite{pnuts-update}.
 
@@
 system~\cite{adya}. None of these weak isolation models guarantees
-serializability, but the benefits of these models can outweigh the
-costs of possible application-level consistency anomalies that arise
+serializability, but the benefits of these models can often outweigh
+costs of possible consistency anomalies that might arise
 from their use.
 
-To understand the prevalence of these weak isolation models, in an
-earlier workshop paper~\cite{hat-hotos}, we surveyed the default and
-maximum isolation models provided by 18 relational databases, often
-claiming to provide ``ACID'' or ``NewSQL'' functionality. As shown in
+To understand the prevalence of these weak isolation models, we
+recently surveyed, we surveyed the default and maximum isolation
+models provided by 18 relational databases, often claiming to provide
+``ACID'' or ``NewSQL'' functionality~\cite{hat-hotos}. As shown in
 Table~\ref{table:existing}, only three out of 18 databases provides
@@
 serializability as an option at all. This is particularly surprising
-when we consider the widespread deployment of many of these databases,
-like Oracle 11g, which are known to power major businesses and product
-functionality. Given that these transactional models are frequently
-used, our inability to provide serializability in arbitrary HATs is
-non-fatal to practical applications. If application writers and
-database vendors have already decided that the benefits of weak
-isolation outweigh potential application inconsistencies, then, as we
-will see, in a highly available environment, similar decisions may be
+when we consider the widespread deployment of many of these
+non-serializable databases, like Oracle 11g, which are known to power
+major businesses and product functionality. Given that these
+transactional models are frequently used, our inability to provide
+serializability in arbitrary HATs is non-fatal for practical
+applications. If application writers and database vendors have already
+decided that the benefits of weak isolation outweigh potential
+application inconsistencies, then, in a highly available environment
+that prevents serializability entirely, similar decisions may be
 appropriate.
@@
 Database & Default & Maximum\\\hline
-Actian Ingres 10.0/10S~\cite{actian-docs} & S & S\\
-Aerospike~\cite{aerospike-docs} & RC & RC\\
-Akiban Persistit~\cite{akiban-docs} & SI & SI\\
-Clustrix CLX 4100~\cite{clustrix-docs} & RR & ?\\
-Greenplum 4.1~\cite{greenplum-docs} & RC & S \\
-IBM DB2 10 for z/OS~\cite{db2-docs} & CS & S\\
-IBM Informix 11.50~\cite{informix-docs} & Depends & RR\\
-MySQL 5.6~\cite{innodb-docs} & RR & S \\
-MemSQL 1b~\cite{memsql-docs} & RC & RC\\
-MS SQL Server 2012~\cite{ms-sql-docs} & RC & S \\
-NuoDB~\cite{nuodb-docs} & CR & CR\\
-Oracle 11g~\cite{oracle-docs} & RC & SI\\
-Oracle Berkeley DB~\cite{bdb-reg-docs} & S & S\\
-Oracle Berkeley DB JE~\cite{bdb-je-docs} & RR & S\\
-Postgres 9.2.2~\cite{postgres-docs} & RC & S\\
-SAP HANA~\cite{hana-docs} & RC & SI\\
-ScaleDB 1.02~\cite{scaledb-docs} & RC & RC\\
-VoltDB~\cite{voltdb-docs} & S & S\\
+Actian Ingres 10.0/10S & S & S\\
+Aerospike & RC & RC\\
+Akiban Persistit & SI & SI\\
+Clustrix CLX 4100 & RR & ?\\
+Greenplum 4.1 & RC & S \\
+IBM DB2 10 for z/OS & CS & S\\
+IBM Informix 11.50 & Depends & RR\\
+MySQL 5.6 & RR & S \\
+MemSQL 1b & RC & RC\\
+MS SQL Server 2012 & RC & S \\
+NuoDB & CR & CR\\
+Oracle 11g & RC & SI\\
+Oracle Berkeley DB & S & S\\
+Oracle Berkeley DB JE & RR & S\\
+Postgres 9.2.2 & RC & S\\
+SAP HANA & RC & SI\\
+ScaleDB 1.02 & RC & RC\\
+VoltDB & S & S\\
 \hline
@@
   databases as of January 2013 (from
-  \protect\cite{needed-hotos}).}\vspace{-1.5em}
+  \protect\cite{hat-hotos}).}\vspace{-1.5em}
 \label{table:existing}
@@
 guarantees can be provided with high availability. The existing
-algorithms for providing weak isolation are typically presented for a
-single-node context and are therefore unavailable. We are not aware of
-any prior literature that provides guidance as to the relationship
-between weak isolation and high availability. Prior work has examined
-the relationship between serializability and high
+algorithms for providing weak isolation are typically designed for a
+single-node context and are, to the best of our knowledge, unavailable
+due to a reliance on concurrency control mechanisms like locking that
+are not resilient to partial failure. Morever, we are not aware of any
+prior literature that provides guidance as to the relationship between
+weak isolation and high availability: prior work has examined the
+relationship between serializability and high
 availability~\cite{davidson-survey} and weak isolation on a

### 2013-03-26 writeup/intro.tex
intro tweaks

@@
 higher performance than comparable SSI algorithms. Accordingly, weak
-isolation is widely employed (Section~\ref{needed}), overwhelmingly as
-the default setting in today's traditional ``ACID'' and up-and-coming
-``NewSQL'' databases and often as the maximum guarantees offered
+isolation is widely employed, overwhelmingly as the default setting in
+today's traditional ``ACID'' and up-and-coming ``NewSQL'' databases
+and often as the maximum guarantees offered
 (Section~\ref{sec:modernacid}). While the prevalence of these
@@
 in a distributed context. Rather, we have a strong understanding of
-weak isolation in the single-server context from which it originated
-as well as several techniques for providing distributed
-serializability but we have little to no insight into which ACID
-guarantees are compatible with Highly Available Transactions (HATs),
-or can be provided with high availability and low latency.
+weak isolation in the single-server context from which it
+originated~\cite{adya, ansicritique, gray-isolation} as well as
+several techniques for providing distributed
+serializability~\cite{bernstein-concurrency} but we have little to no
+insight into which ACID guarantees are compatible with Highly
+Available Transactions (HATs), or can be provided with high
+availability and low latency.
 
@@
 which models are achievable in a HAT system.  While several
-guarantees, like Serializability and Snapshot Isolation, are
+guarantees, like serializability and Snapshot Isolation, are
 unachievable due to reliance on preventing Lost Update and Write Skew
@@
 ACID database consistency by analyzing the relationship between
-so-called session guarantees, ``register'' semantics, and ACID
-properties. In doing so, we introduce a hierarchy of highly available
-properties as, for example, some guarantees require ``stickiness,'' or
-fate-sharing between clients and database servers. We believe that
-this analysis is the first that has been performed over such a broad
-range of ACID and distributed consistency models.
+so-called session guarantees~\cite{sessionguarantees}, ``register''
+semantics~\cite{herlihy-art}, and ACID properties. In doing so, we
+introduce a hierarchy of highly available properties as, for example,
+some guarantees require ``stickiness,'' or fate-sharing between
+clients and database servers. We believe that this analysis is the
+first that has been performed over such a broad range of ACID and
+distributed consistency models.
 

### 2013-03-26 writeup/intro.tex
blooml and stuff

@@
 range of programs, particularly those with monotonic logic and
-commutative updates. HAT systems can also provide arbitrary foreign
-key constraints on multi-item updates and sometimes provide uniqeness
-guarantees. However, particularly for programs with non-monotonic
-logic, HATs may fall short, requiring more coordination-intensive
-mechanisms. Concisely, we attempt to understand \textit{what benefits
-  do Highly Available Transactions offer and what is their cost?}
+commutative updates~\cite{calm, blooml, crdt}. HAT systems can also provide
+arbitrary foreign key constraints on multi-item updates and sometimes
+provide uniqeness guarantees. However, particularly for programs with
+non-monotonic logic, HATs may fall short, requiring more
+coordination-intensive mechanisms. Concisely, we attempt to understand
+\textit{what benefits do Highly Available Transactions offer and what
+  is their cost?}
 

### 2013-03-27 writeup/intro.tex
neilc comments on intro

@@
 database designers to renounce several guarantees provided by
-traditional database systems. In particular, many modern systems forgo
-providing ACID guarantees---which often provide single-system image
-(SSI) programmability but suffer from unavailability and often
-expensive coordination costs. The CAP Theorem, demonstrating the
-incompatibility of ``always on'' operation with guarantees on data
-item recency in the presence of unreliable networks, has led a large
-number of systems to prioritize high availability and guaranteed low
-latency, which are especially important in geo-replicated contexts. We
-observe a dearth of systems providing general-purpose multi-operation
+traditional database systems~\cite{bigtable, pnuts, dynamo}. In
+particular, many modern systems forgo providing ACID
+guarantees---which often provide single-system image programmability
+but suffer from unavailability and often expensive coordination
+costs~\cite{davidson-survey}. The CAP Theorem~\cite{gilbert-cap},
+demonstrating the incompatibility of ``always on'' operation with
+guarantees on data item recency in the presence of unreliable
+networks, has led a large number of systems to prioritize high
+availability and guaranteed low latency, which are especially
+important in geo-replicated contexts. We observe a dearth of highly
+available systems providing general-purpose multi-operation
 \textit{transactions} over multiple data items with useful semantics.
 
-The relationship of the CAP Theorem and ACID semantics is poorly
-understood. While serializability---the gold standard of SSI
-transactions---is unavailable, databases have a long tradition of
-providing weaker isolation and consistency guarantees. While these
-weaker guarantees do not provide SSI for arbitrary general-purpose
-read/write transactions, they often admit greater concurrency and
-higher performance than comparable SSI algorithms. Accordingly, weak
-isolation is widely employed, overwhelmingly as the default setting in
-today's traditional ``ACID'' and up-and-coming ``NewSQL'' databases
-and often as the maximum guarantees offered
-(Section~\ref{sec:modernacid}). While the prevalence of these
-configurations is encouraging, the current literature does not provide
-guidance as to the operation of weakly consistent isolation guarantees
-in a distributed context. Rather, we have a strong understanding of
-weak isolation in the single-server context from which it
-originated~\cite{adya, ansicritique, gray-isolation} as well as
-several techniques for providing distributed
-serializability~\cite{bernstein-concurrency} but we have little to no
-insight into which ACID guarantees are compatible with Highly
-Available Transactions (HATs), or can be provided with high
-availability and low latency.
+While serializability---the gold standard of traditional ACID
+transactions---is not achievable with high availability for general
+read-write transactions~\cite{davidson-survey}, databases have a long
+tradition of providing weaker isolation and consistency
+guarantees~\cite{adya, ansicritique, gray-isolation}. Due to their
+concurrency and overall performance benefits, weak isolation models
+are widely employed, overwhelmingly as the default setting in today's
+traditional ``ACID'' and up-and-coming ``NewSQL'' databases and often
+as the maximum guarantee offered (Section~\ref{sec:modernacid}). While
+this weak isolation does not provide serializability for
+general-purpose transactions, it is apparently strong enough to
+deliver acceptable behavior to many application programmers and is
+substantially stronger than the typical semantics of current highly
+available systems. This raises the question: which of these semantics
+can we provide while remaining highly available?
+
+Unfortunately, the relationship of the CAP Theorem and ACID semantics
+is poorly understood. We have a strong understanding of weak isolation
+in the single-server context from which it originated~\cite{adya,
+  ansicritique, gray-isolation} as well as several techniques for
+providing distributed
+serializability~\cite{bernstein-concurrency}. However, the current
+literature does not provide guidance as to the operation of weakly
+consistent transactional guarantees in a highly available distributed
+context. Accordingly, in this work, we consider the problem of
+providing Highly Available Transactions (HATs), or semantic guarantees
+regarding multiple operations operating on multiple data items that
+are achievable with high availability and low latency, albeit with
+some cost to achievable semantics. While our preliminary results
+indicate that some models are achievable~\cite{hat-hotos}, the
+literature lacks a deep understanding of what models are available,
+which are not, and when this matters.
 
 This work attempts to bridge the gap between ACID properties and
-highly available systems. We analyze a range of existing ACID and
-distributed consistency models to provide a rigorous taxonomization of
-which models are achievable in a HAT system.  While several
-guarantees, like serializability and Snapshot Isolation, are
-unachievable due to reliance on preventing Lost Update and Write Skew
-phenomena, many other ACID guarantees are HAT compliant, such as
-providing Read Committed isolation, preventing Fuzzy Read phenomena,
-and ensuring atomicity of transactional updates. We also attempt to
-bridge the often distant worlds of distributed system consistency and
-ACID database consistency by analyzing the relationship between
-so-called session guarantees~\cite{sessionguarantees}, ``register''
-semantics~\cite{herlihy-art}, and ACID properties. In doing so, we
-introduce a hierarchy of highly available properties as, for example,
-some guarantees require ``stickiness,'' or fate-sharing between
-clients and database servers. We believe that this analysis is the
-first that has been performed over such a broad range of ACID and
-distributed consistency models.
+highly available systems. We introduce a hierarchy of consistency
+models and rigorously taxonomize their availability characteristics;
+in doing so, we unify previous work on ACID properties~\cite{adya},
+session guarantees~\cite{sessionguarantees}, and traditional
+disributed ``register'' semantics~\cite{herlihy-art}. By identifying
+the availability cost of a wide range of popular consistency
+guarantees, this work allows developers to weigh the costs and
+benefits of relying on these guarantees in their applications. While
+several guarantees, like serializability and Snapshot Isolation, are
+unachievable due to update conflict avoidance (preventing Lost Update
+and Write Skew phenomena), many other ACID guarantees are HAT
+compliant, such as providing Read Committed isolation, preventing
+Fuzzy Read phenomena, and ensuring atomicity of transactional
+updates. Several other guarantees, like reading one's writes, fall
+into a new availability class---``sticky'' high availability---which
+provides an intermediate ground between strict high availability and
+majority availability.
 
-While this model classification is useful, we also seek to understand
-the implications for end-users of HAT systems. HATs provide high
-availability and low latency but cannot offer all of the semantic
-guarantees of unavailable systems---when is this trade-off worthwhile?
-We attempt to objectively study these virtues and limitations of both
-classic and HAT systems by performing an extensive survey of
+This taxonomy allows us to weigh the costs and benefits of a wide
+range of semantic guarantees, but which guarantees are worthwhile in
+practice? We attempt to objectively study the virtues and limitations
+of both classic and HAT systems by performing an extensive survey of
 practitioner accounts and research literature, performing experimental
@@
 range of programs, particularly those with monotonic logic and
-commutative updates~\cite{calm, blooml, crdt}. HAT systems can also provide
-arbitrary foreign key constraints on multi-item updates and sometimes
-provide uniqeness guarantees. However, particularly for programs with
-non-monotonic logic, HATs may fall short, requiring more
+commutative updates~\cite{calm, blooml, crdt}. HAT systems can also
+provide arbitrary foreign key constraints on multi-item updates and
+sometimes provide uniqeness guarantees. However, particularly for
+programs with non-monotonic logic, HATs may fall short, requiring more
 coordination-intensive mechanisms. Concisely, we attempt to understand
@@
 \begin{itemize}
-\item We present a model for high availability in a transactional
-  environment, replica-based High Availability, Sticky High
-  Availability, and Transactional High Availability.
+\item We model high availability in a transactional environment,
+  including traditional and sticky high availability.
 
-\item We analyze the existing literature on ACID properties and
-  replicated data semantics to show which properties are achievable in
-  HAT systems. Achievable properties include transactional atomicity,
-  variants of repeatable read isolation, and session guarantees. We
-  also describe unavailable properties, like preventing Lost Update
-  and Write Skew. We believe that this is the first such
-  characterization in the literature.
+\item We taxonomize ACID and distributed consistency properties
+  according to their availability characteristics.
 
-\item We further motivate the study of HATs via the analysis of
-  existing concurrency control algorithms, a case-study of running a
-  transactional application with HATs, and via deployment of a HAT
-  database prototype on public cloud infrastructure.
+\item We analyze existing concurrency control algorithms, perform a
+  case-study of an existing transactional application, and
+  experimentally evaluate a HAT database prototype.
 \end{itemize}

### 2013-03-27 writeup/motivation.tex
neilc comments on intro

@@
 and latency can ``cause big trouble and painful learning
-experiences''~\cite{fallacies-deutsch}. As an example of an actual system rearchitecture,
-less than one year after its announcement, Yahoo!'s PNUTS developers
-explicitly added a support for weaker, highly available operation,
-explaining that ``strict adherence [to strong consistency] leads to
-difficult situations under network partitioning or server
-failures...in many circumstances, applications need a relaxed
-approach''~\cite{pnuts-update}.
+experiences''~\cite{fallacies-deutsch}.
 
@@
 2010~\cite{research-experiment-partition}---can cause widespread
-partitions. Many of our discussions with practitioners---especially
-those operating on public cloud infrastructure---as well as reports
-from large-scale operators like Google~\cite{dean-keynote} confirm
-that partitions are a reality for service operators today.
+partitions.
+
+Many of our discussions with practitioners---especially those
+operating on public cloud infrastructure---as well as reports from
+large-scale operators like Google~\cite{dean-keynote} confirm that
+partition management is an important consideration service operators
+today. As an example of an actual system rearchitecture, less than one
+year after its announcement, Yahoo!'s PNUTS developers explicitly
+added a support for weaker, highly available operation. The engineers
+explained that ``strict adherence [to strong consistency] leads to
+difficult situations under network partitioning or server
+failures...in many circumstances, applications need a relaxed
+approach''~\cite{pnuts-update}.
 
@@
 guarantees, as evidenced by the lack of highly available systems
-providing them.  As an example of an production database redesign
-explicitly attributable to partition behavior, less than one year
-after its announcement, Yahoo!'s PNUTS system developers explicitly
-added a support for weaker, highly available operation, explaining
-that ``strict adherence [to strong consistency] leads to difficult
-situations under network partitioning or server failures...in many
-circumstances, applications need a relaxed
-approach''~\cite{pnuts-update}.
+providing them.
 
@@
 \begin{table}
+\begin{center}
 \begin{small}
@@
 \end{small}
+\end{center}
 \end{table}

### 2013-03-27 writeup/intro.tex
more intro text

@@
 traditional database systems~\cite{bigtable, pnuts, dynamo}. In
-particular, many modern systems forgo providing ACID
-guarantees---which often provide single-system image programmability
-but suffer from unavailability and often expensive coordination
-costs~\cite{davidson-survey}. The CAP Theorem~\cite{gilbert-cap},
-demonstrating the incompatibility of ``always on'' operation with
-guarantees on data item recency in the presence of unreliable
-networks, has led a large number of systems to prioritize high
-availability and guaranteed low latency, which are especially
-important in geo-replicated contexts. We observe a dearth of highly
-available systems providing general-purpose multi-operation
-\textit{transactions} over multiple data items with useful semantics.
+particular, many modern systems forgo providing ACID guarantees, which
+often provide single-system image programmability but suffer from
+unavailability and often expensive coordination
+costs~\cite{davidson-survey}. In particular, the July 2000
+announcement of the CAP Theorem~\cite{gilbert-cap}, which states that
+a highly available system cannot provide recency guarantees in the
+presence of network partitions, proved a rallying call for a
+generation of scalable but semantically weak systems. Over a decade
+later, we observe a dearth of highly available systems providing
+general-purpose multi-operation \textit{transactions} over multiple
+data items with useful semantics.
 
-While serializability---the gold standard of traditional ACID
-transactions---is not achievable with high availability for general
+While serializability, the gold standard of traditional ACID
+transactions, is not achievable with high availability for general
 read-write transactions~\cite{davidson-survey}, databases have a long

### 2013-03-27 writeup/evaluation.tex
more writing

@@
 
-\begin{enumerate}
+\begin{myenumerate}
 \item We revisit traditional database concurrency control with a focus
@@
   properties on public cloud infrastructure.
-\end{enumerate}
+\end{myenumerate}
 
@@
 
-\vspace{.5em}\noindent\textbf{Serializablity} We begin by considering several
-existing algorithms that provide serializability but are not highly
-available. In summary, all algorithms in this
-class~\cite{bernstein-book} require at least one round trip time (RTT)
-before committing.
+\vspace{.5em}\noindent\textbf{Serializablity} To establish a serial
+order on transactions, all existing algorithms for achieving
+serializability of general-purpose read-write transactions in a
+distributed setting~\cite{bernstein-book} may require at least one
+round trip time (RTT) before committing. As an example, traditional
+two-phase locking for a transaction of length $T$ requires $T$
+\texttt{lock} operations and at least one unlock operation.  Each of
+these operations requires coordination, either with other database
+replicas or with a lock service. If the coordination mechanism for
+locking is unavailable, transactions cannot make progress. Similarly,
+optimistic concurrency control requires contacting a validation
+service, while deterministic transaction
+scheduling~\cite{deterministic-scheduling} requires a
+scheduler. Multi-version concurrency control fares somewhat better but
+must still be checked for write conflicts.
 
-\vspace{.5em}\noindent\textit{Two-phase locking} (ignoring deadlocks) requires $T$ lock
-operations and 1 unlock operation for a transaction of length $T$. If
-all items are located on a single master server, this requires one RTT
-(for requests originating from servers other than the master). If
-items are located on $S$ ($S \leq T$) different servers, two-phase
-locking requires, at minimum, $2(S-1)$ RTTs---one for lock and another
-for unlock.
+Perhaps unsurprisingly, the reliance on a globally agreed total order
+necessitates a minimum of one round-trip to a designated master or
+coordination service for each of these classic algorithms.  This is
+intuitively a consequence of serializability's lack of high
+availability~\cite{davidson-survey} (or, as we have shown, reliance on
+preventing Lost Update and Write Skew).  The cost of the (minimum one)
+round trip will be determined by the deployment environment, as we saw
+in Section~\ref{sec:motivation}.
 
-\vspace{.5em}\noindent\textit{Optimistic concurrency control} allows
-reads from any server but transactions must be checked by a scheduler
-before commit. With one scheduler per database, this requires one RTT
-specifically for concurrency control. If schedulers are sharded by
-database partition, a multi-partition transaction may require
-contacting several different schedulers and, possibly, an atomic
-commit protocol, which becomes substantially more
-expensive.
-
-\vspace{.5em}\noindent\textit{Deterministic transaction scheduling} is
-effectively the inverse of optimistic concurrency control: a scheduler
-first orders the transactions, then sends them to database replicas
-for in-order execution. These techniques, pioneered in the context of
-totally ordered atomic broadcast
-systems~\cite{deterministic-scheduling}, have recently seen increased
-attention~\cite{calvin} and effectively require communication overhead
-similar to optimistic concurrency control. Finally,
-
-\vspace{.5em}\noindent\textit{Multiversion concurrency control} allows readers to read older
-versions of data items but, like OCC, must eventually be checked for
-write-write conflicts, requiring a scheduler.
-
-\vspace{.5em} Perhaps unsurprisingly, the reliance on a global total
-order necessitates a minimum of one round-trip to a designated master
-or coordination service for each of these classic algorithms. This is
-simply a cost of providing serializability, and the cost of the
-(minimum one) round trip will be determined by the deployment
-environment, as we saw in Section~\ref{sec:motivation}.
-
-\vspace{.5em}\noindent\textbf{Non-serializablity} While it is possible to use
-several of the prior mechanisms for HAT systems---for example, a
-multi-version storage system is a reasonable candidate for the basis
-of a HAT database, albeit with write conflict checking
-disabled---several techniques, like locking, are not suitable for
-HATs. In fact, there are several mechanisms that do not provide
-serializability but provide transactional semantics with varying
-availability properties:
-
-\vspace{.5em}\noindent\textit{Lock-based protocols} such as those used to provide weak
-  isolation in Gray's original proposal~\cite{gray-isolation} require
-  at least one RTT to the a lock server (or, alternatively, a majority
-  quorum). In the presence of failure, distributed locking is
-  challenging.
-
-\vspace{.5em}\noindent The \textit{MDCC} commit protocol uses a Paxos
-variant and escrow transactions to provide read committed (without
-Lost Update) and Non-monotonic Snapshot Isolation with 1-2 RTTs but is
-unavailable due to Lost Update~\cite{mdcc}.
-
-\vspace{.5em}\noindent\textit{Parallel Snapshot Isolation} as
-developed by the Walter system provides (unavailable) snapshot
-isolation variant and, even when writes commute, requires 1 RTT to
-enforce read recency~\cite{walter}. However, like other Snapshot
-Isolation variants, PSI could be made highly available by giving up
-recency requirements and Lost Update prevention.
-
-\vspace{.5em}\noindent\textit\textbf{Eiger} provides read-only and
-write-only transactions with item cut isolation and causal
-consistency. It relies on sticky high availability with a linearizable
-set of replicas within a datacenter. This provides low latency across
-data centers. In the presence of server failure during write
-transactions, reads and writes may become unavailable. Eiger's
-predecessor \textit{COPS} provides similar semantics for read-only
-transactions but does not suffer from unavailability as it does not
-rely have a coordinator.
-
-\vspace{.5em}\noindent\textit{Chan and Gray's Read-only transactions}
-use a similar algorithm to Eiger to provide read-only transactions
-with item cut isolation and causal consistency but client version
-selection coordination is performed on reads instead of
-writes~\cite{readonly}.
-
-\vspace{.5em}\noindent\textit{Bolt-on Causal
-  Consistency}~\cite{bolton} and \textit{Swift}~\cite{swift} can
-provide read-only and write-only transactions with item cut isolation
-and causal consistency with sticky high availability and both rely on
-client-side caching.
-
-\vspace{.5em}\noindent\textit{Bayou} provides several session
-guarantees but, in its original implementation, may face unavailabilty
-in the presence of failure~\cite{sessionguarantees}.
-
-\vspace{.5em}\noindent\textit{Brantner's S3 Database} uses caching to
-provide several session and transactional guarantees on top of cloud
-infrastructure, often relying on linearizable queues and, accordingly,
-are often unavailable~\cite{kraska-s3}
-
-\vspace{.5em}
+\vspace{.5em}\noindent\textbf{Non-serializablity} Perhaps
+interestingly, many existing distributed implementations of weak
+isolation are not highly available. Lock-based proposals such as those
+used to provide weak isolation in Gray's original
+proposal~\cite{gray-isolation} do not degrade gracefully in the
+presence of partial failure. However, lock-based protocols \textit{do}
+provide recency guarantees. While multi-versioned storage systems
+allow for a variety of transactional guarantees, we have not seen
+proposals for truly weak isolation (e.g., non-``tentative update''
+schemes) in this context. Chan and Gray's read-only transactions
+provide read-only transactions with item-cut isolation, causal
+consistency, and transactional atomicity but are unavailable in the
+presence of coordinator failure~\cite{readonly}, similar to read-only
+and write-only transactions more recently proposed by
+Eiger~\cite{eiger}.  Bolt-on causal consistency~\cite{bolton},
+Swift~\cite{swift}, Brantner's S3 database~\cite{kraska-s3}, and
+Bayou~\cite{sessionguarantees} can all provide variants of the above
+model with high availability, but none provide this HAT functionality
+without modification. Similarly, the MDCC~\cite{mdcc} protocol offers
+Read Committed isolation with Lost Update avoidance but is similarly
+unavailable due to its reliance on preventing write conflicts.
 
@@
 than serializability---including guarantees that are achievable with
-high availability---and not achieve high availability. These
-designs---though often unavailable---offer benefits unavailable to
-HATs: for example, lock-based read committed, provide recency
-guarantees in addition to read committed isolation! We expect that
+high availability---and not achieve high availability. We expect that
 further understanding the trade-offs between semantic strength,

### 2013-03-27 writeup/hat-analysis.tex
more writing

@@
 transaction's updates are present on their respective
-replicas. However, servers do \textit{not} need to agree exactly when
-all replicas are aware that all of a transaction's updates are present
-on their respective replicas. This subtle distinction corresponds to
-the difference between the distributed systems concepts of reliable
-broadcast and consensus. The former requires that, if one server
-reveals a write, then all servers reveal a write, while the latter
-requires explicit agreement across servers (and is unavailable). We
-omit a full discussion of the algorithm and instead refer the reader
-to the extended version of this paper~\cite{hat-tr}.
+replicas. However, replicas do not need to agree on when to reveal a
+new value to clients. In distributed systems terminology,
+transactional atomicity only requires reliable broadcast (with some
+additional metadata for clients). We omit a full discussion of the
+algorithm and instead refer the reader to the extended version of this
+paper~\cite{hat-tr}.
 
@@
 
-\vspace{.5em}\noindent\textit{\textbf{Monotonic reads}} stipulates
-that a client's subsequent reads to a given object ``never return any
+\vspace{.5em}\noindent\textit{\textbf{Monotonic reads}} requires that
+a client's subsequent reads to a given object ``never return any
 previous values''; reads from each item progress forward in a total
@@
 reveal new writes to a data item when all other servers responsible
-for the item have also seen them. Otherwise, a client might read a
-newer version, subsequently become partitioned from its prior server,
-and contact a comparatively out-of-date server. In the case of
-Monotonic Writes, each new write should become visible only when each
-client's prior write is visible everywhere, along with Monotonic Reads
-guarantees. For Writes Follow Reads, a servers should similarly wait
-to reveal items until their dependencies are present everywhere.
-
-Effectively, this mechanism requires that clients only read from a
-globally agreed upon lower bound on the versions written to the
-system. This is highly available as a client will never block
-due to inability to find a server with a sufficiently up-to-date
-version of a data item. However, it does not imply that transactions
-will read their own writes or provide much confidence in making
-forward progress through the version history in the presence of
-partitions. The problem is that, if a server is partitioned, under the
-highly available model, we cannot discount the possibility that an
+for the item have also seen them. In the case of Monotonic Writes and
+Writes Follow Reads, each new write should become visible only when
+each write's dependencies are visible everywhere. This mechanism
+effectively requires that all clients read from a globally agreed upon
+lower bound on the versions written to the system. This is highly
+available as a client will never block due to inability to find a
+server with a sufficiently up-to-date version of a data item. However,
+it does not imply that transactions will read their own writes or, in
+the presence of partitions, make forward progress through the version
+history. The problem is that, if a server becomes partitioned, under
+the highly available model, we must handle the possibility that an
 unfortunate client will be forced to issue her next requests against

### 2013-03-27 writeup/motivation.tex
more writing

@@
 highest priority tickets ranges from 114--188 minutes and up to a full
-day for all tickets~\cite{turner2012failure}. The incidence of network
-failures is further confirmed by several prior studies showing for
-example, on Sprint's WANs, a median time to repair between 2 and 1000
-seconds, and a median time between failures of approximately 3000
-seconds~\cite{ip-backbone-failures}, as well as frequent path routing
+day for all tickets~\cite{turner2012failure}. Other studies confirm
+these results, showing median time between failures of approximately
+3000 seconds with a median time to repair between 2 and 1000
+seconds~\cite{ip-backbone-failures} as well as frequent path routing
 failures on the Internet~\cite{labovitz-failures}. Isolating,
-quantifying, and designing for these network failures is an area of active
-research in networking community~\cite{surviving-failures-bodik,
-  uw-failure-networks}.
+quantifying, and designing for these network failures is an area of
+active research in networking
+community~\cite{surviving-failures-bodik, uw-failure-networks}.
 

### 2013-03-28 writeup/evaluation.tex
Moar writing

@@
 
-\vspace{.5em}\noindent\textbf{TPC-C} Absent a necessary and sufficient result describing exactly which
-application logics can be achieved with high availability and to make
-these requirements more concrete, we consider an example application:
-the TPC-C benchmark. Our goal is \textit{not} to provide a performance
-analysis, as any HAT configuration will be non-compliant. Rather, we
-examine the five queries contained in the benchmarks---ostensibly
-representative of a standard transaction processing application---to
-understand which are achievable in a HAT context. TPC-C consists of
-five transactions, capturing operation of a real-world warehouse,
-including sales, payments, and deliveries. We analyze each kind of
-transaction and its interactions below.
+\vspace{.5em}\noindent\textbf{TPC-C} Absent a necessary and sufficient
+result describing exactly which application logics can be achieved
+with high availability and to make these requirements more concrete,
+we consider an example application: the TPC-C benchmark. Our goal is
+\textit{not} to provide a performance analysis, as any HAT
+configuration will be non-compliant. Rather, we examine the five
+queries contained in the benchmarks---ostensibly representative of a
+standard transaction processing application---to understand which are
+achievable in a HAT context. Surprisingly, four of five can be
+executed with HATs, while the fifth potentially requires unavailablity.
 
-\vspace{.5em}\noindent\textit{New-Order} places an order for a
+TPC-C consists of five transactions, capturing operation of a
+real-world warehouse, including sales, payments, and deliveries. Two
+transactions---Order-Status and Stock-Level---are read-only and can be
+executed safely under HATs. Clients may read stale data, but this does
+not violate TPC-C requirements and clients will read their writes if
+they are sticky-available. Another transaction type, Payment, updates
+running balances for warehouses, districts, and customer records as
+well as providing an audit trail. The transaction is increment-only,
+so all balance increase operations can commute, and TA allows the
+maintainence of any foreign-key constraints.
+
+While three out of five transactions have been easily achievable with
+HATs, the remaining two transactions---New-Order and Delivery---are
+not so simple. The New-Order transaction places an order for a
 variable quantity of data items, updating warehouse stock as
-needed. The transaction selects a sales district, assigns the order an
-ID number, adjusts the remaining warehouse stock, and writes a
-placeholder entry for the pending order. There are two mutations we
-must be consider in detail.
+needed. It selects a sales district, assigns the order an ID number,
+adjusts the remaining warehouse stock, and writes a placeholder entry
+for the pending order. The Delivery transaction represents the
+fulfillment of a New-Order: it deletes the order from the pending
+list, updates the customer's balance, updates the order's carrier ID
+and delivery time, and updates the customer balance.
 
-First, assigning an order ID number is challenging. We can easily
-assign a \textit{unique} order ID, by, say, concatenating the customer
-ID with a timestamp and server ID. However, the TPC-C specification
-requires that order numbers are assigned sequentially, which we cannot
-guarantee in HATs because assigning sequential IDs requires preventing
-Lost Update. Accordingly, HATs can allow placement of new orders with
-unique identifiers but cannot provide compliant TPC-C execution due to
-the requirement for a non-monotonic counter increase.
+There are three challenges that only allow HATs to place non-compliant
+New-Orders and to fail on Payments. First, we must assign each new
+order an ID number. We can assign a unique number by, say,
+concatenating the client ID and a timestamp. However, the TPC-C
+specification requires order numbers to be assigned sequentially,
+which requires Lost Update. Accordingly, HATs cannot provide compliant
+TPC-C execution but can maintain uniqueness constraints. Second, the
+New-Order transaction decrements inventory counts: what if the count
+becomes negative? Fortunately, TPC-C New-Order restocks the inventory
+count (increments by 91) if it would become negative as the result of
+placing an order. This means that, even in the presence of concurrent
+New-Orders, an item's stock will never fall below zero. This is TPC-C
+compliant, but a HAT system might end up with substantially more stock
+than in a single-site implementation. Third, delivery events are
+non-monotonic---deleting a pending New-Order transaction---and should
+be idempotent in order to avoid billing a customer twice; this implies
+a need to prevent Lost Update. We can avoid this issue by moving the
+non-monotonicity to the real world---the carrier that picks up the
+package for an order can ensure that she is the only carrier who has
+done so---but cannot provide a correct execution with HATs alone.
 
-Second, decrementing inventory counts is, in general, a potentially
-risky operation because, without coordination, inventory counts may
-become negative. However, TPC-C automatically ``restocks'' inventory
-counts (by 91 items) when there are 10 or fewer items
-remaining. Accordingly, even in the presence of concurrent new orders
-to the same item, the stock will never fall below zero. However, an
-$R$ replica HAT system might end up with $91(R-1)$ more stock than in
-a single-site implementation.
-
-\vspace{.5em}\noindent\textit{Payment} updates running balances for
-warehouses, districts, and customer records as well as providing an
-audit trail. All balance increase operations can be made commutative,
-so the transaction can execute under HATs.
-
-\vspace{.5em}\noindent\textit{Order-Status} is a read-only transaction
-and can be safely executed under HATs. The TPC-C specification does
-not require recency bounds but, in practice, they may be desirable. If
-a client is sticky with a set of replicas, she will read her own
-writes, which may be sufficient.
-
-\vspace{.5em}\noindent\textit{Delivery} represents the fulfillment of
-a pending order created by a New Order transaction: it deletes the
-order from the pending list, updates the customer's balance, updates
-the order's carrier ID and delivery time, and updates the customer
-balance. The challenge in delivery comes in deletion, which is
-non-monotonic: an item might be removed twice from the pending
-table. If delivery can be made idempotent---that is, if multiple
-Delivery transactions have the same effect as one---then multiple
-clients can execute the same Deliveries without fatal consequences.
-
-In practice, given that a physical shipment is likely only delivered
-once, it is unlikely that multiple agents will attempt a Delivery
-transaction for the same order simultaneously. However, in this case,
-we have simply moved the non-monotonicity to the real world: the
-carrier that picks up the package for an order must ensure that she is
-the only carrier who has done so. While we can program around the
-other operations in the order, HATs cannot--without the help of
-business-specific processes--serve Delivery transactions. That said,
-New-Orders may be placed with high availability.
-
-\vspace{.5em}\noindent\textit{Stock-Level} is another read-only
-transaction and can be executed without any non-HAT guarantees, with
-the same caveats regarding recency.
-
-\vspace{.5em} Throughout execution, TPC-C also requires the
-maintainence of several integrity constraints. For example,
-Consistency Condition 1 (3.3.2.1) requires that each warehouse's sales
-count must reflect the sum of its subordinate sales districts. This
-integrity constraint spans two tables but, given the ability to update
-both tables atomically via transactional atomicity, can be easily
-maintained. Consistency Conditions 4 through 12 (3.3.2.4-12) can
-similarly be satisfied by applying updates atomically across
-tables. Consistency Conditions 2 and 3 (3.3.2.2-3) concern order ID
-assignment and are more problematic, as we have discussed.
+Throughout execution, TPC-C also requires the maintainence of several
+integrity constraints. For example, Consistency Condition 1 (3.3.2.1)
+requires that each warehouse's sales count must reflect the sum of its
+subordinate sales districts. This integrity constraint spans two
+tables but, given the ability to update both tables atomically via
+transactional atomicity, can be easily maintained. Consistency
+Conditions 4 through 12 (3.3.2.4-12) can similarly be satisfied by
+applying updates atomically across tables. Consistency Conditions 2
+and 3 (3.3.2.2-3) concern order ID assignment and are more
+problematic, as we have discussed.
 

### 2013-03-28 writeup/hat-analysis.tex
Moar writing

@@
 order is completely arbitrary (i.e., an order must exist), whereas a
-serializable total order must be equivalent to a serial
-execution. Read Uncommitted is similar to the distributed systems
-concept of replica convergence, in which all replicas for a data item
-must end up with the same data item. This convergence is usually
-achieved by picking a ``winning'' update, requiring a means of totally
-ordering updates to each item; one can view Read Uncommitted as a
-cross-item convergence property. It is easily achieved via applying
-the same logical timestamp to each update in a transaction.
+serializable total order must be equivalent to a serial execution. It
+is easily achieved via applying the same logical timestamp to each
+update in a transaction.
 
@@
 operations except predicate-based reads. However, in its description
-of Repeatabable Read, the ANSI SQL specification provides a useful
+of Repeatable Read, the ANSI SQL specification provides a useful
 property which, although is not true to Gray's spirit of Repeatable
@@
 ``Repeatable Read'': unless a transaction modifies a given data item,
-the value of the data item should not change during the
+the observable value of the data item should not change during the
 transaction. By itself, this property is rather weak---we can always
-read $\bot$ for each data item. However, when coupled with
-additional properties, like transactional atomicity, it is stronger:
-the repeated reads must obey additional ordering constraints. In fact,
-distributed systems describe a ``consistent snapshot'' across a set of
-related events as a \textit{cut} across a set of participants or, in
-our case, data items. Accordingly, to capture the notion that the
-transaction should read from a (non-changing) consistent cut over data
-items (and to disambiguate from the aforementioned stronger Repeatable
-Read properties), we call ANSI SQL Repeatable Read (preventing P2)
-\textbf{Cut Isolation}.It is possible to satisfy cut isolation with
+read $\bot$ for each data item. However, when coupled with additional
+properties, like transactional atomicity, it is stronger: the repeated
+reads must obey additional ordering constraints. In fact, distributed
+systems describe a ``consistent snapshot'' across a set of related
+events as a \textit{cut} across a set of participants or, in our case,
+data items. Accordingly, to capture the notion that the transaction
+should read from a (non-changing) consistent cut over data items (and
+to disambiguate from the aforementioned stronger Repeatable Read
+properties), we call ANSI SQL Repeatable Read (preventing P2)
+\textbf{Cut Isolation}. It is possible to satisfy cut isolation with
 high availability by caching appropriate versions for a transaction's
@@
 
-Transactional atomicity is core to ACID guarantees. Although, at least
-by the acronym, it is not an ``isolation'' property, transactional
-atomicity restricts transactions' ability to view the effects of
-partially completed transactions. Under transactional atomicity,
-within each transaction, either all effects of another transaction are
-observed, or none are (equivalently, once some of the effects of a
-transaction are observed, all effects are observed). This is a
-strictly stronger guarantee than Read Committed isolation: in Read
-Committed, we can read a subset of committed writes whereas, with
-transactional atomicity, we must read all of them. For example, $T_2$
-must observe $b=c=1$ (or later versions for each key):
+Transactional atomicity (TA) is core to ACID guarantees. Although, at
+least by the acronym, it is not an ``isolation'' property,
+transactional atomicity restricts transactions' ability to view the
+effects of partially completed transactions. Under transactional
+atomicity, within each transaction, either all effects of another
+transaction are observed, or none are (equivalently, once some of the
+effects of a transaction are observed, all effects are observed). This
+is a strictly stronger guarantee than Read Committed isolation: in
+Read Committed, we can read a subset of committed writes whereas, with
+transactional atomicity, we must read all of them. Together with cut
+isolation, transactional atomicity prevents Read Skew anomalies
+(Berenson et al.'s A5A~\cite{ansicritique}). As an example of
+TA, $T_2$ must observe $b=c=1$ (or later versions
+for each key):
 \vspace{-.5em}
@@
 
-Perhaps perplexingly, discussions of transactional atomicity are
-absent from existing discussions of weak isolation levels. This is
-perhaps again due to the single-node context in which prior work was
-developed: on a single server (or a fully replicated database),
-atomicity is near-trivial to achieve via lightweight locking and/or
-local concurrency control over data items. (In this spirit, multi-item
-atomicity is common in so-called ``entity groups,'' which logically
-represent groups of co-located items~\cite{entitygroup}.) In contrast,
-in a distributed environment, atomicity over arbitrary groups of items
-is considerably more difficult to achieve with high availability. The
-key challenge is that different servers responsible for subsets of a
-transaction's updates must coordinate regarding when to reveal their
-writes to a transaction---without violating availability due to, say,
-locking, or a missing transactional write. Together with cut
-isolation, transactional atomicity prevents Read Skew anomalies
-(Berenson et al.'s A5A~\cite{ansicritique}).
-
-Transactional atomicity is achievable in a HAT environment. To ensure
-transactional atomicity, servers need to know when all of a
+Perhaps perplexingly, discussions of TA are absent from existing
+discussions of weak ACID. This is perhaps again due to the single-node
+context in which prior work was developed: on a single server (or a
+fully replicated database), TA is near-trivial to achieve via
+lightweight locking and/or local concurrency control over data
+items. In contrast, in a distributed environment, TA over arbitrary
+groups of non-colocated items is considerably more difficult to
+achieve with high availability: servers need to know when all of a
 transaction's updates are present on their respective
 replicas. However, replicas do not need to agree on when to reveal a
-new value to clients. In distributed systems terminology,
-transactional atomicity only requires reliable broadcast (with some
-additional metadata for clients). We omit a full discussion of the
-algorithm and instead refer the reader to the extended version of this
-paper~\cite{hat-tr}.
+new value to clients, which would require consensus; accordingly, TA
+only requires reliable broadcast (with some additional metadata for
+clients) and is achievable in a HAT system. We omit a full discussion
+of the algorithm and instead refer the reader to the extended version
+of this paper~\cite{hat-tr}.
 
@@
 groups of operations (here, transactions) issued by the same logical
-user~\cite{sessionguarantees}. Informally, a \textit{session} is an
-application context that consists of logical state that should persist
-between operations. For example, users accessing a web site may want
-to experience a notion of ``forward progress'' for the duration that
-they are logged in; all transactions submitted between ``log in'' and
-``log out'' operations might form a session. We can define sessions at
-several granularities (e.g., per-transaction, per-client, per-time unit)
-but, for simplicity, will refer to client-level guarantees in the
-discussion.
+user~\cite{sessionguarantees}. A \textit{session} is an application
+context that consists of logical state that should persist between
+operations. For example, all of a user's transactions submitted
+between ``log in'' and ``log out'' operations might form a session.
 
@@
 
-\vspace{.5em}\noindent\textit{\textbf{Monotonic reads}} requires that
-a client's subsequent reads to a given object ``never return any
+\vspace{.5em}\noindent\textit{\textbf{Monotonic reads}} requires that,
+with a session, subsequent reads to a given object ``never return any
 previous values''; reads from each item progress forward in a total
@@
 
-\vspace{.5em}\noindent\textit{\textbf{Monotonic writes}} requires
-that transactions from each individual client be serialized. Any order
-on transactions should also respect the order in which each session
-submitted transactions.
+\vspace{.5em}\noindent\textit{\textbf{Monotonic writes}} requires that
+each session's writes be serialized. Any order on transactions should
+also respect the order in which each session submitted transactions.
 
@@
 reveal new writes until their respective dependencies are fulfilled on
-all replicas. In the case of Monotonic Reads, servers should only
-reveal new writes to a data item when all other servers responsible
-for the item have also seen them. In the case of Monotonic Writes and
-Writes Follow Reads, each new write should become visible only when
-each write's dependencies are visible everywhere. This mechanism
-effectively requires that all clients read from a globally agreed upon
-lower bound on the versions written to the system. This is highly
-available as a client will never block due to inability to find a
-server with a sufficiently up-to-date version of a data item. However,
-it does not imply that transactions will read their own writes or, in
-the presence of partitions, make forward progress through the version
-history. The problem is that, if a server becomes partitioned, under
-the highly available model, we must handle the possibility that an
-unfortunate client will be forced to issue her next requests against
-the partitioned server!
+all replicas. This mechanism effectively requires that all clients
+read from a globally agreed upon lower bound on the versions written
+to the system. This is highly available as a client will never block
+due to inability to find a server with a sufficiently up-to-date
+version of a data item. However, it does not imply that transactions
+will read their own writes or, in the presence of partitions, make
+forward progress through the version history. The problem is that, if
+a server becomes partitioned, under the highly available model, we
+must handle the possibility that an unfortunate client will be forced
+to issue her next requests against the partitioned server!
 
@@
 stickiness, and, because causal consistency requires read your writes,
-causal consistency also requires stickiness.
-
-However, read-your-writes can be accomplished via client-side caching
-and alternatively comes ``for free'' when requests are routed to a
-sticky set of servers, while causality and PRAM can be accomplished
-via sticky versions of the prior algorithms.
+PRAM and causal consistency also require stickiness. Read-your-writes
+is ``free'' in a sticky system, while the remaining causality and PRAM
+guarantees can be accomplished with the prior algorithms.
 
@@
 \noindent{\textbf{Consistency}} A HAT system can make limited
-consistency guarantees. It can execute commutative and logically
+consistency guarantees. It can often execute commutative and logically
 monotonic~\cite{calm} operations without the risk of invalidating
@@
 
-\vspace{.5em}\noindent{\textbf{Convergence}} Finally, to
-require that the system propagates writes between replicas, we can
-require convergence, or eventual consistency for each data item: in
-the absence of new mutations to a data item, in the absence of
-partitions, all servers eventually agree on the value for each
-item. This is typically accomplished by any number of anti-entropy
-protocols, which periodically update neighboring servers with the
-latest value for each data item. This convergence property should be
-considered mandatory given the spirit of existing literature on
-isolation protocols: in all models we have encountered, the database
-should behave as if there is some total order over
-transactions. Indeed, non-convergent systems are remarkably difficult
-to reason about.
+\vspace{.5em}\noindent{\textbf{Convergence}} To require that the
+system propagates writes between replicas, we can require convergence,
+or eventual consistency for each data item: in the absence of new
+mutations to a data item, in the absence of partitions, all servers
+eventually agree on the value for each item. This is typically
+accomplished by any number of anti-entropy protocols, which
+periodically update neighboring servers with the latest value for each
+data item. Establishing a final value is related to determining a
+total order on transaction updates, as in Read Uncommitted.
 

### 2013-03-28 writeup/evaluation.tex
moar writin

@@
 HATs. The two problematic transactions---New-Order and Payment---rely
-on non-monotonic state updates. The former transaction can be modified
+on non-monotonic state update. The former transaction can be modified
 to ensure ID uniqueness at the expense of sequential per-district ID
@@
 expect that, especially for read-dominated workloads found in many
-online services, HAT guarantees may suffice for many
+online services, HAT guarantees will suffice for many
 applications. Finally, while TPC-C is not subject to write skew

### 2013-03-28 writeup/futurework.tex
moar writin

--- 2013-03-17 writeup/futurework.tex
@@
 \section{Future Work}
+\label{sec:futurework}
 
-VII. Future Work (.75pg)
+We believe that there are several promising avenues for future work in
+Highly Available Transaction systems:
 
-* What properties are available?
-	* We've focused on ACID Isolation, but there are lots of other models to consider?
-	* Which consistency criteria? Which models? There are infinite, so how do we choose?
-	* Set ourselves up for APEs
+\textbf{What is the space of consistency and isolation properties?} In
+this work, we have taxonomized several well-defined and
+well-documented consistency properties. This is useful as it
+contextualizes decades of existing database and distributed systems
+design and algorithms but does not provide guidance as to what
+consistency models are the ``strongest''---or admit the smallest set
+of executions---within each availability class. A recent report from
+UT Austin shows that no model stronger than causal consistency can be
+achieved in a sticky highly available, one-way convergent
+system~\cite{cac}, but there are infinite highly available and sticky
+highly available models to consider. For example, always returning the
+value $1$ in response to reads is incomparable with (i.e., neither
+stronger nor weaker than) causal consistency: returning $1$ does not
+respect causality, and causal consistency returns values other than
+$1$. Always returning the value $2$ is similarly incomparable, and, as
+an example, we can enumerate models that return all integers and real
+numbers--an uncountably infinite set of models. Given these infinite
+models, are there meaningful properties that delineate the boundaries
+between classes of highly available systems? Is there a
+``meta-property'' that exactly characterizes what properties are
+achievable with high availability or sticky high availability? This is
+an area of ongoing research within the Berkeley HAT project.
 
-* Full exploration of HAT space
-	* Many parameters; $2^10$ models at the least
-	* Need to mitigate the cost of tracking causality with high availability
-		* The cost moving from S-HA (like Eiger) to truly HA (R-HA) seems to be high
-		* e.g., Eiger relies on metadata GC to keep writes small; is O(|keys|) in limit
-			* Not meant as a fault--it's optimized for a particular scenario--but curious to consider
+\textbf{What are the costs and benefits within HAT designs?} There are
+many possible combinations of HAT guarantees, and the performance and
+ease of implementation of a HAT system will vary depending on which
+guarantees it provides. For example, as we have discussed, the cost in
+moving from a sticky high available implementation of highly available
+guarantees to a truly highly available system seems expensive for some
+models (e.g., metadata requirements or visibility requirements may
+become problematic). Further experimentation regarding the set of
+useful and achievable HAT guarantees will shed further light on this
+question.
 
-* Revisit semantics-based concurrency control
-	* Could taxonomize properties like escrow
-	* Can also consider models with SLAs and bounded messaging asynchrony
-	* Our focus is on "classic" distributed systems trade-off
-		* However, lots of interesting hybrids (maybe cite Doug Terry's recent stuff)
+\textbf{Where do semantics-based concurrency control and weakened
+  failure models lie in the HAT landscape?} The vast amount of
+literature on semantics-based concurrency control provides hints as to
+when coordination can be avoided while providing ``strongly
+consistent'' outcomes for limited failure scenarios. We have performed
+a narrow case study in this work, but a general taxonomization of
+techniques such as escrow transactions~\cite{escrow} and
+Sagas~\cite{sagas} could help answer the challenge of programming in a
+weakly consistent environment. Additionally, HAT guarantees can be
+strengthened by considering failure models with, say, bounded
+asynchrony or service-level agreements from underlying
+hardware. Additionally, while we have focused on systems that provide
+a single availability mode, there are several interesting hybrids to
+consider.

### 2013-03-29 thebes-code/common/src/main/java/edu/berkeley/thebes/common/config/Config.java
Make Mastered routing split evenly

@@ -141,7 +141,7 @@ public class Config {
     }
     
     /** Returns the cluster map (based on the current transaction mode). */
-    private static Map<Integer, List<String>> getClusterMap() {
+    public static Map<Integer, List<String>> getClusterMap() {
         return getOption(txnMode.getClusterConfigParam());
     }
 

### 2013-03-29 thebes-code/hat/client/src/main/java/edu/berkeley/thebes/hat/client/clustering/MasteredReplicaRouter.java

@@ -1,0 +1,65 @@
+package edu.berkeley.thebes.hat.client.clustering;
+
+import org.apache.thrift.transport.TTransportException;
+
+import com.google.common.collect.Lists;
+import com.google.common.collect.Maps;
+
+import edu.berkeley.thebes.common.clustering.RoutingHash;
+import edu.berkeley.thebes.common.config.Config;
+import edu.berkeley.thebes.common.thrift.ServerAddress;
+import edu.berkeley.thebes.hat.common.thrift.ReplicaService;
+import edu.berkeley.thebes.hat.common.thrift.ThriftUtil;
+
+import java.io.IOException;
+import java.util.ArrayList;
+import java.util.List;
+import java.util.Map;
+
+public class MasteredReplicaRouter extends ReplicaRouter {
+
+    private final Map<Integer, List<ServerAddress>> replicaAddressesByCluster;
+    private final Map<Integer, List<ReplicaService.Client>> syncReplicasByCluster;
+    private final int numClusters;
+    private final int numNeighbors;
+
+    public MasteredReplicaRouter() throws TTransportException, IOException {
+        assert(Config.shouldRouteToMasters());
+        
+        this.replicaAddressesByCluster = Maps.newHashMap();
+        this.syncReplicasByCluster = Maps.newHashMap();
+        this.numClusters = Config.getSiblingServers().size();
+        this.numNeighbors = Config.getServersInCluster().size();
+        
+        for (int i = 0; i < numClusters; i ++) {
+            List<ServerAddress> neighbors = Config.getServersInCluster();
+            List<ReplicaService.Client> neighborClients = Lists.newArrayList();
+            for (ServerAddress neighbor : neighbors) {
+                neighborClients.add(ThriftUtil.getReplicaServiceSyncClient(neighbor.getIP(),
+                        neighbor.getPort()));
+            }
+            replicaAddressesByCluster.put(i+1, neighbors);
+            syncReplicasByCluster.put(i+1, neighborClients);
+        }
+    }
+
+    @Override
+    public ReplicaService.Client getSyncReplicaByKey(String key) {
+        int hash = RoutingHash.hashKey(key, numNeighbors);
+        int clusterID = (hash % numClusters) + 1;
+        return syncReplicasByCluster.get(clusterID).get(hash);
+    }
+
+    @Override
+    public ReplicaService.AsyncClient getAsyncReplicaByKey(String key) {
[…]

### 2013-03-29 thebes-code/hat/client/src/main/java/edu/berkeley/thebes/hat/client/clustering/StandardReplicaRouter.java

@@ -1,0 +1,48 @@
+package edu.berkeley.thebes.hat.client.clustering;
+
+import org.apache.thrift.transport.TTransportException;
+
+import edu.berkeley.thebes.common.clustering.RoutingHash;
+import edu.berkeley.thebes.common.config.Config;
+import edu.berkeley.thebes.common.thrift.ServerAddress;
+import edu.berkeley.thebes.hat.common.thrift.ReplicaService;
+import edu.berkeley.thebes.hat.common.thrift.ThriftUtil;
+
+import java.io.IOException;
+import java.util.ArrayList;
+import java.util.List;
+
+public class StandardReplicaRouter extends ReplicaRouter {
+
+    private List<ReplicaService.Client> syncReplicas;
+    private List<ReplicaService.AsyncClient> asyncReplicas;
+
+    public StandardReplicaRouter() throws TTransportException, IOException {
+        assert(!Config.shouldRouteToMasters());
+        
+        List<ServerAddress> serverIPs = Config.getServersInCluster();
+        
+        syncReplicas = new ArrayList<ReplicaService.Client>(serverIPs.size());
+        asyncReplicas = new ArrayList<ReplicaService.AsyncClient>(serverIPs.size());
+
+        for (ServerAddress server : serverIPs) {
+            syncReplicas.add(ThriftUtil.getReplicaServiceSyncClient(server.getIP(), server.getPort()));
+            asyncReplicas.add(ThriftUtil.getReplicaServiceAsyncClient(server.getIP(), server.getPort()));
+        }
+    }
+
+    @Override
+    public ReplicaService.Client getSyncReplicaByKey(String key) {
+        return syncReplicas.get(RoutingHash.hashKey(key, syncReplicas.size()));
+    }
+
+    @Override
+    public ReplicaService.AsyncClient getAsyncReplicaByKey(String key) {
+        return asyncReplicas.get(RoutingHash.hashKey(key, asyncReplicas.size()));
+    }
+
+    @Override
+    public ServerAddress getReplicaIPByKey(String key) {
+        return Config.getServersInCluster().get(RoutingHash.hashKey(key, syncReplicas.size()));
+    }
+}

### 2013-03-29 writeup/evaluation.tex

@@
 so all balance increase operations can commute, and TA allows the
-maintainence of any foreign-key constraints.
+maintainence of any foreign-key constraints (e.g.,
+\texttt{UPDATE/DELETE CASCADE}).
 
@@
 
+\begin{figure}
+\includegraphics[width=.8\columnwidth]{figs/macrotracelegend.pdf}\vspace{-2mm}
+\includegraphics[width=\columnwidth]{figs/lat-thru-trace-K0-SCALEMACRO.pdf}
+\includegraphics[width=\columnwidth]{figs/lat-thru-trace-K0-SCALEMACRO.pdf}
+\caption{Latency-throughput for database operation on EC2 (DUMMY FIG)}
+\label{fig:wan-exp}
+\end{figure}
+
 2.) We ran YCSB on EC2 (and maybe TPC-C if we have time) with a few models:

### 2013-03-29 writeup/hat-analysis.tex

@@
 \end{align*}
-$T_2$ can also observe $a=\bot$ or $a=1$.
+$T_2$ can also observe $a=\bot$ or $a=1$. Notably, TA implies Read
+Committed isolation: observing all effects of a transaction implicitly
+requires observing the final (committed) effects of a transaction as
+well.
 
@@
 unavailable. The inability to prevent Lost Update or Write Skew means
-that Repeatable Read, Conflict Serializability, and One-Copy
-Serializability guarantees are unavailable.
+that Repeatable Read and One-Copy Serializability guarantees are
+unavailable.
 
@@
 Figure~\ref{fig:hatcompared}. Many models are simultaneously
-achievable: for example, a HAT system can provide monotonic reads and
-transactional atomicity or transactional atomicity and read your
-writes and, for maximum semantic richness, can provide all of
-transactional atomicity, predicate cut isolation, monotonic reads,
-monotonic writes, and writes follow reads. An unavailable system might
-provide PRAM and Serializability, as advocated by Salem and
-Daudjee~\cite{daudjee-session}.
+achievable, but we find several particularly compelling. If we combine
+all sticky-HAT guarantees, we have transactional, causal snapshot
+reads. If we combine TA and P-CI, we have transactional snapshot
+reads. We can achieve RC, MR, and RYW by simply sticking clients to
+servers. And we can also combine unavailable models---for example, an
+unavailable system might provide PRAM and One-Copy
+Serializability~\cite{daudjee-session}.
 
@@
 subsumes all other models, while considering the (large) power set of
-all compatible models (e.g., the diagram depicts 144 possible highly available
+all compatible models (e.g., the diagram depicts 96 possible highly available
 combinations) hints at the vast expanse of consistency models found in
@@
 Snapshot Isolation (SI)\lostupdate, Repeatable Read
-(RR)\lostupdate\rwskew, Conflict Serializability
-(CSR)\lostupdate\rwskew, One-Copy Serializability
+(RR)\lostupdate\rwskew, One-Copy Serializability
 (1SR)\lostupdate\rwskew, Recency\linearizable, Safe\linearizable,
@@
 
- \node[draw=none,fill=none] (ici) at (0, 0) {I-CI};
- \node[draw=none,fill=none] (pci) at (1.1, 1.3) {P-CI};
+ \node[draw=none,fill=none] (ici) at (1.2, 0) {I-CI};
+ \node[draw=none,fill=none] (pci) at (1.65, 1.2) {P-CI};
  \node[draw=none,fill=none] (rc) at (-1.2, .8) {RC};
@@
 
- \node[draw=none,fill=none] (ta) at (1.2, 0) {TA};
+ \node[draw=none,fill=none] (ta) at (-.2, 1.3) {TA};
  \node[draw=none,fill=none] (mr) at (2.4, 0) {MR};
@@
  \node[noha] (cs) at (-1.2, 1.8) {CS};
- \node[noha] (rr) at (-.8, 2.7) {RR};
+ \node[noha] (rr) at (-.2, 2.7) {RR};
  \node[noha] (si) at (2.2, 2.4) {SI};
- \node[noha] (s) at (0, 3.5) {CSR};
- \node[noha] (sr) at (2.2, 3.6) {1SR};
- \node[noha] (ssr) at (4.95, 3.7) {Strong-1SR};
+ \node[noha] (1sr) at (1.2, 3.2) {1SR};
+ \node[noha] (ssr) at (3.6, 3.6) {Strong-1SR};
 
@@
  \draw [->, red] (linearizable) -- (ssr);
- \draw [->, red] (sr) -- (ssr);
+ \draw [->, red] (1sr) -- (ssr);
 
  \draw [->] (ru) -- (rc);
+ \draw [->] (rc) -- (ta);
  \draw [->] (ici) -- (pci);
@@
  \draw [->, red] (ici) -- (rr);
- \draw [->, red] (rr) -- (s);
- \draw [->, red] (si) -- (s);
+ \draw [->, red] (rr) -- (1sr);
+ \draw [->, red] (si) -- (1sr);
  \draw [->, red] (ta) -- (si);
  \draw [->, red] (ta) -- (rr);
- \draw [->, red] (s) -- (sr);
- \draw [->, red] (causal) -- (sr);
+ \draw [->, red] (causal) -- (ssr);
 

### 2013-03-29 writeup/hat-vldb.tex
draftwatermark
Another pass at writing
more writing

@@
 \documentclass{vldb}
-\usepackage{times, graphicx, url, xcolor, multicol, minipage, subfig, tikz}
+\usepackage{times, graphicx, url, xcolor, multicol, minipage, subfig, tikz, draftwatermarkbottom, draftwatermarktop}
 \usepackage{balance}  % for  \balance command ON LAST PAGE  (only there!)
 
+\renewcommand{\ttdefault}{fi4}
+
 \usetikzlibrary{positioning, snakes, shapes}
+
+\newenvironment{myitemize}
+{
+
+   \vspace{0mm}
+    \begin{list}{$\bullet$}{\leftmargin=1em}
+        \setlength{\topsep}{0em}
+        \setlength{\parskip}{0pt}
+        \setlength{\partopsep}{0pt}
+        \setlength{\parsep}{0pt}         
+        \setlength{\itemsep}{.25em} 
+        \setlength{\itemindent}{0em}
+}
+{
+    \end{list} 
+    \vspace{-.5em}
+}
+
+\newenvironment{myenumerate}
+{
+
+   \vspace{0mm}
+   \newcounter{qcounter}
+    \begin{list}{\arabic{qcounter}.~}{\usecounter{qcounter}\leftmargin=1em}
+        \setlength{\topsep}{0em}
+        \setlength{\parskip}{0pt}
+        \setlength{\partopsep}{0pt}
+        \setlength{\parsep}{0pt}         
+        \setlength{\itemsep}{.25em} 
+        \setlength{\itemindent}{0em}
+}
+{
+    \end{list} 
+    \vspace{-.5em}
+}
 
@@
 \title{Highly Available Transactions: Virtues and Limitations}
-
+{\author{Peter Bailis{\fontsize{12}{14}$^\dagger$}, Aaron Davidson{\fontsize{12}{14}$^\dagger$}, Alan Fekete{\fontsize{12}{14}$^\diamond$}, Ali Ghodsi{\fontsize{12}{14}$^{\dagger,\ddagger}$}, Joseph M.
+Hellerstein{\fontsize{12}{14}$^\dagger$}, Ion Stoica{\fontsize{12}{14}$^\dagger$}\\{{\fontsize{12}{14}$^\dagger$}\hspace{.5mm}UC Berkeley\hspace{4mm}{\fontsize{12}{14}$^\diamond$}\hspace{.5mm}University of Sydney\hspace{4mm}{\fontsize{12}{14}$^\ddagger$}\hspace{.5mm}KTH/Royal Institute of Technology}}}
 \maketitle
@@
 
-foo~\cite{adya}
-
 % The following two commands are all you need in the

### 2013-03-29 writeup/relatedwork.tex
text editing pass

@@
 
-VI. Related Work (.75pg)
+Although we are not aware of any related work specifically addressing
+what properties are available in a highly available system, there are
+several systems and related research results upon which we draw in
+this work.
 
-* Probably do a good job with this one, in addition to the first part of Section V. We can talk at length about the relationship between distributed systems and databases.
+Weak consistency and high availability have been well
+studied. Serializability has long been known to be
+unachievable~\cite{davidson-survey} and Brewer's CAP Theorem has
+attracted considerable attention~\cite{gilbert-cap}. The research has
+begun to offer hints regarding the connection between ``weak
+consistency'' and low latency~\cite{abadi-pacelc}, while several
+studies~\cite{adya, ansicritique} examine weak isolation guarantees,
+grounded in a single-server context. The connection between weak
+consistency models in a distributed setting~\cite{calm, crdt} and high
+availability is relatively unexplored. With the exception of a
+preliminary workshop paper on this work~\cite{hat-hotos}---upon which
+this work builds, providing impossibility results, a full
+taxonomization, experimental and application analysis (basically all
+but Section 4.1)---we believe this paper is the first to explore this
+area.
 
-1.) High Availability and CAP (giving the DB folks due credit here!)
-2.) Highly Available Distributed Systems
-3.) Weak Isolation (there's a lot here, pay plenty of lip service to Adya)
-4.) Available/unavailable systems
-	--Stress that, with the advent of Spanner, transactions are coming back; HATS represents new design space.
-(Can even say "Spanner borrowed traditional database techniques; would benefit from HAT approach (c.f., 100ms geo-replicated latencies)"; also talk about optimization for read-only transactions).
+There has been a recent resurgence of interest in distributed
+multi-object semantics, both in academia~\cite{kraska-s3, granola,
+  gstore, mdcc, redblue, cops, eiger, walter,calvin, swift} and
+industry~\cite{megastore, orleans, spanner}. We can roughly categorize
+related work into two categories: unavailable systems with stronger
+semantics than HATs and highly available systems with weaker
+semantics.
+
+Unavailable, strongly consistent databases have a long tradition. As
+discussed in Section~\ref{sec:motivation}, classic ACID databases
+provide strong semantics but their lock-based and traditional
+multi-versioned implementations are unavailable in the presence of
+partitions~\cite{bernstein-concurrency, bernstein-book,
+  gray-isolation}. Similar semantics are common in recent
+work. Notably, Google's Spanner provides externally serializable
+transactions. While Spanner fits Google's read-mostly workload well,
+read/write transactions use two-phase locking (incurring at least one
+round trip time over wide-area networks), are unavailable to clients
+in a non-majority partition, and cannot read their
+writes~\cite{spanner}. These latency and availability penalties are
+fundamental to providing strong consistency. For users willing to
+tolerate these costs, Spanner, or similar strongly consistent,
+unavailable systems---including Calvin~\cite{calvin},
+G-Store~\cite{gstore}, Gemini~\cite{redblue}, Granola~\cite{granola},
+HBase~\cite{hbase}, MDCC~\cite{mdcc}, Megastore~\cite{megastore},
+Orleans~\cite{orleans}, and Walter~\cite{walter}---are a reasonable
+choice. With HATs, we seek an alternative set of semantics that are
+still useful but do not violate requirements for high availability or
+low latency.

### 2013-03-29 writeup/hat-analysis.tex
finishing pdb edit pass

@@
 HAT systems---providing transactions with transactional availability
-with high availablity ($1$-available) or sticky high availability
+with high availability ($1$-available) or sticky high availability
 ($1$-sticky-available)---offer substantial latency and availability
@@
 read-your-writes, PRAM, and causal consistency) and which cannot
-(e.g., preventing Lost Update and Write Skew).
+(e.g., preventing Lost Update and Write Skew).  We present a full
+summary of these results in Section~\ref{sec:hat-summary}.
 
@@
 straightforward and have been introduced---albeit with greater brevity
-and discussion elided---in preliminary prior work~\cite{hat-hotos}. In
-our examples, we exclusively consider read and write operations. We
-denote a write of value $v$ to data item $d$ as $w_d(v)$ and a read
-from data item $d$ returning $v$ as $r_d(v)$ and assume that all data
-items have the null value, $\bot$, at database initialization. Unless
-otherwise specified, all example transactions commit.
+and discussion elided---in a preliminary workshop
+paper~\cite{hat-hotos}. In our examples, we exclusively consider read
+and write operations. We denote a write of value $v$ to data item $d$
+as $w_d(v)$ and a read from data item $d$ returning $v$ as $r_d(v)$
+and assume that all data items have the null value, $\bot$, at
+database initialization. Unless otherwise specified, all example
+transactions commit.
 
@@
 \end{align*}
-The difference between this total order on transactions and the total
-order required by serializability is that the Read Uncommitted total
-order is completely arbitrary (i.e., an order must exist), whereas a
-serializable total order must be equivalent to a serial execution. It
-is easily achieved via applying the same logical timestamp to each
-update in a transaction.
+We will strengthen this property in later models, which will prevent,
+for instance, reading $a=2, b=1$ at any time (not just
+``eventually''). The difference between this total order on
+transactions and the total order required by serializability is that
+the Read Uncommitted total order is completely arbitrary (i.e., an
+order must exist), whereas a serializable total order must be
+equivalent to a serial execution. It is easily achieved via applying
+the same logical timestamp to each update in a transaction and
+applying a ``last writer wins'' conflict reconciliation policy at each
+replica.
 
@@
 Repeatable Read lock-based implementation provides substantially
-richer guarantees than those that are promised by the ANSI SQL
+richer guarantees than those that are required by the ANSI SQL
 specification~\cite{gray-isolation}. Gray~\cite{gray-isolation},
@@
 of each data item that it did not itself produce (preventing ``Fuzzy
-Read,'' or P2). In the example below, $T_3$'s must see $a=1$:
+Read,'' or P2). In the example below, $T_3$ must read $a=1$:
 \begin{align*}
@@
 reads must obey additional ordering constraints. In fact, distributed
-systems describe a ``consistent snapshot'' across a set of related
-events as a \textit{cut} across a set of participants or, in our case,
-data items. Accordingly, to capture the notion that the transaction
-should read from a (non-changing) consistent cut over data items (and
-to disambiguate from the aforementioned stronger Repeatable Read
-properties), we call ANSI SQL Repeatable Read (preventing P2)
-\textbf{Cut Isolation}. It is possible to satisfy cut isolation with
-high availability by caching appropriate versions for a transaction's
-reads or, alternatively, using multi-versioning on the server and
-ensuring that each transaction's successive reads return the same
-version of each data item.
+systems often describe a ``consistent snapshot'' across a set of
+related events as a \textit{cut} across a set of participants or, in
+our case, data items. Accordingly, to capture the notion that the
+transaction should read from a (non-changing) consistent cut over data
+items (and to disambiguate from the aforementioned stronger Repeatable
+Read properties), we call ANSI SQL ``Repeatable Read'' (i.e.,
+preventing P2) \textbf{Cut Isolation}. It is possible to satisfy cut
+isolation with high availability by caching appropriate versions for
+each of a transaction's reads or, alternatively, using
+multi-versioning on each server and ensuring that each of a
+transaction's successive reads return the same version of each data
+item.
 
@@
 different data~\cite{gray-isolation}. We call predicate-based cut
-isolation, which prevents Phantoms, \textit{Phantom Cut
+isolation, which prevents Phantoms, \textit{Predicate Cut
   Isolation}.\footnote{Oracle provides an isolation level called
-  ``Statement Level Read Consistency''; this is analogous to Phantom
+  ``Statement Level Read Consistency''; this is analogous to Predicate
   Cut Isolation at the level of a single operation within a
@@
 Transactional atomicity (TA) is core to ACID guarantees. Although, at
-least by the acronym, it is not an ``isolation'' property,
+least by the ACID acronym, it is not an ``isolation'' property,
 transactional atomicity restricts transactions' ability to view the
@@
 is a strictly stronger guarantee than Read Committed isolation: in
-Read Committed, we can read a subset of committed writes whereas, with
-transactional atomicity, we must read all of them. Together with cut
-isolation, transactional atomicity prevents Read Skew anomalies
-(Berenson et al.'s A5A~\cite{ansicritique}). As an example of
-TA, $T_2$ must observe $b=c=1$ (or later versions
-for each key):
+Read Committed, we can read a subset of another transaction's
+committed writes whereas, with TA, we must read all or none of
+them. Together with item cut isolation, TA prevents Read Skew
+anomalies (Berenson et al.'s A5A~\cite{ansicritique}). As an example
+of TA, $T_2$ must observe $b=c=1$ (or later versions for each key):
 \vspace{-.5em}
@@
 \end{align*}
-$T_2$ can also observe $a=\bot$ or $a=1$. Notably, TA implies Read
-Committed isolation: observing all effects of a transaction implicitly
-requires observing the final (committed) effects of a transaction as
-well.
+$T_2$ can also observe $a=\bot$, $a=1$, or a later version of
+$a$. Notably, TA requires Read Committed isolation: observing all
+effects of a transaction implicitly requires observing the final
+(committed) effects of a transaction as well.
 
 Perhaps perplexingly, discussions of TA are absent from existing
-discussions of weak ACID. This is perhaps again due to the single-node
+discussions of weak isolation. This is perhaps again due to the single-node
 context in which prior work was developed: on a single server (or a
@@
 
-So far, our models have not considered what happens across
-transactions: we have not guaranteed any ordering or continuity
-between transactions (other than that there exists some arbitrary
-ordering). However, in the distributed systems context, many useful
-\textit{safety} guarantees span multiple application actions.
-\textit{Session guarantees} are used to describe guarantees across
+Our models have not yet considered interactions across transactions:
+we have not guaranteed any ordering or continuity between transactions
+(other than that there exists some arbitrary ordering). However, in
+the distributed systems context, many useful \textit{safety}
+guarantees span multiple application actions. In particular,
+\textit{session guarantees} are used to describe guarantees across
 groups of operations (here, transactions) issued by the same logical
@@
 context that consists of logical state that should persist between
-operations. For example, all of a user's transactions submitted
-between ``log in'' and ``log out'' operations might form a session.
+transactions. For example, on a social networking site, all of a
+user's transactions submitted between ``log in'' and ``log out''
+operations might form a session.
 
@@
 
-\vspace{.5em}\noindent\textit{\textbf{Monotonic reads}} requires that,
+\vspace{.5em}\noindent\textit{{Monotonic reads}} requires that,
 with a session, subsequent reads to a given object ``never return any
@@
 
-\vspace{.5em}\noindent\textit{\textbf{Monotonic writes}} requires that
+\vspace{.5em}\noindent\textit{{Monotonic writes}} requires that
 each session's writes be serialized. Any order on transactions should
@@
 
-\vspace{.5em}\noindent\textit{\textbf{Writes Follow Reads}} requires
-that, if a session reads a write originating from transaction $T_1$
-then performs transaction $T_2$, then a session can only read $T_2$'s
-writes if it can also read $T_1$'s prior write (or overwritten values
-for $T_1$'s write). This requirement forms the basis for potential
-causality: writes follow reads obeys the ``reads-from'' relation that
-captures all events that influenced each
-transaction~\cite{causalmemory}. Any transaction order should respect
-the reads-from order (note that the total order on transactions
-precludes reads-from cycles).\vspace{.5em}
+\vspace{.5em}\noindent\textit{{Writes Follow Reads}} requires that, if
+a session observes an effect of transaction $T_2$ and subsequently
+commits transaction $T_2$, then another session can only observe
+effects of $T_2$ if it can also observe $T_1$'s (or later values for
+$T_1$'s effects). Any order on transactions should respect the
+reads-from order.\vspace{.5em}
 
@@
 The solution to this conundrum is to give up high availability in
-favor of sticky availability. Sticky availability permits two
+favor of sticky availability. Sticky availability permits three
 additional models, which we first define and then prove are
@@
 
-\vspace{.5em}\noindent\textit{\textbf{Read your writes}} requires
+\vspace{.5em}\noindent\textit{{Read your writes}} requires
 that whenever a client reads a given data item after updating it, the
@@
 
-\vspace{.5em}\noindent\textit{\textbf{PRAM}} (Pipelined Random Access
+\vspace{.5em}\noindent\textit{{PRAM}} (Pipelined Random Access
 Memory) provides the illusion of serializing each session's operations
@@
 
-\vspace{.5em}\noindent\textit{\textbf{Causal consistency}} is the combination of
-all of the session guarantees~\cite{sessiontocausal} (alternatively,
-PRAM and writes-follow-reads) and is also referred to as PL-2L
+\vspace{.5em}\noindent\textit{{Causal
+    consistency}}~\cite{causalmemory} is the combination of all of the
+session guarantees~\cite{sessiontocausal} (alternatively, PRAM with
+writes-follow-reads) and is also referred to as PL-2L
 isolation~\cite{adya}).\vspace{.5em}
@@
 network topology shifts and the client can only contact a different
-server which is partitioned from the rest of the cluster, then the
-client will not be able to read its own writes and the system will
+server which is partitioned from the server that executed $T_1$, then
+the client will be inable to read its own writes and the system will
 have to either stall indefinitely to allow the client to read her
 writes (violating transactional availability) or will have to
-sacrifice read your writes. Accordingly, read your writes requires
-stickiness, and, because causal consistency requires read your writes,
-PRAM and causal consistency also require stickiness. Read-your-writes
-is ``free'' in a sticky system, while the remaining causality and PRAM
+sacrifice read your writes guarantees. Accordingly, read your writes
+requires stickiness, and, because causal consistency and PRAM require
+read your writes, they also require stickiness. Read your writes is
+``free'' in a sticky system, while the remaining causality and PRAM
 guarantees can be accomplished with the prior algorithms.
@@
 this paper is not to sketch the entire space of consistency models
-that are achievable (see Section~\ref{sec:futurework}, however we
-specifically evaluate TPC-C transaction semantics under HAT
-consistency guarantees in Section~\ref{sec:evaluation}.
+that are achievable (see Section~\ref{sec:futurework}. We specifically
+evaluate TPC-C transaction semantics under HAT consistency guarantees
+in Section~\ref{sec:evaluation}.
 
 \vspace{.5em}\noindent{\textbf{Durability}} As we briefly discussed in
-Section~\ref{sec:availability}, a client requiring durability
-of surviving $F$ server faults requires at least ($F+1$)-availability.
+Section~\ref{sec:availability}, a client requiring that its
+transactions' effects survive $F$ server faults requires at least
+($F+1$)-availability.
 
@@
 mutations to a data item, in the absence of partitions, all servers
-eventually agree on the value for each item. This is typically
+should eventually agree on the value for each item. This is typically
 accomplished by any number of anti-entropy protocols, which
 periodically update neighboring servers with the latest value for each
-data item. Establishing a final value is related to determining a
-total order on transaction updates, as in Read Uncommitted.
+data item~\cite{antientropy}. Establishing a final value is related to
+determining a total order on transaction updates, as in Read
+Uncommitted.
 
@@
 (Section~\ref{sec:futurework}), at this point, we have largely
-exhausted the range of achievable, previously defined (useful) HAT
+exhausted the range of achievable, previously defined (and useful) HAT
 semantics. Before summarizing our possibility results, we will present
@@
 Skew---and therefore providing Snapshot Isolation, Repeatable Read,
-and serializability---requires unavailability.
+and one-copy serializability---requires unavailability.
 
-In the words of Berenson et al., \textbf{Lost Update} occurs when one
+In the words of Berenson et al., \textit{Lost Update} occurs when one
 transaction $T1$ reads a given data item, a second transaction $T2$
@@
 If $T_1$ reads $a=1$ but $T_2$'s write to $x$ precedes $T_1$'s write
-operation, then the database will end up with $a=3$---a state that
-could not have resulted in a serial execution, arising from $T_2$'s
+operation, then the database will end up with $a=3$, a state that
+could not have resulted in a serial execution due to from $T_2$'s
 ``Lost Update.''
@@
 The final state of the database will converge to an execution that
-could not have resulted from a serial excecution. To prevent this from
-happening, either $T_1$ or $T_2$ should not have committed. Each
-client's respective server might try to detect that another write
-occurred, but this requires knowing the version of the latest write to
-$x$. In this example, this reduces to a requirement for
-linearizability, which is, via Gilbert and Lynch's proof of the CAP
-Theorem, provably unachievable with high
-availabilty~\cite{gilbert-cap}.
+could not have resulted from a serial excecution.\footnote{In this
+  example, we assume that, as is standard in modern databases,
+  replicas accept values as they are written (i.e., register
+  semantics). This particular example could be made serializable via
+  the use of commutative updates (Section~\ref{sec:eval}) but, as we
+  show in the extended version of the paper~\cite{hat-tr}, the problem
+  persists in the general case.} To prevent this from happening,
+  either $T_1$ or $T_2$ should not have committed. Each client's
+  respective server might try to detect that another write occurred,
+  but this requires knowing the version of the latest write to $x$. In
+  our example, this reduces to a requirement for linearizability,
+  which is, via Gilbert and Lynch's proof of the CAP Theorem, provably
+  unachievable with high availabilty~\cite{gilbert-cap}.
 
@@
 TPC-C~\cite{snapshot-serializable}---but, as a generalization of Lost
-Update, is also unavailable.
+Update, it is also unavailable to HATs.
 
@@
 
-Data storage engines make various recency guarantees on reads of data
+Data storage systems make various recency guarantees on reads of data
 items. As we have discussed, one of the most famous is
@@
 one-copy serializability and linearizability is called \textit{strong
-  (or strict) one-copy serializability} or \textit{externally
-  consistent transactions} (e.g., Google's
-Spanner~\cite{spanner})~\cite{adya}. It is also common, particularly
-in systems that allow reading from masters and slaves, to provide a
-guarantee such as ``read a version that is no more than five seconds
-out of date'' or similar. Unfortunately, an indefinitely long
-partition can cause violation of any recency bound, so recency bounds
-are unavailable to HAT systems.
+  (or strict) one-copy serializability}~\cite{adya} (e.g., Google's
+Spanner~\cite{spanner}). It is also common, particularly in systems
+that allow reading from masters and slaves, to provide a guarantee
+such as ``read a version that is no more than five seconds out of
+date'' or similar. Unfortunately, an indefinitely long partition can
+force a system to violate any recency bound, so recency bounds are
+unavailable to HAT systems.
 
@@
 We summarize our results in Table~\ref{table:hatcompared}. A wide
-range of isolation levels are achievable in a highly available system,
-including transactional atomicity, cut isolation, and several session
-guarantees. With sticky availability, a system can achieve PRAM
-consistency, read your writes guarantees, and causal
-consistency. However, many other prominent models, such as Snapshot
-Isolation, One-Copy Serializability, and Strict Serializability cannot
-be achieved due to the inability to prevent Lost Update and Write Skew phenomena.
+range of isolation levels are achievable in a HAT systems, including
+transactional atomicity, cut isolation, and several session
+guarantees. With sticky availability, a system can achieve read your
+writes guarantees and PRAM and causal consistency. However, many other
+prominent models, such as Snapshot Isolation, One-Copy
+Serializability, and Strict Serializability cannot be achieved due to
+the inability to prevent Lost Update and Write Skew phenomena.
 
 We illustrate the hierarchy of available, sticky available, and
-unavailble consistency models we have discussed in
+unavailable consistency models we have discussed in
 Figure~\ref{fig:hatcompared}. Many models are simultaneously
@@
 large subset of the prominent ACID and distributed consistency
-moels. Additional models of read/write transactions that we have
+models. Additional read/write transaction semantics that we have
 omitted should be easily classifiable based on the available
@@
 
-In this section, we discuss several subleties we have mostly elided up
-to this point, specifically addressing model composition,
-transactional atomicity versus linearizability, and stickiness
-requirements.
+In this section, we discuss several subleties in our results,
+specifically addressing model composition, transactional atomicity
+versus linearizability, and stickiness requirements.
 
@@
 of keys before each transaction begins to execute. Using client-side
-caching can alleviate some of these challenges, but then the system
-becomes sticky high available.
+caching can alleviate some of these challenges~\cite{bolton, swift},
+but then the system becomes sticky high available.
 
@@
 $O(N)$ storage per write~\cite{charron-bost}. This means that, with
-$100K$ clients, each write might be accompanied by $100K$
-timestamps per vector. This is difficult to scale. By compromising on
+$100K$ clients, each write might be accompanied by $100K$ timestamps
+per vector. This is difficult to scale. By compromising on
 availability (e.g., treating a datacenter as a linearizable cluster),
 this overhead can be reduced~\cite{cops, eiger}, but it is much
-cheaper to provide, say, read your writes, compared to causality.
+cheaper to provide, say, read your writes, than full causal
+consistency.
 
@@
 key on multiple servers are made visible to all readers at once. The
-difference is two-old. First, in linearizable (and safe and regular)
+difference is two-fold. First, in linearizable (and safe and regular)
 systems, writes are made visible to all clients \textit{immediately}
@@
 several properties like monotonic reads in a highly available system
-by waiting until all servers have seen a write. However, this incurs
-severe visibility penalties---new writes will not become visible to
-clients in the presence of partitions. A client that does not want to
-guarantee read-your-writes (due to the sticky availability
-requirement) may still wish to read other clients' writes with
-timeliness. Accordingly, a system may forgo high availability in order
-to provide better visibility. We also note that a system may wish to
-provide stickiness in the absence of partitions, then, in the presence
-of partitions, ``lose'' session guarantees by having clients reconnect
-to an available replica. This latter strategy provides session
-guarantees in the absence of partitions but, to provide availability,
-sacrifices them in the event of partitions.
+by waiting to reveal a write until all servers have seen it and its
+relevant dependencies. However, this incurs severe visibility
+penalties---new writes will not become visible to clients in the
+presence of partitions. A client that does not want to guarantee
+read-your-writes (due to the sticky availability requirement) may
+still wish to read other clients' writes with timeliness. Accordingly,
+a system may forgo high availability in order to provide better
+visibility. We also note that a system may wish to provide stickiness
+in the absence of partitions, but, in the presence of partitions,
+``lose'' session guarantees by having clients reconnect to an
+available replica. This latter strategy provides session guarantees in
+the absence of partitions but, to provide availability, sacrifices
+them in the event of partitions.

### 2013-03-29 writeup/intro.tex
finishing pdb edit pass

@@
 simplicity of coordination protocols have led many recent distributed
-database designers to renounce several guarantees provided by
+database designers to renounce several properties provided by
 traditional database systems~\cite{bigtable, pnuts, dynamo}. In
 particular, many modern systems forgo providing ACID guarantees, which
-often provide single-system image programmability but suffer from
-unavailability and often expensive coordination
+often provide single-system image programmability but traditionally
+suffer from unavailability and often expensive coordination
 costs~\cite{davidson-survey}. In particular, the July 2000
@@
 presence of network partitions, proved a rallying call for a
-generation of scalable but semantically weak systems. Over a decade
-later, we observe a dearth of highly available systems providing
-general-purpose multi-operation \textit{transactions} over multiple
-data items with useful semantics.
+generation of scalable but semantically weak system designs. Over a
+decade later, we observe a dearth of highly available systems
+providing general-purpose multi-operation \textit{transactions} over
+multiple data items with useful semantics.
 
@@
 transactions, is not achievable with high availability for general
-read-write transactions~\cite{davidson-survey}, databases have a long
-tradition of providing weaker isolation and consistency
+read-write transactions~\cite{davidson-survey}, database systems have
+a long tradition of providing weaker isolation and consistency
 guarantees~\cite{adya, ansicritique, gray-isolation}. Due to their
@@
 available systems. This raises the question: which of these semantics
-can we provide while remaining highly available?
+can be provided while remaining highly available?
 
@@
   ansicritique, gray-isolation} as well as several techniques for
-providing distributed
-serializability~\cite{bernstein-concurrency}. However, the current
+providing distributed serializability~\cite{bernstein-concurrency,
+  spanner, granola, daudjee-session, calvin}. However, the current
 literature does not provide guidance as to the operation of weakly
@@
 are achievable with high availability and low latency, albeit with
-some cost to achievable semantics. While our preliminary results
-indicate that some models are achievable~\cite{hat-hotos}, the
-literature lacks a deep understanding of what models are available,
-which are not, and when this matters.
+some cost to achievable semantics. While our prior, preliminary
+results indicated that some models are achievable~\cite{hat-hotos},
+the literature lacks a deep understanding of what models are
+available, which are not, and when this distinction matters.
 
@@
 guarantees, this work allows developers to weigh the costs and
-benefits of relying on these guarantees in their applications. While
-several guarantees, like serializability and Snapshot Isolation, are
+benefits of relying on particular guarantees in their applications. While
+several models, like serializability and Snapshot Isolation, are
 unachievable due to update conflict avoidance (preventing Lost Update
@@
 practice? We attempt to objectively study the virtues and limitations
-of both classic and HAT systems by performing an extensive survey of
-practitioner accounts and research literature, performing experimental
-analyses on modern cloud infrastructure, and analyzing representative
-applications for their semantic requirements. While our results are
-not definitive for all applications, they suggest that HATs offer a
-one to three order of magnitude latency decrease compared to
-traditional protocols and provide acceptable semantics for a wide
-range of programs, particularly those with monotonic logic and
-commutative updates~\cite{calm, blooml, crdt}. HAT systems can also
-provide arbitrary foreign key constraints on multi-item updates and
-sometimes provide uniqeness guarantees. However, particularly for
+of both classic and HAT systems by surveying practitioner accounts and
+research literature, performing experimental analyses on modern cloud
+infrastructure, and analyzing representative applications for their
+semantic requirements. While our results are not definitive for all
+applications, they suggest that HATs offer a one to three order of
+magnitude latency decrease compared to traditional protocols and can
+provide acceptable semantics for a wide range of programs,
+particularly those with monotonic logic and commutative
+updates~\cite{calm, blooml, crdt}. HAT systems can also enforce
+arbitrary foreign key constraints on multi-item updates and sometimes
+provide limited uniqueness guarantees. However, particularly for
 programs with non-monotonic logic, HATs may fall short, requiring more
-coordination-intensive mechanisms. Concisely, we attempt to understand
-\textit{what benefits do Highly Available Transactions offer and what
-  is their cost?}
+coordination-intensive, unavailable mechanisms. Concisely, we attempt
+to understand \textit{what benefits do Highly Available Transactions
+  offer and what is their cost?}
 
 In this paper, we make the following contributions:
-
-\begin{itemize}
+\begin{myitemize}
 \item We model high availability in a transactional environment,
@@
   case-study of an existing transactional application, and
-  experimentally evaluate a HAT database prototype.
-\end{itemize}
+  briefly evaluate a HAT database prototype.
+\end{myitemize}

### 2013-03-29 writeup/model.tex
finishing pdb edit pass

@@
 able to contact a (set of) server(s) in a system, they are guaranteed
-a response as the (set of) server(s) does not need to communicate with
-any others. If servers are partitioned from one another, they do not
-need to stall in order to provide a ``safe'' query response. This lack
-of fast-path coordination also means that a highly available system
-also provides low latency~\cite{abadi-pacelc}. However, to properly
-describe whether a \textit{transactional} system is highly available,
-we need both a way to describe what servers a client must contact as
-well as what kinds of responses a server can provide, especially given
-the possibility of aborts.
+a response because the (set of) server(s) will not need to communicate
+with any others. If servers are partitioned from one another, they do
+not need to stall in order to provide a ``safe'' query response. This
+lack of fast-path coordination also means that a highly available
+system also provides low latency~\cite{abadi-pacelc}; in a wide-area
+setting, this means that clients need not need to wait for
+cross-datacenter communication. However, to properly describe whether
+a \textit{transactional} system is highly available, we need both a
+way to describe what servers a client must contact as well as what
+kinds of responses a server can provide, especially given the
+possibility of aborts.
 
@@
 distributed database, designated servers might handle operations on
-different data items;\footnote{There is a further distinction between a
-  fully replicated system, in which all servers are replicas for all
-  data items, and a partially replicated system, in which at least one
-  server acts as a replica for only a subset of all data items. For
-  generality, and, for applicability to many modern ``sharded'' or
-  ``partitioned'' data storage systems like Dynamo, we consider
-  partially replicated systems.} a server that can handle an
-operation for a given data item is called a \textit{replica} for that
-item. This is a useful definition: if a client can contact at least
-one replica for an item on which it wishes to perform an operation,
-then the replica will provide a response to the client. However, there
-are several other cases we should consider.
+different data items;\footnote{There is a further distinction between
+  a \textit{fully replicated} system, in which all servers are
+  replicas for all data items, and a \textit{partially replicated}
+  system, in which at least one server acts as a replica for only a
+  subset of all data items. For generality, and, for applicability to
+  many modern ``sharded'' or ``partitioned'' data storage systems like
+  Dynamo, we consider partially replicated systems.} a server that can
+handle an operation for a given data item is called a \textit{replica}
+for that item. This is a useful definition: if a client can contact at
+least one replica for an item on which it wishes to perform an
+operation, then the replica will provide a response to the
+client. However, there are several other cases we should consider.
 
 First, clients may wish to contact multiple servers. For example, to
-ensure that the effects of operations will persist in the face of
+ensure that the effects of operations will persist in the event of
 server failures, clients may wish to wait until operations reach more
-than one server before they consider operations completed. However,
-according to our definition of high availability, server-fault
-tolerance is unavailable. To allow flexibility in the number of
-servers required for an operation, we propose
+than one server before they consider the operations
+completed. However, according to our definition of high availability,
+server-fault tolerance is unavailable. To allow flexibility in the
+number of participating servers required for an operation, we propose
 \textit{$K$-availability}: if a client can contact $K$ servers, then
@@
 latency and may experience more unsuccessful operations in the
-presence of network and server failures.  Additionally, as is commmon
-in standard consensus algorithms like Paxos, clients may want to
-contact a majority of $N$ servers in a system. This \textit{majority
+presence of network and server failures.  As is commmon in standard
+consensus algorithms like Paxos, clients may want to contact a
+majority of $N$ servers in a system. This \textit{majority
   availability} (a special case of $K$-availability with $K=\lceil
@@
 unavailability. We say that a system provides \textit{sticky
-  availability} if a client's subsequent operations are executed
-against a server that has observed all effects of its prior
-operations, then it eventually receives a response, even in the
-presence of indefinitely long partitions. As above, it may be useful
-for a client to achieve stickiness \textit{and} contact many replicas
-(i.e., for fault tolerance), so we generalize sticky availability to
-\textit{$K$-sticky availability}, which provides sticky availability
-with $K$ distinct servers. A client may choose to become $1$-sticky
-available by acting as a server itself; for example, a client might
-cache its reads and writes. However, a single client cannot achieve
-$K$-sticky availability with $K>1$ on its own. Moreover, ``sticking''
-all clients to a designated set of servers (i.e., master availability)
-is a special case of sticky availability and is less available than
-the equivalent non-mastered sticky availability.
+  availability} if, whenever a client's operation is executed against
+a server that has observed all of its prior operations, it eventually
+receives a response, even in the presence of indefinitely long
+partitions. As above, it may be useful for a client to achieve
+stickiness \textit{and} contact many replicas (i.e., for fault
+tolerance), so we generalize sticky availability to \textit{$K$-sticky
+  availability}, which provides sticky availability with $K$ distinct
+servers. A client may choose to become $1$-sticky available by acting
+as a server itself; for example, a client might cache its reads and
+writes. However, a single client cannot achieve $K$-sticky
+availability with $K>1$ without contacting other replicas. Moreover, ``sticking'' all clients
+to a designated set of servers (i.e., master availability) is a
+special case of sticky availability and is less available than the
+equivalent non-mastered sticky availability.
 
@@
 availability. This is standard in distributed systems literature
-(e.g., atomic, regular, and safe register models all concern single
-objects~\cite{herlihy-art}), yet the database literature largely
-focuses on transactions: groups of multiple operations over multiple
-objects. Accordingly, by itself, replica availability is insufficient
-to describe availability guarantees for transactions. Additionally,
-given the choice of \textit{commit} and \textit{abort}
-responses---which signal transaction success or failure to a
-client---we must be careful in how we define availability.
+(e.g., linearizable, regular, and safe register models all concern
+single objects~\cite{herlihy-art}), yet the database literature
+largely focuses on transactions: groups of multiple operations over
+multiple objects. Accordingly, by itself, replica availability is
+insufficient to describe availability guarantees for
+transactions. Additionally, given the choice of \textit{commit} and
+\textit{abort} responses---which signal transaction success or failure
+to a client---we must be careful in how we define availability.
 
@@
 may result in lower availability than a non-transactional
-requirement. Additionally, given the possiblity of aborts, we need to
-ensure useful forward progress. A system can trivially guarantee
-clients a response by always aborting their
+requirement. Additionally, given the possibility of aborts, we need to
+ensure that the system makes useful forward progress: a system can
+trivially guarantee clients a response by always aborting all
 transactions~\cite{transaction-liveness}. However, this is an
 unsatisfactory system because nothing good (transaction commit) ever
-happens; instead, we require a \textit{liveness} property. We cannot
-guarantee that every transaction will commit---transactions may choose
-to abort themselves---but we need to make sure that the system will
-not indefinitely abort transactions on its own volition. We call a
-transaction abort due to a transaction's own choosing (e.g., as part
-of the transaction itself or due to a would-be integrity constraint
-violation) an \textit{internal abort} and an abort due to system
-implementation an \textit{external abort}.
+happens; instead, we should require a \textit{liveness} property. We
+cannot guarantee that every transaction will commit---transactions may
+choose to abort themselves---but we need to make sure that the system
+will not indefinitely abort transactions on its own volition. We call
+a transaction abort due to a transaction's own choosing (e.g., as an
+operation of the transaction itself or due to a would-be declared
+integrity constraint violation) an \textit{internal abort} and an
+abort due to system implementation an \textit{external abort}.
 
@@
 given replica availability for every data item in a transaction, the
-transaction eventually commits or internally aborts~\cite{hat-hotos}. For example,
-system will violate transactional $2$-availability if client that can
-contact two servers for each of data items in its transaction does not
+transaction eventually commits or internally
+aborts~\cite{hat-hotos}. For example, system will violate
+transactional $2$-availability if a client that can contact two
+servers for each of the data items in its transaction does not
 eventually commit in the absence of internal aborts.

### 2013-03-29 thebes-code/hat/server/src/main/java/edu/berkeley/thebes/hat/server/dependencies/DependencyResolver.java
Add some debug messages

@@ -76,7 +76,11 @@ public class DependencyResolver {
         PendingWrite newPendingWrite = new PendingWrite(key, value);
         
         TransactionQueue transQueue = pendingTransactionsMap.get(version);
-        transQueue.add(newPendingWrite);
+        try {
+            transQueue.add(newPendingWrite);
+        } catch (Exception e) {
+            logger.error("Error on version " + version + ": ", e);
+        }
         
         if (transQueue.shouldAnnounceTransactionReady()) {
             router.announceTransactionReady(version, transQueue.replicaIndicesInvolved);
@@ -100,6 +104,7 @@ public class DependencyResolver {
         }
         
         if (shouldCommit) {
+            logger.debug("Committing via unresolved: " + version + " / " + transQueue.numReplicasInvolved + " / " + newPendingWrite.getReplicaIndicesInvolved().size());
             commit(transQueue);
         }
         
@@ -136,6 +141,7 @@ public class DependencyResolver {
     public void ackTransactionPending(Version transactionId) throws TException {
         TransactionQueue transactionQueue = pendingTransactionsMap.get(transactionId);
         if (transactionQueue != null && transactionQueue.serverAcked()) {
+            logger.debug("Committing via ack: " + transactionId + " / " + transactionQueue.numReplicasInvolved);
             commit(transactionQueue);
             return;
         }

### 2013-03-29 writeup/conclusion.tex

@@
 database designs. While the CAP Theorem has had a large impact on
-distributed data storage design, our knowledge of database systems has
-not kept pace. The key distributed systems challenge of designing
-distributed algorithms and architectures for partial failure and
-network asynchrony is largely unaddressed in the literature on
-transaction processing systems. We believe that, in order to fully
-realize the potential of HAT systems, further rearchitecture and study
+highly available distributed data storage design, our knowledge of
+database systems has not kept pace. The key distributed systems
+challenge of designing distributed algorithms and architectures for
+partial failure and network asynchrony is largely unaddressed in the
+literature on transaction processing systems. To fully realizing the
+potential of HAT systems will require further rearchitecture and study
 of existing database designs as well as better understanding of
 application semantics and their relationship to distributed
-consistency will be necessary.
+consistency.

### 2013-03-29 writeup/evaluation.tex

@@
-\section{Analysis}
+\section{Application and System Analysis}
 \label{sec:evaluation}
@@
 \item We examine the properties required by a realistic OLTP
-  application patterned on TPC-C.
+  application patterned on the TPC-C benchmark.
 \item We perform a brief experimental evaluation of HAT versus non-HAT
@@
 two-phase locking for a transaction of length $T$ requires $T$
-\texttt{lock} operations and at least one unlock operation.  Each of
-these operations requires coordination, either with other database
-replicas or with a lock service. If the coordination mechanism for
-locking is unavailable, transactions cannot make progress. Similarly,
-optimistic concurrency control requires contacting a validation
-service, while deterministic transaction
+\texttt{lock} operations and at least one \texttt{unlock} operation.
+Each of these operations requires coordination, either with other
+database replicas or with a lock service. If the coordination
+mechanism for locking is unavailable, transactions cannot
+commit. Similarly, optimistic concurrency control requires contacting
+a validation service, while deterministic transaction
 scheduling~\cite{deterministic-scheduling} requires a
-scheduler. Multi-version concurrency control fares somewhat better but
-must still be checked for write conflicts.
+scheduler. Serializability under multi-version concurrency control
+fares somewhat better but still requires checking for update
+conflicts.
 
@@
 proposal~\cite{gray-isolation} do not degrade gracefully in the
-presence of partial failure. However, lock-based protocols \textit{do}
-provide recency guarantees. While multi-versioned storage systems
-allow for a variety of transactional guarantees, we have not seen
-proposals for truly weak isolation (e.g., non-``tentative update''
-schemes) in this context. Chan and Gray's read-only transactions
-provide read-only transactions with item-cut isolation, causal
-consistency, and transactional atomicity but are unavailable in the
-presence of coordinator failure~\cite{readonly}, similar to read-only
-and write-only transactions more recently proposed by
-Eiger~\cite{eiger}.  Bolt-on causal consistency~\cite{bolton},
-Swift~\cite{swift}, Brantner's S3 database~\cite{kraska-s3}, and
-Bayou~\cite{sessionguarantees} can all provide variants of the above
-model with high availability, but none provide this HAT functionality
-without modification. Similarly, the MDCC~\cite{mdcc} protocol offers
-Read Committed isolation with Lost Update avoidance but is similarly
+presence of partial failure. (Note, however, that lock-based protocols
+\textit{do} offer the benefit of recency guarantees.) While
+multi-versioned storage systems allow for a variety of transactional
+guarantees, we have not seen proposals for truly weak isolation (e.g.,
+non-``tentative update'' schemes) in this context. Chan and Gray's
+read-only transactions provide read-only transactions with item-cut
+isolation, causal consistency, and transactional atomicity but are
+unavailable in the presence of coordinator failure~\cite{readonly},
+similar to read-only and write-only transactions more recently
+proposed by Eiger~\cite{eiger}.  Bolt-on causal
+consistency~\cite{bolton}, Brantner's S3 database~\cite{kraska-s3},
+and Bayou~\cite{sessionguarantees} can all provide variants of the
+above model with high availability, but none provide this HAT
+functionality without substantial modification; Swift~\cite{swift} is
+closest to providing full HAT operation in a sticky
+context. Similarly, the MDCC~\cite{mdcc} protocol offers Read
+Committed isolation with Lost Update avoidance but is similarly
 unavailable due to its reliance on preventing write conflicts.
@@
 inability to prevent Lost Update, Write Skew, and provide recency
-bounds. In this Section, we attempt to understand when these
+bounds. In this section, we attempt to understand when these
 guarantees matter both abstractly and by the use of a representative
-customer order transaction application patterned off of TPC-C.
+customer order transaction application patterned off of the TPC-C
+benchmark~\cite{tpcc}.
 
-\vspace{.5em}\noindent\textbf{Commutativity and Monotonicity} Recent work on
-the CALM Theorem~\cite{calm}, Lattice-based state
+\vspace{.5em}\noindent\textbf{Commutativity and Monotonicity} Recent
+work on the CALM Theorem~\cite{calm}, Lattice-based state
 update~\cite{blooml}, and Commutative and Replicated Data
 Types~\cite{crdt} demonstrates that, if updates logically commute,
-then they can be performed in different orders at different locations
-in a distributed system. Accordingly, as long as all writes are
-delivered to all replicas, then a system executing monotonic logic
-with commutative operators should not suffer from application-level
-consistency anomalies as a result of Lost Update or Write Skew
-anomalies. However, applications with non-monotonic state mutation
-will not, in general, be able to maintain application-level
+then they can often be safely performed in different orders at
+different locations in a distributed system. Accordingly, as long as
+all writes are delivered to all replicas, then a system executing
+monotonic logic with commutative operators may not suffer from
+application-level consistency anomalies as a result of Lost Update or
+Write Skew anomalies. However, applications with non-monotonic state
+mutation will not, in general, be able to maintain application-level
 consistency constraints. Similarly, applications requiring write
@@
 we consider an example application: the TPC-C benchmark. Our goal is
-\textit{not} to provide a performance analysis, as any HAT
+\textit{not} to provide a performance analysis, as a HAT
 configuration will be non-compliant. Rather, we examine the five
@@
 real-world warehouse, including sales, payments, and deliveries. Two
-transactions---Order-Status and Stock-Level---are read-only and can be
-executed safely under HATs. Clients may read stale data, but this does
-not violate TPC-C requirements and clients will read their writes if
-they are sticky-available. Another transaction type, Payment, updates
-running balances for warehouses, districts, and customer records as
-well as providing an audit trail. The transaction is increment-only,
-so all balance increase operations can commute, and TA allows the
-maintainence of any foreign-key constraints (e.g.,
-\texttt{UPDATE/DELETE CASCADE}).
+transactions---\textit{Order-Status} and \textit{Stock-Level}---are
+read-only and can be executed safely under HATs. Clients may read
+stale data, but this does not violate TPC-C requirements and clients
+will read their writes if they are sticky-available. Another
+transaction type, \textit{Payment}, updates running balances for
+warehouses, districts, and customer records as well as providing an
+audit trail. The transaction is increment-only, so all balance
+increase operations can commute, and TA allows the maintainence of any
+foreign-key constraints (e.g., \texttt{UPDATE/DELETE CASCADE}).
 
 While three out of five transactions have been easily achievable with
-HATs, the remaining two transactions---New-Order and Delivery---are
-not so simple. The New-Order transaction places an order for a
-variable quantity of data items, updating warehouse stock as
-needed. It selects a sales district, assigns the order an ID number,
-adjusts the remaining warehouse stock, and writes a placeholder entry
-for the pending order. The Delivery transaction represents the
-fulfillment of a New-Order: it deletes the order from the pending
-list, updates the customer's balance, updates the order's carrier ID
-and delivery time, and updates the customer balance.
+HATs, the remaining two transactions---\textit{New-Order} and
+\textit{Delivery}---are not so simple. The New-Order transaction
+places an order for a variable quantity of data items, updating
+warehouse stock as needed. It selects a sales district, assigns the
+order an ID number, adjusts the remaining warehouse stock, and writes
+a placeholder entry for the pending order. The Delivery transaction
+represents the fulfillment of a New-Order: it deletes the order from
+the pending list, updates the customer's balance, updates the order's
+carrier ID and delivery time, and updates the customer balance.
 
 There are three challenges that only allow HATs to place non-compliant
-New-Orders and to fail on Payments. First, we must assign each new
-order an ID number. We can assign a unique number by, say,
+New-Orders and to fail on Payments. First, New-Order requires a unique
+ID number for each order. We can create a unique number by, say,
 concatenating the client ID and a timestamp. However, the TPC-C
-specification requires order numbers to be assigned sequentially,
-which requires Lost Update. Accordingly, HATs cannot provide compliant
-TPC-C execution but can maintain uniqueness constraints. Second, the
-New-Order transaction decrements inventory counts: what if the count
-becomes negative? Fortunately, TPC-C New-Order restocks the inventory
-count (increments by 91) if it would become negative as the result of
-placing an order. This means that, even in the presence of concurrent
-New-Orders, an item's stock will never fall below zero. This is TPC-C
-compliant, but a HAT system might end up with substantially more stock
-than in a single-site implementation. Third, delivery events are
-non-monotonic---deleting a pending New-Order transaction---and should
-be idempotent in order to avoid billing a customer twice; this implies
-a need to prevent Lost Update. We can avoid this issue by moving the
-non-monotonicity to the real world---the carrier that picks up the
-package for an order can ensure that she is the only carrier who has
-done so---but cannot provide a correct execution with HATs alone.
+specification requires order numbers to be \textit{sequentially}
+assigned, which requires Lost Update. Accordingly, HATs cannot provide
+compliant TPC-C execution but can maintain uniqueness
+constraints. Second, the New-Order transaction decrements inventory
+counts: what if the count becomes negative? Fortunately, TPC-C
+New-Order restocks each item's inventory count (increments by 91) if
+it would become negative as the result of placing an order. This means
+that, even in the presence of concurrent New-Orders, an item's stock
+will never fall below zero. This is TPC-C compliant, but a HAT system
+might end up with substantially more stock than in a single-site
+implementation. Third, delivery events are non-monotonic---deleting a
+pending New-Order transaction---and should be idempotent in order to
+avoid billing a customer twice; this implies a need to prevent Lost
+Update. We can avoid this issue by moving the non-monotonicity to the
+real world---the carrier that picks up the package for an order can
+ensure that she is the only carrier who has done so---but cannot
+provide a correct execution with HATs alone.
 
@@
 subordinate sales districts. This integrity constraint spans two
-tables but, given the ability to update both tables atomically via
-transactional atomicity, can be easily maintained. Consistency
+tables but, given the ability to update rows in both tables atomically
+via transactional atomicity, can be easily maintained. Consistency
 Conditions 4 through 12 (3.3.2.4-12) can similarly be satisfied by
@@
 expect that, especially for read-dominated workloads found in many
-online services, HAT guarantees will suffice for many
-applications. Finally, while TPC-C is not subject to write skew
+online services, HAT guarantees will suffice for many application
+transactions. Finally, while TPC-C is not subject to multi-key
 anomalies, we note that many TPC-E isolation tests, such as one that
 requires simultaneously modifying a product description and its image,
-are satisfied by HATs.
+are also passable using HATs.
 

### 2013-03-29 writeup/futurework.tex

@@
 
-\textbf{What is the space of consistency and isolation properties?} In
-this work, we have taxonomized several well-defined and
+\textbf{What are the limits of HAT systems, and how are they defined?}
+In this work, we have taxonomized several well-defined and
 well-documented consistency properties. This is useful as it

### 2013-03-29 writeup/hat-vldb.tex

@@
 transactional functionality, which provides strong semantic guarantees
-on groups of multiple operations over multiple data items. In this
+for groups of multiple operations over multiple data items. In this
 work, we consider the problem of providing transactional guarantees

### 2013-03-29 writeup/motivation.tex

@@
 need to account for a possibility of partial system failure. However,
-both of these assumptions are no longer valid in a distributed
-setting. To extend Jim Gray's storage latency analogy, if memory is
-1.5 hours from the CPU and disk is 2 years away, then, especially
-across wide-area networks (WANs), remote server accesses can exceed 30
-years, with the added concern that messages can be lost in transit or
-delayed indefinitely by network
-partitions~\cite{gray-rules}. Accordingly, ignoring network failures
-and latency can ``cause big trouble and painful learning
-experiences''~\cite{fallacies-deutsch}.
+in a distributed setting, both of these assumptions are no longer
+valid. To extend Jim Gray's storage latency analogy, if memory is 1.5
+hours from the CPU and disk is 2 years away, then, especially across
+wide-area networks (WANs), remote server accesses can exceed 30 years,
+with the added concern that messages can be lost in transit or delayed
+indefinitely by network partitions~\cite{gray-rules}. Accordingly,
+ignoring network failures and latency can ``cause big trouble and
+painful learning experiences''~\cite{fallacies-deutsch}.
 
-In this section, we quantify the impact of network partitions and
+In this section, we quantify the effect of network partitions and
 latency in several real-world deployments. While the importance of
@@
 evidence---both anecdotal and experimental---that partitions occur and
-network latencies are non-negligible.
+network latencies are non-negligible. We subsequently discuss the
+impact of these findings on modern database operation and motivate the
+case for HATs.
 
@@
 simultaneously maintaining a correct ``global'' view of system state
-while simultaneously providing operation to all of its
+while simultaneously providing online operation to all of its
 clients~\cite{davidson-survey}. For networks that do not partition,
 maintaining semantic guarantees and available operation is
-simple~\cite{stonebraker2010errors}. However, given the
+straightforward~\cite{stonebraker2010errors}. However, given the
 \textit{possibility} of partitions, a system must be prepared to
@@
 partition management is an important consideration service operators
-today. As an example of an actual system rearchitecture, less than one
-year after its announcement, Yahoo!'s PNUTS developers explicitly
-added a support for weaker, highly available operation. The engineers
-explained that ``strict adherence [to strong consistency] leads to
-difficult situations under network partitioning or server
-failures...in many circumstances, applications need a relaxed
-approach''~\cite{pnuts-update}.
+today. As an example of an actual system rearchitecture resulting from
+partitions, less than one year after its announcement, Yahoo!'s PNUTS
+developers explicitly added support for weaker, highly available
+operation. The engineers explained that ``strict adherence [to strong
+  consistency] leads to difficult situations under network
+partitioning or server failures...in many circumstances, applications
+need a relaxed approach''~\cite{pnuts-update}.
 
@@
 day for all tickets~\cite{turner2012failure}. Other studies confirm
-these results, showing median time between failures of approximately
-3000 seconds with a median time to repair between 2 and 1000
-seconds~\cite{ip-backbone-failures} as well as frequent path routing
-failures on the Internet~\cite{labovitz-failures}. Isolating,
+these results, showing median time between failures over a WAN network
+of approximately 3000 seconds with a median time to repair between 2
+and 1000 seconds~\cite{ip-backbone-failures} as well as frequent path
+routing failures on the Internet~\cite{labovitz-failures}. Isolating,
 quantifying, and designing for these network failures is an area of
@@
 form of system partition that has also been studied at
-scale~\cite{google-availability})---indicate that partitions
+scale~\cite{google-availability})---indicate that network partitions
 \textit{do} occur within and across modern datacenters. We do not
@@
 
-\subsection{Latency: An Omnipresent Concern}
+\subsection{Latency and Planet Earth}
 
@@
 cloud. We measured one week of ping times between all seven EC2
-regions, across three ``availability zones'', or closely co-located
-datacenters, and within a single ``availabilty zone,'' at a
-granularity of 1s (dataset to be released).
+geographic ``regions'', across three ``availability zones'' (closely
+co-located datacenters), and within a single ``availabilty zone''
+(datacenter), at a granularity of 1s (dataset to be released).
 
@@
 communicating from S\~{a}o Paulo to Singapore is lower-bounded by
-speed-of-light transit RTT of 106.7 ms but, on average, incurs a 362.8
-ms RTT (95th percentile: 649ms). As shown in Figure~\ref{fig:rtt}, the
-distribution of latencies varies between links, but the trend---in
-line with other recent studies~\cite{mdcc, redblue}---is clear: remote
-communication has a substantial cost.
+speed-of-light transit RTT of 106.7ms but, on average, incurs a
+362.8ms RTT (95th percentile: 649ms). As shown in
+Figure~\ref{fig:rtt}, the distribution of latencies varies between
+links, but the trend---in line with other recent studies~\cite{mdcc,
+  redblue}---is clear: remote communication has a substantial cost.
 
@@
 Mitigating the effects of partitions and network latency is a
-difficult task, requiring systems designers to make fundamental
-trade-offs. In this section, we briefly discuss known semantic limits
-for systems wishing to maintain high availability and their impact on
-modern database systems. While the gold standard of ACID databases is
-unavailable, we show that many database systems operate under much
-weaker models, motivating the study of these models in a highly
-available context in the remainder of this paper.
+difficult task, requiring systems designers to choose between
+fundamental trade-offs. In this section, we briefly discuss known
+semantic limits for systems wishing to maintain high availability and
+their impact on modern database systems. While the gold standard of
+ACID semantics is unavailable, we show that many database systems
+operate under much weaker models, motivating further study of these
+models in a highly available context in the remainder of this paper.
 
-One of the most popularly celebrated impossibility results for
+One of the most popularly celebrated impossibility results in
 distributed systems is Brewer's CAP Theorem, which, as formally proven
 by Gilbert and Lynch, states that it is impossible to maintain
-linearizability (``C,'' or ``consistency''---albeit poorly named), or
-the ability to read the last completed write, and replica availability
-(``A'') in the presence of network partitions
-(``P'')~\cite{gilbert-cap}. Brewer's Theorem has roots in decades of
-distributed database designs~\cite{davidson-survey} and has had a
-substantial impact on recent large-scale distributed systems design:
-while CAP is narrowly scoped to the impossibility of building highly
-available linearizable systems linearizability, it is often
-interpreted to preclude a larger set of guarantees, including ACID, or
-transactional atomicity, consistency, isolation and durability
-guarantees, as evidenced by the lack of highly available systems
-providing them.
+linearizability , or the ability to read the last completed write,
+(``C,'' or ``consistency''---albeit poorly named) and replica
+availability, or a guaranteed response, (``A'') in the presence of
+network partitions (``P'')~\cite{gilbert-cap}. Brewer's Theorem has
+roots in decades of distributed database
+designs~\cite{davidson-survey} and has had a substantial impact on
+recent large-scale distributed systems design: while CAP is narrowly
+scoped to the impossibility of building highly available linearizable
+systems, it is often interpreted to preclude a larger set of
+guarantees, including ACID (transactional atomicity, consistency,
+isolation and durability guarantees), as evidenced by the lack of
+highly available systems providing them.
 
-Database designers and researchers have long realized that
+Database researchers and designers have long realized that
 serializability---the gold standard of ACID isolation, which
@@
 penalties associated with ensuring equivalence with a serial execution
-can be severe and manifest themselves in the form of concurrency
-control contention (and, subsequently, performance degradation,
-scalability limitations, and, often, external aborts like
+can be severe and manifest themselves in the form of decreased
+concurrency (and, subsequently, performance degradation, scalability
+limitations, and, often, external aborts like
 deadlocks)~\cite{gray-isolation}. Instead, databases offer a host of
@@
 serializability, but the benefits of these models can often outweigh
-costs of possible consistency anomalies that might arise
-from their use.
+costs of possible consistency anomalies that might arise from their
+use.
 
 To understand the prevalence of these weak isolation models, we
-recently surveyed, we surveyed the default and maximum isolation
-models provided by 18 relational databases, often claiming to provide
-``ACID'' or ``NewSQL'' functionality~\cite{hat-hotos}. As shown in
-Table~\ref{table:existing}, only three out of 18 databases provides
-serializability by default, and at least eight do not provide
+recently surveyed the default and maximum isolation models provided by
+18 relational databases, often claiming to provide ``ACID'' or
+``NewSQL'' functionality~\cite{hat-hotos}. As shown in
+Table~\ref{table:existing}, only three out of 18 databases provided
+serializability by default, and at least eight did not provide
 serializability as an option at all. This is particularly surprising
@@
 transactional models are frequently used, our inability to provide
-serializability in arbitrary HATs is non-fatal for practical
+serializability in arbitrary HATs appears non-fatal for practical
 applications. If application writers and database vendors have already
@@
 application inconsistencies, then, in a highly available environment
-that prevents serializability entirely, similar decisions may be
-appropriate.
+that prohibits serializability, similar decisions may be tenable.
 
@@
 The primary challenge in providing weak isolation in a HAT context is
-that, unfortunately, it is unclear \textit{which} of these isolation
-guarantees can be provided with high availability. The existing
-algorithms for providing weak isolation are typically designed for a
-single-node context and are, to the best of our knowledge, unavailable
-due to a reliance on concurrency control mechanisms like locking that
-are not resilient to partial failure. Morever, we are not aware of any
-prior literature that provides guidance as to the relationship between
-weak isolation and high availability: prior work has examined the
+that, unfortunately, it is unclear \textit{which} of these ACID
+(particularly isolation) guarantees can be provided with high
+availability. The existing algorithms for providing weak isolation are
+typically designed for a single-node context and are, to the best of
+our knowledge, unavailable due to a reliance on concurrency control
+mechanisms like locking that are not resilient to partial failure
+(Section~\ref{sec:evaluation}). Morever, we are not aware of any prior
+literature that provides guidance as to the relationship between weak
+isolation and high availability: prior work has examined the
 relationship between serializability and high
@@
 the database literature is a possible contributing factor to the
-often-apparent attitude that general-purpose transactional semantics
-are either too expensive to provide in many modern distributed
-databases---except under special circumstances, such as providing
-operations over groups of co-located data items.
+often-apparent attitude~\cite{hn, foundation-article} that
+general-purpose transactional semantics are either too expensive to
+provide in many modern distributed databases---except under special
+circumstances, such as providing operations over groups of co-located
+data items~\cite{entitygroup}.

### 2013-03-29 writeup/relatedwork.tex

@@
   gstore, mdcc, redblue, cops, eiger, walter,calvin, swift} and
-industry~\cite{megastore, orleans, spanner}. We can roughly categorize
-related work into two categories: unavailable systems with stronger
-semantics than HATs and highly available systems with weaker
-semantics.
-
-Unavailable, strongly consistent databases have a long tradition. As
-discussed in Section~\ref{sec:motivation}, classic ACID databases
-provide strong semantics but their lock-based and traditional
-multi-versioned implementations are unavailable in the presence of
+industry~\cite{megastore, orleans, spanner}. As discussed in
+Section~\ref{sec:motivation}, classic ACID databases provide strong
+semantics but their lock-based and traditional multi-versioned
+implementations are unavailable in the presence of
 partitions~\cite{bernstein-concurrency, bernstein-book,
@@
 still useful but do not violate requirements for high availability or
-low latency.
+low latency. Recent systems proposals such as Swift~\cite{swift},
+Eiger~\cite{eiger}, and Bolt-on Causal Consistency~\cite{bolton}
+provide transactional causal consistency guarantees with varying
+availability and represent a new class of (sticky) HAT systems.

### 2013-03-29 thebes-code/common/src/main/java/edu/berkeley/thebes/common/config/Config.java
Fix NPE in Clinet

@@ -342,4 +342,8 @@ public class Config {
             return ip;
         }
     }
+
+    public static int getNumClusters() {
+        return getClusterMap().size();
+    }
 }

### 2013-03-29 thebes-code/hat/client/src/main/java/edu/berkeley/thebes/hat/client/clustering/MasteredReplicaRouter.java

@@ -28,7 +28,7 @@ public class MasteredReplicaRouter extends ReplicaRouter {
         
         this.replicaAddressesByCluster = Maps.newHashMap();
         this.syncReplicasByCluster = Maps.newHashMap();
-        this.numClusters = Config.getSiblingServers().size();
+        this.numClusters = Config.getNumClusters();
         this.numNeighbors = Config.getServersInCluster().size();
         
         for (int i = 0; i < numClusters; i ++) {

### 2013-03-29 thebes-code/hat/server/src/main/java/edu/berkeley/thebes/hat/server/dependencies/PendingWrite.java

@@ -46,6 +46,9 @@ public class PendingWrite implements Comparable<PendingWrite> {
                 numKeysForThisReplica ++;
             }
         }
+        if (!replicaIndicesInvolved.contains(myIndex)) {
+            System.err.println("WARNINGG: " + replicaIndicesInvolved + " does not include me! " + myIndex + " (" + key + " / " + transactionKeys + ")");
+        }
         replicaIndicesInvolved = serversInTransaction;
     }
 

### 2013-03-29 thebes-code/common/src/main/java/edu/berkeley/thebes/common/config/Config.java
too tired to think of commit message

@@ -149,7 +149,7 @@ public class Config {
         return getOption(ConfigParameters.TWOPL_CLUSTER_CONFIG);
     }
 
-    private static List<ServerAddress> getServersInCluster(int clusterID) {
+    public static List<ServerAddress> getServersInCluster(int clusterID) {
         List<String> serverIPs = getClusterMap().get(clusterID);
         List<ServerAddress> servers = Lists.newArrayList();
         

### 2013-03-29 thebes-code/hat/client/src/main/java/edu/berkeley/thebes/hat/client/clustering/MasteredReplicaRouter.java

@@ -32,7 +32,7 @@ public class MasteredReplicaRouter extends ReplicaRouter {
         this.numNeighbors = Config.getServersInCluster().size();
         
         for (int i = 0; i < numClusters; i ++) {
-            List<ServerAddress> neighbors = Config.getServersInCluster();
+            List<ServerAddress> neighbors = Config.getServersInCluster(i+1);
             List<ReplicaService.Client> neighborClients = Lists.newArrayList();
             for (ServerAddress neighbor : neighbors) {
                 neighborClients.add(ThriftUtil.getReplicaServiceSyncClient(neighbor.getIP(),

### 2013-03-29 writeup/hat-vldb.tex
Joe pass

--- 2013-03-29 writeup/hat-vldb.tex
@@
 \begin{abstract}
-Given the presence of network latency and the possibility of partial
+To minimize network latency and handle the possibility of partial
 failure and network partitions, many modern distributed systems eschew
 transactional functionality, which provides strong semantic guarantees
-on groups of multiple operations over multiple data items. In this
-work, we consider the problem of providing transactional guarantees
-that are highly available, or do not suffer from unavailability during
-system partitions or typically incur high network latency.  We
-introduce a new taxonomy of highly available algorithms and analyze a
+for groups of multiple operations over multiple data items. In this
+work, we consider the problem of achieving transactional guarantees
+with high availability, or without suffering from unavailability
+during system partitions or incurring high network latency.  We
+introduce a new taxonomy of high availability and analyze a
 combination of ACID and distributed consistency guarantees to identify
@@
 benefit of providing high availability as well as the associated
-semantic costs.
+semantic compromises.
 \end{abstract}

### 2013-03-29 writeup/intro.tex
Joe pass

@@
 literature does not provide guidance as to the operation of weakly
-consistent transactional guarantees in a highly available distributed
+consistent transactional guarantees in a replicated
 context. Accordingly, in this work, we consider the problem of
@@
 disributed ``register'' semantics~\cite{herlihy-art}. By identifying
-the availability cost of a wide range of popular consistency
-guarantees, this work allows developers to weigh the costs and
-benefits of relying on particular guarantees in their applications. While
-several models, like serializability and Snapshot Isolation, are
-unachievable due to update conflict avoidance (preventing Lost Update
-and Write Skew phenomena), many other ACID guarantees are HAT
-compliant, such as providing Read Committed isolation, preventing
-Fuzzy Read phenomena, and ensuring atomicity of transactional
-updates. Several other guarantees, like reading one's writes, fall
-into a new availability class---``sticky'' high availability---which
-provides an intermediate ground between strict high availability and
-majority availability.
+the availability cost of a wide range of guarantees, this work allows
+developers to weigh the costs and benefits of relying on particular
+guarantees in their applications. While several models, like
+serializability and Snapshot Isolation, are unachievable due to update
+conflict avoidance (preventing Lost Update and Write Skew phenomena),
+many other ACID guarantees are HAT compliant, such as providing Read
+Committed isolation, preventing Fuzzy Read phenomena, and ensuring
+atomicity of transactional updates. Several other guarantees, like
+reading one's writes, fall into a new availability class---``sticky''
+high availability---which provides an intermediate ground between
+strict high availability and majority availability.
 
@@
 magnitude latency decrease compared to traditional protocols and can
-provide acceptable semantics for a wide range of programs,
-particularly those with monotonic logic and commutative
-updates~\cite{calm, blooml, crdt}. HAT systems can also enforce
-arbitrary foreign key constraints on multi-item updates and sometimes
-provide limited uniqueness guarantees. However, particularly for
-programs with non-monotonic logic, HATs may fall short, requiring more
+provide acceptable semantics for a wide range of programs, especially
+those with monotonic logic and commutative updates~\cite{calm, blooml,
+  crdt}. HAT systems can also enforce arbitrary foreign key
+constraints on multi-item updates and sometimes provide limited
+uniqueness guarantees. However, particularly for programs with
+non-monotonic logic, HATs may fall short, requiring more
 coordination-intensive, unavailable mechanisms. Concisely, we attempt

### 2013-03-29 writeup/model.tex
Joe pass

@@
 able to contact a (set of) server(s) in a system, they are guaranteed
-a response because the (set of) server(s) will not need to communicate
-with any others. If servers are partitioned from one another, they do
-not need to stall in order to provide a ``safe'' query response. This
-lack of fast-path coordination also means that a highly available
-system also provides low latency~\cite{abadi-pacelc}; in a wide-area
-setting, this means that clients need not need to wait for
-cross-datacenter communication. However, to properly describe whether
-a \textit{transactional} system is highly available, we need both a
-way to describe what servers a client must contact as well as what
-kinds of responses a server can provide, especially given the
-possibility of aborts.
+a response; this means servers will not need to communicate with any
+others. If servers are partitioned from one another, they do not need
+to stall in order to provide a ``safe'' query response. This lack of
+fast-path coordination also means that a highly available system also
+provides low latency~\cite{abadi-pacelc}; in a wide-area setting, this
+means that clients need not need to wait for cross-datacenter
+communication. However, to properly describe whether a
+\textit{transactional} system is highly available, we need both a way
+to describe what servers a client must contact as well as what kinds
+of responses a server can provide, especially given the possibility of
+aborts.
 
@@
 handle an operation for a given data item is called a \textit{replica}
-for that item. This is a useful definition: if a client can contact at
-least one replica for an item on which it wishes to perform an
-operation, then the replica will provide a response to the
-client. However, there are several other cases we should consider.
+for that item. While this standard model is useful, there are several
+other cases we should consider.
 
@@
   availability} (a special case of $K$-availability with $K=\lceil
-\frac{N}{2} \rceil$) is more available than full $N$-availability.
+\frac{N}{2} \rceil$) is ``more available'' than full $N$-availability.
 
@@
 servers~\cite{vogels-defs}. Simultaneously maintaining availability
-and stickiness requires fate-sharing between the client and its set of
+and stickiness entails fate-sharing between the client and its set of
 servers: if a client's sticky servers fail, or, equivalently, if the
@@
 to a designated set of servers (i.e., master availability) is a
-special case of sticky availability and is less available than the
+special case of sticky availability and is ``less available'' than the
 equivalent non-mastered sticky availability.
@@
 \textit{abort} responses---which signal transaction success or failure
-to a client---we must be careful in how we define availability.
+to a client---we must be careful in how we define transactional availability.
 
@@
 receive a response from at least one replica for each data item. This
-may result in lower availability than a non-transactional
+may result in ``lower availability'' than a non-transactional
 requirement. Additionally, given the possibility of aborts, we need to
@@
 unsatisfactory system because nothing good (transaction commit) ever
-happens; instead, we should require a \textit{liveness} property. We
+happens; we should require a \textit{liveness} property. We
 cannot guarantee that every transaction will commit---transactions may
@@
 transaction eventually commits or internally
-aborts~\cite{hat-hotos}. For example, system will violate
+aborts~\cite{hat-hotos}. For example, a system will violate
 transactional $2$-availability if a client that can contact two

### 2013-03-30 thebes-code/common/src/main/java/edu/berkeley/thebes/common/interfaces/IThebesClient.java
Adding commit/abort

@@ -12,7 +12,9 @@ public interface IThebesClient {
     public void open() throws TTransportException, ConfigurationException, IOException;
 
     public void beginTransaction() throws TException;
-    public boolean endTransaction() throws TException;
+
+    public void abortTransaction() throws TException;
+    public boolean commitTransaction() throws TException;
 
     public boolean put(String key, ByteBuffer value) throws TException;
     public ByteBuffer get(String key) throws TException;

### 2013-03-30 thebes-code/thebes-client/src/main/java/edu/berkeley/thebes/client/ThebesClient.java

@@ -52,8 +52,13 @@ public class ThebesClient implements IThebesClient {
     }
 
     @Override
-    public boolean endTransaction() throws TException {
-        return internalClient.endTransaction();
+    public boolean commitTransaction() throws TException {
+        return internalClient.commitTransaction();
+    }
+
+    @Override
+    public void abortTransaction() throws TException {
+        internalClient.abortTransaction();
     }
 
     @Override

### 2013-03-30 thebes-code/hat/client/src/main/java/edu/berkeley/thebes/hat/client/clustering/QuorumReplicaRouter.java
Quorum debugging

@@ -88,15 +88,20 @@ public class QuorumReplicaRouter extends ReplicaRouter {
     
     /** Performs the request by queueing N requests and waiting for Q responses. */
     public <E> E performRequest(String key, Request<E> request) {
+        int numSent = 0;
         int replicaIndex = RoutingHash.hashKey(key, numNeighbors);
         for (List<ServerAddress> replicasInCluster : replicaAddressesByCluster.values()) {
             ServerAddress replicaAddress = replicasInCluster.get(replicaIndex);
             ReplicaClient replica = replicaRequestQueues.get(replicaAddress);
             if (!replica.inUse.getAndSet(true)) {
+                numSent ++;
                 request.process(replica);
             }
         }
-        return request.getResponseWhenReady();
+        logger.debug("Sent " + numSent + " messages for key " + key);
+        E ret = request.getResponseWhenReady();
+        logger.debug("Received response " + ret + " for key " + key);
+        return ret;
     }
     
     private abstract class Request<E> {

### 2013-03-30 writeup/evaluation.tex

@@
 
-With a firm understanding of which of several ACID and distributed
-consistency properties are available and which are not, we begin to
-understand their implications for existing algorithms, applications,
-and deployments. Specifically:
+We now have an understanding of which guarantees are and are not
+HAT-compliant. In this section, we begin to understand the impact of
+this high availability on existing systems and applications as well as
+briefly demonstrate HAT algorithms on public cloud
+infrastructure. Specifically:
 
@@
 \vspace{.5em}\noindent\textbf{Serializablity} To establish a serial
-order on transactions, all existing algorithms for achieving
-serializability of general-purpose read-write transactions in a
-distributed setting~\cite{bernstein-book} may require at least one
-round trip time (RTT) before committing. As an example, traditional
-two-phase locking for a transaction of length $T$ requires $T$
-\texttt{lock} operations and at least one \texttt{unlock} operation.
-Each of these operations requires coordination, either with other
-database replicas or with a lock service. If the coordination
-mechanism for locking is unavailable, transactions cannot
-commit. Similarly, optimistic concurrency control requires contacting
-a validation service, while deterministic transaction
+order on transactions, algorithms for achieving serializability of
+general-purpose read-write transactions in a distributed
+setting~\cite{bernstein-book} may require at least one round trip time
+(RTT) before committing. As an example, traditional two-phase locking
+for a transaction of length $T$ requires $T$ \texttt{lock} operations
+and at least one \texttt{unlock} operation.  In a distributed
+environment, each of these operations requires coordination, either
+with other database replicas or with a lock service. If the
+coordination mechanism for locking is unavailable, transactions cannot
+commit. Similarly, optimistic concurrency control requires
+coordinating via a validation step, while deterministic transaction
 scheduling~\cite{deterministic-scheduling} requires a
 scheduler. Serializability under multi-version concurrency control
-fares somewhat better but still requires checking for update
-conflicts.
+also requires checking for update conflicts.
 
-Perhaps unsurprisingly, the reliance on a globally agreed total order
-necessitates a minimum of one round-trip to a designated master or
-coordination service for each of these classic algorithms.  This is
-intuitively a consequence of serializability's lack of high
+In summary, perhaps unsurprisingly, the reliance on a globally agreed
+total order necessitates a minimum of one round-trip to a designated
+master or coordination service for each of these classic algorithms.
+This is intuitively a consequence of serializability's lack of high
 availability~\cite{davidson-survey} (or, as we have shown, reliance on
@@
 
-\vspace{.5em}\noindent\textbf{Non-serializablity} Perhaps
-interestingly, many existing distributed implementations of weak
-isolation are not highly available. Lock-based proposals such as those
-used to provide weak isolation in Gray's original
-proposal~\cite{gray-isolation} do not degrade gracefully in the
-presence of partial failure. (Note, however, that lock-based protocols
-\textit{do} offer the benefit of recency guarantees.) While
-multi-versioned storage systems allow for a variety of transactional
-guarantees, we have not seen proposals for truly weak isolation (e.g.,
-non-``tentative update'' schemes) in this context. Chan and Gray's
-read-only transactions provide read-only transactions with item-cut
-isolation, causal consistency, and transactional atomicity but are
-unavailable in the presence of coordinator failure~\cite{readonly},
-similar to read-only and write-only transactions more recently
-proposed by Eiger~\cite{eiger}.  Bolt-on causal
-consistency~\cite{bolton}, Brantner's S3 database~\cite{kraska-s3},
-and Bayou~\cite{sessionguarantees} can all provide variants of the
-above model with high availability, but none provide this HAT
-functionality without substantial modification; Swift~\cite{swift} is
-closest to providing full HAT operation in a sticky
-context. Similarly, the MDCC~\cite{mdcc} protocol offers Read
-Committed isolation with Lost Update avoidance but is similarly
-unavailable due to its reliance on preventing write conflicts.
+\vspace{.5em}\noindent\textbf{Non-serializablity} Many existing
+distributed implementations of weak isolation are not highly
+available. Lock-based proposals such as those used to provide weak
+isolation in Gray's original proposal~\cite{gray-isolation} do not
+degrade gracefully in the presence of partial failure. (Note, however,
+that lock-based protocols \textit{do} offer the benefit of recency
+guarantees.) While multi-versioned storage systems allow for a variety
+of transactional guarantees, we have not seen proposals for truly weak
+isolation (e.g., non-``tentative update'' schemes) in this
+context. Chan and Gray's read-only transactions provide read-only
+transactions with item-cut isolation, causal consistency, and
+transactional atomicity (session PL-2L) but are unavailable in the
+presence of coordinator failure~\cite{readonly}, similar to read-only
+and write-only transactions more recently proposed by
+Eiger~\cite{eiger}.  Bolt-on causal consistency~\cite{bolton},
+Brantner's S3 database~\cite{kraska-s3}, and
+Bayou~\cite{sessionguarantees} can all provide variants of session
+PL-2L with high availability, but none provide this HAT functionality
+without substantial modification; Swift~\cite{swift} is closest to
+providing full HAT operation in a sticky context. Similarly, the
+MDCC~\cite{mdcc} protocol offers Read Committed isolation with Lost
+Update avoidance but is similarly unavailable due to its reliance on
+preventing write conflicts.
 
@@
 than serializability---including guarantees that are achievable with
-high availability---and not achieve high availability. We expect that
-further understanding the trade-offs between semantic strength,
-performance, and storage and metadata requirements will remain an area
-of active research for the foreseeable future.
+high availability---and still not achieve high availability.
 
@@
 Thus far, we have largely ignored the question of which applications
-require semantics which are unavailable to HATs. As we showed in
+require semantics that are unavailable to HATs. As we showed in
 Section~\ref{sec:hats}, the main cost of choosing HATs comes in the
@@
 mutation will not, in general, be able to maintain application-level
-consistency constraints. Similarly, applications requiring write
-visibility between clients should opt for unavailability instead of
-possibly missing a write visibility deadline.
+consistency constraints with HATs alone. Similarly, applications
+requiring write visibility between clients should opt for
+unavailability instead of possibly missing a write visibility
+deadline.
 
@@
 achievable in a HAT context. Surprisingly, four of five can be
-executed with HATs, while the fifth potentially requires unavailablity.
+executed with HATs, while the fifth may require unavailablity.
 
 TPC-C consists of five transactions, capturing operation of a
-real-world warehouse, including sales, payments, and deliveries. Two
+wholesale warehouse, including sales, payments, and deliveries. Two
 transactions---\textit{Order-Status} and \textit{Stock-Level}---are
@@
 specification requires order numbers to be \textit{sequentially}
-assigned, which requires Lost Update. Accordingly, HATs cannot provide
-compliant TPC-C execution but can maintain uniqueness
+assigned, which requires preventing Lost Update. Accordingly, HATs
+cannot provide compliant TPC-C execution but can maintain uniqueness
 constraints. Second, the New-Order transaction decrements inventory
@@
 ensure that she is the only carrier who has done so---but cannot
-provide a correct execution with HATs alone.
+provide a correct execution with HATs alone. However, according to
+distributed transaction architects~\cite{helland}, these compensatory
+actions are relatively common in real-world business processes.
 

### 2013-03-30 writeup/futurework.tex

@@
 a single availability mode, there are several interesting hybrids to
-consider.
+consider. For example, a system may provide ``best-effort'' stickiness
+(or serializability!) in the absence of partitions, but, in the
+presence of partitions, ``lose'' guarantees by having clients
+reconnect to an available replica. This strategy provides stronger
+guarantees in the absence of partitions but, to provide availability,
+sacrifices them in the event of partitions.

### 2013-03-30 writeup/motivation.tex

@@
 
-Peter Deutsch begins his classic list of ``Fallacies of Distributed
-Computing'' with two concerns fundamental to distributed database
-systems: ``\textit{1.)}  The network is reliable. \textit{2.)} Latency
-is zero''~\cite{fallacies-deutsch}. On a single server, communication
+Distributed databases face a host of challenges that their
+single-server counterparts are exempted from, largely stemming from
+network asynchrony and partial failure. These challenges mean that
+maintaining semantic guarantees has a cost, the exact price of which
+is a subject of considerable speculation in the database
+community~\cite{stonebraker2010errors}. In this section, we attempt to
+quantify the prevalence and cost of network partitions and latency in
+several real-world deployments. This data and a recent survey of
+existing database semantics motivates the study of highly available
+transaction systems.
+
+(TODO-PDB: keep this?) We will begin by studying the behavior of
+network partitions and latency. Peter Deutsch begins his classic list
+of ``Fallacies of Distributed Computing'' with two concerns
+fundamental to distributed database systems: ``\textit{1.)}  The
+network is reliable. \textit{2.)} Latency is
+zero''~\cite{fallacies-deutsch}. On a single server, communication
 channels are robust and relatively fast, and few single-server systems
@@
 in a distributed setting, both of these assumptions are no longer
-valid. To extend Jim Gray's storage latency analogy, if memory is 1.5
-hours from the CPU and disk is 2 years away, then, especially across
-wide-area networks (WANs), remote server accesses can exceed 30 years,
-with the added concern that messages can be lost in transit or delayed
-indefinitely by network partitions~\cite{gray-rules}. Accordingly,
+valid. To extend Jim Gray's storage latency analogy~\cite{gray-rules},
+if memory is 1.5 hours from the CPU and disk is 2 years away, then,
+especially across wide-area networks (WANs), remote server accesses
+can exceed 30 years, with the added concern that messages can be lost
+in transit or delayed indefinitely by network partitions. Accordingly,
 ignoring network failures and latency can ``cause big trouble and
@@
 
-In this section, we quantify the effect of network partitions and
-latency in several real-world deployments. While the importance of
-network behavior is a topic of considerable speculation within the
-database community~\cite{stonebraker2010errors}, there is mounting
-evidence---both anecdotal and experimental---that partitions occur and
-network latencies are non-negligible. We subsequently discuss the
-impact of these findings on modern database operation and motivate the
-case for HATs.
-
 \subsection{Partitions in the Real World}
 
-Network partitions---or, informally, communication link failures
-between participants in a distributed system---are challenging. A
-system facing network partitions faces the impossibility of
-simultaneously maintaining a correct ``global'' view of system state
-while simultaneously providing online operation to all of its
+Network partitions---informally, communication failures between
+participants in a distributed system---are challenging. A system
+facing network partitions faces the impossibility of simultaneously
+maintaining a correct ``global'' view of system state while
+simultaneously providing online operation to all of its
 clients~\cite{davidson-survey}. For networks that do not partition,
-maintaining semantic guarantees and available operation is
-straightforward~\cite{stonebraker2010errors}. However, given the
+maintaining semantic guarantees and providing ``always on'' operation
+is straightforward~\cite{stonebraker2010errors}. However, given the
 \textit{possibility} of partitions, a system must be prepared to
@@
 geographically distinct sites, the cost of communication increases to
-this upper bound.
-
-In real deployments, latencies are actually higher than the speed of
-light dictates due to routing, congestion, and processing times; if
-not, servers a kilometer apart would enjoy a 6.7 $\mu$s RTT. To
-illustrate the difference between intra-datacenter, inter-datacenter,
-and inter-planetary networks, we performed a measurement study of
-network behavior on Amazon's EC2, a widely-used public compute
-cloud. We measured one week of ping times between all seven EC2
-geographic ``regions'', across three ``availability zones'' (closely
-co-located datacenters), and within a single ``availabilty zone''
-(datacenter), at a granularity of 1s (dataset to be released).
+this lower bound. In real deployments, transit times are actually
+slower than the speed of light due to routing, congestion, and
+processing times. To illustrate the difference between
+intra-datacenter, inter-datacenter, and inter-planetary networks, we
+performed a measurement study of network behavior on Amazon's EC2, a
+widely-used public compute cloud. We measured one week of ping times
+between all seven EC2 geographic ``regions'', across three
+``availability zones'' (closely co-located datacenters), and within a
+single ``availabilty zone'' (datacenter), at a granularity of 1s
+(dataset to be released).
 
 We summarize the results of our network measurement study in
-Table~\ref{table:rtt}. On average, inter-datacenter communication (1c)
+Table~\ref{table:rtt}. On average, intra-datacenter communication (1c)
 is between 1.82 and 6.38 times faster than across geographically
@@
 
-Mitigating the effects of partitions and network latency is a
-difficult task, requiring systems designers to choose between
-fundamental trade-offs. In this section, we briefly discuss known
-semantic limits for systems wishing to maintain high availability and
-their impact on modern database systems. While the gold standard of
-ACID semantics is unavailable, we show that many database systems
-operate under much weaker models, motivating further study of these
-models in a highly available context in the remainder of this paper.
+Systems designers must make fundamental trade-offs between latency,
+system behavior during partitions, and achievable semantics. In this
+section, we briefly discuss known semantic limits for systems wishing
+to maintain high availability and their impact on modern database
+systems. We show that many database systems operate under weak model
+whose availability characteristics are unknown, motivating further
+study of these models in a highly available context in the remainder
+of this paper.
 
@@
 roots in decades of distributed database
-designs~\cite{davidson-survey} and has had a substantial impact on
+research~\cite{davidson-survey} and has had a substantial impact on
 recent large-scale distributed systems design: while CAP is narrowly
@@
 guarantees, including ACID (transactional atomicity, consistency,
-isolation and durability guarantees), as evidenced by the lack of
-highly available systems providing them.
+isolation and durability guarantees)~\cite{hn, foundation-article} as
+also evidenced by the lack of highly available systems providing them.
 
@@
 guarantees application-level consistency---is not achievable in a
-highly available system~\cite{davidson-survey}.  However, database
-systems offer a range of ACID properties besides
-serializability. Within a single-node database, the coordination
-penalties associated with ensuring equivalence with a serial execution
-can be severe and manifest themselves in the form of decreased
-concurrency (and, subsequently, performance degradation, scalability
-limitations, and, often, external aborts like
-deadlocks)~\cite{gray-isolation}. Instead, databases offer a host of
-so-called \textit{weak isolation} models that allow varying
+highly available system~\cite{davidson-survey}.  However, to increase
+concurrency, database systems offer a range of ACID properties weaker
+than serializability. Within a single-node database, the coordination
+penalties associated with serializability can be severe and manifest
+themselves in the form of decreased concurrency (and, subsequently,
+performance degradation, scalability limitations, and, often, external
+aborts like deadlocks)~\cite{gray-isolation}. Instead, databases offer
+a host of so-called \textit{weak isolation} models that allow varying
 restrictions on the space of schedules that are allowable by the
 system~\cite{adya}. None of these weak isolation models guarantees
-serializability, but the benefits of these models can often outweigh
-costs of possible consistency anomalies that might arise from their
-use.
+serializability, but, as we see below, the benefits of these models
+are often considered to outweigh costs of possible consistency
+anomalies that might arise from their use.
 
@@
 recently surveyed the default and maximum isolation models provided by
-18 relational databases, often claiming to provide ``ACID'' or
+18 databases, often claiming to provide ``ACID'' or
 ``NewSQL'' functionality~\cite{hat-hotos}. As shown in
@@
 
-The primary challenge in providing weak isolation in a HAT context is
-that, unfortunately, it is unclear \textit{which} of these ACID
-(particularly isolation) guarantees can be provided with high
-availability. The existing algorithms for providing weak isolation are
-typically designed for a single-node context and are, to the best of
-our knowledge, unavailable due to a reliance on concurrency control
-mechanisms like locking that are not resilient to partial failure
+It is unknown \textit{which} of these ACID (particularly isolation)
+guarantees can be provided with high availability. The existing
+algorithms for providing weak isolation are typically designed for a
+single-node context and are, to the best of our knowledge, unavailable
+due to a reliance on concurrency control mechanisms like locking that
+are not resilient to partial failure
 (Section~\ref{sec:evaluation}). Morever, we are not aware of any prior
@@
 single-server~\cite{adya, ansicritique} but never weak isolation and
-high availability taken together. We believe that this apparent gap in
-the database literature is a possible contributing factor to the
-often-apparent attitude~\cite{hn, foundation-article} that
-general-purpose transactional semantics are either too expensive to
-provide in many modern distributed databases---except under special
-circumstances, such as providing operations over groups of co-located
-data items~\cite{entitygroup}.
+high availability taken together.

### 2013-03-30 thebes-code/hat/client/src/main/java/edu/berkeley/thebes/hat/client/clustering/QuorumReplicaRouter.java
More debugging

@@ -78,11 +78,13 @@ public class QuorumReplicaRouter extends ReplicaRouter {
     
     @Override
     public boolean put(String key, DataItem value) throws TException {
+        logger.error("PUT for key " + key);
         return performRequest(key, new WriteRequest(key, value));
     }
 
     @Override
     public ThriftDataItem get(String key, Version requiredVersion) throws TException {
+        logger.error("GET for key " + key);
         return performRequest(key, new ReadRequest(key, requiredVersion));
     }
     
@@ -98,9 +100,9 @@ public class QuorumReplicaRouter extends ReplicaRouter {
                 request.process(replica);
             }
         }
-        logger.debug("Sent " + numSent + " messages for key " + key);
+        logger.error("Sent " + numSent + " messages for key " + key);
         E ret = request.getResponseWhenReady();
-        logger.debug("Received response " + ret + " for key " + key);
+        logger.error("Received response " + ret + " for key " + key);
         return ret;
     }
     

### 2013-03-30 thebes-code/hat/client/src/main/java/edu/berkeley/thebes/hat/client/clustering/ReplicaRouter.java
More debugg

@@ -2,6 +2,7 @@ package edu.berkeley.thebes.hat.client.clustering;
 
 import org.apache.thrift.TException;
 import org.apache.thrift.transport.TTransportException;
+import org.slf4j.LoggerFactory;
 
 import edu.berkeley.thebes.common.config.ConfigParameterTypes.RoutingMode;
 import edu.berkeley.thebes.common.data.DataItem;
@@ -13,6 +14,7 @@ import java.io.IOException;
 public abstract class ReplicaRouter {
     public static ReplicaRouter newInstance(RoutingMode routingMode)
             throws IOException, TTransportException {
+        LoggerFactory.getLogger(ReplicaRouter.class).error("HELLO REPLICA ROUTER??" + routingMode);
         switch (routingMode) {
         case NEAREST:
             return new NearestReplicaRouter();

### 2013-03-30 thebes-code/thebes-client/src/main/java/edu/berkeley/thebes/client/ThebesClient.java

@@ -4,11 +4,13 @@ import edu.berkeley.thebes.common.config.Config;
 import edu.berkeley.thebes.common.interfaces.IThebesClient;
 import edu.berkeley.thebes.common.log4j.Log4JConfig;
 import edu.berkeley.thebes.hat.client.ThebesHATClient;
+import edu.berkeley.thebes.hat.client.clustering.QuorumReplicaRouter;
 import edu.berkeley.thebes.twopl.client.ThebesTwoPLClient;
 import edu.berkeley.thebes.twopl.common.ThebesTwoPLTransactionClient;
 
 import org.apache.thrift.TException;
 import org.apache.thrift.transport.TTransportException;
+import org.slf4j.LoggerFactory;
 
 import javax.naming.ConfigurationException;
 import java.io.FileNotFoundException;
@@ -27,6 +29,7 @@ public class ThebesClient implements IThebesClient {
     public void open() throws TTransportException, ConfigurationException, IOException {
         Config.initializeClient();
         Log4JConfig.configureLog4J();
+        LoggerFactory.getLogger(ThebesClient.class).error("Hello from ThebesClient!");
 
         switch (Config.getThebesTxnMode()) {
         case HAT:

### 2013-03-30 thebes-code/hat/client/src/main/java/edu/berkeley/thebes/hat/client/clustering/QuorumReplicaRouter.java
Debugggy

@@ -158,6 +158,7 @@ public class QuorumReplicaRouter extends ReplicaRouter {
             
             @Override
             public void onComplete(put_call response) {
+                logger.error("Response for " + key + "!");
                 if (numAcks.incrementAndGet() > quorum) {
                     request.sendResponse(true);
                 }

### 2013-03-30 thebes-code/hat/client/src/main/java/edu/berkeley/thebes/hat/client/clustering/ReplicaRouter.java

@@ -14,7 +14,6 @@ import java.io.IOException;
 public abstract class ReplicaRouter {
     public static ReplicaRouter newInstance(RoutingMode routingMode)
             throws IOException, TTransportException {
-        LoggerFactory.getLogger(ReplicaRouter.class).error("HELLO REPLICA ROUTER??" + routingMode);
         switch (routingMode) {
         case NEAREST:
             return new NearestReplicaRouter();

### 2013-03-30 thebes-code/hat/client/src/main/java/edu/berkeley/thebes/hat/client/clustering/QuorumReplicaRouter.java
Debuggery

@@ -144,8 +144,11 @@ public class QuorumReplicaRouter extends ReplicaRouter {
         public void process(ReplicaClient replica) {
             PutCallback callback = new PutCallback(replica);
             try {
+                logger.error("Calling put...");
                 replica.client.put(key, value, callback);
+                logger.error("Called put...");
             } catch (TException e) {
+                logger.error("Exception happened!");
                 callback.onError(e);
             }
         }

### 2013-03-30 thebes-code/hat/client/src/main/java/edu/berkeley/thebes/hat/client/clustering/QuorumReplicaRouter.java
Build

@@ -146,6 +146,7 @@ public class QuorumReplicaRouter extends ReplicaRouter {
             try {
                 logger.error("Calling put...");
                 replica.client.put(key, value, callback);
+                logger.error(replica.client.hasTimeout() + " / " + replica.client.getTimeout());
                 logger.error("Called put...");
             } catch (TException e) {
                 logger.error("Exception happened!");

### 2013-03-30 thebes-code/hat/client/src/main/java/edu/berkeley/thebes/hat/client/clustering/QuorumReplicaRouter.java
sync replicarouter

@@ -45,37 +45,55 @@ public class QuorumReplicaRouter extends ReplicaRouter {
     private final int numClusters;
     private final int numNeighbors;
     private int quorum;
-    
+
     private final Map<ServerAddress, ReplicaClient> replicaRequestQueues = Maps.newHashMap();
 
     private class ReplicaClient {
-        private ReplicaService.AsyncClient client;
+        private Client client;
         private AtomicBoolean inUse;
+        BlockingQueue<Request<?>> requestBlockingQueue;
 
-        public ReplicaClient(ReplicaService.AsyncClient client) {
+        public ReplicaClient(Client client) {
             this.client = client;
             this.inUse = new AtomicBoolean(false);
+            requestBlockingQueue = Queues.newLinkedBlockingQueue();
+
+            new Thread(new Runnable() {
+                @Override
+                public void run() {
+                    while(true) {
+                        Request<?> request = Uninterruptibles.takeUninterruptibly(requestBlockingQueue);
+                        request.process(ReplicaClient.this);
+                    }
+                }
+            }).start();
+        }
+
+        public void executeRequest(Request<?> request) {
+            if(!inUse.getAndSet(true)) {
+                requestBlockingQueue.add(request);
+            }
         }
     }
 
     public QuorumReplicaRouter() throws TTransportException, IOException {
         assert(Config.getRoutingMode() == RoutingMode.QUORUM);
-        
+
         this.replicaAddressesByCluster = Maps.newHashMap();
         this.numClusters = Config.getNumClusters();
         this.numNeighbors = Config.getServersInCluster().size();
         this.quorum = (int) Math.ceil((numNeighbors+1)/2);
-        
+
         for (int i = 0; i < numClusters; i ++) {
             List<ServerAddress> neighbors = Config.getServersInCluster(i+1);
             for (ServerAddress neighbor : neighbors) {
                 replicaRequestQueues.put(neighbor, new ReplicaClient(
-                        ThriftUtil.getReplicaServiceAsyncClient(neighbor.getIP(), neighbor.getPort())));
+                        ThriftUtil.getReplicaServiceSyncClient(neighbor.getIP(), neighbor.getPort())));
             }
             replicaAddressesByCluster.put(i+1, neighbors);
         }
     }
-    
+
     @Override
     public boolean put(String key, DataItem value) throws TException {
         logger.error("PUT for key " + key);
[…]

### 2013-03-30 thebes-code/common/src/main/java/edu/berkeley/thebes/common/config/Config.java
cache size for leveldb

@@ -351,4 +351,8 @@ public class Config {
     public static int getNumClusters() {
         return getClusterMap().size();
     }
+
+    public static int getDatabaseCacheSize() {
+        return getOption(ConfigParameters.DATABASE_CACHE_SIZE);
+    }
 }

### 2013-03-30 thebes-code/common/src/main/java/edu/berkeley/thebes/common/config/ConfigParameters.java

@@ -48,7 +48,8 @@ public enum ConfigParameters {
     ATOMICITY_LEVEL(AtomicityLevel.class, AtomicityLevel.NO_ATOMICITY),
     LOGGER_LEVEL(String.class, "WARN"),
     DISK_DATABASE_FILE(String.class, "/tmp/thebes.db"),
-    DO_CLEAN_DATABASE_FILE(Boolean.class, true);
+    DO_CLEAN_DATABASE_FILE(Boolean.class, true),
+    DATABASE_CACHE_SIZE(Integer.class, -1);
     
     /** Note that defaultValue and reqLevels are mutually exclusive. */
     private Class<?> type;

### 2013-03-30 thebes-code/common/src/main/java/edu/berkeley/thebes/common/persistence/disk/LevelDBPersistenceEngine.java

@@ -38,6 +38,9 @@ public class LevelDBPersistenceEngine implements IPersistenceEngine {
 
     public void open() throws IOException {
         Options options = new Options();
+        options.blockSize(1024);
+        if(Config.getDatabaseCacheSize() != -1)
+            options.cacheSize(Config.getDatabaseCacheSize());
         options.createIfMissing(true);
         lockManager = new LockManager();
 

### 2013-03-30 thebes-code/hat/client/src/main/java/edu/berkeley/thebes/hat/client/clustering/QuorumReplicaRouter.java
printf

@@ -62,7 +62,10 @@ public class QuorumReplicaRouter extends ReplicaRouter {
                 @Override
                 public void run() {
                     while(true) {
+                        logger.error("Before Take");
                         Request<?> request = Uninterruptibles.takeUninterruptibly(requestBlockingQueue);
+                        logger.error("After Take");
+
                         request.process(ReplicaClient.this);
                     }
                 }
@@ -70,9 +73,15 @@ public class QuorumReplicaRouter extends ReplicaRouter {
         }
 
         public void executeRequest(Request<?> request) {
+            logger.error("In execute");
+
             if(!inUse.getAndSet(true)) {
+                logger.error("In execute; inside getandset");
+
                 requestBlockingQueue.add(request);
             }
+            else
+                logger.error("In execute; outside getandset");
         }
     }
 

### 2013-03-30 thebes-code/common/src/main/java/edu/berkeley/thebes/common/config/Config.java
unbreak

@@ -353,6 +353,6 @@ public class Config {
     }
 
     public static int getDatabaseCacheSize() {
-        return getOption(ConfigParameters.DATABASE_CACHE_SIZE);
+        return (Integer) getOption(ConfigParameters.DATABASE_CACHE_SIZE);
     }
 }

### 2013-03-30 thebes-code/common/src/main/java/edu/berkeley/thebes/common/config/Config.java
openjdk

@@ -352,7 +352,7 @@ public class Config {
         return getClusterMap().size();
     }
 
-    public static int getDatabaseCacheSize() {
-        return (Integer) getOption(ConfigParameters.DATABASE_CACHE_SIZE);
+    public static Integer getDatabaseCacheSize() {
+        return getOption(ConfigParameters.DATABASE_CACHE_SIZE);
     }
 }

### 2013-03-30 thebes-code/hat/client/src/main/java/edu/berkeley/thebes/hat/client/clustering/QuorumReplicaRouter.java
remove double reservation

@@ -72,16 +72,19 @@ public class QuorumReplicaRouter extends ReplicaRouter {
             }).start();
         }
 
-        public void executeRequest(Request<?> request) {
+        public boolean executeRequest(Request<?> request) {
             logger.error("In execute");
 
             if(!inUse.getAndSet(true)) {
                 logger.error("In execute; inside getandset");
 
                 requestBlockingQueue.add(request);
+                return true;
             }
             else
                 logger.error("In execute; outside getandset");
+
+            return false;
         }
     }
 
@@ -123,10 +126,8 @@ public class QuorumReplicaRouter extends ReplicaRouter {
         for (List<ServerAddress> replicasInCluster : replicaAddressesByCluster.values()) {
             ServerAddress replicaAddress = replicasInCluster.get(replicaIndex);
             ReplicaClient replica = replicaRequestQueues.get(replicaAddress);
-            if (!replica.inUse.getAndSet(true)) {
-                numSent ++;
-                replica.executeRequest(request);
-            }
+            if(replica.executeRequest(request))
+                numSent++;
         }
         logger.error("Sent " + numSent + " messages for key " + key);
         E ret = request.getResponseWhenReady();

### 2013-03-30 thebes-code/hat/client/src/main/java/edu/berkeley/thebes/hat/client/clustering/QuorumReplicaRouter.java
>=

@@ -175,13 +175,13 @@ public class QuorumReplicaRouter extends ReplicaRouter {
                 replica.client.put(key, value);
                 logger.error("Called put...");
 
-                if (numAcks.incrementAndGet() > quorum) {
+                if (numAcks.incrementAndGet() >= quorum) {
                     sendResponse(true);
                 }
             } catch (TException e) {
                 logger.error("Exception happened!");
 
-                if (numNacks.incrementAndGet() > quorum) {
+                if (numNacks.incrementAndGet() >= quorum) {
                     sendResponse(false);
                 }
             } finally {
@@ -214,7 +214,7 @@ public class QuorumReplicaRouter extends ReplicaRouter {
                     returnedDataItems.add(new DataItem(resp));
                 }
 
-                if (numAcks.incrementAndGet() > quorum) {
+                if (numAcks.incrementAndGet() >= quorum) {
                     if (returnedDataItems.isEmpty()) {
                         sendResponse(new ThriftDataItem()); // "null"
                     } else {
@@ -226,7 +226,7 @@ public class QuorumReplicaRouter extends ReplicaRouter {
                 logger.error("Exception happened!");
 
                 numNacks.incrementAndGet();
-                if (numNacks.incrementAndGet() > quorum) {
+                if (numNacks.incrementAndGet() >= quorum) {
                     sendResponse(new ThriftDataItem()); // "null"
                 }
             } finally {

### 2013-03-30 thebes-code/hat/client/src/main/java/edu/berkeley/thebes/hat/client/clustering/QuorumReplicaRouter.java
Cleaning debugs, fixing NPE

@@ -62,9 +62,7 @@ public class QuorumReplicaRouter extends ReplicaRouter {
                 @Override
                 public void run() {
                     while(true) {
-                        logger.error("Before Take");
                         Request<?> request = Uninterruptibles.takeUninterruptibly(requestBlockingQueue);
-                        logger.error("After Take");
 
                         request.process(ReplicaClient.this);
                     }
@@ -73,17 +71,10 @@ public class QuorumReplicaRouter extends ReplicaRouter {
         }
 
         public boolean executeRequest(Request<?> request) {
-            logger.error("In execute");
-
             if(!inUse.getAndSet(true)) {
-                logger.error("In execute; inside getandset");
-
                 requestBlockingQueue.add(request);
                 return true;
             }
-            else
-                logger.error("In execute; outside getandset");
-
             return false;
         }
     }
@@ -99,7 +90,7 @@ public class QuorumReplicaRouter extends ReplicaRouter {
         for (int i = 0; i < numClusters; i ++) {
             List<ServerAddress> neighbors = Config.getServersInCluster(i+1);
             for (ServerAddress neighbor : neighbors) {
-                logger.error("Connecting to " + neighbor);
+                logger.debug("Connecting to " + neighbor);
                 replicaRequestQueues.put(neighbor, new ReplicaClient(
                         ThriftUtil.getReplicaServiceSyncClient(neighbor.getIP(), neighbor.getPort())));
             }
@@ -109,13 +100,11 @@ public class QuorumReplicaRouter extends ReplicaRouter {
 
     @Override
     public boolean put(String key, DataItem value) throws TException {
-        logger.error("PUT for key " + key);
         return performRequest(key, new WriteRequest(key, value));
     }
 
     @Override
     public ThriftDataItem get(String key, Version requiredVersion) throws TException {
-        logger.error("GET for key " + key);
         return performRequest(key, new ReadRequest(key, requiredVersion));
     }
 
@@ -129,9 +118,7 @@ public class QuorumReplicaRouter extends ReplicaRouter {
             if(replica.executeRequest(request))
                 numSent++;
         }
-        logger.error("Sent " + numSent + " messages for key " + key);
         E ret = request.getResponseWhenReady();
-        logger.error("Received response " + ret + " for key " + key);
         return ret;
     }
 
@@ -171,15 +158,12 @@ public class QuorumReplicaRouter extends ReplicaRouter {
 
         public void process(ReplicaClient replica) {
             try {
[…]

### 2013-03-30 thebes-code/hat/server/src/main/java/edu/berkeley/thebes/hat/server/replica/ReplicaServiceHandler.java

@@ -32,8 +32,8 @@ public class ReplicaServiceHandler implements ReplicaService.Iface {
     public boolean put(String key,
                        ThriftDataItem valueThrift) throws TException {
         DataItem value = new DataItem(valueThrift);
-//        if(logger.isTraceEnabled())
-            logger.error("received PUT request for key: '"+key+
+        if(logger.isTraceEnabled())
+            logger.trace("received PUT request for key: '"+key+
                          "' value: '"+value+
                          "' transactionKeys: "+value.getTransactionKeys());
 

### 2013-03-30 thebes-code/scripts/setup_hosts.py
Quorum stuffs~

@@ -231,7 +231,7 @@ def wait_all_hosts_up(regions, tag):
 
     # Since ssh takes some time to come up
     pprint("Waiting for instances to warm up... ")
-    sleep(60)
+    sleep(20)
     pprint("Awake!")
 
 def claim_instances(regions, tag):
@@ -496,7 +496,7 @@ def start_ycsb_clients(clusters, use2PL, thebesArgString, **kwargs):
 #                           'rm *.log;' \
                            'bin/ycsb %s thebes -p hosts=%s -threads %d -p fieldlength=%d -p histogram.buckets=10000 -p fieldcount=1 -p operationcount=100000000 -p recordcount=%d -t ' \
                            ' -p requestdistribution=%s -p maxexecutiontime=%d -P %s -Dsocket_timeout=%d ' \
-                           ' -DtransactionLengthDistributionType=%s -DtransactionLengthDistributionParameter=%d -Dclientid=%d -Dtxn_mode=%s -Dclusterid=%d -Dhat_isolation_level=%s -Datomicity_level=%s -Dconfig_file=../thebes-code/conf/thebes.yaml -Droute_to_masters=%s %s' \
+                           ' -DtransactionLengthDistributionType=%s -DtransactionLengthDistributionParameter=%d -Dclientid=%d -Dtxn_mode=%s -Dclusterid=%d -Dhat_isolation_level=%s -Datomicity_level=%s -Dconfig_file=../thebes-code/conf/thebes.yaml -Drouting_mode=%s %s' \
                            ' 1>%s_out.log 2>%s_err.log' % (runType,
                                                            hosts,
                                                            kwargs.get("threads", 10) if runType != 'load' else 50,
@@ -513,7 +513,7 @@ def start_ycsb_clients(clusters, use2PL, thebesArgString, **kwargs):
                                                            cluster.clusterID,
                                                            kwargs.get("isolation_level", "NO_ISOLATION"),
                                                            kwargs.get("atomicity_level", "NO_ATOMICITY"),
-                                                           kwargs.get("route_to_masters", "false"),
+                                                           kwargs.get("routing_mode", "NEAREST"),
                                                            thebesArgString,
                                                            runType,
                                                            runType))
@@ -792,14 +792,14 @@ if __name__ == "__main__":
     if args.restart:
         run_ycsb_trial(False, tag, runid="DEFAULT_RUN",
                        threads=60,
-                       distributionparameter=2,
+                       distributionparameter=8,
[…]

### 2013-03-30 writeup/hat-analysis.tex

@@
 
-\begin{quote}
-``...one should not throw out the C so quickly, since there are real
-  error scenarios where CAP does not apply and it seems like a bad
-  tradeoff in many of the other situations.''---Michael
-  Stonebraker~\cite{stonebraker2010errors}
-\end{quote}
-
-HAT systems---providing transactions with transactional availability
-with high availability ($1$-available) or sticky high availability
-($1$-sticky-available)---offer substantial latency and availability
-benefits, yet they come with a cost to achievable semantics. In this
-section, we delineate which of several ACID, distributed consistency,
-and session consistency levels can be achieved with high availability
-(e.g., transactional atomicity, variants of Repeatable Read isolation,
-and many session guarantees) and sticky high availability (e.g.,
-read-your-writes, PRAM, and causal consistency) and which cannot
-(e.g., preventing Lost Update and Write Skew).  We present a full
-summary of these results in Section~\ref{sec:hat-summary}.
+HAT systems, or systems providing transactions with transactional
+availability with high availability ($1$-available) or sticky high
+availability ($1$-sticky-available), offer substantial latency and
+availability benefits, yet they come with a cost to achievable
+semantics. In this section, we delineate which of several ACID,
+distributed consistency, and session consistency levels can be
+achieved with high availability (transactional atomicity, variants of
+Repeatable Read isolation, and many session guarantees) and sticky
+high availability (read your writes, PRAM and causal consistency) and
+which cannot (preventing Lost Update and Write Skew, recency).  We
+present a full summary of these results in
+Section~\ref{sec:hat-summary}.
 
@@
 
-To begin, Adya's \textit{Read Uncommitted} isolation requires that
-transactions are totally ordered and that writes within transactions
+To begin, Adya's \textit{Read Uncommitted} isolation requires a total
+order among transactions and requires that each transaction's writes
 are ordered consistently with this order (prohibiting ``Dirty
@@
 order must exist), whereas a serializable total order must be
-equivalent to a serial execution. It is easily achieved via applying
-the same logical timestamp to each update in a transaction and
-applying a ``last writer wins'' conflict reconciliation policy at each
-replica.
+equivalent to a serial execution. Read Uncommitted is easily achieved
+via applying the same logical timestamp to each update in a
+transaction and applying a ``last writer wins'' conflict
+reconciliation policy at each replica.
 
@@
 writes uncommitted data to the database, then transactions will never
-read each others' dirty data. Clients can buffer their writes until
-they commit, or, alternatively, can send them to servers, who will not
-serve new writes until clients notify them that the writes have been
-committed.
+read each others' dirty data. As a simple solution, clients can buffer
+their writes until they commit, or, alternatively, can send them to
+servers, who will not serve new writes until clients notify them that
+the writes have been committed.
 
@@
 properties, like transactional atomicity, it is stronger: the repeated
-reads must obey additional ordering constraints. In fact, distributed
-systems often describe a ``consistent snapshot'' across a set of
-related events as a \textit{cut} across a set of participants or, in
-our case, data items. Accordingly, to capture the notion that the
-transaction should read from a (non-changing) consistent cut over data
-items (and to disambiguate from the aforementioned stronger Repeatable
-Read properties), we call ANSI SQL ``Repeatable Read'' (i.e.,
-preventing P2) \textbf{Cut Isolation}. It is possible to satisfy cut
-isolation with high availability by caching appropriate versions for
-each of a transaction's reads or, alternatively, using
-multi-versioning on each server and ensuring that each of a
-transaction's successive reads return the same version of each data
-item.
+reads must obey additional ordering constraints
+(Section~\ref{sec:discussion}). In fact, distributed systems often
+describe a ``consistent snapshot'' across a set of related events as a
+\textit{cut} across a set of participants or, in our case, data
+items. Accordingly, to capture the notion that the transaction should
+read from a (non-changing) consistent cut over data items (and to
+disambiguate from the aforementioned stronger Repeatable Read
+properties), we call ANSI SQL ``Repeatable Read'' (i.e., preventing
+P2) \textbf{Cut Isolation}. It is possible to satisfy cut isolation
+with high availability by caching appropriate versions for each of a
+transaction's reads or, alternatively, using multi-versioning on each
+replica and ensuring that each of a transaction's successive reads
+return the same version of each data item.
 
@@
 items. We will call this \textit{Item Cut Isolation}. The second
-provides a cut over individual data items plus logical ranges of data
-items, or predicate-based reads. This prevents the \textit{Phantom
-  Problem}, whereby two successive predicate-based reads return
-different data~\cite{gray-isolation}. We call predicate-based cut
-isolation, which prevents Phantoms, \textit{Predicate Cut
+provides a cut over individual data items and over logical ranges of
+data items, or predicate-based reads (e.g., \texttt{SELECT
+  WHERE}). This prevents the \textit{Phantom Problem}, whereby two
+successive predicate-based reads return different
+data~\cite{gray-isolation}. We call predicate-based cut isolation,
+which prevents Phantoms, \textit{Predicate Cut
   Isolation}.\footnote{Oracle provides an isolation level called
@@
 Perhaps perplexingly, discussions of TA are absent from existing
-discussions of weak isolation. This is perhaps again due to the single-node
-context in which prior work was developed: on a single server (or a
-fully replicated database), TA is near-trivial to achieve via
+discussions of weak isolation. This is perhaps again due to the
+single-node context in which prior work was developed: on a single
+server (or a fully replicated database), TA is achievable via
 lightweight locking and/or local concurrency control over data
@@
 
-Our models have not yet considered interactions across transactions:
-we have not guaranteed any ordering or continuity between transactions
-(other than that there exists some arbitrary ordering). However, in
-the distributed systems context, many useful \textit{safety}
+Our models have not yet considered interactions \textit{across}
+transactions: we have not guaranteed any ordering or continuity
+between transactions (other than that there exists some arbitrary
+ordering). This is a natural concern in a distributed context, and, in
+the distributed systems literature, many useful \textit{safety}
 guarantees span multiple application actions. In particular,
 \textit{session guarantees} are used to describe guarantees across
-groups of operations (here, transactions) issued by the same logical
-user~\cite{sessionguarantees}. A \textit{session} is an application
-context that consists of logical state that should persist between
-transactions. For example, on a social networking site, all of a
-user's transactions submitted between ``log in'' and ``log out''
-operations might form a session.
+transactions within a given \textit{session}, ``an abstraction for the
+sequence of...operations performed during the execution of an
+application''~\cite{sessionguarantees}. Informally a session describes
+a context that should persist between transactions: for example, on a
+social networking site, all of a user's transactions submitted between
+``log in'' and ``log out'' operations might form a session.
 
-There are several guarantees that we can make with high availability:
+There are several session guarantees that we can make with high availability:
 
 \vspace{.5em}\noindent\textit{{Monotonic reads}} requires that,
-with a session, subsequent reads to a given object ``never return any
+within a session, subsequent reads to a given object ``never return any
 previous values''; reads from each item progress forward in a total
@@
 We can achieve the above guarantees by forcing servers to wait to
-reveal new writes until their respective dependencies are fulfilled on
-all replicas. This mechanism effectively requires that all clients
-read from a globally agreed upon lower bound on the versions written
-to the system. This is highly available as a client will never block
-due to inability to find a server with a sufficiently up-to-date
-version of a data item. However, it does not imply that transactions
-will read their own writes or, in the presence of partitions, make
-forward progress through the version history. The problem is that, if
-a server becomes partitioned, under the highly available model, we
-must handle the possibility that an unfortunate client will be forced
-to issue her next requests against the partitioned server!
+reveal new writes until each write's respective dependencies are
+fulfilled on all replicas. This mechanism effectively ensures that all
+clients read from a globally agreed upon lower bound on the versions
+written to the system. This \textit{is} highly available as a client
+will never block due to inability to find a server with a sufficiently
+up-to-date version of a data item. However, it does not imply that
+transactions will read their own writes or, in the presence of
+partitions, make forward progress through the version history. The
+problem is that, if a server becomes partitioned, under the highly
+available model, we must handle the possibility that an unfortunate
+client will be forced to issue her next requests against the
+partitioned server!
 
@@
 
-Read your writes is unavailable in a highly available system. Consider
-a client that executes the following two transactions in succession:
+Read your writes is not achievable in a highly available
+($1$-available) system. Consider a client that executes the following
+two transactions in succession:
 \vspace{-.5em}
@@
 server which is partitioned from the server that executed $T_1$, then
-the client will be inable to read its own writes and the system will
+the client will be unable to read its own writes and the system will
 have to either stall indefinitely to allow the client to read her
 writes (violating transactional availability) or will have to
-sacrifice read your writes guarantees. Accordingly, read your writes
-requires stickiness, and, because causal consistency and PRAM require
-read your writes, they also require stickiness. Read your writes is
-``free'' in a sticky system, while the remaining causality and PRAM
-guarantees can be accomplished with the prior algorithms.
+sacrifice read your writes guarantees. However, if the client remains
+sticky with the server that executed $T_1$, then we can disallow this
+scenario. Accordingly, read your writes, and, by proxy, causal
+consistency and PRAM require stickiness. Read your writes is ``free''
+in a sticky system, while the remaining causality and PRAM guarantees
+can be accomplished with the algorithms for achieving the remaining
+session guarantees.
 
-\subsubsection{Additional Guarantees}
+\subsubsection{Additional HAT Guarantees}
+
+In this section, we briefly discuss additional noteworthy guarantees
+achievable by HAT systems.
 
@@
 (Section~\ref{sec:futurework}), at this point, we have largely
-exhausted the range of achievable, previously defined (and useful) HAT
-semantics. Before summarizing our possibility results, we will present
-impossibility results for HATs, also defined in terms of previously
-identified isolation and consistency anomalies. Perhaps most notably,
-it is impossible to prevent Lost Update or Write Skew in a HAT system.
+exhausted the range of achievable, previously defined (and useful)
+semantics that are available to HAT systems. Before summarizing our
+possibility results, we will present impossibility results for HATs,
+also defined in terms of previously identified isolation and
+consistency anomalies. Perhaps most notably, it is impossible to
+prevent Lost Update or Write Skew in a HAT system.
 
@@
 \end{align*}
-The final state of the database will converge to an execution that
-could not have resulted from a serial excecution.\footnote{In this
-  example, we assume that, as is standard in modern databases,
-  replicas accept values as they are written (i.e., register
-  semantics). This particular example could be made serializable via
-  the use of commutative updates (Section~\ref{sec:eval}) but, as we
-  show in the extended version of the paper~\cite{hat-tr}, the problem
-  persists in the general case.} To prevent this from happening,
-  either $T_1$ or $T_2$ should not have committed. Each client's
-  respective server might try to detect that another write occurred,
-  but this requires knowing the version of the latest write to $x$. In
-  our example, this reduces to a requirement for linearizability,
-  which is, via Gilbert and Lynch's proof of the CAP Theorem, provably
-  unachievable with high availabilty~\cite{gilbert-cap}.
+Regardless of whether $x=120$ or $x=130$ is chosen by a replica, the
+database state could not have arisen serial execution of $T_1$ and
+$T_2$.\footnote{In this example, we assume that, as is standard in
+  modern databases, replicas accept values as they are written (i.e.,
+  register semantics). This particular example could be made
+  serializable via the use of commutative updates
+  (Section~\ref{sec:eval}) but, as we show in the extended version of
+  the paper~\cite{hat-tr}, the problem persists in the general case.}
+To prevent this from happening, either $T_1$ or $T_2$ should not have
+committed. Each client's respective server might try to detect that
+another write occurred, but this requires knowing the version of the
+latest write to $x$. In our example, this reduces to a requirement for
+linearizability, which is, via Gilbert and Lynch's proof of the CAP
+Theorem, provably unachievable with high
+availabilty~\cite{gilbert-cap}.
 
@@
 that Repeatable Read and One-Copy Serializability guarantees are
-unavailable.
+unavailable~\cite{adya,ansicritique}.
 
@@
 date'' or similar. Unfortunately, an indefinitely long partition can
-force a system to violate any recency bound, so recency bounds are
-unavailable to HAT systems.
+force a system to violate any recency bound, so recency bounds are not
+enforceable in HAT systems.
 
@@
 database ACID, distributed consistency, and session guarantee
-models. Perhaps satisfactorily, strong one-copy serializability
-subsumes all other models, while considering the (large) power set of
-all compatible models (e.g., the diagram depicts 96 possible highly available
-combinations) hints at the vast expanse of consistency models found in
-the literature. This taxonimization is \textit{not} exhaustive, but we
-believe it lends substantial clarity into the relationships between a
-large subset of the prominent ACID and distributed consistency
-models. Additional read/write transaction semantics that we have
-omitted should be easily classifiable based on the available
-primitives and unavailable anomalies we have already discussed.
+models. Interestingly, strong one-copy serializability subsumes all
+other models, while considering the (large) power set of all
+compatible models (e.g., the diagram depicts 96 possible highly
+available combinations) hints at the vast expanse of consistency
+models found in the literature. This taxonimization is \textit{not}
+exhaustive, but we believe it lends substantial clarity into the
+relationships between a large subset of the prominent ACID and
+distributed consistency models. Additional read/write transaction
+semantics that we have omitted should be easily classifiable based on
+the available primitives and HAT-incompatible anomaly prevention we
+have already discussed.
 
@@
 
-Composition cost also varies by property. For instance, Charron-Bost
+Composition cost also varies by combination. For instance, Charron-Bost
 has proven that, to capture causality between $N$ communicating
@@
   Atomicity} The relationship between linearizability and
-transactional atomicity is non-obvious. One might attempt to find a
-mapping between the two: transactional atomicity dictates that writes
-to multiple keys across multiple servers are made visible to readers
-all at once, while linearizability dictates that writes to a single
-key on multiple servers are made visible to all readers at once. The
-difference is two-fold. First, in linearizable (and safe and regular)
-systems, writes are made visible to all clients \textit{immediately}
-after they finish. With transactional atomicity, there is no recency
-guarantee. Second, in linearizabile systems, all clients see all
-writes at the same time. With transactional atomicity, clients see
-writes at different times depending on which replicas they contact. We
-are not aware of an analogous model in the distributed systems
-literature. Accordingly, despite apparent similarities, transactional
-atomicity is much weaker than linearizability.
+transactional atomicity is non-obvious. Transactional atomicity
+dictates that writes to multiple keys across multiple servers are made
+visible to readers all at once, while linearizability dictates that
+writes to a single key on multiple servers are made visible to all
+readers at once---what is different? First, in linearizable (and safe
+and regular) systems, writes are made visible to all clients
+\textit{immediately} after they finish. With transactional atomicity,
+there is no recency guarantee. Second, in linearizabile systems, all
+clients see all writes at the same time. With transactional atomicity,
+clients may see writes at different times depending on which replicas
+they contact. We are not aware of an analogous model in the
+distributed systems literature. Accordingly, despite apparent
+similarities, transactional atomicity incomparable with and much
+cheaper (by availability standards) than linearizability.
 
@@
 read-your-writes (due to the sticky availability requirement) may
-still wish to read other clients' writes with timeliness. Accordingly,
-a system may forgo high availability in order to provide better
-visibility. We also note that a system may wish to provide stickiness
-in the absence of partitions, but, in the presence of partitions,
-``lose'' session guarantees by having clients reconnect to an
-available replica. This latter strategy provides session guarantees in
-the absence of partitions but, to provide availability, sacrifices
-them in the event of partitions.
+still wish to read other clients' writes with timeliness.

### 2013-03-30 thebes-code/scripts/setup_hosts.py
quorum in constants

@@ -731,7 +731,7 @@ def run_tpcc_trial(use2PL, tag, serverArgs="", **kwargs):
     write_config(clusters, graphiteRegion)
     restart_graphite(graphiteRegion)
     run_cmd("all-servers", "rm -rf /mnt/md0/thebes.db; tar -xf /home/ubuntu/thebes.db.tar.gz /mnt/md0")
-    start_servers(clusters, use2PL, thebesArgString+" "+serverArgs+"-Ddo_clean_database_file=false -Dpersistence_engine=leveldb -Ddisk_database_file=/mnt/md0/thebes.db")
+    start_servers(clusters, use2PL, thebesArgString+" "+serverArgs+"-Ddo_clean_database_file=false -Dpersistence_engine=leveldb -Ddisk_database_file=/mnt/md0/thebes.db -Ddatabase_cache_size=1073741824")
     start_tpcc_clients(clusters, use2PL, thebesArgString, **kwargs)
     runid = kwargs.get("runid", str(datetime.now()).replace(' ', '_'))
     fetch_logs(runid, clusters)
@@ -913,6 +913,22 @@ if __name__ == "__main__":
                                    timeout=120*10000,
                                    keydistribution="uniform",
                                    routing_mode="MASTERED")
+
+
+                    isolation_level = "NO_ISOLATION"
+                    atomicity_level = "NO_ATOMICITY"
+                    run_ycsb_trial(False, tag, runid=("QUORUM_EVENTUAL-%d-THREADS%d-IT%d" % (transaction_length, 
+                                                                                           threads,
+                                                                                          iteration)),
+                                   threads=threads,
+                                   distributionparameter=transaction_length,
+                                   atomicity_level=atomicity_level,
+                                   isolation_level=isolation_level,
+                                   recordcount=100000,
+                                   time=120,
+                                   timeout=120*10000,
+                                   keydistribution="uniform",
+                                   routing_mode="QUORUM")
 
                     isolation_level = "NO_ISOLATION"
                     atomicity_level = "NO_ATOMICITY"

### 2013-03-30 thebes-code/hat/client/src/main/java/edu/berkeley/thebes/hat/client/clustering/QuorumReplicaRouter.java
fixing exception print and double nack count

@@ -205,9 +205,8 @@ public class QuorumReplicaRouter extends ReplicaRouter {
                 }
 
             } catch (TException e) {
-                logger.error("Exception happened!");
+                logger.error("Exception:", e);
 
-                numNacks.incrementAndGet();
                 if (numNacks.incrementAndGet() >= quorum) {
                     sendResponse(new ThriftDataItem()); // "null"
                 }

### 2013-03-30 thebes-code/common/src/main/java/edu/berkeley/thebes/common/interfaces/IThebesClient.java
start of batching

@@ -7,6 +7,8 @@ import javax.naming.ConfigurationException;
 import java.io.FileNotFoundException;
 import java.io.IOException;
 import java.nio.ByteBuffer;
+import java.util.List;
+import java.util.Map;
 
 public interface IThebesClient {
     public void open() throws TTransportException, ConfigurationException, IOException;
@@ -17,7 +19,11 @@ public interface IThebesClient {
     public boolean commitTransaction() throws TException;
 
     public boolean put(String key, ByteBuffer value) throws TException;
+    public boolean put_all(Map<String, ByteBuffer> pairs) throws TException;
+
     public ByteBuffer get(String key) throws TException;
+    public Map<String, ByteBuffer> get_all(List<String> keys) throws TException;
+
     /** Sends an arbitrary, parseable command. */  
     public void sendCommand(String cmd) throws TException;
 

### 2013-03-30 thebes-code/hat/client/src/main/java/edu/berkeley/thebes/hat/client/ThebesHATClient.java

@@ -320,27 +320,6 @@ public class ThebesHATClient implements IThebesClient {
         }
         return ret;
     }
-
-//    private boolean doPutAsync(String key,
-//                               DataItem value,
-//                               TransactionMultiPutCallback callback) throws TException {
-//        TimerContext timer = latencyPerOperationMetric.time();
-//
-//        try {
-//            router.put(key,
-//                                                 value.toThrift(),
-//                                                 callback);
-//        } catch (RuntimeException e) {
-//            errorMetric.mark();
-//            throw e;
-//        } catch (TException e) {
-//            errorMetric.mark();
-//            throw new TException("exception on replica "+router.getReplicaIPByKey(key)+" "+e.getMessage());
-//        } finally {
-//            timer.stop();
-//        }
-//        return true;
-//    }
     
     private DataItem doGet(String key) throws TException {
         TimerContext timer = latencyPerOperationMetric.time();

### 2013-03-30 thebes-code/hat/client/src/main/java/edu/berkeley/thebes/hat/client/clustering/ReplicaRouter.java

@@ -10,6 +10,8 @@ import edu.berkeley.thebes.common.data.Version;
 import edu.berkeley.thebes.common.thrift.ThriftDataItem;
 
 import java.io.IOException;
+import java.util.List;
+import java.util.Map;
 
 public abstract class ReplicaRouter {
     public static ReplicaRouter newInstance(RoutingMode routingMode)
@@ -27,5 +29,8 @@ public abstract class ReplicaRouter {
     }
     
     abstract public boolean put(String key, DataItem value) throws TException;
+    abstract public boolean put_all(Map<String, DataItem> pairs) throws TException;
+
     abstract public ThriftDataItem get(String key, Version requiredVersion) throws TException;
+    abstract public List<ThriftDataItem> get_all(List<String> keys) throws TException;
 }

### 2013-03-30 thebes-code/hat/common/src/main/java/edu/berkeley/thebes/hat/common/thrift/ReplicaService.java

@@ -36,7 +36,11 @@ public class ReplicaService {
 
     public edu.berkeley.thebes.common.thrift.ThriftDataItem get(String key, edu.berkeley.thebes.common.thrift.ThriftVersion requiredVersion) throws org.apache.thrift.TException;
 
+    public Map<String,edu.berkeley.thebes.common.thrift.ThriftDataItem> get_all(List<String> keys) throws org.apache.thrift.TException;
+
     public boolean put(String key, edu.berkeley.thebes.common.thrift.ThriftDataItem value) throws org.apache.thrift.TException;
+
+    public boolean put_all(Map<String,edu.berkeley.thebes.common.thrift.ThriftDataItem> pairs) throws org.apache.thrift.TException;
 
   }
 
@@ -44,7 +48,11 @@ public class ReplicaService {
 
     public void get(String key, edu.berkeley.thebes.common.thrift.ThriftVersion requiredVersion, org.apache.thrift.async.AsyncMethodCallback<AsyncClient.get_call> resultHandler) throws org.apache.thrift.TException;
 
+    public void get_all(List<String> keys, org.apache.thrift.async.AsyncMethodCallback<AsyncClient.get_all_call> resultHandler) throws org.apache.thrift.TException;
+
     public void put(String key, edu.berkeley.thebes.common.thrift.ThriftDataItem value, org.apache.thrift.async.AsyncMethodCallback<AsyncClient.put_call> resultHandler) throws org.apache.thrift.TException;
+
+    public void put_all(Map<String,edu.berkeley.thebes.common.thrift.ThriftDataItem> pairs, org.apache.thrift.async.AsyncMethodCallback<AsyncClient.put_all_call> resultHandler) throws org.apache.thrift.TException;
 
   }
 
@@ -92,6 +100,29 @@ public class ReplicaService {
       throw new org.apache.thrift.TApplicationException(org.apache.thrift.TApplicationException.MISSING_RESULT, "get failed: unknown result");
     }
 
+    public Map<String,edu.berkeley.thebes.common.thrift.ThriftDataItem> get_all(List<String> keys) throws org.apache.thrift.TException
+    {
+      send_get_all(keys);
+      return recv_get_all();
+    }
+
+    public void send_get_all(List<String> keys) throws org.apache.thrift.TException
+    {
+      get_all_args args = new get_all_args();
+      args.setKeys(keys);
+      sendBase("get_all", args);
+    }
+
[…]

### 2013-03-30 thebes-code/thebes-client/src/main/java/edu/berkeley/thebes/client/ThebesClient.java

@@ -16,6 +16,8 @@ import javax.naming.ConfigurationException;
 import java.io.FileNotFoundException;
 import java.io.IOException;
 import java.nio.ByteBuffer;
+import java.util.List;
+import java.util.Map;
 
 // Wrapper class for various thebes clients
 
@@ -67,7 +69,16 @@ public class ThebesClient implements IThebesClient {
     public boolean put(String key, ByteBuffer value) throws TException {
         return internalClient.put(key, value);
     }
-    
+
+    @Override
+    public boolean put_all(Map<String, ByteBuffer> pairs) throws TException {
+        return internalClient.put_all(pairs);
+    }
+
+    public Map<String, ByteBuffer> get_all(List<String> keys) throws TException {
+        return internalClient.get_all(keys);
+    }
+
     public boolean unsafe_load(String key, ByteBuffer value) throws TException {
         if (internalClient instanceof ThebesTwoPLTransactionClient) {
             return ((ThebesTwoPLTransactionClient) internalClient).unsafe_load(key, value);

### 2013-03-30 thebes-code/hat/client/src/main/java/edu/berkeley/thebes/hat/client/clustering/QuorumReplicaRouter.java
quorum final

@@ -44,7 +44,7 @@ public class QuorumReplicaRouter extends ReplicaRouter {
     private final Map<Integer, List<ServerAddress>> replicaAddressesByCluster;
     private final int numClusters;
     private final int numNeighbors;
-    private int quorum;
+    private final int quorum;
 
     private final Map<ServerAddress, ReplicaClient> replicaRequestQueues = Maps.newHashMap();
 
@@ -86,6 +86,8 @@ public class QuorumReplicaRouter extends ReplicaRouter {
         this.numClusters = Config.getNumClusters();
         this.numNeighbors = Config.getServersInCluster().size();
         this.quorum = (int) Math.ceil((numNeighbors+1)/2);
+
+        logger.debug("Quorum is set to "+this.quorum);
 
         for (int i = 0; i < numClusters; i ++) {
             List<ServerAddress> neighbors = Config.getServersInCluster(i+1);

### 2013-03-30 writeup/intro.tex
New intro

@@
 
-Requirements for high availability, predictable low latency, and
-simplicity of coordination protocols have led many recent distributed
-database designers to renounce several properties provided by
-traditional database systems~\cite{bigtable, pnuts, dynamo}. In
-particular, many modern systems forgo providing ACID guarantees, which
-often provide single-system image programmability but traditionally
-suffer from unavailability and often expensive coordination
-costs~\cite{davidson-survey}. In particular, the July 2000
-announcement of the CAP Theorem~\cite{gilbert-cap}, which states that
-a highly available system cannot provide recency guarantees in the
-presence of network partitions, proved a rallying call for a
-generation of scalable but semantically weak system designs. Over a
-decade later, we observe a dearth of highly available systems
-providing general-purpose multi-operation \textit{transactions} over
-multiple data items with useful semantics.
+In July 2000, the CAP Theorem~\cite{gilbert-cap} was crisply
+formulated that a highly available system cannot provide strong
+consistency guarantees in the presence of network partitions. This
+became a rallying call for a generation of highly scalable but
+semantically weak system designs. The NoSQL movement embraced this
+mantra, leading to many production quality databases that eschewed
+strong consistency guarantees, but could scale-out to many
+servers~\cite{bigtable, pnuts, dynamo}. The consistency concept used
+in the CAP theorem was proposed by the distributed systems community,
+and concerned providing strong recency guarantees of a data item that
+is replicated across servers.  However, the NoSQL movement implicitly
+took it to imply that transactions are also unavailable in the
+presence of network partitions. Over a decade later, we observe a
+dearth of highly available systems providing general-purpose
+multi-operation \textit{transactions} over multiple data items with
+useful semantics.
 
-While serializability, the gold standard of traditional ACID
-transactions, is not achievable with high availability for general
-read-write transactions~\cite{davidson-survey}, database systems have
-a long tradition of providing weaker isolation and consistency
-guarantees~\cite{adya, ansicritique, gray-isolation}. Due to their
-concurrency and overall performance benefits, weak isolation models
-are widely employed, overwhelmingly as the default setting in today's
-traditional ``ACID'' and up-and-coming ``NewSQL'' databases and often
-as the maximum guarantee offered (Section~\ref{sec:modernacid}). While
-this weak isolation does not provide serializability for
-general-purpose transactions, it is apparently strong enough to
-deliver acceptable behavior to many application programmers and is
-substantially stronger than the typical semantics of current highly
-available systems. This raises the question: which of these semantics
-can be provided while remaining highly available?
+ Serializability, the gold standard of traditional ACID transactions,
+ is indeed not achievable with the requirement of high availability
+ the in the presence of partitions~\cite{davidson-survey}. However,
+ database systems have a long tradition of providing weaker isolation
+ and consistency guarantees~\cite{adya, ansicritique,
+   gray-isolation}. Due to their concurrency and overall performance
+ benefits, weak isolation models are widely employed, overwhelmingly
+ as the default setting in today's traditional ``ACID'' and
+ up-and-coming ``NewSQL'' databases and often as the maximum guarantee
+ offered (Section~\ref{sec:modernacid}). While this weak isolation
+ does not provide serializability for general-purpose transactions, it
+ is apparently strong enough to deliver acceptable behavior to many
+ application programmers and is substantially stronger than the
+ typical semantics of current highly available systems. This raises
+ the question: which of these semantics can be provided while always
+ remaining highly available?
 
@@
 literature does not provide guidance as to the operation of weakly
-consistent transactional guarantees in a replicated
-context. Accordingly, in this work, we consider the problem of
-providing Highly Available Transactions (HATs), or semantic guarantees
-regarding multiple operations operating on multiple data items that
-are achievable with high availability and low latency, albeit with
-some cost to achievable semantics. While our prior, preliminary
-results indicated that some models are achievable~\cite{hat-hotos},
-the literature lacks a deep understanding of what models are
-available, which are not, and when this distinction matters.
+consistent transactional guarantees in a replicated context. To make
+matters worse, the distributed computing literature has produced a
+large body of work on weak data consistency models for single
+replicated objects. Accordingly, in this work, we consider the problem
+of providing understanding semantic guarantees regarding multiple
+operations operating on data items that are achievable with high
+availability and low latency, albeit with some cost to achievable
+semantics. While our prior, preliminary results indicated that some
+models are achievable~\cite{hat-hotos}, the literature lacks a deep
+understanding of what models are available, which are not, and when
+this distinction matters.
 
-This work attempts to bridge the gap between ACID properties and
+This work attempts to bridge the gap between, one the one hand, the
+plethora of isolation and consistency models proposed in the
+transactions and distributed systems literature, and, on the other hand
 highly available systems. We introduce a hierarchy of consistency
@@
 session guarantees~\cite{sessionguarantees}, and traditional
-disributed ``register'' semantics~\cite{herlihy-art}. By identifying
-the availability cost of a wide range of guarantees, this work allows
-developers to weigh the costs and benefits of relying on particular
-guarantees in their applications. While several models, like
-serializability and Snapshot Isolation, are unachievable due to update
-conflict avoidance (preventing Lost Update and Write Skew phenomena),
-many other ACID guarantees are HAT compliant, such as providing Read
-Committed isolation, preventing Fuzzy Read phenomena, and ensuring
-atomicity of transactional updates. Several other guarantees, like
-reading one's writes, fall into a new availability class---``sticky''
-high availability---which provides an intermediate ground between
-strict high availability and majority availability.
+distributed ``register'' semantics~\cite{herlihy-art}.
+
+Our main contributions are the following. We classify among the wide
+array of isolation and consistency models those that are highly
+available, denoting them as {\em Highly Available Transactions
+  (HATs)}.  In doing so, we show that many implementations that
+provide HAT-semantics are not available, but that the unavailability
+is {\em not} inherent to their semantic guarantees. Thus, we provide
+highly available implementations for HATs.  Furthermore, our
+investigation shows, somewhat surprisingly, that besides
+serializability, only Snapshot Isolation (SI) and Repeatable Reads
+(RR) are unavailable. The reason for their unavailability stems from
+their requirement to avoid update conflict (preventing Lost Updates
+and Write Skew phenomena). Finally, we show that many of the weaker
+data consistency models from distributed systems can be implemented in
+a highly available manner as well. For that, the model of the system
+has to be changed to include clients, which cannot be mobile, but
+rather have to {\em stick} to one server, something which is implicit
+in much of the previous work on distributed systems.
 
@@
 
-In this paper, we make the following contributions:
-\begin{myitemize}
-\item We model high availability in a transactional environment,
-  including traditional and sticky high availability.
+%% In this paper, we make the following contributions:
+%% \begin{myitemize}
+%% \item We model high availability in a transactional environment,
+%%   including traditional and sticky high availability.
 
-\item We taxonomize ACID and distributed consistency properties
-  according to their availability characteristics.
+%% \item We taxonomize ACID and distributed consistency properties
+%%   according to their availability characteristics.
 
-\item We analyze existing concurrency control algorithms, perform a
-  case-study of an existing transactional application, and
-  briefly evaluate a HAT database prototype.
-\end{myitemize}
+%% \item We analyze existing concurrency control algorithms, perform a
+%%   case-study of an existing transactional application, and
+%%   briefly evaluate a HAT database prototype.
+%% \end{myitemize}

### 2013-03-30 writeup/hat-analysis.tex
Couple very minor fixes

@@
 straightforward and have been introduced---albeit with greater brevity
-and discussion elided---in a preliminary workshop
+and discussion omitted---in a preliminary workshop
 paper~\cite{hat-hotos}. In our examples, we exclusively consider read

### 2013-03-31 writeup/availability-appendix.tex
checkpoint

\section{Availability}

\begin{definition}[$K$-Availability]
A system provides $K$-Availability if
\end{definition}

\begin{definition}[High Availability]
A system provides High Availability if it provides $1$-Availability.
\end{definition}

\begin{definition}[Sticky $K$-Availability]
A system provides Sticky $K$-Availability if 
\end{definition}

\begin{definition}[Sticky High Availability]
A system provides Sticky High Availability if it provides Sticky
$1$-Availabilty.
\end{definition}

\begin{definition}[Availability Strength]
An availability model $M_1$ is \textit{more available} than another
availability model $M_2$ if, whenever $M_2$ is satisfied, $M_1$ is
satisfied, and there is at least one configuration under which $M_1$
is satisfied but $M_2$ is not. If $M_1$ is more available than $M_2$,
then $M_2$ is \textit{less available} than $M_1$.
\end{definition}

\begin{observation}
Majority availability is less available than high sticky availability.
\end{observation}

### 2013-03-31 ycsb-0.1.4/core/src/main/java/com/yahoo/ycsb/DBWrapper.java
Don't record failed xacts

@@ -56,6 +56,7 @@ public class DBWrapper extends DB
 
         if(_transactionalDB != null && _transactionalDB.getNextTransactionLength() == requests.size()) {
         	long txStart = System.nanoTime();
+        	boolean isFucked = false;
         	
         	// TODO: Decide if we want to order the thingies
         	Collections.sort(requests);
@@ -76,6 +77,7 @@ public class DBWrapper extends DB
                             System.err.println("[IR on " + req.getKey() + "] FAILED HERE");
                         }
                     }
+                    isFucked = true;
         			break;
         		}
         	}
@@ -83,8 +85,10 @@ public class DBWrapper extends DB
         	
         	requests.clear();
         	
-            _measurements.measure("TRANSACTION", (int)((System.nanoTime()-txStart)/1000));
-            _measurements.reportReturnCode("TRANSACTION", 1);
+        	if (!isFucked) {
+                _measurements.measure("TRANSACTION", (int)((System.nanoTime()-txStart)/1000));
+                _measurements.reportReturnCode("TRANSACTION", 1);
+        	}
         }
     }
 

### 2013-03-31 writeup/hat-analysis.tex

@@
 (1SR)\lostupdate\rwskew, Recency\linearizable, Safe\linearizable,
-Regular\linearizable, Strong 1SR\lostupdate\rwskew\linearizable
-\\\hline
+Regular\linearizable, Linearizability\linearizable, Strong
+1SR\lostupdate\rwskew\linearizable \\\hline
 \end{tabular}
@@
  \node[draw=none,fill=none] (ta) at (-.2, 1.3) {TA};
- \node[draw=none,fill=none] (mr) at (2.4, 0) {MR};
- \node[draw=none,fill=none] (mw) at (3.6, 0) {MW};
- \node[draw=none,fill=none] (wfr) at (6.3,0) {WFR};
- \node at (5,0) [sticky] (ryw) {RYW};
+
+ \node[draw=none,fill=none] (mr) at (3.6, 0) {MR};
+ \node[draw=none,fill=none] (mw) at (4.8, 0) {MW};
+ \node[draw=none,fill=none] (wfr) at (2.4,0) {WFR};
+ \node at (6.1,0) [sticky] (ryw) {RYW};
 
@@
  \node[noha](linearizable) at (7.7, 3) {linearizable};
- \node at (3.6, 2) [sticky] (causal) {causal};
- \node at (3.6, 1) [sticky] (pram) {PRAM};
+ \node at (4.8, 2) [sticky] (causal) {causal};
+ \node at (4.8, 1) [sticky] (pram) {PRAM};
  \node[noha] (cs) at (-1.2, 1.8) {CS};
@@
  \node[noha] (1sr) at (1.2, 3.2) {1SR};
- \node[noha] (ssr) at (3.6, 3.6) {Strong-1SR};
+ \node[noha] (ssr) at (3.85, 3.6) {Strong-1SR};
 
@@
  \draw [->, red] (ta) -- (rr);
- \draw [->, red] (causal) -- (ssr);
+ \draw [->, red] (causal) -- (linearizable);
+ \draw [->, red] (ryw) -- (safe);
 

### 2013-03-31 writeup/hat-appendix.tex

\section{HAT Guarantees}

\begin{definition}[Read Uncommitted]
\end{definition}

\begin{definition}[Read Committed]
\end{definition}

\begin{definition}[Item Cut Isolation]
\end{definition}

\begin{definition}[Predicate Cut Isolation]
\end{definition}

\begin{definition}[Transactional Atomicity]
\end{definition}

\begin{definition}[Lost Update]
\end{definition}

\begin{definition}[Write Skew]
\end{definition}

\begin{definition}[Snapshot Isolation]
\end{definition}

\begin{definition}[Repeatable Read]
\end{definition}

\begin{definition}[Monotonic Reads]
\end{definition}

\begin{definition}[Monotonic Writes]
\end{definition}

\begin{definition}[Writes Follow Reads]
\end{definition}

\begin{definition}[Read Your Writes]
\end{definition}

\begin{observation}
Read Your Writes is not highly available.
\end{observation}

\begin{corollary}
PRAM and Causal Consistency are not highly available.
\end{corollary}

\begin{definition}[PRAM Consistency]
\end{definition}

\begin{definition}[Causal Consistency]
\end{definition}

\begin{definition}[Recency Guarantees]
\end{definition}

\begin{definition}[Safe Register]
\end{definition}

\begin{definition}[Regular Register]
\end{definition}

\begin{definition}[Linearizable Register]
\end{definition}

\begin{definition}[Strong One-copy Serializability]
\end{definition}

\begin{definition}[Model Strength]
\end{definition}

\begin{observation}
Transactional Atomicity is weaker than Repeatable Read.
\end{observation}

\begin{observation}
Transactional Atomicity is weaker than Snapshot Isolation.
\end{observation}

\begin{observation}
Causality is weaker than linearizability.
\end{observation}

\begin{observation}
Read Your Writes is weaker than safe register semantics.
\end{observation}

### 2013-03-31 writeup/hat-vldb.tex
New intro

@@
 \documentclass{vldb}
-\usepackage{times, graphicx, url, xcolor, multicol, minipage, subfig, tikz, draftwatermarkbottom, draftwatermarktop}
+\let\proof\relax
+\let\endproof\relax
+
+\usepackage{times, graphicx, url, xcolor, multicol, minipage, subfig, tikz, draftwatermarkbottom, draftwatermarktop, amsthm}
 \usepackage{balance}  % for  \balance command ON LAST PAGE  (only there!)
+\usepackage{xspace}
+
+\newcommand{\etal}{{\em et al.}\xspace}
+\newcommand{\etc}{{\em etc}\xspace}
+\newcommand{\cf}{{\em c.f.,}\xspace}
+\newcommand{\versus}{{\em vs.}\xspace}
+\newcommand{\ie}{{\em i.e.,}\xspace}
+\newcommand{\eg}{{\em e.g.,}\xspace}
+
+\theoremstyle{definition}
+\newtheorem{observation}{Observation}
+\newtheorem{theorem}{Theorem}
+\newtheorem{definition}{Definition}
+\newtheorem{corollary}{Corollary}
 
@@
 \begin{abstract}
-To minimize network latency and handle the possibility of partial
-failure and network partitions, many modern distributed systems eschew
-transactional functionality, which provides strong semantic guarantees
-for groups of multiple operations over multiple data items. In this
-work, we consider the problem of achieving transactional guarantees
-with high availability, or without suffering from unavailability
-during system partitions or incurring high network latency.  We
-introduce a new taxonomy of high availability and analyze a
-combination of ACID and distributed consistency guarantees to identify
-which can and cannot be achieved by Highly Available Transaction (HAT)
-systems. In doing so, we analytically and experimentally quantify the
-benefit of providing high availability as well as the associated
-semantic compromises.
+To minimize network latency and remain online during partial failure
+and network partitions, many modern distributed data storage systems
+eschew transactional functionality, which provides strong semantic
+guarantees for groups of multiple operations over multiple data
+items. In this work, we consider the problem of providing Highly
+Available Transactions (HATs), or transactional guarantees that do not
+suffer from unavailability during system partitions or incur high
+network latency.  We introduce a new taxonomy of high availability and
+analyze a existing ACID and distributed consistency guarantees to
+identify which can and cannot be achieved in HAT systems. In doing so,
+we analytically and experimentally quantify the benefit of providing
+high availability as well as the associated semantic compromises.
 \end{abstract}
+
+In the debate over NoSQL systems versus traditional DBMS technologies for distributed data management, one key issue is that many NoSQL systems aim for availability in the face of partitions, and provide no, or restricted, support for transactions that group multiple operations over multiple data items.
+The authors recently advocated Highly Available Transaction (HAT) systems, where general transactions are provided with ACID-like semantic guarantees such as weak isolation levels, in ways that remain available during partitions and avoid high network latency.
+Here we introduce a new taxonomy of partition-availability properties.
+We analyse existing ACID and distributed consistency guarantees, showing the availability that can be achieved for each.
+We analytically and experimentally estimate the performance gains and semantic compromises of providing high availability, compared to traditional serializable transactions in an internet-scale setting.
 
@@
 % Example of an appendix; typically would start on a new page
-%pagebreak
+\pagebreak
 
-%\begin{appendix}
+\begin{appendix}
+\input{availability-appendix.tex}
 
-%\end{appendix}
-
+\input{hat-appendix.tex}
+\end{appendix}
 \end{document}

### 2013-03-31 writeup/alan-intro.tex
alan intro

\section{Introduction}
In July 2000, the CAP Theorem \cite{40} was formulated, stating that a highly available system cannot provide strong consistency guarantees in the presence of network partitions. This became a rallying call for a generation of highly scalable but semantically weak system designs. The NoSQL movement embraced this approach, offering production-use data platforms that eschewed strong consistency guarantees, but could scale-out to many servers \cite{21, 25, 33}. The consistency concept used in the CAP theorem came from the distributed systems community, and concerned reading the most recent value of a data item that is replicated across servers. The NoSQL movement implicitly used this to justify a lack of  transaction support, with either no way to group multiple operations  over multiple data items, or else restricting transactions to operations on items known to be colocated.

Serializability, the gold standard of traditional ACID transactions, is indeed not achievable with the requirement of high availability in the presence of partitions \cite{31}. However, database systems have a long tradition of providing weaker isolation and consistency guarantees \cite{4, 13, 43}. Due to their concurrency and overall performance benefits, weak isolation models are widely employed, overwhelmingly as the default setting in today’s traditional “ACID” and up-and-coming “NewSQL” databases and often as the maximum guarantee offered (Section 2.3). While weak isolation levels do not provide serializability for general-purpose transactions, they are apparently strong enough to deliver acceptable behavior to many application programmers. As they are stronger than the semantics of many distributed data platforms, this led to our recent proposal \cite{9} to offer \emph{Highly Available Transactions (HATs)}: general transactions with semantics similar to weak isolation databases, in ways that remain available in the face of partitions. We now build on \cite{9} by determining the partition-availability status of a wide range of previously identified isolation levels and other consistency conditions.

The relationship of the CAP Theorem and ACID semantics has been poorly understood. In the DB community, weak isolation is well studied in the single-server context from which it originated \cite{4, 13, 43} and many papers offer techniques for providing global serializability or snapshot isolation \cite{14,  30, 65; also cite several SI papers from PC members (Kemme, Elnikety, Salem, Daudgee)}. The distributed computing, programming language, and parallel hardware literature has produced many weak data consistency models for single replicated objects, or ways to build fully isolated transactions \cite{26, 28, also 45 (Herlihy-Shavit)}. Little attention has been given to weak isolation models on replicas of multiple items.

The contributions of this paper fill the gap between, on the one hand, the plethora of previously proposed isolation and consistency models, and, on the other hand, the goal of highly available systems. We introduce a hierarchy of consistency models; in doing so, we unify previous work on ACID properties \cite{4}, session guarantees \cite{64}, and traditional distributed “register” semantics \cite{45}. We rigorously taxonomize the availability characteristics of these consistency models: we determine which can be offered as Highly Available Transactions (HATs). In doing so, we argue that many implementations that provide these semantics are not partition-available, but that often the unavailability is not inherent to their semantic guarantees. Our investigation shows, somewhat surprisingly, that besides serializability, only fairly strong models like Snapshot Isolation (SI) and Repeatable Reads (RR) are \emph{inherently} unavailable, stemming from their requirement to detect conflict between concurrent operations as needed for preventing Lost Updates or Write Skew phenomena. In contrast, Read Committed isolation can be provided in a HAT system. We show that many of the weaker data consistency models from distributed systems can be implemented in a highly available manner as well, in a slightly different definition with clients which stick to (i.e. have affinity with) one server (something which is implicitly assumed in much of the previous work on distributed systems, and is sometimes reasonable in practice).

This taxonomy allows us to determine that a wide range of semantic guarantees can be delivered with availability in face of partitions, but which guarantees are worthwhile in practice? We attempt to objectively study the virtues and limitations of both classic and HAT systems by surveying practitioner accounts and research literature, performing experimental analyses on modern cloud infrastructure, and analyzing representative applications for their semantic requirements. While our results are not definitive, they suggest that HATs offer a one to three order of magnitude latency decrease compared to traditional distributed serializability protocols, and they can provide acceptable semantics for a wide range of programs, especially those with monotonic logic and commutative updates \cite{6, 24, 61}. HAT systems can also enforce arbitrary foreign key constraints on multi-item updates and sometimes provide limited uniqueness guarantees. However, particularly for programs with non-monotonic logic, HATs may fall short, requiring sometimes-unavailable mechanisms that are more coordination-intensive. Overall, this paper clarifies the benefits  Highly Available Transactions can offer and the restrictions they place on applications.

### 2013-03-31 writeup/hat-vldb.tex
alan intro

@@
 suffer from unavailability during system partitions or incur high
-network latency.  We introduce a new taxonomy of high availability and
-analyze a existing ACID and distributed consistency guarantees to
-identify which can and cannot be achieved in HAT systems. In doing so,
-we analytically and experimentally quantify the benefit of providing
-high availability as well as the associated semantic compromises.
+network latency.  We introduce a new taxonomy of highly available
+systems and analyze existing ACID and distributed consistency
+guarantees to identify which can and cannot be achieved in HAT
+systems. We analytically and experimentally quantify the performance
+and availability benefits of HAT systems as well as the associated
+semantic compromises.
 \end{abstract}
-
-In the debate over NoSQL systems versus traditional DBMS technologies for distributed data management, one key issue is that many NoSQL systems aim for availability in the face of partitions, and provide no, or restricted, support for transactions that group multiple operations over multiple data items.
-The authors recently advocated Highly Available Transaction (HAT) systems, where general transactions are provided with ACID-like semantic guarantees such as weak isolation levels, in ways that remain available during partitions and avoid high network latency.
-Here we introduce a new taxonomy of partition-availability properties.
-We analyse existing ACID and distributed consistency guarantees, showing the availability that can be achieved for each.
-We analytically and experimentally estimate the performance gains and semantic compromises of providing high availability, compared to traditional serializable transactions in an internet-scale setting.
 

### 2013-03-31 writeup/intro.tex
alan intro

@@
 
-In July 2000, the CAP Theorem~\cite{gilbert-cap} was crisply
-formulated that a highly available system cannot provide strong
-consistency guarantees in the presence of network partitions. This
-became a rallying call for a generation of highly scalable but
-semantically weak system designs. The NoSQL movement embraced this
-mantra, leading to many production quality databases that eschewed
-strong consistency guarantees, but could scale-out to many
-servers~\cite{bigtable, pnuts, dynamo}. The consistency concept used
-in the CAP theorem was proposed by the distributed systems community,
-and concerned providing strong recency guarantees of a data item that
-is replicated across servers.  However, the NoSQL movement implicitly
-took it to imply that transactions are also unavailable in the
-presence of network partitions. Over a decade later, we observe a
-dearth of highly available systems providing general-purpose
-multi-operation \textit{transactions} over multiple data items with
-useful semantics.
+Brewer's 2000 CAP Theorem stated that a highly available system cannot
+provide strong consistency guarantees in the presence of network
+partitions~\cite{gilbert-cap}. This result became a rallying call for
+a generation of highly scalable but semantically weak system
+designs. Most notably, the NoSQL movement embraced this approach,
+offering production-use data platforms that eschewed strong
+consistency guarantees but offered graceful scale-out to many
+servers~\cite{bigtable, pnuts, dynamo}. The consistency concept
+employed in the CAP Theorem came from the distributed systems
+community and concerned reading the most recent version of a data item
+that is replicated across servers.  However, the NoSQL movement
+implicitly used this fairly narrow result to justify a lack of
+general-purpose transactional functionality, or useful semantics for
+groups of operations over multiple data items~\cite{hn,
+  foundation-article}. Over a decade after CAP, we observe a dearth of
+highly available systems providing transactional functionality.
 
 Serializability, the gold standard of traditional ACID transactions,
- is indeed not achievable with the requirement of high availability
- the in the presence of partitions~\cite{davidson-survey}. However,
- database systems have a long tradition of providing weaker isolation
- and consistency guarantees~\cite{adya, ansicritique,
-   gray-isolation}. Due to their concurrency and overall performance
- benefits, weak isolation models are widely employed, overwhelmingly
- as the default setting in today's traditional ``ACID'' and
- up-and-coming ``NewSQL'' databases and often as the maximum guarantee
- offered (Section~\ref{sec:modernacid}). While this weak isolation
- does not provide serializability for general-purpose transactions, it
- is apparently strong enough to deliver acceptable behavior to many
- application programmers and is substantially stronger than the
- typical semantics of current highly available systems. This raises
- the question: which of these semantics can be provided while always
+is indeed not achievable given a requirement for high availability in
+the presence of partitions~\cite{davidson-survey}. However, database
+systems have a long tradition of providing weaker isolation and
+consistency guarantees~\cite{adya, ansicritique, gray-isolation}. Due
+to their concurrency and overall performance benefits, weak isolation
+models are widely employed, overwhelmingly as the default setting in
+today's traditional ``ACID'' and up-and-coming ``NewSQL'' databases
+and often as the maximum guarantee offered
+(Section~\ref{sec:modernacid}). While weak isolation levels do not
+provide serializability for general-purpose transactions, they are
+apparently strong enough to deliver acceptable behavior to many
+application programmers and are substantially stronger than the
+semantics of current highly available systems. This raises a natural
+question: which of these alaternative semantics can be provided while
 remaining highly available?
@@
 Unfortunately, the relationship of the CAP Theorem and ACID semantics
-is poorly understood. We have a strong understanding of weak isolation
-in the single-server context from which it originated~\cite{adya,
-  ansicritique, gray-isolation} as well as several techniques for
-providing distributed serializability~\cite{bernstein-concurrency,
-  spanner, granola, daudjee-session, calvin}.  However, the current
-literature does not provide guidance as to the operation of weakly
-consistent transactional guarantees in a replicated context. To make
-matters worse, the distributed computing literature has produced a
-large body of work on weak data consistency models for single
-replicated objects. Accordingly, in this work, we consider the problem
-of providing understanding semantic guarantees regarding multiple
-operations operating on data items that are achievable with high
-availability and low latency, albeit with some cost to achievable
-semantics. While our prior, preliminary results indicated that some
-models are achievable~\cite{hat-hotos}, the literature lacks a deep
-understanding of what models are available, which are not, and when
-this distinction matters.
+has been poorly understood. We have a strong understanding of weak
+isolation in the single-server context from which it
+originated~\cite{adya, ansicritique, gray-isolation} and many papers
+offer techniques for providing distributed
+serializability~\cite{bernstein-concurrency, spanner, granola,
+  daudjee-session, krikellas-bargain, calvin} or snapshot
+isolation~\cite{daudjee-snapshot,generalizedsnapshot, kemme-snapshot,
+  walter}. The distributed computing, programming language, and
+parallel hardware literature also contains many consistency models for
+single operations on replicated objects~\cite{pnuts, herlihy-art,
+  cops, eiger, cac, sessionguarantees}. However, the literature has
+largely ignored semantic guarantees for multiple operations operating
+on multiple data items in a highly available distributed environment.
 
-This work attempts to bridge the gap between, one the one hand, the
-plethora of isolation and consistency models proposed in the
-transactions and distributed systems literature, and, on the other hand
-highly available systems. We introduce a hierarchy of consistency
-models and rigorously taxonomize their availability characteristics;
-in doing so, we unify previous work on ACID properties~\cite{adya},
-session guarantees~\cite{sessionguarantees}, and traditional
-distributed ``register'' semantics~\cite{herlihy-art}.
+The contributions of this paper fill the gap between, on one hand, the
+plethora of previously proposed isolation and consistency models, and,
+on the other hand, the goal of high availability. We introduce a
+hierarchy of consistency models and, in doing so, unify previous work
+on ACID properties~\cite{adya}, session
+guarantees~\cite{sessionguarantees}, and traditional distributed
+``register'' semantics~\cite{herlihy-art}. We rigorously taxonomize
+the availability characteristics of these consistency models: we
+determine which can be offered as Highly Available Transactions
+(HATs). In doing so, we demonstrate that many \textit{implementations}
+that provide these semantics are not highly available but that their
+unavailability is not \textit{inherently} required.
 
-Our main contributions are the following. We classify among the wide
-array of isolation and consistency models those that are highly
-available, denoting them as {\em Highly Available Transactions
-  (HATs)}.  In doing so, we show that many implementations that
-provide HAT-semantics are not available, but that the unavailability
-is {\em not} inherent to their semantic guarantees. Thus, we provide
-highly available implementations for HATs.  Furthermore, our
-investigation shows, somewhat surprisingly, that besides
-serializability, only Snapshot Isolation (SI) and Repeatable Reads
-(RR) are unavailable. The reason for their unavailability stems from
-their requirement to avoid update conflict (preventing Lost Updates
-and Write Skew phenomena). Finally, we show that many of the weaker
-data consistency models from distributed systems can be implemented in
-a highly available manner as well. For that, the model of the system
-has to be changed to include clients, which cannot be mobile, but
-rather have to {\em stick} to one server, something which is implicit
-in much of the previous work on distributed systems.
+Our investigation shows, somewhat surprisingly, that besides
+serializability, only fairly strong models like Snapshot Isolation
+(SI) and Repeatable Reads (RR) are not highly available, stemming from
+their requirement to detect conflicts between concurrent updates (as
+needed for preventing Lost Updates or Write Skew phenomena). In
+contrast, Read Committed isolation, Transactional Atomicity, and many
+of the weaker data consistency models from database and distributed
+systems can be implemented in a highly available manner. For several
+of these guarantees, such as Transactional Causal Predicate Cut
+Isolation, we consider a modified form of high availiabity in which
+clients ``stick to'' (i.e. have affinity with) at least one server (a
+property which is implicit in the distributed systems literature but
+which requires \textit{explicit} consideration in a client-server
+replicated database context).
 
-This taxonomy allows us to weigh the costs and benefits of a wide
-range of semantic guarantees, but which guarantees are worthwhile in
-practice? We attempt to objectively study the virtues and limitations
-of both classic and HAT systems by surveying practitioner accounts and
-research literature, performing experimental analyses on modern cloud
-infrastructure, and analyzing representative applications for their
-semantic requirements. While our results are not definitive for all
-applications, they suggest that HATs offer a one to three order of
-magnitude latency decrease compared to traditional protocols and can
+This taxonomy allows us to determine that a wide range of semantic
+guarantees can be delivered with availability in face of partitions,
+but which guarantees are worthwhile in practice? We attempt to
+objectively study the virtues and limitations of both classic and HAT
+systems by surveying practitioner accounts and research literature,
+performing experimental analyses on modern cloud infrastructure, and
+analyzing representative applications for their semantic
+requirements. While our results are not definitive, they suggest that
+HATs offer a one to three order of magnitude latency decrease compared
+to traditional distributed serializability protocols, and they can
 provide acceptable semantics for a wide range of programs, especially
@@
 uniqueness guarantees. However, particularly for programs with
-non-monotonic logic, HATs may fall short, requiring more
-coordination-intensive, unavailable mechanisms. Concisely, we attempt
-to understand \textit{what benefits do Highly Available Transactions
-  offer and what is their cost?}
+non-monotonic logic, HATs may fall short, requiring
+sometimes-unavailable mechanisms that are more
+coordination-intensive. Overall, this paper quantifies the benefits
+Highly Available Transactions can offer as well as the necessary
+semantic restrictions they place on applications.
 

### 2013-03-31 writeup/futurework.tex
joe comments

@@
 a narrow case study in this work, but a general taxonomization of
-techniques such as escrow transactions~\cite{escrow} and
-Sagas~\cite{sagas} could help answer the challenge of programming in a
-weakly consistent environment. Additionally, HAT guarantees can be
-strengthened by considering failure models with, say, bounded
-asynchrony or service-level agreements from underlying
+techniques such as escrow transactions~\cite{escrow},
+Sagas~\cite{sagas}, and other long-running transaction
+support~\cite{gray-virtues} could help answer the challenge of
+programming in a weakly consistent environment. Additionally, HAT
+guarantees can be strengthened by considering failure models with,
+say, bounded asynchrony or service-level agreements from underlying
 hardware. Additionally, while we have focused on systems that provide

### 2013-03-31 writeup/intro.tex
joe comments

@@
 
-Brewer's 2000 CAP Theorem stated that a highly available system cannot
-provide strong consistency guarantees in the presence of network
-partitions~\cite{gilbert-cap}. This result became a rallying call for
-a generation of highly scalable but semantically weak system
-designs. Most notably, the NoSQL movement embraced this approach,
-offering production-use data platforms that eschewed strong
-consistency guarantees but offered graceful scale-out to many
-servers~\cite{bigtable, pnuts, dynamo}. The consistency concept
-employed in the CAP Theorem came from the distributed systems
-community and concerned reading the most recent version of a data item
-that is replicated across servers.  However, the NoSQL movement
-implicitly used this fairly narrow result to justify a lack of
-general-purpose transactional functionality, or useful semantics for
-groups of operations over multiple data items~\cite{hn,
-  foundation-article}. Over a decade after CAP, we observe a dearth of
-highly available systems providing transactional functionality.
+The prior decade saw enormous churn in the design of widely-deployed
+database backends, from the de facto use of transactional
+RDBMSs~\cite{bernstein-concurrency, bernstein-book, gray-isolation,
+  gray-virtues} to the widespread adoption of loosely-consistent
+distributed key-value stores~\cite{bigtable, pnuts, dynamo}, with
+increasing interest in returning to transactional
+designs~\cite{spanner, walter, foundation-article, krikellas-bargain,
+  eiger}. Instrumental in these trends was the 2000 introduction of
+Brewer's CAP Theorem, which stated that a highly available system
+cannot provide strong consistency guarantees in the presence of
+network partitions~\cite{brewer-slides}. For distributed systems
+designers with requirements for ``always on,'' highly available and
+low latency operation, CAP became an easily understandable indication
+of the competing goals of semantic strength and availability. However,
+as formally proven~\cite{gilbert-cap}, the CAP Theorem concerns
+linearizability, the ability to read the most recent write to a data
+item that is replicated across servers. Yet this result---regarding a
+fairly narrow form of distributed consistency---is often conflated
+with ACID database properties~\cite{brewer-slides, hn,
+  foundation-article}, leading to substantial confusion about the
+relationships between replica consistency, transactional isolation, and high
+availability.
 
@@
 systems have a long tradition of providing weaker isolation and
-consistency guarantees~\cite{adya, ansicritique, gray-isolation}. Due
-to their concurrency and overall performance benefits, weak isolation
-models are widely employed, overwhelmingly as the default setting in
-today's traditional ``ACID'' and up-and-coming ``NewSQL'' databases
-and often as the maximum guarantee offered
+consistency guarantees~\cite{adya, ansicritique, gray-isolation,
+  gray-virtues}. Due to their concurrency and overall performance
+benefits, weak isolation models are widely employed, overwhelmingly as
+the default setting in today's traditional ``ACID'' and up-and-coming
+``NewSQL'' databases and often as the maximum guarantee offered
 (Section~\ref{sec:modernacid}). While weak isolation levels do not
@@
 Unfortunately, the relationship of the CAP Theorem and ACID semantics
-has been poorly understood. We have a strong understanding of weak
+has not been well explored. We have a strong understanding of weak
 isolation in the single-server context from which it
@@
 clients ``stick to'' (i.e. have affinity with) at least one server (a
-property which is implicit in the distributed systems literature but
-which requires \textit{explicit} consideration in a client-server
-replicated database context).
+property which is often implicit in the distributed systems
+literature~\cite{herlihy-art, cops, eiger, cac} but which requires
+\textit{explicit} consideration in a client-server replicated database
+context). This sticky availability is easily implementatable and
+widely employed~\cite{cops, eiger, vogels-defs} but is strictly less
+available than traditional definitions of high availability.
 

### 2013-03-31 writeup/alig.tex
checkpoint

For that, the model of the system has to be changed to include
clients, which cannot be mobile, but rather have to {\em stick} to one
server, something which is implicit in much of the previous work on
distributed systems.

This taxonomy allows us to weigh the costs and benefits of a wide
range of semantic guarantees, but which guarantees are worthwhile in
practice? We attempt to objectively study the virtues and limitations
of both classic and HAT systems by surveying practitioner accounts and
research literature, performing experimental analyses on modern cloud
infrastructure, and analyzing representative applications for their
semantic requirements. While our results are not definitive for all
applications, they suggest that HATs offer a one to three order of
magnitude latency decrease compared to traditional protocols and can
provide acceptable semantics for a wide range of programs, especially
those with monotonic logic and commutative updates~\cite{calm, blooml,
  crdt}. HAT systems can also enforce arbitrary foreign key
constraints on multi-item updates and sometimes provide limited
uniqueness guarantees. However, particularly for programs with
non-monotonic logic, HATs may fall short, requiring more
coordination-intensive, unavailable mechanisms. Concisely, we attempt
to understand \textit{what benefits do Highly Available Transactions
  offer and what is their cost?}

### 2013-03-31 writeup/evaluation.tex
checkpoint

@@
-\section{Application and System Analysis}
+\section{HAT Implications}
 \label{sec:evaluation}
@@
 
-\vspace{.5em}\noindent\textbf{Serializablity} To establish a serial
+\vspace{.5em}\noindent\textbf{Serializability} To establish a serial
 order on transactions, algorithms for achieving serializability of
@@
 
-\vspace{.5em}\noindent\textbf{Non-serializablity} Many existing
+\vspace{.5em}\noindent\textbf{Non-serializability} Many existing
 distributed implementations of weak isolation are not highly

### 2013-03-31 writeup/hat-analysis.tex
checkpoint

@@
 
-When possible, we draw on existing properties and definitions from the
-database and distributed systems literature, providing a brief,
-informal explanation and example for each guarantee. In particular,
-for our ACID guarantees, we draw largely on Atul Adya's
+As Brewer states, ``systems and database communities are separate but
+overlapping (with distinct vocabulary)''~\cite{brewer-slides}. With
+this challenge in mind, we draw on existing properties and definitions
+from the database and distributed systems literature, providing a
+brief, informal explanation and example for each guarantee. In
+particular, for our ACID guarantees, we draw largely on Atul Adya's
 dissertation~\cite{adya}, Berenson et al.'s 1995 critique of the ANSI

### 2013-03-31 writeup/intro.tex
checkpoint

@@
 
-The prior decade saw enormous churn in the design of widely-deployed
-database backends, from the de facto use of transactional
+The prior decade saw a large shift in the design of large-scale
+database backends, from the use of transactional
 RDBMSs~\cite{bernstein-concurrency, bernstein-book, gray-isolation,
   gray-virtues} to the widespread adoption of loosely-consistent
-distributed key-value stores~\cite{bigtable, pnuts, dynamo}, with
-increasing interest in returning to transactional
-designs~\cite{spanner, walter, foundation-article, krikellas-bargain,
-  eiger}. Instrumental in these trends was the 2000 introduction of
+distributed key-value stores~\cite{bigtable, pnuts,
+  dynamo}. Instrumental in this trend was the 2000 introduction of
 Brewer's CAP Theorem, which stated that a highly available system
 cannot provide strong consistency guarantees in the presence of
-network partitions~\cite{brewer-slides}. For distributed systems
-designers with requirements for ``always on,'' highly available and
-low latency operation, CAP became an easily understandable indication
-of the competing goals of semantic strength and availability. However,
-as formally proven~\cite{gilbert-cap}, the CAP Theorem concerns
-linearizability, the ability to read the most recent write to a data
-item that is replicated across servers. Yet this result---regarding a
-fairly narrow form of distributed consistency---is often conflated
-with ACID database properties~\cite{brewer-slides, hn,
-  foundation-article}, leading to substantial confusion about the
-relationships between replica consistency, transactional isolation, and high
-availability.
+network partitions~\cite{brewer-slides}. However, as formally
+proven~\cite{gilbert-cap}, the CAP Theorem concerns
+linearizability~\cite{herlihy-art}, the ability to read the most
+recent write to a data item that is replicated across servers. Yet
+this result---regarding a fairly narrow form of distributed
+consistency---is often conflated with ACID database
+properties~\cite{brewer-slides, hn, foundation-article}, leading to
+substantial confusion about the relationships between replica
+consistency, transactional isolation, and high availability. However,
+transactions \textit{are} useful for programmers as evidenced by the
+recent resurgence of transactional data stores such as Google's
+Spanner~\cite{spanner}. Unfortunately, these systems are not highly
+available in the presence of partitions~\cite{walter, mdcc,
+  foundation-article, krikellas-bargain, eiger}. (TODO:ADD more)
 
-Serializability, the gold standard of traditional ACID transactions,
-is indeed not achievable given a requirement for high availability in
-the presence of partitions~\cite{davidson-survey}. However, database
-systems have a long tradition of providing weaker isolation and
-consistency guarantees~\cite{adya, ansicritique, gray-isolation,
-  gray-virtues}. Due to their concurrency and overall performance
+Indeed, serializable transactions, the gold standard of traditional
+ACID databases, is indeed not achievable given a requirement for high
+availability in the presence of
+partitions~\cite{davidson-survey}. However, database systems have a
+long tradition of providing weaker isolation and consistency
+guarantees~\cite{adya, ansicritique, gray-isolation,
+  gray-virtues} (TODO: kemme). Due to their concurrency and overall performance
 benefits, weak isolation models are widely employed, overwhelmingly as
@@
 
-The contributions of this paper fill the gap between, on one hand, the
-plethora of previously proposed isolation and consistency models, and,
-on the other hand, the goal of high availability. We introduce a
-hierarchy of consistency models and, in doing so, unify previous work
-on ACID properties~\cite{adya}, session
-guarantees~\cite{sessionguarantees}, and traditional distributed
-``register'' semantics~\cite{herlihy-art}. We rigorously taxonomize
-the availability characteristics of these consistency models: we
-determine which can be offered as Highly Available Transactions
-(HATs). In doing so, we demonstrate that many \textit{implementations}
-that provide these semantics are not highly available but that their
-unavailability is not \textit{inherently} required.
+Our main contributions are as follows. We fill gap between, on one
+hand, the plethora of previously proposed isolation and consistency
+models, and, on the other hand, the goal of high availability.  We
+classify which, among the wide array models, are highly available,
+denoting them as {\em Highly Available Transactions (HATs)}. In doing
+so, we demonstrate that many \textit{implementations} of HAT semantics
+are not highly available but that their unavailability is not
+\textit{inherently} required. Our investigation shows, somewhat
+surprisingly, that besides serializability, only Snapshot Isolation
+(SI) and Repeatable Reads (RR) are unavailable. We also show that many
+of the weaker data consistency models from distributed systems are
+HAT-compliant.
 
-Our investigation shows, somewhat surprisingly, that besides
-serializability, only fairly strong models like Snapshot Isolation
-(SI) and Repeatable Reads (RR) are not highly available, stemming from
-their requirement to detect conflicts between concurrent updates (as
-needed for preventing Lost Updates or Write Skew phenomena). In
-contrast, Read Committed isolation, Transactional Atomicity, and many
-of the weaker data consistency models from database and distributed
-systems can be implemented in a highly available manner. For several
-of these guarantees, such as Transactional Causal Predicate Cut
-Isolation, we consider a modified form of high availiabity in which
-clients ``stick to'' (i.e. have affinity with) at least one server (a
-property which is often implicit in the distributed systems
-literature~\cite{herlihy-art, cops, eiger, cac} but which requires
-\textit{explicit} consideration in a client-server replicated database
-context). This sticky availability is easily implementatable and
-widely employed~\cite{cops, eiger, vogels-defs} but is strictly less
-available than traditional definitions of high availability.
+More precisely, highly available systems cannot detect conflicts
+between concurrent updates (as needed for preventing Lost Updates or
+Write Skew phenomena). However, Read Committed isolation,
+Transactional Atomicity, and many of the weaker data consistency
+models from database and distributed systems can be implemented in a
+highly available manner. For several of these guarantees, such as
+Transactional Causal Predicate Cut Isolation, we consider a modified
+form of high availiabity in which clients ``stick to'' (i.e., have
+affinity with) at least one server (a property which is often implicit
+in the distributed systems literature~\cite{herlihy-art, cops, eiger,
+  cac} but which requires \textit{explicit} consideration in a
+client-server replicated database context). This sticky availability
+is easily implementatable and widely employed~\cite{cops, eiger,
+  vogels-defs} but is strictly less available than traditional
+definitions of high availability.
 

### 2013-03-31 writeup/motivation.tex
Couple very minor fixes
checkpoint

@@
-\section{Motivation}
+\section{Partitions and Latency}
 \label{sec:motivation}
@@
 \begin{table}
-\subfloat[Cross-region (OR:~Oregon, VA:~Virginia, TO:~Tokyo, IR:~Ireland, SY:~Sydney, SP:~S\~{a}o Paulo, SI:~Singapore)] {
+\subfloat[Cross-region (CA:~California, OR:~Oregon, VA:~Virginia, TO:~Tokyo, IR:~Ireland, SY:~Sydney, SP:~S\~{a}o Paulo, SI:~Singapore)] {
   \begin{tabular}{|c|c|c|c|c|c|c|c|c|}
@@
 
-\subsection{CAP, Isolation, and Modern Systems}
+\section{CAP and Modern Systems}
 \label{sec:modernacid}
@@
 One of the most popularly celebrated impossibility results in
-distributed systems is Brewer's CAP Theorem, which, as formally proven
-by Gilbert and Lynch, states that it is impossible to maintain
-linearizability , or the ability to read the last completed write,
-(``C,'' or ``consistency''---albeit poorly named) and replica
-availability, or a guaranteed response, (``A'') in the presence of
-network partitions (``P'')~\cite{gilbert-cap}. Brewer's Theorem has
-roots in decades of distributed database
+distributed systems is Brewer's CAP Theorem~\cite{brewer-slides},
+which, as formally proven by Gilbert and Lynch~\cite{gilbert-cap},
+states that it is impossible to maintain linearizability , or the
+ability to read the last completed write, (``C,'' or
+``consistency''---albeit poorly named) and replica availability, or a
+guaranteed response, (``A'') in the presence of network partitions
+(``P''). Brewer's Theorem has roots in decades of distributed database
 research~\cite{davidson-survey} and has had a substantial impact on
-recent large-scale distributed systems design: while CAP is narrowly
-scoped to the impossibility of building highly available linearizable
-systems, it is often interpreted to preclude a larger set of
-guarantees, including ACID (transactional atomicity, consistency,
-isolation and durability guarantees)~\cite{hn, foundation-article} as
-also evidenced by the lack of highly available systems providing them.
+recent large-scale distributed systems design~\cite{bigtable, pnuts,
+  dynamo}: while CAP is narrowly scoped to the impossibility of
+building highly available linearizable systems, it is often
+interpreted to preclude a larger set of guarantees, including ACID
+(transactional atomicity, consistency, isolation and durability
+guarantees). We are unlikely to determine the exact cause of this
+confusion, but it appears as early as the first public announcement of
+Brewer's result, in which ''distributed databases'' appear as an
+example of unavailable systems~\cite{brewer-slides}. While, in the
+same document, Brewer is careful to note that consistency and
+availability are a ``spectrum,'' all of ACID is still often labeled
+with a binary categorization of ``unavailable''~\cite{hn,
+  foundation-article}.
 
@@
 availability~\cite{davidson-survey} and weak isolation on a
-single-server~\cite{adya, ansicritique} but never weak isolation and
+single server~\cite{adya, ansicritique, gray-isolation} but never weak isolation and
 high availability taken together.

### 2013-03-31 writeup/relatedwork.tex
checkpoint

@@
 availability and represent a new class of (sticky) HAT systems.
+
+Wisemann et al. propose a three parameter classification for one-copy
+serializable database systems, performing a more extensive
+classification than our overview in Setion XX, including eager
+database replication~\cite{kemme-classification}. Additional, several
+surveys discuss distributed mechanisms for achieving one-copy
+serializability~\cite{wisemann-survey}, linearizability, consistent
+snapshots, and recency bounds~\cite{ceri-mechanism, chen-mechanism}.
+
+% Ceri discuss distributed database update mechanisms with respect to
+% linearizability, consistent snapshots (i.e., cut isolation), and
+% recency bounds~\cite{ceri-mechanism}.  Chen and Pu also classify
+% distributed replica maintenance from perspectives of
+% linearizability,% recency guarantees, and replica
+% divergence~\cite{chen-mechanism}.
+
+, with increasing interest in returning to transactional
+designs~\cite{spanner, walter, foundation-article, krikellas-bargain,
+  eiger}

### 2013-03-31 writeup/intro.tex
new intro

@@
 surprisingly, that besides serializability, only Snapshot Isolation
-(SI) and Repeatable Reads (RR) are unavailable. We also show that many
-of the weaker data consistency models from distributed systems are
-HAT-compliant.
+(SI) and Repeatable Reads (RR) are unavailable, with most other
+isolation levels being available. We also show that many of the weaker
+data consistency models from distributed systems are HAT-compliant.
 
-More precisely, highly available systems cannot detect conflicts
-between concurrent updates (as needed for preventing Lost Updates or
-Write Skew phenomena). However, Read Committed isolation,
-Transactional Atomicity, and many of the weaker data consistency
-models from database and distributed systems can be implemented in a
-highly available manner. For several of these guarantees, such as
-Transactional Causal Predicate Cut Isolation, we consider a modified
-form of high availiabity in which clients ``stick to'' (i.e., have
-affinity with) at least one server (a property which is often implicit
-in the distributed systems literature~\cite{herlihy-art, cops, eiger,
-  cac} but which requires \textit{explicit} consideration in a
-client-server replicated database context). This sticky availability
-is easily implementatable and widely employed~\cite{cops, eiger,
-  vogels-defs} but is strictly less available than traditional
-definitions of high availability.
+More precisely, the reason why fairly strong models like Snapshot
+Isolation (SI) and Repeatable Reads (RR) are not highly available is
+that they require detecting conflicts between concurrent updates (as
+needed for preventing Lost Updates or Write Skew phenomena). However,
+Read Committed isolation, Transactional Atomicity, and many of the
+weaker data consistency models from database and distributed systems
+can be implemented in a highly available manner. For several of these
+guarantees, such as Transactional Causal Predicate Cut Isolation, we
+consider a modified form of high availiabity in which clients ``stick
+to'' (i.e., have affinity with) at least one server (a property which
+is often implicit in the distributed systems
+literature~\cite{herlihy-art, cops, eiger, cac} but which requires
+\textit{explicit} consideration in a client-server replicated database
+context). This sticky availability is easily implementatable and
+widely employed~\cite{cops, eiger, vogels-defs} but is strictly less
+available than traditional definitions of high availability.
 

### 2013-03-31 writeup/intro.tex
typos

@@
 semantics of current highly available systems. This raises a natural
-question: which of these alaternative semantics can be provided while
+question: which of these alternative semantics can be provided while
 remaining highly available?
@@
 guarantees, such as Transactional Causal Predicate Cut Isolation, we
-consider a modified form of high availiabity in which clients ``stick
+consider a modified form of high availability in which clients ``stick
 to'' (i.e., have affinity with) at least one server (a property which
@@
 \textit{explicit} consideration in a client-server replicated database
-context). This sticky availability is easily implementatable and
+context). This sticky availability is easily implementable and
 widely employed~\cite{cops, eiger, vogels-defs} but is strictly less

### 2013-03-31 writeup/intro.tex
more writing

--- 2013-03-31 writeup/intro.tex
@@
 offer techniques for providing distributed
-serializability~\cite{bernstein-concurrency, spanner, granola,
-  daudjee-session, krikellas-bargain, calvin} or snapshot
-isolation~\cite{daudjee-snapshot,generalizedsnapshot, kemme-snapshot,
-  walter}. The distributed computing, programming language, and
-parallel hardware literature also contains many consistency models for
-single operations on replicated objects~\cite{pnuts, herlihy-art,
-  cops, eiger, cac, sessionguarantees}. However, the literature has
-largely ignored semantic guarantees for multiple operations operating
-on multiple data items in a highly available distributed environment.
+serializability~\cite{bernstein-concurrency, kemme-classification,
+  spanner, granola, daudjee-session, krikellas-bargain, calvin} or
+snapshot isolation~\cite{daudjee-snapshot,generalizedsnapshot,
+  kemme-snapshot, walter}. The distributed computing, programming
+language, and parallel hardware literature also contains many
+consistency models for single operations on replicated
+objects~\cite{pnuts, herlihy-art, cops, eiger, cac, sessionguarantees,
+  ceri-mechanism, chen-mechanism}. However, the literature has largely
+ignored semantic guarantees for multiple operations operating on
+multiple data items in a highly available distributed environment.
 

### 2013-03-31 writeup/relatedwork.tex
more writing

@@
 \section{Related Work}
-
-Although we are not aware of any related work specifically addressing
-what properties are available in a highly available system, there are
-several systems and related research results upon which we draw in
-this work.
 
@@
 consistency'' and low latency~\cite{abadi-pacelc}, while several
-studies~\cite{adya, ansicritique} examine weak isolation guarantees,
-grounded in a single-server context. The connection between weak
-consistency models in a distributed setting~\cite{calm, crdt} and high
-availability is relatively unexplored. With the exception of a
-preliminary workshop paper on this work~\cite{hat-hotos}---upon which
-this work builds, providing impossibility results, a full
-taxonomization, experimental and application analysis (basically all
-but Section 4.1)---we believe this paper is the first to explore this
-area.
+studies examine weak isolation guarantees~\cite{adya,
+  ansicritique}. However, as we have discussed, the connection between
+weak consistency models in a distributed setting~\cite{calm, crdt} and
+high availability is relatively unexplored. With the exception of a
+preliminary workshop paper introducing HATs~\cite{hat-hotos}--- which
+this work expands, providing impossibility results, a full
+taxonomization, and experimental and application analysis (basically
+all but Section 4.1)---we believe this paper is the first to explore
+this area. Nonetheless, several studies rigorously classify non-HAT
+semantics: Wisemann et al. have proposed a three parameter
+classification for one-copy serializable database systems, performing
+a more extensive classification than our overview in
+Section~\ref{sec:evaluation}, including eager database replication
+techniques~\cite{kemme-classification}. Additionally several earlier
+surveys discuss distributed mechanisms for achieving one-copy
+serializability~\cite{wisemann-survey}, linearizability, consistent
+snapshots, and recency bounds~\cite{ceri-mechanism, chen-mechanism}.
 
@@
 industry~\cite{megastore, orleans, spanner}. As discussed in
-Section~\ref{sec:motivation}, classic ACID databases provide strong
+Section~\ref{sec:modernacid}, classic ACID databases provide strong
 semantics but their lock-based and traditional multi-versioned
@@
 partitions~\cite{bernstein-concurrency, bernstein-book,
-  gray-isolation}. Similar semantics are common in recent
-work. Notably, Google's Spanner provides externally serializable
-transactions. While Spanner fits Google's read-mostly workload well,
-read/write transactions use two-phase locking (incurring at least one
-round trip time over wide-area networks), are unavailable to clients
-in a non-majority partition, and cannot read their
-writes~\cite{spanner}. These latency and availability penalties are
-fundamental to providing strong consistency. For users willing to
+  gray-isolation}. Notably, Google's Spanner provides strong one-copy
+serializable transactions. While Spanner is highly specialized for
+Google's read-heavy workload, it relies on two-phase commit and
+two-phase locking for read/write transactions~\cite{spanner}. As we
+have discussed, the latency penalties associated with this design
+choice are fundamental to serializability. For users willing to
 tolerate these costs, Spanner, or similar strongly consistent,
 unavailable systems---including Calvin~\cite{calvin},
-G-Store~\cite{gstore}, Gemini~\cite{redblue}, Granola~\cite{granola},
+G-Store~\cite{gstore}, Gemini~\cite{redblue}, Generalized Snapshot
+Isolation~\cite{generalizedsnapshot}, Granola~\cite{granola},
 HBase~\cite{hbase}, MDCC~\cite{mdcc}, Megastore~\cite{megastore},
-Orleans~\cite{orleans}, and Walter~\cite{walter}---are a reasonable
-choice. With HATs, we seek an alternative set of semantics that are
+Orleans~\cite{orleans}, Postgres-R~\cite{kemme-thesis}, and
+Walter~\cite{walter}---are a reasonable choice. We note that many
+replication techniques target variants of Snapshot
+Isolation~\cite{middleware-db}.
+
+With HATs, we seek an alternative set of semantics that are
 still useful but do not violate requirements for high availability or
@@
 
-Wisemann et al. propose a three parameter classification for one-copy
-serializable database systems, performing a more extensive
-classification than our overview in Setion XX, including eager
-database replication~\cite{kemme-classification}. Additional, several
-surveys discuss distributed mechanisms for achieving one-copy
-serializability~\cite{wisemann-survey}, linearizability, consistent
-snapshots, and recency bounds~\cite{ceri-mechanism, chen-mechanism}.
+Middleware-based Database Replication: The Gaps Between Theory and Practice
+Most uses Snapshot Isolation
 

### 2013-03-31 writeup/conclusion.tex
PDB pass plus Alan comments

--- 2013-03-17 writeup/conclusion.tex
@@
 
-VIII. Conclusion (.5pg)
+In this paper, we have explored the relationship between
+multi-operation, multi-object transactions and high availability. We
+have outlined requirements for high availability in a transactional
+context, or Highly Available Transactions (HATs) and taxonomized
+guarantees spanning both distributed systems and ACID database
+literature. Many transactional guarantees are achievable with high
+availability, while a few, like preventing Lost Update and Write Skew,
+are not. To understand the benefits of high availability and low
+latency, we classified the prevalence and cost of partitions and
+network latency both in the literature and via experiment, analyzing
+existing transactions to determine whether HATs were sufficient to
+ensure application consistency, and validated intuitions about the
+performance benefits of HAT on public cloud infrastructure.
 
-HATS represent a new class of database system:
-* Can't get serializability, but can get other properties
-* Key challenge: designing distributed algorithms and architectures for partial failure, asynchrony
-* Rearchitecture of DBMSs and better understanding of applications are key
-
-References (1pg)
+Highly available operation represents a serious but often overlooked
+consideration in distributed database designs. While the CAP Theorem
+has had a large impact on highly available distributed data storage
+design, our treatment of high availability and low latency in
+transactional database systems has not kept pace. The key distributed
+systems challenge of designing distributed algorithms and
+architectures for partial failure and network asynchrony is largely
+unaddressed in the literature on transaction processing systems. Fully
+realizing the potential of HAT systems and bringing transactional
+functionality to a large class of scalable data stores will require
+further rearchitecture and study of existing database designs as well
+as better understanding of application semantics and their
+relationship to distributed consistency.

### 2013-03-31 writeup/evaluation.tex
PDB pass plus Alan comments

@@
 
-We now have an understanding of which guarantees are and are not
-HAT-compliant. In this section, we begin to understand the impact of
-this high availability on existing systems and applications as well as
-briefly demonstrate HAT algorithms on public cloud
-infrastructure. Specifically:
+With an understanding of which guarantees are and are not
+HAT-compliant, in this section, we analyze the implications of these
+results for existing systems and applications and briefly study HAT
+systems on public cloud infrastructure. Specifically:
 
@@
 environment, each of these operations requires coordination, either
-with other database replicas or with a lock service. If the
-coordination mechanism for locking is unavailable, transactions cannot
+with other database replicas or with a lock service. If this
+coordination mechanism is unavailable, transactions cannot safely
 commit. Similarly, optimistic concurrency control requires
@@
 scheduler. Serializability under multi-version concurrency control
-also requires checking for update conflicts.
-
-In summary, perhaps unsurprisingly, the reliance on a globally agreed
-total order necessitates a minimum of one round-trip to a designated
-master or coordination service for each of these classic algorithms.
-This is intuitively a consequence of serializability's lack of high
-availability~\cite{davidson-survey} (or, as we have shown, reliance on
-preventing Lost Update and Write Skew).  The cost of the (minimum one)
-round trip will be determined by the deployment environment, as we saw
-in Section~\ref{sec:motivation}.
+also requires checking for update conflicts. The reliance on a
+globally agreed total order necessitates a minimum of one round-trip
+to a designated master or coordination service for each of these
+classic algorithms.  The cost of the (minimum one) round trip will be
+determined by the deployment environment, as we saw in
+Section~\ref{sec:motivation}; we will demonstrate this cost on public
+cloud infrastructure in the next section.
 
@@
 of transactional guarantees, we have not seen proposals for truly weak
-isolation (e.g., non-``tentative update'' schemes) in this
-context. Chan and Gray's read-only transactions provide read-only
-transactions with item-cut isolation, causal consistency, and
-transactional atomicity (session PL-2L) but are unavailable in the
-presence of coordinator failure~\cite{readonly}, similar to read-only
-and write-only transactions more recently proposed by
-Eiger~\cite{eiger}.  Bolt-on causal consistency~\cite{bolton},
+isolation (e.g., non-``tentative update'' schemes) in this context.
+The MDCC~\cite{mdcc} protocol offers Read Committed isolation with
+Lost Update avoidance but is similarly unavailable due to its reliance
+on preventing write conflicts. Chan and Gray's read-only transactions
+provide read-only transactions with item-cut isolation, causal
+consistency, and transactional atomicity (session PL-2L~\cite{adya})
+but are unavailable in the presence of coordinator
+failure~\cite{readonly}, similar to read-only and write-only
+transactions more recently proposed by Eiger~\cite{eiger}; Causal
+Serializability offers a similar model (with unavailable
+implementation): causal consistency with a variant of Read Uncommitted
+between transactions that write to the same data
+item~\cite{raynal-causal}.  Bolt-on causal consistency~\cite{bolton},
 Brantner's S3 database~\cite{kraska-s3}, and
@@
 without substantial modification; Swift~\cite{swift} is closest to
-providing full HAT operation in a sticky context. Similarly, the
-MDCC~\cite{mdcc} protocol offers Read Committed isolation with Lost
-Update avoidance but is similarly unavailable due to its reliance on
-preventing write conflicts.
-
-As we have seen, it is possible to implement many guarantees weaker
-than serializability---including guarantees that are achievable with
-high availability---and still not achieve high availability.
+providing full HAT operation in a sticky context. As we have seen, it
+is possible to implement many guarantees weaker than
+serializability---including guarantees that are achievable with high
+availability---and still not achieve high availability.
 
@@
 bounds. In this section, we attempt to understand when these
-guarantees matter both abstractly and by the use of a representative
-customer order transaction application patterned off of the TPC-C
-benchmark~\cite{tpcc}.
+guarantees matter both abstractly and in a representative
+transactional application based on the TPC-C benchmark~\cite{tpcc}.
 
@@
 then they can often be safely performed in different orders at
-different locations in a distributed system. Accordingly, as long as
-all writes are delivered to all replicas, then a system executing
-monotonic logic with commutative operators may not suffer from
-application-level consistency anomalies as a result of Lost Update or
-Write Skew anomalies. However, applications with non-monotonic state
-mutation will not, in general, be able to maintain application-level
-consistency constraints with HATs alone. Similarly, applications
-requiring write visibility between clients should opt for
-unavailability instead of possibly missing a write visibility
-deadline.
+different replicas. Accordingly, as long as all writes are delivered
+to all replicas, then a system executing monotonic logic with
+commutative operators may not suffer from application-level
+consistency anomalies as a result of Lost Update or Write Skew
+anomalies. However, applications with non-monotonic state mutation
+will not, in general, be able to maintain application-level
+consistency constraints with HATs alone. Moreover, applications with
+requiring bounded update visibility latency should opt for
+unavailability.
 
-\vspace{.5em}\noindent\textbf{TPC-C} Absent a necessary and sufficient
-result describing exactly which application logics can be achieved
-with high availability and to make these requirements more concrete,
-we consider an example application: the TPC-C benchmark. Our goal is
-\textit{not} to provide a performance analysis, as a HAT
-configuration will be non-compliant. Rather, we examine the five
-queries contained in the benchmarks---ostensibly representative of a
-standard transaction processing application---to understand which are
-achievable in a HAT context. Surprisingly, four of five can be
-executed with HATs, while the fifth may require unavailablity.
+\vspace{.5em}\noindent\textbf{TPC-C} To better understand the impact
+of HAT-compliance in an application context, we consider a concrete
+application: the TPC-C benchmark. Surprisingly, four of five
+transactions can be executed with HATs, while the fifth may require
+unavailablity.
 
-TPC-C consists of five transactions, capturing operation of a
+TPC-C consists of five transactions, capturing the operation of a
 wholesale warehouse, including sales, payments, and deliveries. Two
@@
 warehouses, districts, and customer records as well as providing an
-audit trail. The transaction is increment-only, so all balance
-increase operations can commute, and TA allows the maintainence of any
-foreign-key constraints (e.g., \texttt{UPDATE/DELETE CASCADE}).
+audit trail. The transaction is increment- and append-only, so all
+balance increase operations commute, and TA allows the
+maintainence of any foreign-key constraints (e.g.,
+\texttt{UPDATE/DELETE CASCADE}).
 
-While three out of five transactions have been easily achievable with
+While three out of five transactions are easily achievable with
 HATs, the remaining two transactions---\textit{New-Order} and
-\textit{Delivery}---are not so simple. The New-Order transaction
+\textit{Delivery}---are not as simple. The New-Order transaction
 places an order for a variable quantity of data items, updating
@@
 
-There are three challenges that only allow HATs to place non-compliant
-New-Orders and to fail on Payments. First, New-Order requires a unique
-ID number for each order. We can create a unique number by, say,
+New-Order presents two challenges: ID assignment and stock
+maintainence. First, each New-Order transaction requires a unique ID
+number for the order. We can create a unique number by, say,
 concatenating the client ID and a timestamp. However, the TPC-C
 specification requires order numbers to be \textit{sequentially}
-assigned, which requires preventing Lost Update. Accordingly, HATs
-cannot provide compliant TPC-C execution but can maintain uniqueness
-constraints. Second, the New-Order transaction decrements inventory
-counts: what if the count becomes negative? Fortunately, TPC-C
-New-Order restocks each item's inventory count (increments by 91) if
-it would become negative as the result of placing an order. This means
-that, even in the presence of concurrent New-Orders, an item's stock
-will never fall below zero. This is TPC-C compliant, but a HAT system
-might end up with substantially more stock than in a single-site
-implementation. Third, delivery events are non-monotonic---deleting a
-pending New-Order transaction---and should be idempotent in order to
-avoid billing a customer twice; this implies a need to prevent Lost
-Update. We can avoid this issue by moving the non-monotonicity to the
-real world---the carrier that picks up the package for an order can
-ensure that she is the only carrier who has done so---but cannot
-provide a correct execution with HATs alone. However, according to
-distributed transaction architects~\cite{helland}, these compensatory
-actions are relatively common in real-world business processes.
+assigned within a district, which requires preventing Lost
+Update. Accordingly, HATs cannot provide compliant TPC-C execution but
+can maintain uniqueness constraints. Second, the New-Order transaction
+decrements inventory counts: what if the count becomes negative?
+Fortunately, TPC-C New-Order restocks each item's inventory count
+(increments by 91) if it would become negative as the result of
+placing an order. This means that, even in the presence of concurrent
+New-Orders, an item's stock will never fall below zero. This is TPC-C
+compliant, but a HAT system might end up with substantially more stock
+than in a non-HAT-compliant implementation.
+
+Delivery is challenging due to non-monotonicity. Each Delivery deletes
+a pending order from the NewOrder Table and should be idempotent in
+order to avoid billing a customer twice; this implies a need to
+prevent Lost Update. We can avoid this issue by moving the
+non-monotonicity to the real world---the carrier that picks up the
+package for an order can ensure that she is the only carrier who has
+done so---but cannot provide a correct execution with HATs
+alone. However, according to distributed transaction
+architects~\cite{entitygroup}, these compensatory actions are
+relatively common in real-world business processes.
 
@@
 and 3 (3.3.2.2-3) concern order ID assignment and are more
-problematic, as we have discussed.
+problematic, as we have discussed.  Finally, while TPC-C is not
+subject to multi-key anomalies, we note that many TPC-E isolation
+tests, such as one that requires simultaneously modifying a product
+description and its image, are also passable using HATs.
 
@@
 online services, HAT guarantees will suffice for many application
-transactions. Finally, while TPC-C is not subject to multi-key
-anomalies, we note that many TPC-E isolation tests, such as one that
-requires simultaneously modifying a product description and its image,
-are also passable using HATs.
+transactions.
 

### 2013-03-31 writeup/futurework.tex
PDB pass plus Alan comments

@@
 
-We believe that there are several promising avenues for future work in
-Highly Available Transaction systems:
+There are several promising avenues for future work in HAT systems:
 
-\textbf{What are the limits of HAT systems, and how are they defined?}
-In this work, we have taxonomized several well-defined and
-well-documented consistency properties. This is useful as it
+\textbf{What are the limits of achievable HAT models, and how are they
+  defined?}  In this work, we have taxonomized several well-defined
+and well-documented semantic properties. This is useful as it
 contextualizes decades of existing database and distributed systems
@@
 support~\cite{gray-virtues} could help answer the challenge of
-programming in a weakly consistent environment. Additionally, HAT
-guarantees can be strengthened by considering failure models with,
-say, bounded asynchrony or service-level agreements from underlying
-hardware. Additionally, while we have focused on systems that provide
-a single availability mode, there are several interesting hybrids to
-consider. For example, a system may provide ``best-effort'' stickiness
-(or serializability!) in the absence of partitions, but, in the
-presence of partitions, ``lose'' guarantees by having clients
-reconnect to an available replica. This strategy provides stronger
-guarantees in the absence of partitions but, to provide availability,
-sacrifices them in the event of partitions.
+programming in a weakly consistent environment. HAT guarantees can
+also be strengthened by considering weaker failure models such as
+bounded asynchrony or models in which the underlying hardware provides
+explicit service-level agreements. Additionally, while we have focused
+on systems that provide a single availability mode, there are several
+interesting hybrids to consider. For example, a system may provide
+``best-effort'' stickiness (or serializability) in the absence of
+partitions, but, in the presence of partitions, ``lose'' guarantees by
+having clients reconnect to an available replica. This strategy
+provides stronger guarantees in the absence of partitions but, to
+provide availability, sacrifices them in the event of partitions.

### 2013-03-31 writeup/hat-analysis-af.tex
PDB pass plus Alan comments

--- 2013-03-31 writeup/hat-analysis.tex
@@
-\section{Highly Available Transactions}
-\label{sec:hats}
-
-HAT systems, or systems providing transactions with transactional
-availability with high availability ($1$-available) or sticky high
-availability ($1$-sticky-available), offer substantial latency and
-availability benefits, yet they come with a cost to achievable
-semantics. In this section, we delineate which of several ACID,
-distributed consistency, and session consistency levels can be
-achieved with high availability (transactional atomicity, variants of
-Repeatable Read isolation, and many session guarantees) and sticky
-high availability (read your writes, PRAM and causal consistency) and
-which cannot (preventing Lost Update and Write Skew, recency).  We
-present a full summary of these results in
-Section~\ref{sec:hat-summary}.
-
-When possible, we draw on existing properties and definitions from the
-database and distributed systems literature, providing a brief,
-informal explanation and example for each guarantee. In particular,
-for our ACID guarantees, we draw largely on Atul Adya's
-dissertation~\cite{adya}, Berenson et al.'s 1995 critique of the ANSI
-SQL Specification~\cite{ansicritique}, and the ANSI SQL
-specification~\cite{ansi-sql}. We provide a set of formal definitions
-and semantics in our extended Technical Report~\cite{hat-tr}.
-
-\subsection{Achievable HAT Semantics}
-
-To begin, we present achievable semantics, offering proof-of-concept
-algorithms to demonstrate feasibility. These guarantees are fairly
-straightforward and have been introduced---albeit with greater brevity
-and discussion omitted---in a preliminary workshop
-paper~\cite{hat-hotos}. In our examples, we exclusively consider read
-and write operations. We denote a write of value $v$ to data item $d$
-as $w_d(v)$ and a read from data item $d$ returning $v$ as $r_d(v)$
-and assume that all data items have the null value, $\bot$, at
-database initialization. Unless otherwise specified, all example
-transactions commit.
-
-\subsubsection{ACID Isolation Guarantees}
-
-To begin, Adya's \textit{Read Uncommitted} isolation requires a total
-order among transactions and requires that each transaction's writes
-are ordered consistently with this order (prohibiting ``Dirty
-Writes,'' or $G0$)~\cite{adya}. If two transactions write to the same
-set of data items, then the final database state cannot contain the
-``earlier'' transaction's writes to the data items. For example, in
-the below example, $T_3$ should eventually only read $a=b=1$ or
-$a=b=2$ but not $a=2, b=1$ or $a=1, b=2$:
-\begin{align*}
-\small
-T_1 &: w_x(1)~w_y(1)
-\\T_2 &: w_x(2)~w_y(2)
-\\T_3 &: r_x(a)~r_y(b)\\[-2em]
-\end{align*}
-We will strengthen this property in later models, which will prevent,
-for instance, reading $a=2, b=1$ at any time (not just
-``eventually''). The difference between this total order on
-transactions and the total order required by serializability is that
-the Read Uncommitted total order is completely arbitrary (i.e., an
-order must exist), whereas a serializable total order must be
-equivalent to a serial execution. Read Uncommitted is easily achieved
-via applying the same logical timestamp to each update in a
-transaction and applying a ``last writer wins'' conflict
-reconciliation policy at each replica.
-
-\textit{Read Committed} isolation requires that transactions do not
-read uncommitted versions of data items (prohibiting both ``Dirty
-Writes''---as above---and ``Dirty Reads'' phenomena; captured by
-Adya's $G1a-c$, ANSI's $P1$, and ``broad'' $P1$ (2.2) from Berenson et
-al.). For instance, in the example below, $T_3$ should never see
-$a=1$, and, if $T_2$ aborts, $T_3$ should never read $a=3$:
-\vspace{-.5em}
-\begin{align*}
-\small
-T_1 &: w_x(1)~w_x(2)
-\\T_2 &: w_x(3)\\
-T_3 &: r_x(a)\\[-2em]
-\end{align*}
-It is fairly easy to prvent ``Dirty Reads'': if each transaction never
-writes uncommitted data to the database, then transactions will never
-read each others' dirty data. As a simple solution, clients can buffer
-their writes until they commit, or, alternatively, can send them to
-servers, who will not serve new writes until clients notify them that
-the writes have been committed.
-
-\textit{Repeatable Read} isolation is a contentious property. As
-Berenson et al. discuss~\cite{ansicritique}, Gray's original
-Repeatable Read lock-based implementation provides substantially
-richer guarantees than those that are required by the ANSI SQL
-specification~\cite{gray-isolation}. Gray~\cite{gray-isolation},
-Berenson et al.~\cite{ansicritique}, and Adya~\cite{adya} all
-interpret Repeatable Read as providing serializability for all
-operations except predicate-based reads. However, in its description
-of Repeatable Read, the ANSI SQL specification provides a useful
-property which, although is not true to Gray's spirit of Repeatable
-Read, is also commonly found in distributed consistency models: the
-ANSI Repeatable Read guarantee requires that, along with observing
-Read Committed isolation, each transaction will only read one version
-of each data item that it did not itself produce (preventing ``Fuzzy
-Read,'' or P2). In the example below, $T_3$ must read $a=1$:
-\begin{align*}
-\small
-T_1 &: w_x(1)
-\\T_2 &: w_x(2)
-\\T_3 &: r_x(1)~r_x(a)
-\end{align*}
-This isolation property is a literal interpretation of the phrase
-``Repeatable Read'': unless a transaction modifies a given data item,
-the observable value of the data item should not change during the
-transaction. By itself, this property is rather weak---we can always
-read $\bot$ for each data item. However, when coupled with additional
-properties, like transactional atomicity, it is stronger: the repeated
-reads must obey additional ordering constraints
-(Section~\ref{sec:discussion}). In fact, distributed systems often
-describe a ``consistent snapshot'' across a set of related events as a
-\textit{cut} across a set of participants or, in our case, data
-items. Accordingly, to capture the notion that the transaction should
-read from a (non-changing) consistent cut over data items (and to
-disambiguate from the aforementioned stronger Repeatable Read
-properties), we call ANSI SQL ``Repeatable Read'' (i.e., preventing
-P2) \textbf{Cut Isolation}. It is possible to satisfy cut isolation
-with high availability by caching appropriate versions for each of a
-transaction's reads or, alternatively, using multi-versioning on each
-replica and ensuring that each of a transaction's successive reads
-return the same version of each data item.
-
-We can further consider two variants of Cut Isolation. The first is
-relatively straightforward and provides a cut across individual data
-items. We will call this \textit{Item Cut Isolation}. The second
-provides a cut over individual data items and over logical ranges of
-data items, or predicate-based reads (e.g., \texttt{SELECT
-  WHERE}). This prevents the \textit{Phantom Problem}, whereby two
-successive predicate-based reads return different
-data~\cite{gray-isolation}. We call predicate-based cut isolation,
-which prevents Phantoms, \textit{Predicate Cut
-  Isolation}.\footnote{Oracle provides an isolation level called
-  ``Statement Level Read Consistency''; this is analogous to Predicate
-  Cut Isolation at the level of a single operation within a
-  multi-operation transaction. We do not consider this isolation model
-  here as it is subsumed by standard Snapshot Isolation and we believe
-  our discussion is easily extended to incorporate it.}
-
-\subsubsection{ACID Atomicity Guarantees}
-
-Transactional atomicity (TA) is core to ACID guarantees. Although, at
-least by the ACID acronym, it is not an ``isolation'' property,
-transactional atomicity restricts transactions' ability to view the
-effects of partially completed transactions. Under transactional
-atomicity, within each transaction, either all effects of another
-transaction are observed, or none are (equivalently, once some of the
-effects of a transaction are observed, all effects are observed). This
-is a strictly stronger guarantee than Read Committed isolation: in
-Read Committed, we can read a subset of another transaction's
-committed writes whereas, with TA, we must read all or none of
-them. Together with item cut isolation, TA prevents Read Skew
-anomalies (Berenson et al.'s A5A~\cite{ansicritique}). As an example
-of TA, $T_2$ must observe $b=c=1$ (or later versions for each key):
-\vspace{-.5em}
-\begin{align*}
-\small
-T_1 &: w_x(1)~w_y(1)~w_z(1)
-\\T_2 &: r_x(a)~r_y(1)~r_x(b)~r_z(c)~\\[-1.5em]
-\end{align*}
-$T_2$ can also observe $a=\bot$, $a=1$, or a later version of
-$a$. Notably, TA requires Read Committed isolation: observing all
-effects of a transaction implicitly requires observing the final
-(committed) effects of a transaction as well.
-
-Perhaps perplexingly, discussions of TA are absent from existing
-discussions of weak isolation. This is perhaps again due to the
-single-node context in which prior work was developed: on a single
-server (or a fully replicated database), TA is achievable via
-lightweight locking and/or local concurrency control over data
-items. In contrast, in a distributed environment, TA over arbitrary
-groups of non-colocated items is considerably more difficult to
-achieve with high availability: servers need to know when all of a
-transaction's updates are present on their respective
-replicas. However, replicas do not need to agree on when to reveal a
-new value to clients, which would require consensus; accordingly, TA
-only requires reliable broadcast (with some additional metadata for
-clients) and is achievable in a HAT system. We omit a full discussion
-of the algorithm and instead refer the reader to the extended version
-of this paper~\cite{hat-tr}.
-
-\subsubsection{Session Guarantees}
-
-Our models have not yet considered interactions \textit{across}
-transactions: we have not guaranteed any ordering or continuity
-between transactions (other than that there exists some arbitrary
-ordering). This is a natural concern in a distributed context, and, in
-the distributed systems literature, many useful \textit{safety}
-guarantees span multiple application actions. In particular,
-\textit{session guarantees} are used to describe guarantees across
-transactions within a given \textit{session}, ``an abstraction for the
-sequence of...operations performed during the execution of an
-application''~\cite{sessionguarantees}. Informally a session describes
-a context that should persist between transactions: for example, on a
-social networking site, all of a user's transactions submitted between
-``log in'' and ``log out'' operations might form a session.
-
-There are several session guarantees that we can make with high availability:
-
-\vspace{.5em}\noindent\textit{{Monotonic reads}} requires that,
-within a session, subsequent reads to a given object ``never return any
-previous values''; reads from each item progress forward in a total
-order. The ordering of reads should respect any total ordering on
-transactions.
-
-\vspace{.5em}\noindent\textit{{Monotonic writes}} requires that
-each session's writes be serialized. Any order on transactions should
-also respect the order in which each session submitted transactions.
-
-\vspace{.5em}\noindent\textit{{Writes Follow Reads}} requires that, if
-a session observes an effect of transaction $T_2$ and subsequently
-commits transaction $T_2$, then another session can only observe
-effects of $T_2$ if it can also observe $T_1$'s (or later values for
-$T_1$'s effects). Any order on transactions should respect the
-reads-from order.\vspace{.5em}
-
-We can achieve the above guarantees by forcing servers to wait to
-reveal new writes until each write's respective dependencies are
-fulfilled on all replicas. This mechanism effectively ensures that all
-clients read from a globally agreed upon lower bound on the versions
-written to the system. This \textit{is} highly available as a client
-will never block due to inability to find a server with a sufficiently
-up-to-date version of a data item. However, it does not imply that
-transactions will read their own writes or, in the presence of
-partitions, make forward progress through the version history. The
-problem is that, if a server becomes partitioned, under the highly
-available model, we must handle the possibility that an unfortunate
-client will be forced to issue her next requests against the
-partitioned server!
-
-The solution to this conundrum is to give up high availability in
-favor of sticky availability. Sticky availability permits three
-additional models, which we first define and then prove are
-unachievable in a generic highly available system:
-
-\vspace{.5em}\noindent\textit{{Read your writes}} requires
-that whenever a client reads a given data item after updating it, the
-read returns the updated value (or a value with a higher ID).
-
-\vspace{.5em}\noindent\textit{{PRAM}} (Pipelined Random Access
-Memory) provides the illusion of serializing each session's operations
-and is the combination of monotonic reads, monotonic writes, and read
-your writes~\cite{herlihy-art}.
-
-\vspace{.5em}\noindent\textit{{Causal
-    consistency}}~\cite{causalmemory} is the combination of all of the
-session guarantees~\cite{sessiontocausal} (alternatively, PRAM with
-writes-follow-reads) and is also referred to as PL-2L
-isolation~\cite{adya}).\vspace{.5em}
-
 Read your writes is not achievable in a highly available
@@
 network topology shifts and the client can only contact a different
-server which is partitioned from the server that executed $T_1$, then
+server that is partitioned from the server that executed $T_1$, then
 the client will be unable to read its own writes and the system will
@@
 this paper is not to sketch the entire space of consistency models
-that are achievable (see Section~\ref{sec:futurework}. We specifically
-evaluate TPC-C transaction semantics under HAT consistency guarantees
-in Section~\ref{sec:evaluation}.
+that are achievable (see Section~\ref{sec:futurework}). We
+specifically evaluate TPC-C transaction semantics under HAT
+consistency guarantees in Section~\ref{sec:evaluation}.
 
@@
 Skew---and therefore providing Snapshot Isolation, Repeatable Read,
-and one-copy serializability---requires unavailability.
+and one-copy serializability---inherently requires unavailability.
 
@@
 between $x$ and $y$ such that only one of $x$ or $y$ should have value
-$1$ at any given time, then this write skew would lead to a
-non-serializable execution. Write skew in particular is a somewhat
+$1$ at any given time, then this write skew would violate the constraint (which is preserved in serializable executions). Write skew is a somewhat
 esoteric anomaly---for example, it does not appear in
@@
 
-The inability to prevent Lost Update means that Consistent Read,
-Snapshot Isolation, and Cursor Stability guarantees are all
-unavailable. The inability to prevent Lost Update or Write Skew means
-that Repeatable Read and One-Copy Serializability guarantees are
-unavailable~\cite{adya,ansicritique}.
+Their need to prevent Lost Update means that Consistent Read, Snapshot
+Isolation, and Cursor Stability guarantees are all unavailable.
+Repeatable Read (in the sense of Gray~\cite{gray-isolation}, Berenson
+et al.~\cite{ansicritique}, or Adya~\cite{adya}) and One-Copy
+Serializability need to prevent both Lost Update and Write Skew; this
+means that they are also inherently unavailable.
 
@@
 date'' or similar. Unfortunately, an indefinitely long partition can
-force a system to violate any recency bound, so recency bounds are not
-enforceable in HAT systems.
+force an available system to violate any recency bound, so recency
+bounds are not enforceable in HAT systems.
 
@@
 available combinations) hints at the vast expanse of consistency
-models found in the literature. This taxonimization is \textit{not}
+models found in the literature. This taxonomy  is \textit{not}
 exhaustive, but we believe it lends substantial clarity into the

### 2013-03-31 writeup/model.tex
PDB pass plus Alan comments

@@
 means that clients need not need to wait for cross-datacenter
-communication. However, to properly describe whether a
-\textit{transactional} system is highly available, we need both a way
-to describe what servers a client must contact as well as what kinds
-of responses a server can provide, especially given the possibility of
-aborts.
+communication. To properly describe whether a \textit{transactional}
+system is highly available, we need to describe what servers a client
+must contact as well as what kinds of responses a server can provide,
+especially given the possibility of aborts.
 
@@
   subset of all data items. For generality, and, for applicability to
-  many modern ``sharded'' or ``partitioned'' data storage systems like
-  Dynamo, we consider partially replicated systems.} a server that can
-handle an operation for a given data item is called a \textit{replica}
-for that item. While this standard model is useful, there are several
-other cases we should consider.
+  many modern ``sharded'' or ``partitioned'' data storage
+  systems~\cite{dynamo, pnuts, bigtable, spanner}, we consider partially
+  replicated systems.} a server that can handle an operation for a
+given data item is called a \textit{replica} for that item. While this
+standard model is useful, there are several other cases we should
+consider.
 
@@
 presence of indefinitely long network partitions. Compared to a system
-providing high availability ($1$-availability), a system with
-$2$-availability will require more messaging and possibly higher
-latency and may experience more unsuccessful operations in the
-presence of network and server failures.  As is commmon in standard
-consensus algorithms like Paxos, clients may want to contact a
-majority of $N$ servers in a system. This \textit{majority
-  availability} (a special case of $K$-availability with $K=\lceil
-\frac{N}{2} \rceil$) is ``more available'' than full $N$-availability.
+providing \textit{high availability} ($1$-availability), a system with
+$2$-availability will be \textit{less available}, or require more
+messaging and possibly higher latency and may experience more
+unsuccessful operations in the presence of network and server
+failures.  As is commmon in standard consensus algorithms like Paxos,
+clients may want to contact a majority of $N$ servers in a
+system. This \textit{majority availability} (a special case of
+$K$-availability with $K=\lceil \frac{N}{2} \rceil$) is \textit{more
+  available} than full $N$-availability.
 
@@
 as a server itself; for example, a client might cache its reads and
-writes. However, a single client cannot achieve $K$-sticky
-availability with $K>1$ without contacting other replicas. Moreover, ``sticking'' all clients
-to a designated set of servers (i.e., master availability) is a
-special case of sticky availability and is ``less available'' than the
-equivalent non-mastered sticky availability.
+writes~\cite{bolton, sessionguarantees, swift}. However, a single client
+cannot achieve $K$-sticky availability with $K>1$ without contacting
+other replicas. Moreover, ``sticking'' all clients to a designated set
+of servers (i.e., master availability) is a special case of sticky
+availability and is ``less available'' than the equivalent
+non-mastered sticky availability.
 
@@
 datacenters) but believe there are several avenues for further
-taxnomization.
+taxonomization.
 
@@
 \textit{abort} responses---which signal transaction success or failure
-to a client---we must be careful in how we define transactional availability.
+to a client---we must carefully define transactional availability.
 

### 2013-03-31 writeup/motivation.tex
PDB pass plus Alan comments

@@
-\section{Partitions and Latency}
+\section{Why Availability?}
 \label{sec:motivation}
 
-Distributed databases face a host of challenges that their
-single-server counterparts are exempted from, largely stemming from
-network asynchrony and partial failure. These challenges mean that
-maintaining semantic guarantees has a cost, the exact price of which
-is a subject of considerable speculation in the database
-community~\cite{stonebraker2010errors}. In this section, we attempt to
-quantify the prevalence and cost of network partitions and latency in
-several real-world deployments. This data and a recent survey of
-existing database semantics motivates the study of highly available
-transaction systems.
+Why does availability matter? Peter Deutsch starts his classic list of
+``Fallacies of Distributed Computing'' with two concerns fundamental
+to distributed database systems: ``\textit{1.)}  The network is
+reliable. \textit{2.)} Latency is zero''~\cite{fallacies-deutsch}; in
+a distributed setting, network failures may prevent database servers
+from communicating, and, in the absence of failures, communication is
+slowed by factors like physical distance and, more often, congestion
+and routing. As we will see (Section~\ref{sec:availability}), highly
+available systems minimize the effect of these network
+\textit{partitions} and \textit{latency}. In this Section, we draw on
+a range of evidence that indicates that partitions occur relatively
+frequently in real-world deployments and best-case latencies between
+datacenters are significant.
 
-(TODO-PDB: keep this?) We will begin by studying the behavior of
-network partitions and latency. Peter Deutsch begins his classic list
-of ``Fallacies of Distributed Computing'' with two concerns
-fundamental to distributed database systems: ``\textit{1.)}  The
-network is reliable. \textit{2.)} Latency is
-zero''~\cite{fallacies-deutsch}. On a single server, communication
-channels are robust and relatively fast, and few single-server systems
-need to account for a possibility of partial system failure. However,
-in a distributed setting, both of these assumptions are no longer
-valid. To extend Jim Gray's storage latency analogy~\cite{gray-rules},
-if memory is 1.5 hours from the CPU and disk is 2 years away, then,
-especially across wide-area networks (WANs), remote server accesses
-can exceed 30 years, with the added concern that messages can be lost
-in transit or delayed indefinitely by network partitions. Accordingly,
-ignoring network failures and latency can ``cause big trouble and
-painful learning experiences''~\cite{fallacies-deutsch}.
-
-\subsection{Partitions in the Real World}
-
-Network partitions---informally, communication failures between
-participants in a distributed system---are challenging. A system
-facing network partitions faces the impossibility of simultaneously
-maintaining a correct ``global'' view of system state while
-simultaneously providing online operation to all of its
-clients~\cite{davidson-survey}. For networks that do not partition,
-maintaining semantic guarantees and providing ``always on'' operation
-is straightforward~\cite{stonebraker2010errors}. However, given the
-\textit{possibility} of partitions, a system must be prepared to
-either lose availability of operations or its ability to reason about
-global state. In this section, we present evidence for the frequency
-and non-triviality of WAN and LAN partitions.
+\subsection{Network Partitions at Scale}
 
@@
 
-Several recent studies quantify partition behavior more rigorously. A
-2011 study of several Microsoft datacenters found a mean of 40.8
-network link failures per day (95th percentile: 136), with a median
-time to repair of around five minutes (and up to one week). Perhaps
+Several recent studies rigorously quantify partition behavior. A 2011
+study of several Microsoft datacenters found a mean of 40.8 network
+link failures per day (95th percentile: 136), with a median time to
+repair of around five minutes (and up to one week). Perhaps
 surprisingly, provisioning redundant networks only reduces impact of
@@
 of communication latency, Deutsch's second ``Fallacy.'' In this
-section, we quantify expected latencies, which are often
-huge---hundreds of milliseconds---in a geo-replicated context.
-
+section, we quantify round-trip latencies, which are often
+large---hundreds of milliseconds---in a geo-replicated context.
 Fundamentally, the speed at which two servers can communicate is
@@
 geographically distinct sites, the cost of communication increases to
-this lower bound. In real deployments, transit times are actually
-slower than the speed of light due to routing, congestion, and
-processing times. To illustrate the difference between
-intra-datacenter, inter-datacenter, and inter-planetary networks, we
-performed a measurement study of network behavior on Amazon's EC2, a
-widely-used public compute cloud. We measured one week of ping times
-between all seven EC2 geographic ``regions'', across three
-``availability zones'' (closely co-located datacenters), and within a
-single ``availabilty zone'' (datacenter), at a granularity of 1s
-(dataset to be released).
-
-We summarize the results of our network measurement study in
-Table~\ref{table:rtt}. On average, intra-datacenter communication (1c)
-is between 1.82 and 6.38 times faster than across geographically
-co-located datacenters (1b) and between 40 and 647 times faster than
-across geographically distributed datacenters (1a). The cost of
-wide-area communication exceeds the speed of light: for example,
-communicating from S\~{a}o Paulo to Singapore is lower-bounded by
-speed-of-light transit RTT of 106.7ms but, on average, incurs a
-362.8ms RTT (95th percentile: 649ms). As shown in
-Figure~\ref{fig:rtt}, the distribution of latencies varies between
-links, but the trend---in line with other recent studies~\cite{mdcc,
-  redblue}---is clear: remote communication has a substantial cost.
+this lower bound.
 
@@
 \begin{table}
-\subfloat[Cross-region (CA:~California, OR:~Oregon, VA:~Virginia, TO:~Tokyo, IR:~Ireland, SY:~Sydney, SP:~S\~{a}o Paulo, SI:~Singapore)] {
-  \begin{tabular}{|c|c|c|c|c|c|c|c|c|}
-\hline
-& \multicolumn{1}{c}{OR} & \multicolumn{1}{c}{VA} & \multicolumn{1}{c}{TO} & \multicolumn{1}{c}{IR} & \multicolumn{1}{c}{SY} & \multicolumn{1}{c}{SP} & \multicolumn{1}{c|}{SI} \\\hline
-CA & \colorbox{min-lat-color}{22.5}   & 84.5   & 143.7   & 169.8   & 179.1   & 185.9   & 186.9  \\
-OR &  & 82.9   & 135.1   & 170.6   & 200.6   & 207.8   & 234.4  \\
-VA & &  & 202.4   & 107.9   & 265.6   & 163.4   & 253.5  \\
-TO & & &  & 278.3   & 144.2   & 301.4   & 90.6  \\
-IR & & & &  & 346.2   & 239.8   & 234.1  \\
-SY & & & & &  & 333.6   & 243.1  \\
-SP & & & & & &  & \colorbox{max-lat-color}{362.8}  \\
-\hline
-  \end{tabular}
-}\vspace{.5em}
+\vspace{.5em}
 
@@
  }}
+
+\subfloat[Cross-region (CA:~California, OR:~Oregon, VA:~Virginia, TO:~Tokyo, IR:~Ireland, SY:~Sydney, SP:~S\~{a}o Paulo, SI:~Singapore)] {
+  \begin{tabular}{|c|c|c|c|c|c|c|c|c|}
+\hline
+& \multicolumn{1}{c}{OR} & \multicolumn{1}{c}{VA} & \multicolumn{1}{c}{TO} & \multicolumn{1}{c}{IR} & \multicolumn{1}{c}{SY} & \multicolumn{1}{c}{SP} & \multicolumn{1}{c|}{SI} \\\hline
+CA & \colorbox{min-lat-color}{22.5}   & 84.5   & 143.7   & 169.8   & 179.1   & 185.9   & 186.9  \\
+OR &  & 82.9   & 135.1   & 170.6   & 200.6   & 207.8   & 234.4  \\
+VA & &  & 202.4   & 107.9   & 265.6   & 163.4   & 253.5  \\
+TO & & &  & 278.3   & 144.2   & 301.4   & 90.6  \\
+IR & & & &  & 346.2   & 239.8   & 234.1  \\
+SY & & & & &  & 333.6   & 243.1  \\
+SP & & & & & &  & \colorbox{max-lat-color}{362.8}  \\
+\hline
+  \end{tabular}
+}
+
 \caption{Mean RTT times on EC2 (min and max highlighted)}
@@
 
-\section{CAP and Modern Systems}
+In real deployments, transit times are actually slower than the speed
+of light due to routing, congestion, and processing times. To
+illustrate the difference between intra-datacenter, inter-datacenter,
+and inter-planetary networks, we performed a measurement study of
+network behavior on Amazon's EC2, a widely-used public compute
+cloud. We measured one week of ping times between all seven EC2
+geographic ``regions'', across three ``availability zones'' (closely
+co-located datacenters), and within a single ``availabilty zone''
+(datacenter), at a granularity of 1s (dataset to be released). We
+summarize the results of our network measurement study in
+Table~\ref{table:rtt}. On average, intra-datacenter communication (1a)
+is between 1.82 and 6.38 times faster than across geographically
+co-located datacenters (1b) and between 40 and 647 times faster than
+across geographically distributed datacenters (1c). The cost of
+wide-area communication exceeds the speed of light: for example,
+communicating from S\~{a}o Paulo to Singapore is lower-bounded by
+speed-of-light transit RTT of 106.7ms but, on average, incurs a
+362.8ms RTT (95th percentile: 649ms). As shown in
+Figure~\ref{fig:rtt}, the distribution of latencies varies between
+links, but the trend---in line with other recent studies~\cite{mdcc,
+  redblue}---is clear: remote communication has a substantial cost.
+
+\section{When is ``ACID'' ACID?}
 \label{sec:modernacid}
 
-Systems designers must make fundamental trade-offs between latency,
-system behavior during partitions, and achievable semantics. In this
-section, we briefly discuss known semantic limits for systems wishing
-to maintain high availability and their impact on modern database
-systems. We show that many database systems operate under weak model
-whose availability characteristics are unknown, motivating further
-study of these models in a highly available context in the remainder
-of this paper.
-
-One of the most popularly celebrated impossibility results in
-distributed systems is Brewer's CAP Theorem~\cite{brewer-slides},
-which, as formally proven by Gilbert and Lynch~\cite{gilbert-cap},
-states that it is impossible to maintain linearizability , or the
-ability to read the last completed write, (``C,'' or
-``consistency''---albeit poorly named) and replica availability, or a
-guaranteed response, (``A'') in the presence of network partitions
-(``P''). Brewer's Theorem has roots in decades of distributed database
-research~\cite{davidson-survey} and has had a substantial impact on
-recent large-scale distributed systems design~\cite{bigtable, pnuts,
-  dynamo}: while CAP is narrowly scoped to the impossibility of
-building highly available linearizable systems, it is often
-interpreted to preclude a larger set of guarantees, including ACID
-(transactional atomicity, consistency, isolation and durability
-guarantees). We are unlikely to determine the exact cause of this
-confusion, but it appears as early as the first public announcement of
-Brewer's result, in which ''distributed databases'' appear as an
-example of unavailable systems~\cite{brewer-slides}. While, in the
-same document, Brewer is careful to note that consistency and
-availability are a ``spectrum,'' all of ACID is still often labeled
-with a binary categorization of ``unavailable''~\cite{hn,
-  foundation-article}.
-
-Database researchers and designers have long realized that
-serializability---the gold standard of ACID isolation, which
-guarantees application-level consistency---is not achievable in a
-highly available system~\cite{davidson-survey}.  However, to increase
-concurrency, database systems offer a range of ACID properties weaker
-than serializability. Within a single-node database, the coordination
-penalties associated with serializability can be severe and manifest
-themselves in the form of decreased concurrency (and, subsequently,
-performance degradation, scalability limitations, and, often, external
-aborts like deadlocks)~\cite{gray-isolation}. Instead, databases offer
-a host of so-called \textit{weak isolation} models that allow varying
-restrictions on the space of schedules that are allowable by the
-system~\cite{adya}. None of these weak isolation models guarantees
-serializability, but, as we see below, the benefits of these models
-are often considered to outweigh costs of possible consistency
-anomalies that might arise from their use.
+The previous section demonstrated that distributed systems must
+address partitions and latency: what does this mean for distributed
+databases? Database researchers and designers have long realized that
+serializability is not achievable in a highly available
+system~\cite{davidson-survey}, meaning it will be expensive in the
+environments from Section~\ref{sec:motivation}. However, even a
+single-node database, the coordination penalties associated with
+serializability can be severe, manifesting themselves in the form of
+decreased concurrency (and, subsequently, performance degradation,
+scalability limitations, and, often, external aborts like
+deadlocks)~\cite{gray-isolation}. Accordingly, o increase concurrency,
+database systems offer a range of ACID properties weaker than
+serializability.  Instead, databases offer a host of so-called
+\textit{weak isolation} models that allow varying restrictions on the
+space of schedules that are allowable by the system~\cite{adya,
+  ansicritique, ansi-sql}. None of these weak isolation models
+guarantees serializability, but, as we see below, the benefits of
+these models are often considered to outweigh costs of possible
+consistency anomalies that might arise from their use.
 
@@
 recently surveyed the default and maximum isolation models provided by
-18 databases, often claiming to provide ``ACID'' or
-``NewSQL'' functionality~\cite{hat-hotos}. As shown in
+18 databases, often claiming to provide ``ACID'' or ``NewSQL''
+functionality~\cite{hat-hotos}. As shown in
 Table~\ref{table:existing}, only three out of 18 databases provided
-serializability by default, and at least eight did not provide
-serializability as an option at all. This is particularly surprising
-when we consider the widespread deployment of many of these
-non-serializable databases, like Oracle 11g, which are known to power
-major businesses and product functionality. Given that these
-transactional models are frequently used, our inability to provide
-serializability in arbitrary HATs appears non-fatal for practical
-applications. If application writers and database vendors have already
-decided that the benefits of weak isolation outweigh potential
-application inconsistencies, then, in a highly available environment
-that prohibits serializability, similar decisions may be tenable.
+serializability by default, and eight did not provide serializability
+as an option at all. This is particularly surprising when we consider
+the widespread deployment of many of these non-serializable databases,
+like Oracle 11g, which are known to power major businesses and product
+functionality. Given that these transactional models are frequently
+used, our inability to provide serializability in arbitrary HATs
+appears non-fatal for practical applications. If application writers
+and database vendors have already decided that the benefits of weak
+isolation outweigh potential application inconsistencies, then, in a
+highly available environment that prohibits serializability, similar
+decisions may be tenable.
 
@@
 Akiban Persistit & SI & SI\\
-Clustrix CLX 4100 & RR & ?\\
+Clustrix CLX 4100 & RR & RR\\
 Greenplum 4.1 & RC & S \\
 IBM DB2 10 for z/OS & CS & S\\
-IBM Informix 11.50 & Depends & RR\\
+IBM Informix 11.50 & Depends & S\\
 MySQL 5.6 & RR & S \\
@@
 relationship between serializability and high
-availability~\cite{davidson-survey} and weak isolation on a
-single server~\cite{adya, ansicritique, gray-isolation} but never weak isolation and
-high availability taken together.
+availability~\cite{davidson-survey} and weak isolation on a single
+server~\cite{adya, ansicritique, gray-isolation} but never weak
+isolation and high availability taken together.  A primary goal in the
+remainder of this paper is to understand which models are
+HAT-compliant.

### 2013-03-31 writeup/relatedwork.tex
PDB pass plus Alan comments

@@
 Isolation~\cite{generalizedsnapshot}, Granola~\cite{granola},
-HBase~\cite{hbase}, MDCC~\cite{mdcc}, Megastore~\cite{megastore},
-Orleans~\cite{orleans}, Postgres-R~\cite{kemme-thesis}, and
-Walter~\cite{walter}---are a reasonable choice. We note that many
-replication techniques target variants of Snapshot
-Isolation~\cite{middleware-db}.
+HBase, MDCC~\cite{mdcc}, Megastore~\cite{megastore},
+Orleans~\cite{orleans}, Postgres-R~\cite{kemme-thesis},
+Walter~\cite{walter}, and a range of snapshot isolation
+techniques~\cite{middleware-db, kemme-snapshot,
+  daudjee-snapshot}---are a reasonable choice.
 
-With HATs, we seek an alternative set of semantics that are
-still useful but do not violate requirements for high availability or
-low latency. Recent systems proposals such as Swift~\cite{swift},
+With HATs, we seek an alternative set of semantics that are still
+useful but do not violate requirements for high availability or low
+latency. Recent systems proposals such as Swift~\cite{swift},
 Eiger~\cite{eiger}, and Bolt-on Causal Consistency~\cite{bolton}
 provide transactional causal consistency guarantees with varying
-availability and represent a new class of (sticky) HAT systems.
+availability and represent a new class of (sticky) HAT
+systems.
+
+From Serializable to Causal Transactions for Collaborative Applications: Causal consistency with lost update
 

### 2013-03-31 writeup/futurework.tex
References purge

@@
 techniques such as escrow transactions~\cite{escrow},
-Sagas~\cite{sagas}, and other long-running transaction
-support~\cite{gray-virtues} could help answer the challenge of
-programming in a weakly consistent environment. HAT guarantees can
-also be strengthened by considering weaker failure models such as
-bounded asynchrony or models in which the underlying hardware provides
-explicit service-level agreements. Additionally, while we have focused
-on systems that provide a single availability mode, there are several
-interesting hybrids to consider. For example, a system may provide
-``best-effort'' stickiness (or serializability) in the absence of
-partitions, but, in the presence of partitions, ``lose'' guarantees by
-having clients reconnect to an available replica. This strategy
-provides stronger guarantees in the absence of partitions but, to
-provide availability, sacrifices them in the event of partitions.
+Sagas~\cite{sagas}, and other long-running transaction support could
+help answer the challenge of programming in a weakly consistent
+environment. HAT guarantees can also be strengthened by considering
+weaker failure models such as bounded asynchrony or models in which
+the underlying hardware provides explicit service-level
+agreements. Additionally, while we have focused on systems that
+provide a single availability mode, there are several interesting
+hybrids to consider. For example, a system may provide ``best-effort''
+stickiness (or serializability) in the absence of partitions, but, in
+the presence of partitions, ``lose'' guarantees by having clients
+reconnect to an available replica. This strategy provides stronger
+guarantees in the absence of partitions but, to provide availability,
+sacrifices them in the event of partitions.

### 2013-03-31 writeup/hat-analysis.tex
References purge

@@
 
-HAT systems, or systems providing transactions with transactional
-availability with high availability ($1$-available) or sticky high
-availability ($1$-sticky-available), offer substantial latency and
-availability benefits, yet they come with a cost to achievable
-semantics. In this section, we delineate which of several ACID,
-distributed consistency, and session consistency levels can be
-achieved with high availability (transactional atomicity, variants of
-Repeatable Read isolation, and many session guarantees) and sticky
-high availability (read your writes, PRAM and causal consistency) and
-which cannot (preventing Lost Update and Write Skew, recency).  We
-present a full summary of these results in
-Section~\ref{sec:hat-summary}.
+HAT systems provide transactions with transactional availability and
+either high availability ($1$-available) or sticky high availability
+($1$-sticky-available). They offer substantial latency and
+availability benefits compared to traditional distributed databases,
+yet they cannot achieve all the traditional semantics. In this
+section, we delineate ACID, distributed consistency, and session
+consistency levels which can be achieved with high availability
+(transactional atomicity, variants of Repeatable Read isolation, and
+many session guarantees), those with sticky high availability (read
+your writes, PRAM and causal consistency) and the properties which
+cannot be provided in a HAT system (those preventing Lost Update and
+Write Skew, or with recency).  We present a full summary of these
+results in Section~\ref{sec:hat-summary}.
 
@@
 overlapping (with distinct vocabulary)''~\cite{brewer-slides}. With
-this challenge in mind, we draw on existing properties and definitions
-from the database and distributed systems literature, providing a
-brief, informal explanation and example for each guarantee. In
-particular, for our ACID guarantees, we draw largely on Atul Adya's
-dissertation~\cite{adya}, Berenson et al.'s 1995 critique of the ANSI
-SQL Specification~\cite{ansicritique}, and the ANSI SQL
-specification~\cite{ansi-sql}. We provide a set of formal definitions
-and semantics in our extended Technical Report~\cite{hat-tr}.
+this challenge in mind, when possible, we build on existing properties
+and definitions from the database and distributed systems literature,
+providing a brief, informal explanation and example for each
+guarantee. The database isolation guarantees require particular care,
+since different DBMSs often use the same terminology for different
+mechanisms and may provide additional guarantees in addition to our
+implementation-agnostic definitions.  We draw largely on Atul Adya's
+dissertation~\cite{adya} and somewhat on its predecessor work: the
+ANSI SQL specification~\cite{ansi-sql} and Berenson et al.'s
+subsequent 1995 critique~\cite{ansicritique}. We provide a set of
+formal definitions, semantics, proofs and observations in our extended
+Technical Report and opt for a more informal presentation in this
+paper~\cite{hat-tr}.
 
@@
 
-To begin, we present achievable semantics, offering proof-of-concept
-algorithms to demonstrate feasibility. These guarantees are fairly
-straightforward and have been introduced---albeit with greater brevity
-and discussion omitted---in a preliminary workshop
-paper~\cite{hat-hotos}. In our examples, we exclusively consider read
-and write operations. We denote a write of value $v$ to data item $d$
-as $w_d(v)$ and a read from data item $d$ returning $v$ as $r_d(v)$
-and assume that all data items have the null value, $\bot$, at
-database initialization. Unless otherwise specified, all example
-transactions commit.
+To begin, we present well-known semantics that can be achieved in HAT
+systems. We offer proof-of-concept highly available algorithms but our
+goal is \textit{not} to provide optimal or even efficient
+implementations. These guarantees are fairly straightforward and have
+been introduced---albeit with greater brevity and discussion
+omitted---in a preliminary workshop paper~\cite{hat-hotos}. In our
+examples, we exclusively consider read and write operations. We denote
+a write of value $v$ to data item $d$ as $w_d(v)$ and a read from data
+item $d$ returning $v$ as $r_d(v)$ and assume that all data items have
+the null value, $\bot$, at database initialization. Unless otherwise
+specified, all example transactions commit.
 
@@
 
-To begin, Adya's \textit{Read Uncommitted} isolation requires a total
-order among transactions and requires that each transaction's writes
-are ordered consistently with this order (prohibiting ``Dirty
-Writes,'' or $G0$)~\cite{adya}. If two transactions write to the same
-set of data items, then the final database state cannot contain the
-``earlier'' transaction's writes to the data items. For example, in
-the below example, $T_3$ should eventually only read $a=b=1$ or
-$a=b=2$ but not $a=2, b=1$ or $a=1, b=2$:
+To begin, \textbf{Read Uncommitted} isolation is captured by Adya as
+PL-1,requiring that each transaction's writes are ordered consistently
+according to a single total order on transactions. This prohibits
+Adya's phenomenon $G0$, also called ``Dirty Writes''~\cite{adya}. If
+two transactions write to the same set of data items, then the final
+database state cannot contain the ``earlier'' transaction's writes to
+the data items. For example, in the below example, $T_3$ should
+\emph{eventually} only read $a=b=1$ or $a=b=2$ but not $a=2, b=1$ or
+$a=1, b=2$:
 \begin{align*}
@@
 \end{align*}
-We will strengthen this property in later models, which will prevent,
-for instance, reading $a=2, b=1$ at any time (not just
-``eventually''). The difference between this total order on
-transactions and the total order required by serializability is that
-the Read Uncommitted total order is completely arbitrary (i.e., an
-order must exist), whereas a serializable total order must be
-equivalent to a serial execution. Read Uncommitted is easily achieved
-via applying the same logical timestamp to each update in a
-transaction and applying a ``last writer wins'' conflict
-reconciliation policy at each replica.
+All later properties will strengthen PL-1.
 
-\textit{Read Committed} isolation requires that transactions do not
-read uncommitted versions of data items (prohibiting both ``Dirty
-Writes''---as above---and ``Dirty Reads'' phenomena; captured by
-Adya's $G1a-c$, ANSI's $P1$, and ``broad'' $P1$ (2.2) from Berenson et
-al.). For instance, in the example below, $T_3$ should never see
-$a=1$, and, if $T_2$ aborts, $T_3$ should never read $a=3$:
+Unlike the total order required by serializability, the order in PL-1
+does not constrain the values seen in by a transaction's read
+operations except once the database has reached a ``final
+state.'' Read Uncommitted is easily achieved via applying the same
+logical timestamp to each update in a transaction and applying a
+``last writer wins'' conflict reconciliation policy at each replica.
+
+\textbf{Read Committed} isolation is particularly important in
+practice as it is the default of many DBMSs. Implementations differ,
+with some based on long-duration exclusive locks and short-duration
+read locks~\cite{gray-isolation} and others based on multiple
+versions. The implementations typically have recency and monotonicity
+properties beyond the simple meaning of the name, which is what is
+expressed in the implementation-agnostic definition: under Read
+Committed, transactions should do not access uncommitted or
+intermediate versions of data items. This prohibits both ``Dirty
+Writes'', as above, and also ``Dirty Reads'' phenomena.  This
+isolation is Adya's PL-2, and formalised by prohibiting Adya's $G1a-c$
+(or ANSI's $P1$, or ``broad'' $P1$ (2.2) from Berenson et al.). For
+instance, in the example below, $T_3$ should never see $a=1$, and, if
+$T_2$ aborts, $T_3$ should never read $a=3$:
 \vspace{-.5em}
@@
 \end{align*}
-It is fairly easy to prvent ``Dirty Reads'': if each transaction never
-writes uncommitted data to the database, then transactions will never
-read each others' dirty data. As a simple solution, clients can buffer
-their writes until they commit, or, alternatively, can send them to
-servers, who will not serve new writes until clients notify them that
-the writes have been committed.
+It is fairly easy for a HAT system to prevent ``Dirty Reads'': if each
+transaction never writes uncommitted data to the database, then
+transactions will never read each others' dirty data. As a simple
+solution, clients can buffer their writes until they commit, or,
+alternatively, can send them to servers, who will not deliver their
+value to other readers until notified that the writes have been
+committed. This implementation does not provide recency or
+monotonicity guarantees but satisfies the implementation-agnostic
+definition.
 
-\textit{Repeatable Read} isolation is a contentious property. As
-Berenson et al. discuss~\cite{ansicritique}, Gray's original
-Repeatable Read lock-based implementation provides substantially
-richer guarantees than those that are required by the ANSI SQL
-specification~\cite{gray-isolation}. Gray~\cite{gray-isolation},
-Berenson et al.~\cite{ansicritique}, and Adya~\cite{adya} all
-interpret Repeatable Read as providing serializability for all
-operations except predicate-based reads. However, in its description
-of Repeatable Read, the ANSI SQL specification provides a useful
-property which, although is not true to Gray's spirit of Repeatable
-Read, is also commonly found in distributed consistency models: the
-ANSI Repeatable Read guarantee requires that, along with observing
-Read Committed isolation, each transaction will only read one version
-of each data item that it did not itself produce (preventing ``Fuzzy
-Read,'' or P2). In the example below, $T_3$ must read $a=1$:
+``Repeatable Read'' isolation is a confusing property. Some IBM
+products use the term for fully serializable
+isolation~\cite{hat-hotos}. Gray~\cite{gray-isolation}, Berenson et
+al.~\cite{ansicritique}, and Adya's PL-2.99~\cite{adya} all interpret
+Repeatable Read as providing serializability for all operations except
+predicate-based reads (and so Repeatable Read is identical to
+serializability in a key-value store without predicate-based
+access). In this model, the \textit{Phantom Problem}, whereby two
+successive predicate-based reads return different
+data~\cite{gray-isolation}, is the only kind of non-serializable
+behavior allowed. When referring to ``Repeatable Read,'' we will
+follow these definitions, which, as we will soon show, are for a
+property that be offered in a highly available way.
+
+However, As Berenson et al.~\cite{ansicritique} discuss, the ANSI SQL
+specification allows many additional behaviors under the term
+Repeatable Read isolation.  \textbf{ANSI Repeatable Read} requires
+that, along with respecting Read Committed isolation, each transaction
+will only read one version of each data item that it did not itself
+produce (preventing ``Fuzzy Read,'' or P2). In the example below,
+$T_3$ must read $a=1$:
 \begin{align*}
@@
 \end{align*}
-This isolation property is a literal interpretation of the phrase
-``Repeatable Read'': unless a transaction modifies a given data item,
-the observable value of the data item should not change during the
-transaction. By itself, this property is rather weak---we can always
-read $\bot$ for each data item. However, when coupled with additional
-properties, like transactional atomicity, it is stronger: the repeated
-reads must obey additional ordering constraints
-(Section~\ref{sec:discussion}). In fact, distributed systems often
-describe a ``consistent snapshot'' across a set of related events as a
-\textit{cut} across a set of participants or, in our case, data
-items. Accordingly, to capture the notion that the transaction should
-read from a (non-changing) consistent cut over data items (and to
-disambiguate from the aforementioned stronger Repeatable Read
-properties), we call ANSI SQL ``Repeatable Read'' (i.e., preventing
-P2) \textbf{Cut Isolation}. It is possible to satisfy cut isolation
-with high availability by caching appropriate versions for each of a
-transaction's reads or, alternatively, using multi-versioning on each
-replica and ensuring that each of a transaction's successive reads
-return the same version of each data item.
+This is a literal interpretation of the phrase ``Repeatable Read'':
+unless a transaction modifies a given data item, the observable value
+of the data item should not change during the transaction. By itself,
+this property is extremely weak---we can always read $\bot$ for each
+data item---and easily achieved in a HAT system by caching the values
+read.
 
-We can further consider two variants of Cut Isolation. The first is
-relatively straightforward and provides a cut across individual data
-items. We will call this \textit{Item Cut Isolation}. The second
-provides a cut over individual data items and over logical ranges of
-data items, or predicate-based reads (e.g., \texttt{SELECT
-  WHERE}). This prevents the \textit{Phantom Problem}, whereby two
-successive predicate-based reads return different
-data~\cite{gray-isolation}. We call predicate-based cut isolation,
-which prevents Phantoms, \textit{Predicate Cut
-  Isolation}.\footnote{Oracle provides an isolation level called
-  ``Statement Level Read Consistency''; this is analogous to Predicate
-  Cut Isolation at the level of a single operation within a
-  multi-operation transaction. We do not consider this isolation model
-  here as it is subsumed by standard Snapshot Isolation and we believe
-  our discussion is easily extended to incorporate it.}
+Between our Repeatable Read (PL-2.99) and the ANSI definition, is a
+concept we will call \textbf{Item Cut Isolation}: that all the
+transaction's reads should see values from a non-changing consistent
+cut or snapshot over the data items. It is possible to satisfy Item
+Cut Isolation with high availability by using multi-versioning on each
+replica and ensuring that each of a transaction's reads return an
+appropriately chosen version (e.g., identified by vector clock).
+
+A stronger achievable property asks that a consistent cut spans both a
+transaction's predicate reads (e.g., \texttt{SELECT WHERE}) as well as
+its item reads.  We call this \textit{Predicate Cut
+  Isolation}\footnote{Oracle provides an isolation level called
+  ``Statement Level Read Consistency''~\cite{adya}; this is analogous
+  to Predicate Cut Isolation at the level of a single operation within
+  a multi-operation transaction. We do not consider this isolation
+  model here as it is subsumed by standard Snapshot Isolation and we
+  believe our discussion is easily extended to incorporate it.} and it
+prohibits Phantoms. Predict Cut Isolation is also achievable in HAT
+system via similar multi-versioning as above.
 
@@
 least by the ACID acronym, it is not an ``isolation'' property,
-transactional atomicity restricts transactions' ability to view the
-effects of partially completed transactions. Under transactional
-atomicity, within each transaction, either all effects of another
-transaction are observed, or none are (equivalently, once some of the
-effects of a transaction are observed, all effects are observed). This
-is a strictly stronger guarantee than Read Committed isolation: in
-Read Committed, we can read a subset of another transaction's
-committed writes whereas, with TA, we must read all or none of
-them. Together with item cut isolation, TA prevents Read Skew
+transactional atomicity restricts the ability to view the effects of
+partially completed transactions. Under transactional atomicity, once
+some of the effects of a transaction $T_1$ are observed by another
+transaction $T_2$, thereafter all effects of $T_1$ are observed by
+$T_2$. Together with item cut isolation, TA prevents Read Skew
 anomalies (Berenson et al.'s A5A~\cite{ansicritique}). As an example
-of TA, $T_2$ must observe $b=c=1$ (or later versions for each key):
+of TA, because $T_2$ has read $T_1$'s write to $y$, $T_2$ must observe
+$b=c=1$ (or later versions for each key):
 \vspace{-.5em}
@@
 
-Perhaps perplexingly, discussions of TA are absent from existing
-discussions of weak isolation. This is perhaps again due to the
-single-node context in which prior work was developed: on a single
-server (or a fully replicated database), TA is achievable via
-lightweight locking and/or local concurrency control over data
-items. In contrast, in a distributed environment, TA over arbitrary
-groups of non-colocated items is considerably more difficult to
-achieve with high availability: servers need to know when all of a
-transaction's updates are present on their respective
-replicas. However, replicas do not need to agree on when to reveal a
-new value to clients, which would require consensus; accordingly, TA
-only requires reliable broadcast (with some additional metadata for
-clients) and is achievable in a HAT system. We omit a full discussion
-of the algorithm and instead refer the reader to the extended version
-of this paper~\cite{hat-tr}.
+Perplexingly, discussions of TA are absent from existing discussions
+of weak isolation. This is perhaps again due to the single-node
+context in which prior work was developed: on a single server (or a
+fully replicated database), TA is achievable via lightweight locking
+and/or local concurrency control over data items~\cite{gstore}. In
+contrast, in a distributed environment, TA over arbitrary groups of
+non-colocated items is considerably more difficult to achieve with
+high availability: servers need to know when all of a transaction's
+updates are present on their respective replicas. However, replicas do
+not need to agree on when to reveal a new value to clients, which
+would require consensus; accordingly, TA only requires reliable
+broadcast (with some additional metadata for clients) and is
+achievable in a HAT system. We omit a full discussion of the algorithm
+and instead refer the reader to the extended version of this
+paper~\cite{hat-tr}.  space{.5em}
 
@@
 transactions: we have not guaranteed any ordering or continuity
-between transactions (other than that there exists some arbitrary
-ordering). This is a natural concern in a distributed context, and, in
-the distributed systems literature, many useful \textit{safety}
-guarantees span multiple application actions. In particular,
-\textit{session guarantees} are used to describe guarantees across
-transactions within a given \textit{session}, ``an abstraction for the
-sequence of...operations performed during the execution of an
+between transactions, other than that there exists some arbitrary
+ordering (via Read Uncommitted). In the distributed systems
+literature, many useful \textit{safety} guarantees span multiple
+application actions. In particular, \textit{session guarantees} are
+used to describe guarantees across transactions within a given
+\textit{session}, ``an abstraction for the sequence of...operations
+performed during the execution of an
 application''~\cite{sessionguarantees}. Informally a session describes
@@
 
-There are several session guarantees that we can make with high availability:
+There are several session guarantees that we can make with high
+availability:
 
-\vspace{.5em}\noindent\textit{{Monotonic reads}} requires that,
-within a session, subsequent reads to a given object ``never return any
+\vspace{.5em}\noindent\textit{{Monotonic reads}} requires that, within
+a session, subsequent reads to a given object ``never return any
 previous values''; reads from each item progress forward in a total
-order. The ordering of reads should respect any total ordering on
-transactions.
+order. The ordering of reads should respect any externally observable
+total ordering on transactions.
 
-\vspace{.5em}\noindent\textit{{Monotonic writes}} requires that
-each session's writes be serialized. Any order on transactions should
-also respect the order in which each session submitted transactions.
+\vspace{.5em}\noindent\textit{{Monotonic writes}} requires that each
+session's writes be applied to any replica in the order they were
+submitted by the client. Any order on transactions should also be
+consistent with any precedence that a global observer would see.
 
 \vspace{.5em}\noindent\textit{{Writes Follow Reads}} requires that, if
-a session observes an effect of transaction $T_2$ and subsequently
+a session observes an effect of transaction $T_1$ and subsequently
 commits transaction $T_2$, then another session can only observe
-effects of $T_2$ if it can also observe $T_1$'s (or later values for
-$T_1$'s effects). Any order on transactions should respect the
-reads-from order.\vspace{.5em}
+effects of $T_2$ if it can also observe $T_1$'s effects (or later
+values that supercede $T_1$'s).  Any order on transactions should
+respect the reads-from order.\vspace{.5em}
 
@@
 client will be forced to issue her next requests against the
-partitioned server!
+partitioned server.
 
@@
 session guarantees~\cite{sessiontocausal} (alternatively, PRAM with
-writes-follow-reads) and is also referred to as PL-2L
+writes-follow-reads) and is also referred to by Adya as PL-2L
 isolation~\cite{adya}).\vspace{.5em}

### 2013-03-31 writeup/intro.tex
PDB pass plus Alan comments
References purge

--- 2013-03-31 writeup/intro.tex
@@
 database backends, from the use of transactional
-RDBMSs~\cite{bernstein-concurrency, bernstein-book, gray-isolation,
+RDBMSs~\cite{bernstein-book, gray-isolation,
   gray-virtues} to the widespread adoption of loosely-consistent
@@
 offer techniques for providing distributed
-serializability~\cite{bernstein-concurrency, spanner, granola,
-  daudjee-session, krikellas-bargain, calvin} or snapshot
-isolation~\cite{daudjee-snapshot,generalizedsnapshot, kemme-snapshot,
-  walter}. The distributed computing, programming language, and
-parallel hardware literature also contains many consistency models for
-single operations on replicated objects~\cite{pnuts, herlihy-art,
-  cops, eiger, cac, sessionguarantees}. However, the literature has
-largely ignored semantic guarantees for multiple operations operating
-on multiple data items in a highly available distributed environment.
+serializability~\cite{bernstein-book, hstore, kemme-classification,
+  spanner, daudjee-session, krikellas-bargain, calvin} or
+snapshot isolation~\cite{daudjee-snapshot,generalizedsnapshot,
+  kemme-snapshot, walter}. The distributed computing, programming
+language, and parallel hardware literature also contains many
+consistency models for single operations on replicated
+objects~\cite{pnuts, herlihy-art,  eiger, cac, sessionguarantees,
+  ceri-mechanism, chen-mechanism}. However, the literature has largely
+ignored semantic guarantees for multiple operations operating on
+multiple data items in a highly available distributed environment.
 
@@
 models, and, on the other hand, the goal of high availability.  We
-classify which, among the wide array models, are highly available,
+classify which, among the wide array of models, are highly available,
 denoting them as {\em Highly Available Transactions (HATs)}. In doing
@@
 is often implicit in the distributed systems
-literature~\cite{herlihy-art, cops, eiger, cac} but which requires
+literature~\cite{herlihy-art,  eiger, cac} but which requires
 \textit{explicit} consideration in a client-server replicated database
 context). This sticky availability is easily implementable and
-widely employed~\cite{cops, eiger, vogels-defs} but is strictly less
+widely employed~\cite{eiger, vogels-defs} but is strictly less
 available than traditional definitions of high availability.

### 2013-03-31 writeup/model.tex
References purge

@@
   many modern ``sharded'' or ``partitioned'' data storage
-  systems~\cite{dynamo, pnuts, bigtable, spanner}, we consider partially
+  systems~\cite{dynamo, pnuts, bigtable, spanner, hstore}, we consider partially
   replicated systems.} a server that can handle an operation for a

### 2013-03-31 writeup/motivation.tex
References purge

@@
 active research in networking
-community~\cite{surviving-failures-bodik, uw-failure-networks}.
+community~\cite{uw-failure-networks}.
 
-These results---which do not consider server-level failures (another
-form of system partition that has also been studied at
-scale~\cite{google-availability})---indicate that network partitions
-\textit{do} occur within and across modern datacenters. We do not
-intend to fixate on any particular statistical behavior but instead
-observe that, as a general trend, partitions exist and must
-accordingly be met with either unavailability at some servers or, as
-we will discuss, relaxed semantic guarantees.
+These results---which do not even consider server-level
+failures---indicate that network partitions \textit{do} occur within
+and across modern datacenters. We do not intend to fixate on any
+particular statistical behavior but instead observe that, as a general
+trend, partitions exist and must accordingly be met with either
+unavailability at some servers or, as we will discuss, relaxed
+semantic guarantees.
 
@@
 Figure~\ref{fig:rtt}, the distribution of latencies varies between
-links, but the trend---in line with other recent studies~\cite{mdcc,
-  redblue}---is clear: remote communication has a substantial cost.
+links, but the trend is clear: remote communication has a substantial
+cost.
 

### 2013-03-31 writeup/relatedwork.tex
References purge

@@
 multi-object semantics, both in academia~\cite{kraska-s3, granola,
-  gstore, mdcc, redblue, cops, eiger, walter,calvin, swift} and
-industry~\cite{megastore, orleans, spanner}. As discussed in
+  gstore, mdcc, eiger, walter,calvin, swift} and
+industry~\cite{orleans, spanner}. As discussed in
 Section~\ref{sec:modernacid}, classic ACID databases provide strong
@@
 implementations are unavailable in the presence of
-partitions~\cite{bernstein-concurrency, bernstein-book,
-  gray-isolation}. Notably, Google's Spanner provides strong one-copy
-serializable transactions. While Spanner is highly specialized for
-Google's read-heavy workload, it relies on two-phase commit and
-two-phase locking for read/write transactions~\cite{spanner}. As we
-have discussed, the latency penalties associated with this design
-choice are fundamental to serializability. For users willing to
-tolerate these costs, Spanner, or similar strongly consistent,
-unavailable systems---including Calvin~\cite{calvin},
-G-Store~\cite{gstore}, Gemini~\cite{redblue}, Generalized Snapshot
-Isolation~\cite{generalizedsnapshot}, Granola~\cite{granola},
-HBase, MDCC~\cite{mdcc}, Megastore~\cite{megastore},
-Orleans~\cite{orleans}, Postgres-R~\cite{kemme-thesis},
-Walter~\cite{walter}, and a range of snapshot isolation
-techniques~\cite{middleware-db, kemme-snapshot,
+partitions~\cite{bernstein-book, gray-isolation}. Notably, Google's
+Spanner provides strong one-copy serializable transactions. While
+Spanner is highly specialized for Google's read-heavy workload, it
+relies on two-phase commit and two-phase locking for read/write
+transactions~\cite{spanner}. As we have discussed, the latency
+penalties associated with this design choice are fundamental to
+serializability. For users willing to tolerate these costs, Spanner,
+or similar strongly consistent, unavailable systems---including
+Calvin~\cite{calvin}, G-Store~\cite{gstore}, Generalized Snapshot
+Isolation~\cite{generalizedsnapshot}, HBase, HStore~\cite{hstore},
+MDCC~\cite{mdcc}, Orleans~\cite{orleans},
+Postgres-R~\cite{kemme-thesis}, Walter~\cite{walter}, and a range of
+snapshot isolation techniques~\cite{middleware-db, kemme-snapshot,
   daudjee-snapshot}---are a reasonable choice.

### 2013-03-31 writeup/relatedwork.tex
checkpoin

@@
 
-From Serializable to Causal Transactions for Collaborative Applications: Causal consistency with lost update
-
-Middleware-based Database Replication: The Gaps Between Theory and Practice
-Most uses Snapshot Isolation
-
 % Ceri discuss distributed database update mechanisms with respect to
@@
 
-, with increasing interest in returning to transactional
-designs~\cite{spanner, walter, foundation-article, krikellas-bargain,
-  eiger}
+%, with increasing interest in returning to transactional
+%designs~\cite{spanner, walter, foundation-article, krikellas-bargain,
+%  eiger}

### 2013-03-31 writeup/conclusion.tex
Respin

--- 2013-03-29 writeup/conclusion.tex
@@
 \section{Conclusion}
+The current state of database software offers uncomfortable and unnecessary
+choices between availability and transactional semantics.
+In this paper we expose a broad design space of Highly Available Transactional 
+Systems (HATs), which can offer the key benefits of
+highly available distributed systems---partition tolerance and low-latency 
+writes---while also providing a family of transactional isolation 
+levels and replica semantics that have been widely used in practice. 
+In taxonomizing this design space, we show that many transactional 
+guarantees are achievable with high availability, while a few, 
+like preventing Lost Update and Write Skew, are not.
 
-In this paper, we have begun to understand the relationship between
-multi-operation, multi-object transactions and high availability. We
-have outlined requirements for high availability in a transactional
-context, or Highly Available Transactions (HATs) and taxonomized
-guarantees spanning both distributed systems and ACID database
-literature accordingly. Many transactional guarantees are achievable
-with high availability, while a few, like preventing Lost Update and
-Write Skew, are not. To understand the benefits of high availability
-and low latency, we classified the prevalence and cost of partitions
-and network latency both in the literature and via experiment, while
-analyzing existing transactions to determine whether HATs were
-sufficient to ensure application consistency.
+There has been some debate in the database community regarding the importance
+of availability in modern databases; there is also concern about the utility of
+relaxed isolation.  Our analysis and experiments in Sections~\ref{sec:motivation} 
+and~\ref{sec:evaluation} convince us that the availability goals of the NoSQL
+movement will remain a critical aspect of many important systems going forward, and 
+that the isolation levels available via HATs can have significant practical impact
+in delivering useful transactional semantics to application developers.
 
-Fundamentally, high availability represents a new constraint for
-database designs. While the CAP Theorem has had a large impact on
-highly available distributed data storage design, our knowledge of
-database systems has not kept pace. The key distributed systems
-challenge of designing distributed algorithms and architectures for
-partial failure and network asynchrony is largely unaddressed in the
-literature on transaction processing systems. To fully realizing the
-potential of HAT systems will require further rearchitecture and study
-of existing database designs as well as better understanding of
-application semantics and their relationship to distributed
-consistency.
+Much of the prior research in distributed transactions has focused on prototyping 
+individual points in the design space, with a focus on providing serializability
+at the expense of availability.  Our goal in this paper has been to take a broader view. 
+We taxonomize the joint design space of replica consistency and transactional isolation,
+and identify what is possible and impossible to achieve semantically via HATs.  From
+a research perspective, this offers a long-overdue connection between storage 
+semantics studied in the database and distributed systems communities.
+With this understanding in hand, the next challenge is to return to issues of 
+performance and systems architecture within the HATs design space.  Based on our 
+understanding of what is desired in the field and what is possible to achieve, we 
+believe this is an area of great opportunity in the coming years.

### 2013-03-31 writeup/evaluation.tex

@@
 \vspace{.5em}\noindent\textbf{Commutativity and Monotonicity} Recent
-work on the CALM Theorem~\cite{calm}, Lattice-based state
-update~\cite{blooml}, and Commutative and Replicated Data
+work on the CALM Theorem~\cite{calm} and Commutative and Replicated Data
 Types~\cite{crdt} demonstrates that, if updates logically commute,

### 2013-03-31 writeup/conclusion.tex
new conclusion

@@
 \section{Conclusion}
-The current state of database software offers uncomfortable and unnecessary
-choices between availability and transactional semantics.
-In this paper we expose a broad design space of Highly Available Transactional 
-Systems (HATs), which can offer the key benefits of
-highly available distributed systems---partition tolerance and low-latency 
-writes---while also providing a family of transactional isolation 
-levels and replica semantics that have been widely used in practice. 
-In taxonomizing this design space, we show that many transactional 
-guarantees are achievable with high availability, while a few, 
-like preventing Lost Update and Write Skew, are not.
+The current state of database software offers uncomfortable and
+unnecessary choices between availability and transactional semantics.
+We show, however, through analysis and experiments, that the
+availability goals of the NoSQL movement will remain a critical aspect
+of many important systems going forward. Thus, we expose a broad
+design space of Highly Available Transactional Systems (HATs), which
+can offer the key benefits of highly available distributed
+systems---partition tolerance and low-latency writes---while also
+providing a family of transactional isolation levels and replica
+semantics that have been widely used in practice.  In taxonomizing
+this design space, we show that many transactional guarantees are
+achievable with high availability, while a few, like preventing Lost
+Update and Write Skew, are not. From a research perspective, this
+offers a long-overdue connection between storage semantics studied in
+the database and distributed systems communities.
 
-There has been some debate in the database community regarding the importance
-of availability in modern databases; there is also concern about the utility of
-relaxed isolation.  Our analysis and experiments in Sections~\ref{sec:motivation} 
-and~\ref{sec:evaluation} convince us that the availability goals of the NoSQL
-movement will remain a critical aspect of many important systems going forward, and 
-that the isolation levels available via HATs can have significant practical impact
-in delivering useful transactional semantics to application developers.
+With the above understanding in hand, the next challenge is to return to
+issues of performance and systems architecture within the HATs design
+space.  Based on our understanding of what is desired in the field and
+what is possible to achieve, we believe this is an area of great
+opportunity in the coming years.
 
-Much of the prior research in distributed transactions has focused on prototyping 
-individual points in the design space, with a focus on providing serializability
-at the expense of availability.  Our goal in this paper has been to take a broader view. 
-We taxonomize the joint design space of replica consistency and transactional isolation,
-and identify what is possible and impossible to achieve semantically via HATs.  From
-a research perspective, this offers a long-overdue connection between storage 
-semantics studied in the database and distributed systems communities.
-With this understanding in hand, the next challenge is to return to issues of 
-performance and systems architecture within the HATs design space.  Based on our 
-understanding of what is desired in the field and what is possible to achieve, we 
-believe this is an area of great opportunity in the coming years.
+%% In this paper we expose a broad design space of Highly Available Transactional 
+%% Systems (HATs), which can offer the key benefits of
+%% highly available distributed systems---partition tolerance and low-latency 
+%% writes---while also providing a family of transactional isolation 
+%% levels and replica semantics that have been widely used in practice. 
+%% In taxonomizing this design space, we show that many transactional 
+%% guarantees are achievable with high availability, while a few, 
+%% like preventing Lost Update and Write Skew, are not.
+
+%% There has been some debate in the database community regarding the importance
+%% of availability in modern databases; there is also concern about the utility of
+%% relaxed isolation.  Our analysis and experiments in Sections~\ref{sec:motivation} 
+%% and~\ref{sec:evaluation} convince us that the availability goals of the NoSQL
+%% movement will remain a critical aspect of many important systems going forward, and 
+%% that the isolation levels available via HATs can have significant practical impact
+%% in delivering useful transactional semantics to application developers.
+
+%% Much of the prior research in distributed transactions has focused on prototyping 
+%% individual points in the design space, with a focus on providing serializability
+%% at the expense of availability.  Our goal in this paper has been to take a broader view. 
+%% We taxonomize the joint design space of replica consistency and transactional isolation,
+%% and identify what is possible and impossible to achieve semantically via HATs.  From
+%% a research perspective, this offers a long-overdue connection between storage 
+%% semantics studied in the database and distributed systems communities.
+%% With this understanding in hand, the next challenge is to return to issues of 
+%% performance and systems architecture within the HATs design space.  Based on our 
+%% understanding of what is desired in the field and what is possible to achieve, we 
+%% believe this is an area of great opportunity in the coming years.

### 2013-04-01 writeup/model.tex
killed the hierarchy stuff

@@
 
-Traditionally, a system provides high availability if every user that
-can contact a server eventually receives a response from that server,
+Traditionally, a system provides {\em  availability} if every user that
+can contact a correct server eventually receives a response from that server,
 even in the presence of arbitrary, indefinitely long network
@@
 
-First, clients may wish to contact multiple servers. For example, to
-ensure that the effects of operations will persist in the event of
-server failures, clients may wish to wait until operations reach more
-than one server before they consider the operations
-completed. However, according to our definition of high availability,
-server-fault tolerance is unavailable. To allow flexibility in the
-number of participating servers required for an operation, we propose
-\textit{$K$-availability}: if a client can contact $K$ servers, then
-it eventually receives a response from all of them, even in the
-presence of indefinitely long network partitions. Compared to a system
-providing \textit{high availability} ($1$-availability), a system with
-$2$-availability will be \textit{less available}, or require more
-messaging and possibly higher latency and may experience more
-unsuccessful operations in the presence of network and server
-failures.  As is commmon in standard consensus algorithms like Paxos,
-clients may want to contact a majority of $N$ servers in a
-system. This \textit{majority availability} (a special case of
-$K$-availability with $K=\lceil \frac{N}{2} \rceil$) is \textit{more
-  available} than full $N$-availability.
-
-Clients may also wish to contact the same server (or set of servers)
-across subsequent operations. As we will discuss in
+Some algorithms assume a model in which the client always contacts the
+same server across subsequent operations. As we will discuss in
 Section~\ref{sec:hats}, clients can ensure continuity between
@@
 servers~\cite{vogels-defs}. Simultaneously maintaining availability
-and stickiness entails fate-sharing between the client and its set of
-servers: if a client's sticky servers fail, or, equivalently, if the
-client is partitioned from its sticky servers, then it may face
+and stickiness entails fate-sharing between the client and its 
+server: if a client's sticky server fail, or, equivalently, if the
+client is partitioned from its sticky server, then it may face
 unavailability. We say that a system provides \textit{sticky
   availability} if, whenever a client's operation is executed against
-a server that has observed all of its prior operations, it eventually
+a correct server that has observed all of its prior operations, it eventually
 receives a response, even in the presence of indefinitely long
-partitions. As above, it may be useful for a client to achieve
-stickiness \textit{and} contact many replicas (i.e., for fault
-tolerance), so we generalize sticky availability to \textit{$K$-sticky
-  availability}, which provides sticky availability with $K$ distinct
-servers. A client may choose to become $1$-sticky available by acting
+partitions. A client may choose to become sticky available by acting
 as a server itself; for example, a client might cache its reads and
-writes~\cite{bolton, sessionguarantees, swift}. However, a single client
-cannot achieve $K$-sticky availability with $K>1$ without contacting
-other replicas. Moreover, ``sticking'' all clients to a designated set
-of servers (i.e., master availability) is a special case of sticky
-availability and is ``less available'' than the equivalent
-non-mastered sticky availability.
+writes~\cite{bolton, sessionguarantees, swift}.
 
-\begin{figure}
-\centering
-\begin{tikzpicture}[scale=0.8]
-  \tikzstyle{every node}=[font=\small]
- \node[draw=none,fill=none] (1) at (0,0) {$1$-availability (high availability)}; 
- \node[draw=none,fill=none] (2) at (0,1) {$2$-availability}; 
- \node[draw=none,fill=none] (m-1) at (0, 2) {\ldots}; 
- \node[draw=none,fill=none] (m) at (0, 3) {majority availability}; 
- \node[draw=none,fill=none] (m+1) at (0, 4) {\ldots}; 
- \node[draw=none,fill=none] (n) at (0, 5) {$N$-availability}; 
- \node[draw=none,fill=none] (1s) at (4, 1.5) {$1$-sticky availability};
- \node[draw=none,fill=none] (2s) at (4, 2.5) {$2$-sticky availability};
- \node[draw=none,fill=none] (n2s) at (4, 3.5) {\ldots};
- \node[draw=none,fill=none] (n1s) at (4, 4.5) {$(N$$-$$1)$-sticky availability};
+Availability implies sticky availability, since availability does not
+require that the contacted server has seen all the previous operations
+of a client, whereas sticky availability is the special case in which
+the system only guarantees correctness if the contacted server has
+seen all the prior operations from the client.
 
- \draw [->] (1) -- (2);
- \draw [->] (2) -- (m-1);
- \draw [->] (m-1) -- (m);
- \draw [->] (m) -- (m+1);
- \draw [->] (m+1) -- (n);
- \draw [->] (1) -- (1s);
- \draw [->] (1s) -- (m);
- \draw [->] (2s) -- (m+1);
- \draw [->] (1s) -- (2s);
- \draw [->] (2) -- (2s);
- \draw [->] (2s) -- (n2s);
- \draw [->] (n2s) -- (n1s);
- \draw [->] (n1s) -- (n);
+%% \begin{figure}
+%% \centering
+%% \begin{tikzpicture}[scale=0.8]
+%%   \tikzstyle{every node}=[font=\small]
+%%  \node[draw=none,fill=none] (1) at (0,0) {$1$-availability (high availability)}; 
+%%  \node[draw=none,fill=none] (2) at (0,1) {$2$-availability}; 
+%%  \node[draw=none,fill=none] (m-1) at (0, 2) {\ldots}; 
+%%  \node[draw=none,fill=none] (m) at (0, 3) {majority availability}; 
+%%  \node[draw=none,fill=none] (m+1) at (0, 4) {\ldots}; 
+%%  \node[draw=none,fill=none] (n) at (0, 5) {$N$-availability}; 
+%%  \node[draw=none,fill=none] (1s) at (4, 1.5) {$1$-sticky availability};
+%%  \node[draw=none,fill=none] (2s) at (4, 2.5) {$2$-sticky availability};
+%%  \node[draw=none,fill=none] (n2s) at (4, 3.5) {\ldots};
+%%  \node[draw=none,fill=none] (n1s) at (4, 4.5) {$(N$$-$$1)$-sticky availability};
 
-\end{tikzpicture}
-\caption{Hierarchy of replica availability levels for $N>3$ servers.}
-\label{fig:availability-order}
-\end{figure}
+%%  \draw [->] (1) -- (2);
+%%  \draw [->] (2) -- (m-1);
+%%  \draw [->] (m-1) -- (m);
+%%  \draw [->] (m) -- (m+1);
+%%  \draw [->] (m+1) -- (n);
+%%  \draw [->] (1) -- (1s);
+%%  \draw [->] (1s) -- (m);
+%%  \draw [->] (2s) -- (m+1);
+%%  \draw [->] (1s) -- (2s);
+%%  \draw [->] (2) -- (2s);
+%%  \draw [->] (2s) -- (n2s);
+%%  \draw [->] (n2s) -- (n1s);
+%%  \draw [->] (n1s) -- (n);
 
-We show a hierarchy of replica availability levels in
-Figure~\ref{fig:availability-order}. $K$-sticky availability subsumes
-$K$-availability, but $K$-sticky availability is incomparable with
-$(K+1)$-availability. $N$-sticky availability is equivalent to
-$N$-availability, while majority availability subsumes $1$-sticky
-availability. More generally, $\lceil \frac{N}{2} \rceil$$+$$K$$-$$1$
-availability subsumes $K$-sticky availability. We have omitted
-discussion of operation-specific availability levels (e.g.,
-$N$-availability for writes and $1$-availability for reads in a
-write-all, read-one data store), system membership changes, or
-heterogeneous replicas (e.g., servers in a local and remote
-datacenters) but believe there are several avenues for further
-taxonomization.
+%% \end{tikzpicture}
+%% \caption{Hierarchy of replica availability levels for $N>3$ servers.}
+%% \label{fig:availability-order}
+%% \end{figure}
+
+%% We show a hierarchy of replica availability levels in
+%% Figure~\ref{fig:availability-order}. $K$-sticky availability subsumes
+%% $K$-availability, but $K$-sticky availability is incomparable with
+%% $(K+1)$-availability. $N$-sticky availability is equivalent to
+%% $N$-availability, while majority availability subsumes $1$-sticky
+%% availability. More generally, $\lceil \frac{N}{2} \rceil$$+$$K$$-$$1$
+%% availability subsumes $K$-sticky availability. We have omitted
+%% discussion of operation-specific availability levels (e.g.,
+%% $N$-availability for writes and $1$-availability for reads in a
+%% write-all, read-one data store), system membership changes, or
+%% heterogeneous replicas (e.g., servers in a local and remote
+%% datacenters) but believe there are several avenues for further
+%% taxonomization.
 
@@
 aborts~\cite{hat-hotos}. For example, a system will violate
-transactional $2$-availability if a client that can contact two
+transactional availability if a client that can contact two
 servers for each of the data items in its transaction does not

### 2013-04-01 thebes-code/scripts/graph/plot.py
better plot

@@ -1,6 +1,9 @@
 import matplotlib
 matplotlib.use('Agg')
 
+matplotlib.rcParams['figure.figsize'] = 5, 1.6
+matplotlib.rcParams['lines.linewidth'] = 2
+
 from pylab import *
 from os import listdir
 from sys import argv
@@ -87,6 +90,7 @@ def processYCSB(d):
                     print d, line
 
     if ops is None or time is None or len(lats) == 0 or avglat is None:
+        print d
         print ops, time, len(lats), avglat
         return None
 
@@ -111,9 +115,16 @@ results = {}
 txnlens = []
 
 for config in listdir(rootdir):
+    '''
+    p = processConfig(rootdir+"/"+config)
+    print config, p.get_thru(), p.get_average_latency(), p.get_pctile_latency(.999)
+    continue
+    '''
+
     configsplit=config.split("-THREADS")
-    nthreads = int(configsplit[1])
+    nthreads = int(configsplit[1].split("-IT")[0])
     txnlength = configsplit[0].split('-')[1]
+    it = int(config.split("-IT")[1])
 
     if int(txnlength) not in txnlens:
         txnlens.append(int(txnlength))
@@ -124,49 +135,127 @@ for config in listdir(rootdir):
     r = processConfig(rootdir+"/"+config)
     if r is None:
         print "SKIPPING", config
+        '''
+        if nthreads in results[configstr]:
+            del results[configstr][nthreads]
+        '''
         continue
     r.threads = nthreads
     r.txnlength = int(txnlength)
 
     if configstr not in results:
-        results[configstr] = []
+        results[configstr] = {}
+    if nthreads not in results[configstr]:
+        results[configstr][nthreads] = []
 
-    results[configstr].append(r)
+    results[configstr][nthreads].append(r)
 
 txnlens.sort()
 
 PCT = 0
 AVG = 1
 
-for latplot in [PCT, AVG]:
+fmtdict = {
+"CONSTANT_TRANSACTIONREAD_COMMITTED-NO_ATOMICITY" : ["blue", '^-'],
+"QUORUM_EVENTUAL" : ["black", 'v-'],
+"MASTERED_EVENTUAL" : ["teal", 's-'],
+"CONSTANT_TRANSACTIONREAD_COMMITTED-CLIENT" : ["green", 'o-'],
+"EVENTUAL" : ["red", 'x-']}
+
+threadresultthru = {}
+threadresultlat = {}
+
+for latplot in [AVG]:
[…]

### 2013-04-01 writeup/hat-analysis.tex

@@
 HAT systems provide transactions with transactional availability and
-either high availability ($1$-available) or sticky high availability
-($1$-sticky-available). They offer substantial latency and
-availability benefits compared to traditional distributed databases,
-yet they cannot achieve all the traditional semantics. In this
-section, we delineate ACID, distributed consistency, and session
-consistency levels which can be achieved with high availability
-(transactional atomicity, variants of Repeatable Read isolation, and
-many session guarantees), those with sticky high availability (read
-your writes, PRAM and causal consistency) and the properties which
-cannot be provided in a HAT system (those preventing Lost Update and
-Write Skew, or with recency).  We present a full summary of these
-results in Section~\ref{sec:hat-summary}.
+either high availability or sticky high availability. They offer
+substantial latency and availability benefits compared to traditional
+distributed databases, yet they cannot achieve all the traditional
+semantics. In this section, we delineate ACID, distributed
+consistency, and session consistency levels which can be achieved with
+high availability (transactional atomicity, variants of Repeatable
+Read isolation, and many session guarantees), those with sticky high
+availability (read your writes, PRAM and causal consistency) and the
+properties which cannot be provided in a HAT system (those preventing
+Lost Update and Write Skew, or with recency).  We present a full
+summary of these results in Section~\ref{sec:hat-summary}.
 
@@
 Read your writes is not achievable in a highly available
-($1$-available) system. Consider a client that executes the following
-two transactions in succession:
+system. Consider a client that executes the following two transactions
+in succession:
 \vspace{-.5em}
@@
 
-\vspace{.5em}\noindent{\textbf{Durability}} As we briefly discussed in
-Section~\ref{sec:availability}, a client requiring that its
+\vspace{.5em}\noindent{\textbf{Durability}} A client requiring that its
 transactions' effects survive $F$ server faults requires at least
-($F+1$)-availability.
+$F+1$ correct nodes.
 

### 2013-04-01 writeup/evaluation.tex
eval

@@
 
-\begin{figure}
-\includegraphics[width=.8\columnwidth]{figs/macrotracelegend.pdf}\vspace{-2mm}
-\includegraphics[width=\columnwidth]{figs/lat-thru-trace-K0-SCALEMACRO.pdf}
-\includegraphics[width=\columnwidth]{figs/lat-thru-trace-K0-SCALEMACRO.pdf}
-\caption{Latency-throughput for database operation on EC2 (DUMMY FIG)}
+To better understand the impact of HAT versus non-HAT algorithms and
+validate the latency-based analysis in Section~\ref{sec:latency}, we
+deployed a basic distributed database prototype on public cloud
+infrastructure. We compared the behavior of eventual consistency,
+per-key mastering (wherein all operations required a RTT to a
+designated master server), Read Committed, and TA with read your
+writes guarantees. We deployed an experimental Java-based distributed
+datastore backed by LevelDB both within and across EC2 regions and ran
+the YCSB benchmark to determine the resulting performance impact. We
+partitioned the database into two fully replicated clusters of three
+\texttt{m1.xlarge} instances. For HAT, we stuck YCSB clients to their
+local cluster and, to provide transactional functionality, grouped
+every six YCSB operations from the default workload to form a
+transaction. We report the median of five trials per experiment.
+
+\begin{figure}[t!]
+\begin{center}
+\hspace{2em}\includegraphics[width=.8\columnwidth]{figs/strategylegend.pdf}
+\end{center}\vspace{-2em}
+\begin{center}\textbf{Within \texttt{us-east}}\end{center}\vspace{-1.5em}
+\includegraphics[width=\columnwidth]{figs/lan-threads-lats.pdf}\vspace{-1em}
+\includegraphics[width=\columnwidth]{figs/lan-threads-thru.pdf}
+\begin{center}\textbf{Between \texttt{us-east} and \texttt{us-west-2}}\end{center}\vspace{-1.5em}
+\includegraphics[width=\columnwidth]{figs/wan-threads-lats.pdf}\vspace{-1em}
+\includegraphics[width=\columnwidth]{figs/wan-threads-thru.pdf}
+\caption{YCSB performance for two clusters of three servers each
+  deployed within a single datacenter and cross-datacenters.}
 \label{fig:wan-exp}
@@
 
-2.) We ran YCSB on EC2 (and maybe TPC-C if we have time) with a few models:
-	- Eventual Consistency
-	- Eventual consistency with a "master" for updates--simulates the lower bound on a non-HAT system.
-	- Transactional Atomicity and Read Committed
+As Figure~\ref{fig:wan-exp} shows, within a single datacenter, a
+mastered implementation achieves approximately double the latency and
+slightly over half of the throughput of a HAT implementation. This is
+because, given two replicas for each key, the mastered implementation
+serves requests from only one. The mastered implementation still
+achieves $19,797$ transactions per second---over 6,000 per
+master---and is only $15.56$ms slower at peak compared to eventual
+consistency. However, eventual consistency and HAT models achieve
+substantially higher throughput--$37,575$ and $36,178$ transactions
+per second, respectively--and lower latency---$48.3\%$ and $47.2\%$
+respectively. This difference is unsurprising because, in these weaker
+models, all servers can serve both reads and writes, effectively
+doubling the capacity of the system. The weaker models still propagate
+new values between replicas, but this process is asynchronous and,
+aside from consuming disk I/O and network bandwidth, does not
+substantially impact clients. As expected, HAT models perform
+similarly to eventual consistency: our Read Committed implementation
+buffers client writes and adds no additional server-side overhead,
+while TA requires additional metadata with each write but has less
+than 5\% performance impact compared to eventual consistency.
 
-	-This is going to look a lot like the ping tests: much higher latency for "non-HATs."
-	-Our Naive 2PL implementation bottlenecks extremely quickly.
-	-On, say, 5 servers, can get hundreds of thousands of TPS
+In comparison, over multiple datacenters, the cost of coordination
+increases. The HAT and eventual consistency implementations continue
+to enjoy local RTTs only, and, with the exception of a slight
+performance degradation in TA---which we believe is attributable to
+Java garbage collection overheads in TA's inbound anti-entropy
+buffers---performance remains unchanged. However, each YCSB update to
+a remote replica ($\frac{2}{3}$ in one datacenter, $\frac{1}{3}$ in
+the other) of requiring a remote RTT between Virginia and Oregon (or
+vice-versa) to reach the master. This leads to a latency increase of
+around $166$ms ($746\%$) per transaction compared to a
+single-datacenter deployment. In order to attain reasonable throughput
+for the mastered deployment, we had to greatly increase the number of
+concurrent clients accessing the database, and we found that threading
+and socket overheads quickly overwhelmed the database server. Over a
+wide-area network, the increased cost of RTTs dramatically affects
+system efficiency.
 
-	-Not meant as an exhaustive study, but validates our intuitions. Coordination costs of non-HAT systems are unaccounted for, will only go up, and we haven't even talked about availability.
+This brief set of experiments is not meant as an exhaustive study, but
+it validates our earlier intuition regarding the cost of violating
+HAT-compliance. As Deutsch points out, ignoring factors such as
+latency can ``cause big trouble and painful learning
+experiences''~\cite{fallacies-deutsch}---in a single-site context,
+paying the cost of coordination may be tenable, but, especially as
+services are geo-replicated, the costs increase.

### 2013-04-01 writeup/conclusion.tex
editing pass

@@
 availability goals of the NoSQL movement will remain a critical aspect
-of many important systems going forward. Thus, we expose a broad
+of many future data storage systems. Accordingly, we expose a broad
 design space of Highly Available Transactional Systems (HATs), which
 can offer the key benefits of highly available distributed
-systems---partition tolerance and low-latency writes---while also
-providing a family of transactional isolation levels and replica
-semantics that have been widely used in practice.  In taxonomizing
-this design space, we show that many transactional guarantees are
-achievable with high availability, while a few, like preventing Lost
-Update and Write Skew, are not. From a research perspective, this
-offers a long-overdue connection between storage semantics studied in
-the database and distributed systems communities.
+systems---''always on'' operation during partitions and low-latency
+operations---while also providing a family of transactional isolation
+levels and replica semantics that have been adopted in practice.  In
+taxonomizing this design space, we show that many transactional
+guarantees are achievable with high availability, while a few, like
+preventing Lost Update and Write Skew, are not. From a research
+perspective, this offers a long-overdue connection between storage
+semantics studied in the database and distributed systems communities.
 
-With the above understanding in hand, the next challenge is to return to
-issues of performance and systems architecture within the HATs design
-space.  Based on our understanding of what is desired in the field and
-what is possible to achieve, we believe this is an area of great
-opportunity in the coming years.
+With the above understanding in hand, the next challenge is to return
+to issues of performance and systems architecture within the HAT
+design space.  Based on our understanding of what is desired in the
+field and newfound knowledge of what is possible to achieve, we
+believe this is an area of great opportunity in the coming years.
 

### 2013-04-01 writeup/hat-analysis.tex
editing pass

@@
 follow these definitions, which, as we will soon show, are for a
-property that be offered in a highly available way.
+property that can be offered in a highly available way.
 
@@
   believe our discussion is easily extended to incorporate it.} and it
-prohibits Phantoms. Predict Cut Isolation is also achievable in HAT
-system via similar multi-versioning as above.
+prohibits Phantoms. Predicate Cut Isolation is also achievable in HAT
+systems via similar multi-versioning as above.
 
@@
 broadcast (with some additional metadata for clients) and is
-achievable in a HAT system. We omit a full discussion of the algorithm
+achievable in HAT systems. We omit a full discussion of the algorithm
 and instead refer the reader to the extended version of this
@@
 
-In this section, we discuss several subleties in our results,
+In this section, we discuss several subtleties in our results,
 specifically addressing model composition, transactional atomicity
@@
 distributed systems literature. Accordingly, despite apparent
-similarities, transactional atomicity incomparable with and much
+similarities, transactional atomicity is incomparable with and much
 cheaper (by availability standards) than linearizability.

### 2013-04-01 writeup/intro.tex
editing pass

@@
 database backends, from the use of transactional
-RDBMSs~\cite{bernstein-book, gray-isolation,
-  gray-virtues} to the widespread adoption of loosely-consistent
-distributed key-value stores~\cite{bigtable, pnuts,
-  dynamo}. Instrumental in this trend was the 2000 introduction of
-Brewer's CAP Theorem, which stated that a highly available system
-cannot provide strong consistency guarantees in the presence of
-network partitions~\cite{brewer-slides}. However, as formally
-proven~\cite{gilbert-cap}, the CAP Theorem concerns
+RDBMSs~\cite{bernstein-book, gray-isolation, gray-virtues} to the
+widespread adoption of loosely-consistent distributed key-value
+stores~\cite{bigtable, pnuts, dynamo}. Instrumental in this trend was
+the 2000 introduction of Brewer's CAP Theorem, which stated that a
+highly available system cannot provide strong consistency guarantees
+in the presence of network partitions~\cite{brewer-slides}. However,
+as formally proven~\cite{gilbert-cap}, the CAP Theorem concerns
 linearizability~\cite{herlihy-art}, the ability to read the most
@@
 Spanner~\cite{spanner}. Unfortunately, these systems are not highly
-available in the presence of partitions~\cite{walter, mdcc,
-  foundation-article, krikellas-bargain, eiger}. (TODO:ADD more)
+available in the presence of partitions~\cite{ foundation-article,
+  mdcc, krikellas-bargain, eiger, walter}. (TODO:ADD more)
 
@@
 long tradition of providing weaker isolation and consistency
-guarantees~\cite{adya, ansicritique, gray-isolation,
-  gray-virtues} (TODO: kemme). Due to their concurrency and overall performance
+guarantees~\cite{adya, ansicritique, gray-virtues, gray-isolation,
+  kemme-thesis}. Due to their concurrency and overall performance
 benefits, weak isolation models are widely employed, overwhelmingly as
@@
 offer techniques for providing distributed
-serializability~\cite{bernstein-book, hstore, kemme-classification,
-  spanner, daudjee-session, krikellas-bargain, calvin} or
-snapshot isolation~\cite{daudjee-snapshot,generalizedsnapshot,
-  kemme-snapshot, walter}. The distributed computing, programming
-language, and parallel hardware literature also contains many
-consistency models for single operations on replicated
-objects~\cite{pnuts, herlihy-art,  eiger, cac, sessionguarantees,
-  ceri-mechanism, chen-mechanism}. However, the literature has largely
-ignored semantic guarantees for multiple operations operating on
-multiple data items in a highly available distributed environment.
+serializability~\cite{bernstein-book, spanner, daudjee-session,
+  hstore, krikellas-bargain, calvin, kemme-classification} or snapshot
+isolation~\cite{daudjee-snapshot,generalizedsnapshot, kemme-snapshot,
+  walter}. The distributed computing, programming language, and
+parallel hardware literature also contains many consistency models for
+single operations on replicated objects~\cite{ ceri-mechanism,
+  chen-mechanism, pnuts, herlihy-art, eiger, cac, sessionguarantees}. However, the literature has largely ignored semantic guarantees for
+multiple operations operating on multiple data items in a highly
+available distributed environment.
 
-Our main contributions are as follows. We fill gap between, on one
+Our main contributions are as follows. We fill the gap between, on one
 hand, the plethora of previously proposed isolation and consistency

### 2013-04-01 writeup/model.tex
editing pass

@@
 will formulate a model that captures a range of availability models,
-including ``true'' high availability, availablility with stickiness,
+including ``true'' high availability, availability with stickiness,
 and several other useful variants.
@@
 provides low latency~\cite{abadi-pacelc}; in a wide-area setting, this
-means that clients need not need to wait for cross-datacenter
+means that clients need not wait for cross-datacenter
 communication. To properly describe whether a \textit{transactional}

### 2013-04-01 writeup/motivation.tex
eval
editing pass

@@
 large-scale operators like Google~\cite{dean-keynote} confirm that
-partition management is an important consideration service operators
+partition management is an important consideration for service operators
 today. As an example of an actual system rearchitecture resulting from
@@
 \subsection{Latency and Planet Earth}
+\label{sec:latency}
 
@@
 system~\cite{davidson-survey}, meaning it will be expensive in the
-environments from Section~\ref{sec:motivation}. However, even a
+environments from Section~\ref{sec:motivation}. However, even in a
 single-node database, the coordination penalties associated with
@@
 scalability limitations, and, often, external aborts like
-deadlocks)~\cite{gray-isolation}. Accordingly, o increase concurrency,
+deadlocks)~\cite{gray-isolation}. Accordingly, to increase concurrency,
 database systems offer a range of ACID properties weaker than
@@
 are not resilient to partial failure
-(Section~\ref{sec:evaluation}). Morever, we are not aware of any prior
+(Section~\ref{sec:evaluation}). Moreover, we are not aware of any prior
 literature that provides guidance as to the relationship between weak

### 2013-04-01 writeup/intro.tex
cleaning

@@
 
-The prior decade saw a large shift in the design of large-scale
+The prior decade has seen a large shift in the design of large-scale
 database backends, from the use of transactional
@@
 widespread adoption of loosely-consistent distributed key-value
-stores~\cite{bigtable, pnuts, dynamo}. Instrumental in this trend was
+stores~\cite{bigtable, pnuts, dynamo}. Fundamental to this trend was
 the 2000 introduction of Brewer's CAP Theorem, which stated that a
 highly available system cannot provide strong consistency guarantees
-in the presence of network partitions~\cite{brewer-slides}. However,
-as formally proven~\cite{gilbert-cap}, the CAP Theorem concerns
-linearizability~\cite{herlihy-art}, the ability to read the most
-recent write to a data item that is replicated across servers. Yet
-this result---regarding a fairly narrow form of distributed
+in the presence of network partitions~\cite{brewer-slides}. As
+formally proven~\cite{gilbert-cap}, the CAP Theorem concerns
+linearizability: the ability to read the most recent write to a data
+item that is replicated across servers~\cite{herlihy-art}. Yet this
+result---regarding a fairly narrow form of distributed
 consistency---is often conflated with ACID database
@@
 Spanner~\cite{spanner}. Unfortunately, these systems are not highly
-available in the presence of partitions~\cite{ foundation-article,
-  mdcc, krikellas-bargain, eiger, walter}. (TODO:ADD more)
+available in the presence of partitions~\cite{middleware-db,
+  foundation-article, hstore,generalizedsnapshot, mdcc,
+  krikellas-bargain, eiger, walter, calvin}.
 
 Indeed, serializable transactions, the gold standard of traditional
-ACID databases, is indeed not achievable given a requirement for high
+ACID databases, are indeed not achievable given a requirement for high
 availability in the presence of
@@
 surprisingly, that besides serializability, only Snapshot Isolation
-(SI) and Repeatable Reads (RR) are unavailable, with most other
-isolation levels being available. We also show that many of the weaker
-data consistency models from distributed systems are HAT-compliant.
+(SI) and Repeatable Reads (RR) are inherently unavailable, with most
+other isolation levels being available. We also show that many of the
+weaker data consistency models from distributed systems are
+HAT-compliant.
 
@@
 weaker data consistency models from database and distributed systems
-can be implemented in a highly available manner. For several of these
+can be implemented in a highly available manner. For several
 guarantees, such as Transactional Causal Predicate Cut Isolation, we
@@
 \textit{explicit} consideration in a client-server replicated database
-context). This sticky availability is easily implementable and
-widely employed~\cite{eiger, vogels-defs} but is strictly less
-available than traditional definitions of high availability.
+context). This sticky availability is easily implementable and widely
+employed~\cite{eiger, vogels-defs} but is strictly less available than
+traditional definitions of high availability.
 
@@
 sometimes-unavailable mechanisms that are more
-coordination-intensive. Overall, this paper quantifies the benefits
+coordination intensive. Overall, this paper quantifies the benefits
 Highly Available Transactions can offer as well as the necessary

### 2013-04-01 writeup/motivation.tex
cleaning

@@
 North America and Europe due a router
-bug~\cite{juniper-partition}---and misconfigurations---such as BGP
+bug~\cite{juniper-partition}---and misconfigurations such as BGP
 faults in 2008~\cite{pakistan-youtube} and
-2010~\cite{research-experiment-partition}---can cause widespread
+2010~\cite{research-experiment-partition} can cause widespread
 partitions.
@@
 Fundamentally, the speed at which two servers can communicate is
-(according to modern physics) bounded by the speed of light. Two
-servers located on opposite sides of the Earth face, in the best
-case---communicating via a hypothetical link crossing through the
-center of the planet---require a minimum 85.1 ms RTT (133.7 ms if sent
-at ground level). As services are replicated to multiple,
-geographically distinct sites, the cost of communication increases to
-this lower bound.
+(according to modern physics) bounded by the speed of light. In the
+best case, two servers on opposite sides of the Earth's
+surface---communicating via a hypothetical link traversing the
+planet's diamater---require a minimum 85.1 ms RTT (133.7 ms if sent at
+ground level). As services are replicated to multiple, geographically
+distinct sites, the cost of communication increases to this lower
+bound.
 

### 2013-04-01 writeup/relatedwork.tex
cleaning

@@
 There has been a recent resurgence of interest in distributed
-multi-object semantics, both in academia~\cite{kraska-s3, granola,
-  gstore, mdcc, eiger, walter,calvin, swift} and
-industry~\cite{orleans, spanner}. As discussed in
-Section~\ref{sec:modernacid}, classic ACID databases provide strong
-semantics but their lock-based and traditional multi-versioned
-implementations are unavailable in the presence of
-partitions~\cite{bernstein-book, gray-isolation}. Notably, Google's
-Spanner provides strong one-copy serializable transactions. While
-Spanner is highly specialized for Google's read-heavy workload, it
-relies on two-phase commit and two-phase locking for read/write
-transactions~\cite{spanner}. As we have discussed, the latency
-penalties associated with this design choice are fundamental to
-serializability. For users willing to tolerate these costs, Spanner,
-or similar strongly consistent, unavailable systems---including
-Calvin~\cite{calvin}, G-Store~\cite{gstore}, Generalized Snapshot
-Isolation~\cite{generalizedsnapshot}, HBase, HStore~\cite{hstore},
-MDCC~\cite{mdcc}, Orleans~\cite{orleans},
+multi-object semantics, both in academia~\cite{kraska-s3, gstore,
+  mdcc, eiger, walter,calvin, swift} and industry~\cite{orleans,
+  spanner}. As discussed in Section~\ref{sec:modernacid}, classic ACID
+databases provide strong semantics but their lock-based and
+traditional multi-versioned implementations are unavailable in the
+presence of partitions~\cite{bernstein-book, gray-isolation}. Notably,
+Google's Spanner provides strong one-copy serializable
+transactions. While Spanner is highly specialized for Google's
+read-heavy workload, it relies on two-phase commit and two-phase
+locking for read/write transactions~\cite{spanner}. As we have
+discussed, the latency penalties associated with this design choice
+are fundamental to serializability. For users willing to tolerate
+these costs, Spanner, or similar strongly consistent, unavailable
+systems---including Calvin~\cite{calvin}, G-Store~\cite{gstore},
+Generalized Snapshot Isolation~\cite{generalizedsnapshot}, HBase,
+HStore~\cite{hstore}, MDCC~\cite{mdcc}, Orleans~\cite{orleans},
 Postgres-R~\cite{kemme-thesis}, Walter~\cite{walter}, and a range of

### 2013-04-01 writeup/hat-analysis.tex
cleaning
tweaks

@@
 all sticky-HAT guarantees, we have transactional, causal snapshot
-reads. If we combine TA and P-CI, we have transactional snapshot
-reads. We can achieve RC, MR, and RYW by simply sticking clients to
-servers. And we can also combine unavailable models---for example, an
-unavailable system might provide PRAM and One-Copy
+reads (i.e., Causal Transactional Predicate Cut Isolation). If we
+combine TA and P-CI, we have transactional snapshot reads. We can
+achieve RC, MR, and RYW by simply sticking clients to servers. And we
+can also combine unavailable models---for example, an unavailable
+system might provide PRAM and One-Copy
 Serializability~\cite{daudjee-session}.
@@
 availability (e.g., treating a datacenter as a linearizable cluster),
-this overhead can be reduced~\cite{cops, eiger}, but it is much
+this overhead can be reduced~\cite{eiger}, but it is much
 cheaper to provide, say, read your writes, than full causal

### 2013-04-01 writeup/hat-analysis.tex
fixed

@@
 
-Perplexingly, discussions of TA are absent from existing discussions
-of weak isolation. This is perhaps again due to the single-node
+Perplexingly, discussions of TA are absent from existing ones of
+weak isolation. This is perhaps again due to the single-node
 context in which prior work was developed: on a single server (or a
@@
 would require consensus; accordingly, TA only requires reliable
-broadcast (with some additional metadata for clients) and is
+broadcast (with some additional client metadata) and is
 achievable in HAT systems. We omit a full discussion of the algorithm
 and instead refer the reader to the extended version of this
-paper~\cite{hat-tr}.  space{.5em}
+paper~\cite{hat-tr}.  
+%space{.5em}
 
@@
 
-We can achieve the above guarantees by forcing servers to wait to
+The above guarantees can be achieved by forcing servers to wait to
 reveal new writes until each write's respective dependencies are
@@
 clients read from a globally agreed upon lower bound on the versions
-written to the system. This \textit{is} highly available as a client
+written. This \textit{is} highly available as a client
 will never block due to inability to find a server with a sufficiently
@@
 also defined in terms of previously identified isolation and
-consistency anomalies. Perhaps most notably, it is impossible to
+consistency anomalies. Most notably, it is impossible to
 prevent Lost Update or Write Skew in a HAT system.

### 2013-04-01 writeup/model.tex
fixed

@@
 
-So far, we have focused on single-object, single-operation
-availability. This is standard in distributed systems literature
-(e.g., linearizable, regular, and safe register models all concern
+We have so far focused on single-object, single-operation
+availability. This is standard in the distributed systems literature
+(e.g., linearizable, and regular register models all concern
 single objects~\cite{herlihy-art}), yet the database literature
@@
 requirement. Additionally, given the possibility of aborts, we need to
-ensure that the system makes useful forward progress: a system can
-trivially guarantee clients a response by always aborting all
+ensure useful forward progress: a system can trivially guarantee
+clients a response by always aborting all
 transactions~\cite{transaction-liveness}. However, this is an
 unsatisfactory system because nothing good (transaction commit) ever
-happens; we should require a \textit{liveness} property. We
-cannot guarantee that every transaction will commit---transactions may
-choose to abort themselves---but we need to make sure that the system
-will not indefinitely abort transactions on its own volition. We call
-a transaction abort due to a transaction's own choosing (e.g., as an
+happens; we should require a \textit{liveness} property. We cannot
+guarantee that every transaction will commit---transactions may choose
+to abort themselves---but we need to make sure that the system will
+not indefinitely abort transactions on its own volition. We call a
+transaction abort due to a transaction's own choosing (e.g., as an
 operation of the transaction itself or due to a would-be declared

### 2013-04-01 writeup/evaluation.tex
foo

@@
 datastore backed by LevelDB both within and across EC2 regions and ran
-the YCSB benchmark to determine the resulting performance impact. We
+the YCSB benchmark~\cite{ycsb} to determine the impact of latency. We
 partitioned the database into two fully replicated clusters of three

### 2013-04-01 writeup/relatedwork.tex
foo

@@
 serializability~\cite{wisemann-survey}, linearizability, consistent
-snapshots, and recency bounds~\cite{ceri-mechanism, chen-mechanism}.
+snapshots, and recency bounds~\cite{ceri-mechanism}.
 

### 2013-04-01 writeup/relatedwork.tex
relwork

--- 2013-04-01 writeup/relatedwork.tex
@@
 this work expands, providing impossibility results, a full
-taxonomization, and experimental and application analysis (basically
-all but Section 4.1)---we believe this paper is the first to explore
-this area. Nonetheless, several studies rigorously classify non-HAT
-semantics: Wisemann et al. have proposed a three parameter
-classification for one-copy serializable database systems, performing
-a more extensive classification than our overview in
-Section~\ref{sec:evaluation}, including eager database replication
-techniques~\cite{kemme-classification}. Additionally several earlier
-surveys discuss distributed mechanisms for achieving one-copy
-serializability~\cite{wisemann-survey}, linearizability, consistent
-snapshots, and recency bounds~\cite{ceri-mechanism, chen-mechanism}.
+taxonomization, and experimental and application analysis---we believe
+this paper is the first to explore this area. Nonetheless, several
+studies rigorously classify non-HAT semantics: Wisemann et al. have
+proposed a three parameter classification for one-copy serializable
+database systems, performing a more extensive classification than our
+overview in Section~\ref{sec:evaluation}, including eager database
+replication techniques~\cite{kemme-classification}. Additionally
+several earlier surveys discuss distributed mechanisms for achieving
+one-copy serializability~\cite{wisemann-survey}, linearizability,
+consistent snapshots, and recency bounds~\cite{ceri-mechanism,
+  chen-mechanism}.
 

### 2013-04-01 writeup/evaluation.tex
fixed

@@
 isolation in Gray's original proposal~\cite{gray-isolation} do not
-degrade gracefully in the presence of partial failure. (Note, however,
+degrade gracefully in the presence of partial failures. (Note, however,
 that lock-based protocols \textit{do} offer the benefit of recency
@@
 without substantial modification; Swift~\cite{swift} is closest to
-providing full HAT operation in a sticky context. As we have seen, it
+providing full HAT operation in a sticky context. As  seen, it
 is possible to implement many guarantees weaker than

### 2013-04-01 writeup/evaluation.tex
Foo

@@
 
-With an understanding of which guarantees are and are not
-HAT-compliant, in this section, we analyze the implications of these
-results for existing systems and applications and briefly study HAT
-systems on public cloud infrastructure. Specifically:
+With an understanding of which guarantees are HAT-compliant, in this
+section, we analyze the implications of these results for existing
+systems and briefly study HAT systems on public cloud
+infrastructure. Specifically:
 
@@
 \item We examine the properties required by a realistic OLTP
-  application patterned on the TPC-C benchmark.
+  application based on the TPC-C benchmark.
 \item We perform a brief experimental evaluation of HAT versus non-HAT
@@
 are not designed for high availability. These algorithms often presume
-a single-server deployment or the requirement for serializability. As
+a single-server deployment or a requirement for serializability. As
 a consequence, traditional transaction processing systems are not
@@
 (RTT) before committing. As an example, traditional two-phase locking
-for a transaction of length $T$ requires $T$ \texttt{lock} operations
-and at least one \texttt{unlock} operation.  In a distributed
-environment, each of these operations requires coordination, either
-with other database replicas or with a lock service. If this
-coordination mechanism is unavailable, transactions cannot safely
-commit. Similarly, optimistic concurrency control requires
-coordinating via a validation step, while deterministic transaction
-scheduling~\cite{deterministic-scheduling} requires a
-scheduler. Serializability under multi-version concurrency control
-also requires checking for update conflicts. The reliance on a
+for a transaction of length $T$ may require $T$ \texttt{lock}
+operations and will require at least one \texttt{lock} and
+\texttt{unlock} operation.  In a distributed environment, each of
+these operations requires coordination, either with other database
+replicas or with a lock service. If this coordination mechanism is
+unavailable, transactions cannot safely commit. Similarly, optimistic
+concurrency control requires coordinating via a validation step, while
+deterministic transaction scheduling~\cite{deterministic-scheduling}
+requires a scheduler. Serializability under multi-version concurrency
+control also requires checking for update conflicts. The reliance on a
 globally agreed total order necessitates a minimum of one round-trip
@@
 
-To better understand the impact of HAT versus non-HAT algorithms and
-validate the latency-based analysis in Section~\ref{sec:latency}, we
-deployed a basic distributed database prototype on public cloud
-infrastructure. We compared the behavior of eventual consistency,
-per-key mastering (wherein all operations required a RTT to a
-designated master server), Read Committed, and TA with read your
-writes guarantees. We deployed an experimental Java-based distributed
-datastore backed by LevelDB both within and across EC2 regions and ran
-the YCSB benchmark~\cite{ycsb} to determine the impact of latency. We
-partitioned the database into two fully replicated clusters of three
-\texttt{m1.xlarge} instances. For HAT, we stuck YCSB clients to their
-local cluster and, to provide transactional functionality, grouped
-every six YCSB operations from the default workload to form a
-transaction. We report the median of five trials per experiment.
+To better understand the performance implications of HAT guarantees in
+a real-world environment and to validate the latency-based analysis in
+Section~\ref{sec:latency}, we deployed a distributed database
+prototype on public cloud infrastructure. We compared the behavior of
+eventual consistency, per-item mastering (wherein all operations to a
+given item required contacting a designated master server), Read
+Committed, and TA with read your writes guarantees. We deployed an
+experimental Java-based distributed datastore backed by LevelDB both
+within and across EC2 regions and ran the YCSB benchmark~\cite{ycsb}
+to determine the impact of latency. We partitioned the database into
+two fully replicated clusters of three \texttt{m1.xlarge} instances
+each. For HAT, we stuck YCSB clients to their local cluster and, to
+provide transactional functionality, grouped every six YCSB operations
+from the default workload to form a transaction. We report the median
+of five trials per experiment.
 

### 2013-04-01 writeup/intro.tex
Foo

@@
 
-Indeed, serializable transactions, the gold standard of traditional
-ACID databases, are indeed not achievable given a requirement for high
+Indeed, serializable transactions---the gold standard of traditional
+ACID databases---are not achievable given a requirement for high
 availability in the presence of

### 2013-04-01 writeup/model.tex
Foo

@@
 Traditionally, a system provides {\em availability} if every user that
-can contact a correct server eventually receives a response from that server,
-even in the presence of arbitrary, indefinitely long network
-partitions between servers~\cite{gilbert-cap}. As in a standard
-distributed database, designated servers might handle operations on
-different data items;\footnote{There is a further distinction between
-  a \textit{fully replicated} system, in which all servers are
-  replicas for all data items, and a \textit{partially replicated}
-  system, in which at least one server acts as a replica for only a
-  subset of all data items. For generality, and, for applicability to
-  many modern ``sharded'' or ``partitioned'' data storage
-  systems~\cite{dynamo, pnuts, bigtable, spanner, hstore}, we consider partially
+can contact a correct server eventually receives a response from that
+server, even in the presence of arbitrary, indefinitely long network
+partitions between servers~\cite{gilbert-cap}.\footnote{Under this
+  definition, a system that requires a majority of servers to be
+  online is not availabile.} As in a standard distributed database,
+designated servers might handle operations on different data
+items;\footnote{There is a further distinction between a \textit{fully
+    replicated} system, in which all servers are replicas for all data
+  items, and a \textit{partially replicated} system, in which at least
+  one server acts as a replica for only a subset of all data
+  items. For generality, and, for applicability to many modern
+  ``sharded'' or ``partitioned'' data storage systems~\cite{dynamo,
+    pnuts, bigtable, spanner, hstore}, we consider partially
   replicated systems.} a server that can handle an operation for a

### 2013-04-01 writeup/model.tex
foaosdfs

@@
 availability. This is standard in the distributed systems literature
-(e.g., linearizable, and regular register models all concern
-single objects~\cite{herlihy-art}), yet the database literature
-largely focuses on transactions: groups of multiple operations over
-multiple objects. Accordingly, by itself, replica availability is
-insufficient to describe availability guarantees for
-transactions. Additionally, given the choice of \textit{commit} and
-\textit{abort} responses---which signal transaction success or failure
-to a client---we must carefully define transactional availability.
+(e.g., linearizable and regular register models all concern single
+objects~\cite{herlihy-art}), yet the database literature largely
+focuses on transactions: groups of multiple operations over multiple
+objects. Accordingly, by itself, replica availability is insufficient
+to describe availability guarantees for transactions. Additionally,
+given the choice of \textit{commit} and \textit{abort}
+responses---which signal transaction success or failure to a
+client---we must carefully define transactional availability.
 

### 2013-04-01 writeup/evaluation.tex
fixed

@@
 New-Orders, an item's stock will never fall below zero. This is TPC-C
-compliant, but a HAT system might end up with substantially more stock
+compliant, but a HAT system might end up with much more stock
 than in a non-HAT-compliant implementation.
@@
 order to avoid billing a customer twice; this implies a need to
-prevent Lost Update. We can avoid this issue by moving the
+prevent Lost Update. This issue  can be avoided  by moving the
 non-monotonicity to the real world---the carrier that picks up the
@@
 \begin{center}\textbf{Within \texttt{us-east}}\end{center}\vspace{-1.5em}
-\includegraphics[width=\columnwidth]{figs/lan-threads-lats.pdf}\vspace{-1em}
-\includegraphics[width=\columnwidth]{figs/lan-threads-thru.pdf}
+\includegraphics[width=0.90\columnwidth]{figs/lan-threads-lats.pdf}\vspace{-1em}
+\includegraphics[width=0.90\columnwidth]{figs/lan-threads-thru.pdf}
 \begin{center}\textbf{Between \texttt{us-east} and \texttt{us-west-2}}\end{center}\vspace{-1.5em}
-\includegraphics[width=\columnwidth]{figs/wan-threads-lats.pdf}\vspace{-1em}
-\includegraphics[width=\columnwidth]{figs/wan-threads-thru.pdf}
+\includegraphics[width=0.90\columnwidth]{figs/wan-threads-lats.pdf}\vspace{-1em}
+\includegraphics[width=0.90\columnwidth]{figs/wan-threads-thru.pdf}
 \caption{YCSB performance for two clusters of three servers each

### 2013-04-01 writeup/hat-analysis.tex
fixed

@@
 Repeatable Read as providing serializability for all operations except
-predicate-based reads (and so Repeatable Read is identical to
+for predicate-based reads (and so Repeatable Read is identical to
 serializability in a key-value store without predicate-based
@@
 
-There are several session guarantees that we can make with high
-availability:
+Several session guarantees can be made with high availability:
 
@@
 
+\vspace{0.5em}
 \noindent{\textbf{Consistency}} A HAT system can make limited
@@
 monotonic~\cite{calm} operations without the risk of invalidating
-(also monotonic) application-level integrity constraints. Our goal in
+ application-level integrity constraints. Our goal in
 this paper is not to sketch the entire space of consistency models
@@
   serializable via the use of commutative updates
-  (Section~\ref{sec:eval}) but, as we show in the extended version of
+  (Section~\ref{sec:evaluation}) but, as we show in the extended version of
   the paper~\cite{hat-tr}, the problem persists in the general case.}
@@
 Composition cost also varies by combination. For instance, Charron-Bost
-has proven that, to capture causality between $N$ communicating
+ proved that, to capture causality between $N$ communicating
 processes, standard vector-based approaches face an upper bound of

### 2013-04-01 writeup/evaluation.tex
spell

@@
 transactions can be executed with HATs, while the fifth may require
-unavailablity.
+unavailability.
 
@@
 balance increase operations commute, and TA allows the
-maintainence of any foreign-key constraints (e.g.,
+maintenance of any foreign-key constraints (e.g.,
 \texttt{UPDATE/DELETE CASCADE}).
@@
 New-Order presents two challenges: ID assignment and stock
-maintainence. First, each New-Order transaction requires a unique ID
+maintenance. First, each New-Order transaction requires a unique ID
 number for the order. We can create a unique number by, say,
@@
 
-Throughout execution, TPC-C also requires the maintainence of several
+Throughout execution, TPC-C also requires the maintenance of several
 integrity constraints. For example, Consistency Condition 1 (3.3.2.1)

### 2013-04-01 writeup/model.tex
spell

@@
   definition, a system that requires a majority of servers to be
-  online is not availabile.} As in a standard distributed database,
+  online is not available.} As in a standard distributed database,
 designated servers might handle operations on different data

### 2013-04-01 writeup/motivation.tex
spell

@@
 surface---communicating via a hypothetical link traversing the
-planet's diamater---require a minimum 85.1 ms RTT (133.7 ms if sent at
+planet's diameter---require a minimum 85.1 ms RTT (133.7 ms if sent at
 ground level). As services are replicated to multiple, geographically

### 2013-04-01 writeup/hat-analysis.tex
small bump
fixed
not british or assuie

@@
 Writes'', as above, and also ``Dirty Reads'' phenomena.  This
-isolation is Adya's PL-2, and formalised by prohibiting Adya's $G1a-c$
+isolation is Adya's PL-2, and formalized by prohibiting Adya's $G1a-c$
 (or ANSI's $P1$, or ``broad'' $P1$ (2.2) from Berenson et al.). For
@@
 \label{fig:hatcompared}
-\end{figure}
+\end{figure}\vspace{-1em}
 
@@
 
-Composition cost also varies by combination. For instance, Charron-Bost
+Composition cost may also vary by combination. For instance, Charron-Bost
  proved that, to capture causality between $N$ communicating

### 2013-04-01 writeup/hat-analysis.tex
spell

@@
 effects of $T_2$ if it can also observe $T_1$'s effects (or later
-values that supercede $T_1$'s).  Any order on transactions should
+values that supersede $T_1$'s).  Any order on transactions should
 respect the reads-from order.\vspace{.5em}
@@
 Theorem, provably unachievable with high
-availabilty~\cite{gilbert-cap}.
+availability~\cite{gilbert-cap}.
 

### 2013-04-01 writeup/evaluation.tex
bah

@@
 Delivery is challenging due to non-monotonicity. Each Delivery deletes
-a pending order from the NewOrder Table and should be idempotent in
+a pending order from the New-Order Table and should be idempotent in
 order to avoid billing a customer twice; this implies a need to

### 2013-04-01 writeup/evaluation.tex
fixed

@@
 This brief set of experiments is not meant as an exhaustive study, but
-it validates our earlier intuition regarding the cost of violating
+it validates our earlier intuition about the cost of violating
 HAT-compliance. As Deutsch points out, ignoring factors such as
@@
 paying the cost of coordination may be tenable, but, especially as
-services are geo-replicated, the costs increase.
+services are geo-replicated,  costs increase.

### 2013-04-01 writeup/futurework.tex
editing pass
fixed

@@
 
-There are several promising avenues for future work in HAT systems:
+There are several promising avenues for future work in HATs:
 
@@
 \textbf{Where do semantics-based concurrency control and weakened
-  failure models lie in the HAT landscape?} The vast amount of
+  failure models lie in the HAT landscape?} The vast
 literature on semantics-based concurrency control provides hints as to

### 2013-04-01 writeup/futurework.tex
fixed

@@
   failure models lie in the HAT landscape?} The vast
-literature on semantics-based concurrency control provides hints as to
+literature on semantics-based concurrency control indicates 
 when coordination can be avoided while providing ``strongly

### 2013-04-01 writeup/futurework.tex
fixed

@@
 consistent'' outcomes for limited failure scenarios. We have performed
-a narrow case study in this work, but a general taxonomization of
+a narrow case study, but a general taxonomization of
 techniques such as escrow transactions~\cite{escrow},

### 2013-04-04 thebes-code/common/src/main/java/edu/berkeley/thebes/common/persistence/disk/LevelDBPersistenceEngine.java
More metrics integration

@@ -11,16 +11,38 @@ import org.apache.thrift.TException;
 import org.apache.thrift.TSerializer;
 import org.apache.thrift.protocol.TBinaryProtocol;
 import org.iq80.leveldb.*;
+import com.yammer.metrics.Metrics;
+import com.yammer.metrics.core.Gauge;
+import com.yammer.metrics.core.Meter;
+import com.yammer.metrics.core.Timer;
+import com.yammer.metrics.core.TimerContext;
+
 import org.slf4j.*;
 
 import static org.fusesource.leveldbjni.JniDBFactory.*;
 import java.io.*;
+import java.util.concurrent.TimeUnit;
 
 public class LevelDBPersistenceEngine implements IPersistenceEngine {
     DB db;
     LockManager lockManager;
     private static org.slf4j.Logger logger = LoggerFactory.getLogger(LevelDBPersistenceEngine.class);
 
+    private final Meter putCount = Metrics.newMeter(LevelDBPersistenceEngine.class,
+                                                    "put-requests",
+                                                    "requests",
+                                                    TimeUnit.SECONDS);
+
+    private final Meter getCount = Metrics.newMeter(LevelDBPersistenceEngine.class,
+                                                    "get-requests",
+                                                    "requests",
+                                                    TimeUnit.SECONDS);
+
+    private final Timer putLatencyTimer = Metrics.newTimer(LevelDBPersistenceEngine.class,
+                                                           "leveldb-put-latencies");
+
+    private final Timer getLatencyTimer = Metrics.newTimer(LevelDBPersistenceEngine.class,
+                                                           "leveldb-get-latencies");
 
     ThreadLocal<TSerializer> serializer = new ThreadLocal<TSerializer>() {
         @Override
@@ -54,39 +76,61 @@ public class LevelDBPersistenceEngine implements IPersistenceEngine {
         }
 
         db = factory.open(new File(Config.getDiskDatabaseFile()), options);
+
+        Metrics.newGauge(LevelDBPersistenceEngine.class, "locktable-size", new Gauge<Integer>() {
+            @Override
+            public Integer value() {
+                return lockManager.getSize();
+            }
+        });
     }
 
     public boolean put(String key, DataItem value) throws TException {
-        if(value == null) {
-            logger.warn("NULL write to key "+key);
-            return true;
-        }
+        putCount.mark();
+        TimerContext context = putLatencyTimer.time();
 
-        lockManager.lock(key);
         try {
-            DataItem curItem = get(key);
-
[…]

### 2013-04-04 thebes-code/common/src/main/java/edu/berkeley/thebes/common/persistence/util/LockManager.java

@@ -34,6 +34,10 @@ public class LockManager {
         }
     }
 
+    public int getSize() {
+        return lockTable.size();
+    }
+
     private Map<String, LockWaiting> lockTable;
     private Lock tableLock = new ReentrantLock();
 

### 2013-04-04 thebes-code/hat/client/src/main/java/edu/berkeley/thebes/hat/client/ThebesHATClient.java

@@ -13,6 +13,7 @@ import java.util.concurrent.atomic.AtomicInteger;
 
 import javax.naming.ConfigurationException;
 
+import com.yammer.metrics.core.Histogram;
 import edu.berkeley.thebes.common.thrift.ServerAddress;
 import edu.berkeley.thebes.common.thrift.ThriftDataItem;
 import edu.berkeley.thebes.hat.common.thrift.ReplicaService;
@@ -72,7 +73,7 @@ public class ThebesHATClient implements IThebesClient {
     private final Timer latencyBufferedXactMetric = Metrics.newTimer(ThebesHATClient.class, "hat-latencies-buf-xact",
                                                                      TimeUnit.MILLISECONDS, TimeUnit.SECONDS);
     private final Timer latencyPerOperationMetric = Metrics.newTimer(ThebesHATClient.class, "hat-latencies-per-op", TimeUnit.MILLISECONDS, TimeUnit.SECONDS);
-    
+
     private static final AtomicInteger LOGICAL_CLOCK = new AtomicInteger(0);
 
     private ReplicaRouter router;
@@ -92,45 +93,6 @@ public class ThebesHATClient implements IThebesClient {
     //Atomicity data structures
     private AtomicityLevel atomicityLevel = Config.getThebesAtomicityLevel();
     private VersionVector atomicityVersionVector;
-
-    private class TransactionMultiPutCallback implements AsyncMethodCallback<ReplicaService.AsyncClient.put_call> {
-        private Semaphore blockFor;
-        private final int numWrites;
-        private List<Exception> exceptionList = Lists.newArrayList();
-
-        public TransactionMultiPutCallback(int numWrites) {
-            blockFor = new Semaphore(numWrites);
-            this.numWrites = numWrites;
-        }
-
-        @Override
-        public void onComplete(ReplicaService.AsyncClient.put_call put_call) {
-            blockFor.release();
-        }
-
-        @Override
-        public void onError(Exception e) {
-            exceptionList.add(e);
-            blockFor.release();
-        }
-
-        public void blockForWrites() throws TException {
-            try {
-                blockFor.acquire(numWrites);
-            } catch (InterruptedException e) {
-                exceptionList.add(e);
-            } finally {
-                if(!exceptionList.isEmpty()) {
-                    String exceptionString = "Exceptions occured in processing write: ";
-                    for(Exception e : exceptionList) {
-                        exceptionString += e.getMessage() + e.getStackTrace();
-                    }
-
[…]

### 2013-04-04 thebes-code/hat/server/src/main/java/edu/berkeley/thebes/hat/server/dependencies/DependencyResolver.java

@@ -1,5 +1,8 @@
 package edu.berkeley.thebes.hat.server.dependencies;
 
+import com.yammer.metrics.Metrics;
+import com.yammer.metrics.core.Gauge;
+import com.yammer.metrics.core.Meter;
 import org.apache.thrift.TException;
 import org.slf4j.Logger;
 import org.slf4j.LoggerFactory;
@@ -14,6 +17,7 @@ import edu.berkeley.thebes.hat.server.antientropy.clustering.AntiEntropyServiceR
 import java.util.Set;
 import java.util.concurrent.ConcurrentMap;
 import java.util.concurrent.ConcurrentSkipListSet;
+import java.util.concurrent.TimeUnit;
 import java.util.concurrent.atomic.AtomicBoolean;
 import java.util.concurrent.atomic.AtomicInteger;
 import java.util.concurrent.locks.Lock;
@@ -58,7 +62,17 @@ public class DependencyResolver {
     private final ConcurrentMap<Version, AtomicInteger> unresolvedAcksMap;
     
     private final Lock unresolvedAcksLock;
-    
+
+    Meter commitCount = Metrics.newMeter(DependencyResolver.class,
+                                         "dgood-transaction-total",
+                                         "transactions",
+                                         TimeUnit.SECONDS);
+
+    Meter retrievePendingCount = Metrics.newMeter(DependencyResolver.class,
+                                                 "dpending-retrieval-count",
+                                                 "retrievals",
+                                                 TimeUnit.SECONDS);
+
     public DependencyResolver(AntiEntropyServiceRouter router,
             IPersistenceEngine persistenceEngine) {
         this.persistenceEngine = persistenceEngine;
@@ -66,6 +80,13 @@ public class DependencyResolver {
         this.pendingTransactionsMap = Maps.newConcurrentMap();
         this.unresolvedAcksMap = Maps.newConcurrentMap();
         this.unresolvedAcksLock = new ReentrantLock();
+
+        Metrics.newGauge(DependencyResolver.class, "num-pending-versions", new Gauge<Integer>() {
+            @Override
+            public Integer value() {
+                return pendingTransactionsMap.size();
+            }
+        });
     }
 
     public void addPendingWrite(String key, DataItem value) throws TException {
@@ -104,14 +125,6 @@ public class DependencyResolver {
             logger.debug("Committing via unresolved: " + version + " / " + transQueue.numReplicasInvolved + " / " + newPendingWrite.getReplicaIndicesInvolved().size());
             commit(transQueue);
         }
[…]

### 2013-04-23 writeup/hat-appendix.tex
just a taste of appendix

@@
+\normalsize
+
+\section{Model}
+
+\begin{definition}[Change the Matches of a Predicate-Based Read]
+A transaction $T_i$ changes the matches
+of a predicate-based read $r_j$(P:Vset(P)) if $T_i$ installs $x_i$, $x_h$
+immediately precedes $x_i$ in the version order, and $x_h$ matches
+P whereas $x_i$ does not or vice-versa. In this case, we also
+say that $x_i$ changes the matches of the predicate-based read.
+The above deﬁnition identiﬁes $T_i$ to be a transaction where
+a change occurs for the matched set of $r_j$ (P: Vset(P)).
+\end{definition}
+
+\begin{definition}[Directly Read-Depends]
+$T_j$ directly read-depends on transaction $T_i$ if it directly
+item-read-depends or directly predicate-read-depends on $T_i$.
+\end{definition}
+
+\begin{definition}[Directly item-read-depends by $x$]
+$T_j$ directly item-read-depends on transaction $T_i$ if $T_i$ installs some
+  object version $x_i$ and $T_j$ reads $x_i$.
+\end{definition}
+
+\begin{definition}[Directly item-read-depends]
+$T_j$ directly item-read-depends on transaction $T_i$ if $T_j$
+  directly item-read-depends by $x$ on $T_i$ for some data item $x$.
+\end{definition}
+
+\begin{definition}[Directly predicate-read-depends by $P$]
+Transaction $T_j$ directly predicate-read-depends by $P$ on
+transaction $T_i$ if $T_j$ performs an operation $r_j$(P: Vset(P)),
+$x_k$ $\in$ Vset(P), $i = k$ or $x_i \ll x_k$ , and $x_i$ changes the
+matches of $r_j$ (P: Vset(P)).
+\end{definition}
+
+\begin{definition}[Directly predicate-read-depends]
+$T_j$ directly predicate-read-depends on $T_i$ if $T_j$ directly
+  predicate-read-depends by $P$ on $T_i$ for some predicate $P$.
+\end{definition}
+
+\begin{definition}[Directly Anti-Depends]
+Transaction $T_j$ directly anti-depends on transaction $T_i$ if it
+directly item-anti-depends or directly predicate-anti-depends on
+$T_i$.
+\end{definition}
+
+\begin{definition}[Directly item-anti-depends by $x$]
+$T_j$ directly item-anti-depends by $x$ on transaction $T_i$ if $T_i$
+  reads some object version $x_k$ and $T_j$ installs $x$'s next
+  version (after $x_k$) in the version order. Note that the
+  transaction that wrote the later version directly item-anti-depends
+  on the transaction that read the earlier version.
+\end{definition}
+
+\begin{definition}[Directly item-anti-depends]
+$T_j$ directly item-anti-depends on transaction $T_i$ if $T_j$
+  directly item-anti-depends on transaction $T_i$.
+\end{definition}
+
+\begin{definition}[Directly predicate-anti-depends by $P$]
+$T_j$ directly predicate-anti-depends by $P$ on transaction $T_i$ if
+  $T_j$ overwrites an operation $r_i(P:$ Vset(P)). That is, if $T_j$
+  installs a later version of some object that changes the matches of
+  a predicate based read performed by $T_i$.
+\end{definition}
+
+\begin{definition}[Directly predicate-anti-depends by $P$]
+$T_j$ directly predicate-anti-depends on transaction $T_i$ if $T_j$
+  directly predicate anti-depends by $P$ on $T_i$ for some predicate
+  $P$.
+\end{definition}
+
+\begin{definition}[Directly Write-Depends by $x$]
+A transaction $T_j$ directly write-depends by $x$ on transaction $T_i$
+if $T_i$ installs a version $x_i$ and $T_j$ installs $x$'s next
+version (after $x_i$) in the version order.
+\end{definition}
+
+\begin{definition}[Directly Write-Depends]
+A transaction $T_j$ directly write-depends on transaction $T_i$ if
+$T_i$ directly write-depends by $x$ on $T_j$ for some item $x$.
+\end{definition}
+
+\begin{definition}[Directly Session-Depends by $S$]
+A transaction $T_j$ directly session-depends on transaction $T_i$ if
+$T_i$ immediately precedes $T_j$ in the session commit order for session $S$.
+\end{definition}
+
+\begin{definition}[Directly Session-Depends]
+A transaction $T_j$ directly session-depends on transaction $T_i$ if
+$T_j$ directly session-depends on transaction $T_i$ for some session
+$S$.
+\end{definition}
+
 \section{HAT Guarantees}
@@
 
-\begin{definition}[Item Cut Isolation]
+\begin{definition}[Item Cut Non-Isolation (N-ICI)]
+A history $H$ exhibits phenomenon N-ICI if there exists a transaction
+$T_i$ such that $T_i$ directly item-depends by $x$ on more than one
+other transaction.
 \end{definition}
 
-\begin{definition}[Predicate Cut Isolation]
+\begin{definition}[Item Cut Isolation (ICI)]
+A system that provides Item Cut Isolation prohibits phenomenon N-ICI.
 \end{definition}
 
-\begin{definition}[Transactional Atomicity]
+\begin{definition}[Predicate Cut Non-Isolation (N-ICI)]
+A history $H$ exhibits phenomenon N-ICI if there exists a transaction
+$T_i$ such that $P= \{p \mid T_i$ directly predicate-depends by $p$ on another
+other transaction$\}$ and $\bigcap_{p\in P} P \neq \emptyset$.
+\end{definition}
+
+\begin{definition}[Predicate Cut Isolation (ICI)]
+A system that provides Predicate Cut Isolation prohibits phenomenon N-ICI.
+\end{definition}
+
+\begin{definition}[Non-transactional Atomicity (N-TA)]
+A history exhibits phenomenon N-TA if it contains a directed cycle
+consisting of exactly one read-dependency edge and one anti-dependency
+edge.
+\end{definition}
+
+\begin{definition}[Transactional Atomicity (TA)]
+A system that provides transactional atomicity prohibits phenomenon
+N-TA.
+\end{definition}
+
+\begin{definition}[Non-monotonic Reads (N-MR)]
+A history exhibits phenomenon N-MR if it contains a directed cycle
+consisting of a transitive session-dependency between transactions
+$T_j$ and $T_i$ with an anti-dependency edge by $i$ from $T_j$ and a
+read-dependency edge by $i$ into $T_i$.
+\end{definition}
+
+\begin{definition}[Monotonic Reads (MR)]
+A system provides Monotonic Reads if it prohibits phenomenon N-MR.
+\end{definition}
+
+\begin{definition}[Non-monotonic Writes (N-MW)]
+A history exhibits phenomenon N-MW if it contains a directed cycle
+consisting of a transitive session-dependency between transactions
+$T_j$ and $T_i$ and at least one write-dependency edge.
+\end{definition}
+
+\begin{definition}[Monotonic Writes (MW)]
+A system provides Monotonic Writes if it prohibits phenomenon N-MW.
+\end{definition}
+
+\begin{definition}[Writes Do Not Follow Reads (N-WFR)]
+A history exhibits phenomenon N-WFR if, for all committed transactions
+$T_1$, $T_2$, $T_3$ such that $T_2$ write-depends on $T_1$ and $T_3$
+write-depends on $T_2$, $T_3$ does not directly anti-depend on $T_1$.
+\end{definition}
+
+\begin{definition}[Writes Follow Reads (WFR)]
+A system provides Writes Follow Reads if it prohibits phenomenon N-WFR.
+\end{definition}
+
+\begin{definition}[Not-Read Your Writes (N-RYW)]
+A history exhibits phenomenon N-RYW if it contains a directed cycle
+consisting of a transitive session-dependency between transactions
+$T_j$ and $T_i$, at least one anti-dependency edge, and the remainder
+anti-dependency or write-dependency edges.
+\end{definition}
+
+\begin{definition}[Read Your Writes (RYW)]
+A system provides Read Your Writes if it prohibits phenomenon N-RYW.
+\end{definition}
+
+\begin{definition}[PRAM Consistency]
+A system provides PRAM Consistency if it prohibits phenomenon N-MR,
+N-MW, and N-RYW.
+\end{definition}
+
+\begin{definition}[Causal Consistency]
+A system provides Causal Consistency if it provides PRAM Consistency
+and prohibits phenomenon N-WFR.
+\end{definition}
+
+\begin{observation}
+Read Your Writes is not highly available.
+\end{observation}
+
+\begin{corollary}
+PRAM and Causal Consistency are not highly available.
+\end{corollary}
+
+\begin{definition}[Recency Guarantees]
 \end{definition}
@@
 \begin{definition}[Repeatable Read]
-\end{definition}
-
-\begin{definition}[Monotonic Reads]
-\end{definition}
-
-\begin{definition}[Monotonic Writes]
-\end{definition}
-
-\begin{definition}[Writes Follow Reads]
-\end{definition}
-
-\begin{definition}[Read Your Writes]
-\end{definition}
-
-\begin{observation}
-Read Your Writes is not highly available.
-\end{observation}
-
-\begin{corollary}
-PRAM and Causal Consistency are not highly available.
-\end{corollary}
-
-\begin{definition}[PRAM Consistency]
-\end{definition}
-
-\begin{definition}[Causal Consistency]
-\end{definition}
-
-\begin{definition}[Recency Guarantees]
 \end{definition}
@@
 \begin{definition}[Model Strength]
+We say that a consistency model $M_1$ is stronger than model $M_2$ if
+$M_1$ admits a strict subset of the histories admitted by $M_2$.
 \end{definition}

### 2013-04-23 writeup/conclusion.tex
editing pass

@@
 unnecessary choices between availability and transactional semantics.
-We show, however, through analysis and experiments, that the
-availability goals of the NoSQL movement will remain a critical aspect
-of many future data storage systems. Accordingly, we expose a broad
-design space of Highly Available Transactional Systems (HATs), which
-can offer the key benefits of highly available distributed
+However, through our analysis and experiments, we have demonstrated
+that the NoSQL movement's goals of high availability will remain a
+critical aspect of many future data storage systems. Accordingly, we
+expose a broad design space of Highly Available Transactions (HATs),
+which can offer the key benefits of highly available distributed
 systems---''always on'' operation during partitions and low-latency
@@
 field and newfound knowledge of what is possible to achieve, we
-believe this is an area of great opportunity in the coming years.
+believe HATs will be an area of great opportunity in the coming years.
 

### 2013-04-23 writeup/evaluation.tex
editing pass

@@
 
-\begin{myenumerate}
+\begin{myenumerate}\vspace{-.5em}
 \item We revisit traditional database concurrency control with a focus
@@
   properties on public cloud infrastructure.
-\end{myenumerate}
+\end{myenumerate}\vspace{-.5em}
 
@@
 general-purpose read-write transactions in a distributed
-setting~\cite{bernstein-book} may require at least one round trip time
+setting~\cite{bernstein-book} can require at least one round trip time
 (RTT) before committing. As an example, traditional two-phase locking
-for a transaction of length $T$ may require $T$ \texttt{lock}
+for a transaction of length $T$ can require $T$ \texttt{lock}
 operations and will require at least one \texttt{lock} and
@@
 these operations requires coordination, either with other database
-replicas or with a lock service. If this coordination mechanism is
+servers or with a lock service. If this coordination mechanism is
 unavailable, transactions cannot safely commit. Similarly, optimistic
@@
 to a designated master or coordination service for each of these
-classic algorithms.  The cost of the (minimum one) round trip will be
-determined by the deployment environment, as we saw in
-Section~\ref{sec:motivation}; we will demonstrate this cost on public
-cloud infrastructure in the next section.
+classic algorithms.  The cost of round trips will be determined by the
+deployment environment, as we saw in Section~\ref{sec:motivation}; we
+will demonstrate this cost on public cloud infrastructure in the next
+section.
 
@@
 isolation in Gray's original proposal~\cite{gray-isolation} do not
-degrade gracefully in the presence of partial failures. (Note, however,
-that lock-based protocols \textit{do} offer the benefit of recency
-guarantees.) While multi-versioned storage systems allow for a variety
-of transactional guarantees, we have not seen proposals for truly weak
-isolation (e.g., non-``tentative update'' schemes) in this context.
-The MDCC~\cite{mdcc} protocol offers Read Committed isolation with
-Lost Update avoidance but is similarly unavailable due to its reliance
-on preventing write conflicts. Chan and Gray's read-only transactions
-provide read-only transactions with item-cut isolation, causal
-consistency, and transactional atomicity (session PL-2L~\cite{adya})
-but are unavailable in the presence of coordinator
+degrade gracefully in the presence of partial failures. (Note,
+however, that lock-based protocols \textit{do} offer the benefit of
+recency guarantees.) While multi-versioned storage systems allow for a
+variety of transactional guarantees, we have not seen proposals for
+truly weak isolation (e.g., non-``tentative update'' schemes) in this
+context.  The MDCC~\cite{mdcc} protocol offers Read Committed
+isolation with Lost Update avoidance but is similarly unavailable due
+to its reliance on preventing write conflicts. Chan and Gray's
+read-only transactions provide read-only transactions with item-cut
+isolation, causal consistency, and transactional atomicity (session
+PL-2L~\cite{adya}) but are unavailable in the presence of coordinator
 failure~\cite{readonly}, similar to read-only and write-only
@@
 between transactions that write to the same data
-item~\cite{raynal-causal}.  Bolt-on causal consistency~\cite{bolton},
-Brantner's S3 database~\cite{kraska-s3}, and
-Bayou~\cite{sessionguarantees} can all provide variants of session
+item~\cite{raynal-causal}.  Brantner's S3 database~\cite{kraska-s3}
+and Bayou~\cite{sessionguarantees} can all provide variants of session
 PL-2L with high availability, but none provide this HAT functionality
-without substantial modification; Swift~\cite{swift} is closest to
-providing full HAT operation in a sticky context. As  seen, it
-is possible to implement many guarantees weaker than
-serializability---including guarantees that are achievable with high
-availability---and still not achieve high availability.
+without substantial modification; Swift~\cite{swift} and bolt-on
+causal consistency~\cite{bolton} are closest to providing full HAT
+semantics operation in a sticky context. As seen, it is possible to
+implement many guarantees weaker than serializability---including
+guarantees that are achievable with high availability---and still not
+achieve high availability.
 
@@
 \vspace{.5em}\noindent\textbf{Commutativity and Monotonicity} Recent
-work on the CALM Theorem~\cite{calm} and Commutative and Replicated Data
-Types~\cite{crdt} demonstrates that, if updates logically commute,
-then they can often be safely performed in different orders at
-different replicas. Accordingly, as long as all writes are delivered
-to all replicas, then a system executing monotonic logic with
-commutative operators may not suffer from application-level
+work on the CALM Theorem~\cite{calm} and Commutative and Replicated
+Data Types~\cite{crdt} demonstrates that, if updates logically
+commute, then they can often be safely performed in different orders
+at different replicas. Accordingly, as long as all writes are
+delivered to all replicas, then a system executing monotonic logic
+with commutative operators may not suffer from application-level
 consistency anomalies as a result of Lost Update or Write Skew
@@
 will not, in general, be able to maintain application-level
-consistency constraints with HATs alone. Moreover, applications with
+consistency constraints with HATs alone--in particular, applications
 requiring bounded update visibility latency should opt for
@@
 of HAT-compliance in an application context, we consider a concrete
-application: the TPC-C benchmark. Surprisingly, four of five
-transactions can be executed with HATs, while the fifth may require
-unavailability.
+application: the TPC-C benchmark~\cite{tpcc}. Surprisingly, four of
+five transactions can be executed with HATs, while the fifth may
+require unavailability.
 
@@
 Update. Accordingly, HATs cannot provide compliant TPC-C execution but
-can maintain uniqueness constraints. Second, the New-Order transaction
+can stil maintain uniqueness constraints. Second, the New-Order transaction
 decrements inventory counts: what if the count becomes negative?
@@
 New-Orders, an item's stock will never fall below zero. This is TPC-C
-compliant, but a HAT system might end up with much more stock
+compliant, but a HAT system might end up with more stock
 than in a non-HAT-compliant implementation.
@@
 Delivery is challenging due to non-monotonicity. Each Delivery deletes
-a pending order from the New-Order Table and should be idempotent in
+a pending order from the New-Order table and should be idempotent in
 order to avoid billing a customer twice; this implies a need to
@@
 tables but, given the ability to update rows in both tables atomically
-via transactional atomicity, can be easily maintained. Consistency
-Conditions 4 through 12 (3.3.2.4-12) can similarly be satisfied by
-applying updates atomically across tables. Consistency Conditions 2
-and 3 (3.3.2.2-3) concern order ID assignment and are more
-problematic, as we have discussed.  Finally, while TPC-C is not
+via TA, can be easily maintained. Consistency Conditions 4 through 12
+(3.3.2.4-12) can similarly be satisfied by applying updates atomically
+across tables. Consistency Conditions 2 and 3 (3.3.2.2-3) concern
+order ID assignment and are problematic. Finally, while TPC-C is not
 subject to multi-key anomalies, we note that many TPC-E isolation
-tests, such as one that requires simultaneously modifying a product
-description and its image, are also passable using HATs.
+tests (i.e., simultaneously modifying a product description and its
+thumbnail) are also achievable using HATs.
 

### 2013-04-23 writeup/hat-analysis.tex
bah
fixed
editing pass

@@
 either high availability or sticky high availability. They offer
-substantial latency and availability benefits compared to traditional
-distributed databases, yet they cannot achieve all the traditional
-semantics. In this section, we delineate ACID, distributed
-consistency, and session consistency levels which can be achieved with
-high availability (transactional atomicity, variants of Repeatable
-Read isolation, and many session guarantees), those with sticky high
-availability (read your writes, PRAM and causal consistency) and the
-properties which cannot be provided in a HAT system (those preventing
-Lost Update and Write Skew, or with recency).  We present a full
-summary of these results in Section~\ref{sec:hat-summary}.
+latency and availability benefits over traditional distributed
+databases, yet they cannot achieve all the traditional semantics. In
+this section, we delineate ACID, distributed consistency, and session
+consistency levels which can be achieved with high availability
+(transactional atomicity, variants of Repeatable Read isolation, and
+many session guarantees), those with sticky high availability (read
+your writes, PRAM and causal consistency) and the properties which
+cannot be provided in a HAT system (those preventing Lost Update and
+Write Skew, or with recency).  We present a full summary of these
+results in Section~\ref{sec:hat-summary}.
 
@@
 To begin, we present well-known semantics that can be achieved in HAT
-systems. We offer proof-of-concept highly available algorithms but our
-goal is \textit{not} to provide optimal or even efficient
-implementations. These guarantees are fairly straightforward and have
-been introduced---albeit with greater brevity and discussion
-omitted---in a preliminary workshop paper~\cite{hat-hotos}. In our
-examples, we exclusively consider read and write operations. We denote
-a write of value $v$ to data item $d$ as $w_d(v)$ and a read from data
-item $d$ returning $v$ as $r_d(v)$ and assume that all data items have
-the null value, $\bot$, at database initialization. Unless otherwise
-specified, all example transactions commit.
+systems. We offer proof-of-concept highly available algorithms, but
+our goal is \textit{not} to provide optimal or even efficient
+implementations. In our examples, we exclusively consider read and
+write operations. We denote a write of value $v$ to data item $d$ as
+$w_d(v)$ and a read from data item $d$ returning $v$ as $r_d(v)$ and
+assume that all data items have the null value, $\bot$, at database
+initialization. Unless otherwise specified, all transactions in the
+examples commit.
 
@@
 To begin, \textbf{Read Uncommitted} isolation is captured by Adya as
-PL-1,requiring that each transaction's writes are ordered consistently
-according to a single total order on transactions. This prohibits
-Adya's phenomenon $G0$, also called ``Dirty Writes''~\cite{adya}. If
-two transactions write to the same set of data items, then the final
+$PL-1$, requiring that each transaction's writes are ordered
+consistently according to a single total order on transactions. This
+prohibits Adya's phenomenon $G0$, also called ``Dirty
+Writes''~\cite{adya}. We form a graph of transactions, with edges from
+one transaction to another if the former writes to an object before
+the latter writes to the same object. PL-1 prohibits cycles in this
+graph, requiring some total order on writes. Accordingly, the final
 database state cannot contain the ``earlier'' transaction's writes to
-the data items. For example, in the below example, $T_3$ should
+the data items: for example, in the below example, $T_3$ should
 \emph{eventually} only read $a=b=1$ or $a=b=2$ but not $a=2, b=1$ or
@@
 \begin{align*}
-\small
+\small\vspace{-.5em}
 T_1 &: w_x(1)~w_y(1)
 \\T_2 &: w_x(2)~w_y(2)
-\\T_3 &: r_x(a)~r_y(b)\\[-2em]
+\\T_3 &: r_x(a)~r_y(b)\vspace{-.5em}
 \end{align*}
-All later properties will strengthen PL-1.
+Later properties will strengthen Read Uncommitted.
 
@@
 read locks~\cite{gray-isolation} and others based on multiple
-versions. The implementations typically have recency and monotonicity
-properties beyond the simple meaning of the name, which is what is
+versions. The implementations often provide recency and monotonicity
+properties beyond the simple meaning from the name, which is what is
 expressed in the implementation-agnostic definition: under Read
@@
 $T_2$ aborts, $T_3$ should never read $a=3$:
-\vspace{-.5em}
 \begin{align*}
-\small
+\small\vspace{-.5em}
 T_1 &: w_x(1)~w_x(2)
 \\T_2 &: w_x(3)\\
-T_3 &: r_x(a)\\[-2em]
+T_3 &: r_x(a)\vspace{-.5em}
 \end{align*}
@@
 data~\cite{gray-isolation}, is the only kind of non-serializable
-behavior allowed. When referring to ``Repeatable Read,'' we will
-follow these definitions, which, as we will soon show, are for a
-property that can be offered in a highly available way.
+behavior allowed. In this paper, we reserve the use of ``Repeatable
+Read'' to refer to these definitions, which we will soon show are not
+HAT-compliant.
 
@@
 
-A stronger achievable property asks that a consistent cut spans both a
+A stronger achievable property requires consistent cuts to span both a
 transaction's predicate reads (e.g., \texttt{SELECT WHERE}) as well as
 its item reads.  We call this \textit{Predicate Cut
-  Isolation}\footnote{Oracle provides an isolation level called
-  ``Statement Level Read Consistency''~\cite{adya}; this is analogous
-  to Predicate Cut Isolation at the level of a single operation within
-  a multi-operation transaction. We do not consider this isolation
-  model here as it is subsumed by standard Snapshot Isolation and we
-  believe our discussion is easily extended to incorporate it.} and it
+  Isolation} and it
 prohibits Phantoms. Predicate Cut Isolation is also achievable in HAT
@@
 broadcast (with some additional client metadata) and is
-achievable in HAT systems. We omit a full discussion of the algorithm
-and instead refer the reader to the extended version of this
-paper~\cite{hat-tr}.  
-%space{.5em}
+achievable in HAT systems.
+
+To achieve TA, clients buffer their writes until commit, sending the
+final write to each data item to each respective server. Along with
+each write, clients attach a the transaction ID and a list of all data
+items written to in the transaction, $L_v$. Upon receiving a new write
+to item $d$, a server places the write into a \textit{pending} queue
+and notifies all other servers responsible for a data item in $L_v$
+that it has received a write to $d$ with the respective transaction
+ID. Once a server receives acknowledgments with appropriate
+transaction ID for all items in $L_v$, it moves the write from
+\textit{pending} to the \textit{stable} database (which applies the
+writes according to highest-ID-wins). Accordingly, all writes that
+should be made atomic with a write in \textit{stable} are guaranteed
+to be present on their respective servers, either in \textit{pending}
+or \textit{stable}. In the presence of internal aborts, writes may
+require 2 RTTs: one to check constraints and one to execute the above.
+
+For reads, each client tracks which values it should read. It
+maintains a mapping (vector) from data items to transaction IDs. Upon
+transaction start, the client clears its mapping, and, when it
+performs a read from a data item, it sends the server its current
+value for the item. If the value is empty, the server reads from
+\texttt{stable}. If the value is non-empty, the server either returns
+the matching value from \texttt{pending} or a higher-ID value from
+\texttt{pending}---along with $L_v$ for the write. When the client
+receives a response, it updates its local vector with
+$L_v$. Accordingly, even if transaction's writes to two keys $x$ and
+$y$ are present on the $x$ server's \textit{stable} and the $y$
+server's \textit{pending}, a client's reads from $x$ then $y$ will
+return a transactionally atomic pair of values because the $y$ server
+will know to check pending.
+
+This implementation is entirely masterless and both reads and writes
+do not block for coordination. One key property is that we do not, in
+general, dictate when writes become visible. Accordingly, it is highly
+available. However, we can optimize write visibility delay by ensuring
+that clients are sticky with disjoint groups of servers. Once all
+transaction values are \textit{pending} within a group of servers,
+values can be made \textit{stable} (as opposed to once they are
+present on all servers). For example, in a multi-datacenter setting,
+all clients and servers within a datacenter may form a sticky group.
 
@@
 Our models have not yet considered interactions \textit{across}
-transactions: we have not guaranteed any ordering or continuity
-between transactions, other than that there exists some arbitrary
-ordering (via Read Uncommitted). In the distributed systems
-literature, many useful \textit{safety} guarantees span multiple
-application actions. In particular, \textit{session guarantees} are
-used to describe guarantees across transactions within a given
-\textit{session}, ``an abstraction for the sequence of...operations
-performed during the execution of an
+transactions, other than that there exists some arbitrary ordering on
+them (via Read Uncommitted). In the distributed systems literature,
+many useful \textit{safety} guarantees span multiple,
+non-transactional operations. In particular, \textit{session
+  guarantees} are used to describe guarantees across transactions
+within a given \textit{session}, ``an abstraction for the sequence
+of...operations performed during the execution of an
 application''~\cite{sessionguarantees}. Informally a session describes
@@
 a session, subsequent reads to a given object ``never return any
-previous values''; reads from each item progress forward in a total
+previous values''; reads from each item progress according to a total
 order. The ordering of reads should respect any externally observable
@@
 \vspace{.5em}\noindent\textit{{Monotonic writes}} requires that each
-session's writes be applied to any replica in the order they were
-submitted by the client. Any order on transactions should also be
-consistent with any precedence that a global observer would see.
+session's writes become visible in the order they were submitted by
+the client. Any order on transactions should also be consistent with
+any precedence that a global observer would see.
 
@@
 in succession:
-\vspace{-.5em}
 \begin{align*}
-\small
+\small\vspace{-.5em}
 T_1 &: w_x(1)
-\\T_2 &: r_x(a)
+\\T_2 &: r_x(a)\vspace{-.5em}
 \end{align*}
@@
 scenario. Accordingly, read your writes, and, by proxy, causal
-consistency and PRAM require stickiness. Read your writes is ``free''
-in a sticky system, while the remaining causality and PRAM guarantees
-can be accomplished with the algorithms for achieving the remaining
-session guarantees.
+consistency and PRAM require stickiness. Read your writes is provided
+by default in a sticky system, while the remaining causality and PRAM
+guarantees can be accomplished with the algorithms for achieving the
+remaining session guarantees.
 
@@
 \noindent{\textbf{Consistency}} A HAT system can make limited
-consistency guarantees. It can often execute commutative and logically
-monotonic~\cite{calm} operations without the risk of invalidating
- application-level integrity constraints. Our goal in
-this paper is not to sketch the entire space of consistency models
-that are achievable (see Section~\ref{sec:futurework}). We
-specifically evaluate TPC-C transaction semantics under HAT
-consistency guarantees in Section~\ref{sec:evaluation}.
+application-level consistency guarantees. It can often execute
+commutative and logically monotonic~\cite{calm} operations without the
+risk of invalidating application-level integrity constraints. We do
+not attempt to sketch the entire space of \textit{application-level}
+consistency properties that are achievable in HAT systems (see
+Section~\ref{sec:futurework}) but we specifically evaluate TPC-C
+transaction semantics under HAT consistency guarantees in
+Section~\ref{sec:evaluation}.
 
@@
 transactions' effects survive $F$ server faults requires at least
-$F+1$ correct nodes.
+$F+1$ non-failing nodes.
 
@@
 \begin{align*}
-\small
+\small\vspace{-.5em}
 T_1 &: r_x(a) w_x(a+2)
-\\T_2 &: w_x(2)
+\\T_2 &: w_x(2)\vspace{-.5em}
 \end{align*}
@@
 \begin{align*}
-\small
+\small\vspace{-.5em}
 T_1 &: r_x(100)~w_x(100+20=120)
-\\T_2 &: r_x(100)~w_x(100+30=130)
+\\T_2 &: r_x(100)~w_x(100+30=130)\vspace{-.5em}
 \end{align*}
@@
 Isolation, and Cursor Stability guarantees are all unavailable.
-Repeatable Read (in the sense of Gray~\cite{gray-isolation}, Berenson
-et al.~\cite{ansicritique}, or Adya~\cite{adya}) and One-Copy
-Serializability need to prevent both Lost Update and Write Skew; this
-means that they are also inherently unavailable.
+Repeatable Read (defined by to Gray, Berenson et al., and Adya) and
+One-Copy Serializability need to prevent both Lost Update and Write
+Skew. These prevention requirements mean that these models are
+inherently unavailable.
 
@@
 one-copy serializability and linearizability is called \textit{strong
-  (or strict) one-copy serializability}~\cite{adya} (e.g., Google's
+  (or strict) one-copy serializability}~\cite{adya} (e.g.,
 Spanner~\cite{spanner}). It is also common, particularly in systems
@@
 force an available system to violate any recency bound, so recency
-bounds are not enforceable in HAT systems.
+bounds are not enforceable by HAT systems.
 
@@
 We summarize our results in Table~\ref{table:hatcompared}. A wide
-range of isolation levels are achievable in a HAT systems, including
+range of isolation levels are achievable in HAT systems, including
 transactional atomicity, cut isolation, and several session
@@
 other models, while considering the (large) power set of all
-compatible models (e.g., the diagram depicts 96 possible highly
-available combinations) hints at the vast expanse of consistency
-models found in the literature. This taxonomy  is \textit{not}
-exhaustive, but we believe it lends substantial clarity into the
-relationships between a large subset of the prominent ACID and
-distributed consistency models. Additional read/write transaction
-semantics that we have omitted should be easily classifiable based on
-the available primitives and HAT-incompatible anomaly prevention we
-have already discussed.
+compatible models (e.g., the diagram depicts 96 possible HAT
+combinations) hints at the vast expanse of consistency models found in
+the literature. This taxonomy is not exhaustive, but we believe it
+lends substantial clarity into the relationships between a large
+subset of the prominent ACID and distributed consistency
+models. Additional read/write transaction semantics that we have
+omitted should be easily classifiable based on the available
+primitives and HAT-incompatible anomaly prevention we have already
+discussed.
 
@@
 \label{fig:hatcompared}
-\end{figure}\vspace{-1em}
+\end{figure}
 
@@
 \begin{align*}
-\small
+\small\vspace{-.5em}
 T_1 &: w_x(1)~w_y(1)
 \\T_2 &: w_x(2)~w_y(2)
-\\T_3 &: r_x(a)~r_y(b)
+\\T_3 &: r_x(a)~r_y(b)\vspace{-.5em}
 \end{align*}

### 2013-04-23 writeup/hat-appendix.tex
editing pass

@@
 A history $H$ exhibits phenomenon N-ICI if there exists a transaction
-$T_i$ such that $P= \{p \mid T_i$ directly predicate-depends by $p$ on another
-other transaction$\}$ and $\bigcap_{p\in P} P \neq \emptyset$.
+$T_i$ such that $P= \{p \mid T_i$ directly predicate-read-depends by
+$p$ on another other transaction$\}$ and $\bigcap_{p\in P} P \neq
+\emptyset$. (TODO: this isn't necessarily it)
 \end{definition}

### 2013-04-23 writeup/hat-vldb.tex
References purge
cleaning
tweaks
fixed
editing pass

@@
 \let\endproof\relax
+\usepackage{draftwatermarkbottom, draftwatermarktop}
 
-\usepackage{times, graphicx, url, xcolor, multicol, minipage, subfig, tikz, draftwatermarkbottom, draftwatermarktop, amsthm}
+\newcommand{\pbnote}[1]{{\color{red}{#1 ---P.B.}}\\}
+
+\interfootnotelinepenalty=10000
+
+\usepackage{times, graphicx, url, xcolor, multicol, minipage, subfig, tikz, amsthm}
 \usepackage{balance}  % for  \balance command ON LAST PAGE  (only there!)
@@
 \newtheorem{corollary}{Corollary}
+
+\makeatletter
+\def\th@plain{%
+  \thm@notefont{}% same as heading font
+  \itshape % body font
+}
+\def\th@definition{%
+  \thm@notefont{}% same as heading font
+  \normalfont % body font
+}
+\makeatother
 
@@
 {\author{Peter Bailis{\fontsize{12}{14}$^\dagger$}, Aaron Davidson{\fontsize{12}{14}$^\dagger$}, Alan Fekete{\fontsize{12}{14}$^\diamond$}, Ali Ghodsi{\fontsize{12}{14}$^{\dagger,\ddagger}$}, Joseph M.
-Hellerstein{\fontsize{12}{14}$^\dagger$}, Ion Stoica{\fontsize{12}{14}$^\dagger$}\\{{\fontsize{12}{14}$^\dagger$}\hspace{.5mm}UC Berkeley\hspace{4mm}{\fontsize{12}{14}$^\diamond$}\hspace{.5mm}University of Sydney\hspace{4mm}{\fontsize{12}{14}$^\ddagger$}\hspace{.5mm}KTH/Royal Institute of Technology}}}
+Hellerstein{\fontsize{12}{14}$^\dagger$}, Ion Stoica{\fontsize{12}{14}$^\dagger$}\\{\affaddr{\fontsize{12}{14}$^\dagger$}\hspace{.5mm}UC Berkeley\hspace{4mm}{\fontsize{12}{14}$^\diamond$}\hspace{.5mm}University of Sydney\hspace{4mm}{\fontsize{12}{14}$^\ddagger$}\hspace{.5mm}KTH/Royal Institute of Technology}}}
 \maketitle
@@
 \begin{abstract}
-To minimize network latency and remain online during partial failure
+To minimize network latency and remain online during server failures
 and network partitions, many modern distributed data storage systems
@@
 Available Transactions (HATs), or transactional guarantees that do not
-suffer from unavailability during system partitions or incur high
-network latency.  We introduce a new taxonomy of highly available
-systems and analyze existing ACID and distributed consistency
-guarantees to identify which can and cannot be achieved in HAT
-systems. We analytically and experimentally quantify the performance
-and availability benefits of HAT systems as well as the associated
-semantic compromises.
+suffer unavailability during system partitions or incur high network
+latency.  We introduce a new taxonomy of highly available systems and
+analyze existing ACID and distributed consistency guarantees to
+identify which can and cannot be achieved in HAT systems. We
+analytically and experimentally quantify the performance and
+availability benefits of HATs as well as their necessary semantic
+compromises.
 \end{abstract}
@@
 % vldb_sample.bib is the name of the Bibliography in this case
-\footnotesize
+%\footnotesize
+\scriptsize
+\balance
+
 \bibliography{hat-vldb}
+
 % You must have a proper ".bib" file
@@
 % Example of an appendix; typically would start on a new page
+%\pagebreak
+%
 \pagebreak
-
 \begin{appendix}
-\input{availability-appendix.tex}
+%\input{availability-appendix.tex}
 
@@
 \end{document}
+
+% megastore
+% redblue
+% gemini
+% redblue
+% granola
+% bernstein-concurrency
+% surviving-failures-bodik
+% blooml
+% google-availability
+% cops

### 2013-04-23 writeup/intro.tex
editing pass

@@
 
-The prior decade has seen a large shift in the design of large-scale
+The last decade has seen a large shift in the design of large-scale
 database backends, from the use of transactional
@@
 widespread adoption of loosely-consistent distributed key-value
-stores~\cite{bigtable, pnuts, dynamo}. Fundamental to this trend was
-the 2000 introduction of Brewer's CAP Theorem, which stated that a
+stores~\cite{bigtable, pnuts, dynamo}. A prominent event in this shift
+was the 2000 introduction of Brewer's CAP Theorem, which stated that a
 highly available system cannot provide strong consistency guarantees
@@
 linearizability: the ability to read the most recent write to a data
-item that is replicated across servers~\cite{herlihy-art}. Yet this
-result---regarding a fairly narrow form of distributed
-consistency---is often conflated with ACID database
-properties~\cite{brewer-slides, hn, foundation-article}, leading to
-substantial confusion about the relationships between replica
-consistency, transactional isolation, and high availability. However,
-transactions \textit{are} useful for programmers as evidenced by the
-recent resurgence of transactional data stores such as Google's
-Spanner~\cite{spanner}. Unfortunately, these systems are not highly
-available in the presence of partitions~\cite{middleware-db,
+item that is replicated across servers~\cite{herlihy-art}. Yet the
+implications of CAP---which concerns a fairly narrow form of
+distributed consistency---are often conflated with the ability to
+provide ACID database properties~\cite{brewer-slides, hn,
+  foundation-article}; this has led to substantial confusion regarding
+replica consistency, transactional isolation, and high
+availability. While transactions \textit{are} useful for programmers
+as evidenced by the recent resurgence of transactional data stores
+such as Spanner~\cite{spanner}, these systems are unfortunately not
+highly available in the presence of partitions~\cite{middleware-db,
   foundation-article, hstore,generalizedsnapshot, mdcc,
@@
 Indeed, serializable transactions---the gold standard of traditional
-ACID databases---are not achievable given a requirement for high
-availability in the presence of
-partitions~\cite{davidson-survey}. However, database systems have a
-long tradition of providing weaker isolation and consistency
-guarantees~\cite{adya, ansicritique, gray-virtues, gray-isolation,
-  kemme-thesis}. Due to their concurrency and overall performance
-benefits, weak isolation models are widely employed, overwhelmingly as
-the default setting in today's traditional ``ACID'' and up-and-coming
-``NewSQL'' databases and often as the maximum guarantee offered
-(Section~\ref{sec:modernacid}). While weak isolation levels do not
-provide serializability for general-purpose transactions, they are
-apparently strong enough to deliver acceptable behavior to many
-application programmers and are substantially stronger than the
-semantics of current highly available systems. This raises a natural
-question: which of these alternative semantics can be provided while
-remaining highly available?
+ACID databases---are not achievable with high availability in the
+presence of partitions~\cite{davidson-survey}. However, database
+systems have a long tradition of providing weaker isolation and
+consistency guarantees~\cite{adya, ansicritique, gray-virtues,
+  gray-isolation, kemme-thesis}. Due to their concurrency and overall
+performance benefits, weak isolation models are widely employed,
+overwhelmingly as the default setting in today's traditional ``ACID''
+and up-and-coming ``NewSQL'' databases and often as the maximum
+guarantee offered (Section~\ref{sec:modernacid}). While weak isolation
+levels do not provide serializability for general-purpose
+transactions, they are apparently strong enough to deliver acceptable
+behavior to many application programmers and are substantially
+stronger than the semantics provided by current highly available
+systems. This raises a natural question: which of these alternative
+semantics can be provided while remaining highly available?
 
-Unfortunately, the relationship of the CAP Theorem and ACID semantics
-has not been well explored. We have a strong understanding of weak
-isolation in the single-server context from which it
+Unfortunately, the relationship of high availability and ACID
+semantics has not been well explored. We have a strong understanding
+of weak isolation in the single-server context from which it
 originated~\cite{adya, ansicritique, gray-isolation} and many papers
@@
 isolation~\cite{daudjee-snapshot,generalizedsnapshot, kemme-snapshot,
-  walter}. The distributed computing, programming language, and
-parallel hardware literature also contains many consistency models for
-single operations on replicated objects~\cite{ ceri-mechanism,
-  chen-mechanism, pnuts, herlihy-art, eiger, cac, sessionguarantees}. However, the literature has largely ignored semantic guarantees for
-multiple operations operating on multiple data items in a highly
-available distributed environment.
+  walter}. Additionally, the distributed computing, programming
+language, and parallel hardware literature contains many
+consistency models for single operations on replicated objects~\cite{
+  ceri-mechanism, chen-mechanism, pnuts, herlihy-art, eiger, cac,
+  sessionguarantees}. However, the literature has largely ignored
+semantic guarantees for multiple operations operating on multiple data
+items in a highly available distributed environment.
 
-Our main contributions are as follows. We fill the gap between, on one
-hand, the plethora of previously proposed isolation and consistency
-models, and, on the other hand, the goal of high availability.  We
-classify which, among the wide array of models, are highly available,
-denoting them as {\em Highly Available Transactions (HATs)}. In doing
-so, we demonstrate that many \textit{implementations} of HAT semantics
-are not highly available but that their unavailability is not
-\textit{inherently} required. Our investigation shows, somewhat
-surprisingly, that besides serializability, only Snapshot Isolation
-(SI) and Repeatable Reads (RR) are inherently unavailable, with most
-other isolation levels being available. We also show that many of the
-weaker data consistency models from distributed systems are
-HAT-compliant.
+Our main contributions in this paper are as follows. We fill the gap
+between, on one hand, the plethora of previously proposed isolation
+and consistency models, and, on the other hand, the goal of high
+availability.  We classify which among the wide array of models are
+achievable with high availability, denoting them as {\em Highly
+  Available Transactions} (HATs). In doing so, we demonstrate that many
+implementations of HAT semantics are not highly available but
+that their unavailability is not inherently required. Our
+investigation shows, somewhat surprisingly, that besides
+serializability, only Snapshot Isolation and Repeatable Reads are not
+HAT-compliant, while most other isolation levels are achievable with
+high availability. We also show that many weak data consistency models
+from distributed systems are also HAT-compliant.
 
-More precisely, the reason why fairly strong models like Snapshot
-Isolation (SI) and Repeatable Reads (RR) are not highly available is
-that they require detecting conflicts between concurrent updates (as
-needed for preventing Lost Updates or Write Skew phenomena). However,
-Read Committed isolation, Transactional Atomicity, and many of the
-weaker data consistency models from database and distributed systems
-can be implemented in a highly available manner. For several
-guarantees, such as Transactional Causal Predicate Cut Isolation, we
-consider a modified form of high availability in which clients ``stick
-to'' (i.e., have affinity with) at least one server (a property which
-is often implicit in the distributed systems
-literature~\cite{herlihy-art, eiger, cac} but which requires
-\textit{explicit} consideration in a client-server replicated database
-context). This sticky availability is easily implementable and widely
-employed~\cite{eiger, vogels-defs} but is strictly less available than
-traditional definitions of high availability.
+Making HAT classifications requires study of several phenomena. For
+example, Snapshot Isolation (SI) and Repeatable Reads (RR) are not
+HAT-compliant because they require detecting conflicts between
+concurrent updates (as needed for preventing Lost Updates or Write
+Skew phenomena). However, Read Committed isolation, Transactional
+Atomicity, and many of the weaker data consistency models from
+database and distributed systems are achievable via novel algorithms
+that rely on multi-versioning and limited client-side caching. For
+several guarantees, such as causal consistency with phantom prevention
+and ANSI Repeatable Read, we consider a modified form of high
+availability in which clients ``stick to'' (i.e., have affinity with)
+at least one server (a property which is often implicit in the
+distributed systems literature~\cite{herlihy-art, eiger, cac} but
+which requires explicit consideration in a client-server replicated
+database context). This sticky availability is easily implementable
+and widely employed~\cite{eiger, vogels-defs} but is strictly less
+available than traditional definitions of high availability.
 
@@
 guarantees can be delivered with availability in face of partitions,
-but which guarantees are worthwhile in practice? We attempt to
-objectively study the virtues and limitations of both classic and HAT
-systems by surveying practitioner accounts and research literature,
-performing experimental analyses on modern cloud infrastructure, and
-analyzing representative applications for their semantic
-requirements. While our results are not definitive, they suggest that
-HATs offer a one to three order of magnitude latency decrease compared
-to traditional distributed serializability protocols, and they can
-provide acceptable semantics for a wide range of programs, especially
-those with monotonic logic and commutative updates~\cite{calm, blooml,
-  crdt}. HAT systems can also enforce arbitrary foreign key
-constraints on multi-item updates and sometimes provide limited
-uniqueness guarantees. However, particularly for programs with
-non-monotonic logic, HATs may fall short, requiring
-sometimes-unavailable mechanisms that are more
+but which guarantees are worthwhile in practice? We study the virtues
+and limitations of both classic and HAT systems by surveying
+practitioner accounts and research literature, performing experimental
+analysis on modern cloud infrastructure, and analyzing representative
+applications for their semantic requirements. While our results are
+not definitive, they suggest that HATs offer a one to three order of
+magnitude latency decrease compared to traditional distributed
+serializability protocols, and they can provide acceptable semantics
+for a wide range of programs, especially those with monotonic logic
+and commutative updates~\cite{calm, blooml, crdt}. HAT systems can
+also enforce arbitrary foreign key constraints on multi-item updates
+and sometimes provide limited uniqueness guarantees. However,
+particularly for programs with non-monotonic logic, HATs can fall
+short, requiring sometimes-unavailable mechanisms that are more
 coordination intensive. Overall, this paper quantifies the benefits

### 2013-04-23 writeup/model.tex
editing pass

@@
 
-Some algorithms assume a model in which the client always contacts the
-same server across subsequent operations. As we will discuss in
-Section~\ref{sec:hats}, clients can ensure continuity between
-operations (e.g., reading their prior updates to a data item) by
-maintaining affinity or ``stickiness'' with a server or set of
-servers~\cite{vogels-defs}. Simultaneously maintaining availability
-and stickiness entails fate-sharing between the client and its 
-server: if a client's sticky server fail, or, equivalently, if the
-client is partitioned from its sticky server, then it may face
+Many distributed algorithms assume a model in which clients always
+contact the same (sets of) server(s) across subsequent operations. As
+we will discuss in Section~\ref{sec:hats}, clients can ensure
+continuity between operations (e.g., reading their prior updates to a
+data item) by maintaining affinity or ``stickiness'' with a server or
+set of servers~\cite{vogels-defs}. Simultaneously maintaining
+availability and stickiness requires fate-sharing between the client
+and its server: if a client's sticky servers fail or if the client is
+partitioned from its sticky servers, then it may face
 unavailability. We say that a system provides \textit{sticky
   availability} if, whenever a client's operation is executed against
-a correct server that has observed all of its prior operations, it eventually
-receives a response, even in the presence of indefinitely long
-partitions. A client may choose to become sticky available by acting
-as a server itself; for example, a client might cache its reads and
-writes~\cite{bolton, sessionguarantees, swift}.
-
-Availability implies sticky availability, since availability does not
-require that the contacted server has seen all the previous operations
-of a client, whereas sticky availability is the special case in which
-the system only guarantees correctness if the contacted server has
-seen all the prior operations from the client.
+a correct server that has observed all of its prior operations, it
+eventually receives a response, even in the presence of indefinitely
+long partitions. A client may choose to become sticky available by
+acting as a server itself; for example, a client might cache its reads
+and writes~\cite{bolton, sessionguarantees, swift}. Any guarantee
+achievable in a highly available system is achievable in a sticky high
+availability system but not vice-versa.
 
@@
 availability. This is standard in the distributed systems literature
-(e.g., linearizable and regular register models all concern single
-objects~\cite{herlihy-art}), yet the database literature largely
-focuses on transactions: groups of multiple operations over multiple
-objects. Accordingly, by itself, replica availability is insufficient
-to describe availability guarantees for transactions. Additionally,
-given the choice of \textit{commit} and \textit{abort}
-responses---which signal transaction success or failure to a
-client---we must carefully define transactional availability.
+(e.g., distributed register models such as linearizability all
+concern single objects~\cite{herlihy-art}), yet the database
+literature largely focuses on transactions: groups of multiple
+operations over multiple objects. Accordingly, by itself, replica
+availability is insufficient to describe availability guarantees for
+transactions. Additionally, given the choice of \textit{commit} and
+\textit{abort} responses---which signal transaction success or failure
+to a client---we must take care in defining transactional availability.
 

### 2013-04-23 writeup/relatedwork.tex
editing pass

@@
 unachievable~\cite{davidson-survey} and Brewer's CAP Theorem has
-attracted considerable attention~\cite{gilbert-cap}. The research has
+attracted considerable attention~\cite{gilbert-cap}. Recent work has
 begun to offer hints regarding the connection between ``weak
@@
 high availability is relatively unexplored. With the exception of a
-preliminary workshop paper introducing HATs~\cite{hat-hotos}--- which
-this work expands, providing impossibility results, a full
-taxonomization, and experimental and application analysis---we believe
-this paper is the first to explore this area. Nonetheless, several
-studies rigorously classify non-HAT semantics: Wisemann et al. have
-proposed a three parameter classification for one-copy serializable
-database systems, performing a more extensive classification than our
-overview in Section~\ref{sec:evaluation}, including eager database
-replication techniques~\cite{kemme-classification}. Additionally
-several earlier surveys discuss distributed mechanisms for achieving
-one-copy serializability~\cite{wisemann-survey}, linearizability,
-consistent snapshots, and recency bounds~\cite{ceri-mechanism,
-  chen-mechanism}.
+preliminary workshop paper discussing highly available RC and
+I-CI~\cite{hat-hotos}--- which this work expands, providing additional
+algorithms and impossibility results, a full taxonomization, and
+experimental and application analysis---we believe this paper is the
+first to explore this area. Nonetheless, several studies rigorously
+classify non-HAT semantics: Wisemann et al. have proposed a three
+parameter classification for one-copy serializable database systems,
+performing a more extensive classification than our overview in
+Section~\ref{sec:evaluation}, including eager database replication
+techniques~\cite{kemme-classification}. Additionally several earlier
+surveys discuss distributed mechanisms for achieving one-copy
+serializability~\cite{wisemann-survey}, linearizability, consistent
+snapshots, and recency bounds~\cite{ceri-mechanism, chen-mechanism}.
 
@@
 provide transactional causal consistency guarantees with varying
-availability and represent a new class of (sticky) HAT
-systems.
+availability and represent a new class of sticky HAT systems.
+
+\pbnote{add TA stuff}
 

### 2013-04-25 writeup/intro.tex
adam's comments

@@
 
-The last decade has seen a large shift in the design of large-scale
-database backends, from the use of transactional
+The last decade has seen a significant shift in the design of
+large-scale database backends, from the use of transactional
 RDBMSs~\cite{bernstein-book, gray-isolation, gray-virtues} to the
 widespread adoption of loosely-consistent distributed key-value
-stores~\cite{bigtable, pnuts, dynamo}. A prominent event in this shift
-was the 2000 introduction of Brewer's CAP Theorem, which stated that a
-highly available system cannot provide strong consistency guarantees
-in the presence of network partitions~\cite{brewer-slides}. As
+stores~\cite{bigtable, pnuts, dynamo}. The 2000 introduction of
+Brewer's CAP Theorem, which stated that a highly available system
+cannot provide strong consistency guarantees in the presence of
+network partitions, helped spur this shift~\cite{brewer-slides}. As
 formally proven~\cite{gilbert-cap}, the CAP Theorem concerns
@@
 item that is replicated across servers~\cite{herlihy-art}. Yet the
-implications of CAP---which concerns a fairly narrow form of
-distributed consistency---are often conflated with the ability to
-provide ACID database properties~\cite{brewer-slides, hn,
-  foundation-article}; this has led to substantial confusion regarding
-replica consistency, transactional isolation, and high
-availability. While transactions \textit{are} useful for programmers
-as evidenced by the recent resurgence of transactional data stores
-such as Spanner~\cite{spanner}, these systems are unfortunately not
-highly available in the presence of partitions~\cite{middleware-db,
-  foundation-article, hstore,generalizedsnapshot, mdcc,
-  krikellas-bargain, eiger, walter, calvin}.
+implications of CAP---which concern a fairly narrow form of
+distributed consistency---have often been conflated with the ability
+to provide ACID database properties~\cite{brewer-slides, hn,
+  foundation-article}; this misunderstanding has led to substantial
+confusion regarding replica consistency, transactional isolation, and
+high availability. The recent resurgence of transactional systems like
+Spanner~\cite{spanner} suggests that programmers value these
+semantics, but, unfortunately, existing transactional data stores do
+not provide availability in the presence of
+partitions~\cite{middleware-db, foundation-article,
+  hstore,generalizedsnapshot, mdcc, krikellas-bargain, eiger, walter,
+  calvin}.
 
@@
 consistency guarantees~\cite{adya, ansicritique, gray-virtues,
-  gray-isolation, kemme-thesis}. Due to their concurrency and overall
-performance benefits, weak isolation models are widely employed,
-overwhelmingly as the default setting in today's traditional ``ACID''
-and up-and-coming ``NewSQL'' databases and often as the maximum
-guarantee offered (Section~\ref{sec:modernacid}). While weak isolation
-levels do not provide serializability for general-purpose
-transactions, they are apparently strong enough to deliver acceptable
-behavior to many application programmers and are substantially
-stronger than the semantics provided by current highly available
-systems. This raises a natural question: which of these alternative
-semantics can be provided while remaining highly available?
+  gray-isolation, kemme-thesis}. Today's ACID and NewSQL databases
+often employ weak isolation models due to their concurrency and
+performance benefits; weak isolation is overwhelmingly the default
+setting in these stores and is often the only option offered
+(Section~\ref{sec:modernacid}). While weak isolation levels do not
+provide serializability for general-purpose transactions, they are
+apparently strong enough to deliver acceptable behavior to many
+application programmers and are substantially stronger than the
+semantics provided by current highly available systems. This raises a
+natural question: which of these semantics can be provided while
+remaining highly available?
 
-Unfortunately, the relationship of high availability and ACID
-semantics has not been well explored. We have a strong understanding
+Unfortunately, the relationship between high availability and ACID
+semantics has not been well-explored. We have a strong understanding
 of weak isolation in the single-server context from which it
@@
   walter}. Additionally, the distributed computing, programming
-language, and parallel hardware literature contains many
+languages, and parallel hardware literature contains many
 consistency models for single operations on replicated objects~\cite{
@@
 Our main contributions in this paper are as follows. We fill the gap
-between, on one hand, the plethora of previously proposed isolation
-and consistency models, and, on the other hand, the goal of high
-availability.  We classify which among the wide array of models are
-achievable with high availability, denoting them as {\em Highly
-  Available Transactions} (HATs). In doing so, we demonstrate that many
-implementations of HAT semantics are not highly available but
-that their unavailability is not inherently required. Our
-investigation shows, somewhat surprisingly, that besides
-serializability, only Snapshot Isolation and Repeatable Reads are not
-HAT-compliant, while most other isolation levels are achievable with
-high availability. We also show that many weak data consistency models
-from distributed systems are also HAT-compliant.
+between the plethora of previously proposed isolation and consistency
+models and the goal of high availability.  We classify which among the
+wide array of models are achievable with high availability, denoting
+them as {\em Highly Available Transactions} (HATs). In doing so, we
+demonstrate that although many implementations of HAT semantics are
+not highly available, this is an artifact of the implementations
+rather than an inherent property of the semantics. Our investigation
+shows that, besides serializability, only Snapshot Isolation and
+Repeatable Reads are not HAT-compliant, while most other isolation
+levels are achievable with high availability. We also demonstrate that
+many weak data consistency models from distributed systems are also
+HAT-compliant.
 
-Making HAT classifications requires study of several phenomena. For
-example, Snapshot Isolation (SI) and Repeatable Reads (RR) are not
-HAT-compliant because they require detecting conflicts between
-concurrent updates (as needed for preventing Lost Updates or Write
-Skew phenomena). However, Read Committed isolation, Transactional
+Our investigation is based on both impossibility results and several
+constructive proof-of-concept algorithms. For example, Snapshot
+Isolation (SI) and Repeatable Reads (RR) are not HAT-compliant because
+they require detecting conflicts between concurrent updates (as needed
+for preventing Lost Updates or Write Skew phenomena), which we prove
+is unavailable. However, Read Committed isolation, Transactional
 Atomicity, and many of the weaker data consistency models from
@@
 
-This taxonomy allows us to determine that a wide range of semantic
-guarantees can be delivered with availability in face of partitions,
-but which guarantees are worthwhile in practice? We study the virtues
-and limitations of both classic and HAT systems by surveying
-practitioner accounts and research literature, performing experimental
-analysis on modern cloud infrastructure, and analyzing representative
-applications for their semantic requirements. While our results are
-not definitive, they suggest that HATs offer a one to three order of
-magnitude latency decrease compared to traditional distributed
-serializability protocols, and they can provide acceptable semantics
-for a wide range of programs, especially those with monotonic logic
-and commutative updates~\cite{calm, blooml, crdt}. HAT systems can
-also enforce arbitrary foreign key constraints on multi-item updates
-and sometimes provide limited uniqueness guarantees. However,
-particularly for programs with non-monotonic logic, HATs can fall
-short, requiring sometimes-unavailable mechanisms that are more
-coordination intensive. Overall, this paper quantifies the benefits
+Our results demonstrate that a wide range of semantic guarantees can
+be delivered with availability in face of partitions, but which
+guarantees are worthwhile in practice? We study the virtues and
+limitations of both classic and HAT systems by surveying practitioner
+accounts and research literature, performing experimental analysis on
+modern cloud infrastructure, and analyzing representative applications
+for their semantic requirements. Our experiences with a HAT prototype
+running across multiple geo-replicated datacenters indicate that HATs
+offer a one to three order of magnitude latency decrease compared to
+traditional distributed serializability protocols, and they can
+provide acceptable semantics for a wide range of programs, especially
+those with monotonic logic and commutative updates~\cite{calm, blooml,
+  crdt}. HAT systems can also enforce arbitrary foreign key
+constraints on multi-item updates and sometimes provide limited
+uniqueness guarantees. However, particularly for programs with
+non-monotonic logic, HATs can fall short, requiring
+sometimes-unavailable mechanisms that are more
+coordination-intensive. Overall, this paper quantifies the benefits of
 Highly Available Transactions can offer as well as the necessary

### 2013-04-25 writeup/conclusion.tex
another pass

@@
 unnecessary choices between availability and transactional semantics.
-However, through our analysis and experiments, we have demonstrated
-that the NoSQL movement's goals of high availability will remain a
-critical aspect of many future data storage systems. Accordingly, we
-expose a broad design space of Highly Available Transactions (HATs),
-which can offer the key benefits of highly available distributed
-systems---''always on'' operation during partitions and low-latency
-operations---while also providing a family of transactional isolation
-levels and replica semantics that have been adopted in practice.  In
-taxonomizing this design space, we show that many transactional
-guarantees are achievable with high availability, while a few, like
-preventing Lost Update and Write Skew, are not. From a research
-perspective, this offers a long-overdue connection between storage
-semantics studied in the database and distributed systems communities.
-
-With the above understanding in hand, the next challenge is to return
-to issues of performance and systems architecture within the HAT
-design space.  Based on our understanding of what is desired in the
-field and newfound knowledge of what is possible to achieve, we
-believe HATs will be an area of great opportunity in the coming years.
+Through our analysis and experiments, we have demonstrated how goals
+of high availability will remain a critical aspect of many future data
+storage systems. Accordingly, we expose a broad design space of Highly
+Available Transactions (HATs), which can offer the key benefits of
+highly available distributed systems---''always on'' operation during
+partitions and low-latency operations---while also providing a family
+of transactional isolation levels and replica semantics that have been
+adopted in practice.  In taxonomizing this design space, we show that
+many transactional guarantees are achievable with high availability,
+while a few, like preventing Lost Update and Write Skew, are not. This
+offers a long-overdue connection between storage semantics studied in
+the database and distributed systems communities. The subsequent
+challenge is to return to issues of performance and systems
+architecture within the HAT design space.  Based on our understanding
+of what is desired in the field and newfound knowledge of what is
+possible to achieve, we believe HATs will be an area of great
+opportunity in the coming years.
 

### 2013-04-25 writeup/hat-analysis.tex
another pass

@@
 latency and availability benefits over traditional distributed
-databases, yet they cannot achieve all the traditional semantics. In
-this section, we delineate ACID, distributed consistency, and session
+databases, yet they cannot achieve all possible semantics. In this
+section, we delineate ACID, distributed consistency, and session
 consistency levels which can be achieved with high availability
@@
 many session guarantees), those with sticky high availability (read
-your writes, PRAM and causal consistency) and the properties which
-cannot be provided in a HAT system (those preventing Lost Update and
-Write Skew, or with recency).  We present a full summary of these
-results in Section~\ref{sec:hat-summary}.
+your writes, PRAM and causal consistency), and properties that cannot
+be provided in a HAT system (those preventing Lost Update and Write
+Skew, or with recency).  We present a full summary of these results in
+Section~\ref{sec:hat-summary}.
 
@@
 formal definitions, semantics, proofs and observations in our extended
-Technical Report and opt for a more informal presentation in this
-paper~\cite{hat-tr}.
+Technical Report~\cite{hat-tr} and opt for a more informal
+presentation here.
 
@@
 our goal is \textit{not} to provide optimal or even efficient
-implementations. In our examples, we exclusively consider read and
-write operations. We denote a write of value $v$ to data item $d$ as
-$w_d(v)$ and a read from data item $d$ returning $v$ as $r_d(v)$ and
-assume that all data items have the null value, $\bot$, at database
-initialization. Unless otherwise specified, all transactions in the
-examples commit.
+implementations. As is standard in the distributed database
+literature, in our sample implementations, we consider a set of
+servers acting as replicas for a set of data items with which
+non-co-located clients interact. In our examples, we exclusively
+consider read and write operations, denoting a write of value $v$ to
+data item $d$ as $w_d(v)$ and a read from data item $d$ returning $v$
+as $r_d(v)$. We assume that all data items have the null value,
+$\bot$, at database initialization, and, unless otherwise specified,
+all transactions in the examples commit.
 
@@
 To begin, \textbf{Read Uncommitted} isolation is captured by Adya as
-$PL-1$, requiring that each transaction's writes are ordered
+\textit{PL-1}, requiring that each transaction's writes are ordered
 consistently according to a single total order on transactions. This
 prohibits Adya's phenomenon $G0$, also called ``Dirty
-Writes''~\cite{adya}. We form a graph of transactions, with edges from
-one transaction to another if the former writes to an object before
-the latter writes to the same object. PL-1 prohibits cycles in this
-graph, requiring some total order on writes. Accordingly, the final
-database state cannot contain the ``earlier'' transaction's writes to
-the data items: for example, in the below example, $T_3$ should
-\emph{eventually} only read $a=b=1$ or $a=b=2$ but not $a=2, b=1$ or
-$a=1, b=2$:
+Writes''~\cite{adya}. More formally, we require a total ordering on
+writes to each data item and we create a graph of transactions with
+edges from one transaction to another if the former overwrites the
+latter's write to the same object. \textit{PL-1} prohibits cycles in
+this graph, requiring some total order on writes. Accordingly, the
+final database state cannot contain the ``earlier'' transaction's
+writes to the data items: for example, in the below example, $T_3$
+should \emph{eventually} only read $a=b=1$ or $a=b=2$ but not $a=2,
+b=1$ or $a=1, b=2$:
 \begin{align*}
@@
 Writes'', as above, and also ``Dirty Reads'' phenomena.  This
-isolation is Adya's PL-2, and formalized by prohibiting Adya's $G1a-c$
-(or ANSI's $P1$, or ``broad'' $P1$ (2.2) from Berenson et al.). For
-instance, in the example below, $T_3$ should never see $a=1$, and, if
-$T_2$ aborts, $T_3$ should never read $a=3$:
+isolation is Adya's \textit{PL-2} and is formalized by prohibiting
+Adya's \textit{G1\{a-c\}} (or ANSI's $P1$, or ``broad'' $P1$ [2.2]
+from Berenson et al.). For instance, in the example below, $T_3$
+should never see $a=1$, and, if $T_2$ aborts, $T_3$ should never read
+$a=3$:
 \begin{align*}
@@
 It is fairly easy for a HAT system to prevent ``Dirty Reads'': if each
-transaction never writes uncommitted data to the database, then
+client never writes uncommitted data to shared copies of the databaseq, then
 transactions will never read each others' dirty data. As a simple
@@
 
-``Repeatable Read'' isolation is a confusing property. Some IBM
-products use the term for fully serializable
-isolation~\cite{hat-hotos}. Gray~\cite{gray-isolation}, Berenson et
-al.~\cite{ansicritique}, and Adya's PL-2.99~\cite{adya} all interpret
-Repeatable Read as providing serializability for all operations except
-for predicate-based reads (and so Repeatable Read is identical to
-serializability in a key-value store without predicate-based
-access). In this model, the \textit{Phantom Problem}, whereby two
-successive predicate-based reads return different
-data~\cite{gray-isolation}, is the only kind of non-serializable
-behavior allowed. In this paper, we reserve the use of ``Repeatable
-Read'' to refer to these definitions, which we will soon show are not
-HAT-compliant.
+``Repeatable Read'' isolation is a confusing property. For example,
+some IBM products use the term for fully serializable
+isolation~\cite{hat-hotos}, while Gray~\cite{gray-isolation}, Berenson
+et al.~\cite{ansicritique}, and Adya's PL-2.99~\cite{adya} all
+interpret Repeatable Read as providing serializability for all
+operations except for predicate-based reads (and so Repeatable Read is
+identical to serializability in a key-value store without
+predicate-based access). In this latter model, the \textit{Phantom
+  Problem}, whereby two successive predicate-based reads return
+different data~\cite{gray-isolation}, is the only kind of
+non-serializable behavior allowed. In this paper, we reserve the use
+of ``Repeatable Read'' to refer to these definitions, which we will
+soon show are not HAT-compliant.
 
 However, As Berenson et al.~\cite{ansicritique} discuss, the ANSI SQL
-specification allows many additional behaviors under the term
-Repeatable Read isolation.  \textbf{ANSI Repeatable Read} requires
-that, along with respecting Read Committed isolation, each transaction
-will only read one version of each data item that it did not itself
-produce (preventing ``Fuzzy Read,'' or P2). In the example below,
-$T_3$ must read $a=1$:
+specification allows several additional behaviors under ``Repeatable
+Read'' isolation.  \textbf{ANSI Repeatable Read} requires that, along
+with respecting Read Committed isolation, each transaction will only
+read one version of each data item that it did not itself produce
+(preventing ``Fuzzy Read,'' or P2). In the example below, $T_3$ must
+read $a=1$:
 \begin{align*}
@@
 of the data item should not change during the transaction. By itself,
-this property is extremely weak---we can always read $\bot$ for each
-data item---and easily achieved in a HAT system by caching the values
-read.
+this property is weak---we can always read $\bot$ for each data
+item---and easily achieved in a HAT system by caching the values
+read. To prevent confusion, we call this isolation model \textbf{Item
+  Cut Isolation}: all the transaction's reads should see values from a
+non-changing consistent cut or snapshot over the data items. It is
+possible to satisfy Item Cut Isolation with high availability by using
+multi-versioning on each replica and ensuring that each of a
+transaction's reads return an appropriately chosen version (e.g.,
+identified by vector clock).
 
-Between our Repeatable Read (PL-2.99) and the ANSI definition, is a
-concept we will call \textbf{Item Cut Isolation}: that all the
-transaction's reads should see values from a non-changing consistent
-cut or snapshot over the data items. It is possible to satisfy Item
-Cut Isolation with high availability by using multi-versioning on each
-replica and ensuring that each of a transaction's reads return an
-appropriately chosen version (e.g., identified by vector clock).
-
-A stronger achievable property requires consistent cuts to span both a
-transaction's predicate reads (e.g., \texttt{SELECT WHERE}) as well as
-its item reads.  We call this \textit{Predicate Cut
-  Isolation} and it
-prohibits Phantoms. Predicate Cut Isolation is also achievable in HAT
-systems via similar multi-versioning as above.
+A stronger achievable property that expands on Item Cut Isolation
+requires consistent cuts to span both a transaction's predicate reads
+(e.g., \texttt{SELECT WHERE}) as well as its item reads.  We call this
+\textbf{Predicate Cut Isolation} and it prohibits Phantoms. Predicate
+Cut Isolation is also achievable in HAT systems via similar
+multi-versioning as above.
 
@@
 Transactional atomicity (TA) is core to ACID guarantees. Although, at
-least by the ACID acronym, it is not an ``isolation'' property,
-transactional atomicity restricts the ability to view the effects of
-partially completed transactions. Under transactional atomicity, once
-some of the effects of a transaction $T_1$ are observed by another
-transaction $T_2$, thereafter all effects of $T_1$ are observed by
-$T_2$. Together with item cut isolation, TA prevents Read Skew
-anomalies (Berenson et al.'s A5A~\cite{ansicritique}). As an example
-of TA, because $T_2$ has read $T_1$'s write to $y$, $T_2$ must observe
-$b=c=1$ (or later versions for each key):
+least by the ACID acronym, it is not an ``isolation'' property, TA
+restricts transactions' ability to view the effects of partially
+completed transactions. Under TA, once some of the effects of a
+transaction $T_i$ are observed by another transaction $T_j$,
+thereafter all effects of $T_i$ are observed by $T_2$. Together with
+item cut isolation, TA prevents Read Skew anomalies (Berenson et al.'s
+A5A~\cite{ansicritique}). As an example of TA, because $T_2$ has read
+$T_1$'s write to $y$, $T_2$ must observe $b=c=1$ (or later versions
+for each key):
 \vspace{-.5em}
@@
 
-Perplexingly, discussions of TA are absent from existing ones of
-weak isolation. This is perhaps again due to the single-node
-context in which prior work was developed: on a single server (or a
-fully replicated database), TA is achievable via lightweight locking
-and/or local concurrency control over data items~\cite{gstore}. In
-contrast, in a distributed environment, TA over arbitrary groups of
+Perplexingly, discussions of TA are absent from existing treatment of
+weak isolation. This is perhaps again due to the single-node context
+in which prior work was developed: on a single server (or a fully
+replicated database), TA is achievable via lightweight locking and/or
+local concurrency control over data items~\cite{gstore}. In contrast,
+in a distributed environment, TA over arbitrary groups of
 non-colocated items is considerably more difficult to achieve with
-high availability: servers need to know when all of a transaction's
-updates are present on their respective replicas. However, replicas do
-not need to agree on when to reveal a new value to clients, which
-would require consensus; accordingly, TA only requires reliable
-broadcast (with some additional client metadata) and is
-achievable in HAT systems.
+high availability: servers need to ensure that, regardless of replica
+selection, all of a client's accesses will read a sufficiently
+up-to-date/``consistent'' set of data items. However, replicas do not
+need to agree on when to choose to reveal new sets of values clients,
+which would require consensus. Instead, TA only requires the
+equivalent of reliable broadcast (with some additional client
+metadata) and is achievable in HAT systems.
 
-To achieve TA, clients buffer their writes until commit, sending the
-final write to each data item to each respective server. Along with
-each write, clients attach a the transaction ID and a list of all data
-items written to in the transaction, $L_v$. Upon receiving a new write
-to item $d$, a server places the write into a \textit{pending} queue
-and notifies all other servers responsible for a data item in $L_v$
-that it has received a write to $d$ with the respective transaction
-ID. Once a server receives acknowledgments with appropriate
-transaction ID for all items in $L_v$, it moves the write from
-\textit{pending} to the \textit{stable} database (which applies the
-writes according to highest-ID-wins). Accordingly, all writes that
-should be made atomic with a write in \textit{stable} are guaranteed
-to be present on their respective servers, either in \textit{pending}
-or \textit{stable}. In the presence of internal aborts, writes may
-require 2 RTTs: one to check constraints and one to execute the above.
+We can achieve TA with a two-phase multi-versioning algorithm. Clients
+buffer their writes until commit, sending the final write to each data
+item to each respective server. Along with each write, clients attach
+a the transaction ID and a list of all data items written to in the
+transaction, $L_v$. Upon receiving a new write to item $d$, a server
+places the write into a \texttt{pending} queue and notifies all other
+servers responsible for a data item in $L_v$ that it has received a
+write to $d$ with the respective transaction ID. Once a server
+receives acknowledgments with appropriate transaction ID for all items
+in $L_v$, it moves the write from \texttt{pending} to the
+\texttt{stable} database (which applies the writes according to
+highest-ID-wins). Accordingly, all writes that should be made atomic
+with a write in \texttt{stable} are guaranteed to be present on their
+respective servers, either in \texttt{pending} or
+\texttt{stable}. Given the possibility of internal aborts, writes to
+\texttt{pending} may require 2 RTTs: one to check constraints and one
+to execute the above (but these checks \textit{cannot} cause other
+transactions to block, as would be the case in a two-phase
+\textit{commit} protocol).
 
@@
 transaction start, the client clears its mapping, and, when it
-performs a read from a data item, it sends the server its current
-value for the item. If the value is empty, the server reads from
-\texttt{stable}. If the value is non-empty, the server either returns
-the matching value from \texttt{pending} or a higher-ID value from
-\texttt{pending}---along with $L_v$ for the write. When the client
-receives a response, it updates its local vector with
-$L_v$. Accordingly, even if transaction's writes to two keys $x$ and
-$y$ are present on the $x$ server's \textit{stable} and the $y$
-server's \textit{pending}, a client's reads from $x$ then $y$ will
-return a transactionally atomic pair of values because the $y$ server
-will know to check pending.
+performs a read from a data item, it sends the server its current ID
+for the item. If the client does not supply an ID, the server returns
+the latest value from \texttt{stable}. If the ID is non-empty, the
+server either returns the matching value from \texttt{pending} or a
+value with higher ID from \texttt{pending}---along with $L_v$ for the
+write. When the client receives a response, it updates its local
+vector with $L_v$. Accordingly, even if transaction's writes to two
+keys $x$ and $y$ are present on the replica for $x$'s \texttt{stable}
+and the replica for $y$'s \texttt{pending}, a client's reads from $x$
+then $y$ will return a transactionally atomic pair of values because
+the replica for $y$ will know to check \texttt{pending}.
 
@@
 that clients are sticky with disjoint groups of servers. Once all
-transaction values are \textit{pending} within a group of servers,
-values can be made \textit{stable} (as opposed to once they are
+transaction values are \texttt{pending} within a group of servers,
+values can be made \texttt{stable} (as opposed to once they are
 present on all servers). For example, in a multi-datacenter setting,
@@
 
-In the words of Berenson et al., \textit{Lost Update} occurs when one
+Berenson et al. define \textit{Lost Update} as when one
 transaction $T1$ reads a given data item, a second transaction $T2$
@@
   serializable via the use of commutative updates
-  (Section~\ref{sec:evaluation}) but, as we show in the extended version of
-  the paper~\cite{hat-tr}, the problem persists in the general case.}
-To prevent this from happening, either $T_1$ or $T_2$ should not have
-committed. Each client's respective server might try to detect that
-another write occurred, but this requires knowing the version of the
-latest write to $x$. In our example, this reduces to a requirement for
-linearizability, which is, via Gilbert and Lynch's proof of the CAP
-Theorem, provably unachievable with high
+  (Section~\ref{sec:evaluation}) but the problem persists in the
+  general case.}  To prevent this from happening, either $T_1$ or
+$T_2$ should not have committed. Each client's respective server might
+try to detect that another write occurred, but this requires knowing
+the version of the latest write to $x$. In our example, this reduces
+to a requirement for linearizability, which is, via Gilbert and
+Lynch's proof of the CAP Theorem, provably unachievable with high
 availability~\cite{gilbert-cap}.
@@
 TPC-C~\cite{snapshot-serializable}---but, as a generalization of Lost
-Update, it is also unavailable to HATs.
+Update, it is also unavailable to HAT systems.
 
@@
 
-Data storage systems make various recency guarantees on reads of data
-items. As we have discussed, one of the most famous is
-linearizability~\cite{herlihy-art}, which states that reads will
+Distributed data storage systems often make various recency guarantees
+on reads of data items. As we have discussed, one of the most famous
+is linearizability~\cite{herlihy-art}, which states that reads will
 return the last completed write to a data item, and there are several
@@
 
-We summarize our results in Table~\ref{table:hatcompared}. A wide
-range of isolation levels are achievable in HAT systems, including
+As we summarize in Table~\ref{table:hatcompared}, a wide range of
+isolation levels are achievable in HAT systems, including
 transactional atomicity, cut isolation, and several session
@@
 combine TA and P-CI, we have transactional snapshot reads. We can
-achieve RC, MR, and RYW by simply sticking clients to servers. And we
-can also combine unavailable models---for example, an unavailable
-system might provide PRAM and One-Copy
+achieve RC, MR, and RYW by simply sticking clients to servers. We can
+also combine unavailable models---for example, an unavailable system
+might provide PRAM and One-Copy
 Serializability~\cite{daudjee-session}.
@@
 
-\subsection{Discussion}
+\subsection{Additional Discussion}
 \label{sec:discussion}
@@
 specifically addressing model composition, transactional atomicity
-versus linearizability, and stickiness requirements.
+versus linearizability, and benefits of stickiness.
 
@@
 readers at once---what is different? First, in linearizable (and safe
-and regular) systems, writes are made visible to all clients
+and regular) systems, writes are made visible to clients
 \textit{immediately} after they finish. With transactional atomicity,
@@
 \vspace{.5em}\noindent\textbf{Visibility and Stickiness} Sticky
-availability can result in much better write \textit{visibility}:
-clients will be able to safely read writes more quickly in a sticky
-available system. In the model we discussed, it is possible to achieve
-several properties like monotonic reads in a highly available system
-by waiting to reveal a write until all servers have seen it and its
+availability can improve write \textit{visibility}: clients will be
+able to safely read writes more quickly in a sticky available
+system. In the model we discussed, it is possible to achieve several
+properties like monotonic reads in a highly available system by
+waiting to reveal a write until all servers have seen it and its
 relevant dependencies. However, this incurs severe visibility

### 2013-04-25 writeup/hat-appendix.tex
another pass

@@
 
+\begin{figure}[H]
+\begin{align*}
+\small
+T_1 &: w_x(1)\\
+T_2 &: w_x(2)\\
+T_3 &: r_x(1)~r_x(2)
+\end{align*}
+\caption{Example of \textit{N-ICI} anomaly.}
+\label{fig:nici-history}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\begin{tikzpicture}[node distance=3cm]
+\tikzstyle{tx}=[draw=none, fill=none]
+\node[tx] (T1) at (0,0) {$T_1$};
+\node[tx] (T2) at (0,-1) {$T_2$};
+\node[tx] (T3) at (2,-.5) {$T_3$};
+
+\draw[->] (T1) edge node[above]{$wr_x$} (T3);
+\draw[->] (T2) edge node[below]{$wr_x$} (T3);
+\end{tikzpicture}
+\caption{DSG for Figure~\ref{fig:nici-history}.}
+\label{fig:nici-dsg}
+\end{figure}
+
 \begin{definition}[Item Cut Isolation (ICI)]
@@
 A history exhibits phenomenon N-TA if it contains a directed cycle
-consisting of exactly one read-dependency edge and one anti-dependency
-edge.
+consisting of exactly one read-dependency edge by $x$ from $T_i$ to
+$T_j$ and one anti-dependency edge by $y$ from $T_j$ to $T_i$ and
+$T_i$'s corresponding read from $x$ does not precede its corresponding
+read from $y$.
 \end{definition}
+
+\begin{figure}[H]
+\begin{align*}
+\small
+T_1 &: w_x(1)~w_y(1)\\
+T_2 &: w_x(2)~w_y(2)\\
+T_3 &: r_x(2)~r_y(1)
+\end{align*}
+\caption{Example of \textit{N-TA} anomaly.}
+\label{fig:nta-history}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\begin{tikzpicture}[node distance=3cm]
+\tikzstyle{tx}=[draw=none, fill=none]
+\node[tx] (T1) at (0,0) {$T_1$};
+\node[tx] (T2) at (2,0) {$T_2$};
+\node[tx] (T3) at (2,-1.5) {$T_3$};
+
+\draw[->] (T1) edge node[sloped, above]{$ww_{\{x, y\}}$} (T2);
+\draw[->] (T1) edge [bend right] node[sloped, below]{$wr_y$} (T3);
+\draw[->] (T2) edge [bend right] node[left]{$wr_x$} (T3);
+\draw[dashed, ->] (T3) edge [bend right] node[right]{$rw_y$} (T2);
+\end{tikzpicture}
+\caption{DSG for Figure~\ref{fig:nta-history}.}
+\label{fig:nta-dsg}
+\end{figure}
 
@@
 \end{definition}
+
+\begin{figure}[H]
+\begin{align*}
+\small
+T_1 &: w_x(1)\\
+T_2 &: w_x(2)\\
+T_3 &: r_x(2)\\
+T_4 &: r_x(1)
+\end{align*}
+\caption{Example of \textit{N-MR} violation when $w_x(1) \ll w_x(2)$ and $T_4$ directly session-depends on $T_3$.}
+\label{fig:nmr-history}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\begin{tikzpicture}[node distance=3cm]
+\tikzstyle{tx}=[draw=none, fill=none]
+\node[tx] (T1) at (0,0) {$T_1$};
+\node[tx] (T2) at (2,0) {$T_2$};
+
+\node[tx] (T3) at (0,-1) {$T_3$};
+\node[tx] (T4) at (2,-1) {$T_4$};
+
+\draw[->] (T1) edge node[above]{$ww_x$} (T2);
+\draw[->] (T3) edge node[below]{$s_i$} (T4);
+
+\draw[->] (T2) edge node[sloped, above]{$wr_x$} (T3);
+%\draw[->] (T2) edge [bend right] node[left]{$wr_x$} (T4);
+%\draw[->] (T1) edge [bend right]node[right]{$wr_x$} (T4);
+\draw[dashed, ->] (T4) edge [bend right]node[right]{$rw_x$} (T2);
+\end{tikzpicture}
+\caption{DSG for Figure~\ref{fig:nmr-history}. $wr_x$ dependency from $T_1$ to $T_4$ omitted.} 
+\label{fig:nmr-dsg}
+\end{figure}
 

### 2013-04-25 writeup/hat-vldb.tex
another pass

@@
 
-\usepackage{times, graphicx, url, xcolor, multicol, minipage, subfig, tikz, amsthm}
+\usepackage{times, graphicx, url, xcolor, multicol, minipage, subfig, tikz, amsthm, float}
 \usepackage{balance}  % for  \balance command ON LAST PAGE  (only there!)
@@
 \usetikzlibrary{positioning, snakes, shapes}
+
+\tikzset{>=latex}
 
@@
 items. In this work, we consider the problem of providing Highly
-Available Transactions (HATs), or transactional guarantees that do not
+Available Transactions (HATs): transactional guarantees that do not
 suffer unavailability during system partitions or incur high network
@@
 identify which can and cannot be achieved in HAT systems. We
-analytically and experimentally quantify the performance and
-availability benefits of HATs as well as their necessary semantic
-compromises.
+analytically and experimentally quantify the availability and
+performance benefits of HATs---often several orders of magnitude over
+wide-area networks---as well as their necessary semantic compromises.
 \end{abstract}

### 2013-04-25 thebes-code/common/src/main/java/edu/berkeley/thebes/common/persistence/memory/MemoryPersistenceEngine.java
DepRes check for newer

@@ -57,7 +57,7 @@ public class MemoryPersistenceEngine implements IPersistenceEngine {
                 // If we already have this key, ensure new item is a more recent version
                 if (map.containsKey(key)) {
                     DataItem curItem = map.get(key);
-                    if (curItem.getVersion().compareTo(value.getVersion()) <= 0) {
+                    if (curItem.getVersion().compareTo(value.getVersion()) > 0) {
                         return;
                     }
                 }

### 2013-04-25 thebes-code/hat/server/src/main/java/edu/berkeley/thebes/hat/server/dependencies/DependencyResolver.java

@@ -87,11 +87,16 @@ public class DependencyResolver {
                                              "assertion-violation",
                                              "violations",
                                              TimeUnit.SECONDS);
+    Meter obsoletePuts = Metrics.newMeter(DependencyResolver.class,
+                                          "obsolete-puts",
+                                          "puts",
+                                          TimeUnit.SECONDS);
 
     private final Timer addPendingTimer = Metrics.newTimer(DependencyResolver.class, "add-pending-latency");
     
 
     private final Timer stage1Timer = Metrics.newTimer(DependencyResolver.class, "stage-1");
+    private final Timer stage15Timer = Metrics.newTimer(DependencyResolver.class, "stage-1.5");
     private final Timer stage2Timer = Metrics.newTimer(DependencyResolver.class, "stage-2");
     private final Timer stage3Timer = Metrics.newTimer(DependencyResolver.class, "stage-3");
     private final Timer stage4Timer = Metrics.newTimer(DependencyResolver.class, "stage-4");
@@ -161,14 +166,26 @@ public class DependencyResolver {
         TimerContext context = addPendingTimer.time();
         try {
             Version version = value.getVersion();
+            boolean obsoletedWrite = false;
             
             {
                 TimerContext s1 = stage1Timer.time();
                 pendingTransactionsMap.putIfAbsent(version, new TransactionQueue(version));
                 s1.stop();
             }
+            
             {
+                TimerContext s15 = stage15Timer.time();
+                if (persistenceEngine.get(key).getVersion().compareTo(version) > 0) {
+                    obsoletePuts.count();
+                    obsoletedWrite = true;
+                }
+                s15.stop();
+            }
+            
+            if (!obsoletedWrite) {
                 TimerContext s2 = stage2Timer.time();
+                
                 persistPendingWrite(key, value);
                 s2.stop();
             }
@@ -176,7 +193,7 @@ public class DependencyResolver {
             PendingWrite newPendingWrite;
             {
                 TimerContext s3 = stage3Timer.time();
-                newPendingWrite = new PendingWrite(key, value);
+                newPendingWrite = new PendingWrite(key, value, obsoletedWrite);
                 s3.stop();
             }
 
@@ -193,7 +210,6 @@ public class DependencyResolver {
                 s4.stop();
             }
     
-
             {
                 TimerContext s5 = stage5Timer.time();
                 try {
[…]

### 2013-04-25 thebes-code/hat/server/src/main/java/edu/berkeley/thebes/hat/server/dependencies/PendingWrite.java

@@ -20,8 +20,9 @@ public class PendingWrite implements Comparable<PendingWrite> {
     private int numKeysForThisReplica;
     private Set<Integer> replicaIndicesInvolved;
     private List<String> transactionKeys;
+    private boolean obsoleted;
     
-    public PendingWrite(String key, DataItem value) {
+    public PendingWrite(String key, DataItem value, boolean obsoleted) {
         if (value.getTransactionKeys().isEmpty()) {
             throw new IllegalStateException("Pending writes must be waiting on at least one other key.");
         }
@@ -29,6 +30,7 @@ public class PendingWrite implements Comparable<PendingWrite> {
         this.key = key;
         this.transactionVersion = value.getVersion();
         this.transactionKeys = value.getTransactionKeys();
+        this.obsoleted = obsoleted;
 
         examineReplicasInvolved(key, value.getTransactionKeys());
     }
@@ -66,6 +68,10 @@ public class PendingWrite implements Comparable<PendingWrite> {
 
     public Version getVersion() {
         return transactionVersion;
+    }
+    
+    public boolean isObsoleted() {
+        return obsoleted;
     }
 
     @Override

### 2013-04-25 writeup/evaluation.tex

@@
 general-purpose read-write transactions in a distributed
-setting~\cite{bernstein-book} can require at least one round trip time
+setting~\cite{bernstein-book} requires at least one round trip time
 (RTT) before committing. As an example, traditional two-phase locking
-for a transaction of length $T$ can require $T$ \texttt{lock}
+for a transaction of length $T$ may require $T$ \texttt{lock}
 operations and will require at least one \texttt{lock} and
@@
 requires a scheduler. Serializability under multi-version concurrency
-control also requires checking for update conflicts. The reliance on a
-globally agreed total order necessitates a minimum of one round-trip
-to a designated master or coordination service for each of these
-classic algorithms.  The cost of round trips will be determined by the
-deployment environment, as we saw in Section~\ref{sec:motivation}; we
-will demonstrate this cost on public cloud infrastructure in the next
-section.
+control requires checking for update conflicts. All told, the reliance
+on a globally agreed total order necessitates a minimum of one
+round-trip to a designated master or coordination service for each of
+these classic algorithms.  The cost of round trips will be determined
+by the deployment environment, as we saw in
+Section~\ref{sec:motivation}; we will demonstrate this cost on public
+cloud infrastructure in Section~\ref{sec:prototype}.
 
-\vspace{.5em}\noindent\textbf{Non-serializability} Many existing
+\vspace{.5em}\noindent\textbf{Non-serializability} Most existing
 distributed implementations of weak isolation are not highly
@@
 recency guarantees.) While multi-versioned storage systems allow for a
-variety of transactional guarantees, we have not seen proposals for
-truly weak isolation (e.g., non-``tentative update'' schemes) in this
-context.  The MDCC~\cite{mdcc} protocol offers Read Committed
-isolation with Lost Update avoidance but is similarly unavailable due
-to its reliance on preventing write conflicts. Chan and Gray's
-read-only transactions provide read-only transactions with item-cut
-isolation, causal consistency, and transactional atomicity (session
-PL-2L~\cite{adya}) but are unavailable in the presence of coordinator
+variety of transactional guarantees, few offer traditional weak
+isolation (e.g., non-``tentative update'' schemes) in this context.
+The MDCC~\cite{mdcc} protocol offers Read Committed isolation with
+Lost Update avoidance but is similarly unavailable due to its reliance
+on preventing write conflicts. Chan and Gray's read-only transactions
+provide read-only transactions with item-cut isolation, causal
+consistency, and transactional atomicity (session PL-2L~\cite{adya})
+but are unavailable in the presence of coordinator
 failure~\cite{readonly}, similar to read-only and write-only
-transactions more recently proposed by Eiger~\cite{eiger}; Causal
+transactions more recently proposed by Eiger~\cite{eiger}. Causal
 Serializability offers a similar model (with unavailable
@@
 PL-2L with high availability, but none provide this HAT functionality
-without substantial modification; Swift~\cite{swift} and bolt-on
-causal consistency~\cite{bolton} are closest to providing full HAT
-semantics operation in a sticky context. As seen, it is possible to
-implement many guarantees weaker than serializability---including
-guarantees that are achievable with high availability---and still not
-achieve high availability.
+without substantial modification. Swift~\cite{swift} and bolt-on
+causal consistency~\cite{bolton} are closest to providing maximum
+sticky HAT semantics. As we have seen, it is possible to implement many
+guarantees weaker than serializability---including guarantees that are
+achievable with high availability---and still not achieve high
+availability.
 
@@
 
-Thus far, we have largely ignored the question of which applications
-require semantics that are unavailable to HATs. As we showed in
-Section~\ref{sec:hats}, the main cost of choosing HATs comes in the
-inability to prevent Lost Update, Write Skew, and provide recency
-bounds. In this section, we attempt to understand when these
-guarantees matter both abstractly and in a representative
+Thus far, we have largely ignored the question of when HAT semantics
+are useful (or otherwise are too weak). As we showed in
+Section~\ref{sec:hats}, the main cost of high availability and low
+latency comes in the inability to prevent Lost Update, Write Skew, and
+provide recency bounds. In this section, we attempt to understand when
+these guarantees matter both abstractly and in a representative
 transactional application based on the TPC-C benchmark~\cite{tpcc}.
@@
 will not, in general, be able to maintain application-level
-consistency constraints with HATs alone--in particular, applications
+consistency constraints with HATs alone---in particular, applications
 requiring bounded update visibility latency should opt for
@@
 of HAT-compliance in an application context, we consider a concrete
-application: the TPC-C benchmark~\cite{tpcc}. Surprisingly, four of
-five transactions can be executed with HATs, while the fifth may
-require unavailability.
+application: the TPC-C benchmark. In brief, we find that four of five
+transactions can be executed with HATs, while the fifth may require
+unavailability.
 
@@
 transactions---\textit{Order-Status} and \textit{Stock-Level}---are
-read-only and can be executed safely under HATs. Clients may read
-stale data, but this does not violate TPC-C requirements and clients
-will read their writes if they are sticky-available. Another
-transaction type, \textit{Payment}, updates running balances for
-warehouses, districts, and customer records as well as providing an
-audit trail. The transaction is increment- and append-only, so all
-balance increase operations commute, and TA allows the
-maintenance of any foreign-key constraints (e.g.,
-\texttt{UPDATE/DELETE CASCADE}).
+read-only and can be executed safely with HATs. Clients may read stale
+data, but this does not violate TPC-C requirements and clients will
+read their writes if they are sticky-available. Another transaction
+type, \textit{Payment}, updates running balances for warehouses,
+districts, and customer records and provides an audit trail. The
+transaction is increment- and append-only, so all balance increase
+operations commute, and TA allows the maintenance of foreign-key
+integrity constraints (e.g., via \texttt{UPDATE/DELETE CASCADE}).
 
@@
 
-New-Order presents two challenges: ID assignment and stock
-maintenance. First, each New-Order transaction requires a unique ID
-number for the order. We can create a unique number by, say,
+The New-Order transaction presents two challenges: ID assignment and
+stock maintenance. First, each New-Order transaction requires a unique
+ID number for the order. We can create a unique number by, say,
 concatenating the client ID and a timestamp. However, the TPC-C
@@
 Update. Accordingly, HATs cannot provide compliant TPC-C execution but
-can stil maintain uniqueness constraints. Second, the New-Order transaction
-decrements inventory counts: what if the count becomes negative?
-Fortunately, TPC-C New-Order restocks each item's inventory count
-(increments by 91) if it would become negative as the result of
+can stil maintain uniqueness constraints. Second, the New-Order
+transaction decrements inventory counts: what if the count becomes
+negative?  Fortunately, TPC-C New-Order restocks each item's inventory
+count (increments by 91) if it would become negative as the result of
 placing an order. This means that, even in the presence of concurrent
 New-Orders, an item's stock will never fall below zero. This is TPC-C
-compliant, but a HAT system might end up with more stock
-than in a non-HAT-compliant implementation.
+compliant, but a HAT system might end up with more stock than in a
+non-HAT-compliant implementation.
 
-Delivery is challenging due to non-monotonicity. Each Delivery deletes
-a pending order from the New-Order table and should be idempotent in
-order to avoid billing a customer twice; this implies a need to
-prevent Lost Update. This issue  can be avoided  by moving the
+The Delivery transaction is challenging due to non-monotonicity. Each
+Delivery deletes a pending order from the New-Order table and should
+be idempotent in order to avoid billing a customer twice; this implies
+a need to prevent Lost Update. This issue can be avoided by moving the
 non-monotonicity to the real world---the carrier that picks up the
@@
 HATs. The two problematic transactions---New-Order and Payment---rely
-on non-monotonic state update. The former transaction can be modified
-to ensure ID uniqueness at the expense of sequential per-district ID
-ordering, while the latter is an inherently non-monotonic action
-requiring external compensation or stronger consistency protocols. We
-expect that, especially for read-dominated workloads found in many
-online services, HAT guarantees will suffice for many application
-transactions.
+on non-monotonic state update. The former can be modified to ensure ID
+uniqueness but not sequential ID ordering, while the latter is an
+inherently non-monotonic action requiring external compensation or
+stronger consistency protocols. Based on these experiences and
+discussions with practitioners, we expect that, especially for
+read-dominated workloads found in many online services, HAT guarantees
+will provide useful semantics for many applications.
 
 \subsection{Experimental Costs}
+\label{sec:prototype}
 

### 2013-04-25 writeup/futurework.tex
fixed
final pageshorten
editing pass

@@
 
-\textbf{What are the limits of achievable HAT models, and how are they
-  defined?}  In this work, we have taxonomized several well-defined
-and well-documented semantic properties. This is useful as it
-contextualizes decades of existing database and distributed systems
-design and algorithms but does not provide guidance as to what
+\textbf{What are the limits of HAT semantics?}  In this work, we have
+taxonomized several well-defined and well-documented semantic
+properties. This is useful as it contextualizes existing database and
+distributed systems semantics but does not provide guidance as to what
 consistency models are the ``strongest''---or admit the smallest set
@@
 respect causality, and causal consistency returns values other than
-$1$. Always returning the value $2$ is similarly incomparable, and, as
-an example, we can enumerate models that return all integers and real
-numbers--an uncountably infinite set of models. Given these infinite
-models, are there meaningful properties that delineate the boundaries
-between classes of highly available systems? Is there a
-``meta-property'' that exactly characterizes what properties are
-achievable with high availability or sticky high availability? This is
-an area of ongoing research within the Berkeley HAT project.
+$1$. There are actually infintely many incomparable models (vacuously:
+return $1$, return $2$, and so on). Are there \textit{meaningful}
+properties that delineate the boundaries between availability levels?
+Is there a ``meta-property'' that exactly characterizes what
+properties are achievable with high availability or sticky high
+availability? This is an area of ongoing research within the Berkeley
+HAT project. \pbnote{SHORTEN}
 
@@
 guarantees to a truly highly available system seems expensive for some
-models (e.g., metadata requirements or visibility requirements may
+models (e.g., metadata requirements or visibility penalties may
 become problematic). Further experimentation regarding the set of
@@
 \textbf{Where do semantics-based concurrency control and weakened
-  failure models lie in the HAT landscape?} The vast
-literature on semantics-based concurrency control indicates 
-when coordination can be avoided while providing ``strongly
-consistent'' outcomes for limited failure scenarios. We have performed
-a narrow case study, but a general taxonomization of
-techniques such as escrow transactions~\cite{escrow},
-Sagas~\cite{sagas}, and other long-running transaction support could
-help answer the challenge of programming in a weakly consistent
-environment. HAT guarantees can also be strengthened by considering
-weaker failure models such as bounded asynchrony or models in which
-the underlying hardware provides explicit service-level
-agreements. Additionally, while we have focused on systems that
-provide a single availability mode, there are several interesting
-hybrids to consider. For example, a system may provide ``best-effort''
-stickiness (or serializability) in the absence of partitions, but, in
-the presence of partitions, ``lose'' guarantees by having clients
-reconnect to an available replica. This strategy provides stronger
-guarantees in the absence of partitions but, to provide availability,
-sacrifices them in the event of partitions.
+  failure models lie in the HAT landscape?} The vast literature on
+semantics-based concurrency control indicates when coordination can be
+avoided while providing ``consistent'' operation for limited failure
+scenarios. We have performed a narrow case study, but a general
+taxonomization of techniques such as escrow
+transactions~\cite{escrow}, Sagas~\cite{sagas}, and other long-running
+transaction support could help address the challenge of programming in
+a weakly consistent environment. HAT guarantees can also be
+strengthened by considering weaker failure models such as networks
+with bounded asynchrony. Additionally, while we have focused on
+systems that provide a single availability mode, there are several
+interesting hybrids to consider. For example, a system may provide
+strong semantics in the absence of partitions, but, in the presence of
+partitions, ``lose'' guarantees by having clients reconnect to an
+available replica. This provides ``best effort'' guarantees but, to
+provide availability, sacrifices them in the event of partitions.

### 2013-04-25 writeup/hat-appendix.tex

@@
 
+\begin{figure}[H]
+\begin{align*}
+\small
+T_1 &: w_x(1)\\
+T_2 &: w_y(1)\\
+T_3 &: r_y(1)~r_x(0)
+\end{align*}
+\caption{Example of \textit{N-MW} anomaly if $T_2$ directly session-depends on $T_1$.}
+\label{fig:nmw-history}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\begin{tikzpicture}[node distance=3cm]
+\tikzstyle{tx}=[draw=none, fill=none]
+\node[tx] (T1) at (0,0) {$T_1$};
+\node[tx] (T2) at (2,0) {$T_2$};
+\node[tx] (T3) at (2,-1) {$T_3$};
+
+\draw[->] (T1) edge node[above]{$s_i$} (T2);
+\draw[->] (T2) edge node[right]{$wr_y$} (T3);
+\draw[->, dashed] (T3) edge node[below, sloped]{$rw_x$} (T1);
+\end{tikzpicture}
+\caption{DSG for Figure~\ref{fig:nmw-history}.}
+\label{fig:nmw-dsg}
+\end{figure}
+
 \begin{definition}[Monotonic Writes (MW)]
@@
 \end{definition}
+
+\begin{figure}[H]
+\begin{align*}
+\small
+T_1 &: w_x(1)\\
+T_2 &: r_x(1) w_y(1)\\
+T_3 &: r_y(1)~r_x(0)
+\end{align*}
+\caption{Example of \textit{N-WFR} anomaly.}
+\label{fig:nwfr-history}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\begin{tikzpicture}[node distance=3cm]
+\tikzstyle{tx}=[draw=none, fill=none]
+\node[tx] (T1) at (0,0) {$T_1$};
+\node[tx] (T2) at (2,0) {$T_2$};
+\node[tx] (T3) at (2,-1) {$T_3$};
+
+\draw[->] (T1) edge node[above]{$wr_x$} (T2);
+\draw[->] (T2) edge node[right]{$wr_y$} (T3);
+\draw[->, dashed] (T3) edge node[below, sloped]{$rw_x$} (T1);
+\end{tikzpicture}
+\caption{DSG for Figure~\ref{fig:nwfr-history}.}
+\label{fig:nwfr-dsg}
+\end{figure}
 
@@
 \end{definition}
+
+\begin{figure}[H]
+\begin{align*}
+\small
+T_1 &: w_x(1)\\
+T_2 &: r_x(0)\\
+\end{align*}
+\caption{Example of \textit{N-RYW} anomaly if $T_2$ directly session-depends on $T_1$.}
+\label{fig:nryw-history}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\begin{tikzpicture}[node distance=3cm]
+\tikzstyle{tx}=[draw=none, fill=none]
+\node[tx] (T1) at (0,0) {$T_1$};
+\node[tx] (T2) at (2,0) {$T_2$};
+
+\draw[->] (T1) edge node[above]{$s_i$} (T2);
+\draw[dashed, ->] (T2) edge [bend left] node[below]{$rw_x$} (T1);
+\end{tikzpicture}
+\caption{DSG for Figure~\ref{fig:nwfr-history}.}
+\label{fig:nryw-dsg}
+\end{figure}
 

### 2013-04-25 writeup/model.tex
another pass

@@
 including ``true'' high availability, availability with stickiness,
-and several other useful variants.
+and transactional availability.
 
@@
 others. If servers are partitioned from one another, they do not need
-to stall in order to provide a ``safe'' query response. This lack of
-fast-path coordination also means that a highly available system also
-provides low latency~\cite{abadi-pacelc}; in a wide-area setting, this
-means that clients need not wait for cross-datacenter
-communication. To properly describe whether a \textit{transactional}
-system is highly available, we need to describe what servers a client
-must contact as well as what kinds of responses a server can provide,
-especially given the possibility of aborts.
+to stall in order to provide clients a ``safe'' response to
+operations. This lack of fast-path coordination also means that a
+highly available system also provides low latency~\cite{abadi-pacelc};
+in a wide-area setting, this means that clients need not wait for
+cross-datacenter communication. To properly describe whether a
+\textit{transactional} system is highly available, we need to describe
+what servers a client must contact as well as what kinds of responses
+a server can provide, especially given the possibility of aborts.
 
@@
   online is not available.} As in a standard distributed database,
-designated servers might handle operations on different data
-items;\footnote{There is a further distinction between a \textit{fully
-    replicated} system, in which all servers are replicas for all data
-  items, and a \textit{partially replicated} system, in which at least
-  one server acts as a replica for only a subset of all data
-  items. For generality, and, for applicability to many modern
-  ``sharded'' or ``partitioned'' data storage systems~\cite{dynamo,
-    pnuts, bigtable, spanner, hstore}, we consider partially
-  replicated systems.} a server that can handle an operation for a
-given data item is called a \textit{replica} for that item. While this
-standard model is useful, there are several other cases we should
-consider.
+designated servers might perform operations for different data items;
+a server that can handle an operation for a given data item is called
+a \textit{replica} for that item.\footnote{There is a further
+  distinction between a \textit{fully replicated} system, in which all
+  servers are replicas for all data items, and a \textit{partially
+    replicated} system, in which at least one server acts as a replica
+  for only a subset of all data items. For generality, and, for
+  applicability to many modern ``sharded'' or ``partitioned'' data
+  storage systems~\cite{dynamo, pnuts, bigtable, spanner, hstore}, we
+  consider partially replicated systems.} While this standard model is
+useful, there are several other cases we should consider.
 
@@
 
-We have so far focused on single-object, single-operation
+Until now, we have considered single-object, single-operation
 availability. This is standard in the distributed systems literature
-(e.g., distributed register models such as linearizability all
-concern single objects~\cite{herlihy-art}), yet the database
-literature largely focuses on transactions: groups of multiple
-operations over multiple objects. Accordingly, by itself, replica
-availability is insufficient to describe availability guarantees for
+(e.g., distributed register models such as linearizability all concern
+single objects~\cite{herlihy-art}), yet the database literature
+largely focuses on transactions: groups of multiple operations over
+multiple objects. Accordingly, by itself, replica availability is
+insufficient to describe availability guarantees for
 transactions. Additionally, given the choice of \textit{commit} and
 \textit{abort} responses---which signal transaction success or failure
-to a client---we must take care in defining transactional availability.
+to a client---we must take care in defining transactional
+availability.
 

### 2013-04-25 writeup/motivation.tex

@@
-\section{Why Availability?}
+\section{Why High Availability?}
 \label{sec:motivation}
@@
 and routing. As we will see (Section~\ref{sec:availability}), highly
-available systems minimize the effect of these network
+available system operation minimizes the effect of network
 \textit{partitions} and \textit{latency}. In this Section, we draw on
@@
 frequently in real-world deployments and best-case latencies between
-datacenters are significant.
+datacenters are substantial, often on the order of hundreds of
+milliseconds.
 
@@
 Hamilton's assertion. In April 2011, a network misconfiguration led to
-a twelve-hour long series of outages across the Amazon EC2 and RDS
+a twelve-hour series of outages across the Amazon EC2 and RDS
 services~\cite{amazon-netpartition}. Subsequent misconfigurations and
@@
 North America and Europe due a router
-bug~\cite{juniper-partition}---and misconfigurations such as BGP
+bug~\cite{juniper-partition}---and misconfigurations like the BGP
 faults in 2008~\cite{pakistan-youtube} and
@@
 large-scale operators like Google~\cite{dean-keynote} confirm that
-partition management is an important consideration for service operators
-today. As an example of an actual system rearchitecture resulting from
-partitions, less than one year after its announcement, Yahoo!'s PNUTS
-developers explicitly added support for weaker, highly available
-operation. The engineers explained that ``strict adherence [to strong
-  consistency] leads to difficult situations under network
-partitioning or server failures...in many circumstances, applications
-need a relaxed approach''~\cite{pnuts-update}.
+partition management is an important consideration for service
+operators today. System designs that do not account for partition
+behavior may prove difficult to operate at scale: for example, less
+than one year after its announcement, Yahoo!'s PNUTS developers
+explicitly added support for weaker, highly available operation. The
+engineers explained that ``strict adherence [to strong consistency]
+leads to difficult situations under network partitioning or server
+failures...in many circumstances, applications need a relaxed
+approach''~\cite{pnuts-update}.
 
@@
 routing failures on the Internet~\cite{labovitz-failures}. Isolating,
-quantifying, and designing for these network failures is an area of
-active research in networking
-community~\cite{uw-failure-networks}.
+quantifying, and accounting for these network failures is an area of
+active research in networking community~\cite{uw-failure-networks}.
 
-These results---which do not even consider server-level
-failures---indicate that network partitions \textit{do} occur within
-and across modern datacenters. We do not intend to fixate on any
-particular statistical behavior but instead observe that, as a general
-trend, partitions exist and must accordingly be met with either
-unavailability at some servers or, as we will discuss, relaxed
-semantic guarantees.
+These results---which do not even consider server failures---indicate
+that network partitions \textit{do} occur within and across modern
+datacenters. We observe that, as a general trend, partitions exist and
+must accordingly be met with either unavailability at some servers or,
+as we will discuss, relaxed semantic guarantees.
 
@@
 section, we quantify round-trip latencies, which are often
-large---hundreds of milliseconds---in a geo-replicated context.
-Fundamentally, the speed at which two servers can communicate is
-(according to modern physics) bounded by the speed of light. In the
-best case, two servers on opposite sides of the Earth's
-surface---communicating via a hypothetical link traversing the
-planet's diameter---require a minimum 85.1 ms RTT (133.7 ms if sent at
-ground level). As services are replicated to multiple, geographically
-distinct sites, the cost of communication increases to this lower
-bound.
+large---hundreds of milliseconds---in a geo-replicated,
+multi-datacenter context.  Fundamentally, the speed at which two
+servers can communicate is (according to modern physics) bounded by
+the speed of light. In the best case, two servers on opposite sides of
+the Earth---communicating via a hypothetical link traversing the
+planet's diameter---require a minimum 85.1ms RTT (133.7ms if sent at
+surface level). As services are replicated to multiple, geographically
+distinct sites, this cost of communication increases.
 
@@
  }
-\subfloat[Within \texttt{us-east-a} AZ] {
+\subfloat[Within \texttt{us-east-b} AZ] {
   \makebox[.25\textwidth]{
@@
 
-\caption{Mean RTT times on EC2 (min and max highlighted)}
+\caption{Mean RTT times on EC2 (min and max highlighted)}\vspace{-1em}
 \label{table:rtt}
@@
 \begin{figure}
-\includegraphics[width=\columnwidth]{graphs/ping-plot.pdf}
+\includegraphics[width=\columnwidth]{graphs/ping-plot.pdf}\vspace{-1em}
 \caption{CDF of round-trip times for slowest inter- and intra-
-  availability zone links compared to cross-region links.}
+  availability zone links compared to cross-region links.}\vspace{-1em}
 \label{fig:rtt}
@@
 
-In real deployments, transit times are actually slower than the speed
-of light due to routing, congestion, and processing times. To
-illustrate the difference between intra-datacenter, inter-datacenter,
-and inter-planetary networks, we performed a measurement study of
-network behavior on Amazon's EC2, a widely-used public compute
-cloud. We measured one week of ping times between all seven EC2
-geographic ``regions'', across three ``availability zones'' (closely
-co-located datacenters), and within a single ``availabilty zone''
-(datacenter), at a granularity of 1s (dataset to be released). We
-summarize the results of our network measurement study in
-Table~\ref{table:rtt}. On average, intra-datacenter communication (1a)
-is between 1.82 and 6.38 times faster than across geographically
-co-located datacenters (1b) and between 40 and 647 times faster than
-across geographically distributed datacenters (1c). The cost of
-wide-area communication exceeds the speed of light: for example,
-communicating from S\~{a}o Paulo to Singapore is lower-bounded by
-speed-of-light transit RTT of 106.7ms but, on average, incurs a
-362.8ms RTT (95th percentile: 649ms). As shown in
-Figure~\ref{fig:rtt}, the distribution of latencies varies between
-links, but the trend is clear: remote communication has a substantial
-cost.
+In real deployments, messages travel slower than the speed of light
+due to routing, congestion, and server-side overheads. To illustrate
+the difference between intra-datacenter, inter-datacenter, and
+inter-planetary networks, we performed a measurement study of network
+behavior on Amazon's EC2, a widely-used public compute cloud. We
+measured one week of ping times between all seven EC2 geographic
+``regions'', across three ``availability zones'' (closely co-located
+datacenters), and within a single ``availabilty zone'' (datacenter),
+at a granularity of 1s (dataset to be released). We summarize the
+results of our network measurement study in Table~\ref{table:rtt}. On
+average, intra-datacenter communication (Table 1a) is between 1.82 and
+6.38 times faster than across geographically co-located datacenters
+(Table 1b) and between 40 and 647 times faster than across
+geographically distributed datacenters (Table 1c). The cost of
+wide-area communication exceeds the speed of light: for example, while
+a speed-of-light RTT from S\~{a}o Paulo to Singapore RTT is 106.7ms,
+ICMP packets incur an average 362.8ms RTT (95th percentile: 649ms). As
+shown in Figure~\ref{fig:rtt}, the distribution of latencies varies
+between links, but the trend is clear: remote communication has a
+substantial cost.
 
@@
 serializability is not achievable in a highly available
-system~\cite{davidson-survey}, meaning it will be expensive in the
-environments from Section~\ref{sec:motivation}. However, even in a
+system~\cite{davidson-survey}, meaning that, in environments like
+those in Section~\ref{sec:motivation}, database designs face a choice
+between availability and strong semantics. However, even in a
 single-node database, the coordination penalties associated with
@@
 scalability limitations, and, often, external aborts like
-deadlocks)~\cite{gray-isolation}. Accordingly, to increase concurrency,
-database systems offer a range of ACID properties weaker than
-serializability.  Instead, databases offer a host of so-called
+deadlocks)~\cite{gray-isolation}. Accordingly, to increase
+concurrency, database systems offer a range of ACID properties weaker
+than serializability.  Instead, databases offer a host of so-called
 \textit{weak isolation} models that allow varying restrictions on the
@@
   ansicritique, ansi-sql}. None of these weak isolation models
-guarantees serializability, but, as we see below, the benefits of
-these models are often considered to outweigh costs of possible
-consistency anomalies that might arise from their use.
+guarantees serializability, but, as we see below, their benefits are
+often considered to outweigh costs of possible consistency anomalies
+that might arise from their use.
 
@@
 
-It is unknown \textit{which} of these ACID (particularly isolation)
-guarantees can be provided with high availability. The existing
-algorithms for providing weak isolation are typically designed for a
-single-node context and are, to the best of our knowledge, unavailable
-due to a reliance on concurrency control mechanisms like locking that
-are not resilient to partial failure
-(Section~\ref{sec:evaluation}). Moreover, we are not aware of any prior
-literature that provides guidance as to the relationship between weak
-isolation and high availability: prior work has examined the
-relationship between serializability and high
+It is unknown \textit{which} of these guarantees can be provided with
+high availability. The existing algorithms for providing weak
+isolation are typically designed for a single-node context and are, to
+the best of our knowledge, unavailable due to a reliance on
+concurrency control mechanisms like locking that are not resilient to
+partial failure (Section~\ref{sec:evaluation}). Moreover, we are not
+aware of any prior literature that provides guidance as to the
+relationship between weak isolation and high availability: prior work
+has examined the relationship between serializability and high
 availability~\cite{davidson-survey} and weak isolation on a single

### 2013-04-25 writeup/relatedwork.tex

@@
 \section{Related Work}
+
+There is a growing body related work on highly available semantics,
+mechanisms for concurrency control, and techniques for scalable
+distributed operations.
 
@@
 unachievable~\cite{davidson-survey} and Brewer's CAP Theorem has
-attracted considerable attention~\cite{gilbert-cap}. Recent work has
-begun to offer hints regarding the connection between ``weak
+attracted considerable attention~\cite{gilbert-cap}. Recent work on
+PACELC expands CAP by considering connections between ``weak
 consistency'' and low latency~\cite{abadi-pacelc}, while several
@@
   ansicritique}. However, as we have discussed, the connection between
-weak consistency models in a distributed setting~\cite{calm, crdt} and
-high availability is relatively unexplored. With the exception of a
-preliminary workshop paper discussing highly available RC and
-I-CI~\cite{hat-hotos}--- which this work expands, providing additional
-algorithms and impossibility results, a full taxonomization, and
-experimental and application analysis---we believe this paper is the
-first to explore this area. Nonetheless, several studies rigorously
-classify non-HAT semantics: Wisemann et al. have proposed a three
-parameter classification for one-copy serializable database systems,
-performing a more extensive classification than our overview in
-Section~\ref{sec:evaluation}, including eager database replication
-techniques~\cite{kemme-classification}. Additionally several earlier
-surveys discuss distributed mechanisms for achieving one-copy
-serializability~\cite{wisemann-survey}, linearizability, consistent
-snapshots, and recency bounds~\cite{ceri-mechanism, chen-mechanism}.
+weak, multi-object consistency models in a distributed setting and
+high availability is relatively unexplored. With the exception of our
+preliminary workshop paper discussing transactional availability and
+HAT RC and I-CI~\cite{hat-hotos}--- which this work expands, providing
+additional algorithms and impossibility results, a full
+taxonomization, and experimental and application analysis---we believe
+this paper is the first to explore this area. However, several studies
+rigorously classify \textit{non-HAT} semantics: Wisemann et al. have
+proposed a three parameter classification for one-copy serializable
+database systems, performing a more extensive classification than our
+overview in Section~\ref{sec:evaluation}, including eager database
+replication techniques~\cite{kemme-classification}. Additionally
+several earlier surveys consider distributed mechanisms for achieving
+one-copy serializability~\cite{wisemann-survey}, linearizability,
+consistent snapshots, and recency bounds~\cite{ceri-mechanism,
+  chen-mechanism}.
 

### 2013-04-28 writeup/conclusion.tex
some graphs and joe comments

@@
 Available Transactions (HATs), which can offer the key benefits of
-highly available distributed systems---''always on'' operation during
+highly available distributed systems---``always on'' operation during
 partitions and low-latency operations---while also providing a family
@@
 while a few, like preventing Lost Update and Write Skew, are not. This
-offers a long-overdue connection between storage semantics studied in
-the database and distributed systems communities. The subsequent
-challenge is to return to issues of performance and systems
-architecture within the HAT design space.  Based on our understanding
-of what is desired in the field and newfound knowledge of what is
-possible to achieve, we believe HATs will be an area of great
-opportunity in the coming years.
+offers a useful connection between storage semantics studied in the
+database and distributed systems communities. The next challenge is to
+return to issues of performance and systems architecture within the
+HAT design space.  Based on our understanding of what is desired in
+the field and newfound knowledge of what is possible to achieve, we
+believe HATs provide a large and useful design space to pursue.
 

### 2013-04-28 writeup/evaluation.tex
some graphs and joe comments

@@
 general-purpose read-write transactions in a distributed
-setting~\cite{bernstein-book} requires at least one round trip time
+setting~\cite{bernstein-book} require at least one round trip time
 (RTT) before committing. As an example, traditional two-phase locking
 for a transaction of length $T$ may require $T$ \texttt{lock}
-operations and will require at least one \texttt{lock} and
+operations and will require at least one \texttt{lock} and one
 \texttt{unlock} operation.  In a distributed environment, each of
@@
 districts, and customer records and provides an audit trail. The
-transaction is increment- and append-only, so all balance increase
-operations commute, and TA allows the maintenance of foreign-key
-integrity constraints (e.g., via \texttt{UPDATE/DELETE CASCADE}).
+transaction is monotonic---increment- and append-only---so all balance
+increase operations commute, and TA allows the maintenance of
+foreign-key integrity constraints (e.g., via \texttt{UPDATE/DELETE
+  CASCADE}).
 
-While three out of five transactions are easily achievable with
-HATs, the remaining two transactions---\textit{New-Order} and
-\textit{Delivery}---are not as simple. The New-Order transaction
-places an order for a variable quantity of data items, updating
-warehouse stock as needed. It selects a sales district, assigns the
-order an ID number, adjusts the remaining warehouse stock, and writes
-a placeholder entry for the pending order. The Delivery transaction
+\vspace{.5em}\noindent\textit{New-Order and Delivery.} While three out of
+five transactions are easily achievable with HATs, the remaining two
+transactions are not as simple. The New-Order transaction places an
+order for a variable quantity of data items, updating warehouse stock
+as needed. It selects a sales district, assigns the order an ID
+number, adjusts the remaining warehouse stock, and writes a
+placeholder entry for the pending order. The Delivery transaction
 represents the fulfillment of a New-Order: it deletes the order from
@@
 
-The New-Order transaction presents two challenges: ID assignment and
+\vspace{.5em}\noindent\textit{IDs and decrements.} The New-Order transaction presents two challenges: ID assignment and
 stock maintenance. First, each New-Order transaction requires a unique
@@
 Update. Accordingly, HATs cannot provide compliant TPC-C execution but
-can stil maintain uniqueness constraints. Second, the New-Order
+can still maintain uniqueness constraints. Second, the New-Order
 transaction decrements inventory counts: what if the count becomes
@@
 
-The Delivery transaction is challenging due to non-monotonicity. Each
-Delivery deletes a pending order from the New-Order table and should
-be idempotent in order to avoid billing a customer twice; this implies
-a need to prevent Lost Update. This issue can be avoided by moving the
+\vspace{.5em}\noindent\textit{TPC-C Non-monotonicity.} The Delivery
+transaction is challenging due to non-monotonicity. Each Delivery
+deletes a pending order from the New-Order table and should be
+idempotent in order to avoid billing a customer twice; this implies a
+need to prevent Lost Update. This issue can be avoided by moving the
 non-monotonicity to the real world---the carrier that picks up the
-package for an order can ensure that she is the only carrier who has
-done so---but cannot provide a correct execution with HATs
-alone. However, according to distributed transaction
-architects~\cite{entitygroup}, these compensatory actions are
-relatively common in real-world business processes.
+package for an order can ensure that no other carrier will do so---but
+cannot provide a correct execution with HATs alone. However, according
+to distributed transaction architects~\cite{entitygroup}, these
+compensatory actions are relatively common in real-world business
+processes.
 
-Throughout execution, TPC-C also requires the maintenance of several
+\vspace{.5em}\noindent\textit{Integrity Constraints.} Throughout execution, TPC-C also requires the maintenance of several
 integrity constraints. For example, Consistency Condition 1 (3.3.2.1)
@@
 
-In summary, many---but not all---TPC-C transactions are well-served by
+\vspace{.5em}\noindent\textit{Summary.} Many---but not all---TPC-C transactions are well-served by
 HATs. The two problematic transactions---New-Order and Payment---rely
@@
 \begin{center}
-\hspace{2em}\includegraphics[width=.8\columnwidth]{figs/strategylegend.pdf}
+\includegraphics[width=.8\columnwidth]{figs/strategylegend.pdf}
 \end{center}\vspace{-2em}
-\begin{center}\textbf{Within \texttt{us-east}}\end{center}\vspace{-1.5em}
-\includegraphics[width=0.90\columnwidth]{figs/lan-threads-lats.pdf}\vspace{-1em}
-\includegraphics[width=0.90\columnwidth]{figs/lan-threads-thru.pdf}
-\begin{center}\textbf{Between \texttt{us-east} and \texttt{us-west-2}}\end{center}\vspace{-1.5em}
-\includegraphics[width=0.90\columnwidth]{figs/wan-threads-lats.pdf}\vspace{-1em}
-\includegraphics[width=0.90\columnwidth]{figs/wan-threads-thru.pdf}
-\caption{YCSB performance for two clusters of three servers each
+\begin{center}\small\textbf{Within \texttt{us-east (VA)}}\end{center}\vspace{-1.5em}
+\includegraphics[width=0.90\columnwidth]{figs/finals/2lan-threads-lats-log.pdf}\vspace{-.5em}
+\includegraphics[width=0.90\columnwidth]{figs/finals/2lan-threads-thru.pdf}
+\begin{center}\small\textbf{Between \texttt{us-east (CA)} and \texttt{us-west-2 (OR)}}\end{center}\vspace{-1.5em}
+\includegraphics[width=0.90\columnwidth]{figs/finals/2wan-threads-lats-log.pdf}\vspace{-.5em}
+\includegraphics[width=0.90\columnwidth]{figs/finals/2wan-threads-thru.pdf}
+\begin{center}\small Between \textbf{\texttt{us-east} \texttt{(VA)}, \texttt{us-west-1} \texttt{(CA)},\\ \texttt{us-west-2} \texttt{(OR)}, \texttt{eu-west} \texttt{(IR)}, \texttt{ap-northeast} \texttt{(SI)}}\end{center}\vspace{-1.5em}
+\includegraphics[width=0.90\columnwidth]{figs/finals/5wan-threads-lats-log.pdf}\vspace{--.5em}
+\includegraphics[width=0.90\columnwidth]{figs/finals/5wan-threads-thru.pdf}
+\caption{YCSB performance for two clusters of five servers each
   deployed within a single datacenter and cross-datacenters.}
 \label{fig:wan-exp}
+\end{figure}
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=0.90\columnwidth]{figs/finals/scaleout-thru.pdf}
+%\includegraphics[width=0.90\columnwidth]{figs/finals/scaleout-thru-perserver.pdf}
+\end{center}\vspace{-2em}
+\caption{Scale-out of TA, Eventual, and RC.}
+\label{fig:scaleout}
+\end{figure}
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=0.90\columnwidth]{figs/finals/txnlen-thru.pdf}
+\end{center}\vspace{-2em}
+\caption{Transaction length versus throughput.}
+\label{fig:txlen}
+\end{figure}
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=0.90\columnwidth]{figs/finals/wprop-thru.pdf}
+\end{center}\vspace{-2em}
+\caption{Proportion of reads and writes versus throughput.}
+\label{fig:rprop}
 \end{figure}

### 2013-04-28 writeup/futurework.tex
some graphs and joe comments

@@
 
-\textbf{What are the limits of HAT semantics?}  In this work, we have
-taxonomized several well-defined and well-documented semantic
-properties. This is useful as it contextualizes existing database and
-distributed systems semantics but does not provide guidance as to what
-consistency models are the ``strongest''---or admit the smallest set
-of executions---within each availability class. A recent report from
-UT Austin shows that no model stronger than causal consistency can be
-achieved in a sticky highly available, one-way convergent
-system~\cite{cac}, but there are infinite highly available and sticky
-highly available models to consider. For example, always returning the
-value $1$ in response to reads is incomparable with (i.e., neither
-stronger nor weaker than) causal consistency: returning $1$ does not
-respect causality, and causal consistency returns values other than
-$1$. There are actually infintely many incomparable models (vacuously:
-return $1$, return $2$, and so on). Are there \textit{meaningful}
-properties that delineate the boundaries between availability levels?
-Is there a ``meta-property'' that exactly characterizes what
-properties are achievable with high availability or sticky high
-availability? This is an area of ongoing research within the Berkeley
-HAT project. \pbnote{SHORTEN}
+ A recent report from UT Austin shows that no model stronger than
+ causal consistency can be achieved in a sticky highly available,
+ one-way convergent system~\cite{cac}, but there are infinite highly
+ available and sticky highly available models to consider. For
+ example, always returning the value $1$ in response to reads is
+ incomparable with (i.e., neither stronger nor weaker than) causal
+ consistency: returning $1$ does not respect causality, and causal
+ consistency returns values other than $1$. There are actually
+ infintely many incomparable models (vacuously: return $1$, return
+ $2$, and so on). \textbf{What are the limits of HAT semantics?}  Are
+ there \textit{meaningful} properties that delineate the boundaries
+ between availability levels?  Is there a ``meta-property'' that
+ exactly characterizes what properties are achievable with high
+ availability or sticky high availability?
 
@@
 avoided while providing ``consistent'' operation for limited failure
-scenarios. We have performed a narrow case study, but a general
-taxonomization of techniques such as escrow
+scenarios. We have performed a narrow case study of TPC-C, but a
+general taxonomization of techniques such as escrow
 transactions~\cite{escrow}, Sagas~\cite{sagas}, and other long-running

### 2013-04-28 writeup/hat-analysis.tex
more examples, writing tweaks
some graphs and joe comments

@@
 To begin, we present well-known semantics that can be achieved in HAT
-systems. We offer proof-of-concept highly available algorithms, but
+systems. In this section, our goal is feasibility, not performance. As
+a result, we offer proof-of-concept highly available algorithms, but
 our goal is \textit{not} to provide optimal or even efficient
-implementations. As is standard in the distributed database
-literature, in our sample implementations, we consider a set of
-servers acting as replicas for a set of data items with which
-non-co-located clients interact. In our examples, we exclusively
-consider read and write operations, denoting a write of value $v$ to
-data item $d$ as $w_d(v)$ and a read from data item $d$ returning $v$
-as $r_d(v)$. We assume that all data items have the null value,
-$\bot$, at database initialization, and, unless otherwise specified,
-all transactions in the examples commit.
+implementations: the challenge is to provide algorithms that provide
+high availability. However, we briefly study performance implications
+in Section~\ref{sec:evaluation}.
+
+In our examples, we exclusively consider read and write operations,
+denoting a write of value $v$ to data item $d$ as $w_d(v)$ and a read
+from data item $d$ returning $v$ as $r_d(v)$. We assume that all data
+items have the null value, $\bot$, at database initialization, and,
+unless otherwise specified, all transactions in the examples commit.
 
@@
 To begin, \textbf{Read Uncommitted} isolation is captured by Adya as
-\textit{PL-1}, requiring that each transaction's writes are ordered
+\textit{PL-1}, requiring only that each transaction's writes are ordered
 consistently according to a single total order on transactions. This
@@
 Unlike the total order required by serializability, the order in PL-1
-does not constrain the values seen in by a transaction's read
+does not constrain the values seen by a transaction's read
 operations except once the database has reached a ``final
@@
 \textbf{Read Committed} isolation is particularly important in
-practice as it is the default of many DBMSs. Implementations differ,
-with some based on long-duration exclusive locks and short-duration
-read locks~\cite{gray-isolation} and others based on multiple
-versions. The implementations often provide recency and monotonicity
-properties beyond the simple meaning from the name, which is what is
-expressed in the implementation-agnostic definition: under Read
-Committed, transactions should do not access uncommitted or
+practice as it is the default of many DBMSs. Centralized
+implementations differ, with some based on long-duration exclusive
+locks and short-duration read locks~\cite{gray-isolation} and others
+based on multiple versions. The implementations often provide recency
+and monotonicity properties beyond the simple meaning from the name,
+which is what is expressed in the implementation-agnostic definition:
+under Read Committed, transactions should not access uncommitted or
 intermediate versions of data items. This prohibits both ``Dirty
@@
 It is fairly easy for a HAT system to prevent ``Dirty Reads'': if each
-client never writes uncommitted data to shared copies of the databaseq, then
+client never writes uncommitted data to shared copies of the database, then
 transactions will never read each others' dirty data. As a simple
@@
 
-``Repeatable Read'' isolation is a confusing property. For example,
+\textbf{Repeatable Read} isolation is a confusing property. For example,
 some IBM products use the term for fully serializable
@@
 
-However, As Berenson et al.~\cite{ansicritique} discuss, the ANSI SQL
+However, as Berenson et al.~\cite{ansicritique} discuss, the ANSI SQL
 specification allows several additional behaviors under ``Repeatable
@@
 (e.g., \texttt{SELECT WHERE}) as well as its item reads.  We call this
-\textbf{Predicate Cut Isolation} and it prohibits Phantoms. Predicate
-Cut Isolation is also achievable in HAT systems via similar
-multi-versioning as above.
+\textbf{Predicate Cut Isolation} and it prohibits Phantoms (but does
+not guarantee serializability of item-based reads and
+writes). Predicate Cut Isolation is also achievable in HAT systems via
+similar multi-versioning as above.
 
@@
 
-We can achieve TA with a two-phase multi-versioning algorithm. Clients
-buffer their writes until commit, sending the final write to each data
-item to each respective server. Along with each write, clients attach
-a the transaction ID and a list of all data items written to in the
-transaction, $L_v$. Upon receiving a new write to item $d$, a server
-places the write into a \texttt{pending} queue and notifies all other
-servers responsible for a data item in $L_v$ that it has received a
-write to $d$ with the respective transaction ID. Once a server
-receives acknowledgments with appropriate transaction ID for all items
-in $L_v$, it moves the write from \texttt{pending} to the
-\texttt{stable} database (which applies the writes according to
-highest-ID-wins). Accordingly, all writes that should be made atomic
-with a write in \texttt{stable} are guaranteed to be present on their
-respective servers, either in \texttt{pending} or
-\texttt{stable}. Given the possibility of internal aborts, writes to
-\texttt{pending} may require 2 RTTs: one to check constraints and one
-to execute the above (but these checks \textit{cannot} cause other
-transactions to block, as would be the case in a two-phase
-\textit{commit} protocol).
+TA can be achieved via multi-versioning. If clients attach vector
+clocks to every write (incrementing their own position for each
+transaction) and replicas store all writes via multi-versioning, then
+clients can safely determine which sets of data items to read; this
+approach is adopted by Swift~\cite{swift} and, to a lesser-extent,
+bolt-on causal consistency~\cite{bolton}. These systems also provide
+causal consistency, which requires sticky availability (in each,
+implemented via client-side caching). However, TA is achievable
+without stickiness via an alternate algorithm: if servers wait to
+reveal writes until they are present on all replicas, clients do not
+need to be sticky.
 
-For reads, each client tracks which values it should read. It
-maintains a mapping (vector) from data items to transaction IDs. Upon
-transaction start, the client clears its mapping, and, when it
-performs a read from a data item, it sends the server its current ID
-for the item. If the client does not supply an ID, the server returns
-the latest value from \texttt{stable}. If the ID is non-empty, the
-server either returns the matching value from \texttt{pending} or a
-value with higher ID from \texttt{pending}---along with $L_v$ for the
-write. When the client receives a response, it updates its local
-vector with $L_v$. Accordingly, even if transaction's writes to two
-keys $x$ and $y$ are present on the replica for $x$'s \texttt{stable}
-and the replica for $y$'s \texttt{pending}, a client's reads from $x$
-then $y$ will return a transactionally atomic pair of values because
-the replica for $y$ will know to check \texttt{pending}.
+We sketch a non-sticky TA algorithm and provide greater detail in our
+technical report~\cite{hat-tr}. Each replica waits to reveal write $w$
+from transaction $T$ until all of the writes in $T$ are present on all
+respective replicas (are \textit{pending stable}). Once they are,
+replicas asynchronously reveal their writes. To prevent clients from
+reading writes that have been revealed on some servers but not others,
+clients can attach a vector that uniquely identifies the transaction:
+this can be the transaction ID and a list of keys in the
+transaction. When clients attempt to read a data item from a replica,
+they can send the highest ID they have observed for that item;
+replicas respond with either a matching write (which may or may not be
+pending stable) or a write with higher ID that is pending
+stable. Clients can reset their record of IDs and keys upon
+transaction commit. Servers keep two sets of versions for each data
+item: the value with the highest ID that is pending stable and a set
+of values that are not yet pending stable.
 
-This implementation is entirely masterless and both reads and writes
-do not block for coordination. One key property is that we do not, in
-general, dictate when writes become visible. Accordingly, it is highly
+This non-sticky implementation is entirely masterless and neither reads
+nor writes block for coordination. Moreover, there are several
+possible optimizations that, for brevity, we do not describe
+here. However, one key property is that we do not, in general, dictate
+when writes become visible. Accordingly, it is highly
 available. However, we can optimize write visibility delay by ensuring
@@
 
-Our models have not yet considered interactions \textit{across}
-transactions, other than that there exists some arbitrary ordering on
-them (via Read Uncommitted). In the distributed systems literature,
-many useful \textit{safety} guarantees span multiple,
-non-transactional operations. In particular, \textit{session
-  guarantees} are used to describe guarantees across transactions
-within a given \textit{session}, ``an abstraction for the sequence
-of...operations performed during the execution of an
+As is typical in the database literature (with few
+exceptions~\cite{daudjee-session}), our models have not yet considered
+interactions \textit{across} transactions, other than that there
+exists some arbitrary ordering on them (via Read Uncommitted). In the
+distributed systems literature, many useful \textit{safety} guarantees
+span multiple, non-transactional operations. In particular,
+\textit{session guarantees} are used to describe guarantees across
+transactions within a given \textit{session}, ``an abstraction for the
+sequence of...operations performed during the execution of an
 application''~\cite{sessionguarantees}. Informally a session describes
@@
 operation, then the database will end up with $a=3$, a state that
-could not have resulted in a serial execution due to from $T_2$'s
+could not have resulted in a serial execution due to $T_2$'s
 ``Lost Update.''
@@
 Isolation, and Cursor Stability guarantees are all unavailable.
-Repeatable Read (defined by to Gray, Berenson et al., and Adya) and
+Repeatable Read (defined by Gray, Berenson et al., and Adya) and
 One-Copy Serializability need to prevent both Lost Update and Write
@@
 To the best of our knowledge, this is the first unification of
-database ACID, distributed consistency, and session guarantee
-models. Interestingly, strong one-copy serializability subsumes all
-other models, while considering the (large) power set of all
-compatible models (e.g., the diagram depicts 96 possible HAT
+database ACID transactions, distributed consistency, and session
+guarantee models. Interestingly, strong one-copy serializability
+subsumes all other models, while considering the (large) power set of
+all compatible models (e.g., the diagram depicts 96 possible HAT
 combinations) hints at the vast expanse of consistency models found in
@@
 HA & Transactional Atomicity (TA), Read Uncommitted (RU), Read
-Committed (RC), Item Cut Isolation (P-CI), Predicate Cut Isolation
+Committed (RC), Item Cut Isolation (I-CI), Predicate Cut Isolation
 (P-CI), Monotonic Reads (MR), Monotonic Writes (MW), Writes Follow
@@
 \label{fig:hat-order}
-\caption{HAT, sticky HAT (in boxes), and unavailable models (circled)
-  from Table~\protect\ref{table:hatcompared} compared
+\caption{HAT, sticky HAT (in boxes, blue), and unavailable models
+  (circled, red) from Table~\protect\ref{table:hatcompared} compared
   graphically. Directed edges represent ordering by model
@@
   Atomicity} The relationship between linearizability and
-transactional atomicity is non-obvious. Transactional atomicity
-dictates that writes to multiple keys across multiple servers are made
-visible to readers all at once, while linearizability dictates that
-writes to a single key on multiple servers are made visible to all
-readers at once---what is different? First, in linearizable (and safe
-and regular) systems, writes are made visible to clients
-\textit{immediately} after they finish. With transactional atomicity,
-there is no recency guarantee. Second, in linearizabile systems, all
-clients see all writes at the same time. With transactional atomicity,
-clients may see writes at different times depending on which replicas
-they contact. We are not aware of an analogous model in the
-distributed systems literature. Accordingly, despite apparent
-similarities, transactional atomicity is incomparable with and much
-cheaper (by availability standards) than linearizability.
+transactional atomicity is non-obvious. TA dictates that writes to
+multiple keys across multiple servers are made visible to readers all
+at once, while linearizability dictates that writes to a single key on
+multiple servers are made visible to all readers at once---what is
+different? First, in linearizable (and safe and regular) systems,
+writes are made visible to clients \textit{immediately} after they
+finish. With transactional atomicity, there is no recency
+guarantee. Second, in linearizable systems, all clients see all writes
+at the same time. With TA as defined here, clients may see writes at
+different times depending on which replicas they contact. We are not
+aware of an analogous model in the distributed systems
+literature. Accordingly, despite apparent similarities, TA is
+incomparable with and much cheaper (by availability standards) than
+linearizability.
 

### 2013-04-28 writeup/hat-appendix.tex
some graphs and joe comments

@@
 
-\section{Model}
+\section{Adya Definitions}
 
@@
 \end{definition}
+
+\section{Session Definitions}
 
@@
 \end{observation}
+
+\section{Transactional Atomicity}
+
+In our non-sticky TA algorithm, clients buffer their writes until
+commit, sending the final write to each data item to each respective
+server. Along with each write, clients attach a the transaction ID and
+a list of all data items written to in the transaction, $L_v$. Upon
+receiving a new write to item $d$, a server places the write into a
+\texttt{pending} queue and notifies all other servers responsible for
+a data item in $L_v$ that it has received a write to $d$ with the
+respective transaction ID. Once a server receives acknowledgments with
+appropriate transaction ID for all items in $L_v$, it moves the write
+from \texttt{pending} to the \texttt{stable} database (which applies
+the writes according to highest-ID-wins). Accordingly, all writes that
+should be made atomic with a write in \texttt{stable} are guaranteed
+to be present on their respective servers, either in \texttt{pending}
+or \texttt{stable}. Given the possibility of internal aborts, writes
+to \texttt{pending} may require 2 RTTs: one to check constraints and
+one to execute the above (but these checks \textit{cannot} cause other
+transactions to block, as would be the case in a two-phase
+\textit{commit} protocol).
+
+For reads, each client tracks which values it should read. It
+maintains a mapping (vector) from data items to transaction IDs. Upon
+transaction start, the client clears its mapping, and, when it
+performs a read from a data item, it sends the server its current ID
+for the item. If the client does not supply an ID, the server returns
+the latest value from \texttt{stable}. If the ID is non-empty, the
+server either returns the matching value from \texttt{pending} or a
+value with higher ID from \texttt{pending}---along with $L_v$ for the
+write. When the client receives a response, it updates its local
+vector with $L_v$. Accordingly, even if transaction's writes to two
+keys $x$ and $y$ are present on the replica for $x$'s \texttt{stable}
+and the replica for $y$'s \texttt{pending}, a client's reads from $x$
+then $y$ will return a transactionally atomic pair of values because
+the replica for $y$ will know to check \texttt{pending}.

### 2013-04-28 writeup/model.tex
some graphs and joe comments

@@
 partitions between servers~\cite{gilbert-cap}.\footnote{Under this
-  definition, a system that requires a majority of servers to be
-  online is not available.} As in a standard distributed database,
-designated servers might perform operations for different data items;
-a server that can handle an operation for a given data item is called
-a \textit{replica} for that item.\footnote{There is a further
-  distinction between a \textit{fully replicated} system, in which all
-  servers are replicas for all data items, and a \textit{partially
-    replicated} system, in which at least one server acts as a replica
-  for only a subset of all data items. For generality, and, for
-  applicability to many modern ``sharded'' or ``partitioned'' data
-  storage systems~\cite{dynamo, pnuts, bigtable, spanner, hstore}, we
-  consider partially replicated systems.} While this standard model is
-useful, there are several other cases we should consider.
+  definition, systems that require a majority of servers to be online
+  or that require the database to function correctly with high (but
+  not perfect) probability are not available.} As in a standard
+distributed database, designated servers might perform operations for
+different data items; a server that can handle an operation for a
+given data item is called a \textit{replica} for that
+item.\footnote{There is a further distinction between a \textit{fully
+    replicated} system, in which all servers are replicas for all data
+  items, and a \textit{partially replicated} system, in which at least
+  one server acts as a replica for only a subset of all data
+  items. For generality, and, for applicability to many modern
+  ``sharded'' or ``partitioned'' data storage systems~\cite{dynamo,
+    pnuts, bigtable, spanner, hstore}, we consider partially
+  replicated systems.} While this standard model is useful, there are
+several other cases we should consider.
 
@@
 may result in ``lower availability'' than a non-transactional
-requirement. Additionally, given the possibility of aborts, we need to
-ensure useful forward progress: a system can trivially guarantee
-clients a response by always aborting all
-transactions~\cite{transaction-liveness}. However, this is an
-unsatisfactory system because nothing good (transaction commit) ever
-happens; we should require a \textit{liveness} property. We cannot
+requirement. Additionally, given the possibility of system-initiated
+aborts, we need to ensure useful forward progress: a system can
+trivially guarantee clients a response by always aborting all
+transactions. However, this is an unsatisfactory system because
+nothing good (transaction commit) ever happens; we should require a
+\textit{liveness} property~\cite{transaction-liveness}. We cannot
 guarantee that every transaction will commit---transactions may choose
@@
 given replica availability for every data item in a transaction, the
-transaction eventually commits or internally
-aborts~\cite{hat-hotos}. For example, a system will violate
-transactional availability if a client that can contact two
-servers for each of the data items in its transaction does not
+transaction eventually commits (possibly after multiple client
+retries) or internally aborts~\cite{hat-hotos}. For example, a system
+will violate transactional availability if a client that can contact
+two servers for each of the data items in its transaction does not
 eventually commit in the absence of internal aborts.

### 2013-04-29 writeup/evaluation.tex
eval rewrite

@@
 
-To better understand the performance implications of HAT guarantees in
-a real-world environment and to validate the latency-based analysis in
-Section~\ref{sec:latency}, we deployed a distributed database
-prototype on public cloud infrastructure. We compared the behavior of
-eventual consistency, per-item mastering (wherein all operations to a
-given item required contacting a designated master server), Read
-Committed, and TA with read your writes guarantees. We deployed an
-experimental Java-based distributed datastore backed by LevelDB both
-within and across EC2 regions and ran the YCSB benchmark~\cite{ycsb}
-to determine the impact of latency. We partitioned the database into
-two fully replicated clusters of three \texttt{m1.xlarge} instances
-each. For HAT, we stuck YCSB clients to their local cluster and, to
-provide transactional functionality, grouped every six YCSB operations
-from the default workload to form a transaction. We report the median
-of five trials per experiment.
+To further understand the performance implications of HAT guarantees
+in a real-world environment, we implemented a HAT database
+prototype. Across a range of deployments on public cloud
+infrastructures, we found that, as Section~\ref{sec:latency}'s
+measurements suggested, ``strongly consistent'' algorithms incur
+substantial latency penalties: from 10 to 100 times their HAT
+counterparts. Moreover, HAT algorithms like TA---while more
+resource-intensive than basic eventual consistency---are able to scale
+well to many servers.
+
+\vspace{.5em}\noindent\textit{Implementation.} Our prototype database
+is a partially replicated (hash-based, partitioned) key-value backed
+by LevelDB and implemented in Java using Apache Thrift. Our prototype
+currently supports eventual consistency (via simple anti-entropy) and
+HAT RU, RC, and TA guarantees (as described in
+Section~\ref{sec:hats}). It also supports non-HAT operation whereby
+all operations for a given key are routed to a (randomly) designated
+\textit{master} replica for each key (i.e., PNUTS~\cite{pnuts}) as
+well as distributed two-phase locking. Servers are durable: they
+synchronously write to LevelDB before responding to client requests,
+while pending writes in TA are sent to a write-ahead log, which is
+synchronously flushed to disk.
+
+\vspace{.5em}\noindent\textit{Configuration.} We deploy the database
+in \textit{clusters}---sets of database servers that contain a single,
+fully replicated copy of the data---typically across datacenters,
+sticking all clients within a datacenter to their respective
+cluster. By default, we use 5 Amazon EC2 \texttt{m1.xlarge} instances
+as servers in each cluster. For our workload, we link our client
+library to the YCSB benchmark~\cite{ycsb}, grouping every eight YCSB
+operations from the default workload (50\% reads, 50\% writes) to form
+a transaction. We increase the number of keys in the workload from the
+default $1,000$ to $100,000$ with uniform random key access, keeping
+the default value size of $1KB$, and running YCSB for 180 seconds per
+configuration. Clients trivially achieve read-your-writes and
+monotonic reads guarantees via stickiness.
 
@@
 \begin{center}
-\includegraphics[width=.8\columnwidth]{figs/strategylegend.pdf}
+\hspace{2em}\includegraphics[width=.8\columnwidth]{figs/strategylegend.pdf}
 \end{center}\vspace{-2em}
-\begin{center}\small\textbf{Within \texttt{us-east (VA)}}\end{center}\vspace{-1.5em}
-\includegraphics[width=0.90\columnwidth]{figs/finals/2lan-threads-lats-log.pdf}\vspace{-.5em}
-\includegraphics[width=0.90\columnwidth]{figs/finals/2lan-threads-thru.pdf}
-\begin{center}\small\textbf{Between \texttt{us-east (CA)} and \texttt{us-west-2 (OR)}}\end{center}\vspace{-1.5em}
-\includegraphics[width=0.90\columnwidth]{figs/finals/2wan-threads-lats-log.pdf}\vspace{-.5em}
-\includegraphics[width=0.90\columnwidth]{figs/finals/2wan-threads-thru.pdf}
-\begin{center}\small Between \textbf{\texttt{us-east} \texttt{(VA)}, \texttt{us-west-1} \texttt{(CA)},\\ \texttt{us-west-2} \texttt{(OR)}, \texttt{eu-west} \texttt{(IR)}, \texttt{ap-northeast} \texttt{(SI)}}\end{center}\vspace{-1.5em}
-\includegraphics[width=0.90\columnwidth]{figs/finals/5wan-threads-lats-log.pdf}\vspace{--.5em}
-\includegraphics[width=0.90\columnwidth]{figs/finals/5wan-threads-thru.pdf}
+\begin{center}\small\textbf{A.) Within \texttt{us-east} \texttt{(VA)}}\end{center}\vspace{-1.5em}
+\includegraphics[width=\figfactor\columnwidth]{figs/finals/2lan-threads-lats.pdf}\vspace{-.5em}
+\includegraphics[width=\figfactor\columnwidth]{figs/finals/2lan-threads-thru.pdf}\vspace{-.75em}
+\begin{center}\small\textbf{B.) Between \texttt{us-east} \texttt{(CA)} and \texttt{us-west-2} \texttt{(OR)}}\end{center}\vspace{-1.5em}
+\includegraphics[width=\figfactor\columnwidth]{figs/finals/2wan-threads-lats-log.pdf}\vspace{-.5em}
+\includegraphics[width=\figfactor\columnwidth]{figs/finals/2wan-threads-thru.pdf}\vspace{-.75em}
+\begin{center}\small \textbf{C.) Between}{ \texttt{us-east} \texttt{(VA)}, \texttt{us-west-1} \texttt{(CA)},\\ \texttt{us-west-2} \texttt{(OR)}, \texttt{eu-west} \texttt{(IR)}, \texttt{ap-northeast} \texttt{(SI)}}\end{center}\vspace{-1.5em}
+\includegraphics[width=\figfactor\columnwidth]{figs/finals/5wan-threads-lats-log.pdf}\vspace{--.5em}
+\includegraphics[width=\figfactor\columnwidth]{figs/finals/5wan-threads-thru.pdf}
 \caption{YCSB performance for two clusters of five servers each
@@
 
+\vspace{.5em}\noindent\textit{Geo-replication.} We first deploy the
+database prototype across an increasing number of
+datacenters. Figure~\ref{fig:wan-exp}A shows that, when operating two
+clusters within a single datacenter, employing a single master for
+each data item results in approximately half the throughput and double
+the latency of an eventually consistent configuration: the HAT models
+are able to utilize double the number of servers. RC---essentially
+eventual consistency with buffering---is almost identical to eventual
+consistency, while TA---which incurs two writes for every client-side
+write (i.e., pending writes are sent to a WAL then subsequently moved
+into LevelDB once stable)---achieves approximately 75\% of the
+throughput. Latency increases linearly with the number of YCSB clients
+due to contention within LevelDB.
+
+ In contrast, when the two clusters are deployed across the
+ continental United States (Figure~\ref{fig:wan-exp}B), the average
+ latency of mastered operation increases to $300$ms (a $278$--$4257\%$
+ latency increase; average $37$ms latency per operation). For the same
+ number of YCSB client threads, mastered operation has substantially
+ lower throughput than the HAT configurations. Increasing the number of YCSB
+ clients \textit{does} increase the throughput of mastered operation,
+ but our Thrift-based connection processing did not gracefully handle
+ more than several thousand concurrent connections. In contrast,
+ across two datacenters, the performance of eventual, RC, and TA is
+ near-identical to a single-datacenter deployment.
+
+When we deployed five clusters (as opposed to two, as before) across
+the five AWS datacenters with lowest communication cost
+(Figure~\ref{fig:wan-exp}C), the trend continues: mastered latency
+increases to nearly $800$ms per transaction. As an attempt at reducing
+this overhead, we implemented and benchmarked a variant of
+quorum-based replication as in Dynamo~\cite{dynamo}, where clients
+sent requests to all replicas, which completed as soon as a majority
+of servers responded (guaranteeing regular
+semantics~\cite{herlihy-art}); this strategy (not pictured) did not
+substantially improve performance due to the network topology and
+because worst-case server load was unaffected. In this configuration,
+TA's relative throughput decreased: every YCSB \texttt{put} operation
+resulted in four \texttt{put} operations on remote replicas and,
+accordingly, the cost of anti-entropy increased (e.g., each server
+processed four incoming anti-entropy streams---as opposed to one
+before---reducing the opportunity for batching and decreasing
+available resources for for incoming client requests). This in turn
+increased garbage collection activity and, more importantly, IOPS when
+compared to Eventual and RC, causing throughput to peak at around half
+of Eventual. With an in-memory persistence (i.e., no LevelDB or WAN),
+TA throughput was within 20\% of eventual.
+
+We have intentionally omitted performance data for two-phase
+locking. Mastered configuration performed \textit{far} better than our
+textbook implementation, which, in addition to requiring a WAN
+round-trip per transaction, also incurred substantial overheads due to
+mutual exclusion via locking. We expect that, while techniques like
+those recently proposed in Calvin~\cite{calvin} can reduce the
+overhead of serializable transactions by avoiding locking, our
+master-based implementation and the data from
+Section~\ref{sec:latency} are reasonable lower bounds on latency.
+
 \begin{figure}[t!]
 \begin{center}
-\includegraphics[width=0.90\columnwidth]{figs/finals/scaleout-thru.pdf}
-%\includegraphics[width=0.90\columnwidth]{figs/finals/scaleout-thru-perserver.pdf}
-\end{center}\vspace{-2em}
+\hspace{2em}\includegraphics[width=.8\columnwidth]{figs/strategylegend.pdf}\vspace{-2em}
+\end{center}
+\begin{center}
+\includegraphics[width=\figfactor\columnwidth]{figs/finals/txnlen-thru.pdf}
+\end{center}\vspace{-2.25em}
+\caption{Transaction length versus throughput.}\vspace{-.5em}
+\label{fig:txlen}
+\end{figure}
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=\figfactor\columnwidth]{figs/finals/wprop-thru.pdf}
+\end{center}\vspace{-2.25em}
+\caption{Proportion of reads and writes versus throughput.}\vspace{-.5em}
+\label{fig:rprop}
+\end{figure}
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=\figfactor\columnwidth]{figs/finals/scaleout-thru.pdf}
+%\includegraphics[width=\figfactor\columnwidth]{figs/finals/scaleout-thru-perserver.pdf}
+\end{center}\vspace{-2.25em}
 \caption{Scale-out of TA, Eventual, and RC.}
@@
 
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=0.90\columnwidth]{figs/finals/txnlen-thru.pdf}
-\end{center}\vspace{-2em}
-\caption{Transaction length versus throughput.}
-\label{fig:txlen}
-\end{figure}
+\vspace{.5em}\noindent\textit{Transaction length.} As shown in
+Figure~\ref{fig:txlen} (clusters in Virginia and Oregon), throughput
+of eventual consistency, RC, and mastered operation are unaffected by
+transaction length. In contrast, TA's throughput decreases linearly
+with increased transaction length: with $1$ operation per transaction,
+TA throughput is within 18\% of eventual ($34$ bytes overhead), and
+with $128$ operations per transaction, TA throughput is within $60\%$
+($1898$ bytes overhead). This is primarily because our TA algorithm
+requires metadata proportional to the length of each transaction,
+consuming IOPS and network bandwidth.
 
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=0.90\columnwidth]{figs/finals/wprop-thru.pdf}
-\end{center}\vspace{-2em}
-\caption{Proportion of reads and writes versus throughput.}
-\label{fig:rprop}
-\end{figure}
+\vspace{.5em}\noindent\textit{Read proportion.} The default equal
+proportion of reads and writes is fairly pessimistic: for example,
+Facebook reports $99.8\%$ reads for their workload~\cite{eiger}. As
+shown in Figure~\ref{fig:rprop} (clusters in Virginia and Oregon),
+with all reads, TA is within $4.8\%$ of eventual; with all writes,
+eventual throughput decreases by $288.8\%$ compared to all reads and
+TA is within $33\%$ of eventual. At $99.8\%$ reads, TA incurs a $7\%$
+overhead ($5.8\%$ for in-memory storage).
 
-As Figure~\ref{fig:wan-exp} shows, within a single datacenter, a
-mastered implementation achieves approximately double the latency and
-slightly over half of the throughput of a HAT implementation. This is
-because, given two replicas for each key, the mastered implementation
-serves requests from only one. The mastered implementation still
-achieves $19,797$ transactions per second---over 6,000 per
-master---and is only $15.56$ms slower at peak compared to eventual
-consistency. However, eventual consistency and HAT models achieve
-substantially higher throughput--$37,575$ and $36,178$ transactions
-per second, respectively--and lower latency---$48.3\%$ and $47.2\%$
-respectively. This difference is unsurprising because, in these weaker
-models, all servers can serve both reads and writes, effectively
-doubling the capacity of the system. The weaker models still propagate
-new values between replicas, but this process is asynchronous and,
-aside from consuming disk I/O and network bandwidth, does not
-substantially impact clients. As expected, HAT models perform
-similarly to eventual consistency: our Read Committed implementation
-buffers client writes and adds no additional server-side overhead,
-while TA requires additional metadata with each write but has less
-than 5\% performance impact compared to eventual consistency.
+\vspace{.5em}\noindent\textit{Scale-out.} One of the key benefits of
+our HAT algorithms is that they are shared-nothing, meaning they
+should not compromise scalability. Figure~\ref{fig:scaleout} shows
+that varying the number of servers across two clusters in Virginia and
+Oregon (with $15$ YCSB clients per server) results in linear scale-out
+for both eventual, RC, and TA. RC and eventual scale perfectly with an
+approximately $5$x throughput increase moving from $5$ to $25$ servers
+per cluster. For the same configuration, TA scales by $3.8$x,
+achieving over $260,000$ operations per second. We believe that TA's
+performance is due to resource contention---with a memory-backed
+database (instead of LevelDB), TA scales by $4.25$x (not shown)---and
+increased garbage collection activity.
 
-In comparison, over multiple datacenters, the cost of coordination
-increases. The HAT and eventual consistency implementations continue
-to enjoy local RTTs only, and, with the exception of a slight
-performance degradation in TA---which we believe is attributable to
-Java garbage collection overheads in TA's inbound anti-entropy
-buffers---performance remains unchanged. However, each YCSB update to
-a remote replica ($\frac{2}{3}$ in one datacenter, $\frac{1}{3}$ in
-the other) of requiring a remote RTT between Virginia and Oregon (or
-vice-versa) to reach the master. This leads to a latency increase of
-around $166$ms ($746\%$) per transaction compared to a
-single-datacenter deployment. In order to attain reasonable throughput
-for the mastered deployment, we had to greatly increase the number of
-concurrent clients accessing the database, and we found that threading
-and socket overheads quickly overwhelmed the database server. Over a
-wide-area network, the increased cost of RTTs dramatically affects
-system efficiency.
-
-This brief set of experiments is not meant as an exhaustive study, but
-it validates our earlier intuition about the cost of violating
-HAT-compliance. As Deutsch points out, ignoring factors such as
-latency can ``cause big trouble and painful learning
+\vspace{.5em}\noindent\textit{Summary.} Our experimental prototype
+confirms our earlier analytical intuitions. HAT systems can provide
+useful semantics without substantial performance penalties. In
+particular, our TA algorithm can achieve throughput competitive with
+eventual consistency at the expense of increased disk and network
+utilization. Perhaps more importantly, all HAT algorithms circumvent
+high WAN latencies inevitable with non-HAT implementations. Our results
+highlight Deutcsh's observation that ignoring factors such as latency
+can ``cause big trouble and painful learning
 experiences''~\cite{fallacies-deutsch}---in a single-site context,

### 2013-04-29 writeup/hat-analysis.tex
eval rewrite

@@
 \label{fig:hat-order}
-\caption{HAT, sticky HAT (in boxes, blue), and unavailable models
-  (circled, red) from Table~\protect\ref{table:hatcompared} compared
-  graphically. Directed edges represent ordering by model
-  strength. Models that do not share a common ancestor can be
-  simultaneously achieved, and the resulting availability is that of
-  the weakest available model in the combination.}
+\caption{Partial ordering of HAT, sticky HAT (in boxes, blue), and
+  unavailable models (circled, red) from
+  Table~\protect\ref{table:hatcompared}. Directed edges represent
+  ordering by model strength. Incomparable models can be
+  simultaneously achieved, and the availability of a combination of
+  models has the availability of the least available individual
+  model.}
 \label{fig:hatcompared}

### 2013-04-29 writeup/intro.tex
some graphs and joe comments
eval rewrite

@@
 many weak data consistency models from distributed systems are also
-HAT-compliant.
+both HAT-compliant and simultaneously achievable with traditional ACID
+properties.
 
@@
 database context). This sticky availability is easily implementable
-and widely employed~\cite{eiger, vogels-defs} but is strictly less
-available than traditional definitions of high availability.
+and widely employed~\cite{eiger, vogels-defs} but is more restrictive
+than traditional definitions of high availability.
 

### 2013-04-29 writeup/conclusion.tex
More tweaks

@@
-\section{Conclusion}
+\section{Conclusion and Future Work}
 The current state of database software offers uncomfortable and
@@
 offers a useful connection between storage semantics studied in the
-database and distributed systems communities. The next challenge is to
-return to issues of performance and systems architecture within the
-HAT design space.  Based on our understanding of what is desired in
-the field and newfound knowledge of what is possible to achieve, we
-believe HATs provide a large and useful design space to pursue.
+database and distributed systems communities.
+
+HATs raise several challenges for future study. In this paper, we have
+highlighted standardized models, but we would like a more concise
+description of which models are and are not achievable in HAT
+systems. Similarly, while we have studied the analytical and
+experiment behavior of several HAT models, there is substantial work
+in further understanding the performance and design of systems within
+the large set of HAT models. We also believe it is time to revisit
+techniques such as escrow transactions~\cite{escrow},
+Sagas~\cite{sagas}, and other long-running transaction support in
+order to aid HAT programming. Weakened failure assumptions as in
+escrow or in the form of bounded network asynchrony could enable
+richer HAT semantics at the cost of general-purpose
+availability. Alternatively, there is a range of possible hybrid
+systems providing strong semantics during partition-free periods and
+weakened semantics during partitions. Based on our understanding of
+what is desired in the field and newfound knowledge of what is
+possible to achieve, we believe HATs provide a large and useful design
+space to explore.
 

### 2013-04-29 writeup/relatedwork.tex
More tweaks

@@
 provide transactional causal consistency guarantees with varying
-availability and represent a new class of sticky HAT systems.
-
-\pbnote{add TA stuff}
+availability and represent a new class of sticky HAT systems. There
+are infinitely many HAT models (i.e., always reading value 1 is
+comparable with always returning value 2), but a recent report from UT
+Austin shows that no model stronger than causal consistency can be
+achieved in a sticky highly available, \textit{one-way convergent}
+system~\cite{cac}. This result is promising and complementary to our
+results for general-purpose convergent data stores.
 

### 2013-04-29 writeup/conclusion.tex
smaller, smaller

@@
-\section{Conclusion and Future Work}
+\section{Conclusions and Future Work}
 The current state of database software offers uncomfortable and

### 2013-04-29 writeup/hat-analysis.tex
smaller, smaller

@@
 
-\textbf{Repeatable Read} isolation is a confusing property. For example,
-some IBM products use the term for fully serializable
+\textbf{Repeatable Read} isolation is a confusing property. For
+example, some IBM products use the term for fully serializable
 isolation~\cite{hat-hotos}, while Gray~\cite{gray-isolation}, Berenson
@@
 identical to serializability in a key-value store without
-predicate-based access). In this latter model, the \textit{Phantom
-  Problem}, whereby two successive predicate-based reads return
-different data~\cite{gray-isolation}, is the only kind of
+predicate-based access). In models like Adya's PL-2.99, the
+\textit{Phantom Problem}, whereby two successive predicate-based reads
+return different data~\cite{gray-isolation}, is the only kind of
 non-serializable behavior allowed. In this paper, we reserve the use
@@
 
-Transactional atomicity (TA) is core to ACID guarantees. Although, at
+\textbf{Transactional atomicity (TA)} is core to ACID guarantees. Although, at
 least by the ACID acronym, it is not an ``isolation'' property, TA
@@
 
-\vspace{.5em}\noindent\textit{{Monotonic reads}} requires that, within
+\vspace{.5em}\noindent\textbf{{Monotonic reads}} requires that, within
 a session, subsequent reads to a given object ``never return any
@@
 
-\vspace{.5em}\noindent\textit{{Monotonic writes}} requires that each
+\vspace{.5em}\noindent\textbf{{Monotonic writes}} requires that each
 session's writes become visible in the order they were submitted by
@@
 
-\vspace{.5em}\noindent\textit{{Writes Follow Reads}} requires that, if
+\vspace{.5em}\noindent\textbf{{Writes Follow Reads}} requires that, if
 a session observes an effect of transaction $T_1$ and subsequently
@@
 
-\vspace{.5em}\noindent\textit{{Read your writes}} requires
+\vspace{.5em}\noindent\textbf{{Read your writes}} requires
 that whenever a client reads a given data item after updating it, the
@@
 
-\vspace{.5em}\noindent\textit{{PRAM}} (Pipelined Random Access
+\vspace{.5em}\noindent\textbf{{PRAM}} (Pipelined Random Access
 Memory) provides the illusion of serializing each session's operations
@@
 
-\vspace{.5em}\noindent\textit{{Causal
+\vspace{.5em}\noindent\textbf{{Causal
     consistency}}~\cite{causalmemory} is the combination of all of the

### 2013-04-29 writeup/intro.tex
smaller, smaller

@@
 which requires explicit consideration in a client-server replicated
-database context). This sticky availability is easily implementable
-and widely employed~\cite{eiger, vogels-defs} but is more restrictive
-than traditional definitions of high availability.
+database context). This sticky availability is widely
+employed~\cite{eiger, vogels-defs} but is less restrictive (and
+therefore more easily achievable) than traditional high availability.
 

### 2013-04-29 writeup/conclusion.tex
more writing and polishing

@@
 the large set of HAT models. We also believe it is time to revisit
-techniques such as escrow transactions~\cite{escrow},
-Sagas~\cite{sagas}, and other long-running transaction support in
-order to aid HAT programming. Weakened failure assumptions as in
-escrow or in the form of bounded network asynchrony could enable
-richer HAT semantics at the cost of general-purpose
+techniques such as Sagas~\cite{sagas} and other long-running
+transaction support in order to aid HAT programming. Weakened failure
+assumptions as in escrow or in the form of bounded network asynchrony
+could enable richer HAT semantics at the cost of general-purpose
 availability. Alternatively, there is a range of possible hybrid

### 2013-04-29 writeup/evaluation.tex
more writing and polishing

@@
   properties on public cloud infrastructure.
-\end{myenumerate}\vspace{-.5em}
+\end{myenumerate}
 
@@
 
-Many existing database transaction and concurrency control algorithms
-are not designed for high availability. These algorithms often presume
-a single-server deployment or a requirement for serializability. As
-a consequence, traditional transaction processing systems are not
-well-optimized for a HAT context. In this section, we briefly discuss
-design decisions and algorithmic details that preclude high
-availability.
+Most existing database transaction and concurrency control algorithms
+are not designed for high availability and often presume a
+single-server deployment or a requirement for serializability. In this
+section, we briefly discuss design decisions and algorithmic details
+that preclude high availability.
 
@@
 general-purpose read-write transactions in a distributed
-setting~\cite{bernstein-book} require at least one round trip time
-(RTT) before committing. As an example, traditional two-phase locking
-for a transaction of length $T$ may require $T$ \texttt{lock}
-operations and will require at least one \texttt{lock} and one
-\texttt{unlock} operation.  In a distributed environment, each of
-these operations requires coordination, either with other database
-servers or with a lock service. If this coordination mechanism is
-unavailable, transactions cannot safely commit. Similarly, optimistic
-concurrency control requires coordinating via a validation step, while
-deterministic transaction scheduling~\cite{deterministic-scheduling}
-requires a scheduler. Serializability under multi-version concurrency
-control requires checking for update conflicts. All told, the reliance
-on a globally agreed total order necessitates a minimum of one
-round-trip to a designated master or coordination service for each of
-these classic algorithms.  The cost of round trips will be determined
-by the deployment environment, as we saw in
-Section~\ref{sec:motivation}; we will demonstrate this cost on public
-cloud infrastructure in Section~\ref{sec:prototype}.
+setting~\cite{bernstein-book, davidson-survey, wisemann-survey}
+require at least one round trip time (RTT) before committing. As an
+example, traditional two-phase locking for a transaction of length $T$
+may require $T$ \texttt{lock} operations and will require at least one
+\texttt{lock} and one \texttt{unlock} operation.  In a distributed
+environment, each of these operations requires coordination, either
+with other database servers or with a lock service. If this
+coordination mechanism is unavailable, transactions cannot safely
+commit. Similarly, optimistic concurrency control requires
+coordinating via a validation step, while deterministic transaction
+scheduling~\cite{deterministic-scheduling} requires a
+scheduler. Serializability under multi-version concurrency control
+requires checking for update conflicts. All told, the reliance on a
+globally agreed total order necessitates a minimum of one round-trip
+to a designated master or coordination service for each of these
+classic algorithms.  The cost of round trips will be determined by the
+deployment environment, as we saw in Section~\ref{sec:motivation}; we
+will demonstrate this cost on public cloud infrastructure in
+Section~\ref{sec:prototype}.
 
@@
 
-\vspace{.5em}\noindent\textit{Summary.} Many---but not all---TPC-C transactions are well-served by
-HATs. The two problematic transactions---New-Order and Payment---rely
-on non-monotonic state update. The former can be modified to ensure ID
-uniqueness but not sequential ID ordering, while the latter is an
-inherently non-monotonic action requiring external compensation or
-stronger consistency protocols. Based on these experiences and
-discussions with practitioners, we expect that, especially for
-read-dominated workloads found in many online services, HAT guarantees
-will provide useful semantics for many applications.
+\vspace{.5em}\noindent\textit{Summary.} Many---but not all---TPC-C
+transactions are well-served by HATs. The two problematic
+transactions, New-Order and Payment, rely on non-monotonic state
+update. The former can be modified to ensure ID uniqueness but not
+sequential ID ordering, while the latter is inherently
+non-monotonic, requiring external compensation or stronger
+consistency protocols. Based on these experiences and discussions with
+practitioners, we expect that, especially for read-dominated workloads
+found in many online services, HAT guarantees will provide useful
+semantics for many applications.
 
@@
 fully replicated copy of the data---typically across datacenters,
-sticking all clients within a datacenter to their respective
-cluster. By default, we use 5 Amazon EC2 \texttt{m1.xlarge} instances
-as servers in each cluster. For our workload, we link our client
-library to the YCSB benchmark~\cite{ycsb}, grouping every eight YCSB
-operations from the default workload (50\% reads, 50\% writes) to form
-a transaction. We increase the number of keys in the workload from the
-default $1,000$ to $100,000$ with uniform random key access, keeping
-the default value size of $1KB$, and running YCSB for 180 seconds per
-configuration. Clients trivially achieve read-your-writes and
-monotonic reads guarantees via stickiness.
+sticking all clients within a datacenter to their respective cluster
+(trivially providing read-your-writes and monotonic reads
+guarantees). By default, we use 5 Amazon EC2 \texttt{m1.xlarge}
+instances as servers in each cluster. For our workload, we link our
+client library to the YCSB benchmark~\cite{ycsb}, grouping every eight
+YCSB operations from the default workload (50\% reads, 50\% writes) to
+form a transaction. We increase the number of keys in the workload
+from the default $1,000$ to $100,000$ with uniform random key access,
+keeping the default value size of $1KB$, and running YCSB for 180
+seconds per configuration.
 
@@
 \hspace{2em}\includegraphics[width=.8\columnwidth]{figs/strategylegend.pdf}
-\end{center}\vspace{-2em}
+\end{center}\vspace{-2.5em}
 \begin{center}\small\textbf{A.) Within \texttt{us-east} \texttt{(VA)}}\end{center}\vspace{-1.5em}
@@
 \caption{YCSB performance for two clusters of five servers each
-  deployed within a single datacenter and cross-datacenters.}
+  deployed within a single datacenter and cross-datacenters.}\vspace{-1.5em}
 \label{fig:wan-exp}
@@
 write (i.e., pending writes are sent to a WAL then subsequently moved
-into LevelDB once stable)---achieves approximately 75\% of the
+into LevelDB once stable)---achieves ~75\% of the
 throughput. Latency increases linearly with the number of YCSB clients
@@
 \end{center}\vspace{-2.25em}
-\caption{Transaction length versus throughput.}\vspace{-.5em}
+\caption{Transaction length versus throughput.}\vspace{-1em}
 \label{fig:txlen}
-\end{figure}
-\begin{figure}[t!]
 \begin{center}
@@
 \end{center}\vspace{-2.25em}
-\caption{Proportion of reads and writes versus throughput.}\vspace{-.5em}
+\caption{Proportion of reads and writes versus throughput.}\vspace{-1em}
 \label{fig:rprop}
-\end{figure}
-\begin{figure}[t!]
 \begin{center}
@@
 \end{center}\vspace{-2.25em}
-\caption{Scale-out of TA, Eventual, and RC.}
+\caption{Scale-out of TA, Eventual, and RC.}\vspace{-1.5em}
 \label{fig:scaleout}
@@
 with $128$ operations per transaction, TA throughput is within $60\%$
-($1898$ bytes overhead). This is primarily because our TA algorithm
-requires metadata proportional to the length of each transaction,
-consuming IOPS and network bandwidth.
+($1898$ bytes overhead). This reflects our TA algorithm's metadata
+requirements, which are proportional to transaction length and consume
+IOPS and network bandwidth.
 

### 2013-04-29 writeup/hat-analysis.tex
more writing and polishing

@@
 \begin{align*}
-\small\vspace{-.5em}
+\small\vspace{-1em}
 T_1 &: w_x(1)~w_y(1)
 \\T_2 &: w_x(2)~w_y(2)
-\\T_3 &: r_x(a)~r_y(b)\vspace{-.5em}
+\\T_3 &: r_x(a)~r_y(b)\vspace{-1em}
 \end{align*}
@@
 \begin{align*}
-\small\vspace{-.5em}
+\small\vspace{-1em}
 T_1 &: w_x(1)~w_x(2)
 \\T_2 &: w_x(3)\\
-T_3 &: r_x(a)\vspace{-.5em}
+T_3 &: r_x(a)\vspace{-1em}
 \end{align*}
@@
 for each key):
-\vspace{-.5em}
+\vspace{-1em}
 \begin{align*}
@@
 Read your writes is not achievable in a highly available
-system. Consider a client that executes the following two transactions
-in succession:
+system. Consider a client that executes the following two transactions:
 \begin{align*}
-\small\vspace{-.5em}
+\small\vspace{-1em}
 T_1 &: w_x(1)
-\\T_2 &: r_x(a)\vspace{-.5em}
+\\T_2 &: r_x(a)\vspace{-1em}
 \end{align*}
@@
 the rest of the other servers, the server must allow $T_1$ to
-commit. If the client executes $T_2$ against the same (partitioned)
+commit. If the client subsequently executes $T_2$ against the same (partitioned)
 server, then it will be able to read its writes. However, if the
@@
 transactions' effects survive $F$ server faults requires at least
-$F+1$ non-failing nodes.
+$F+1$ non-failing replicas.
 
@@
 While there are infinitely many HAT models
-(Section~\ref{sec:futurework}), at this point, we have largely
+(Section~\ref{sec:relatedwork}), at this point, we have largely
 exhausted the range of achievable, previously defined (and useful)
@@
 \begin{align*}
-\small\vspace{-.5em}
+\small\vspace{-1em}
 T_1 &: r_x(a) w_x(a+2)
-\\T_2 &: w_x(2)\vspace{-.5em}
+\\T_2 &: w_x(2)\vspace{-1em}
 \end{align*}
@@
 \begin{align*}
-\small\vspace{-.5em}
+\small\vspace{-1em}
 T_1 &: r_x(100)~w_x(100+20=120)
-\\T_2 &: r_x(100)~w_x(100+30=130)\vspace{-.5em}
+\\T_2 &: r_x(100)~w_x(100+30=130)\vspace{-1em}
 \end{align*}
@@
   update\lostupdate, preventing write skew\rwskew, and requiring
-  recency guarantees\linearizable.}
+  recency guarantees\linearizable.}\vspace{-1.5em}
 \label{table:hatcompared}
@@
   models has the availability of the least available individual
-  model.}
+  model.}\vspace{-1em}
 \label{fig:hatcompared}

### 2013-04-29 writeup/hat-appendix.tex
more writing and polishing

@@
 
-\section{Adya Definitions}
+\section{Formal Definitions}
+
+In this section, we formally define HAT transactional semantics. Our
+formalism is based off of that of Adya~\cite{adya}. For the reader
+familiar with his formalism, this is a mostly-straightforward exercise
+combining transactional models with distributed systems
+semantics. While the novel aspects of this work largely pertain
+to \textit{using} these definitions (e.g., in Section~\ref{sec:hats}),
+we believe it is instructive to accompany them by appropriate
+definitions to both eliminate any ambiguity in prose and for the
+enjoyment of more theoretically-inclined readers.
+
+\subsection{Model}
+
+Here, we briefly describe our database model. It is, with the
+exception of sessions, identical to that of Adya~\cite{adya}. We omit
+a full duplication of his formalism here but highlight several salient
+criteria. We refer the interested reader to Adya's Ph.D. thesis,
+Section 3.1 (pp. 33--43).
+
+Users submit transactions to a database system that contains multiple
+versions of sets of objects. Each transaction is composed of writes,
+which create new versions of an object, and reads, which return a
+written version or the initial version of the object. A transaction's
+last operation is either \texttt{commit} or \texttt{abort}, and there
+is exactly one invocation of either these two operations per
+transaction. Transactions can either read individual items or read
+based on predicates, or logical ranges of data items.
+
+\begin{definition}[Version set of a predicate-based operation]
+When a transaction executes a read or write based on a predicate P,
+the system selects a version for each tuple in P’s relations. The set
+of selected versions is called the \textit{Version set} of this predicate-based
+operation and is denoted by Vset(P).
+\end{definition}
+
+A history over transactions has two parts: a partial order of events
+reflects the ordering of operations with respect to each transaction
+and a (total) version order ($\ll$) on the committed versions of each
+object.
+
+As a departure from Adya's formalism, to capture the use of session
+guarantees, we allow transactions to be grouped into sessions. We
+represent sessions as a partial ordering on committed transactions
+such that each transaction in the history appears in at most one
+session.
+
+\subsection{Conflict and Serialization Graphs}
+
+To reason about isolation anomalies, we use Adya's concept of a
+conflict graph, which is composed of dependencies between
+transaction. The definitions in this section are directly from Adya,
+with two difference. First, we expand Adya's formalism to deal
+with \textit{per-item} dependencies. Second, we define session
+dependencies (Definitions~\ref{def:sd1} and~\ref{def:sd2}).
 
@@
 
-\section{Session Definitions}
-
 \begin{definition}[Directly Session-Depends by $S$]
+\label{def:sd1}
 A transaction $T_j$ directly session-depends on transaction $T_i$ if
@@
 \begin{definition}[Directly Session-Depends]
+\label{def:sd2}
 A transaction $T_j$ directly session-depends on transaction $T_i$ if
@@
 
-\section{HAT Guarantees}
+The dependencies for a history $H$ form a graph called its Directed
+Serialization Graph ($DSG(H)$).
+
+\subsection{Transactional Anomalies and Isolation Levels}
+
+Following Adya, we define isolation levels according to
+possible \textit{anomalies}--typically represented by cycles in the
+serialization graphs. Definitions~\ref{def:n-ici}--\ref{def:lostupdate}
+are not found in Adya but are found (albeit not in this formalism) in
+Berenson et al.~\cite{ansicritique} and the literature on session
+guarantees~\cite{sessionguarantees, vogels-defs}.
+
+\begin{definition}[Write Cycles (G0)]
+A history $H$ exhibits phenomenon G0 if $DSG(H)$ contains a directed
+cycle consisting entirely of write-dependency edges.
+\end{definition}
 
 \begin{definition}[Read Uncommitted]
+A system that provides Read Uncommitted isolation prohibits phenomenon G0.
+\end{definition}
+
+\begin{definition}[Aborted Reads (G1a)]
+A history $H$ exhibits phenomenon G1a if it contains an aborted
+transaction $T_1$ and a committed transaction $T_2$ such that $T_2$ has read
+some object (maybe via a predicate) modiﬁed by $T_1$.
+\end{definition}
+
+\begin{definition}[Intermediate Reads (G1b)]
+A history $H$ exhibits phenomenon G1b if it contains a committed
+transaction $T_2$ that has read a version of object $x$ (maybe via a
+predicate) written by transaction $T_1$ that was not $T_1$’s ﬁnal
+modiﬁcation of $x$.
+\end{definition}
+
+\begin{definition}[Circular Information Flow (G1c)]
+A history $H$ exhibits phenomenon G1c if $DSG(H)$ contains a directed
+cycle consisting entirely of dependency edges.
 \end{definition}
@@
 \begin{definition}[Read Committed]
+A system that provides Read Committed isolation prohibits phenomenon G0, G1a, G1b, and G1c.
 \end{definition}
 
-\begin{definition}[Item Cut Non-Isolation (N-ICI)]
-A history $H$ exhibits phenomenon N-ICI if there exists a transaction
-$T_i$ such that $T_i$ directly item-depends by $x$ on more than one
-other transaction.
+\begin{definition}[Item Cut Non-Isolation (N-I-CI)]
+\label{def:n-ici}
+A history $H$ exhibits phenomenon N-ICI if $DSG(H)$ contains a
+transaction $T_i$ such that $T_i$ directly item-depends by $x$ on more
+than one other transaction.
 \end{definition}
@@
 
-\begin{definition}[Item Cut Isolation (ICI)]
-A system that provides Item Cut Isolation prohibits phenomenon N-ICI.
+\begin{definition}[Item Cut Isolation (I-CI)]
+A system that provides Item Cut Isolation prohibits phenomenon N-I-CI.
 \end{definition}
 
-\begin{definition}[Predicate Cut Non-Isolation (N-ICI)]
-A history $H$ exhibits phenomenon N-ICI if there exists a transaction
-$T_i$ such that $P= \{p \mid T_i$ directly predicate-read-depends by
-$p$ on another other transaction$\}$ and $\bigcap_{p\in P} P \neq
-\emptyset$. (TODO: this isn't necessarily it)
+\begin{definition}[Predicate Cut Non-Isolation (N-P-CI)]
+A history $H$ exhibits phenomenon N-I-CI if $DSG(H)$ contains a
+transaction $T_i$ such that, for two predicate-based reads $p_k$,
+$p_j$ in $P$ such that $p_k \subseteq p_j$ and $T_i$ directly
+predicate-read-depends by $p_k$ on another transaction $T_l$, but
+$T_i$ does not directly predicate-read-depend by $p_j$ on $T_l$.
 \end{definition}
 
-\begin{definition}[Predicate Cut Isolation (ICI)]
-A system that provides Predicate Cut Isolation prohibits phenomenon N-ICI.
+\begin{definition}[Predicate Cut Isolation (P-CI)]
+A system that provides Predicate Cut Isolation prohibits phenomenon N-P-CI.
 \end{definition}
@@
 \begin{definition}[Non-transactional Atomicity (N-TA)]
-A history exhibits phenomenon N-TA if it contains a directed cycle
+A history $H$ exhibits phenomenon N-TA if $DSG(H)$ contains a directed cycle
 consisting of exactly one read-dependency edge by $x$ from $T_i$ to
@@
 \begin{definition}[Non-monotonic Reads (N-MR)]
-A history exhibits phenomenon N-MR if it contains a directed cycle
+A history $H$ exhibits phenomenon N-MR if $DSG(H)$ contains a directed cycle
 consisting of a transitive session-dependency between transactions
@@
 \begin{definition}[Non-monotonic Writes (N-MW)]
-A history exhibits phenomenon N-MW if it contains a directed cycle
+A history $H$ exhibits phenomenon N-MW if $DSG(H)$ contains a directed cycle
 consisting of a transitive session-dependency between transactions
@@
 \begin{definition}[Writes Do Not Follow Reads (N-WFR)]
-A history exhibits phenomenon N-WFR if, for all committed transactions
-$T_1$, $T_2$, $T_3$ such that $T_2$ write-depends on $T_1$ and $T_3$
-write-depends on $T_2$, $T_3$ does not directly anti-depend on $T_1$.
+A history $H$ exhibits phenomenon N-WFR if, in $DSG(H)$, for all
+committed transactions $T_1$, $T_2$, $T_3$ such that $T_2$
+write-depends on $T_1$ and $T_3$ write-depends on $T_2$, $T_3$ does
+not directly anti-depend on $T_1$.
 \end{definition}
@@
 \begin{definition}[Not-Read Your Writes (N-RYW)]
-A history exhibits phenomenon N-RYW if it contains a directed cycle
+A history $H$ exhibits phenomenon N-RYW if $DSG(H)$ contains a directed cycle
 consisting of a transitive session-dependency between transactions
@@
 
-\begin{observation}
-Read Your Writes is not highly available.
-\end{observation}
-
-\begin{corollary}
-PRAM and Causal Consistency are not highly available.
-\end{corollary}
-
-\begin{definition}[Recency Guarantees]
+\begin{definition}[Lost Update]
+A history $H$ exhibits phenomenon Lost if $DSG(H)$ contains a directed
+cycle having one or more item-antidependency edges and all edges are
+by the same data item $x$.
+\label{def:lostupdate}
 \end{definition}
 
-\begin{definition}[Lost Update]
+\begin{definition}[Write Skew (Adya G2-item)]
+A history $H$ exhibits phenomenon Write Skew if $DSG(H)$ contains a directed
+cycle having one or more item-antidependency edges.
 \end{definition}
 
-\begin{definition}[Write Skew]
+For Snapshot Isolation, we depart from Adya's recency-based definition
+(see Adya Section 4.3). Nonetheless, implementations of this
+definition will still be unavailable due to reliance of preventing
+Lost Update.
+
+\begin{definition}[Snapshot Isolation]
+A system that provides Snapshot Isolation prevents phenomena G0, G1a,
+G1b, G1c, N-P-CI, N-TA, and Lost Update.
 \end{definition}
 
-\begin{definition}[Snapshot Isolation]
+For Repeatable Read, we return to Adya.
+
+\begin{definition}[Repeatable Read]
+A system that provides Repeatable Read Isolation prohibits phenomena
+G0, G1a, G1b, G1c, and Write Skew.
 \end{definition}
 
-\begin{definition}[Repeatable Read]
-\end{definition}
-
-\begin{definition}[Safe Register]
-\end{definition}
-
-\begin{definition}[Regular Register]
-\end{definition}
-
-\begin{definition}[Linearizable Register]
-\end{definition}
-
-\begin{definition}[Strong One-copy Serializability]
-\end{definition}
-
-\begin{definition}[Model Strength]
-We say that a consistency model $M_1$ is stronger than model $M_2$ if
-$M_1$ admits a strict subset of the histories admitted by $M_2$.
-\end{definition}
-
-\begin{observation}
-Transactional Atomicity is weaker than Repeatable Read.
-\end{observation}
-
-\begin{observation}
-Transactional Atomicity is weaker than Snapshot Isolation.
-\end{observation}
-
-\begin{observation}
-Causality is weaker than linearizability.
-\end{observation}
-
-\begin{observation}
-Read Your Writes is weaker than safe register semantics.
-\end{observation}
+For definitions of safe, regular, and linearizable register semantics,
+we refer the reader to Herlihy's textbook~\cite{herlihy-art}.
 

### 2013-04-29 writeup/intro.tex
more writing and polishing

@@
 semantics provided by current highly available systems. This raises a
-natural question: which of these semantics can be provided while
-remaining highly available?
+natural question: which of these semantics can be provided with high
+availability?
 
-Unfortunately, the relationship between high availability and ACID
-semantics has not been well-explored. We have a strong understanding
-of weak isolation in the single-server context from which it
-originated~\cite{adya, ansicritique, gray-isolation} and many papers
-offer techniques for providing distributed
+Unfortunately, the relationship between ACID semantics and high
+availability has not been well-explored. We have a strong
+understanding of weak isolation in the single-server context from
+which it originated~\cite{adya, ansicritique, gray-isolation} and many
+papers offer techniques for providing distributed
 serializability~\cite{bernstein-book, spanner, daudjee-session,
@@
 isolation~\cite{daudjee-snapshot,generalizedsnapshot, kemme-snapshot,
-  walter}. Additionally, the distributed computing, programming
-languages, and parallel hardware literature contains many
-consistency models for single operations on replicated objects~\cite{
-  ceri-mechanism, chen-mechanism, pnuts, herlihy-art, eiger, cac,
-  sessionguarantees}. However, the literature has largely ignored
-semantic guarantees for multiple operations operating on multiple data
-items in a highly available distributed environment.
+  walter}. Additionally, the distributed computing and parallel
+hardware literature contains many consistency models for single
+operations on replicated objects~\cite{pnuts, herlihy-art, eiger, cac,
+  sessionguarantees}. However, the literature lends few clues for
+providing semantic guarantees for multiple operations operating on
+multiple data items in a highly available distributed environment.
 
@@
 rather than an inherent property of the semantics. Our investigation
-shows that, besides serializability, only Snapshot Isolation and
-Repeatable Reads are not HAT-compliant, while most other isolation
-levels are achievable with high availability. We also demonstrate that
-many weak data consistency models from distributed systems are also
-both HAT-compliant and simultaneously achievable with traditional ACID
+shows that, besides serializability, Snapshot Isolation and Repeatable
+Reads are not HAT-compliant, while most other isolation levels are
+achievable with high availability. We also demonstrate that many weak
+data consistency models from distributed systems are also both
+HAT-compliant and simultaneously achievable with traditional ACID
 properties.
@@
 provide acceptable semantics for a wide range of programs, especially
-those with monotonic logic and commutative updates~\cite{calm, blooml,
-  crdt}. HAT systems can also enforce arbitrary foreign key
+those with monotonic logic and commutative updates~\cite{calm, crdt}. HAT systems can also enforce arbitrary foreign key
 constraints on multi-item updates and sometimes provide limited

### 2013-04-29 writeup/motivation.tex
some graphs and joe comments
more writing and polishing

@@
 
-\begin{table}
+\begin{table}[t!]
 \vspace{.5em}
@@
 
-\begin{figure}
-\includegraphics[width=\columnwidth]{graphs/ping-plot.pdf}\vspace{-1em}
+\begin{figure}[t!]
+\includegraphics[width=\columnwidth]{figs/ping-plot.pdf}\vspace{-1em}
 \caption{CDF of round-trip times for slowest inter- and intra-
@@
 \hline
-\multicolumn{3}{|p{7cm}|}{{\footnotesize RC: read committed, RR: repeatable read, SI: snapshot isolation, S: serializability, CS: cursor stability, CR: consistent read}}\\\hline
+\multicolumn{3}{|p{7cm}|}{{\begin{small}{RC: read committed, RR: repeatable read, SI: snapshot isolation, S: serializability, CS: cursor stability, CR: consistent read}\end{small}}}\\\hline
 
@@
   databases as of January 2013 (from
-  \protect\cite{hat-hotos}).}\vspace{-1.5em}
+  \protect\cite{hat-hotos}).}\vspace{-2.5em}
 \label{table:existing}

### 2013-04-29 writeup/relatedwork.tex
more writing and polishing

@@
 \section{Related Work}
+\label{sec:relatedwork}
 
@@
 overview in Section~\ref{sec:evaluation}, including eager database
-replication techniques~\cite{kemme-classification}. Additionally
-several earlier surveys consider distributed mechanisms for achieving
-one-copy serializability~\cite{wisemann-survey}, linearizability,
-consistent snapshots, and recency bounds~\cite{ceri-mechanism,
-  chen-mechanism}.
+replication techniques~\cite{kemme-classification} and distributed
+mechanisms for achieving one-copy
+serializability~\cite{wisemann-survey}.
+
+%, linearizability, consistent snapshots, and recency
+%bounds~\cite{ceri-mechanism, chen-mechanism}.
 

### 2013-04-29 writeup/evaluation.tex
small nits

@@
 is a partially replicated (hash-based, partitioned) key-value backed
-by LevelDB and implemented in Java using Apache Thrift. Our prototype
-currently supports eventual consistency (via simple anti-entropy) and
-HAT RU, RC, and TA guarantees (as described in
+by LevelDB and implemented in Java using Apache Thrift. It currently
+supports eventual consistency (last-writer-wins RU, via simple
+anti-entropy) and HAT RC and TA guarantees (as described in
 Section~\ref{sec:hats}). It also supports non-HAT operation whereby
@@
 \vspace{.5em}\noindent\textit{Configuration.} We deploy the database
-in \textit{clusters}---sets of database servers that contain a single,
-fully replicated copy of the data---typically across datacenters,
-sticking all clients within a datacenter to their respective cluster
-(trivially providing read-your-writes and monotonic reads
-guarantees). By default, we use 5 Amazon EC2 \texttt{m1.xlarge}
+in \textit{clusters}---disjoint sets of database servers that each
+contain a single, fully replicated copy of the data---typically across
+datacenters, sticking all clients within a datacenter to their
+respective cluster (trivially providing read-your-writes and monotonic
+reads guarantees). By default, we use 5 Amazon EC2 \texttt{m1.xlarge}
 instances as servers in each cluster. For our workload, we link our

### 2013-04-29 writeup/hat-analysis.tex
small nits

@@
   update\lostupdate, preventing write skew\rwskew, and requiring
-  recency guarantees\linearizable.}\vspace{-1.5em}
+  recency guarantees\linearizable.}\vspace{-.5em}
 \label{table:hatcompared}

### 2013-05-01 writeup/evaluation.tex
more writin

@@
 measurements suggested, ``strongly consistent'' algorithms incur
-substantial latency penalties: from 10 to 100 times their HAT
-counterparts. Moreover, HAT algorithms like TA---while more
-resource-intensive than basic eventual consistency---are able to scale
-well to many servers.
+substantial latency penalties: over WAN, 10 to 100 times higher than
+their HAT counterparts. In light of recent algorithms for providing
+causally consistent get transactions~\cite{bolton, cops}, we focus our
+HAT analysis on TA, which we believe to be the most novel of our
+algorithms. We find that HAT properties---while more
+resource-intensive than basic eventual consistency---scale well to
+many servers.
 
@@
 by LevelDB and implemented in Java using Apache Thrift. It currently
-supports eventual consistency (last-writer-wins RU, via simple
-anti-entropy) and HAT RC and TA guarantees (as described in
-Section~\ref{sec:hats}). It also supports non-HAT operation whereby
-all operations for a given key are routed to a (randomly) designated
-\textit{master} replica for each key (i.e., PNUTS~\cite{pnuts}) as
-well as distributed two-phase locking. Servers are durable: they
-synchronously write to LevelDB before responding to client requests,
-while pending writes in TA are sent to a write-ahead log, which is
-synchronously flushed to disk.
+supports eventual consistency (last-writer-wins RU with standard
+all-to-all anti-entropy between replicas) and HAT RC and TA guarantees
+(as described in Section~\ref{sec:hats}). It also supports non-HAT
+operation whereby all operations for a given key are routed to a
+(randomly) designated \texttt{master} replica for each key (i.e.,
+PNUTS~\cite{pnuts}) as well as distributed two-phase locking. Servers
+are durable: they synchronously write to LevelDB before responding to
+client requests, while pending writes in TA are sent to a write-ahead
+log, which is synchronously flushed to disk.
 
@@
 instances as servers in each cluster. For our workload, we link our
-client library to the YCSB benchmark~\cite{ycsb}, grouping every eight
-YCSB operations from the default workload (50\% reads, 50\% writes) to
-form a transaction. We increase the number of keys in the workload
-from the default $1,000$ to $100,000$ with uniform random key access,
-keeping the default value size of $1KB$, and running YCSB for 180
-seconds per configuration.
+client library to the YCSB benchmark~\cite{ycsb}, which is well-suited
+to LevelDB's key-value schema, grouping every eight YCSB operations
+from the default workload (50\% reads, 50\% writes) to form a
+transaction. We increase the number of keys in the workload from the
+default $1,000$ to $100,000$ with uniform random key access, keeping
+the default value size of $1KB$, and running YCSB for 180 seconds per
+configuration.
 
@@
  number of YCSB client threads, mastered operation has substantially
- lower throughput than the HAT configurations. Increasing the number of YCSB
- clients \textit{does} increase the throughput of mastered operation,
- but our Thrift-based connection processing did not gracefully handle
- more than several thousand concurrent connections. In contrast,
- across two datacenters, the performance of eventual, RC, and TA is
- near-identical to a single-datacenter deployment.
+ lower throughput than the HAT configurations. Increasing the number
+ of YCSB clients \textit{does} increase the throughput of mastered
+ operation, but our Thrift-based server-side connection processing did
+ not gracefully handle more than several thousand concurrent
+ connections. In contrast, across two datacenters, the performance of
+ eventual, RC, and TA is near-identical to a single-datacenter
+ deployment.
 
@@
 substantially improve performance due to the network topology and
-because worst-case server load was unaffected. In this configuration,
+because worst-case server load was unaffected. With five clusters,
 TA's relative throughput decreased: every YCSB \texttt{put} operation
@@
 increased garbage collection activity and, more importantly, IOPS when
-compared to Eventual and RC, causing throughput to peak at around half
-of Eventual. With an in-memory persistence (i.e., no LevelDB or WAN),
+compared to Eventual and RC, causing TA throughput to peak at around half
+of Eventual. With an in-memory persistence (i.e., no LevelDB or WAL),
 TA throughput was within 20\% of eventual.
@@
 
-\vspace{.5em}\noindent\textit{Read proportion.} The default equal
+\vspace{.5em}\noindent\textit{Read proportion.} Our default (equal)
 proportion of reads and writes is fairly pessimistic: for example,
@@
 shown in Figure~\ref{fig:rprop} (clusters in Virginia and Oregon),
-with all reads, TA is within $4.8\%$ of eventual; with all writes,
-eventual throughput decreases by $288.8\%$ compared to all reads and
-TA is within $33\%$ of eventual. At $99.8\%$ reads, TA incurs a $7\%$
-overhead ($5.8\%$ for in-memory storage).
+with all reads, TA is within $4.8\%$ of eventual; with all writes, TA
+is within $33\%$, and Eventual throughput decreases by $288.8\%$
+compared to all reads and TA is within $33\%$ of eventual. At $99.8\%$
+reads, TA incurs a $7\%$ overhead ($5.8\%$ for in-memory storage).
 
@@
 database (instead of LevelDB), TA scales by $4.25$x (not shown)---and
-increased garbage collection activity.
+TA-related performance heterogeneity across servers (Calvin's authors
+report similar heterogeneity on EC2~\cite{calvin}). Nevertheless, this
+linear scalability is not typical of traditional database system
+designs~\cite{gray-isolation}.
 

### 2013-05-01 writeup/hat-analysis.tex
more writin

@@
 To begin, \textbf{Read Uncommitted} isolation is captured by Adya as
-\textit{PL-1}, requiring only that each transaction's writes are ordered
-consistently according to a single total order on transactions. This
-prohibits Adya's phenomenon $G0$, also called ``Dirty
-Writes''~\cite{adya}. More formally, we require a total ordering on
-writes to each data item and we create a graph of transactions with
-edges from one transaction to another if the former overwrites the
-latter's write to the same object. \textit{PL-1} prohibits cycles in
-this graph, requiring some total order on writes. Accordingly, the
-final database state cannot contain the ``earlier'' transaction's
-writes to the data items: for example, in the below example, $T_3$
-should \emph{eventually} only read $a=b=1$ or $a=b=2$ but not $a=2,
-b=1$ or $a=1, b=2$:
+\textit{PL-1}. In this model, writes to each object are totally
+ordered, corresponding to the order in which they are installed in the
+database. In a distributed database, different replicas may receive
+writes to their local copies of data at different times, but the order
+in which replicas overwrite versions should be consistent with the total order for
+each item. \textit{PL-1} requires that writes to different objects are
+ordered consistently across transactions, prohibiting Adya's
+phenomenon $G0$, also called ``Dirty Writes''~\cite{adya}. If we
+create a graph of transactions with edges from one transaction to
+another if the former overwrites the latter's write to the same
+object, then, under Read Uncommitted, there should be no cycles in
+this graph. Consider the following example:
 \begin{align*}
@@
 \\T_2 &: w_x(2)~w_y(2)
-\\T_3 &: r_x(a)~r_y(b)\vspace{-1em}
 \end{align*}
-Later properties will strengthen Read Uncommitted.
-
-Unlike the total order required by serializability, the order in PL-1
-does not constrain the values seen by a transaction's read
-operations except once the database has reached a ``final
-state.'' Read Uncommitted is easily achieved via applying the same
-logical timestamp to each update in a transaction and applying a
-``last writer wins'' conflict reconciliation policy at each replica.
+In this example, under Read Uncommitted, it is impossible for the
+database to order $T_1$'s $w_x(1)$ before $T_2$'s $w_x(2)$ but order
+$T_2$'s $w_y(2)$ before $T_1$'s $w_y(1)$.  Read Uncommitted is easily
+achieved via applying the same logical timestamp to each update in a
+transaction and applying a ``last writer wins'' conflict
+reconciliation policy at each replica. Later properties will
+strengthen Read Uncommitted.
 
@@
 transaction $T_i$ are observed by another transaction $T_j$,
-thereafter all effects of $T_i$ are observed by $T_2$. Together with
+thereafter, all effects of $T_i$ are observed by $T_j$. Together with
 item cut isolation, TA prevents Read Skew anomalies (Berenson et al.'s
@@
 \small\vspace{-1em}
-T_1 &: r_x(a) w_x(a+2)
+T_1 &: r_x(a)~w_x(a+2)
 \\T_2 &: w_x(2)\vspace{-1em}

### 2013-05-01 writeup/hat-appendix.tex
more writin

@@
 the replica for $y$ will know to check \texttt{pending}.
+
+\begin{algorithm}[t!]
+\newcommand{\myindent}{\hspace{-1em}}
+\begin{addmargin}[-1em]{0em}% 1em left, 2em right
+
+\begin{algorithmic}
+\caption*{\textbf{Algorithm} HAT Transactional Atomicity}\vspace{.5em}
+\State{\textit{\textbf{Shared Data Types}}\\\vspace{.5em}
+\texttt{timestamp}: unique, per transaction identifier\\
+$\texttt{write}:[\texttt{key:}k, \texttt{value}:v, \texttt{timestamp}:ts, \texttt{set<key>}:tx\_keys]$\\\hrulefill}\vspace{.5em}
+
+  \State{\textbf{\textit{Server-side Data Structures and Methods}\\\vspace{.5em}
+$\texttt{set<write>:pending}$\\
+$\texttt{set<write>:good}$\\
+$\texttt{map<timestamp, int>:acks}$}}\\
+
+\Procedure{put}{\texttt{write}:$w$}
+  \State \texttt{pending}.add($w$)
+  \For{key $k_t \in w.tx\_keys$}
+  \State \{all replicas for $sib$\}.notify($w.ts$)
+  \EndFor
+
+  \State \textit{asynchronously send $w$ to other replicas via anti-entropy}
+  \State \Return
+
+  \EndProcedure\vspace{.5em}
+
+  \Procedure{notify}{\texttt{timestamp}:$ts$}
+  \State \texttt{acks}.get($ts$).increment()
+  \If{all \texttt{acks} received for all replicas for $ts$'s $tx\_keys$}
+  \State \texttt{good}.add($w$)
+  \State \texttt{pending}.remove($w$)  
+  \EndIf  
+  \EndProcedure\vspace{.5em}
+
+  \Procedure{get}{\texttt{key}:$k$, \texttt{timestamp}:$ts_{required}$}
+  \If{$ts_{required} = \bot$}
+  \State \Return $w \in \texttt{good}$ s.t. $w.key = key$ with highest timestamp
+  \ElsIf{$\exists~w \in \texttt{good}$ s.t. $w.key = key, w.ts \geq ts_{required}$}
+  \State \Return $w$
+  \Else
+  \State \Return $w \in \texttt{pending}$ s.t. $w.key = key,  w.ts = ts_{required}$
+  \EndIf
+  \EndProcedure\\\vspace{-.5em}\hrulefill\vspace{.25em}
+
+  \State{\textbf{\textit{Client-side Data Structures and Methods}\\\vspace{.5em}
+$\texttt{int:cur\_txn}$\\
+$\texttt{map<key, value>:write\_buffer}$\\
+$\texttt{map<key, timestamp>:required}$}}\\
+
+\Procedure{begin\_transaction}{}
+  \State \texttt{cur\_txn} = \textit{new txn ID} \commentt{// (e.g., clientID+logicalClock)}
+  \EndProcedure\vspace{.5em}
+
+\Procedure{put}{\texttt{key:$k$}, \texttt{value:$v$}}
+  \State \texttt{write\_buffer}.put($k$, $v$)
+\EndProcedure\vspace{.5em}
+
+\Procedure{get}{\texttt{key:$k$}}
+  \If {$k \in \texttt{write\_buffer}$}
+  \State \Return \texttt{write\_buffer}.get($k$) \commentt{// if we want per-TxN RYW}
+  \EndIf
+
+  \State $w\_{ret}$ = (available replica for $k$).get($k$, \texttt{required}.get($k$))
+  \For {$(tx_{key} \in w_{ret}.tx\_keys$}
+  \If {$w_{ret}.ts > \texttt{required}.get(tx_{key})$}
+  \State $\texttt{required}$.put$(tx_{key}, w_{ret}.ts)$
+  \EndIf
+  \EndFor
+
+  \State \Return $w\_{ret}.value$
+\EndProcedure\vspace{.5em}
+
+\Procedure{commit}{}
+ \For{$(tx_{key}, v) \in \texttt{write\_buffer}$}
+ \State $r$ = (available replica for $tx_{key}$)
+ \State $r$.put($[tx_{key}, v, \texttt{cur\_txn}, \texttt{write\_buffer}.keys()]$)
+ \EndFor
+  \State \textsc{cleanup()}
+\EndProcedure\vspace{.5em}
+
+\Procedure{abort}{}
+  \State \textsc{cleanup()}
+\EndProcedure\vspace{.5em}
+
+\Procedure{cleanup}{}
+  \State \texttt{write\_buffer}.clear()
+  \State \texttt{required\_map}.clear()
+\EndProcedure\vspace{.5em}
+
+\end{algorithmic}
+\end{addmargin}
+\end{algorithm}

### 2013-05-01 writeup/evaluation.tex
aaron, ali, alan comments

@@
 on preventing write conflicts. Chan and Gray's read-only transactions
-provide read-only transactions with item-cut isolation, causal
-consistency, and transactional atomicity (session PL-2L~\cite{adya})
-but are unavailable in the presence of coordinator
-failure~\cite{readonly}, similar to read-only and write-only
-transactions more recently proposed by Eiger~\cite{eiger}. Causal
-Serializability offers a similar model (with unavailable
-implementation): causal consistency with a variant of Read Uncommitted
-between transactions that write to the same data
-item~\cite{raynal-causal}.  Brantner's S3 database~\cite{kraska-s3}
-and Bayou~\cite{sessionguarantees} can all provide variants of session
-PL-2L with high availability, but none provide this HAT functionality
-without substantial modification. Swift~\cite{swift} and bolt-on
-causal consistency~\cite{bolton} are closest to providing maximum
-sticky HAT semantics. As we have seen, it is possible to implement many
+have item-cut isolation with causal consistency and TA (session
+PL-2L~\cite{adya}) but are unavailable in the presence of coordinator
+failure and assume serializable update transactions~\cite{readonly};
+this is similar to read-only and write-only transactions more recently
+proposed by Eiger~\cite{eiger}. Causal Serializability offers a
+similar model (with unavailable implementation): causal consistency
+with a variant of Read Uncommitted between transactions that write to
+the same data item~\cite{raynal-causal}.  Brantner's S3
+database~\cite{kraska-s3} and Bayou~\cite{sessionguarantees} can all
+provide variants of session PL-2L with high availability, but none
+provide this HAT functionality without substantial
+modification. Swift~\cite{swift} and bolt-on causal
+consistency~\cite{bolton} are closest to providing maximum sticky HAT
+semantics. As we have seen, it is possible to implement many
 guarantees weaker than serializability---including guarantees that are

### 2013-05-01 writeup/hat-analysis.tex
aaron, ali, alan comments

@@
 ANSI SQL specification~\cite{ansi-sql} and Berenson et al.'s
-subsequent 1995 critique~\cite{ansicritique}. We provide a set of
-formal definitions, semantics, proofs and observations in our extended
-Technical Report~\cite{hat-tr} and opt for a more informal
-presentation here.
+subsequent 1995 critique~\cite{ansicritique}.
+
+For brevity, we provide an informal presentation of each guarantee
+here (accompanied by appropriate references) but give a full set of
+formal definitions in our extended Technical Report~\cite{hat-tr}. In
+our examples, we exclusively consider read and write operations,
+denoting a write of value $v$ to data item $d$ as $w_d(v)$ and a read
+from data item $d$ returning $v$ as $r_d(v)$. We assume that all data
+items have the null value, $\bot$, at database initialization, and,
+unless otherwise specified, all transactions in the examples commit.
 
@@
 in Section~\ref{sec:evaluation}.
-
-In our examples, we exclusively consider read and write operations,
-denoting a write of value $v$ to data item $d$ as $w_d(v)$ and a read
-from data item $d$ returning $v$ as $r_d(v)$. We assume that all data
-items have the null value, $\bot$, at database initialization, and,
-unless otherwise specified, all transactions in the examples commit.
 
@@
 $T_2$'s $w_y(2)$ before $T_1$'s $w_y(1)$.  Read Uncommitted is easily
-achieved via applying the same logical timestamp to each update in a
-transaction and applying a ``last writer wins'' conflict
+achieved via timestamping each update in a transaction with the same
+logical timestamp (which is unique across transactions; e.g., client
+ID and logical clock) and applying a ``last writer wins'' conflict
 reconciliation policy at each replica. Later properties will
@@
 
-\textbf{Repeatable Read} isolation is a confusing property. For
-example, some IBM products use the term for fully serializable
-isolation~\cite{hat-hotos}, while Gray~\cite{gray-isolation}, Berenson
-et al.~\cite{ansicritique}, and Adya's PL-2.99~\cite{adya} all
-interpret Repeatable Read as providing serializability for all
-operations except for predicate-based reads (and so Repeatable Read is
-identical to serializability in a key-value store without
-predicate-based access). In models like Adya's PL-2.99, the
-\textit{Phantom Problem}, whereby two successive predicate-based reads
-return different data~\cite{gray-isolation}, is the only kind of
-non-serializable behavior allowed. In this paper, we reserve the use
-of ``Repeatable Read'' to refer to these definitions, which we will
-soon show are not HAT-compliant.
-
-However, as Berenson et al.~\cite{ansicritique} discuss, the ANSI SQL
-specification allows several additional behaviors under ``Repeatable
-Read'' isolation.  \textbf{ANSI Repeatable Read} requires that, along
-with respecting Read Committed isolation, each transaction will only
-read one version of each data item that it did not itself produce
-(preventing ``Fuzzy Read,'' or P2). In the example below, $T_3$ must
-read $a=1$:
+Several different properties have been labeled \textbf{Repeatable
+  Read} Isolation. As we will show in
+Section~\ref{sec:unachievable-acid}, some of these are not achievable
+in a HAT system. However, the ANSI standardized
+implementation-agnostic definition~\cite{ansi-sql} \textit{is}
+achievable and directly captures directly the spirit of the term: if a
+transaction reads the same data twice, it sees the same value each
+time (preventing ``Fuzzy Read,'' or $P2$). In this paper, to
+disambiguate between other definitions of ``Repeatable Read,'' we will
+call this property ``cut isolation,'' since each transaction reads
+from a non-changing cut, or snapshot, over the data items. If this
+property holds over reads from discrete data items, we call it
+\textbf{Item Cut Isolation (I-CI)}, and if we also expect a cut over
+predicate-based reads (preventing Phantoms~\cite{gray-isolation}, or
+Berenson et al.'s $P3/A3$~\cite{ansicritique}), we have the stronger
+property of \textbf{Predicate Cut-Isolation}. In the example below,
+under both levels of cut isolation, $T_3$ must read $a=1$:
 \begin{align*}
@@
 \end{align*}
-This is a literal interpretation of the phrase ``Repeatable Read'':
-unless a transaction modifies a given data item, the observable value
-of the data item should not change during the transaction. By itself,
-this property is weak---we can always read $\bot$ for each data
-item---and easily achieved in a HAT system by caching the values
-read. To prevent confusion, we call this isolation model \textbf{Item
-  Cut Isolation}: all the transaction's reads should see values from a
-non-changing consistent cut or snapshot over the data items. It is
-possible to satisfy Item Cut Isolation with high availability by using
-multi-versioning on each replica and ensuring that each of a
-transaction's reads return an appropriately chosen version (e.g.,
-identified by vector clock).
-
-A stronger achievable property that expands on Item Cut Isolation
-requires consistent cuts to span both a transaction's predicate reads
-(e.g., \texttt{SELECT WHERE}) as well as its item reads.  We call this
-\textbf{Predicate Cut Isolation} and it prohibits Phantoms (but does
-not guarantee serializability of item-based reads and
-writes). Predicate Cut Isolation is also achievable in HAT systems via
-similar multi-versioning as above.
+It is possible to satisfy Item Cut Isolation with high availability by
+having clients cache their reads such that the value that they read
+never changes unless they overwrite it themselves. The cache can be
+cleared at the end of each transaction and can alternatively be
+accomplished on (sticky) replicas via multi-versioning. Predicate Cut
+Isolation is also achievable in HAT systems via similar caching or
+multi-versioning that tracks predicate ranges instead of item-based
+reads.
 
@@
 bolt-on causal consistency~\cite{bolton}. These systems also provide
-causal consistency, which requires sticky availability (in each,
+causal consistency and are subsequently sticky available (each is
 implemented via client-side caching). However, TA is achievable
@@
 favor of sticky availability. Sticky availability permits three
-additional models, which we first define and then prove are
+additional guarantees, which we first define and then prove are
 unachievable in a generic highly available system:
@@
 the rest of the other servers, the server must allow $T_1$ to
-commit. If the client subsequently executes $T_2$ against the same (partitioned)
-server, then it will be able to read its writes. However, if the
-network topology shifts and the client can only contact a different
-server that is partitioned from the server that executed $T_1$, then
-the client will be unable to read its own writes and the system will
-have to either stall indefinitely to allow the client to read her
-writes (violating transactional availability) or will have to
-sacrifice read your writes guarantees. However, if the client remains
-sticky with the server that executed $T_1$, then we can disallow this
-scenario. Accordingly, read your writes, and, by proxy, causal
-consistency and PRAM require stickiness. Read your writes is provided
-by default in a sticky system, while the remaining causality and PRAM
-guarantees can be accomplished with the algorithms for achieving the
-remaining session guarantees.
+commit. If the client subsequently executes $T_2$ against the same
+(partitioned) server, then it will be able to read its
+writes. However, if the network topology shifts and the client can
+only contact a different server that is partitioned from the server
+that executed $T_1$, then the client will be unable to read its own
+writes and the system will have to either stall indefinitely to allow
+the client to read her writes (violating transactional availability)
+or will have to sacrifice read your writes guarantees. However, if the
+client remains sticky with the server that executed $T_1$, then we can
+disallow this scenario. Accordingly, read your writes, and, by proxy,
+causal consistency and PRAM require stickiness. Read your writes is
+provided by default in a sticky system. Causality and PRAM guarantees
+can be accomplished with well-known variants~\cite{causalmemory,
+  bolton, cops, sessionguarantees, swift} of the prior session
+guarantee algorithms we presented earlier.
 
@@
 consistency properties that are achievable in HAT systems (see
-Section~\ref{sec:futurework}) but we specifically evaluate TPC-C
+Section~\ref{sec:relatedwork}) but we specifically evaluate TPC-C
 transaction semantics under HAT consistency guarantees in
@@
 transactions' effects survive $F$ server faults requires at least
-$F+1$ non-failing replicas.
+$F+1$ non-failing replicas; this affects availability.
 
@@
 \subsubsection{Unachievable ACID Isolation}
+\label{sec:unachievable-acid}
 
@@
 Isolation, and Cursor Stability guarantees are all unavailable.
-Repeatable Read (defined by Gray, Berenson et al., and Adya) and
-One-Copy Serializability need to prevent both Lost Update and Write
-Skew. These prevention requirements mean that these models are
-inherently unavailable.
+Repeatable Read (defined by Gray~\cite{gray-isolation}, Berenson et
+al.~\cite{ansicritique}, and Adya~\cite{adya}) and One-Copy
+Serializability need to prevent both Lost Update and Write Skew. These
+prevention requirements mean that these guarantees are inherently
+unavailable.
 
@@
 writes guarantees and PRAM and causal consistency. However, many other
-prominent models, such as Snapshot Isolation, One-Copy
+prominent semantics, such as Snapshot Isolation, One-Copy
 Serializability, and Strict Serializability cannot be achieved due to
@@
 collection), or use pre-declared read sets to fetch a consistent cut
-of keys before each transaction begins to execute. Using client-side
-caching can alleviate some of these challenges~\cite{bolton, swift},
-but then the system becomes sticky high available.
+of keys before each transaction begins to execute (e.g.,
+COPS~\cite{cops}). Using client-side caching can alleviate some of
+these challenges~\cite{bolton, swift}, but then the system becomes
+sticky high available.
 

### 2013-05-01 writeup/hat-appendix.tex
aaron, ali, alan comments

@@
 
-\begin{definition}[Directly Session-Depends by $S$]
+\begin{definition}[Session-Depends]
 \label{def:sd1}
-A transaction $T_j$ directly session-depends on transaction $T_i$ if
-$T_i$ immediately precedes $T_j$ in the session commit order for session $S$.
-\end{definition}
-
-\begin{definition}[Directly Session-Depends]
-\label{def:sd2}
-A transaction $T_j$ directly session-depends on transaction $T_i$ if
-$T_j$ directly session-depends on transaction $T_i$ for some session
-$S$.
+A transaction $T_j$ session-depends on transaction $T_i$ if $T_i$ and
+$T_j$ occur in the same session and $T_i$ precedes $T_j$ in the
+session commit order.
 \end{definition}
@@
 The dependencies for a history $H$ form a graph called its Directed
-Serialization Graph ($DSG(H)$).
+Serialization Graph ($DSG(H)$). If $T_j$ directly write-depends on
+$T_i$ by $x$, we draw $T_i\overset{ww_x}\longrightarrow T_j$. If $T_j$
+read-depends on $T_i$ by $x$, we draw
+$T_i \overset{rw_x}\longrightarrow T_j$. If $T_j$ directly
+anti-depends on transaction $T_j$ by $x$, we draw
+$T_i \overset{wr_x}\DashedArrow T_j$. If $T_j$ session-depends on
+$T_i$ in session $S$, we draw $T_i \overset{S}\longrightarrow T_j$.
 
@@
 
-\begin{definition}[Item Cut Non-Isolation (N-I-CI)]
+\begin{definition}[Item-Many-Preceders (N-I-CI)]
 \label{def:n-ici}
 A history $H$ exhibits phenomenon N-ICI if $DSG(H)$ contains a
-transaction $T_i$ such that $T_i$ directly item-depends by $x$ on more
+transaction $T_i$ such that $T_i$ directly item-read-depends by $x$ on more
 than one other transaction.
@@
 
-\begin{definition}[Predicate Cut Non-Isolation (N-P-CI)]
-A history $H$ exhibits phenomenon N-I-CI if $DSG(H)$ contains a
-transaction $T_i$ such that, for two predicate-based reads $p_k$,
-$p_j$ in $P$ such that $p_k \subseteq p_j$ and $T_i$ directly
-predicate-read-depends by $p_k$ on another transaction $T_l$, but
-$T_i$ does not directly predicate-read-depend by $p_j$ on $T_l$.
+\begin{definition}[Predicate-Many-Preceders (N-P-CI)]
+A history $H$ exhibits phenomenon N-I-CI if, for all predicate-based
+reads $r_i(P_i:Vset(P_i))$ and $r_j(P_j:Vset(P_j)$ in $T_k$ such that
+the logical ranges of $P_i$ and $P_j$ overlap (call it $P_o$), the set
+of transactions that change the matches of $P_o$ for $r_i$ and $r_j$
+differ.
 \end{definition}
@@
 
-\begin{definition}[Non-transactional Atomicity (N-TA)]
+\begin{definition}[Observed Transaction Vanishes (N-TA)]
 A history $H$ exhibits phenomenon N-TA if $DSG(H)$ contains a directed cycle
@@
 $T_j$ and one anti-dependency edge by $y$ from $T_j$ to $T_i$ and
-$T_i$'s corresponding read from $x$ does not precede its corresponding
-read from $y$.
+$T_i$'s read from $y$ precedes its read from $x$.
 \end{definition}
@@
 \end{definition}
+
+The following session guarantees are directly adapted from Terry et
+al.'s original definitions~\cite{sessionguarantees}:
 
@@
 
-\begin{definition}[Not-Read Your Writes (N-RYW)]
+\begin{definition}[Missing Your Writes (N-RYW)]
 A history $H$ exhibits phenomenon N-RYW if $DSG(H)$ contains a directed cycle

### 2013-05-01 writeup/hat-vldb.tex
some graphs and joe comments
eval rewrite
More tweaks
more writing and polishing
more writin
aaron, ali, alan comments

@@
 \let\endproof\relax
-\usepackage{draftwatermarkbottom, draftwatermarktop}
+\usepackage{draftwatermarkbottom, draftwatermarktop, scrextend, MnSymbol}
 
-\newcommand{\pbnote}[1]{{\color{red}{#1 ---P.B.}}\\}
+\newcommand{\pbnote}[1]{{\color{red}{#1 ---P.B.}}}
 
@@
 \usepackage{balance}  % for  \balance command ON LAST PAGE  (only there!)
-\usepackage{xspace}
+\newcommand{\subparagraph}{}
+
+\usepackage{xspace, titlesec}
 
@@
 \newcommand{\eg}{{\em e.g.,}\xspace}
+
+\newcommand{\figfactor}{1.0}
+
+\usepackage{algorithm}% http://ctan.org/pkg/algorithms
+\usepackage[noend]{algpseudocode}% http://ctan.org/pkg/algorithmicx
+\newcommand{\commentt}[1]{{\small\texttt{#1}}}
 
@@
 \tikzset{>=latex}
+
+\titleformat*{\section}{\fontsize{12pt}{12}\selectfont\bfseries}
+\titleformat*{\subsection}{\fontsize{10pt}{11}\selectfont\bfseries}
+\titleformat*{\subsubsection}{\fontsize{10pt}{11}\itshape}
+
+\titlespacing{\section}{1ex}{2ex}{1ex}
+
+\titlespacing{\subsection}{1ex}{1ex}{0.25ex}
+\titlespacing{\subsubsection}{1ex}{1ex}{0.25ex}
+
+\let\footnotesize\small
+
+\newcommand*{\DashedArrow}[1][]{\mathbin{\tikz [baseline=-.25ex,-latex, dashed,#1] \draw [#1] (0pt,0.5ex) -- (2em,0.5ex);}}%
 
@@
 analyze existing ACID and distributed consistency guarantees to
-identify which can and cannot be achieved in HAT systems. We
-analytically and experimentally quantify the availability and
-performance benefits of HATs---often several orders of magnitude over
-wide-area networks---as well as their necessary semantic compromises.
-\end{abstract}
+identify which can and cannot be achieved in HAT systems. This unifies
+the literature on weak isolation, data consistency, and highly
+available systems. We analytically and experimentally quantify the
+availability and performance benefits of HATs---often several orders
+of magnitude over wide-area networks---as well as their necessary
+semantic compromises.
+\end{abstract}\vspace{-.5em}
 
@@
 
-\input{futurework.tex}
+%\input{futurework.tex}
 
@@
 \scriptsize
-\balance
 

### 2013-05-01 writeup/model.tex
aaron, ali, alan comments

@@
 
-Traditionally, a system provides {\em availability} if every user that
-can contact a correct server eventually receives a response from that
-server, even in the presence of arbitrary, indefinitely long network
-partitions between servers~\cite{gilbert-cap}.\footnote{Under this
-  definition, systems that require a majority of servers to be online
-  or that require the database to function correctly with high (but
-  not perfect) probability are not available.} As in a standard
-distributed database, designated servers might perform operations for
-different data items; a server that can handle an operation for a
-given data item is called a \textit{replica} for that
-item.\footnote{There is a further distinction between a \textit{fully
-    replicated} system, in which all servers are replicas for all data
-  items, and a \textit{partially replicated} system, in which at least
-  one server acts as a replica for only a subset of all data
-  items. For generality, and, for applicability to many modern
-  ``sharded'' or ``partitioned'' data storage systems~\cite{dynamo,
-    pnuts, bigtable, spanner, hstore}, we consider partially
-  replicated systems.} While this standard model is useful, there are
-several other cases we should consider.
+Traditionally, a system provides {\em high availability} if every user
+that can contact a correct (non-failing) server eventually receives a
+response from that server, even in the presence of arbitrary,
+indefinitely long network partitions between
+servers~\cite{gilbert-cap}.\footnote{Under this definition, systems
+  that require a majority of servers to be online or that require the
+  database to function correctly with high (but not perfect)
+  probability are not available.} As in a standard distributed
+database, designated servers might perform operations for different
+data items; a server that can handle an operation for a given data
+item is called a \textit{replica} for that item.\footnote{There is a
+  further distinction between a \textit{fully replicated} system, in
+  which all servers are replicas for all data items, and a
+  \textit{partially replicated} system, in which at least one server
+  acts as a replica for only a subset of all data items. For
+  generality, and, for applicability to many modern ``sharded'' or
+  ``partitioned'' data storage systems~\cite{dynamo, pnuts, bigtable,
+    spanner, hstore}, we consider partially replicated systems.} While
+this standard model is useful, there are several other cases we should
+consider.
 

### 2013-05-01 writeup/hat-analysis.tex
PDB pass

@@
 overlapping (with distinct vocabulary)''~\cite{brewer-slides}. With
-this challenge in mind, when possible, we build on existing properties
-and definitions from the database and distributed systems literature,
+this challenge in mind, we build on existing properties and
+definitions from the database and distributed systems literature,
 providing a brief, informal explanation and example for each
@@
 ANSI SQL specification~\cite{ansi-sql} and Berenson et al.'s
-subsequent 1995 critique~\cite{ansicritique}.
+subsequent critique~\cite{ansicritique}.
 
@@
 achieved via timestamping each update in a transaction with the same
-logical timestamp (which is unique across transactions; e.g., client
-ID and logical clock) and applying a ``last writer wins'' conflict
-reconciliation policy at each replica. Later properties will
-strengthen Read Uncommitted.
+logical timestamp (which is unique across transactions; e.g.,
+combining a client ID and logical clock) and applying a ``last writer
+wins'' conflict reconciliation policy at each replica. Later
+properties will strengthen Read Uncommitted.
 
@@
 It is fairly easy for a HAT system to prevent ``Dirty Reads'': if each
-client never writes uncommitted data to shared copies of the database, then
+client never writes uncommitted data to shared copies of data, then
 transactions will never read each others' dirty data. As a simple
@@
 \textbf{Item Cut Isolation (I-CI)}, and if we also expect a cut over
-predicate-based reads (preventing Phantoms~\cite{gray-isolation}, or
-Berenson et al.'s $P3/A3$~\cite{ansicritique}), we have the stronger
-property of \textbf{Predicate Cut-Isolation}. In the example below,
-under both levels of cut isolation, $T_3$ must read $a=1$:
+predicate-based reads (e.g., \texttt{SELECT WHERE}; preventing
+Phantoms~\cite{gray-isolation}, or Berenson et al.'s
+$P3/A3$~\cite{ansicritique}), we have the stronger property of
+\textbf{Predicate Cut-Isolation}. In the example below, under both
+levels of cut isolation, $T_3$ must read $a=1$:
 \begin{align*}
@@
 Isolation is also achievable in HAT systems via similar caching or
-multi-versioning that tracks predicate ranges instead of item-based
-reads.
+multi-versioning that tracks entire logical ranges of predicates
+in addition to item-based reads.
 
@@
 for each key):
-\vspace{-1em}
 \begin{align*}
@@
 technical report~\cite{hat-tr}. Each replica waits to reveal write $w$
-from transaction $T$ until all of the writes in $T$ are present on all
-respective replicas (are \textit{pending stable}). Once they are,
-replicas asynchronously reveal their writes. To prevent clients from
-reading writes that have been revealed on some servers but not others,
-clients can attach a vector that uniquely identifies the transaction:
-this can be the transaction ID and a list of keys in the
-transaction. When clients attempt to read a data item from a replica,
-they can send the highest ID they have observed for that item;
-replicas respond with either a matching write (which may or may not be
-pending stable) or a write with higher ID that is pending
-stable. Clients can reset their record of IDs and keys upon
-transaction commit. Servers keep two sets of versions for each data
-item: the value with the highest ID that is pending stable and a set
-of values that are not yet pending stable.
+from transaction $T$ until all of the writes from $T$ are present on
+all respective replicas (are \textit{pending stable}). Once they are,
+replicas asynchronously reveal their respective writes to readers. To
+prevent clients from reading incomplete portions of a transaction's
+writes as they are being revealed (e.g., a client reads a data item
+that from a replica that has detected pending stability, then attempts
+to read another item from another replica which has not yet detected
+pending stability), clients include additional metadata with each
+write: a list of items in the transaction along with the transaction's
+timestamp (from Read Uncommitted). When a client reads a write, the
+write's list of items and timestamp form a lower bound on the versions
+that the client should read for other items. When a client performs
+subsequent reads, it includes a timestamp with its request that
+corresponds current lower bound on reads from the respective item
+Replicas respond with either a write matching the timestamp or a write
+with a higher timestamp that is pending stable. Clients discard any
+metadata upon transaction commit, and servers keep two sets of writes
+for each data item: the write with the highest timestamp that is
+pending stable and a set of writes that are not yet pending stable.
 
-This non-sticky implementation is entirely masterless and neither reads
-nor writes block for coordination. Moreover, there are several
+This non-sticky implementation is entirely masterless and neither
+reads nor writes block for coordination. Moreover, there are several
 possible optimizations that, for brevity, we do not describe
 here. However, one key property is that we do not, in general, dictate
-when writes become visible. Accordingly, it is highly
-available. However, we can optimize write visibility delay by ensuring
-that clients are sticky with disjoint groups of servers. Once all
-transaction values are \texttt{pending} within a group of servers,
-values can be made \texttt{stable} (as opposed to once they are
-present on all servers). For example, in a multi-datacenter setting,
-all clients and servers within a datacenter may form a sticky group.
+when writes become visible. Accordingly, the algorithm is highly
+available. We can optimize write visibility by sticking clients are
+with disjoint groups of servers: once all of a transaction's writes
+are pending stable within a group of servers, writes can be made
+visible (as opposed to once they are pending stable with respect to
+all servers). For example, in a multi-datacenter setting, all clients
+and servers within a datacenter may form a sticky group.
 
@@
 
-\vspace{.5em}\noindent{\textbf{Durability}} A client requiring that its
-transactions' effects survive $F$ server faults requires at least
-$F+1$ non-failing replicas; this affects availability.
+%\vspace{.5em}\noindent{\textbf{Durability}} A client requiring that its
+%transactions' effects survive $F$ server faults requires at least
+%$F+1$ non-failing replicas; this affects availability.
 

### 2013-05-01 writeup/hat-vldb.tex
PDB pass

@@
 \begin{abstract}
-To minimize network latency and remain online during server failures
-and network partitions, many modern distributed data storage systems
-eschew transactional functionality, which provides strong semantic
-guarantees for groups of multiple operations over multiple data
-items. In this work, we consider the problem of providing Highly
+\noindent To minimize network latency and remain online during server
+failures and network partitions, many modern distributed data storage
+systems eschew transactional functionality, which provides strong
+semantic guarantees for groups of multiple operations over multiple
+data items. In this work, we consider the problem of providing Highly
 Available Transactions (HATs): transactional guarantees that do not
@@
 available systems. We analytically and experimentally quantify the
-availability and performance benefits of HATs---often several orders
-of magnitude over wide-area networks---as well as their necessary
+availability and performance benefits of HATs---two to three orders of
+magnitude over wide-area networks---as well as their necessary
 semantic compromises.

### 2013-05-01 writeup/intro.tex
PDB pass

@@
 shows that, besides serializability, Snapshot Isolation and Repeatable
-Reads are not HAT-compliant, while most other isolation levels are
-achievable with high availability. We also demonstrate that many weak
-data consistency models from distributed systems are also both
-HAT-compliant and simultaneously achievable with traditional ACID
+Read isolation are not HAT-compliant, while most other isolation
+levels are achievable with high availability. We also demonstrate that
+many weak data consistency models from distributed systems are also
+both HAT-compliant and simultaneously achievable with traditional ACID
 properties.
@@
 Our investigation is based on both impossibility results and several
-constructive proof-of-concept algorithms. For example, Snapshot
-Isolation (SI) and Repeatable Reads (RR) are not HAT-compliant because
-they require detecting conflicts between concurrent updates (as needed
-for preventing Lost Updates or Write Skew phenomena), which we prove
-is unavailable. However, Read Committed isolation, Transactional
+constructive, proof-of-concept algorithms. For example, Snapshot
+Isolation and Repeatable Read isolation are not HAT-compliant because they
+require detecting conflicts between concurrent updates (as needed for
+preventing Lost Updates or Write Skew phenomena), which we prove is
+unavailable. However, Read Committed isolation, Transactional
 Atomicity, and many of the weaker data consistency models from

### 2013-05-01 writeup/model.tex
PDB pass

@@
   ``partitioned'' data storage systems~\cite{dynamo, pnuts, bigtable,
-    spanner, hstore}, we consider partially replicated systems.} While
-this standard model is useful, there are several other cases we should
-consider.
+    spanner, hstore}, we consider partially replicated systems.}
 
-Many distributed algorithms assume a model in which clients always
-contact the same (sets of) server(s) across subsequent operations. As
+In addition to high availability, which allows operations on any replica,
+distributed algorithms often assume a model in which clients always
+contact the same (sets of) replica(s) across subsequent operations. As
 we will discuss in Section~\ref{sec:hats}, clients can ensure
@@
 integrity constraint violation) an \textit{internal abort} and an
-abort due to system implementation an \textit{external abort}.
+abort due to system implementation or operation an \textit{external abort}.
 

### 2013-05-01 writeup/motivation.tex
PDB pass

@@
 
-These results---which do not even consider server failures---indicate
+These results---which do not consider server failures---indicate
 that network partitions \textit{do} occur within and across modern
@@
 the speed of light. In the best case, two servers on opposite sides of
-the Earth---communicating via a hypothetical link traversing the
-planet's diameter---require a minimum 85.1ms RTT (133.7ms if sent at
+the Earth---communicating via a hypothetical link along the planet's
+diameter---require a minimum 85.1ms RTT (133.7ms if sent at
 surface level). As services are replicated to multiple, geographically
@@
 a speed-of-light RTT from S\~{a}o Paulo to Singapore RTT is 106.7ms,
-ICMP packets incur an average 362.8ms RTT (95th percentile: 649ms). As
+ping packets incur an average 362.8ms RTT (95th percentile: 649ms). As
 shown in Figure~\ref{fig:rtt}, the distribution of latencies varies
@@
 
-\section{When is ``ACID'' ACID?}
+\section{ACID in the Wild}
 \label{sec:modernacid}

### 2013-05-01 writeup/evaluation.tex
PB more

@@
 When we deployed five clusters (as opposed to two, as before) across
-the five AWS datacenters with lowest communication cost
+the five EC2 datacenters with lowest communication cost
 (Figure~\ref{fig:wan-exp}C), the trend continues: mastered latency
@@
 accordingly, the cost of anti-entropy increased (e.g., each server
-processed four incoming anti-entropy streams---as opposed to one
-before---reducing the opportunity for batching and decreasing
-available resources for for incoming client requests). This in turn
-increased garbage collection activity and, more importantly, IOPS when
-compared to Eventual and RC, causing TA throughput to peak at around half
-of Eventual. With an in-memory persistence (i.e., no LevelDB or WAL),
-TA throughput was within 20\% of eventual.
+processed four replica's anti-entropy as opposed to one before,
+reducing the opportunity for batching and decreasing available
+resources for for incoming client requests). This in turn increased
+garbage collection activity and, more importantly, IOPS when compared
+to Eventual and RC, causing TA throughput to peak at around half of
+Eventual. With an in-memory persistence (i.e., no LevelDB or WAL), TA
+throughput was within 20\% of eventual.
 
@@
 with all reads, TA is within $4.8\%$ of eventual; with all writes, TA
-is within $33\%$, and Eventual throughput decreases by $288.8\%$
-compared to all reads and TA is within $33\%$ of eventual. At $99.8\%$
-reads, TA incurs a $7\%$ overhead ($5.8\%$ for in-memory storage).
+is within $33\%$, and the throughput of eventual consistency decreases
+by $288.8\%$ compared to all reads. At $99.8\%$ reads, TA incurs a $7\%$ overhead ($5.8\%$ for
+in-memory storage).
 

### 2013-05-01 writeup/hat-analysis.tex
PB more

@@
 previous values''; reads from each item progress according to a total
-order. The ordering of reads should respect any externally observable
-total ordering on transactions.
+order. The ordering of versions read should respect any ordering on
+transactions.
 
@@
   update\lostupdate, preventing write skew\rwskew, and requiring
-  recency guarantees\linearizable.}\vspace{-.5em}
+  recency guarantees\linearizable.}
 \label{table:hatcompared}

### 2013-05-01 writeup/conclusion.tex
pass, plus spelling and grammar

@@
 \section{Conclusions and Future Work}
+\label{sec:conclusion}
+
 The current state of database software offers uncomfortable and

### 2013-05-01 writeup/evaluation.tex
pass, plus spelling and grammar

@@
   on coordination costs and on high availability.
-\item We examine the properties required by a realistic OLTP
-  application based on the TPC-C benchmark.
+\item We examine the properties required by an OLTP application based
+  on the TPC-C benchmark.
 \item We perform a brief experimental evaluation of HAT versus non-HAT
@@
 \subsection{Existing Algorithms}
+\label{sec:eval-existing}
 
@@
 setting~\cite{bernstein-book, davidson-survey, wisemann-survey}
-require at least one round trip time (RTT) before committing. As an
+require at least one RTT before committing. As an
 example, traditional two-phase locking for a transaction of length $T$
@@
 \texttt{lock} and one \texttt{unlock} operation.  In a distributed
-environment, each of these operations requires coordination, either
-with other database servers or with a lock service. If this
+environment, each of these lock operations requires coordination,
+either with other database servers or with a lock service. If this
 coordination mechanism is unavailable, transactions cannot safely
@@
 have item-cut isolation with causal consistency and TA (session
-PL-2L~\cite{adya}) but are unavailable in the presence of coordinator
+\textit{PL-2L}~\cite{adya}) but are unavailable in the presence of coordinator
 failure and assume serializable update transactions~\cite{readonly};
@@
 database~\cite{kraska-s3} and Bayou~\cite{sessionguarantees} can all
-provide variants of session PL-2L with high availability, but none
+provide variants of session \textit{PL-2L} with high availability, but none
 provide this HAT functionality without substantial
@@
 
-\vspace{.5em}\noindent\textbf{Commutativity and Monotonicity} Recent
-work on the CALM Theorem~\cite{calm} and Commutative and Replicated
-Data Types~\cite{crdt} demonstrates that, if updates logically
-commute, then they can often be safely performed in different orders
-at different replicas. Accordingly, as long as all writes are
-delivered to all replicas, then a system executing monotonic logic
-with commutative operators may not suffer from application-level
-consistency anomalies as a result of Lost Update or Write Skew
-anomalies. However, applications with non-monotonic state mutation
-will not, in general, be able to maintain application-level
+\vspace{.5em}\noindent\textbf{Commutativity and Monotonicity} As
+evidenced by the CALM Theorem~\cite{calm} and by Commutative and
+Replicated Data Types~\cite{crdt}, if updates logically commute, then
+they can often be safely performed in different orders at different
+replicas. Accordingly, as long as all writes are delivered to all
+replicas, then a system executing monotonic logic may not suffer from
+application-level consistency anomalies as a result of Lost Update or
+Write Skew anomalies. However, applications with non-monotonic state
+mutation will not, in general, be able to maintain application-level
 consistency constraints with HATs alone---in particular, applications
@@
 their HAT counterparts. In light of recent algorithms for providing
-causally consistent get transactions~\cite{bolton, cops}, we focus our
-HAT analysis on TA, which we believe to be the most novel of our
-algorithms. We find that HAT properties---while more
-resource-intensive than basic eventual consistency---scale well to
-many servers.
+causal consistency~\cite{bolton, cops, swift}, we focus our HAT
+analysis on TA (algorithm from Section~\ref{sec:ta}, described in
+greater in our Technical Report~\cite{hat-tr}), which we believe to be
+the most novel of our algorithms. We find that HAT properties---while
+more resource-intensive than basic eventual consistency---scale well
+to many servers.
 
@@
 by LevelDB and implemented in Java using Apache Thrift. It currently
-supports eventual consistency (last-writer-wins RU with standard
-all-to-all anti-entropy between replicas) and HAT RC and TA guarantees
-(as described in Section~\ref{sec:hats}). It also supports non-HAT
-operation whereby all operations for a given key are routed to a
-(randomly) designated \texttt{master} replica for each key (i.e.,
-PNUTS~\cite{pnuts}) as well as distributed two-phase locking. Servers
-are durable: they synchronously write to LevelDB before responding to
-client requests, while pending writes in TA are sent to a write-ahead
-log, which is synchronously flushed to disk.
+supports eventual consistency (hereafter, \texttt{eventual};
+last-writer-wins RU with standard all-to-all anti-entropy between
+replicas) and HAT RC and TA with RC guarantees (hereafter, \texttt{TA}). It also supports
+non-HAT operation whereby all operations for a given key are routed to
+a (randomly) designated \texttt{master} replica for each key
+(hereafter, \texttt{master}; i.e., PNUTS~\cite{pnuts}) as well as
+distributed two-phase locking. Servers are durable: they synchronously
+write to LevelDB before responding to client requests, while new
+writes in TA are sent to a write-ahead log, which is synchronously
+flushed to disk.
 
@@
 transaction. We increase the number of keys in the workload from the
-default $1,000$ to $100,000$ with uniform random key access, keeping
+default 1,000 to 100,000 with uniform random key access, keeping
 the default value size of $1KB$, and running YCSB for 180 seconds per
@@
 datacenters. Figure~\ref{fig:wan-exp}A shows that, when operating two
-clusters within a single datacenter, employing a single master for
-each data item results in approximately half the throughput and double
-the latency of an eventually consistent configuration: the HAT models
-are able to utilize double the number of servers. RC---essentially
-eventual consistency with buffering---is almost identical to eventual
-consistency, while TA---which incurs two writes for every client-side
-write (i.e., pending writes are sent to a WAL then subsequently moved
-into LevelDB once stable)---achieves ~75\% of the
-throughput. Latency increases linearly with the number of YCSB clients
-due to contention within LevelDB.
+clusters within a single datacenter, mastering each data item results
+in approximately half the throughput and double the latency of
+\texttt{eventual}. This is because HAT models are able to utilize
+replicas in both clusters instead of always contacting the (single)
+master. \texttt{RC}---essentially \texttt{eventual} with
+buffering---is almost identical to \texttt{eventual}, while
+\texttt{TA}---which incurs two writes for every client-side write
+(i.e., new writes are sent to the WAL then subsequently moved into
+LevelDB once pending stable)---achieves ~75\% of the throughput. Latency
+increases linearly with the number of YCSB clients due to contention
+within LevelDB.
 
- In contrast, when the two clusters are deployed across the
- continental United States (Figure~\ref{fig:wan-exp}B), the average
- latency of mastered operation increases to $300$ms (a $278$--$4257\%$
- latency increase; average $37$ms latency per operation). For the same
- number of YCSB client threads, mastered operation has substantially
- lower throughput than the HAT configurations. Increasing the number
- of YCSB clients \textit{does} increase the throughput of mastered
- operation, but our Thrift-based server-side connection processing did
- not gracefully handle more than several thousand concurrent
- connections. In contrast, across two datacenters, the performance of
- eventual, RC, and TA is near-identical to a single-datacenter
- deployment.
+In contrast, when the two clusters are deployed across the continental
+United States (Figure~\ref{fig:wan-exp}B), the average latency of
+\texttt{master} increases to $300$ms (a $278$--$4257\%$ latency
+increase; average $37$ms latency per operation). For the same number
+of YCSB client threads, \texttt{master} has substantially lower
+throughput than the HAT configurations. Increasing the number of YCSB
+clients \textit{does} increase the throughput of \texttt{master}, but
+our Thrift-based server-side connection processing did not gracefully
+handle more than several thousand concurrent connections. In contrast,
+across two datacenters, the performance of eventual, RC, and TA are
+near-identical to a single-datacenter deployment.
 
-When we deployed five clusters (as opposed to two, as before) across
+When five clusters (as opposed to two, as before) are deployed across
 the five EC2 datacenters with lowest communication cost
-(Figure~\ref{fig:wan-exp}C), the trend continues: mastered latency
-increases to nearly $800$ms per transaction. As an attempt at reducing
-this overhead, we implemented and benchmarked a variant of
+(Figure~\ref{fig:wan-exp}C), the trend continues: \texttt{master}
+latency increases to nearly $800$ms per transaction. As an attempt at
+reducing this overhead, we implemented and benchmarked a variant of
 quorum-based replication as in Dynamo~\cite{dynamo}, where clients
@@
 reducing the opportunity for batching and decreasing available
-resources for for incoming client requests). This in turn increased
+resources for incoming client requests). This in turn increased
 garbage collection activity and, more importantly, IOPS when compared
 to Eventual and RC, causing TA throughput to peak at around half of
-Eventual. With an in-memory persistence (i.e., no LevelDB or WAL), TA
-throughput was within 20\% of eventual.
+Eventual. With in-memory persistence (i.e., no LevelDB or WAL),
+\texttt{TA} throughput was within 20\% of \texttt{eventual}.
 
 We have intentionally omitted performance data for two-phase
-locking. Mastered configuration performed \textit{far} better than our
+locking. \texttt{master} performed \textit{far} better than our
 textbook implementation, which, in addition to requiring a WAN
-round-trip per transaction, also incurred substantial overheads due to
+round-trip per operation, also incurred substantial overheads due to
 mutual exclusion via locking. We expect that, while techniques like
@@
 overhead of serializable transactions by avoiding locking, our
-master-based implementation and the data from
-Section~\ref{sec:latency} are reasonable lower bounds on latency.
+mastered implementation and the data from Section~\ref{sec:latency}
+are reasonable lower bounds on latency.
 
@@
 Figure~\ref{fig:txlen} (clusters in Virginia and Oregon), throughput
-of eventual consistency, RC, and mastered operation are unaffected by
-transaction length. In contrast, TA's throughput decreases linearly
-with increased transaction length: with $1$ operation per transaction,
-TA throughput is within 18\% of eventual ($34$ bytes overhead), and
-with $128$ operations per transaction, TA throughput is within $60\%$
-($1898$ bytes overhead). This reflects our TA algorithm's metadata
-requirements, which are proportional to transaction length and consume
-IOPS and network bandwidth.
+of \texttt{eventual}, \texttt{RC}, and \texttt{master} operation are
+unaffected by transaction length. In contrast, \texttt{TA} throughput
+decreases linearly with increased transaction length: with $1$
+operation per transaction, \texttt{TA} throughput is within 18\% of
+\texttt{eventual} ($34$ bytes overhead), and with $128$ operations per
+transaction, \texttt{TA} throughput is within $60\%$ ($1898$ bytes
+overhead). This reflects our TA algorithm's metadata requirements,
+which are proportional to transaction length and consume IOPS and
+network bandwidth.
 
@@
 shown in Figure~\ref{fig:rprop} (clusters in Virginia and Oregon),
-with all reads, TA is within $4.8\%$ of eventual; with all writes, TA
-is within $33\%$, and the throughput of eventual consistency decreases
-by $288.8\%$ compared to all reads. At $99.8\%$ reads, TA incurs a $7\%$ overhead ($5.8\%$ for
-in-memory storage).
+with all reads, \texttt{TA} is within $4.8\%$ of \texttt{eventual};
+with all writes, \texttt{TA} is within $33\%$, and the throughput of
+\texttt{eventual} decreases by $288.8\%$ compared to all reads. At
+$99.8\%$ reads, \texttt{TA} incurs a $7\%$ overhead ($5.8\%$ for in-memory
+storage).
 
@@
 Oregon (with $15$ YCSB clients per server) results in linear scale-out
-for both eventual, RC, and TA. RC and eventual scale perfectly with an
-approximately $5$x throughput increase moving from $5$ to $25$ servers
-per cluster. For the same configuration, TA scales by $3.8$x,
-achieving over $260,000$ operations per second. We believe that TA's
-performance is due to resource contention---with a memory-backed
-database (instead of LevelDB), TA scales by $4.25$x (not shown)---and
-TA-related performance heterogeneity across servers (Calvin's authors
-report similar heterogeneity on EC2~\cite{calvin}). Nevertheless, this
-linear scalability is not typical of traditional database system
+for \texttt{eventual}, \texttt{RC}, and \texttt{TA}. \texttt{RC} and
+\texttt{eventual} scale perfectly with an approximately $5$x
+throughput increase moving from $5$ to $25$ servers per cluster. For
+the same configuration, \texttt{TA} scales by $3.8$x, achieving over
+$260,000$ operations per second. \texttt{TA} performance is due to
+resource contention---with a memory-backed database (instead of
+LevelDB), \texttt{TA} scales by $4.25$x (not shown)---and TA-related
+performance heterogeneity across servers (Calvin's authors report
+similar heterogeneity on EC2~\cite{calvin}). Nevertheless, this linear
+scalability is not typical of traditional database system
 designs~\cite{gray-isolation}.
@@
 high WAN latencies inevitable with non-HAT implementations. Our results
-highlight Deutcsh's observation that ignoring factors such as latency
+highlight Deutsch's observation that ignoring factors such as latency
 can ``cause big trouble and painful learning

### 2013-05-01 writeup/hat-analysis.tex
pass, plus spelling and grammar

@@
 databases, yet they cannot achieve all possible semantics. In this
-section, we delineate ACID, distributed consistency, and session
-consistency levels which can be achieved with high availability
-(transactional atomicity, variants of Repeatable Read isolation, and
-many session guarantees), those with sticky high availability (read
-your writes, PRAM and causal consistency), and properties that cannot
-be provided in a HAT system (those preventing Lost Update and Write
-Skew, or with recency).  We present a full summary of these results in
+section, we delineate ACID, distributed replica consistency, and
+session consistency levels which can be achieved with high
+availability (transactional atomicity, variants of Repeatable Read and
+traditionally weaker isolation, and many session guarantees), those
+with sticky high availability (read your writes, PRAM and causal
+consistency), and properties that cannot be provided in a HAT system
+(those preventing Lost Update and Write Skew or guaranteeing recency).
+We present a full summary of these results in
 Section~\ref{sec:hat-summary}.
@@
 To begin, we present well-known semantics that can be achieved in HAT
-systems. In this section, our goal is feasibility, not performance. As
-a result, we offer proof-of-concept highly available algorithms, but
-our goal is \textit{not} to provide optimal or even efficient
-implementations: the challenge is to provide algorithms that provide
-high availability. However, we briefly study performance implications
-in Section~\ref{sec:evaluation}.
+systems. In this section, our primary goal is feasibility, not
+performance. As a result, we offer proof-of-concept highly available
+algorithms that are not necessarily optimal or even efficient: the
+challenge is to prove the existence of algorithms that provide high
+availability. However, we briefly study a subset of their performance
+implications in Section~\ref{sec:evaluation}.
 
 \subsubsection{ACID Isolation Guarantees}
+\label{sec:isolation}
 
@@
 database. In a distributed database, different replicas may receive
-writes to their local copies of data at different times, but the order
-in which replicas overwrite versions should be consistent with the total order for
-each item. \textit{PL-1} requires that writes to different objects are
-ordered consistently across transactions, prohibiting Adya's
-phenomenon $G0$, also called ``Dirty Writes''~\cite{adya}. If we
-create a graph of transactions with edges from one transaction to
-another if the former overwrites the latter's write to the same
-object, then, under Read Uncommitted, there should be no cycles in
-this graph. Consider the following example:
+writes to their local copies of data at different times but should
+handle concurrent updates (i.e., overwrites) in accordance with the
+total order for each item. \textit{PL-1} requires that writes to
+different objects be ordered consistently across transactions,
+prohibiting Adya's phenomenon $G0$ (also called ``Dirty
+Writes''~\cite{ansicritique}). If we build a graph of transactions
+with edges from one transaction to another if the former overwrites
+the latter's write to the same object, then, under Read Uncommitted,
+the graph should not contain cycles~\cite{adya}. Consider the
+following example:
 \begin{align*}
@@
 achieved via timestamping each update in a transaction with the same
-logical timestamp (which is unique across transactions; e.g.,
-combining a client ID and logical clock) and applying a ``last writer
-wins'' conflict reconciliation policy at each replica. Later
-properties will strengthen Read Uncommitted.
+timestamp (unique across transactions; e.g., combining a client's ID
+with a sequence number) and applying a ``last writer wins'' conflict
+reconciliation policy at each replica. Later properties will
+strengthen Read Uncommitted.
 
 \textbf{Read Committed} isolation is particularly important in
-practice as it is the default of many DBMSs. Centralized
-implementations differ, with some based on long-duration exclusive
-locks and short-duration read locks~\cite{gray-isolation} and others
-based on multiple versions. The implementations often provide recency
-and monotonicity properties beyond the simple meaning from the name,
-which is what is expressed in the implementation-agnostic definition:
-under Read Committed, transactions should not access uncommitted or
+practice as it is the default of many DBMSs (as in
+Section~\ref{sec:modernacid}). Centralized implementations differ,
+with some based on long-duration exclusive locks and short-duration
+read locks~\cite{gray-isolation} and others based on multiple
+versions. These implementations often provide recency and monotonicity
+properties beyond what is implied by the name ``Read Committed'' and
+what is captured by the implementation-agnostic definition: under
+Read Committed, transactions should not access uncommitted or
 intermediate versions of data items. This prohibits both ``Dirty
@@
 from Berenson et al.). For instance, in the example below, $T_3$
-should never see $a=1$, and, if $T_2$ aborts, $T_3$ should never read
+should never see $a=1$, and, if $T_2$ aborts, $T_3$ should not read
 $a=3$:
@@
 value to other readers until notified that the writes have been
-committed. This implementation does not provide recency or
-monotonicity guarantees but satisfies the implementation-agnostic
-definition.
+committed. Unlike a lock-based implementation, this implementation
+does not provide recency or monotonicity guarantees but it satisfies
+the implementation-agnostic definition.
 
 Several different properties have been labeled \textbf{Repeatable
-  Read} Isolation. As we will show in
+  Read} isolation. As we will show in
 Section~\ref{sec:unachievable-acid}, some of these are not achievable
@@
 implementation-agnostic definition~\cite{ansi-sql} \textit{is}
-achievable and directly captures directly the spirit of the term: if a
+achievable and directly captures the spirit of the term: if a
 transaction reads the same data twice, it sees the same value each
@@
 property holds over reads from discrete data items, we call it
-\textbf{Item Cut Isolation (I-CI)}, and if we also expect a cut over
+\textbf{Item Cut Isolation}, and, if we also expect a cut over
 predicate-based reads (e.g., \texttt{SELECT WHERE}; preventing
-Phantoms~\cite{gray-isolation}, or Berenson et al.'s
-$P3/A3$~\cite{ansicritique}), we have the stronger property of
-\textbf{Predicate Cut-Isolation}. In the example below, under both
-levels of cut isolation, $T_3$ must read $a=1$:
+Phantoms~\cite{gray-isolation}, or Berenson et al.'s $P3/A3$), we have
+the stronger property of \textbf{Predicate Cut-Isolation}. In the
+example below, under both levels of cut isolation, $T_3$ must read
+$a=1$:
 \begin{align*}
@@
 It is possible to satisfy Item Cut Isolation with high availability by
-having clients cache their reads such that the value that they read
-never changes unless they overwrite it themselves. The cache can be
-cleared at the end of each transaction and can alternatively be
-accomplished on (sticky) replicas via multi-versioning. Predicate Cut
-Isolation is also achievable in HAT systems via similar caching or
-multi-versioning that tracks entire logical ranges of predicates
-in addition to item-based reads.
+having transactions cache their reads at the client such that the
+values that they read for each item never changes unless they
+overwrite it themselves. The cache can be cleared at the end of each
+transaction and can alternatively be accomplished on (sticky) replicas
+via multi-versioning. Predicate Cut Isolation is also achievable in
+HAT systems via similar caching or multi-versioning that track entire
+logical ranges of predicates in addition to item-based reads.
 
 \subsubsection{ACID Atomicity Guarantees}
+\label{sec:ta}
 
-\textbf{Transactional atomicity (TA)} is core to ACID guarantees. Although, at
-least by the ACID acronym, it is not an ``isolation'' property, TA
-restricts transactions' ability to view the effects of partially
-completed transactions. Under TA, once some of the effects of a
-transaction $T_i$ are observed by another transaction $T_j$,
-thereafter, all effects of $T_i$ are observed by $T_j$. Together with
-item cut isolation, TA prevents Read Skew anomalies (Berenson et al.'s
-A5A~\cite{ansicritique}). As an example of TA, because $T_2$ has read
-$T_1$'s write to $y$, $T_2$ must observe $b=c=1$ (or later versions
-for each key):
+\textbf{Transactional atomicity (TA)} is core to ACID
+guarantees. Although, at least by the ACID acronym, it is not an
+``isolation'' property, TA restricts transactions' ability to view the
+effects of partially completed transactions. Under TA, once some of
+the effects of a transaction $T_i$ are observed by another transaction
+$T_j$, thereafter, all effects of $T_i$ are observed by
+$T_j$. Together with item cut isolation, TA prevents Read Skew
+anomalies (Berenson et al.'s A5A). In the example below, under TA,
+because $T_2$ has read $T_1$'s write to $y$, $T_2$ must observe
+$b=c=1$ (or later versions for each key):
 \begin{align*}
@@
 $T_2$ can also observe $a=\bot$, $a=1$, or a later version of
-$a$. Notably, TA requires Read Committed isolation: observing all
-effects of a transaction implicitly requires observing the final
-(committed) effects of a transaction as well.
+$x$. Notably, TA requires disallows reading intermediate writes
+(Adya's $G1b$): observing all effects of a transaction implicitly
+requires observing the final (committed) effects of the transaction as
+well.
 
-Perplexingly, discussions of TA are absent from existing treatment of
+Perplexingly, discussions of TA are absent from existing treatments of
 weak isolation. This is perhaps again due to the single-node context
@@
 up-to-date/``consistent'' set of data items. However, replicas do not
-need to agree on when to choose to reveal new sets of values clients,
-which would require consensus. Instead, TA only requires the
-equivalent of reliable broadcast (with some additional client
-metadata) and is achievable in HAT systems.
+need to agree on when to choose to reveal sets of values to clients, which would require consensus. Instead, as we see below, TA
+only requires the equivalent of reliable broadcast (with some
+additional client metadata) and is achievable in HAT systems.
 
-TA can be achieved via multi-versioning. If clients attach vector
-clocks to every write (incrementing their own position for each
-transaction) and replicas store all writes via multi-versioning, then
-clients can safely determine which sets of data items to read; this
-approach is adopted by Swift~\cite{swift} and, to a lesser-extent,
-bolt-on causal consistency~\cite{bolton}. These systems also provide
-causal consistency and are subsequently sticky available (each is
-implemented via client-side caching). However, TA is achievable
-without stickiness via an alternate algorithm: if servers wait to
-reveal writes until they are present on all replicas, clients do not
-need to be sticky.
+% TA can be achieved via multi-versioning. If clients attach vector
+% clocks to every write (incrementing their own position for each
+% transaction) and replicas store all writes via multi-versioning, then
+% clients can safely determine which sets of data items to read; this
+% approach is adopted by Swift~\cite{swift} and, to a lesser-extent,
+% bolt-on causal consistency~\cite{bolton}. These systems also provide
+% causal consistency and are subsequently sticky available (each is
+% implemented via client-side caching). However, TA is achievable
+% without stickiness via an alternate algorithm: if servers wait to
+% reveal writes until they are present on all replicas, clients do not
+% need to be sticky.
 
-We sketch a non-sticky TA algorithm and provide greater detail in our
-technical report~\cite{hat-tr}. Each replica waits to reveal write $w$
-from transaction $T$ until all of the writes from $T$ are present on
-all respective replicas (are \textit{pending stable}). Once they are,
-replicas asynchronously reveal their respective writes to readers. To
+We sketch a highly available TA algorithm and provide greater detail
+in our technical report~\cite{hat-tr}. We begin with our Read
+Committed algorithm, but replicas wait to reveal new writes to readers
+until all of the replicas for the final writes in the transaction have
+received their respective writes (are \textit{pending
+  stable}). Accordingly, replicas reveal writes asynchronously;
+different replicas may detect pending stability at different times. To
 prevent clients from reading incomplete portions of a transaction's
@@
 that from a replica that has detected pending stability, then attempts
-to read another item from another replica which has not yet detected
+to read another item from another replica that has not yet detected
 pending stability), clients include additional metadata with each
-write: a list of items in the transaction along with the transaction's
-timestamp (from Read Uncommitted). When a client reads a write, the
-write's list of items and timestamp form a lower bound on the versions
-that the client should read for other items. When a client performs
-subsequent reads, it includes a timestamp with its request that
-corresponds current lower bound on reads from the respective item
-Replicas respond with either a write matching the timestamp or a write
-with a higher timestamp that is pending stable. Clients discard any
-metadata upon transaction commit, and servers keep two sets of writes
-for each data item: the write with the highest timestamp that is
-pending stable and a set of writes that are not yet pending stable.
+write: a single timestamp for all writes in the transaction (e.g., as
+in Read Uncommitted) and a list of items written to in the
+transaction. When a client reads a write, the write's timestamp and
+list of items form a lower bound on the versions that the client
+should read for the other items. When a client reads any data item, it
+attaches a timestamp to its request representing the current lower
+bound for that item. Replicas use this timestamp to respond with
+either a write matching the timestamp or a pending stable write with a
+higher timestamp. Clients may discard any metadata upon transaction
+commit, and servers keep two sets of writes for each data item: the
+write with the highest timestamp that is pending stable and a set of
+writes that are not yet pending stable.
 
-This non-sticky implementation is entirely masterless and neither
-reads nor writes block for coordination. Moreover, there are several
-possible optimizations that, for brevity, we do not describe
-here. However, one key property is that we do not, in general, dictate
-when writes become visible. Accordingly, the algorithm is highly
-available. We can optimize write visibility by sticking clients are
-with disjoint groups of servers: once all of a transaction's writes
-are pending stable within a group of servers, writes can be made
-visible (as opposed to once they are pending stable with respect to
-all servers). For example, in a multi-datacenter setting, all clients
-and servers within a datacenter may form a sticky group.
+This implementation is entirely master-less and operations never block
+due to replica coordination. Moreover, there are several possible
+optimizations that, for brevity, we do not describe here. However, one
+key property is that we do not dictate when writes become visible;
+accordingly, the algorithm is highly available. We can optimize write
+visibility via sticky availability by sticking clients are with
+disjoint groups of servers: once all of a transaction's writes are
+pending stable within a group of servers, writes can be made visible
+(as opposed to once they are pending stable with respect to all
+servers). For example, in a multi-datacenter setting, all clients and
+servers within a datacenter may form a sticky group.
 
@@
 exceptions~\cite{daudjee-session}), our models have not yet considered
-interactions \textit{across} transactions, other than that there
-exists some arbitrary ordering on them (via Read Uncommitted). In the
-distributed systems literature, many useful \textit{safety} guarantees
-span multiple, non-transactional operations. In particular,
-\textit{session guarantees} are used to describe guarantees across
-transactions within a given \textit{session}, ``an abstraction for the
-sequence of...operations performed during the execution of an
+real-time or client-centric ordering \textit{across} transactions. In
+the distributed systems literature~\cite{sessionguarantees, vogels-defs}, many
+useful \textit{safety} guarantees span multiple, non-transactional
+operations. In particular, \textit{session guarantees} are used to
+describe guarantees across transactions within a given
+\textit{session}, ``an abstraction for the sequence of...operations
+performed during the execution of an
 application''~\cite{sessionguarantees}. Informally a session describes
@@
 previous values''; reads from each item progress according to a total
-order. The ordering of versions read should respect any ordering on
-transactions.
+order (e.g., the order from Read Uncommitted).
 
@@
 reveal new writes until each write's respective dependencies are
-fulfilled on all replicas. This mechanism effectively ensures that all
+visible on all replicas. This mechanism effectively ensures that all
 clients read from a globally agreed upon lower bound on the versions
-written. This \textit{is} highly available as a client
-will never block due to inability to find a server with a sufficiently
-up-to-date version of a data item. However, it does not imply that
-transactions will read their own writes or, in the presence of
-partitions, make forward progress through the version history. The
-problem is that, if a server becomes partitioned, under the highly
-available model, we must handle the possibility that an unfortunate
-client will be forced to issue her next requests against the
-partitioned server.
+written. This is highly available as a client will never block due to
+inability to find a server with a sufficiently up-to-date version of a
+data item. However, it does not imply that transactions will read
+their own writes or, in the presence of partitions, make forward
+progress through the version history. The problem is that under
+non-sticky high availability, a system must handle the possibility
+that, under a partition, an unfortunate client will be forced to issue
+its next requests against a partitioned, out-of-date server.
 
-The solution to this conundrum is to give up high availability in
-favor of sticky availability. Sticky availability permits three
+The solution to this conundrum is to give up high availability and
+settle for sticky availability. Sticky availability permits three
 additional guarantees, which we first define and then prove are
@@
 
-\vspace{.5em}\noindent\textbf{{Read your writes}} requires
-that whenever a client reads a given data item after updating it, the
-read returns the updated value (or a value with a higher ID).
+\vspace{.5em}\noindent\textbf{{Read your writes}} requires that
+whenever a client reads a given data item after updating it, the read
+returns the updated value (or a value that overwrote the previously
+written value).
 
-\vspace{.5em}\noindent\textbf{{PRAM}} (Pipelined Random Access
-Memory) provides the illusion of serializing each session's operations
-and is the combination of monotonic reads, monotonic writes, and read
-your writes~\cite{herlihy-art}.
+\vspace{.5em}\noindent\textbf{{PRAM}} (Pipelined Random Access Memory)
+provides the illusion of serializing each of the operations (both
+reads and writes) within each session and is the combination of
+monotonic reads, monotonic writes, and read your
+writes~\cite{herlihy-art}.
 
@@
 session guarantees~\cite{sessiontocausal} (alternatively, PRAM with
-writes-follow-reads) and is also referred to by Adya as PL-2L
+writes-follow-reads) and is also referred to by Adya as \textit{PL-2L}
 isolation~\cite{adya}).\vspace{.5em}
@@
 If the client executes $T_1$ against a server that is partitioned from
-the rest of the other servers, the server must allow $T_1$ to
-commit. If the client subsequently executes $T_2$ against the same
-(partitioned) server, then it will be able to read its
-writes. However, if the network topology shifts and the client can
-only contact a different server that is partitioned from the server
-that executed $T_1$, then the client will be unable to read its own
-writes and the system will have to either stall indefinitely to allow
-the client to read her writes (violating transactional availability)
-or will have to sacrifice read your writes guarantees. However, if the
-client remains sticky with the server that executed $T_1$, then we can
+the rest of the other servers, then, for transactional availability,
+the server must allow $T_1$ to commit. If the same client subsequently
+executes $T_2$ against the same (partitioned) server in the same
+session, then it will be able to read its writes. However, if the
+network topology changes and the client can only contact a different
+server that is partitioned from the server that executed $T_1$, then
+the system will have to either stall indefinitely to allow the client
+to read her writes (violating transactional availability) or will have
+to sacrifice read your writes guarantees. However, if the client
+remains sticky with the server that executed $T_1$, then we can
 disallow this scenario. Accordingly, read your writes, and, by proxy,
@@
 
-In this section, we briefly discuss additional noteworthy guarantees
-achievable by HAT systems.
+In this section, we briefly discuss two additional kinds of guarantees
+that are achievable in HAT systems.
 
@@
 commutative and logically monotonic~\cite{calm} operations without the
-risk of invalidating application-level integrity constraints. We do
-not attempt to sketch the entire space of \textit{application-level}
-consistency properties that are achievable in HAT systems (see
+risk of invalidating application-level integrity constraints and can
+maintain limited criteria like foreign key constraints (via TA). We do
+not attempt to describe the entire space of application-level
+consistency properties that are achievable (see
 Section~\ref{sec:relatedwork}) but we specifically evaluate TPC-C
-transaction semantics under HAT consistency guarantees in
+transaction semantics with HAT guarantees in
 Section~\ref{sec:evaluation}.
@@
 system propagates writes between replicas, we can require convergence,
-or eventual consistency for each data item: in the absence of new
-mutations to a data item, in the absence of partitions, all servers
-should eventually agree on the value for each item. This is typically
-accomplished by any number of anti-entropy protocols, which
+or \textit{eventual consistency}: in the absence of new mutations to a
+data item, in the absence of partitions, all servers should eventually
+agree on the value for each item~\cite{cac, vogels-defs}. This is
+typically accomplished by any number of anti-entropy protocols, which
 periodically update neighboring servers with the latest value for each
-data item~\cite{antientropy}. Establishing a final value is related to
-determining a total order on transaction updates, as in Read
-Uncommitted.
+data item~\cite{antientropy}. Establishing a final convergent value is
+related to determining a total order on transaction updates to each
+item, as in Read Uncommitted.
 
@@
 Skew---and therefore providing Snapshot Isolation, Repeatable Read,
-and one-copy serializability---inherently requires unavailability.
+and one-copy serializability---inherently requires foregoing high
+availability guarantees.
 
@@
 Regardless of whether $x=120$ or $x=130$ is chosen by a replica, the
-database state could not have arisen serial execution of $T_1$ and
+database state could not have arisen from a serial execution of $T_1$ and
 $T_2$.\footnote{In this example, we assume that, as is standard in
@@
   (Section~\ref{sec:evaluation}) but the problem persists in the
-  general case.}  To prevent this from happening, either $T_1$ or
+  general case.}  To prevent this, either $T_1$ or
 $T_2$ should not have committed. Each client's respective server might
@@
 to a requirement for linearizability, which is, via Gilbert and
-Lynch's proof of the CAP Theorem, provably unachievable with high
+Lynch's proof of the CAP Theorem, provably at odds with high
 availability~\cite{gilbert-cap}.
@@
 Distributed data storage systems often make various recency guarantees
-on reads of data items. As we have discussed, one of the most famous
-is linearizability~\cite{herlihy-art}, which states that reads will
+on reads of data items.  Unfortunately, an indefinitely long partition
+can force an available system to violate any recency bound, so recency
+bounds are not enforceable by HAT systems~\cite{gilbert-cap}. One of
+the most famous of these guarantees is
+linearizability~\cite{herlihy-art}, which states that reads will
 return the last completed write to a data item, and there are several
@@
 such as ``read a version that is no more than five seconds out of
-date'' or similar. Unfortunately, an indefinitely long partition can
-force an available system to violate any recency bound, so recency
-bounds are not enforceable by HAT systems.
+date'' or similar. None of these guarantees is HAT-compliant.
 
@@
 guarantee models. Interestingly, strong one-copy serializability
-subsumes all other models, while considering the (large) power set of
-all compatible models (e.g., the diagram depicts 96 possible HAT
+entails all other models, while considering the (large) power set of
+all compatible models (e.g., the diagram depicts 144 possible HAT
 combinations) hints at the vast expanse of consistency models found in
-the literature. This taxonomy is not exhaustive, but we believe it
-lends substantial clarity into the relationships between a large
-subset of the prominent ACID and distributed consistency
-models. Additional read/write transaction semantics that we have
-omitted should be easily classifiable based on the available
-primitives and HAT-incompatible anomaly prevention we have already
-discussed.
+the literature. This taxonomy is not exhaustive
+(Section~\ref{sec:conclusion}), but we believe it lends substantial
+clarity to the relationships between a large subset of the prominent
+ACID and distributed consistency models. Additional read/write
+transaction semantics that we have omitted should be classifiable
+based on the available primitives and HAT-incompatible anomaly
+prevention we have already discussed.
 
@@
 \begin{tabular}{| c | p{6cm} | }\hline
-HA & Transactional Atomicity (TA), Read Uncommitted (RU), Read
-Committed (RC), Item Cut Isolation (I-CI), Predicate Cut Isolation
-(P-CI), Monotonic Reads (MR), Monotonic Writes (MW), Writes Follow
-Reads (WFR)\\\hline Sticky-HA & Read Your Writes (RYW), PRAM,
+HA & Read Uncommitted (RU), Read Committed (RC), Transactional
+Atomicity (TA), Item Cut Isolation (I-CI), Predicate Cut Isolation
+(P-CI), Writes Follow Reads (WFR), Monotonic Reads (MR), Monotonic
+Writes (MW)\\\hline Sticky HA & Read Your Writes (RYW), PRAM,
 Causal\\\hline Unavailable & Cursor Stability (CS)\lostupdate,
@@
  \node[draw=none,fill=none] (ici) at (1.2, 0) {I-CI};
- \node[draw=none,fill=none] (pci) at (1.65, 1.2) {P-CI};
- \node[draw=none,fill=none] (rc) at (-1.2, .8) {RC};
+ \node[draw=none,fill=none] (pci) at (1.75, 1.2) {P-CI};
+ \node[draw=none,fill=none] (rc) at (-1.2, .95) {RC};
  \node[draw=none,fill=none] (ru) at (-1.2, 0) {RU};
 
- \node[draw=none,fill=none] (ta) at (-.2, 1.3) {TA};
+ \node[draw=none,fill=none] (ta) at (-.2, 0) {TA};
 
@@
  \node at (4.8, 1) [sticky] (pram) {PRAM};
- \node[noha] (cs) at (-1.2, 1.8) {CS};
+ \node[noha] (cs) at (-1.2, 2) {CS};
  \node[noha] (rr) at (-.2, 2.7) {RR};
- \node[noha] (si) at (2.2, 2.4) {SI};
+ \node[noha] (si) at (1.75, 2.3) {SI};
  \node[noha] (1sr) at (1.2, 3.2) {1SR};
@@
  \draw [->] (ru) -- (rc);
- \draw [->] (rc) -- (ta);
  \draw [->] (ici) -- (pci);
@@
 this overhead can be reduced~\cite{eiger}, but it is much
-cheaper to provide, say, read your writes, than full causal
+cheaper to provide, say, read your writes guarantees, than causal
 consistency.
@@
 literature. Accordingly, despite apparent similarities, TA is
-incomparable with and much cheaper (by availability standards) than
-linearizability.
+incomparable with and less coordination intensive (by availability
+measures) than linearizability.
 

### 2013-05-01 writeup/hat-appendix.tex
pass, plus spelling and grammar

@@
 
-In our non-sticky TA algorithm, clients buffer their writes until
-commit, sending the final write to each data item to each respective
-server. Along with each write, clients attach a the transaction ID and
-a list of all data items written to in the transaction, $L_v$. Upon
-receiving a new write to item $d$, a server places the write into a
-\texttt{pending} queue and notifies all other servers responsible for
-a data item in $L_v$ that it has received a write to $d$ with the
-respective transaction ID. Once a server receives acknowledgments with
-appropriate transaction ID for all items in $L_v$, it moves the write
-from \texttt{pending} to the \texttt{stable} database (which applies
-the writes according to highest-ID-wins). Accordingly, all writes that
-should be made atomic with a write in \texttt{stable} are guaranteed
-to be present on their respective servers, either in \texttt{pending}
-or \texttt{stable}. Given the possibility of internal aborts, writes
-to \texttt{pending} may require 2 RTTs: one to check constraints and
-one to execute the above (but these checks \textit{cannot} cause other
-transactions to block, as would be the case in a two-phase
-\textit{commit} protocol).
+In this section, we provide additional information on our two-phase
+Transactional Atomicity algorithm.
 
-For reads, each client tracks which values it should read. It
-maintains a mapping (vector) from data items to transaction IDs. Upon
-transaction start, the client clears its mapping, and, when it
-performs a read from a data item, it sends the server its current ID
-for the item. If the client does not supply an ID, the server returns
-the latest value from \texttt{stable}. If the ID is non-empty, the
-server either returns the matching value from \texttt{pending} or a
-value with higher ID from \texttt{pending}---along with $L_v$ for the
-write. When the client receives a response, it updates its local
-vector with $L_v$. Accordingly, even if transaction's writes to two
-keys $x$ and $y$ are present on the replica for $x$'s \texttt{stable}
-and the replica for $y$'s \texttt{pending}, a client's reads from $x$
-then $y$ will return a transactionally atomic pair of values because
-the replica for $y$ will know to check \texttt{pending}.
+The below pseudocode describes a straightforward implementation of the
+algorithm from Section~\ref{sec:ta}. Replicas maintain two sets of
+data items: \texttt{pending} (which contains writes that are not yet
+pending stable) and \texttt{good} (which contains writes that are
+pending stable). Incoming writes are added to \texttt{pending} and
+moved to \texttt{good} once they are stable. Stable pending
+calculation is performed by counting the number of acknowledgments
+(\texttt{acks}) of writes from other replicas, signifying their
+receipt of the transaction's writes in their
+own \texttt{pending}. Accordingly, the algorithm maintains the
+invariant that the transactional siblings (or suitable replacements)
+for every write in \texttt{good} are present on their respective
+replicas. Clients use the RC algorithm from
+Section~\ref{sec:isolation} and keep a map of data items to timestamps
+(\texttt{required})---effectively a vector clock whose entries are
+data items (not processes, as is standard).
+
+\vspace{1em}\noindent\textbf{Example.} To illustrate this algorithm consider an execution of the following transactions:
+\begin{align*}
+\small\vspace{-1em}
+T_1 &: w_x(1)~w_y(1)
+\\T_2 &: r_x(1)~r_y(a)
+\end{align*}
+For TA, the system must ensure that $a=1$. Consider a system with one
+replica for each of $x$ and $y$ (denoted $R_x$ and $R_y$). Say client
+$c_1$ executes $T_1$ and commits. The client chooses timestamp $10001$
+(whereby the last $4$ digits are reserved for client ID to ensure
+uniqueness) and sends values to $w_x(1)$ to $R_x$ and $R_y$ with
+$tx\_keys = \{x,y\}$ and $timestamp=10001$. $R_x$ and $R_y$ place
+their respective writes into their individual \texttt{pending} sets
+and send acknowledgment for transaction timestamped $10001$ to the
+other. Consider the case where $r_x$ has seen $r_y$'s acknowledgment
+but not vice versa: $w_x(1)$ will be in \texttt{good} on $R_x$
+but \texttt{pending} on $R_y$. Now, if another client $c_2$ executes
+$T_2$ and reads $x$ from $R_x$, it will read $x=1$. $c_2$ will update
+its \texttt{required} vector so that $\texttt{required}=\{x:10001,
+y:10001\}$. When $c_2$ reads $y$ from $R_y$, it will specify
+$ts=10001$ and $R_y$ will not find a suitable write
+in \texttt{good}. $R_y$ will instead return $y=1$
+from \texttt{pending}.
+
+\vspace{1em}\noindent\textbf{Overheads.} TA requires overhead both on client and server.
+First, clients must store a vector (\texttt{required}) for the duration of a transaction in order to determine what values to read.
+Second, each write contains metadata ($tx\_keys$) linear in the number of writes in the transaction.
+(Note that we could have stored a different timestamp with every write, but a single timestamp suffices and is more efficient) These overheads are modest for short transaction lengths but they increase value sizes on disk, in memory, and over the network.
+Third, on the server side, each client \textsc{put} incurs two back-ends \texttt{put}s: one into \texttt{pending} and \texttt{good}.
+If keeping a single value for \texttt{good}, the second \texttt{put} into \texttt{good} (as is the case with normal, single-valued eventually consistent operation) is really a \texttt{get} then a possible \texttt{put}: the server first checks if a higher-timestamped value has been written and, if not, it overwrites the current value.
+The \texttt{put} into \texttt{pending} is simpler: if message delivery is exactly-once, then \texttt{put}s can be made unconditionally.
+
+\vspace{1em}\noindent\textbf{Optimizations.} The use of \texttt{pending} and \texttt{required} is effectively to
+handle the above race condition, where some but not all replicas have
+realized that a write is pending stable. With this in mind, there are
+several optimizations we can perform:
+
+If a replica serves a read from \texttt{pending}, it must mean one of
+two things: either the write is stable pending or the writing client
+for the write is requesting the data item (this second case does not
+occur in the below pseudocode since own-writes are served out
+of \texttt{write\_buffer}). In the former case, the replica can safely
+install the write into \texttt{good}.
+
+As we mention in Section~\ref{sec:ta}, the size of \texttt{good} can
+be limited to a single item. Recall that any item can be served
+from \texttt{good} as long as its timestamp is greater than the
+required timestamp. Accordingly, if replicas store multiple writes
+in \texttt{good}, any time they can serve a write from \texttt{good},
+the highest-timestamped write is always a valid response candidate
+(and, indeed, returning the highest-timestamped write provides the
+best visibility). Note, however, that serving a higher-than-requested
+timestamp from \texttt{pending} is dangerous: doing so might mean
+returning a write that is not pending stable and, via appropriate
+unfortunate partition behavior, force a client to stall while an
+available replica waits for the write to arrive in \texttt{pending}.
+
+The below pseudocode implements a push-based acknowledgment scheme: replicas
+eagerly notify other replicas about \texttt{pending} writes. This need
+not be the case and, indeed, it is wasteful: if a write
+in \texttt{pending} has lower timestamp than the highest timestamped
+write in \texttt{good}, it can safely be discarded. Especially for
+data items with high temporal locality or data stores that are
+replicated over WAN, this \texttt{pending} invalidation may be a
+frequent event. Instead, replicas can periodically poll other replicas
+for acknowledgments (i.e., asking for either a match
+in \texttt{pending} or a higher timestamp in \texttt{good}); this
+reduces both network utilization and, in the case of
+durable \texttt{pending} and \texttt{good} sets, disk utilization.
+
+There are ample opportunities for batching of pending and good
+acknowledgments. Batching is easier in a push-based acknowledgment
+scheme, as pull-based acknowledgments may require invalidation to
+actually reduce network transfers. Moreover, if a replica is
+responsible for several writes within a transaction, \textrm{notify}
+messages can be coalesced. We have also observed that batching
+multiple background operations together (particularly for
+anti-entropy) allows for more efficient compression---especially
+important in network-constrained environments.
+
+\vspace{1em}\noindent\textbf{Implementation note.} Our implementation from Section~\ref{sec:evaluation} implements the below pseudocode, with batching between \textrm{notify} operations and a single-item \texttt{good}.
+We have not yet investigated the efficiency of a pull-based approach (mostly due to our choice of RPC library, which unfortunately does not currently handle multiple outstanding requests per socket).
+We have also not experimentally determined the visibility effect of the side-channel pending stable notifications as in the first optimization.
+
+\vspace{1em}\noindent\textbf{Algorithmic improvements.} We are actively investigating more efficient alternatives to the algorithm presented here.
+In particular, a lower-overhead (but still highly available) implementation that used less metadata would be appealing.
+A system using server-side vector clocks is feasible but does not necessarily reduce overhead because the number of items written per transaction is likely smaller than the number of servers in a system.
+Similarly, we have observed an algorithm trade-off between the size of $ts_{required}$ and write metadata.
+While it is certainly possible to reduce overheads by reducing availability~\cite{eiger}, we believe high availability is a worthwhile goal for future exploration.
 

### 2013-05-01 writeup/hat-vldb.tex
pass, plus spelling and grammar

@@
 \let\endproof\relax
-\usepackage{draftwatermarkbottom, draftwatermarktop, scrextend, MnSymbol}
+\usepackage{draftwatermarkbottom, draftwatermarktop}
+\usepackage{scrextend, MnSymbol}
 
@@
 identify which can and cannot be achieved in HAT systems. This unifies
-the literature on weak isolation, data consistency, and highly
+the literature on weak isolation, replica consistency, and highly
 available systems. We analytically and experimentally quantify the
-availability and performance benefits of HATs---two to three orders of
-magnitude over wide-area networks---as well as their necessary
-semantic compromises.
+availability and performance benefits of HATs---often two to three
+orders of magnitude over wide-area networks---as well as their
+necessary semantic compromises.
 \end{abstract}\vspace{-.5em}
@@
 \pagebreak
-\begin{appendix}
+%\begin{appendix}
 %\input{availability-appendix.tex}
@@
 \input{hat-appendix.tex}
-\end{appendix}
+%\end{appendix}
 \end{document}

### 2013-05-01 writeup/intro.tex
pass, plus spelling and grammar

@@
 The last decade has seen a significant shift in the design of
-large-scale database backends, from the use of transactional
+large-scale database systems, from the use of transactional
 RDBMSs~\cite{bernstein-book, gray-isolation, gray-virtues} to the
-widespread adoption of loosely-consistent distributed key-value
+widespread adoption of loosely consistent distributed key-value
 stores~\cite{bigtable, pnuts, dynamo}. The 2000 introduction of
@@
 distributed consistency---have often been conflated with the ability
-to provide ACID database properties~\cite{brewer-slides, hn,
+to provide ACID database properties~\cite{hn,brewer-slides,
   foundation-article}; this misunderstanding has led to substantial
@@
 partitions~\cite{middleware-db, foundation-article,
-  hstore,generalizedsnapshot, mdcc, krikellas-bargain, eiger, walter,
-  calvin}.
+  hstore,spanner,generalizedsnapshot, mdcc, krikellas-bargain, eiger,
+  walter, calvin}.
 
@@
   gray-isolation, kemme-thesis}. Today's ACID and NewSQL databases
-often employ weak isolation models due to their concurrency and
-performance benefits; weak isolation is overwhelmingly the default
-setting in these stores and is often the only option offered
+often employ weak isolation models due to concurrency and performance
+benefits; weak isolation is overwhelmingly the default setting in
+these stores and is often the only option offered
 (Section~\ref{sec:modernacid}). While weak isolation levels do not
@@
 
-Unfortunately, the relationship between ACID semantics and high
-availability has not been well-explored. We have a strong
+To date, the relationship between ACID semantics and high
+availability has not been well explored. We have a strong
 understanding of weak isolation in the single-server context from
@@
 Our main contributions in this paper are as follows. We fill the gap
-between the plethora of previously proposed isolation and consistency
-models and the goal of high availability.  We classify which among the
-wide array of models are achievable with high availability, denoting
-them as {\em Highly Available Transactions} (HATs). In doing so, we
+between the many previously proposed isolation and consistency models
+and the goal of high availability.  We classify which among the wide
+array of models are achievable with high availability, denoting them
+as {\em Highly Available Transactions} (HATs). In doing so, we
 demonstrate that although many implementations of HAT semantics are
@@
 levels are achievable with high availability. We also demonstrate that
-many weak data consistency models from distributed systems are also
-both HAT-compliant and simultaneously achievable with traditional ACID
+many weak replica consistency models from distributed systems are both
+HAT-compliant and simultaneously achievable with several ACID
 properties.
@@
 constructive, proof-of-concept algorithms. For example, Snapshot
-Isolation and Repeatable Read isolation are not HAT-compliant because they
-require detecting conflicts between concurrent updates (as needed for
-preventing Lost Updates or Write Skew phenomena), which we prove is
-unavailable. However, Read Committed isolation, Transactional
+Isolation and Repeatable Read isolation are not HAT-compliant because
+they require detecting conflicts between concurrent updates (as needed
+for preventing Lost Updates or Write Skew phenomena), which we show
+is unavailable. However, Read Committed isolation, Transactional
 Atomicity, and many of the weaker data consistency models from
-database and distributed systems are achievable via novel algorithms
-that rely on multi-versioning and limited client-side caching. For
-several guarantees, such as causal consistency with phantom prevention
-and ANSI Repeatable Read, we consider a modified form of high
-availability in which clients ``stick to'' (i.e., have affinity with)
-at least one server (a property which is often implicit in the
-distributed systems literature~\cite{herlihy-art, eiger, cac} but
-which requires explicit consideration in a client-server replicated
-database context). This sticky availability is widely
-employed~\cite{eiger, vogels-defs} but is less restrictive (and
-therefore more easily achievable) than traditional high availability.
+database and distributed systems are achievable via algorithms that
+rely on multi-versioning and limited client-side caching. For several
+guarantees, such as causal consistency with phantom prevention and
+ANSI Repeatable Read, we consider a modified form of high availability
+in which clients ``stick to'' (i.e., have affinity with) at least one
+server---a property which is often implicit in the distributed systems
+literature~\cite{herlihy-art, eiger, cac} but which requires explicit
+consideration in a client-server replicated database context. This
+sticky availability is widely employed~\cite{eiger, vogels-defs} but
+is less restrictive (and therefore more easily achievable) than
+traditional high availability.
 
@@
 provide acceptable semantics for a wide range of programs, especially
-those with monotonic logic and commutative updates~\cite{calm, crdt}. HAT systems can also enforce arbitrary foreign key
-constraints on multi-item updates and sometimes provide limited
+those with monotonic logic and commutative updates~\cite{calm,
+  crdt}. HAT systems can also enforce arbitrary foreign key
+constraints for multi-item updates and, in some cases, provide limited
 uniqueness guarantees. However, particularly for programs with
 non-monotonic logic, HATs can fall short, requiring
-sometimes-unavailable mechanisms that are more
-coordination-intensive. Overall, this paper quantifies the benefits of
-Highly Available Transactions can offer as well as the necessary
-semantic restrictions they place on applications.
+sometimes-unavailable mechanisms that are more coordination
+intensive. Overall, this paper quantifies the benefits that Highly
+Available Transactions offer as well as the necessary restrictions
+that they place on applications.
 

### 2013-05-01 writeup/model.tex
pass, plus spelling and grammar

@@
 
-To understand which guarantees can be provided with high availabilty,
+To understand which guarantees can be provided with high availability,
 we must first define what high availability means. In this section, we
 will formulate a model that captures a range of availability models,
-including ``true'' high availability, availability with stickiness,
-and transactional availability.
+including high availability, availability with stickiness, and
+transactional availability.
 
@@
 highly available system also provides low latency~\cite{abadi-pacelc};
-in a wide-area setting, this means that clients need not wait for
-cross-datacenter communication. To properly describe whether a
-\textit{transactional} system is highly available, we need to describe
-what servers a client must contact as well as what kinds of responses
-a server can provide, especially given the possibility of aborts.
+in a wide-area setting, clients of a highly available system need not
+wait for cross-datacenter communication. To properly describe whether
+a \textit{transactional} system is highly available, we need to
+describe what servers a client must contact as well as what kinds of
+responses a server can provide, especially given the possibility of
+aborts.
 
@@
 servers~\cite{gilbert-cap}.\footnote{Under this definition, systems
-  that require a majority of servers to be online or that require the
-  database to function correctly with high (but not perfect)
-  probability are not available.} As in a standard distributed
-database, designated servers might perform operations for different
-data items; a server that can handle an operation for a given data
-item is called a \textit{replica} for that item.\footnote{There is a
-  further distinction between a \textit{fully replicated} system, in
-  which all servers are replicas for all data items, and a
-  \textit{partially replicated} system, in which at least one server
-  acts as a replica for only a subset of all data items. For
-  generality, and, for applicability to many modern ``sharded'' or
-  ``partitioned'' data storage systems~\cite{dynamo, pnuts, bigtable,
-    spanner, hstore}, we consider partially replicated systems.}
+  that require a majority of servers to be online is not
+  available. Similarly, a system which ensures that clients receive a
+  response with high (but not perfect) probability is not available.}
+As in a standard distributed database, designated servers might
+perform operations for different data items; a server that can handle
+an operation for a given data item is called a \textit{replica} for
+that item.\footnote{There is a further distinction between a
+  \textit{fully replicated} system, in which all servers are replicas
+  for all data items, and a \textit{partially replicated} system, in
+  which at least one server acts as a replica for only a subset of all
+  data items. For generality, and, for applicability to many modern
+  ``sharded'' or ``partitioned'' data storage systems~\cite{
+   bigtable, pnuts,  spanner, dynamo, hstore}, we consider partially
+  replicated systems.}
 
@@
 set of servers~\cite{vogels-defs}. Simultaneously maintaining
-availability and stickiness requires fate-sharing between the client
+availability and stickiness requires fate sharing between the client
 and its server: if a client's sticky servers fail or if the client is

### 2013-05-01 writeup/motivation.tex
pass, plus spelling and grammar

@@
 
-Why does availability matter? Peter Deutsch starts his classic list of
-``Fallacies of Distributed Computing'' with two concerns fundamental
-to distributed database systems: ``\textit{1.)}  The network is
-reliable. \textit{2.)} Latency is zero''~\cite{fallacies-deutsch}; in
-a distributed setting, network failures may prevent database servers
-from communicating, and, in the absence of failures, communication is
-slowed by factors like physical distance and, more often, congestion
-and routing. As we will see (Section~\ref{sec:availability}), highly
-available system operation minimizes the effect of network
-\textit{partitions} and \textit{latency}. In this Section, we draw on
-a range of evidence that indicates that partitions occur relatively
-frequently in real-world deployments and best-case latencies between
-datacenters are substantial, often on the order of hundreds of
-milliseconds.
+Why does high availability matter? Peter Deutsch starts his classic
+list of ``Fallacies of Distributed Computing'' with two concerns
+fundamental to distributed database systems: ``\textit{1.)}  The
+network is reliable. \textit{2.)} Latency is
+zero''~\cite{fallacies-deutsch}; in a distributed setting, network
+failures may prevent database servers from communicating, and, in the
+absence of failures, communication is slowed by factors like physical
+distance and, more often, congestion and routing. As we will see
+(Section~\ref{sec:availability}), highly available system design
+mitigates the effects of network \textit{partitions} and
+\textit{latency}. In this section, we draw on a range of evidence that
+indicates that partitions occur relatively frequently in real-world
+deployments and best-case latencies between datacenters are
+substantial, often on the order of several hundreds of milliseconds.
 
@@
 to full site disruptions for popular web services like Reddit,
-Foursquare, and Heroku~\cite{ec2-downsites}. At a global scale,
+Foursquare, and Heroku~\cite{ec2-downsites}. At global scale,
 hardware failures---like the 2011 outages in Internet backbones in
@@
 Even with fault-free networks, distributed systems face the challenge
-of communication latency, Deutsch's second ``Fallacy.'' In this
+of network communication latency, Deutsch's second ``Fallacy.'' In this
 section, we quantify round-trip latencies, which are often
-large---hundreds of milliseconds---in a geo-replicated,
+large---hundreds of milliseconds in a geo-replicated,
 multi-datacenter context.  Fundamentally, the speed at which two
@@
 
-\subfloat[Cross-region (CA:~California, OR:~Oregon, VA:~Virginia, TO:~Tokyo, IR:~Ireland, SY:~Sydney, SP:~S\~{a}o Paulo, SI:~Singapore)] {
+\subfloat[Cross-region (CA:~California, OR:~Oregon, VA:~Virginia, TO:~Tokyo, IR:~Ireland, SY:~Sydney, SP:~S\~{a}o Paulo, SI:~Singapore)\vspace{-.5em}] {
   \begin{tabular}{|c|c|c|c|c|c|c|c|c|}
@@
 
-\caption{Mean RTT times on EC2 (min and max highlighted)}\vspace{-1em}
+\caption{Mean RTT times on EC2 (min and max highlighted)}\vspace{-.5em}
 \label{table:rtt}
@@
 inter-planetary networks, we performed a measurement study of network
-behavior on Amazon's EC2, a widely-used public compute cloud. We
-measured one week of ping times between all seven EC2 geographic
-``regions'', across three ``availability zones'' (closely co-located
-datacenters), and within a single ``availabilty zone'' (datacenter),
-at a granularity of 1s (dataset to be released). We summarize the
-results of our network measurement study in Table~\ref{table:rtt}. On
-average, intra-datacenter communication (Table 1a) is between 1.82 and
-6.38 times faster than across geographically co-located datacenters
-(Table 1b) and between 40 and 647 times faster than across
-geographically distributed datacenters (Table 1c). The cost of
-wide-area communication exceeds the speed of light: for example, while
-a speed-of-light RTT from S\~{a}o Paulo to Singapore RTT is 106.7ms,
+behavior on Amazon's EC2, a widely used public compute cloud. We
+measured one week of ping times (i.e., round-trip times, or
+RTTs) between all seven EC2 geographic ``regions,'' across three
+``availability zones'' (closely co-located datacenters), and within a
+single ``availability zone'' (datacenter), at a granularity of 1s
+(dataset to be released). We summarize the results of our network
+measurement study in Table~\ref{table:rtt}. On average,
+intra-datacenter communication (Table 1a) is between 1.82 and 6.38
+times faster than across geographically co-located datacenters (Table
+1b) and between 40 and 647 times faster than across geographically
+distributed datacenters (Table 1c). The cost of wide-area
+communication exceeds the speed of light: for example, while a
+speed-of-light RTT from S\~{a}o Paulo to Singapore RTT is 106.7ms,
 ping packets incur an average 362.8ms RTT (95th percentile: 649ms). As
@@
 concurrency, database systems offer a range of ACID properties weaker
-than serializability.  Instead, databases offer a host of so-called
-\textit{weak isolation} models that allow varying restrictions on the
-space of schedules that are allowable by the system~\cite{adya,
-  ansicritique, ansi-sql}. None of these weak isolation models
-guarantees serializability, but, as we see below, their benefits are
-often considered to outweigh costs of possible consistency anomalies
-that might arise from their use.
+than serializability: the host of so-called \textit{weak isolation}
+models describe varying restrictions on the space of schedules that
+are allowable by the system~\cite{adya, ansi-sql, ansicritique}. None
+of these weak isolation models guarantees serializability, but, as we see below,
+their benefits are often considered to outweigh costs of possible
+consistency anomalies that might arise from their use.
 
-To understand the prevalence of these weak isolation models, we
-recently surveyed the default and maximum isolation models provided by
+To understand the prevalence of weak isolation, we
+recently surveyed the default and maximum isolation guarantees provided by
 18 databases, often claiming to provide ``ACID'' or ``NewSQL''
@@
 It is unknown \textit{which} of these guarantees can be provided with
-high availability. The existing algorithms for providing weak
-isolation are typically designed for a single-node context and are, to
-the best of our knowledge, unavailable due to a reliance on
-concurrency control mechanisms like locking that are not resilient to
-partial failure (Section~\ref{sec:evaluation}). Moreover, we are not
-aware of any prior literature that provides guidance as to the
+high availability, or are HAT-compliant. Existing algorithms for
+providing weak isolation are often designed for a single-node context
+and are, to the best of our knowledge, unavailable due to reliance
+on concurrency control mechanisms like locking that are not resilient
+to partial failure (Section~\ref{sec:eval-existing}). Moreover, we are
+not aware of any prior literature that provides guidance as to the
 relationship between weak isolation and high availability: prior work
 has examined the relationship between serializability and high
-availability~\cite{davidson-survey} and weak isolation on a single
-server~\cite{adya, ansicritique, gray-isolation} but never weak
-isolation and high availability taken together.  A primary goal in the
+availability~\cite{davidson-survey} and weak isolation in
+general~\cite{adya, ansicritique, gray-isolation} but not weak
+isolation and high availability together.  A primary goal in the
 remainder of this paper is to understand which models are

### 2013-05-01 writeup/relatedwork.tex
PDB pass
pass, plus spelling and grammar

@@
 
-There is a growing body related work on highly available semantics,
-mechanisms for concurrency control, and techniques for scalable
-distributed operations.
+While we have discussed traditional mechanisms for distributed
+coordination and several related systems in
+Section~\ref{sec:eval-existing}, in this section, we further discuss
+related work. In particular, we discuss related work on highly
+available semantics, mechanisms for concurrency control, and
+techniques for scalable distributed operations.
 
@@
 high availability is relatively unexplored. With the exception of our
-preliminary workshop paper discussing transactional availability and
-HAT RC and I-CI~\cite{hat-hotos}--- which this work expands, providing
-additional algorithms and impossibility results, a full
-taxonomization, and experimental and application analysis---we believe
-this paper is the first to explore this area. However, several studies
-rigorously classify \textit{non-HAT} semantics: Wisemann et al. have
-proposed a three parameter classification for one-copy serializable
-database systems, performing a more extensive classification than our
-overview in Section~\ref{sec:evaluation}, including eager database
-replication techniques~\cite{kemme-classification} and distributed
-mechanisms for achieving one-copy
-serializability~\cite{wisemann-survey}.
+preliminary workshop paper discussing transactional availability,
+real-world ACID, and HAT RC and I-CI~\cite{hat-hotos}---which this
+work expands, providing additional algorithms and impossibility
+results, a full taxonomization, experimental and application analysis,
+and much greater discussion of implications and related work---we
+believe this paper is the first to explore this area. However, several
+studies rigorously classify \textit{non-HAT} semantics: Wisemann et
+al. have proposed a three parameter classification for one-copy
+serializable database systems, performing a more extensive
+classification than our overview in Section~\ref{sec:evaluation},
+including eager database replication
+techniques~\cite{kemme-classification} and distributed mechanisms for
+achieving one-copy serializability~\cite{wisemann-survey}.
 
@@
 system~\cite{cac}. This result is promising and complementary to our
-results for general-purpose convergent data stores.
+results for general-purpose convergent data stores. Finally, Burkhardt
+et al. have concurrently developed an axiomatic specification for
+eventually consistent stores; their work-in-progress report contains
+alternate formalism for several HAT guarantees and (snapshot)
+transactions~\cite{burkhardt-txns}.
 

### 2013-05-01 writeup/conclusion.tex
Final submitted

@@
 possible to achieve, we believe HATs provide a large and useful design
-space to explore.
+space for exploration.
 

### 2013-05-01 writeup/evaluation.tex
Final submitted

@@
 substantial latency penalties: over WAN, 10 to 100 times higher than
-their HAT counterparts. In light of recent algorithms for providing
+their HAT counterparts. Given recent algorithms for providing
 causal consistency~\cite{bolton, cops, swift}, we focus our HAT
 analysis on TA (algorithm from Section~\ref{sec:ta}, described in
-greater in our Technical Report~\cite{hat-tr}), which we believe to be
-the most novel of our algorithms. We find that HAT properties---while
-more resource-intensive than basic eventual consistency---scale well
-to many servers.
+greater detail
+in Appendix B
+%in our Technical Report~\cite{hat-tr}
+), which we believe to be the
+most interesting of our algorithms. We find that HAT
+properties---while more resource-intensive than basic eventual
+consistency---scale well to many servers.
 

### 2013-05-01 writeup/hat-analysis.tex
Final submitted

@@
 here (accompanied by appropriate references) but give a full set of
-formal definitions in our extended Technical Report~\cite{hat-tr}. In
-our examples, we exclusively consider read and write operations,
+formal definitions in
+ Appendix A.
+% our extended Technical Report~\cite{hat-tr}.
+In our examples, we exclusively consider read and write operations,
 denoting a write of value $v$ to data item $d$ as $w_d(v)$ and a read
@@
 We sketch a highly available TA algorithm and provide greater detail
-in our technical report~\cite{hat-tr}. We begin with our Read
-Committed algorithm, but replicas wait to reveal new writes to readers
+in Appendix B.
+%our extended technical report~\cite{hat-tr}.
+We begin with our Read Committed algorithm, but replicas wait to reveal new writes to readers
 until all of the replicas for the final writes in the transaction have

### 2013-05-01 writeup/hat-vldb.tex
dat appendix
Final submitted

@@
 \let\endproof\relax
-\usepackage{draftwatermarkbottom, draftwatermarktop}
+%\usepackage{draftwatermarkbottom, draftwatermarktop}
 \usepackage{scrextend, MnSymbol}
+
+\usepackage[hidelinks]{hyperref}
 
@@
 
-\title{Highly Available Transactions: Virtues and Limitations}
+\title{Highly Available Transactions: Virtues and Limitations\\(Extended Version)}
 {\author{Peter Bailis{\fontsize{12}{14}$^\dagger$}, Aaron Davidson{\fontsize{12}{14}$^\dagger$}, Alan Fekete{\fontsize{12}{14}$^\diamond$}, Ali Ghodsi{\fontsize{12}{14}$^{\dagger,\ddagger}$}, Joseph M.
@@
 % Example of an appendix; typically would start on a new page
-%\pagebreak
-%
 \pagebreak
-%\begin{appendix}
-%\input{availability-appendix.tex}
+
+\pagebreak
+\begin{appendix}
 
 \input{hat-appendix.tex}
-%\end{appendix}
+\end{appendix}
 \end{document}

### 2013-07-04 writeup/evaluation.tex
Start of reviewer comments.

@@
 update. The former can be modified to ensure ID uniqueness but not
-sequential ID ordering, while the latter is inherently
-non-monotonic, requiring external compensation or stronger
-consistency protocols. Based on these experiences and discussions with
-practitioners, we expect that, especially for read-dominated workloads
-found in many online services, HAT guarantees will provide useful
-semantics for many applications.
+sequential ID ordering, while the latter is inherently non-monotonic,
+requiring external compensation or stronger consistency
+protocols. Based on these experiences and discussions with
+practitioners, we believe that HAT guarantees can provide useful
+semantics for a large class of application functionality, while a
+(possibly small) subset of operations will require stronger,
+unavailable properties.
 
@@
 last-writer-wins RU with standard all-to-all anti-entropy between
-replicas) and HAT RC and TA with RC guarantees (hereafter, \texttt{TA}). It also supports
-non-HAT operation whereby all operations for a given key are routed to
-a (randomly) designated \texttt{master} replica for each key
-(hereafter, \texttt{master}; i.e., PNUTS~\cite{pnuts}) as well as
+replicas) and HAT RC and TA with RC guarantees (hereafter,
+\texttt{TA}). It also supports non-HAT operation whereby all
+operations for a given key are routed to a (randomly) designated
+\texttt{master} replica for each key (guaranteeing single-key
+linearizability, as in Gilbert and Lynch's CAP Theorem
+proof~\cite{gilbert-cap} and in PNUTS~\cite{pnuts}'s
+\texttt{read-latest} operation; hereafter, \texttt{master}) as well as
 distributed two-phase locking. Servers are durable: they synchronously

### 2013-07-04 writeup/hat-analysis.tex
Start of reviewer comments.

@@
 replicated database), TA is achievable via lightweight locking and/or
-local concurrency control over data items~\cite{gstore}. In contrast,
-in a distributed environment, TA over arbitrary groups of
-non-colocated items is considerably more difficult to achieve with
-high availability: servers need to ensure that, regardless of replica
-selection, all of a client's accesses will read a sufficiently
-up-to-date/``consistent'' set of data items. However, replicas do not
-need to agree on when to choose to reveal sets of values to clients, which would require consensus. Instead, as we see below, TA
-only requires the equivalent of reliable broadcast (with some
-additional client metadata) and is achievable in HAT systems.
+local concurrency control over data items~\cite{gstore,
+  kemme-thesis}. In contrast, in a distributed environment, TA over
+arbitrary groups of non-colocated items is considerably more difficult
+to achieve with high availability: servers need to ensure that,
+regardless of replica selection, all of a client's accesses will read
+a sufficiently up-to-date/``consistent'' set of data items. However,
+replicas do not need to agree on when to choose to reveal sets of
+values to clients, which would require consensus. Instead, as we see
+below, TA only requires the equivalent of reliable broadcast (with
+some additional client metadata) and is achievable in HAT systems.
 
@@
 effects of $T_2$ if it can also observe $T_1$'s effects (or later
-values that supersede $T_1$'s).  Any order on transactions should
-respect the reads-from order.\vspace{.5em}
+values that supersede $T_1$'s); this corresponds to Lamport's
+``happens-before'' relation~\cite{lamportclocks}.  Any order on
+transactions should respect the transitive reads-from
+order.\vspace{.5em}
 
 The above guarantees can be achieved by forcing servers to wait to
-reveal new writes until each write's respective dependencies are
-visible on all replicas. This mechanism effectively ensures that all
-clients read from a globally agreed upon lower bound on the versions
-written. This is highly available as a client will never block due to
-inability to find a server with a sufficiently up-to-date version of a
-data item. However, it does not imply that transactions will read
-their own writes or, in the presence of partitions, make forward
-progress through the version history. The problem is that under
-non-sticky high availability, a system must handle the possibility
-that, under a partition, an unfortunate client will be forced to issue
-its next requests against a partitioned, out-of-date server.
+reveal new writes (say, by buffering them in separate local storage)
+until each write's respective dependencies are visible on all
+replicas. This mechanism effectively ensures that all clients read
+from a globally agreed upon lower bound on the versions written. This
+is highly available as a client will never block due to inability to
+find a server with a sufficiently up-to-date version of a data
+item. However, it does not imply that transactions will read their own
+writes or, in the presence of partitions, make forward progress
+through the version history. The problem is that under non-sticky high
+availability, a system must handle the possibility that, under a
+partition, an unfortunate client will be forced to issue its next
+requests against a partitioned, out-of-date server.
 
@@
 Section~\ref{sec:evaluation}.
-
-%\vspace{.5em}\noindent{\textbf{Durability}} A client requiring that its
-%transactions' effects survive $F$ server faults requires at least
-%$F+1$ non-failing replicas; this affects availability.
 
@@
 Their need to prevent Lost Update means that Consistent Read, Snapshot
-Isolation, and Cursor Stability guarantees are all unavailable.
-Repeatable Read (defined by Gray~\cite{gray-isolation}, Berenson et
+Isolation (including Parallel Snapshot Isolation~\cite{walter}), and
+Cursor Stability guarantees are all unavailable.  Repeatable Read
+(defined by Gray~\cite{gray-isolation}, Berenson et
 al.~\cite{ansicritique}, and Adya~\cite{adya}) and One-Copy
-Serializability need to prevent both Lost Update and Write Skew. These
+Serializability~\cite{1sr} need to prevent both Lost Update and Write Skew. These
 prevention requirements mean that these guarantees are inherently
@@
 date'' or similar. None of these guarantees is HAT-compliant.
+
+\subsubsection{Durability}
+
+A client requiring that its transactions' effects survive $F$ server
+faults requires at least $F+1$ non-failing replicas; this affects
+availability and, according to the Gilbert and Lynch definition we
+have adopted, $F>1$ fault tolerance is not achievable with high
+availability.
 

### 2013-07-04 writeup/hat-appendix.tex
Start of reviewer comments.

@@
 \texttt{timestamp}: unique, per transaction identifier\\
-$\texttt{write}:[\texttt{key:}k, \texttt{value}:v, \texttt{timestamp}:ts, \texttt{set<key>}:tx\_keys]$\\\hrulefill}\vspace{.5em}
+$\texttt{write}:[\texttt{key:}k, \texttt{value}:v, \texttt{timestamp}:ts, \texttt{set<key>}:sibs]$\\\hrulefill}\vspace{.5em}
 
@@
   \State \texttt{pending}.add($w$)
-  \For{key $k_t \in w.tx\_keys$}
+  \For{key $k_t \in w.sibs$}
   \State \{all replicas for $sib$\}.notify($w.ts$)
@@
   \State \texttt{acks}.get($ts$).increment()
-  \If{all \texttt{acks} received for all replicas for $ts$'s $tx\_keys$}
+  \If{all \texttt{acks} received for all replicas for $ts$'s $sibs$}
   \State \texttt{good}.add($w$)
@@
   \State $w\_{ret}$ = (available replica for $k$).get($k$, \texttt{required}.get($k$))
-  \For {$(tx_{key} \in w_{ret}.tx\_keys$}
+  \For {$(tx_{key} \in w_{ret}.sibs$}
   \If {$w_{ret}.ts > \texttt{required}.get(tx_{key})$}

### 2013-07-04 writeup/model.tex
Start of reviewer comments.

@@
 Informally, highly available algorithms ensure ``always on'' operation
-and guaranteed low latency. If users of a highly available system are
-able to contact a (set of) server(s) in a system, they are guaranteed
-a response; this means servers will not need to communicate with any
-others. If servers are partitioned from one another, they do not need
-to stall in order to provide clients a ``safe'' response to
-operations. This lack of fast-path coordination also means that a
-highly available system also provides low latency~\cite{abadi-pacelc};
-in a wide-area setting, clients of a highly available system need not
-wait for cross-datacenter communication. To properly describe whether
-a \textit{transactional} system is highly available, we need to
-describe what servers a client must contact as well as what kinds of
-responses a server can provide, especially given the possibility of
-aborts.
+and, as a side effect, guarantee low latency. If users of a highly
+available system are able to contact a (set of) server(s) in a system,
+they are guaranteed a response; this means servers will not need to
+communicate with any others. If servers are partitioned from one
+another, they do not need to stall in order to provide clients a
+``safe'' response to operations. This lack of fast-path coordination
+also means that a highly available system also provides low
+latency~\cite{abadi-pacelc}; in a wide-area setting, clients of a
+highly available system need not wait for cross-datacenter
+communication. To properly describe whether a \textit{transactional}
+system is highly available, we need to describe what servers a client
+must contact as well as what kinds of responses a server can provide,
+especially given the possibility of aborts.
 
@@
   available. Similarly, a system which ensures that clients receive a
-  response with high (but not perfect) probability is not available.}
-As in a standard distributed database, designated servers might
-perform operations for different data items; a server that can handle
-an operation for a given data item is called a \textit{replica} for
-that item.\footnote{There is a further distinction between a
-  \textit{fully replicated} system, in which all servers are replicas
-  for all data items, and a \textit{partially replicated} system, in
-  which at least one server acts as a replica for only a subset of all
-  data items. For generality, and, for applicability to many modern
-  ``sharded'' or ``partitioned'' data storage systems~\cite{
-   bigtable, pnuts,  spanner, dynamo, hstore}, we consider partially
-  replicated systems.}
+  response with high (but not perfect) probability is not
+  available. This is a stringent definition but matches the
+  assumptions made by the CAP Theorem~\cite{gilbert-cap} and
+  guarantees low latency~\cite{abadi-pacelc}.} As in a standard
+distributed database, designated servers might perform operations for
+different data items; a server that can handle an operation for a
+given data item is called a \textit{replica} for that
+item.\footnote{There is a further distinction between a \textit{fully
+    replicated} system, in which all servers are replicas for all data
+  items, and a \textit{partially replicated} system, in which at least
+  one server acts as a replica for only a subset of all data
+  items. For generality, and, for applicability to many modern
+  ``sharded'' or ``partitioned'' data storage systems~\cite{ bigtable,
+    pnuts, spanner, dynamo, hstore}, we consider partially replicated
+  systems.}
 
-In addition to high availability, which allows operations on any replica,
-distributed algorithms often assume a model in which clients always
-contact the same (sets of) replica(s) across subsequent operations. As
-we will discuss in Section~\ref{sec:hats}, clients can ensure
-continuity between operations (e.g., reading their prior updates to a
-data item) by maintaining affinity or ``stickiness'' with a server or
-set of servers~\cite{vogels-defs}. Simultaneously maintaining
-availability and stickiness requires fate sharing between the client
-and its server: if a client's sticky servers fail or if the client is
-partitioned from its sticky servers, then it may face
+\subsection{Sticky Availability}
+
+In addition to high availability, which allows operations on any
+replica, distributed algorithms often assume a model in which clients
+always contact the same logical (sets of) replica(s) across subsequent
+operations. As we will discuss in Section~\ref{sec:hats}, clients can
+ensure continuity between operations (e.g., reading their prior
+updates to a data item) by maintaining affinity or ``stickiness'' with
+a server or set of servers~\cite{vogels-defs}. Simultaneously
+maintaining availability and stickiness requires fate sharing between
+the client and its server: if a client's sticky servers fail or if the
+client is partitioned from its sticky servers, then it may face
 unavailability. We say that a system provides \textit{sticky
-  availability} if, whenever a client's operation is executed against
-a correct server that has observed all of its prior operations, it
-eventually receives a response, even in the presence of indefinitely
-long partitions. A client may choose to become sticky available by
-acting as a server itself; for example, a client might cache its reads
-and writes~\cite{bolton, sessionguarantees, swift}. Any guarantee
-achievable in a highly available system is achievable in a sticky high
-availability system but not vice-versa.
+  availability} if, whenever a client's operation on an item is
+executed against a correct replica that has observed all of its prior
+operations for that item, it eventually receives a response, even in
+the presence of indefinitely long partitions. A client may choose to
+become sticky available by acting as a server itself; for example, a
+client might cache its reads and writes~\cite{bolton,
+  sessionguarantees, swift}. Any guarantee achievable in a highly
+available system is achievable in a sticky high availability system
+but not vice-versa.
 
@@
 largely focuses on transactions: groups of multiple operations over
-multiple objects. Accordingly, by itself, replica availability is
-insufficient to describe availability guarantees for
-transactions. Additionally, given the choice of \textit{commit} and
-\textit{abort} responses---which signal transaction success or failure
-to a client---we must take care in defining transactional
+multiple objects. Accordingly, by itself, traditional definitions of
+high availability are insufficient to describe availability guarantees
+for transactions. Additionally, given the choice of \textit{commit}
+and \textit{abort} responses---which signal transaction success or
+failure to a client---we must take care in defining transactional
 availability.
@@
 on multiple data items, then the client must be able to contact and
-receive a response from at least one replica for each data item. This
-may result in ``lower availability'' than a non-transactional
-requirement. Additionally, given the possibility of system-initiated
-aborts, we need to ensure useful forward progress: a system can
-trivially guarantee clients a response by always aborting all
-transactions. However, this is an unsatisfactory system because
+receive a response from at least one replica for each data item. We
+say that a transaction has \emph{replica availability} if it can
+contact at least one replica for every item it attempts to
+access. This may result in ``lower availability'' than a
+non-transactional requirement. Additionally, given the possibility of
+system-initiated aborts, we need to ensure useful forward progress: a
+system can trivially guarantee clients a response by always aborting
+all transactions. However, this is an unsatisfactory system because
 nothing good (transaction commit) ever happens; we should require a
@@
 integrity constraint violation) an \textit{internal abort} and an
-abort due to system implementation or operation an \textit{external abort}.
-
-We say that a system provides \textit{transactional availability} if,
-given replica availability for every data item in a transaction, the
-transaction eventually commits (possibly after multiple client
-retries) or internally aborts~\cite{hat-hotos}. For example, a system
-will violate transactional availability if a client that can contact
-two servers for each of the data items in its transaction does not
-eventually commit in the absence of internal aborts.
+abort due to system implementation or operation an \textit{external
+  abort}. We say that a system provides \textit{transactional
+  availability} if, given replica availability for every data item in
+a transaction, the transaction eventually commits (possibly after
+multiple client retries) or internally aborts~\cite{hat-hotos}.

### 2013-07-04 writeup/motivation.tex
fix labels
Start of reviewer comments.

@@
 Several recent studies rigorously quantify partition behavior. A 2011
-study of several Microsoft datacenters found a mean of 40.8 network
-link failures per day (95th percentile: 136), with a median time to
-repair of around five minutes (and up to one week). Perhaps
-surprisingly, provisioning redundant networks only reduces impact of
-failures by up to 40\%, meaning network providers cannot easily
-curtail partition behavior~\cite{sigcomm-dc}. A 2010 study of over 200
-wide-area routers found an average of 16.2--302.0 failures per link
-per year with an average annual downtime of 24--497 minutes per link
-per year (95th percentile at least 34 hours)~\cite{sigcomm-wan}. In
-HP's managed enterprise networks, WAN, LAN, and connectivity problems
-account for 28.1\% of all customer support tickets while 39\% of
-tickets relate to network hardware.  The median incident duration for
-highest priority tickets ranges from 114--188 minutes and up to a full
-day for all tickets~\cite{turner2012failure}. Other studies confirm
-these results, showing median time between failures over a WAN network
+study of several Microsoft datacenters observed over 13,300 network
+failures with end-user impact, with an estimated median 59,000 packets
+lost per failure. The study found a mean of 40.8 network link failures
+per day (95th percentile: 136), with a median time to repair of around
+five minutes (and up to one week). Perhaps surprisingly, provisioning
+redundant networks only reduces impact of failures by up to 40\%,
+meaning network providers cannot easily curtail partition
+behavior~\cite{sigcomm-dc}. A 2010 study of over 200 wide-area routers
+found an average of 16.2--302.0 failures per link per year with an
+average annual downtime of 24--497 minutes per link per year (95th
+percentile at least 34 hours)~\cite{sigcomm-wan}. In HP's managed
+enterprise networks, WAN, LAN, and connectivity problems account for
+28.1\% of all customer support tickets while 39\% of tickets relate to
+network hardware.  The median incident duration for highest priority
+tickets ranges from 114--188 minutes and up to a full day for all
+tickets~\cite{turner2012failure}. Other studies confirm these results,
+showing median time between connectivity failures over a WAN network
 of approximately 3000 seconds with a median time to repair between 2
 and 1000 seconds~\cite{ip-backbone-failures} as well as frequent path
-routing failures on the Internet~\cite{labovitz-failures}. Isolating,
-quantifying, and accounting for these network failures is an area of
-active research in networking community~\cite{uw-failure-networks}.
+routing failures on the Internet~\cite{labovitz-failures}. Finally,
+Kingsbury and Bailis have recently collected an informal survey of
+network failures~\cite{aphyr-post}. Isolating, quantifying, and
+accounting for these network failures is an area of active research in
+networking community~\cite{uw-failure-networks}.
 
@@
 
+\subfloat[Within \texttt{us-east-b} AZ] {
+  \makebox[.2\textwidth]{
+ \begin{tabular}{|c|c|c|}\hline
+ & \multicolumn{1}{c}{H2} & \multicolumn{1}{c|}{H3}\\\hline
+H1  & 0.55   & \colorbox{max-lat-color}{0.56} \\
+H2 &  & \colorbox{min-lat-color}{0.50}  \\
+\hline
+  \end{tabular}}
+ }
 \subfloat[Across \texttt{us-east} AZs]{
-  \makebox[.2\textwidth]{
+  \makebox[.25\textwidth]{
     \begin{tabular}{|c|c|c|}\hline
@@
 }
-\subfloat[Within \texttt{us-east-b} AZ] {
-  \makebox[.25\textwidth]{
-  \begin{tabular}{|c|c|c|}\hline
- & \multicolumn{1}{c}{H2} & \multicolumn{1}{c|}{H3}\\\hline
-H1  & 0.55   & \colorbox{max-lat-color}{0.56} \\
-H2 &  & \colorbox{min-lat-color}{0.50}  \\
-\hline
-  \end{tabular}
- }}
 
@@
 behavior on Amazon's EC2, a widely used public compute cloud. We
-measured one week of ping times (i.e., round-trip times, or
-RTTs) between all seven EC2 geographic ``regions,'' across three
+measured one week of ping times (i.e., round-trip times, or RTTs)
+between all seven EC2 geographic ``regions,'' across three
 ``availability zones'' (closely co-located datacenters), and within a
@@
 between links, but the trend is clear: remote communication has a
-substantial cost.
+substantial cost. Like failures, quantifying and delays communication
+delays is an active area of research in the networking
+community~\cite{bobtail}.
 

### 2013-07-19 writeup/model.tex
Discuss stickiness

@@
 updates to a data item) by maintaining affinity or ``stickiness'' with
-a server or set of servers~\cite{vogels-defs}. Simultaneously
-maintaining availability and stickiness requires fate sharing between
-the client and its server: if a client's sticky servers fail or if the
-client is partitioned from its sticky servers, then it may face
-unavailability. We say that a system provides \textit{sticky
-  availability} if, whenever a client's operation on an item is
-executed against a correct replica that has observed all of its prior
-operations for that item, it eventually receives a response, even in
-the presence of indefinitely long partitions. A client may choose to
+a server or set of servers~\cite{vogels-defs}. In a fully-replicated
+system, where all servers are replicas for all data items, stickiness
+is simple: a client can maintain stickiness by contacting the same
+server for each of its requests. However, in a partially-replicated
+system, where servers are replicas for subsets of the set of data
+items (which we consider in this paper), we say a client must remain
+sticky with a single \textit{logical} server that may consist of
+multiple physical servers.. We say that a system provides
+\textit{sticky availability} if, whenever a client's operation on an
+item is executed against a logical server that has observed all of its
+prior operations, it eventually receives a response, even in the
+presence of indefinitely long partitions. A client may choose to
 become sticky available by acting as a server itself; for example, a

### 2013-07-19 writeup/hat-analysis.tex
Fix style, proof

@@
 $T_j$. Together with item cut isolation, TA prevents Read Skew
-anomalies (Berenson et al.'s A5A). In the example below, under TA,
-because $T_2$ has read $T_1$'s write to $y$, $T_2$ must observe
-$b=c=1$ (or later versions for each key):
+anomalies (Berenson et al.'s A5A) and is useful in several contexts
+such as maintaining foreign key constraints, consistent global
+secondary indexing, and maintenance of derived data. In the example
+below, under TA, because $T_2$ has read $T_1$'s write to $y$, $T_2$
+must observe $b=c=1$ (or later versions for each key):
 \begin{align*}
@@
   bolton, cops, sessionguarantees, swift} of the prior session
-guarantee algorithms we presented earlier.
+guarantee algorithms we presented earlier: only reveal new writes to
+clients when their dependencies have been revealed.
 

### 2013-07-19 writeup/model.tex
Fix style, proof

@@
 
-Traditionally, a system provides {\em high availability} if every user
+Traditionally, a system provides {\textbf{high availability}} if every user
 that can contact a correct (non-failing) server eventually receives a
@@
 multiple physical servers.. We say that a system provides
-\textit{sticky availability} if, whenever a client's operation on an
+\textbf{sticky availability} if, whenever a client's operation on an
 item is executed against a logical server that has observed all of its
@@
 receive a response from at least one replica for each data item. We
-say that a transaction has \emph{replica availability} if it can
+say that a transaction has \textbf{replica availability} if it can
 contact at least one replica for every item it attempts to
@@
 abort due to system implementation or operation an \textit{external
-  abort}. We say that a system provides \textit{transactional
+  abort}. We say that a system provides \textbf{transactional
   availability} if, given replica availability for every data item in

### 2013-07-22 writeup/evaluation.tex
Starting to address TA -> RA

@@
 on preventing write conflicts. Chan and Gray's read-only transactions
-have item-cut isolation with causal consistency and TA (session
+have item-cut isolation with causal consistency and RA (session
 \textit{PL-2L}~\cite{adya}) but are unavailable in the presence of coordinator
@@
 transaction is monotonic---increment- and append-only---so all balance
-increase operations commute, and TA allows the maintenance of
+increase operations commute, and RA allows the maintenance of
 foreign-key integrity constraints (e.g., via \texttt{UPDATE/DELETE
@@
 tables but, given the ability to update rows in both tables atomically
-via TA, can be easily maintained. Consistency Conditions 4 through 12
+via RA, can be easily maintained. Consistency Conditions 4 through 12
 (3.3.2.4-12) can similarly be satisfied by applying updates atomically
@@
 causal consistency~\cite{bolton, cops, swift}, we focus our HAT
-analysis on TA (algorithm from Section~\ref{sec:ta}, described in
+analysis on RA (algorithm from Section~\ref{sec:ta}, described in
 greater detail
@@
 last-writer-wins RU with standard all-to-all anti-entropy between
-replicas) and HAT RC and TA with RC guarantees (hereafter,
-\texttt{TA}). It also supports non-HAT operation whereby all
+replicas) and HAT RC and RA with RC guarantees (hereafter,
+\texttt{RA}). It also supports non-HAT operation whereby all
 operations for a given key are routed to a (randomly) designated
@@
 write to LevelDB before responding to client requests, while new
-writes in TA are sent to a write-ahead log, which is synchronously
+writes in RA are sent to a write-ahead log, which is synchronously
 flushed to disk.
@@
 buffering---is almost identical to \texttt{eventual}, while
-\texttt{TA}---which incurs two writes for every client-side write
+\texttt{RA}---which incurs two writes for every client-side write
 (i.e., new writes are sent to the WAL then subsequently moved into
@@
 handle more than several thousand concurrent connections. In contrast,
-across two datacenters, the performance of eventual, RC, and TA are
+across two datacenters, the performance of eventual, RC, and RA are
 near-identical to a single-datacenter deployment.
@@
 because worst-case server load was unaffected. With five clusters,
-TA's relative throughput decreased: every YCSB \texttt{put} operation
+RA's relative throughput decreased: every YCSB \texttt{put} operation
 resulted in four \texttt{put} operations on remote replicas and,
@@
 garbage collection activity and, more importantly, IOPS when compared
-to Eventual and RC, causing TA throughput to peak at around half of
+to Eventual and RC, causing RA throughput to peak at around half of
 Eventual. With in-memory persistence (i.e., no LevelDB or WAL),
-\texttt{TA} throughput was within 20\% of \texttt{eventual}.
+\texttt{RA} throughput was within 20\% of \texttt{eventual}.
 
@@
 \end{center}\vspace{-2.25em}
-\caption{Scale-out of TA, Eventual, and RC.}\vspace{-1.5em}
+\caption{Scale-out of RA, Eventual, and RC.}\vspace{-1.5em}
 \label{fig:scaleout}
@@
 of \texttt{eventual}, \texttt{RC}, and \texttt{master} operation are
-unaffected by transaction length. In contrast, \texttt{TA} throughput
+unaffected by transaction length. In contrast, \texttt{RA} throughput
 decreases linearly with increased transaction length: with $1$
-operation per transaction, \texttt{TA} throughput is within 18\% of
+operation per transaction, \texttt{RA} throughput is within 18\% of
 \texttt{eventual} ($34$ bytes overhead), and with $128$ operations per
-transaction, \texttt{TA} throughput is within $60\%$ ($1898$ bytes
-overhead). This reflects our TA algorithm's metadata requirements,
+transaction, \texttt{RA} throughput is within $60\%$ ($1898$ bytes
+overhead). This reflects our RA algorithm's metadata requirements,
 which are proportional to transaction length and consume IOPS and
@@
 shown in Figure~\ref{fig:rprop} (clusters in Virginia and Oregon),
-with all reads, \texttt{TA} is within $4.8\%$ of \texttt{eventual};
-with all writes, \texttt{TA} is within $33\%$, and the throughput of
+with all reads, \texttt{RA} is within $4.8\%$ of \texttt{eventual};
+with all writes, \texttt{RA} is within $33\%$, and the throughput of
 \texttt{eventual} decreases by $288.8\%$ compared to all reads. At
-$99.8\%$ reads, \texttt{TA} incurs a $7\%$ overhead ($5.8\%$ for in-memory
+$99.8\%$ reads, \texttt{RA} incurs a $7\%$ overhead ($5.8\%$ for in-memory
 storage).
@@
 Oregon (with $15$ YCSB clients per server) results in linear scale-out
-for \texttt{eventual}, \texttt{RC}, and \texttt{TA}. \texttt{RC} and
+for \texttt{eventual}, \texttt{RC}, and \texttt{RA}. \texttt{RC} and
 \texttt{eventual} scale perfectly with an approximately $5$x
 throughput increase moving from $5$ to $25$ servers per cluster. For
-the same configuration, \texttt{TA} scales by $3.8$x, achieving over
-$260,000$ operations per second. \texttt{TA} performance is due to
+the same configuration, \texttt{RA} scales by $3.8$x, achieving over
+$260,000$ operations per second. \texttt{RA} performance is due to
 resource contention---with a memory-backed database (instead of
-LevelDB), \texttt{TA} scales by $4.25$x (not shown)---and TA-related
+LevelDB), \texttt{RA} scales by $4.25$x (not shown)---and RA-related
 performance heterogeneity across servers (Calvin's authors report
@@
 useful semantics without substantial performance penalties. In
-particular, our TA algorithm can achieve throughput competitive with
+particular, our RA algorithm can achieve throughput competitive with
 eventual consistency at the expense of increased disk and network

### 2013-07-22 writeup/hat-analysis.tex
Starting to address TA -> RA

@@
 
-\textbf{Transactional atomicity (TA)} is core to ACID
+Atomicity, informally guaranteeing that either all or none of
+transactions' effects should succeed, is core to ACID
 guarantees. Although, at least by the ACID acronym, it is not an
-``isolation'' property, TA restricts transactions' ability to view the
-effects of partially completed transactions. Under TA, once some of
-the effects of a transaction $T_i$ are observed by another transaction
-$T_j$, thereafter, all effects of $T_i$ are observed by
-$T_j$. Together with item cut isolation, TA prevents Read Skew
-anomalies (Berenson et al.'s A5A) and is useful in several contexts
-such as maintaining foreign key constraints, consistent global
-secondary indexing, and maintenance of derived data. In the example
-below, under TA, because $T_2$ has read $T_1$'s write to $y$, $T_2$
-must observe $b=c=1$ (or later versions for each key):
+``isolation'' property, atomicity properties also restrict the updates
+visible to other transactions. Accordingly, here, we consider the
+\textit{isolation} effects of atomicity, which we call \textbf{Read
+  Atomic (RA)} isolation.  Under RA, once some of the effects of a
+transaction $T_i$ are observed by another transaction $T_j$,
+thereafter, all effects of $T_i$ are observed by $T_j$. That is, if a
+transaction $T_j$ reads a version of an object that transaction $T_i$
+wrote, then a later read by $T_j$ cannot return value whose next
+version is installed by $T_i$. Together with item cut isolation, RA
+prevents Read Skew anomalies (Berenson et al.'s A5A) and is useful in
+several contexts such as maintaining foreign key constraints,
+consistent global secondary indexing, and maintenance of derived
+data. In the example below, under RA, because $T_2$ has read $T_1$'s
+write to $y$, $T_2$ must observe $b=c=1$ (or later versions for each
+key):
 \begin{align*}
@@
 \end{align*}
-$T_2$ can also observe $a=\bot$, $a=1$, or a later version of
-$x$. Notably, TA requires disallows reading intermediate writes
-(Adya's $G1b$): observing all effects of a transaction implicitly
-requires observing the final (committed) effects of the transaction as
-well.
+$T_2$ can also observe $a=\bot$, $a=1$, or a later version of $x$. In
+the hierarchy of existing isolation properties, we place RA below
+Adya's $PL-2L$ (as it does not necessarily enforce transitive
+read-write dependencies) but above Read Committed ($PL-2$). Notably,
+RA requires disallows reading intermediate writes (Adya's $G1b$):
+observing all effects of a transaction implicitly requires observing
+the final (committed) effects of the transaction as well.
 
-Perplexingly, discussions of TA are absent from existing treatments of
+Perplexingly, discussions of RA are absent from existing treatments of
 weak isolation. This is perhaps again due to the single-node context
 in which prior work was developed: on a single server (or a fully
-replicated database), TA is achievable via lightweight locking and/or
+replicated database), RA is achievable via lightweight locking and/or
 local concurrency control over data items~\cite{gstore,
-  kemme-thesis}. In contrast, in a distributed environment, TA over
+  kemme-thesis}. In contrast, in a distributed environment, RA over
 arbitrary groups of non-colocated items is considerably more difficult
@@
 values to clients, which would require consensus. Instead, as we see
-below, TA only requires the equivalent of reliable broadcast (with
+below, RA only requires the equivalent of reliable broadcast (with
 some additional client metadata) and is achievable in HAT systems.
 
-% TA can be achieved via multi-versioning. If clients attach vector
+% RA can be achieved via multi-versioning. If clients attach vector
 % clocks to every write (incrementing their own position for each
@@
 % causal consistency and are subsequently sticky available (each is
-% implemented via client-side caching). However, TA is achievable
+% implemented via client-side caching). However, RA is achievable
 % without stickiness via an alternate algorithm: if servers wait to
@@
 
-We sketch a highly available TA algorithm and provide greater detail
-in Appendix B.
-%our extended technical report~\cite{hat-tr}.
-We begin with our Read Committed algorithm, but replicas wait to reveal new writes to readers
+As a straw-man, replicas can store all versions ever written for all
+items. Servers can gossip information about versions they have seen
+and construct a lower bound on the versions that can be found on every
+replica. At the start of each transaction, clients choose a bound that
+is guaranteed to be lower than or equal to the this global lower
+bound, and, during transaction execution, replicas return the latest
+version of each item that is not greater than the client's chosen
+bound. If the lower bound is advanced along transactional boundaries,
+clients will observe RA. This algorithm has several variants in the
+literature~\cite{readonly, swift}: bounds can be represented by vector
+clocks, and older versions can be asynchronously garbage collected.
+
+We have developed a more efficient RA algorithm, which we sketch here
+and provide greater detail in Appendix B. We begin with our Read
+Committed algorithm, but replicas wait to reveal new writes to readers
 until all of the replicas for the final writes in the transaction have
 received their respective writes (are \textit{pending
-  stable}). Accordingly, replicas reveal writes asynchronously;
-different replicas may detect pending stability at different times. To
-prevent clients from reading incomplete portions of a transaction's
-writes as they are being revealed (e.g., a client reads a data item
-that from a replica that has detected pending stability, then attempts
-to read another item from another replica that has not yet detected
-pending stability), clients include additional metadata with each
-write: a single timestamp for all writes in the transaction (e.g., as
-in Read Uncommitted) and a list of items written to in the
-transaction. When a client reads a write, the write's timestamp and
-list of items form a lower bound on the versions that the client
-should read for the other items. When a client reads any data item, it
-attaches a timestamp to its request representing the current lower
-bound for that item. Replicas use this timestamp to respond with
-either a write matching the timestamp or a pending stable write with a
-higher timestamp. Clients may discard any metadata upon transaction
-commit, and servers keep two sets of writes for each data item: the
-write with the highest timestamp that is pending stable and a set of
-writes that are not yet pending stable.
+  stable}). Clients include additional metadata with each write: a
+single timestamp for all writes in the transaction (e.g., as in Read
+Uncommitted) and a list of items written to in the transaction. When a
+client reads, the return value's timestamp and list of items form a
+lower bound on the versions that the client should read for the other
+items. When a client reads, it attaches a timestamp to its request
+representing the current lower bound for that item. Replicas use this
+timestamp to respond with either a write matching the timestamp or a
+pending stable write with a higher timestamp. Servers keep two sets of
+writes for each data item: the write with the highest timestamp that
+is pending stable and a set of writes that are not yet pending
+stable. This is entirely master-less and operations never block due to
+replica coordination. Moreover, there are several possible
+optimizations that, for brevity, we do not describe here.
 
-This implementation is entirely master-less and operations never block
-due to replica coordination. Moreover, there are several possible
-optimizations that, for brevity, we do not describe here. However, one
-key property is that we do not dictate when writes become visible;
-accordingly, the algorithm is highly available. We can optimize write
-visibility via sticky availability by sticking clients are with
-disjoint groups of servers: once all of a transaction's writes are
-pending stable within a group of servers, writes can be made visible
-(as opposed to once they are pending stable with respect to all
-servers). For example, in a multi-datacenter setting, all clients and
-servers within a datacenter may form a sticky group.
+However, one key property with RA is that it do not dictate when
+writes become visible; accordingly, the algorithm is highly
+available. We can optimize write visibility via sticky availability by
+sticking clients are with disjoint groups of servers: once all of a
+transaction's writes are pending stable within a group of servers,
+writes can be made visible (as opposed to once they are pending stable
+with respect to all servers). For example, in a multi-datacenter
+setting, all clients and servers within a datacenter may form a sticky
+group.
 
@@
 reads (i.e., Causal Transactional Predicate Cut Isolation). If we
-combine TA and P-CI, we have transactional snapshot reads. We can
+combine RA and P-CI, we have transactional snapshot reads. We can
 achieve RC, MR, and RYW by simply sticking clients to servers. We can
@@
 \begin{tabular}{| c | p{6cm} | }\hline
-HA & Read Uncommitted (RU), Read Committed (RC), Transactional
-Atomicity (TA), Item Cut Isolation (I-CI), Predicate Cut Isolation
+HA & Read Uncommitted (RU), Read Committed (RC), Read
+Atomic (RA), Item Cut Isolation (I-CI), Predicate Cut Isolation
 (P-CI), Writes Follow Reads (WFR), Monotonic Reads (MR), Monotonic
@@
  \node[draw=none,fill=none] (ici) at (1.2, 0) {I-CI};
- \node[draw=none,fill=none] (pci) at (1.75, 1.2) {P-CI};
+ \node[draw=none,fill=none] (pci) at (1.75, .95) {P-CI};
  \node[draw=none,fill=none] (rc) at (-1.2, .95) {RC};
@@
 
- \node[draw=none,fill=none] (ta) at (-.2, 0) {TA};
+ \node[draw=none,fill=none] (ra) at (0, 1.475) {RA};
 
@@
  \node[noha] (cs) at (-1.2, 2) {CS};
- \node[noha] (rr) at (-.2, 2.7) {RR};
- \node[noha] (si) at (1.75, 2.3) {SI};
- \node[noha] (1sr) at (1.2, 3.2) {1SR};
+ \node[noha] (rr) at (0.2, 2.7) {RR};
+ \node[noha] (si) at (1.75, 2) {SI};
+ \node[noha] (1sr) at (1.75, 3.2) {1SR};
  \node[noha] (ssr) at (3.85, 3.6) {Strong-1SR};
@@
  \draw [->] (ru) -- (rc);
+ \draw [->] (rc) -- (ra);
  \draw [->] (ici) -- (pci);
@@
  \draw [->, red] (si) -- (1sr);
- \draw [->, red] (ta) -- (si);
- \draw [->, red] (ta) -- (rr);
+ \draw [->, red] (ra) -- (si);
+ \draw [->, red] (ra) -- (rr);
  \draw [->, red] (causal) -- (linearizable);
@@
   Atomicity} The relationship between linearizability and
-transactional atomicity is non-obvious. TA dictates that writes to
+transactional atomicity is non-obvious. RA dictates that writes to
 multiple keys across multiple servers are made visible to readers all
@@
 guarantee. Second, in linearizable systems, all clients see all writes
-at the same time. With TA as defined here, clients may see writes at
+at the same time. With RA as defined here, clients may see writes at
 different times depending on which replicas they contact. We are not
 aware of an analogous model in the distributed systems
-literature. Accordingly, despite apparent similarities, TA is
+literature. Accordingly, despite apparent similarities, RA is
 incomparable with and less coordination intensive (by availability

### 2013-07-22 writeup/hat-appendix.tex
Starting to address TA -> RA

@@
 
-\begin{definition}[Observed Transaction Vanishes (N-TA)]
-A history $H$ exhibits phenomenon N-TA if $DSG(H)$ contains a directed cycle
-consisting of exactly one read-dependency edge by $x$ from $T_i$ to
-$T_j$ and one anti-dependency edge by $y$ from $T_j$ to $T_i$ and
-$T_i$'s read from $y$ precedes its read from $x$.
+\begin{definition}[Observed Transaction Vanishes (N-RA)]
+A history $H$ exhibits phenomenon N-RA if $USG(H)$ contains a directed
+cycle consisting of exactly one read-dependency edge by $x$ from $T_j$
+to $T_i$ and a set of edges by $y$ containing at least one
+anti-dependency edge from $T_i$ to $T_j$ and $T_j$'s read from $y$
+precedes its read from $x$.
 \end{definition}
@@
 
-\begin{definition}[Transactional Atomicity (TA)]
-A system that provides transactional atomicity prohibits phenomenon
-N-TA.
+\begin{definition}[Read Atomic (RA)]
+A system that provides Read Atomic isolation prohibits phenomenon
+N-RA in addition to providing Read Committed isolation.
 \end{definition}
@@
 A system that provides Snapshot Isolation prevents phenomena G0, G1a,
-G1b, G1c, N-P-CI, N-TA, and Lost Update.
+G1b, G1c, N-P-CI, N-RA, and Lost Update.
 \end{definition}
@@
 \end{align*}
-For TA, the system must ensure that $a=1$. Consider a system with one
+For RA, the system must ensure that $a=1$. Consider a system with one
 replica for each of $x$ and $y$ (denoted $R_x$ and $R_y$). Say client
@@
 
-\vspace{1em}\noindent\textbf{Overheads.} TA requires overhead both on client and server.
+\vspace{1em}\noindent\textbf{Overheads.} RA requires overhead both on client and server.
 First, clients must store a vector (\texttt{required}) for the duration of a transaction in order to determine what values to read.

### 2013-07-24 writeup/evaluation.tex
Editing pass

@@
 
-With an understanding of which guarantees are HAT-compliant, in this
+With an understanding of which semantics are HAT-compliant, in this
 section, we analyze the implications of these results for existing
@@
 coordinating via a validation step, while deterministic transaction
-scheduling~\cite{deterministic-scheduling} requires a
+scheduling~\cite{deterministic-scheduling} requires contacting a
 scheduler. Serializability under multi-version concurrency control
@@
 distributed implementations of weak isolation are not highly
-available. Lock-based proposals such as those used to provide weak
-isolation in Gray's original proposal~\cite{gray-isolation} do not
+available. Lock-based mechanisms such as those in Gray's original proposal~\cite{gray-isolation} do not
 degrade gracefully in the presence of partial failures. (Note,
@@
 modification. Swift~\cite{swift} and bolt-on causal
-consistency~\cite{bolton} are closest to providing maximum sticky HAT
-semantics. As we have seen, it is possible to implement many
+consistency~\cite{bolton} are closest to providing maximal sticky HAT
+semantics. Accordingly, it is possible to implement many
 guarantees weaker than serializability---including guarantees that are
@@
 application-level consistency anomalies as a result of Lost Update or
-Write Skew anomalies. However, applications with non-monotonic state
+Write Skew anomalies. However, applications with non-monotone state
 mutation will not, in general, be able to maintain application-level
 consistency constraints with HATs alone---in particular, applications
-requiring bounded update visibility latency should opt for
-unavailability.
+requiring visibility bounds should opt for unavailability.
 
@@
 Update. Accordingly, HATs cannot provide compliant TPC-C execution but
-can still maintain uniqueness constraints. Second, the New-Order
+can maintain uniqueness constraints. Second, the New-Order
 transaction decrements inventory counts: what if the count becomes
-negative?  Fortunately, TPC-C New-Order restocks each item's inventory
+negative?  Fortunately, New-Order restocks each item's inventory
 count (increments by 91) if it would become negative as the result of
@@
 substantial latency penalties: over WAN, 10 to 100 times higher than
-their HAT counterparts. Given recent algorithms for providing
-causal consistency~\cite{bolton, cops, swift}, we focus our HAT
-analysis on RA (algorithm from Section~\ref{sec:ta}, described in
-greater detail
-in Appendix B
-%in our Technical Report~\cite{hat-tr}
-), which we believe to be the
-most interesting of our algorithms. We find that HAT
+their HAT counterparts. We find that HAT
 properties---while more resource-intensive than basic eventual
@@
 (i.e., new writes are sent to the WAL then subsequently moved into
-LevelDB once pending stable)---achieves ~75\% of the throughput. Latency
+LevelDB once stable)---achieves ~75\% of the throughput. Latency
 increases linearly with the number of YCSB clients due to contention
@@
 because worst-case server load was unaffected. With five clusters,
-RA's relative throughput decreased: every YCSB \texttt{put} operation
-resulted in four \texttt{put} operations on remote replicas and,
-accordingly, the cost of anti-entropy increased (e.g., each server
-processed four replica's anti-entropy as opposed to one before,
+\texttt{RA}'s relative throughput decreased: every YCSB \texttt{put}
+operation resulted in four \texttt{put} operations on remote replicas
+and, accordingly, the cost of anti-entropy increased (e.g., each
+server processed four replicas' anti-entropy as opposed to one before,
 reducing the opportunity for batching and decreasing available
@@
 garbage collection activity and, more importantly, IOPS when compared
-to Eventual and RC, causing RA throughput to peak at around half of
-Eventual. With in-memory persistence (i.e., no LevelDB or WAL),
-\texttt{RA} throughput was within 20\% of \texttt{eventual}.
+to \texttt{eventual} and \texttt{RC}, causing \texttt{RA} throughput
+to peak at around half of \texttt{eventual}. With in-memory
+persistence (i.e., no LevelDB or WAL), \texttt{RA} throughput was
+within 20\% of \texttt{eventual}.
 
@@
 transaction, \texttt{RA} throughput is within $60\%$ ($1898$ bytes
-overhead). This reflects our RA algorithm's metadata requirements,
+overhead). This reflects our \texttt{RA} algorithm's metadata requirements,
 which are proportional to transaction length and consume IOPS and

### 2013-07-24 writeup/hat-analysis.tex
Editing pass

@@
 mechanisms and may provide additional guarantees in addition to our
-implementation-agnostic definitions.  We draw largely on Atul Adya's
+implementation-agnostic definitions.  We draw largely on Adya's
 dissertation~\cite{adya} and somewhat on its predecessor work: the
@@
 Writes''~\cite{ansicritique}). If we build a graph of transactions
-with edges from one transaction to another if the former overwrites
+with edges from one transaction to another and, if the former overwrites
 the latter's write to the same object, then, under Read Uncommitted,
@@
 achievable and directly captures the spirit of the term: if a
-transaction reads the same data twice, it sees the same value each
+transaction reads the same data more than once, it sees the same value each
 time (preventing ``Fuzzy Read,'' or $P2$). In this paper, to
@@
 transactions' effects should succeed, is core to ACID
-guarantees. Although, at least by the ACID acronym, it is not an
+guarantees. Although, at least by the ACID acronym, atomicity is not an
 ``isolation'' property, atomicity properties also restrict the updates
@@
 the hierarchy of existing isolation properties, we place RA below
-Adya's $PL-2L$ (as it does not necessarily enforce transitive
+Adya's \textit{PL-2L} (as it does not necessarily enforce transitive
 read-write dependencies) but above Read Committed ($PL-2$). Notably,
@@
 arbitrary groups of non-colocated items is considerably more difficult
-to achieve with high availability: servers need to ensure that,
-regardless of replica selection, all of a client's accesses will read
-a sufficiently up-to-date/``consistent'' set of data items. However,
-replicas do not need to agree on when to choose to reveal sets of
-values to clients, which would require consensus. Instead, as we see
-below, RA only requires the equivalent of reliable broadcast (with
-some additional client metadata) and is achievable in HAT systems.
+to achieve with high availability.
 
@@
 
-As a straw-man, replicas can store all versions ever written for all
-items. Servers can gossip information about versions they have seen
-and construct a lower bound on the versions that can be found on every
-replica. At the start of each transaction, clients choose a bound that
-is guaranteed to be lower than or equal to the this global lower
-bound, and, during transaction execution, replicas return the latest
-version of each item that is not greater than the client's chosen
-bound. If the lower bound is advanced along transactional boundaries,
-clients will observe RA. This algorithm has several variants in the
-literature~\cite{readonly, swift}: bounds can be represented by vector
-clocks, and older versions can be asynchronously garbage collected.
+As a straw-man, replicas can store all versions ever written to each
+data item. Replicas can gossip information about versions they have
+observed and construct a lower bound on the versions that can be found
+on each replica. At the start of each transaction, clients choose a
+bound that is guaranteed to be lower than or equal to the this global
+lower bound, and, during transaction execution, replicas return the
+latest version of each item that is not greater than the client's
+chosen bound. If the lower bound is advanced along transactional
+boundaries, clients will observe RA. This algorithm has several
+variants in the literature~\cite{readonly, swift}: bounds can be
+represented by vector clocks, and older versions can be asynchronously
+garbage collected.
 
@@
 optimizations that, for brevity, we do not describe here.
-
-However, one key property with RA is that it do not dictate when
-writes become visible; accordingly, the algorithm is highly
-available. We can optimize write visibility via sticky availability by
-sticking clients are with disjoint groups of servers: once all of a
-transaction's writes are pending stable within a group of servers,
-writes can be made visible (as opposed to once they are pending stable
-with respect to all servers). For example, in a multi-datacenter
-setting, all clients and servers within a datacenter may form a sticky
-group.
 
@@
 session, then it will be able to read its writes. However, if the
-network topology changes and the client can only contact a different
-server that is partitioned from the server that executed $T_1$, then
-the system will have to either stall indefinitely to allow the client
-to read her writes (violating transactional availability) or will have
-to sacrifice read your writes guarantees. However, if the client
-remains sticky with the server that executed $T_1$, then we can
+network topology changes and the client can only execute $T_2$ on a
+different replica that is partitioned from the replica that executed
+$T_1$, then the system will have to either stall indefinitely to allow
+the client to read her writes (violating transactional availability)
+or will have to sacrifice read your writes guarantees. However, if the
+client remains sticky with the server that executed $T_1$, then we can
 disallow this scenario. Accordingly, read your writes, and, by proxy,
@@
 maintain limited criteria like foreign key constraints (via TA). We do
-not attempt to describe the entire space of application-level
-consistency properties that are achievable (see
-Section~\ref{sec:relatedwork}) but we specifically evaluate TPC-C
-transaction semantics with HAT guarantees in
-Section~\ref{sec:evaluation}.
+not describe the entire space of application-level consistency
+properties that are achievable (see Section~\ref{sec:relatedwork}) but
+we specifically evaluate TPC-C transaction semantics with HAT
+guarantees in Section~\ref{sec:evaluation}.
 
@@
 \end{align*}
-If we want to guarantee both cut isolation and transactional atomicity
-and the system only executes $T_1$, $T_2$, and $T_3$, then $T_3$ needs
-to read $a=b=\bot$, $a=b=1$, or $a=b=2$. This means that either the
+If we want to guarantee both cut isolation and RA isolation and the
+system only executes $T_1$, $T_2$, and $T_3$, then $T_3$ needs to read
+$a=b=\bot$, $a=b=1$, or $a=b=2$. This means that either the
 implementation should frequently return $\bot$ (definitely undesirable
@@
 COPS~\cite{cops}). Using client-side caching can alleviate some of
-these challenges~\cite{bolton, swift}, but then the system becomes
-sticky high available.
+these challenges~\cite{bolton, swift} at the expense of requiring
+sticky availability.
 
@@
 
-\vspace{.5em}\noindent\textbf{Linearizability and Transactional
+\vspace{.5em}\noindent\textbf{Linearizability and Read
   Atomicity} The relationship between linearizability and

### 2013-07-24 writeup/intro.tex
Editing pass

@@
 
-The last decade has seen a significant shift in the design of
-large-scale database systems, from the use of transactional
+The last decade has seen a shift in the design of popular large-scale
+database systems, from the use of transactional
 RDBMSs~\cite{bernstein-book, gray-isolation, gray-virtues} to the
 widespread adoption of loosely consistent distributed key-value
-stores~\cite{bigtable, pnuts, dynamo}. The 2000 introduction of
-Brewer's CAP Theorem, which stated that a highly available system
-cannot provide strong consistency guarantees in the presence of
-network partitions, helped spur this shift~\cite{brewer-slides}. As
+stores~\cite{bigtable, pnuts, dynamo}. Core to this shift was the 2000
+introduction of Brewer's CAP Theorem, which stated that a highly
+available system cannot provide ``strong'' consistency guarantees in
+the presence of network partitions~\cite{brewer-slides}. Yet, as
 formally proven~\cite{gilbert-cap}, the CAP Theorem concerns
@@
 implications of CAP---which concern a fairly narrow form of
-distributed consistency---have often been conflated with the ability
-to provide ACID database properties~\cite{hn,brewer-slides,
+distributed data consistency---have often been conflated with the
+ability to provide ACID database properties~\cite{hn,brewer-slides,
   foundation-article}; this misunderstanding has led to substantial
@@
 high availability. The recent resurgence of transactional systems like
-Spanner~\cite{spanner} suggests that programmers value these
+Spanner~\cite{spanner} suggests that programmers value transactional
 semantics, but, unfortunately, existing transactional data stores do
@@
 semantics provided by current highly available systems. This raises a
-natural question: which of these semantics can be provided with high
+natural question: which semantics can be provided with high
 availability?
@@
 they require detecting conflicts between concurrent updates (as needed
-for preventing Lost Updates or Write Skew phenomena), which we show
-is unavailable. However, Read Committed isolation, Transactional
-Atomicity, and many of the weaker data consistency models from
-database and distributed systems are achievable via algorithms that
-rely on multi-versioning and limited client-side caching. For several
+for preventing Lost Updates or Write Skew phenomena), which we show is
+unavailable. However, Read Committed isolation, transactionally atomic
+ioslation, and many of the weak consistency models from database and
+distributed systems are achievable via algorithms that rely on
+multi-versioning and limited client-side caching. For several
 guarantees, such as causal consistency with phantom prevention and
@@
 
-Our results demonstrate that a wide range of semantic guarantees can
-be delivered with availability in face of partitions, but which
-guarantees are worthwhile in practice? We study the virtues and
-limitations of both classic and HAT systems by surveying practitioner
-accounts and research literature, performing experimental analysis on
-modern cloud infrastructure, and analyzing representative applications
-for their semantic requirements. Our experiences with a HAT prototype
-running across multiple geo-replicated datacenters indicate that HATs
-offer a one to three order of magnitude latency decrease compared to
+At a high level, the virtues of HATs are guaranteed responses from any
+replica, low latency, and a range of semantic guarantees including
+``snapshot reads'' to transactional atomicity. However, highly
+available systems are fundamentally unable to prevent concurrent
+updates to shared data items and cannot provide recency guarantees for
+reads. To understand when these virtues and limitations are relevant
+in practice, we survey both practitioner accounts and academic
+literature, perform experimental analysis on modern cloud
+infrastructure, and analyze representative applications for their
+semantic requirements. Our experiences with a HAT prototype running
+across multiple geo-replicated datacenters indicate that HATs offer a
+one to three order of magnitude latency decrease compared to
 traditional distributed serializability protocols, and they can
@@
 non-monotonic logic, HATs can fall short, requiring
-sometimes-unavailable mechanisms that are more coordination
-intensive. Overall, this paper quantifies the benefits that Highly
-Available Transactions offer as well as the necessary restrictions
-that they place on applications.
+sometimes-unavailable mechanisms that are more
+coordination-intensive. Overall, this paper quantifies the benefits
+that Highly Available Transactions offer as well as the necessary
+restrictions that they place on applications.
 

### 2013-07-24 writeup/model.tex
Editing pass

@@
 they are guaranteed a response; this means servers will not need to
-communicate with any others. If servers are partitioned from one
+synchronously communicate with others. If servers are partitioned from one
 another, they do not need to stall in order to provide clients a
@@
 
-Traditionally, a system provides {\textbf{high availability}} if every user
-that can contact a correct (non-failing) server eventually receives a
-response from that server, even in the presence of arbitrary,
-indefinitely long network partitions between
-servers~\cite{gilbert-cap}.\footnote{Under this definition, systems
-  that require a majority of servers to be online is not
-  available. Similarly, a system which ensures that clients receive a
-  response with high (but not perfect) probability is not
-  available. This is a stringent definition but matches the
-  assumptions made by the CAP Theorem~\cite{gilbert-cap} and
+Traditionally, a system provides {\textbf{high availability}} if every
+user that can contact a correct (non-failing) server eventually
+receives a response from that server, even in the presence of
+arbitrary, indefinitely long network partitions between
+servers~\cite{gilbert-cap}.\footnote{Under this definition from the
+  distributed systems literuter, systems that require a majority of
+  servers to be online is not available. Similarly, a system which
+  ensures that clients receive a response with high (but not perfect)
+  probability is not available. This is a stringent but matches the
+  assumptions made in the CAP Theorem~\cite{gilbert-cap} and
   guarantees low latency~\cite{abadi-pacelc}.} As in a standard
@@
 is simple: a client can maintain stickiness by contacting the same
-server for each of its requests. However, in a partially-replicated
-system, where servers are replicas for subsets of the set of data
-items (which we consider in this paper), we say a client must remain
-sticky with a single \textit{logical} server that may consist of
-multiple physical servers.. We say that a system provides
+server for each of its requests. However, to stay ``sticky'' in a
+partially-replicated system, where servers are replicas for subsets of
+the set of data items (which we consider in this paper), a client must
+maintain stickiness with a single \textit{logical} server that may
+consist of multiple physical servers.. We say that a system provides
 \textbf{sticky availability} if, whenever a client's operation on an
@@
 transaction abort due to a transaction's own choosing (e.g., as an
-operation of the transaction itself or due to a would-be declared
-integrity constraint violation) an \textit{internal abort} and an
+operation of the transaction itself or due to a would-be violation of a declared integrity constraint) an \textit{internal abort} and an
 abort due to system implementation or operation an \textit{external

### 2013-07-24 writeup/motivation.tex
Editing pass

@@
 absence of failures, communication is slowed by factors like physical
-distance and, more often, congestion and routing. As we will see
+distance, network congestion, and routing. As we will see
 (Section~\ref{sec:availability}), highly available system design
@@
 ``availability zones'' (closely co-located datacenters), and within a
-single ``availability zone'' (datacenter), at a granularity of 1s
-(dataset to be released). We summarize the results of our network
-measurement study in Table~\ref{table:rtt}. On average,
-intra-datacenter communication (Table 1a) is between 1.82 and 6.38
-times faster than across geographically co-located datacenters (Table
-1b) and between 40 and 647 times faster than across geographically
-distributed datacenters (Table 1c). The cost of wide-area
-communication exceeds the speed of light: for example, while a
-speed-of-light RTT from S\~{a}o Paulo to Singapore RTT is 106.7ms,
-ping packets incur an average 362.8ms RTT (95th percentile: 649ms). As
-shown in Figure~\ref{fig:rtt}, the distribution of latencies varies
-between links, but the trend is clear: remote communication has a
-substantial cost. Like failures, quantifying and delays communication
-delays is an active area of research in the networking
-community~\cite{bobtail}.
+single ``availability zone'' (datacenter), at a granularity of
+1s\footnote{\url{http://www.bailis.org/blog/communication-costs-in-real-world-networks/}}. We
+summarize the results of our network measurement study in
+Table~\ref{table:rtt}. On average, intra-datacenter communication
+(Table 1a) is between 1.82 and 6.38 times faster than across
+geographically co-located datacenters (Table 1b) and between 40 and
+647 times faster than across geographically distributed datacenters
+(Table 1c). The cost of wide-area communication exceeds the speed of
+light: for example, while a speed-of-light RTT from S\~{a}o Paulo to
+Singapore RTT is 106.7ms, ping packets incur an average 362.8ms RTT
+(95th percentile: 649ms). As shown in Figure~\ref{fig:rtt}, the
+distribution of latencies varies between links, but the trend is
+clear: remote communication has a substantial cost. As with network
+failures, quantifying and delays communication delays is an active
+area of research in the networking community~\cite{bobtail}.
 
@@
 like Oracle 11g, which are known to power major businesses and product
-functionality. Given that these transactional models are frequently
+functionality. Given that these weak transactional models are frequently
 used, our inability to provide serializability in arbitrary HATs

### 2013-07-24 writeup/relatedwork.tex
Start of reviewer comments.
Editing pass

@@
 
-While we have discussed traditional mechanisms for distributed
+We have discussed traditional mechanisms for distributed
 coordination and several related systems in
-Section~\ref{sec:eval-existing}, in this section, we further discuss
+Section~\ref{sec:eval-existing}, but, in this section, we further discuss
 related work. In particular, we discuss related work on highly
@@
 techniques~\cite{kemme-classification} and distributed mechanisms for
-achieving one-copy serializability~\cite{wisemann-survey}.
+achieving one-copy serializability~\cite{wisemann-survey}. Two recent
+studies survey strategies for programming and reasoning about weak
+consistency~\cite{bailis-ec, bernstein-survey}.
 
@@
 are infinitely many HAT models (i.e., always reading value 1 is
-comparable with always returning value 2), but a recent report from UT
-Austin shows that no model stronger than causal consistency can be
+incomparable with always returning value 2), but a recent report from
+UT Austin shows that no model stronger than causal consistency can be
 achieved in a sticky highly available, \textit{one-way convergent}
@@
 eventually consistent stores; their work-in-progress report contains
-alternate formalism for several HAT guarantees and (snapshot)
+alternate formalism for several HAT guarantees and (snapshot-based)
 transactions~\cite{burkhardt-txns}.

### 2013-07-25 writeup/evaluation.tex
Removing a few references, squeezing space

@@
 distributed implementations of weak isolation are not highly
-available. Lock-based mechanisms such as those in Gray's original proposal~\cite{gray-isolation} do not
-degrade gracefully in the presence of partial failures. (Note,
-however, that lock-based protocols \textit{do} offer the benefit of
-recency guarantees.) While multi-versioned storage systems allow for a
-variety of transactional guarantees, few offer traditional weak
-isolation (e.g., non-``tentative update'' schemes) in this context.
-The MDCC~\cite{mdcc} protocol offers Read Committed isolation with
-Lost Update avoidance but is similarly unavailable due to its reliance
-on preventing write conflicts. Chan and Gray's read-only transactions
-have item-cut isolation with causal consistency and RA (session
-\textit{PL-2L}~\cite{adya}) but are unavailable in the presence of coordinator
-failure and assume serializable update transactions~\cite{readonly};
-this is similar to read-only and write-only transactions more recently
-proposed by Eiger~\cite{eiger}. Causal Serializability offers a
-similar model (with unavailable implementation): causal consistency
-with a variant of Read Uncommitted between transactions that write to
-the same data item~\cite{raynal-causal}.  Brantner's S3
-database~\cite{kraska-s3} and Bayou~\cite{sessionguarantees} can all
-provide variants of session \textit{PL-2L} with high availability, but none
-provide this HAT functionality without substantial
-modification. Swift~\cite{swift} and bolt-on causal
-consistency~\cite{bolton} are closest to providing maximal sticky HAT
-semantics. Accordingly, it is possible to implement many
-guarantees weaker than serializability---including guarantees that are
-achievable with high availability---and still not achieve high
-availability.
+available. Lock-based mechanisms such as those in Gray's original
+proposal~\cite{gray-isolation} do not degrade gracefully in the
+presence of partial failures. (Note, however, that lock-based
+protocols \textit{do} offer the benefit of recency guarantees.) While
+multi-versioned storage systems allow for a variety of transactional
+guarantees, few offer traditional weak isolation (e.g.,
+non-``tentative update'' schemes) in this context.  Chan and Gray's
+read-only transactions have item-cut isolation with causal consistency
+and RA (session \textit{PL-2L}~\cite{adya}) but are unavailable in the
+presence of coordinator failure and assume serializable update
+transactions~\cite{readonly}; this is similar to read-only and
+write-only transactions more recently proposed by
+Eiger~\cite{eiger}.  Brantner's S3 database~\cite{kraska-s3}
+and Bayou~\cite{sessionguarantees} can all provide variants of session
+\textit{PL-2L} with high availability, but none provide this HAT
+functionality without substantial modification. Swift~\cite{swift} and
+bolt-on causal consistency~\cite{bolton} are closest to providing
+maximal sticky HAT semantics. Accordingly, it is possible to implement
+many guarantees weaker than serializability---including guarantees
+that are achievable with high availability---and still not achieve
+high availability.
 

### 2013-07-25 writeup/hat-analysis.tex
Removing a few references, squeezing space

@@
 can be accomplished with well-known variants~\cite{causalmemory,
-  bolton, cops, sessionguarantees, swift} of the prior session
+  bolton, eiger, sessionguarantees, swift} of the prior session
 guarantee algorithms we presented earlier: only reveal new writes to
@@
 collection), or use pre-declared read sets to fetch a consistent cut
-of keys before each transaction begins to execute (e.g.,
-COPS~\cite{cops}). Using client-side caching can alleviate some of
-these challenges~\cite{bolton, swift} at the expense of requiring
-sticky availability.
+of keys before each transaction begins to execute. Using client-side
+caching can alleviate some of these challenges~\cite{bolton, swift} at
+the expense of requiring sticky availability.
 

### 2013-07-25 writeup/intro.tex
Removing a few references, squeezing space

@@
 distributed data consistency---have often been conflated with the
-ability to provide ACID database properties~\cite{hn,brewer-slides,
-  foundation-article}; this misunderstanding has led to substantial
-confusion regarding replica consistency, transactional isolation, and
-high availability. The recent resurgence of transactional systems like
-Spanner~\cite{spanner} suggests that programmers value transactional
-semantics, but, unfortunately, existing transactional data stores do
-not provide availability in the presence of
-partitions~\cite{middleware-db, foundation-article,
-  hstore,spanner,generalizedsnapshot, mdcc, krikellas-bargain, eiger,
-  walter, calvin}.
+ability to provide ACID database
+properties~\cite{hat-hotos,brewer-slides, foundation-article}; this
+misunderstanding has led to substantial confusion regarding replica
+consistency, transactional isolation, and high availability. The
+recent resurgence of transactional systems like Spanner~\cite{spanner}
+suggests that programmers value transactional semantics, but,
+unfortunately, existing transactional data stores do not provide
+availability in the presence of partitions~\cite{foundation-article,
+  hstore,spanner,eiger,  walter, calvin}.
 
@@
 serializability~\cite{bernstein-book, spanner, daudjee-session,
-  hstore, krikellas-bargain, calvin, kemme-classification} or snapshot
-isolation~\cite{daudjee-snapshot,generalizedsnapshot, kemme-snapshot,
-  walter}. Additionally, the distributed computing and parallel
+  hstore, calvin, kemme-classification} or snapshot
+isolation~\cite{daudjee-snapshot, walter}. Additionally, the distributed computing and parallel
 hardware literature contains many consistency models for single

### 2013-07-25 writeup/relatedwork.tex
Removing a few references, squeezing space

@@
 multi-object semantics, both in academia~\cite{kraska-s3, gstore,
-  mdcc, eiger, walter,calvin, swift} and industry~\cite{orleans,
-  spanner}. As discussed in Section~\ref{sec:modernacid}, classic ACID
-databases provide strong semantics but their lock-based and
-traditional multi-versioned implementations are unavailable in the
-presence of partitions~\cite{bernstein-book, gray-isolation}. Notably,
-Google's Spanner provides strong one-copy serializable
-transactions. While Spanner is highly specialized for Google's
-read-heavy workload, it relies on two-phase commit and two-phase
-locking for read/write transactions~\cite{spanner}. As we have
-discussed, the latency penalties associated with this design choice
-are fundamental to serializability. For users willing to tolerate
-these costs, Spanner, or similar strongly consistent, unavailable
-systems---including Calvin~\cite{calvin}, G-Store~\cite{gstore},
-Generalized Snapshot Isolation~\cite{generalizedsnapshot}, HBase,
-HStore~\cite{hstore}, MDCC~\cite{mdcc}, Orleans~\cite{orleans},
-Postgres-R~\cite{kemme-thesis}, Walter~\cite{walter}, and a range of
-snapshot isolation techniques~\cite{middleware-db, kemme-snapshot,
-  daudjee-snapshot}---are a reasonable choice.
+  eiger, walter,calvin, swift} and industry~\cite{spanner}. As
+discussed in Section~\ref{sec:modernacid}, classic ACID databases
+provide strong semantics but their lock-based and traditional
+multi-versioned implementations are unavailable in the presence of
+partitions~\cite{bernstein-book, gray-isolation}. Notably, Google's
+Spanner provides strong one-copy serializable transactions. While
+Spanner is highly specialized for Google's read-heavy workload, it
+relies on two-phase commit and two-phase locking for read/write
+transactions~\cite{spanner}. As we have discussed, the latency
+penalties associated with this design choice are fundamental to
+serializability. For users willing to tolerate these costs, Spanner,
+or similar strongly consistent, unavailable systems---including
+Calvin~\cite{calvin}, G-Store~\cite{gstore}, HBase, HStore~\cite{hstore}, Postgres-R~\cite{kemme-thesis},
+Walter~\cite{walter}, and a range of snapshot isolation
+techniques~\cite{daudjee-snapshot}---are a reasonable choice.
 

### 2013-07-25 writeup/hat-analysis.tex
next -> later

@@
 transaction $T_j$ reads a version of an object that transaction $T_i$
-wrote, then a later read by $T_j$ cannot return value whose next
+wrote, then a later read by $T_j$ cannot return value whose later
 version is installed by $T_i$. Together with item cut isolation, RA

### 2013-07-26 writeup/evaluation.tex
Cutting down length

@@
 general-purpose read-write transactions in a distributed
-setting~\cite{bernstein-book, davidson-survey, wisemann-survey}
+setting~\cite{bernstein-book, davidson-survey}
 require at least one RTT before committing. As an

### 2013-07-26 writeup/hat-analysis.tex
Cutting down length

@@
 caching can alleviate some of these challenges~\cite{bolton, swift} at
-the expense of requiring sticky availability.
-
-Composition cost may also vary by combination. For instance, Charron-Bost
- proved that, to capture causality between $N$ communicating
-processes, standard vector-based approaches face an upper bound of
-$O(N)$ storage per write~\cite{charron-bost}. This means that, with
-$100K$ clients, each write might be accompanied by $100K$ timestamps
-per vector. This is difficult to scale. By compromising on
-availability (e.g., treating a datacenter as a linearizable cluster),
-this overhead can be reduced~\cite{eiger}, but it is much
-cheaper to provide, say, read your writes guarantees, than causal
-consistency.
+the expense of requiring sticky availability. Study of the
+architectural interaction and overheads of various combinations is a
+promising area for future work.
 

### 2013-07-26 writeup/intro.tex
Cutting down length

@@
 serializability~\cite{bernstein-book, spanner, daudjee-session,
-  hstore, calvin, kemme-classification} or snapshot
+  hstore, calvin} or snapshot
 isolation~\cite{daudjee-snapshot, walter}. Additionally, the distributed computing and parallel

### 2013-07-26 writeup/motivation.tex
Cutting down length

@@
 single ``availability zone'' (datacenter), at a granularity of
-1s\footnote{\url{http://www.bailis.org/blog/communication-costs-in-real-world-networks/}}. We
+1s\footnote{Data available at \url{http://bailis.org/blog/communication-costs-in-real-world-networks/}}. We
 summarize the results of our network measurement study in

### 2013-07-26 writeup/relatedwork.tex
Cutting down length

@@
 studies examine weak isolation guarantees~\cite{adya,
-  ansicritique}. However, as we have discussed, the connection between
-weak, multi-object consistency models in a distributed setting and
-high availability is relatively unexplored. With the exception of our
+  ansicritique}. There are a wide range of coordination-avoiding
+``optimistic replication'' strategies~\cite{optimistic}, and there
+have been several recent efforts at further taxonomizing these
+strategies in light of current practice and the proliferation of
+``eventually consistent'' stores~\cite{bailis-ec, bernstein-survey}
+(notably, Bernstein and Das~\cite{bernstein-survey} specifically
+highlight the importance of stickiness~\cite{sessionguarantees,
+  vogels-defs}---which we formalize in Section~\ref{sec:sticky}).
+However, as we have discussed, the connection between weak,
+multi-object consistency models in a distributed setting and high
+availability is relatively unexplored. With the exception of our
 preliminary workshop paper discussing transactional availability,
@@
 and much greater discussion of implications and related work---we
-believe this paper is the first to explore this area. However, several
-studies rigorously classify \textit{non-HAT} semantics: Wisemann et
-al. have proposed a three parameter classification for one-copy
-serializable database systems, performing a more extensive
-classification than our overview in Section~\ref{sec:evaluation},
-including eager database replication
-techniques~\cite{kemme-classification} and distributed mechanisms for
-achieving one-copy serializability~\cite{wisemann-survey}. Two recent
-studies survey strategies for programming and reasoning about weak
-consistency~\cite{bailis-ec, bernstein-survey}.
+believe this paper is the first to explore this area.
 
@@
 or similar strongly consistent, unavailable systems---including
-Calvin~\cite{calvin}, G-Store~\cite{gstore}, HBase, HStore~\cite{hstore}, Postgres-R~\cite{kemme-thesis},
+Calvin~\cite{calvin}, G-Store~\cite{gstore}, HBase,
+HStore~\cite{hstore}, Postgres-R~\cite{kemme-thesis},
 Walter~\cite{walter}, and a range of snapshot isolation
-techniques~\cite{daudjee-snapshot}---are a reasonable choice.
+techniques~\cite{daudjee-session, daudjee-snapshot}---are a reasonable
+choice.
 
-With HATs, we seek an alternative set of semantics that are still
-useful but do not violate requirements for high availability or low
-latency. Recent systems proposals such as Swift~\cite{swift},
+With HATs, we seek an alternative set of transactional semantics that
+are still useful but do not violate requirements for high availability
+or low latency. Recent systems proposals such as Swift~\cite{swift},
 Eiger~\cite{eiger}, and Bolt-on Causal Consistency~\cite{bolton}

### 2013-07-31 writeup/reviewer-response/response.tex
Last round of comments from Alan and Aaron

\documentclass[10pt]{article}
\usepackage[margin=1.5in]{geometry}

\usepackage{times}

\renewcommand{\ttdefault}{fi4}

\newcommand{\reviewer}[1] {\noindent\texttt{#1}\\}

\newcommand{\response}[1] {\noindent\textit{Response: } #1\\}

\newenvironment{myitemize}
{
   \vspace{0mm}
    \begin{list}{$\bullet$ }{}
        \setlength{\topsep}{0em}
        \setlength{\parskip}{0pt}
        \setlength{\partopsep}{0pt}
        \setlength{\parsep}{0pt}         
        \setlength{\itemsep}{1mm} 
}
{
    \end{list} 
}

\thispagestyle{empty}
\begin{document}

\noindent\textbf{To the VLDB 2014 reviewers and editor:} Thank you for your
feedback. We have integrated your suggestions and believe that they
have strengthened the article, particularly with respect to clarity of
definitions and applicability of results. We address individual
comments in order below.\\

\noindent For your convenience, we have also included a diff of the text with our revision.\\

\noindent Thank you again for your consideration of our manuscript.\\

\noindent Peter Bailis\\
Aaron Davidson\\
Alan Fekete\\
Ali Ghodsi\\
Joseph M. Hellerstein\\
Ion Stoica

\newpage
\noindent\textbf{Reviewer 1 Comments}\\

\reviewer{W1. While the understanding and insights are useful, this paper still leaves it at an abstraction level. For instance, how can a practitioner use this knowledge to design systems? If availability is a requirement, do we chose the strongest form of consistency that is highly available or is there another tradeoff with performance (aka request latency). That is, reduction and application of these ideas to practice is unclear. A few compelling application scenarios (beyond TPC-C, which is not a great application for the target environment for HAT) would be beneficial.}

\response{We have added several use cases to the RA section (see below) and have
clarified takeaways both in the introduction. Additionally, we believe
that addressing several of the below comments has improved practical
importance.}

\reviewer{W2. Certain parts of the paper are unclear and/or hard to follow (details below). Clarifications will considerably improve the presentation.}

\response{We have addressed the specific suggestions below.}

\reviewer{D1. The stickiness of sessions is an important concept. While it appears repeatedly in the latter parts of the paper, the first definition is very easy to skip. It would be beneficial to somehow highlight this definition.}

\response{We have made "Sticky Availability" a separate subsection.}

\reviewer{[Elaborating W2]\\
D2. Section 5 mentions that stickiness can also be implemented with client-side caching. However, the implications of actual server stickiness and client-side caching are not the same. Ensuring causal consistency is a good differentiator. For instance, if the client session stays connected to the same server even after a network partition, causality can still be guaranteed. However, if stickiness is emulated by client-side caching, a client might never have seen a write on which it causally depends due to transitivity. Caching such writes at the client becomes impractical. This needs to be clarified.}

\response{Both the Swift system [Zawirski et al.] and Bolt-on Causal Consistency
[Bailis et al. 2013] achieve causality with client-side caching; we
have references to these systems.}

\reviewer{D3.I did not entirely follow the discussion on transaction atomicity. In my understanding, atomicity just says all or nothing. It does have a subtle relationship with isolation in the non-failure case. For a non-distributed transaction, if the isolation level allows dirty reads, it implies that if T2 read from what T1 (an uncommitted transaction) wrote, then T2's cannot commit until T1 commits as it would violate atomicity and if T1 aborts, T2 must also abort (causing cascading aborts), which is why dirty reads are often avoided. However, atomicity also has an implication with failures such that when a server fails, all in-flight transaction's work must be undone. Similarly, in the distributed case, atomicity means that either the transaction commits at all participating sites or at none of the sites. 2PC just gives you atomicity while an additional concurrency control mechanism, such as 2PL or OCC, is still needed for isolation. A clarification along these lines will be useful.}

\response{We have added a discussion clarifying the difference between the
Atomicity concept and the "subtle relationship with isolation."
Namely, we have refactored "Transactional Atomicity (TA)" into a
separate isolation property, "Read Atomic (RA)," which we think will
help avoid confusion by avoiding (further) overloading of the term
"Atomicity." This change has been made throughout the paper.}

\reviewer{D4. In section 5.1.2, the protocol to achieve replication with atomic broadcast is similar to [Kemme and Alonso, 2000]. Add a citation to it. Bettina Kemme, Gustavo Alonso: Don't Be Lazy, Be Consistent: Postgres-R, A New Way to Implement Database Replication. VLDB 2000: 134-143}

\response{We have added a citation to Kemme's thesis in this section.}

\reviewer{D5. For the discussion following Writes Follows Reads, is there an assumption that the underlying storage layer is multi-version? If not, then isn't holding back a client's write equivalent to blocking it, which will result in unavailability as the blocking can be indefinite. If multi-version, then wouldn't holding back writes potentially prevent the system from surfacing any recent writes? That is, is there a possibility that even though the system continues accepting writes, none of the subsequent reads from other sessions see those writes? In that case, the system might be available but not "alive." Elaborate on this.}

\response{We have updated the text.

Writes Follow Reads (and, more broadly, causal consistency) does not
presume a multi-version storage system. Rather, in both traditional
and recent implementations, incoming writes are *buffered* when they
are delivered to replicas but reads are never served from these
buffers [Ahamad et al. 1995, Lloyd et al. 2011, Lloyd et
al. 2013]. Perhaps this is a philosophical distinction, but we do not
believe the authors of these systems would characterize their systems
as multi-version in the tradition of classic database systems. There
*is*, however, a possible visibility penalty--writes may indeed not
become visible to other readers--which we explicity address in Section
5.2.2.}

\reviewer{D6. For causal consistency, why not use to standard and well-understood definition by Lamport, 1978? Leslie Lamport: Time, Clocks, and the Ordering of Events in a Distributed System. Commun. ACM 21(7): 558-565 (1978)\\
D7. Also, the discussion about causal consistency does not capture one essential property of causality: transitive dependencies due to the reads-from constraint.}

\response{We have clarified the discussion of "writes follow reads" to include a
connection to Lamport's "happens-before" relation. We have opted to
present causal consistency as it is treated in the literature on
session guarantees and believe that, with the suggested modifications,
this strikes a reasonable compromise.}

\reviewer{D8. In 5.2.1, it would be useful to add a discussion about PSI which is also not achievable. [Sovran et al, 2011]}

\response{We have added a reference to PSI in 5.2.1.}

\reviewer{D9. An almost concurrently-published survey seems very relevant to this paper. It might be worthwhile to contrast with that, especially since the current paper lacks much details on eventual consistency. Phil Bernstein and Sudipto Das: Rethinking Eventual Consistency, (to appear) SIGMOD 2013}

\response{We have added a reference and discussion in Related Work.}

\reviewer{D10. While this paper tries to sort out the clutter, there also seems to be a number of related papers published by the authors. Clearly articulating how this paper's contributions differ from that of the other papers will be beneficial. Papers I am referring to are: \\
* P. Bailis, A. Fekete, A. Ghodsi, J. M. Hellerstein, and I. Stoica. HAT, not CAP: \\
 Introducing Highly Available Transactions. In To appear in HotOS 2013.\\
* P. Bailis, A. Ghodsi, J. M. Hellerstein, and I. Stoica. Bolt-on causal consistency. In SIGMOD 2013. \\
* P. Bailis, A. Ghodsi: Eventual consistency today: limitations, extensions, and beyond, In CACM, May 2013.}

\response{We have placed references to the above in the related work section.}

\reviewer{D11. The paper is missing another citation, the paper that introduces 1SR. 
Rony Attar, Philip A. Bernstein, Nathan Goodman: Site Initialization, Recovery, and Backup in a Distributed Database System. IEEE Trans. Software Eng. 10(6): 645-650 (1984)}

\response{We have added a citation in Section 5.2.1.}

\reviewer{Minor: 
* Table 1 seems to be missing the latency numbers for the Singapore region which appears in the legend.}

\response{The Singapore numbers are located in the far right-hand side.}

\reviewer{* Midway through the paper, the style changes from emphasis to bold face. Make it consistent.}

\response{The style in the first part of the paper is now bold.}

\reviewer{Address D2--D11. If possible, W1.}

\response{We believe that the inline comments address this concern.}

\newpage
\noindent\textbf{Reviewer 2 Comments}\\

\reviewer{ - Feasibility solutions/arguments could be improved. }

\response{We believe that, in addressing the reviewers' comments, the presentation of feasibility has been improved.}

\reviewer{ - Detail/description of the TA algorithm would improve the paper. }

\response{We have added a simpler but less efficient TA algorithm to demonstrate
feasibility. We subsequently expand on this algorithm to explain our
more efficient algorithm (as originally presented). The relevant text
has been re-written.}

\reviewer{ - Better "proofs" would help keep paper self-contained.}

\response{We have revised several sections of Section 5 for better readability
and self-containment.}

\reviewer{ - Section 2 discusses why high availability is important, providing both anecdotal and experimental evidence for mitigating the effects of network partitions and inter-data center latencies. The latency data study provided in Section 2.2 is very nice and summarizes/motivates the paper nicely. While this paper is replete with references, but it might be appropriate to reference the Bobtail paper here (Xu et al., NSDI 2013). }

\response{We have added a reference to the Bobtail paper in Section 2.2.}

\reviewer{ - Section 3 discusses the need for various isolation levels in practice and provides a survey of the highest (and default) isolation levels provided by 18 "traditional" and NewSQL database systems. This work appeared in a previous workshop paper and the paper clearly mention this. While this survey helps motivate this work, it could be shortened in favor of more novel content (e.g., details of the TA implementation). }

\response{With this suggestion in mind, we have revised large parts of Section 5
(see below and other review comments). However, we believe Section 3
is a useful motivation for non-serializable isolation and have left
the table intact.}

\reviewer{ - Section 5 contains the main contribution of the paper by providing a discussion of what isolation guarantees are possible (and not possible) in a highly available distributed system. The paper uses classifications and terminology from from Adya (as well as Berenson et al.) in this discussion. This section lends clarity to the design space for highly-available transactions; a contribution that has potential to influence the future design of distributed storage systems. The classification of "sticky" and "non-sticky" availability is quite nice. The explanations and examples are incredibly clear and digestible. While some of the guarantee discussions sacrifice formality for brevity, it might be a good idea to bring in some formality to help the paper be better self-contained (that, or provide more implementation details to get the idea across). As is, the paper provides a single-paragraph algorithmic sketch for each HAT guarantee. Adding more detail about the TA algorithm would help greatly, since TA seems to be a contribution itself (reducing some of the "motivation" sections 2 and 3 could help). This would also help interpretation of the experimental results. This paper does have an associated technical report - and I expect a future journal article will combine this content (and more). }

\response{As we have noted above, "We have added a simpler but less efficient TA
algorithm to demonstrate feasibility. We subsequently expand on this
algorithm to explain our more efficient algorithm (as originally
presented). The relevant text has been re-written.}

\reviewer{ - Section 6 provides a nice discussion about the implications of HATs. The TPC-C analysis helps provide practical evidence for HAT guarantees in the "real-world" application space. }

\response{Thank you.}

\reviewer{ - A better "proof" for causality and PRAM guarantees would be helpful. As it stands, the paper refers to several references - a short sketch could go a long way to help the paper stay self-contained.}

\response{We have added a short sketch to the relevant section (within 5.1.3).}

\newpage
\noindent\textbf{Reviewer 3 Comments}\\

\reviewer{W1. Some definitions need to be clarified - see detailed comments. \\
W2. The experimental results are not very surprising (and not particularly compelling). For the 
record, I would have been in favor of accepting this paper even if it had had no evaluation. }

\reviewer{D1. The biggest weakness of this paper is that the its definition of high-availability is very strict - so strict that it excludes all so-called highly-available database systems that I am aware of: Spanner, Megastore, NoSQL systems like Cassandra, log shipping to hot standbys in RDBMS, and more. It is interesting to consider what kinds of consistency are achievable under such a strict HA requirement (as this paper does), but I wonder whether this will have any impact on practice.}

\response{We have expanded our explanation in Section 4.1. In brief, we have
focused on this stringent definition of high availability as it is the
definition of availability that appears in the distributed systems
literature and the proof of the CAP Theorem by Gilbert and Lynch. It
is indeed a restrictive model but guarantees low latency (see Abadi's
PACELC) and a guaranteed response under a large variety of failure
scenarios.}

\reviewer{D2. This paper focuses on the tradeoffs between availability and consistency. It does not mention durability of updates. You can't have a HAT system and also ensure that committed updates are geo-replicated (or replicated at all). It seems that HAT is a very demanding god... }

\response{This is an excellent point. We have added an additional Section
(5.2.3) to address this.}

\reviewer{D3. Section 2.1 is supposed to be about partitions, but many the
statistics being reported (e.g., in paragraph 3) are for link
failures, router failures, and other things (WAN and LAN) that may or
may not result in partitioning. Can you relate any of this stuff to
partitioning? Right now, it feels like you are hand-waving about
network-related failures in general.}

\response{We have added additional
statistics from the Microsoft paper, clarified wording on several
other references, and added an additional reference to a recent
survey.}

\reviewer{D4. Section (2nd paragraph) says that HA algs "ensure...guaranteed low latency". I see nothing about latency in any of the subsequent definitions. I understand that reduced latency can be a beneficial side-effect of designing for HA, but here you seem to be suggesting that the *definition* of HA has to do with low latency.}

\response{We have modified this sentence to reflect the fact that low latency is
a side effect of choosing high availability.}

\reviewer{D5. I'm puzzled by the definition of "sticky" availability. You indicate that a client be "sticky" by maintaining affinity with a set of servers, rather than a single server. However, the sticky availability definition seems to a require that a single server observe all of a client's requests. So, which is it? On the one hand, if not all servers can handle requests for all objects, then it would seem that you need to allow a client to be sticky with a set of servers. On the other hand, this means that no server observes all requests. Also, if a client can be sticky by sending its requests to a set of servers, then aren't all clients sticky (considering the set consisting of all servers). Since you are later going to actually classify algorithms based on this definition, I think it needs to be clarified.}

\response{This is an excellent point. In the case of fully-replicated systems,
the a client can remain sticky with a single server. But, as you point
out, in the case of partially-replicated systems, the definition is
more complex. We have amended the text to first handle the simple
fully-replicated case, then adapted the concept of stickiness for the
partially-replicated case: the distinction is that, in the latter,
clients interact with a single "logical" server that is split over
multiple physical servers. This is now fixed.}

\reviewer{D6. Transactional availability (Section 4.2) requires "replica availability", which is not defined. Does it correspond to what you previously called "traditional availability"? Also, why does the last sentence in Section 4.2 talk about the client's ability to contact *two* servers per data item?}

\response{We have added a definition of "replica availability" and removed the
problematic example.}

\reviewer{D7. Section 5.1.2 says that item cut isolation plus TA prevents read skew. I don't understand how this can be. TA allows a transaction that is reading multiple data items to observe only part of the effect of another transaction that updates those items - isn't that read skew? (And item cut isolation doesn't help.) More generally, I'm not really convinced of the utility of TA to applications - perhaps that explains its perplexing absence from existing treatments of weak isolation? }

\response{We have provided three scenarios in which we believe TA is useful.}

\reviewer{D8. It seems that the TA algorithm you propose (Section 5.1.2) will not play well with enforcing item cut isolation using caching, as you propose. If a transaction reads an item a second time, the latter may be trying to feed it an old value while former is compelled to feed a newer value (to preserve TA). You presumably want these two algs to play nicely together if TA plus item cut isolation somehow prevent read skew (in some way that I do not understand) as you claim. }

\response{This is an excellent point. We believe that presenting the simplified
TA algorithm presentation in Section 5.1.2 (see responses to reviewers
above) is easily integrated with Item Cut Isolation and should address
any concerns regarding feasibility. However, we have also noted your
concerns in Section 5.4: more efficient algorithms for I-CI plus RA
(nee TA) are an area for future work.}

\reviewer{D9. I don't see how the observations you make about application requirements 
(Section 6.2) support your conclusion about the utility of HAT at the end 
 of Section 6.2. }

\response{We have tempered the conclusion at the end of Section 6.2.}

\reviewer{D10. I don't understand the Master system you are using as a baseline in Section 6. There's a separate master per item? Which consistency guarantees is this baseline enforcing? I can't find that in the paper.}

\response{Our original wording was confusing; the "master" system is equivalent
to PNUTS's read-latest architecture. This represents single-key
linearizability, the guarantee addressed by the CAP Theorem.}

\end{document}

### 2013-07-31 writeup/hat-analysis.tex
Last round of comments from Alan and Aaron

@@
 HAT systems provide transactions with transactional availability and
-either high availability or sticky high availability. They offer
-latency and availability benefits over traditional distributed
-databases, yet they cannot achieve all possible semantics. In this
-section, we delineate ACID, distributed replica consistency, and
-session consistency levels which can be achieved with high
-availability (transactional atomicity, variants of Repeatable Read and
-traditionally weaker isolation, and many session guarantees), those
-with sticky high availability (read your writes, PRAM and causal
-consistency), and properties that cannot be provided in a HAT system
-(those preventing Lost Update and Write Skew or guaranteeing recency).
-We present a full summary of these results in
-Section~\ref{sec:hat-summary}.
+either high availability or sticky availability. They offer latency
+and availability benefits over traditional distributed databases, yet
+they cannot achieve all possible semantics. In this section, we
+delineate ACID, distributed replica consistency, and session
+consistency levels which can be achieved with high availability (Read
+Committed isolation, variants of Repeatable Read, atomic reads, and
+many session guarantees), those with sticky availability (read your
+writes, PRAM and causal consistency), and properties that cannot be
+provided in a HAT system (those preventing Lost Update and Write Skew
+or guaranteeing recency).  We present a full summary of these results
+in Section~\ref{sec:hat-summary}.
 
@@
 Writes''~\cite{ansicritique}). If we build a graph of transactions
-with edges from one transaction to another and, if the former overwrites
-the latter's write to the same object, then, under Read Uncommitted,
-the graph should not contain cycles~\cite{adya}. Consider the
-following example:
+with edges from one transaction to another and, when the former
+overwrites the latter's write to the same object, then, under Read
+Uncommitted, the graph should not contain cycles~\cite{adya}. Consider
+the following example:
 \begin{align*}
@@
 transactions' effects should succeed, is core to ACID
-guarantees. Although, at least by the ACID acronym, atomicity is not an
-``isolation'' property, atomicity properties also restrict the updates
-visible to other transactions. Accordingly, here, we consider the
-\textit{isolation} effects of atomicity, which we call \textbf{Read
-  Atomic (RA)} isolation.  Under RA, once some of the effects of a
-transaction $T_i$ are observed by another transaction $T_j$,
-thereafter, all effects of $T_i$ are observed by $T_j$. That is, if a
-transaction $T_j$ reads a version of an object that transaction $T_i$
-wrote, then a later read by $T_j$ cannot return value whose later
-version is installed by $T_i$. Together with item cut isolation, RA
-prevents Read Skew anomalies (Berenson et al.'s A5A) and is useful in
-several contexts such as maintaining foreign key constraints,
-consistent global secondary indexing, and maintenance of derived
-data. In the example below, under RA, because $T_2$ has read $T_1$'s
-write to $y$, $T_2$ must observe $b=c=1$ (or later versions for each
-key):
+guarantees. Although, at least by the ACID acronym, atomicity is not
+an ``isolation'' property, atomicity properties also restrict the
+updates visible to other transactions. Accordingly, here, we consider
+the \textit{isolation} effects of atomicity, which we call
+\textbf{Read Atomic (RA)} isolation.  Under RA, once some of the
+effects of a transaction $T_i$ are observed by another transaction
+$T_j$, thereafter, all effects of $T_i$ are observed by $T_j$. That
+is, if a transaction $T_j$ reads a version of an object that
+transaction $T_i$ wrote, then a later read by $T_j$ cannot return a
+value whose later version is installed by $T_i$. Together with item
+cut isolation, RA prevents Read Skew anomalies (Berenson et al.'s A5A)
+and is useful in several contexts such as maintaining foreign key
+constraints, consistent global secondary indexing, and maintenance of
+derived data. In the example below, under RA, because $T_2$ has read
+$T_1$'s write to $y$, $T_2$ must observe $b=c=1$ (or later versions
+for each key):
 \begin{align*}
@@
 
-As is typical in the database literature (with few
-exceptions~\cite{daudjee-session}), our models have not yet considered
-real-time or client-centric ordering \textit{across} transactions. In
-the distributed systems literature~\cite{sessionguarantees, vogels-defs}, many
-useful \textit{safety} guarantees span multiple, non-transactional
-operations. In particular, \textit{session guarantees} are used to
-describe guarantees across transactions within a given
-\textit{session}, ``an abstraction for the sequence of...operations
-performed during the execution of an
-application''~\cite{sessionguarantees}. Informally a session describes
-a context that should persist between transactions: for example, on a
-social networking site, all of a user's transactions submitted between
-``log in'' and ``log out'' operations might form a session.
+A useful class of safety guarantees refer to real-time or
+client-centric ordering within a \textit{session}, ``an abstraction
+for the sequence of...operations performed during the execution of an
+application''~\cite{sessionguarantees}. These ``session guarantees''
+have been explored in the distributed systems
+literature~\cite{sessionguarantees,vogels-defs} and sometimes in the
+database literature~\cite{daudjee-session}. For us, a session
+describes a context that should persist between transactions: for
+example, on a social networking site, all of a user's transactions
+submitted between ``log in'' and ``log out'' operations might form a
+session.
 
@@
 writes or, in the presence of partitions, make forward progress
-through the version history. The problem is that under non-sticky high
+through the version history. The problem is that under non-sticky
 availability, a system must handle the possibility that, under a
@@
 risk of invalidating application-level integrity constraints and can
-maintain limited criteria like foreign key constraints (via TA). We do
+maintain limited criteria like foreign key constraints (via RA). We do
 not describe the entire space of application-level consistency
@@
 
-Their need to prevent Lost Update means that Consistent Read, Snapshot
-Isolation (including Parallel Snapshot Isolation~\cite{walter}), and
-Cursor Stability guarantees are all unavailable.  Repeatable Read
-(defined by Gray~\cite{gray-isolation}, Berenson et
+Consistent Read, Snapshot Isolation (including Parallel Snapshot
+Isolation~\cite{walter}), and Cursor Stability guarantees are all
+unavailable because they require preventing Lost Update phenomena.
+Repeatable Read (defined by Gray~\cite{gray-isolation}, Berenson et
 al.~\cite{ansicritique}, and Adya~\cite{adya}) and One-Copy
-Serializability~\cite{1sr} need to prevent both Lost Update and Write Skew. These
-prevention requirements mean that these guarantees are inherently
-unavailable.
+Serializability~\cite{1sr} need to prevent both Lost Update and Write
+Skew. Their phenomena prevention requirements mean that these
+guarantees are inherently unachievable in a HAT system.
 
-\subsubsection{Unavailable Recency Guarantees}
+\subsubsection{Unachievable Recency Guarantees}
 
@@
 such as ``read a version that is no more than five seconds out of
-date'' or similar. None of these guarantees is HAT-compliant.
+date'' or similar. None of these guarantees are HAT-compliant.
 
@@
 A client requiring that its transactions' effects survive $F$ server
-faults requires at least $F+1$ non-failing replicas; this affects
+faults requires that the client be able to contact at least $F+1$
+non-failing replicas before committing. This affects
 availability and, according to the Gilbert and Lynch definition we
@@
 As we summarize in Table~\ref{table:hatcompared}, a wide range of
-isolation levels are achievable in HAT systems, including
-transactional atomicity, cut isolation, and several session
-guarantees. With sticky availability, a system can achieve read your
-writes guarantees and PRAM and causal consistency. However, many other
-prominent semantics, such as Snapshot Isolation, One-Copy
-Serializability, and Strict Serializability cannot be achieved due to
-the inability to prevent Lost Update and Write Skew phenomena.
+isolation levels are achievable in HAT systems. With sticky
+availability, a system can achieve read your writes guarantees and
+PRAM and causal consistency. However, many other prominent semantics,
+such as Snapshot Isolation, One-Copy Serializability, and Strong
+Serializability cannot be achieved due to the inability to prevent
+Lost Update and Write Skew phenomena.
 
@@
 achievable, but we find several particularly compelling. If we combine
-all sticky-HAT guarantees, we have transactional, causal snapshot
+all HAT and sticky guarantees, we have transactional, causal snapshot
 reads (i.e., Causal Transactional Predicate Cut Isolation). If we
@@
 (P-CI), Writes Follow Reads (WFR), Monotonic Reads (MR), Monotonic
-Writes (MW)\\\hline Sticky HA & Read Your Writes (RYW), PRAM,
+Writes (MW)\\\hline Sticky & Read Your Writes (RYW), PRAM,
 Causal\\\hline Unavailable & Cursor Stability (CS)\lostupdate,
@@
 \end{tabular}
-\caption{Summary of highly available, sticky highly available, and
+\caption{Summary of highly available, sticky available, and
   unavailable models considered in this paper. Unavailable models are
@@
   update\lostupdate, preventing write skew\rwskew, and requiring
-  recency guarantees\linearizable.}
+  recency guarantees\linearizable.}\vspace{-2em}
 \label{table:hatcompared}
@@
 \label{fig:hat-order}
-\caption{Partial ordering of HAT, sticky HAT (in boxes, blue), and
+\caption{Partial ordering of HAT, sticky available (in boxes, blue), and
   unavailable models (circled, red) from
@@
 In this section, we discuss several subtleties in our results,
-specifically addressing model composition, transactional atomicity
+specifically addressing model composition, read atomicity
 versus linearizability, and benefits of stickiness.
@@
   Atomicity} The relationship between linearizability and
-transactional atomicity is non-obvious. RA dictates that writes to
+RA is non-obvious. RA dictates that writes to
 multiple keys across multiple servers are made visible to readers all
@@
 writes are made visible to clients \textit{immediately} after they
-finish. With transactional atomicity, there is no recency
+finish. With RA, there is no recency
 guarantee. Second, in linearizable systems, all clients see all writes

### 2013-07-31 writeup/hat-appendix.tex
Last round of comments from Alan and Aaron

@@
 transaction. The definitions in this section are directly from Adya,
-with two difference. First, we expand Adya's formalism to deal
+with two differences. First, we expand Adya's formalism to deal
 with \textit{per-item} dependencies. Second, we define session
-dependencies (Definitions~\ref{def:sd1} and~\ref{def:sd2}).
+dependencies (Definition~\ref{def:sd1}).
 
@@
 
-\begin{definition}[Directly Read-Depends]
+\begin{definition}[Directly Read-Depends{~\cite[Definition 2]{adya}}]
 $T_j$ directly read-depends on transaction $T_i$ if it directly
@@
 
-\begin{definition}[Directly Anti-Depends]
+\begin{definition}[Directly Anti-Depends{~\cite[Definition 4]{adya}}]
 Transaction $T_j$ directly anti-depends on transaction $T_i$ if it
@@
 
-\begin{definition}[Directly Write-Depends]
+\begin{definition}[Directly Write-Depends{~\cite[Definition 5]{adya}}]
 A transaction $T_j$ directly write-depends on transaction $T_i$ if
@@
 $T_i \overset{wr_x}\DashedArrow T_j$. If $T_j$ session-depends on
-$T_i$ in session $S$, we draw $T_i \overset{S}\longrightarrow T_j$.
+$T_i$ in session $S$, we draw $T_i \overset{S}\longrightarrow
+T_j$~\cite[Definition 8]{adya}.
+
+We also consider the Unfolded Serialization Graph ($USG(H)$) that is a
+variation of the $DSG$. The USG is speciﬁed for the transaction of
+interest, $T_i$, and a history, $H$, and is denoted by $USG(H,
+T_i)$. For the USG, we retain all nodes and edges of the $DSG$ except
+for $T_i$ and the edges incident on it. Instead, we split the node for
+$T_i$ into multiple nodes---one node for every read/write event in
+$T_i$. The edges are now incident on the relevant event of $T_i$.
+
+$USG(H, T_i)$ is obtained by transforming $DSG(H)$ as follows:.  For
+each node $p$ ($p \neq T_i$) in $DSG(H)$, we add a node to
+$USG(H,T_i)$. For each edge from node $p$ to node $q$ in $DSG(H)$,
+where p and q are different from $T_i$, we draw a corresponding edge
+in $USG(H,T_i)$. Now we add a node corresponding to every read and
+write performed by $T_i$. Any edge that was incident on $T_i$ in the
+$DSG$ is now incident on the relevant event of $T_i$ in the
+$USG$. Finally, consecutive events in $T_i$ are connected
+by \textit{order edges}, e.g., if an action (e.g., SQL statement)
+reads object $y_j$ and immediately follows a write on object $x$ in
+transaction $T_i$, we add an order-edge from $w_i(x_i)$ to
+$r_i(y_j)$~\cite[Section 4.2.1]{adya}.
 
@@
 
-\begin{definition}[Item-Many-Preceders (N-I-CI)]
-\label{def:n-ici}
-A history $H$ exhibits phenomenon N-ICI if $DSG(H)$ contains a
+\begin{definition}[Item-Many-Preceders (IMP)]
+\label{def:imp}
+A history $H$ exhibits phenomenon IMP if $DSG(H)$ contains a
 transaction $T_i$ such that $T_i$ directly item-read-depends by $x$ on more
@@
 \end{align*}
-\caption{Example of \textit{N-ICI} anomaly.}
+\caption{Example of \textit{IMP} anomaly.}
 \label{fig:nici-history}
@@
 \begin{definition}[Item Cut Isolation (I-CI)]
-A system that provides Item Cut Isolation prohibits phenomenon N-I-CI.
+A system that provides Item Cut Isolation prohibits phenomenon IMP.
 \end{definition}
 
-\begin{definition}[Predicate-Many-Preceders (N-P-CI)]
-A history $H$ exhibits phenomenon N-I-CI if, for all predicate-based
+\begin{definition}[Predicate-Many-Preceders (PMP)]
+A history $H$ exhibits phenomenon PMP if, for all predicate-based
 reads $r_i(P_i:Vset(P_i))$ and $r_j(P_j:Vset(P_j)$ in $T_k$ such that
@@
 \begin{definition}[Predicate Cut Isolation (P-CI)]
-A system that provides Predicate Cut Isolation prohibits phenomenon N-P-CI.
+A system that provides Predicate Cut Isolation prohibits phenomenon PMP.
 \end{definition}
 
-\begin{definition}[Observed Transaction Vanishes (N-RA)]
-A history $H$ exhibits phenomenon N-RA if $USG(H)$ contains a directed
+\begin{definition}[Observed Transaction Vanishes (OTV)]
+A history $H$ exhibits phenomenon OTV if $USG(H)$ contains a directed
 cycle consisting of exactly one read-dependency edge by $x$ from $T_j$
@@
 \end{align*}
-\caption{Example of \textit{N-TA} anomaly.}
+\caption{Example of \textit{OTV} anomaly.}
 \label{fig:nta-history}
@@
 A system that provides Read Atomic isolation prohibits phenomenon
-N-RA in addition to providing Read Committed isolation.
+OTV in addition to providing Read Committed isolation.
 \end{definition}
@@
 
-\begin{definition}[Writes Do Not Follow Reads (N-WFR)]
-A history $H$ exhibits phenomenon N-WFR if, in $DSG(H)$, for all
+\begin{definition}[Missing Read-Write Dependency (MRWD)]
+A history $H$ exhibits phenomenon MRWD if, in $DSG(H)$, for all
 committed transactions $T_1$, $T_2$, $T_3$ such that $T_2$
@@
 \end{align*}
-\caption{Example of \textit{N-WFR} anomaly.}
+\caption{Example of \textit{MRWD} anomaly.}
 \label{fig:nwfr-history}
@@
 \begin{definition}[Writes Follow Reads (WFR)]
-A system provides Writes Follow Reads if it prohibits phenomenon N-WFR.
+A system provides Writes Follow Reads if it prohibits phenomenon MWRD.
 \end{definition}
 
-\begin{definition}[Missing Your Writes (N-RYW)]
-A history $H$ exhibits phenomenon N-RYW if $DSG(H)$ contains a directed cycle
+\begin{definition}[Missing Your Writes (MYR)]
+A history $H$ exhibits phenomenon MYR if $DSG(H)$ contains a directed cycle
 consisting of a transitive session-dependency between transactions
@@
 \end{align*}
-\caption{Example of \textit{N-RYW} anomaly if $T_2$ directly session-depends on $T_1$.}
+\caption{Example of \textit{MYR} anomaly if $T_2$ directly session-depends on $T_1$.}
 \label{fig:nryw-history}
@@
 \begin{definition}[Read Your Writes (RYW)]
-A system provides Read Your Writes if it prohibits phenomenon N-RYW.
+A system provides Read Your Writes if it prohibits phenomenon MYR.
 \end{definition}
@@
 A system provides PRAM Consistency if it prohibits phenomenon N-MR,
-N-MW, and N-RYW.
+N-MW, and MYR.
 \end{definition}
@@
 A system provides Causal Consistency if it provides PRAM Consistency
-and prohibits phenomenon N-WFR.
+and prohibits phenomenon MWRD.
 \end{definition}
@@
 A system that provides Snapshot Isolation prevents phenomena G0, G1a,
-G1b, G1c, N-P-CI, N-RA, and Lost Update.
+G1b, G1c, PMP, OTV, and Lost Update.
 \end{definition}
@@
 
-\section{Transactional Atomicity}
+\section{Read Atomic Isolation}
 
 In this section, we provide additional information on our two-phase
-Transactional Atomicity algorithm.
+algorithm for providing Read Atomic isolation.
 
@@
 \begin{algorithmic}
-\caption*{\textbf{Algorithm} HAT Transactional Atomicity}\vspace{.5em}
+\caption*{\textbf{Algorithm} HAT Read Atomic Isolation}\vspace{.5em}
 \State{\textit{\textbf{Shared Data Types}}\\\vspace{.5em}

### 2013-07-31 writeup/hat-vldb.tex
Editing pass
Removing a few references, squeezing space
Last round of comments from Alan and Aaron

@@
 \usepackage{scrextend, MnSymbol}
+\usepackage[hyphens]{url}
 
 \usepackage[hidelinks]{hyperref}
+
+\usepackage{breakurl}
 
@@
 
-\usepackage{times, graphicx, url, xcolor, multicol, minipage, subfig, tikz, amsthm, float}
+\usepackage{times, graphicx, xcolor, multicol, minipage, subfig, tikz, amsthm, float}
 \usepackage{balance}  % for  \balance command ON LAST PAGE  (only there!)
@@
 
-\titlespacing{\section}{1ex}{2ex}{1ex}
+\titlespacing{\section}{1ex}{1ex}{0.25ex}
 
-\titlespacing{\subsection}{1ex}{1ex}{0.25ex}
-\titlespacing{\subsubsection}{1ex}{1ex}{0.25ex}
+\titlespacing{\subsection}{1ex}{0.5ex}{0.25ex}
+\titlespacing{\subsubsection}{1ex}{0.25ex}{0.25ex}
 
@@
 identify which can and cannot be achieved in HAT systems. This unifies
-the literature on weak isolation, replica consistency, and highly
-available systems. We analytically and experimentally quantify the
-availability and performance benefits of HATs---often two to three
-orders of magnitude over wide-area networks---as well as their
-necessary semantic compromises.
-\end{abstract}\vspace{-.5em}
+the literature on weak transactional isolation, replica consistency,
+and highly available systems. We analytically and experimentally
+quantify the availability and performance benefits of HATs---often two
+to three orders of magnitude over wide-area networks---as well as
+their necessary semantic compromises.
+\end{abstract}
 

### 2013-07-31 writeup/intro.tex
Last round of comments from Alan and Aaron

@@
 available system cannot provide ``strong'' consistency guarantees in
-the presence of network partitions~\cite{brewer-slides}. Yet, as
-formally proven~\cite{gilbert-cap}, the CAP Theorem concerns
-linearizability: the ability to read the most recent write to a data
-item that is replicated across servers~\cite{herlihy-art}. Yet the
-implications of CAP---which concern a fairly narrow form of
-distributed data consistency---have often been conflated with the
-ability to provide ACID database
-properties~\cite{hat-hotos,brewer-slides, foundation-article}; this
+the presence of network partitions~\cite{brewer-slides}. As formally
+proven~\cite{gilbert-cap}, the CAP Theorem pertains to a data
+consistency model called linearizability, or the ability to read the
+most recent write to a data item that is replicated across
+servers~\cite{herlihy-art}. However, despite its narrow scope, the CAP
+Theorem is often misconstrued as a broad result regarding the ability
+to provide ACID database properties with high
+availability~\cite{hat-hotos,brewer-slides, foundation-article}; this
 misunderstanding has led to substantial confusion regarding replica
@@
 unfortunately, existing transactional data stores do not provide
-availability in the presence of partitions~\cite{foundation-article,
-  hstore,spanner,eiger,  walter, calvin}.
+availability in the presence of partitions~\cite{orleans,foundation-article,
+  hstore,spanner,eiger, walter,calvin}.
 
@@
 
-Our main contributions in this paper are as follows. We fill the gap
-between the many previously proposed isolation and consistency models
-and the goal of high availability.  We classify which among the wide
-array of models are achievable with high availability, denoting them
-as {\em Highly Available Transactions} (HATs). In doing so, we
-demonstrate that although many implementations of HAT semantics are
-not highly available, this is an artifact of the implementations
-rather than an inherent property of the semantics. Our investigation
-shows that, besides serializability, Snapshot Isolation and Repeatable
-Read isolation are not HAT-compliant, while most other isolation
-levels are achievable with high availability. We also demonstrate that
-many weak replica consistency models from distributed systems are both
-HAT-compliant and simultaneously achievable with several ACID
-properties.
+Our main contributions in this paper are as follows. We relate the
+many previously proposed isolation and consistency models to the goal
+high availability, which guarantees a response from every data replica
+in the presence of arbitrary partitions between them.  We classify
+which among the wide array of models are achievable with high
+availability, denoting them as {\em Highly Available Transactions}
+(HATs). In doing so, we demonstrate that although many implementations
+of HAT semantics are not highly available, this is an artifact of the
+implementations rather than an inherent property of the semantics. Our
+investigation shows that, besides serializability, Snapshot Isolation
+and Repeatable Read isolation are not HAT-compliant, while most other
+isolation levels are achievable with high availability. We also
+demonstrate that many weak replica consistency models from distributed
+systems are both HAT-compliant and simultaneously achievable with
+several ACID properties.
 
@@
 for preventing Lost Updates or Write Skew phenomena), which we show is
-unavailable. However, Read Committed isolation, transactionally atomic
-ioslation, and many of the weak consistency models from database and
-distributed systems are achievable via algorithms that rely on
-multi-versioning and limited client-side caching. For several
-guarantees, such as causal consistency with phantom prevention and
-ANSI Repeatable Read, we consider a modified form of high availability
-in which clients ``stick to'' (i.e., have affinity with) at least one
-server---a property which is often implicit in the distributed systems
-literature~\cite{herlihy-art, eiger, cac} but which requires explicit
-consideration in a client-server replicated database context. This
-sticky availability is widely employed~\cite{eiger, vogels-defs} but
-is less restrictive (and therefore more easily achievable) than
-traditional high availability.
+unavailable. However, Read Committed isolation, transactional
+atomicity (Read Atomic isolation), and many of the weak consistency
+models from database and distributed systems are achievable via
+algorithms that rely on multi-versioning and limited client-side
+caching. For several guarantees, such as causal consistency with
+phantom prevention and ANSI Repeatable Read, we consider a modified
+form of high availability in which clients ``stick to'' (i.e., have
+affinity with) at least one server---a property which is often
+implicit in the distributed systems literature~\cite{herlihy-art,
+  eiger, cac} but which requires explicit consideration in a
+client-server replicated database context. This sticky availability is
+widely employed~\cite{eiger, vogels-defs} but is a less restrictive
+model (and therefore more easily achievable) than traditional high
+availability.
 
@@
 replica, low latency, and a range of semantic guarantees including
-``snapshot reads'' to transactional atomicity. However, highly
-available systems are fundamentally unable to prevent concurrent
-updates to shared data items and cannot provide recency guarantees for
-reads. To understand when these virtues and limitations are relevant
-in practice, we survey both practitioner accounts and academic
-literature, perform experimental analysis on modern cloud
-infrastructure, and analyze representative applications for their
-semantic requirements. Our experiences with a HAT prototype running
-across multiple geo-replicated datacenters indicate that HATs offer a
-one to three order of magnitude latency decrease compared to
+several whose usefulness is widely acepted such as Read
+Committed. However, highly available systems are fundamentally unable
+to prevent concurrent updates to shared data items and cannot provide
+recency guarantees for reads. To understand when these virtues and
+limitations are relevant in practice, we survey both practitioner
+accounts and academic literature, perform experimental analysis on
+modern cloud infrastructure, and analyze representative applications
+for their semantic requirements. Our experiences with a HAT prototype
+running across multiple geo-replicated datacenters indicate that HATs
+offer a one to three order of magnitude latency decrease compared to
 traditional distributed serializability protocols, and they can

### 2013-07-31 writeup/model.tex
Cutting down length
Last round of comments from Alan and Aaron

@@
 servers~\cite{gilbert-cap}.\footnote{Under this definition from the
-  distributed systems literuter, systems that require a majority of
-  servers to be online is not available. Similarly, a system which
+  distributed systems literature, systems that require a majority of
+  servers to be online are not available. Similarly, a system which
   ensures that clients receive a response with high (but not perfect)
-  probability is not available. This is a stringent but matches the
-  assumptions made in the CAP Theorem~\cite{gilbert-cap} and
-  guarantees low latency~\cite{abadi-pacelc}.} As in a standard
+  probability is not available. This is a stringent requirement but it
+  matches the assumptions made in the CAP Theorem~\cite{gilbert-cap}
+  and guarantees low latency~\cite{abadi-pacelc}.} As in a standard
 distributed database, designated servers might perform operations for
@@
 \subsection{Sticky Availability}
+\label{sec:sticky}
 
@@
 maintain stickiness with a single \textit{logical} server that may
-consist of multiple physical servers.. We say that a system provides
+consist of multiple physical servers. We say that a system provides
 \textbf{sticky availability} if, whenever a client's operation on an

### 2013-07-31 writeup/relatedwork.tex
Last round of comments from Alan and Aaron

@@
   ansicritique}. There are a wide range of coordination-avoiding
-``optimistic replication'' strategies~\cite{optimistic}, and there
-have been several recent efforts at further taxonomizing these
-strategies in light of current practice and the proliferation of
-``eventually consistent'' stores~\cite{bailis-ec, bernstein-survey}
-(notably, Bernstein and Das~\cite{bernstein-survey} specifically
-highlight the importance of stickiness~\cite{sessionguarantees,
-  vogels-defs}---which we formalize in Section~\ref{sec:sticky}).
-However, as we have discussed, the connection between weak,
-multi-object consistency models in a distributed setting and high
-availability is relatively unexplored. With the exception of our
-preliminary workshop paper discussing transactional availability,
-real-world ACID, and HAT RC and I-CI~\cite{hat-hotos}---which this
-work expands, providing additional algorithms and impossibility
-results, a full taxonomization, experimental and application analysis,
-and much greater discussion of implications and related work---we
-believe this paper is the first to explore this area.
+``optimistic replication'' strategies~\cite{optimistic} and several
+recent efforts at further taxonomizing these strategies in light of
+current practice and the proliferation of ``eventually consistent''
+stores~\cite{bailis-ec, bernstein-survey} (notably, Bernstein and
+Das~\cite{bernstein-survey} specifically highlight the importance of
+stickiness~\cite{sessionguarantees, vogels-defs}---which we formalize
+in Section~\ref{sec:sticky}).  However, as we have discussed, the
+connection between weak, multi-object consistency models in a
+distributed setting and high availability is relatively
+unexplored. With the exception of our preliminary workshop paper
+discussing transactional availability, real-world ACID, and HAT RC and
+I-CI~\cite{hat-hotos}---which this work expands, providing additional
+algorithms and impossibility results, a full taxonomization,
+experimental and application analysis, and a greater discussion of
+implications and related work---we believe this paper is the first to
+explore this area.
 
@@
 multi-object semantics, both in academia~\cite{kraska-s3, gstore,
-  eiger, walter,calvin, swift} and industry~\cite{spanner}. As
+  eiger, walter,calvin, swift} and industry~\cite{orleans,spanner}. As
 discussed in Section~\ref{sec:modernacid}, classic ACID databases
@@
 Calvin~\cite{calvin}, G-Store~\cite{gstore}, HBase,
-HStore~\cite{hstore}, Postgres-R~\cite{kemme-thesis},
+HStore~\cite{hstore}, Orleans~\cite{orleans}, Postgres-R~\cite{kemme-thesis},
 Walter~\cite{walter}, and a range of snapshot isolation
@@
 et al. have concurrently developed an axiomatic specification for
-eventually consistent stores; their work-in-progress report contains
-alternate formalism for several HAT guarantees and (snapshot-based)
-transactions~\cite{burkhardt-txns}.
+eventual consistency; their work-in-progress report contains alternate
+formalism for several HAT guarantees~\cite{burkhardt-txns}.
 

### 2013-08-01 writeup/hat-vldb.tex
Last edits

@@
 \newcommand{\pbnote}[1]{{\color{red}{#1 ---P.B.}}}
+
+\newif\iftechreport
+%\techreporttrue
+\techreportfalse
 
@@
 
-\title{Highly Available Transactions: Virtues and Limitations\\(Extended Version)}
+\title{Highly Available Transactions: Virtues and Limitations\iftechreport\\(Extended Version)\fi}
 {\author{Peter Bailis{\fontsize{12}{14}$^\dagger$}, Aaron Davidson{\fontsize{12}{14}$^\dagger$}, Alan Fekete{\fontsize{12}{14}$^\diamond$}, Ali Ghodsi{\fontsize{12}{14}$^{\dagger,\ddagger}$}, Joseph M.
@@
 % Example of an appendix; typically would start on a new page
-\pagebreak
 
+\iftechreport
 \pagebreak
@@
 \end{appendix}
+\fi
 \end{document}

### 2013-08-01 writeup/model.tex
Last edits

@@
   servers to be online are not available. Similarly, a system which
-  ensures that clients receive a response with high (but not perfect)
-  probability is not available. This is a stringent requirement but it
-  matches the assumptions made in the CAP Theorem~\cite{gilbert-cap}
-  and guarantees low latency~\cite{abadi-pacelc}.} As in a standard
+  guarantees that servers provide a response with high probability is
+  not available. This admittedly stringent requirement matches the
+  assumptions made in the CAP Theorem~\cite{gilbert-cap} and
+  guarantees low latency~\cite{abadi-pacelc}.} As in a standard
 distributed database, designated servers might perform operations for

### 2013-08-01 writeup/relatedwork.tex
Last edits

@@
 serializability. For users willing to tolerate these costs, Spanner,
-or similar strongly consistent, unavailable systems---including
-Calvin~\cite{calvin}, G-Store~\cite{gstore}, HBase,
-HStore~\cite{hstore}, Orleans~\cite{orleans}, Postgres-R~\cite{kemme-thesis},
+or similar strongly consistent, unavailable
+systems~\cite{kemme-classification}---including Calvin~\cite{calvin},
+G-Store~\cite{gstore}, HBase, HStore~\cite{hstore},
+Orleans~\cite{orleans}, Postgres-R~\cite{kemme-thesis},
 Walter~\cite{walter}, and a range of snapshot isolation
@@
 incomparable with always returning value 2), but a recent report from
-UT Austin shows that no model stronger than causal consistency can be
-achieved in a sticky highly available, \textit{one-way convergent}
+UT Austin shows that no model stronger than causal consistency is
+achievable in a sticky highly available, \textit{one-way convergent}
 system~\cite{cac}. This result is promising and complementary to our

### 2013-08-01 writeup/reviewer-response/response.tex

@@
 
-\noindent For your convenience, we have also included a diff of the text with our revision.\\
-
 \noindent Thank you again for your consideration of our manuscript.\\
@@
 
-\response{We have added several use cases to the RA section (see below) and have
-clarified takeaways both in the introduction. Additionally, we believe
-that addressing several of the below comments has improved practical
-importance.}
+\response{We have added several use cases to the RA section (see
+  below) and have clarified takeaways both in the introduction and in
+  the analysis section. Additionally, we believe that addressing
+  several of the below comments has improved practical importance.}
 
@@
 
-\response{We have made "Sticky Availability" a separate subsection.}
+\response{We have made "Sticky Availability" a separate subsection (now Section 4.2).}
 
@@
 [Bailis et al. 2013] achieve causality with client-side caching; we
-have references to these systems.}
+have provided references to these systems.}
 
@@
 
-\response{We have updated the text.
-
-Writes Follow Reads (and, more broadly, causal consistency) does not
-presume a multi-version storage system. Rather, in both traditional
-and recent implementations, incoming writes are *buffered* when they
-are delivered to replicas but reads are never served from these
-buffers [Ahamad et al. 1995, Lloyd et al. 2011, Lloyd et
-al. 2013]. Perhaps this is a philosophical distinction, but we do not
-believe the authors of these systems would characterize their systems
-as multi-version in the tradition of classic database systems. There
+\response{We have updated the text. In general, writes Follow Reads
+  (and, more broadly, causal consistency) does not presume a
+  multi-version storage system. Rather, in both traditional and recent
+  implementations, incoming writes are *buffered* when they are
+  delivered to replicas but reads are never served from these buffers
+  [Ahamad et al. 1995, Lloyd et al. 2011, Lloyd et al. 2013]. Perhaps
+  this is a philosophical distinction, but we do not believe the
+  authors of these systems would characterize their systems as
+  multi-version in the tradition of classic database systems. There
   *is*, however, a possible visibility penalty--writes may indeed not
-become visible to other readers--which we explicity address in Section
-5.2.2.}
+  become visible to other readers--which we address in
+  Section 5.2.2.}
 
@@
 
-\response{We have clarified the discussion of "writes follow reads" to include a
-connection to Lamport's "happens-before" relation. We have opted to
-present causal consistency as it is treated in the literature on
-session guarantees and believe that, with the suggested modifications,
-this strikes a reasonable compromise.}
+\response{We have clarified the discussion of ``writes follow reads''
+  to include a connection to Lamport's ``happens-before'' relation. We
+  have opted to present causal consistency as it is treated in the
+  literature on session guarantees (i.e., Terry et al.) and believe
+  that the suggested modifications clarify this decision.}
 
@@
 
-\response{We have added a reference to PSI in 5.2.1.}
+\response{We have added a reference to PSI in 5.2.1 and also discuss it in Related Work.}
 
@@
 
-\response{We have placed references to the above in the related work section.}
+\response{We discuss and have placed references to the above in the related work section.}
 
@@
 
-\response{The Singapore numbers are located in the far right-hand side.}
+\response{The Singapore numbers are located in the far right-hand side of the table.}
 
@@
 
-\response{As we have noted above, "We have added a simpler but less efficient TA
+\response{As we have noted above, we have added a simpler but less efficient TA
 algorithm to demonstrate feasibility. We subsequently expand on this
@@
 \noindent\textbf{Reviewer 3 Comments}\\
-
-\reviewer{W1. Some definitions need to be clarified - see detailed comments. \\
-W2. The experimental results are not very surprising (and not particularly compelling). For the 
-record, I would have been in favor of accepting this paper even if it had had no evaluation. }
 
@@
 
-\response{This is an excellent point. We have added an additional Section
-(5.2.3) to address this.}
+\response{This is an excellent point. We have added an additional section (Section
+5.2.3) to specifically address this point.}
 
@@
 \response{We have modified this sentence to reflect the fact that low latency is
-a side effect of choosing high availability.}
+a side effect of implementations that guarantee high availability.}
 
@@
 \reviewer{D6.
-Transactional availability (Section 4.2) requires "replica availability", which is not defined.
+Transactional availability (Section 4.2) requires ``replica availability'', which is not defined.
 Does it correspond to what you previously called "traditional availability"?
@@
 
-\response{We have added a definition of "replica availability" and removed the
+\response{We have added a definition of ``replica availability'' and removed the
 problematic example.}
@@
 
-\response{This is an excellent point. We believe that presenting the simplified
-TA algorithm presentation in Section 5.1.2 (see responses to reviewers
-above) is easily integrated with Item Cut Isolation and should address
-any concerns regarding feasibility. However, we have also noted your
-concerns in Section 5.4: more efficient algorithms for I-CI plus RA
-(nee TA) are an area for future work.}
+\response{This is an excellent point. We believe that presenting the
+  simplified TA algorithm presentation in Section 5.1.2 (see responses
+  to reviewers above) is easily integrated with Item Cut Isolation and
+  should address any concerns regarding feasibility. However, we have
+  also noted your concerns in Section 5.4: more efficient algorithms
+  for I-CI plus RA (previously called TA---see above) are an area for
+  future work.}
 

### 2013-08-01 writeup/evaluation.tex

@@
 last-writer-wins RU with standard all-to-all anti-entropy between
-replicas) and HAT RC and RA with RC guarantees (hereafter,
-\texttt{RA}). It also supports non-HAT operation whereby all
-operations for a given key are routed to a (randomly) designated
-\texttt{master} replica for each key (guaranteeing single-key
-linearizability, as in Gilbert and Lynch's CAP Theorem
-proof~\cite{gilbert-cap} and in PNUTS~\cite{pnuts}'s
+replicas) and the efficient HAT RA algorithm as sketched in
+Section~\ref{sec:ta}. (hereafter, \texttt{RA}). It also supports
+non-HAT operation whereby all operations for a given key are routed to
+a (randomly) designated \texttt{master} replica for each key
+(guaranteeing single-key linearizability, as in Gilbert and Lynch's
+CAP Theorem proof~\cite{gilbert-cap} and in PNUTS~\cite{pnuts}'s
 \texttt{read-latest} operation; hereafter, \texttt{master}) as well as

### 2013-08-01 writeup/motivation.tex

--- 2013-07-24 writeup/motivation.tex
@@
 Even with fault-free networks, distributed systems face the challenge
-of network communication latency, Deutsch's second ``Fallacy.'' In this
-section, we quantify round-trip latencies, which are often
-large---hundreds of milliseconds in a geo-replicated,
-multi-datacenter context.  Fundamentally, the speed at which two
-servers can communicate is (according to modern physics) bounded by
-the speed of light. In the best case, two servers on opposite sides of
-the Earth---communicating via a hypothetical link along the planet's
-diameter---require a minimum 85.1ms RTT (133.7ms if sent at
-surface level). As services are replicated to multiple, geographically
-distinct sites, this cost of communication increases.
+of network communication latency, Deutsch's second ``Fallacy.'' In
+this section, we quantify round-trip latencies, which are often
+large---hundreds of milliseconds in a geo-replicated, multi-datacenter
+context.  Fundamentally, the speed at which two servers can
+communicate is (according to modern physics) bounded by the speed of
+light. In the best case, two servers on opposite sides of the
+Earth---communicating via a hypothetical link along the planet's
+diameter---require a minimum 85.1ms round-trip time (RTT; 133.7ms if
+sent at surface level). As services are replicated to multiple,
+geographically distinct sites, this cost of communication increases.
 
@@
 single ``availability zone'' (datacenter), at a granularity of
-1s\footnote{\url{http://www.bailis.org/blog/communication-costs-in-real-world-networks/}}. We
+1s\footnote{Data available at
+  \url{http://bailis.org/blog/communication-costs-in-real-world-networks/}}. We
 summarize the results of our network measurement study in
@@
 clear: remote communication has a substantial cost. As with network
-failures, quantifying and delays communication delays is an active
+failures, quantifying and minimizing communication delays is an active
 area of research in the networking community~\cite{bobtail}.
@@
 
-It is unknown \textit{which} of these guarantees can be provided with
-high availability, or are HAT-compliant. Existing algorithms for
+It has been unknown \textit{which} of these guarantees can be provided
+with high availability, or are HAT-compliant. Existing algorithms for
 providing weak isolation are often designed for a single-node context
-and are, to the best of our knowledge, unavailable due to reliance
-on concurrency control mechanisms like locking that are not resilient
-to partial failure (Section~\ref{sec:eval-existing}). Moreover, we are
+and are, to the best of our knowledge, unavailable due to reliance on
+concurrency control mechanisms like locking that are not resilient to
+partial failure (Section~\ref{sec:eval-existing}). Moreover, we are
 not aware of any prior literature that provides guidance as to the

### 2013-08-01 writeup/hat-analysis.tex

@@
 formal definitions in
+\iftechreport
  Appendix A.
-% our extended Technical Report~\cite{hat-tr}.
+\else
+ our extended Technical Report~\cite{hat-tr}.
+\fi
 In our examples, we exclusively consider read and write operations,
@@
 $T_2$'s $w_y(2)$ before $T_1$'s $w_y(1)$. Read Uncommitted is easily
-achieved via timestamping each update in a transaction with the same
+achieved by marking each of a transactions' writes with the same
 timestamp (unique across transactions; e.g., combining a client's ID
@@
 We have developed a more efficient RA algorithm, which we sketch here
-and provide greater detail in Appendix B. We begin with our Read
+and provide greater detail in
+\iftechreport
+ Appendix B.
+\else
+ our extended Technical Report~\cite{hat-tr}.
+\fi
+We begin with our Read
 Committed algorithm, but replicas wait to reveal new writes to readers
@@
 Serializability~\cite{1sr} need to prevent both Lost Update and Write
-Skew. Their phenomena prevention requirements mean that these
-guarantees are inherently unachievable in a HAT system.
+Skew. Their prevention requirements mean that these guarantees are
+inherently unachievable in a HAT system.
 

### 2013-10-01 writeup/hat-vldb.tex
Intro tweaks for CR

@@
 \let\endproof\relax
-%\usepackage{draftwatermarkbottom, draftwatermarktop}
+\usepackage{draftwatermarkbottom, draftwatermarktop}
 \usepackage{scrextend, MnSymbol}
@@
 
+\newenvironment{introenumerate}
+{
+
+   \vspace{-.5em}
+   \newcounter{qdcounter}
+    \begin{list}{\arabic{qdcounter}.~}{\usecounter{qdcounter}\leftmargin=1em}
+        \setlength{\topsep}{0em}
+        \setlength{\parskip}{0pt}
+        \setlength{\partopsep}{0pt}
+        \setlength{\parsep}{0pt}         
+        \setlength{\itemsep}{.25em} 
+        \setlength{\itemindent}{0em}
+}
+{
+    \end{list} 
+    \vspace{-.5em}
+}
+
 \begin{document}
@@
 \title{Highly Available Transactions: Virtues and Limitations\iftechreport\\(Extended Version)\fi}
-{\author{Peter Bailis{\fontsize{12}{14}$^\dagger$}, Aaron Davidson{\fontsize{12}{14}$^\dagger$}, Alan Fekete{\fontsize{12}{14}$^\diamond$}, Ali Ghodsi{\fontsize{12}{14}$^{\dagger,\ddagger}$}, Joseph M.
-Hellerstein{\fontsize{12}{14}$^\dagger$}, Ion Stoica{\fontsize{12}{14}$^\dagger$}\\{\affaddr{\fontsize{12}{14}$^\dagger$}\hspace{.5mm}UC Berkeley\hspace{4mm}{\fontsize{12}{14}$^\diamond$}\hspace{.5mm}University of Sydney\hspace{4mm}{\fontsize{12}{14}$^\ddagger$}\hspace{.5mm}KTH/Royal Institute of Technology}}}
+{\author{Peter Bailis, Aaron Davidson, Alan Fekete{\fontsize{12}{14}$^\dagger$}, Ali Ghodsi, Joseph M.
+Hellerstein, Ion Stoica \\{\affaddr{UC Berkeley and {\fontsize{12}{14}$^\dagger$}University of Sydney}}}}
 \maketitle

### 2013-10-01 writeup/intro.tex
Intro tweaks for CR

@@
 uniqueness guarantees. However, particularly for programs with
-non-monotonic logic, HATs can fall short, requiring
-sometimes-unavailable mechanisms that are more
-coordination-intensive. Overall, this paper quantifies the benefits
-that Highly Available Transactions offer as well as the necessary
-restrictions that they place on applications.
+non-monotonic logic or that are otherwise concurrency-sensitive, HATs
+can fall short, requiring non-HA coordination.
+
+We recognize that the large variety of ACID isolation levels and
+distributed consistency models (and therefore in our taxonomization)
+can be confusing and that the subtle distinctions between them may
+appear to be primarily of academic concern. Accordingly, we offer the
+following pragmatic takeaways:
+\begin{introenumerate}
+\item The default (and sometimes strongest) configurations of most
+  widely-deployed database systems expose a range of anomalies that
+  can compromise application-level consistency.
+
+\item Most of these ``weak isolation'' models are achievable without
+  sacrificing high availability if implemented correctly. However,
+  none of these models prevents concurrent modifications.
+
+\item In addition to providing a guaranteed response and horizontal
+  scale-out, these HAT models allow for one to three order of
+  magnitude lower latencies on current infrastructure.
+
+\item For correct behavior, most applications will require a
+  combination of HAT and (ideally sparing use of) non-HAT isolation levels;
+  future database designers should act accordingly.
+\end{introenumerate}
 

### 2013-10-01 writeup/conclusion.tex
Editing pass

@@
 many transactional guarantees are achievable with high availability,
-while a few, like preventing Lost Update and Write Skew, are not. This
-offers a useful connection between storage semantics studied in the
-database and distributed systems communities.
+while a few, like preventing Lost Update and Write Skew, are
+not. Surprisingly, many of the default (and sometimes strongest)
+semantics provided by today's traditional database systems are
+achievable as HATs, hinting that distributed databases need not
+compromise availability in order to serve many existing applications.
 
 HATs raise several challenges for future study. In this paper, we have
-highlighted standardized models, but we would like a more concise
-description of which models are and are not achievable in HAT
-systems. Similarly, while we have studied the analytical and
-experiment behavior of several HAT models, there is substantial work
-in further understanding the performance and design of systems within
-the large set of HAT models. We also believe it is time to revisit
-techniques such as Sagas~\cite{sagas} and other long-running
-transaction support in order to aid HAT programming. Weakened failure
-assumptions as in escrow or in the form of bounded network asynchrony
-could enable richer HAT semantics at the cost of general-purpose
-availability. Alternatively, there is a range of possible hybrid
-systems providing strong semantics during partition-free periods and
-weakened semantics during partitions. Based on our understanding of
-what is desired in the field and newfound knowledge of what is
-possible to achieve, we believe HATs provide a large and useful design
-space for exploration.
+largely focused on previously defined isolation and data consistency
+models from the consistency from the database and distributed systems
+communities. Their previous definitions and, in many cases, widespread
+adoption hints at their utility to end-users. However, we believe
+there is considerable work to be done to improve the programmability
+of highly available systems. Isolation and data consistency are means
+by which \textit{application-level} consistency is achieved but are
+typically not end goals for end-user applications. Our results hint
+that a mix of HAT and non-HAT semantics (the latter used sparingly) is
+required for practical applications, but the decision to employ each
+and the system architectures for a hybrid approach remain open
+problems. While we have studied the analytical and experiment behavior
+of several HAT models, there is substantial work in further
+understanding the performance and design of systems within the large
+set of HAT models. Weakened failure assumptions as in escrow or in the
+form of bounded network asynchrony could enable richer HAT semantics
+at the cost of general-purpose availability. Alternatively, there is a
+range of possible solutions providing strong semantics during
+partition-free periods and weakened semantics during partitions. Based
+on our understanding of what is desired in the field and newfound
+knowledge of what is possible to achieve, we believe HATs provide a
+large and useful design space for exploration.
 

### 2013-10-01 writeup/evaluation.tex
Editing pass

@@
 read-only transactions have item-cut isolation with causal consistency
-and RA (session \textit{PL-2L}~\cite{adya}) but are unavailable in the
+and MAV (session \textit{PL-2L}~\cite{adya}) but are unavailable in the
 presence of coordinator failure and assume serializable update
@@
 provide recency bounds. In this section, we attempt to understand when
-these guarantees matter both abstractly and in a representative
-transactional application based on the TPC-C benchmark~\cite{tpcc}.
-
-\vspace{.5em}\noindent\textbf{Commutativity and Monotonicity} As
-evidenced by the CALM Theorem~\cite{calm} and by Commutative and
-Replicated Data Types~\cite{crdt}, if updates logically commute, then
-they can often be safely performed in different orders at different
-replicas. Accordingly, as long as all writes are delivered to all
-replicas, then a system executing monotonic logic may not suffer from
-application-level consistency anomalies as a result of Lost Update or
-Write Skew anomalies. However, applications with non-monotone state
-mutation will not, in general, be able to maintain application-level
-consistency constraints with HATs alone---in particular, applications
-requiring visibility bounds should opt for unavailability.
+these guarantees matter in a representative transactional application
+based on the TPC-C benchmark~\cite{tpcc}.
 
@@
 transaction is monotonic---increment- and append-only---so all balance
-increase operations commute, and RA allows the maintenance of
+increase operations commute, and MAV allows the maintenance of
 foreign-key integrity constraints (e.g., via \texttt{UPDATE/DELETE
@@
 tables but, given the ability to update rows in both tables atomically
-via RA, can be easily maintained. Consistency Conditions 4 through 12
+via MAV, can be easily maintained. Consistency Conditions 4 through 12
 (3.3.2.4-12) can similarly be satisfied by applying updates atomically
@@
 last-writer-wins RU with standard all-to-all anti-entropy between
-replicas) and the efficient HAT RA algorithm as sketched in
-Section~\ref{sec:ta}. (hereafter, \texttt{RA}). It also supports
+replicas) and the efficient HAT MAV algorithm as sketched in
+Section~\ref{sec:ta}. (hereafter, \texttt{MAV}). It also supports
 non-HAT operation whereby all operations for a given key are routed to
@@
 write to LevelDB before responding to client requests, while new
-writes in RA are sent to a write-ahead log, which is synchronously
+writes in MAV are sent to a write-ahead log, which is synchronously
 flushed to disk.
@@
 buffering---is almost identical to \texttt{eventual}, while
-\texttt{RA}---which incurs two writes for every client-side write
+\texttt{MAV}---which incurs two writes for every client-side write
 (i.e., new writes are sent to the WAL then subsequently moved into
@@
 handle more than several thousand concurrent connections. In contrast,
-across two datacenters, the performance of eventual, RC, and RA are
+across two datacenters, the performance of eventual, RC, and MAV are
 near-identical to a single-datacenter deployment.
@@
 because worst-case server load was unaffected. With five clusters,
-\texttt{RA}'s relative throughput decreased: every YCSB \texttt{put}
+\texttt{MAV}'s relative throughput decreased: every YCSB \texttt{put}
 operation resulted in four \texttt{put} operations on remote replicas
@@
 garbage collection activity and, more importantly, IOPS when compared
-to \texttt{eventual} and \texttt{RC}, causing \texttt{RA} throughput
+to \texttt{eventual} and \texttt{RC}, causing \texttt{MAV} throughput
 to peak at around half of \texttt{eventual}. With in-memory
-persistence (i.e., no LevelDB or WAL), \texttt{RA} throughput was
+persistence (i.e., no LevelDB or WAL), \texttt{MAV} throughput was
 within 20\% of \texttt{eventual}.
@@
 \end{center}\vspace{-2.25em}
-\caption{Scale-out of RA, Eventual, and RC.}\vspace{-1.5em}
+\caption{Scale-out of MAV, Eventual, and RC.}\vspace{-1.5em}
 \label{fig:scaleout}
@@
 of \texttt{eventual}, \texttt{RC}, and \texttt{master} operation are
-unaffected by transaction length. In contrast, \texttt{RA} throughput
+unaffected by transaction length. In contrast, \texttt{MAV} throughput
 decreases linearly with increased transaction length: with $1$
-operation per transaction, \texttt{RA} throughput is within 18\% of
+operation per transaction, \texttt{MAV} throughput is within 18\% of
 \texttt{eventual} ($34$ bytes overhead), and with $128$ operations per
-transaction, \texttt{RA} throughput is within $60\%$ ($1898$ bytes
-overhead). This reflects our \texttt{RA} algorithm's metadata requirements,
+transaction, \texttt{MAV} throughput is within $60\%$ ($1898$ bytes
+overhead). This reflects our \texttt{MAV} algorithm's metadata requirements,
 which are proportional to transaction length and consume IOPS and
@@
 shown in Figure~\ref{fig:rprop} (clusters in Virginia and Oregon),
-with all reads, \texttt{RA} is within $4.8\%$ of \texttt{eventual};
-with all writes, \texttt{RA} is within $33\%$, and the throughput of
+with all reads, \texttt{MAV} is within $4.8\%$ of \texttt{eventual};
+with all writes, \texttt{MAV} is within $33\%$, and the throughput of
 \texttt{eventual} decreases by $288.8\%$ compared to all reads. At
-$99.8\%$ reads, \texttt{RA} incurs a $7\%$ overhead ($5.8\%$ for in-memory
+$99.8\%$ reads, \texttt{MAV} incurs a $7\%$ overhead ($5.8\%$ for in-memory
 storage).
@@
 Oregon (with $15$ YCSB clients per server) results in linear scale-out
-for \texttt{eventual}, \texttt{RC}, and \texttt{RA}. \texttt{RC} and
+for \texttt{eventual}, \texttt{RC}, and \texttt{MAV}. \texttt{RC} and
 \texttt{eventual} scale perfectly with an approximately $5$x
 throughput increase moving from $5$ to $25$ servers per cluster. For
-the same configuration, \texttt{RA} scales by $3.8$x, achieving over
-$260,000$ operations per second. \texttt{RA} performance is due to
+the same configuration, \texttt{MAV} scales by $3.8$x, achieving over
+$260,000$ operations per second. \texttt{MAV} performance is due to
 resource contention---with a memory-backed database (instead of
-LevelDB), \texttt{RA} scales by $4.25$x (not shown)---and RA-related
+LevelDB), \texttt{MAV} scales by $4.25$x (not shown)---and MAV-related
 performance heterogeneity across servers (Calvin's authors report
@@
 useful semantics without substantial performance penalties. In
-particular, our RA algorithm can achieve throughput competitive with
+particular, our MAV algorithm can achieve throughput competitive with
 eventual consistency at the expense of increased disk and network

### 2013-10-01 writeup/hat-analysis.tex
Editing pass

@@
 the \textit{isolation} effects of atomicity, which we call
-\textbf{Read Atomic (RA)} isolation.  Under RA, once some of the
+\textbf{Monotonic Atomic View (MAV)} isolation.  Under MAV, once some of the
 effects of a transaction $T_i$ are observed by another transaction
@@
 value whose later version is installed by $T_i$. Together with item
-cut isolation, RA prevents Read Skew anomalies (Berenson et al.'s A5A)
+cut isolation, MAV prevents Read Skew anomalies (Berenson et al.'s A5A)
 and is useful in several contexts such as maintaining foreign key
 constraints, consistent global secondary indexing, and maintenance of
-derived data. In the example below, under RA, because $T_2$ has read
+derived data. In the example below, under MAV, because $T_2$ has read
 $T_1$'s write to $y$, $T_2$ must observe $b=c=1$ (or later versions
@@
 $T_2$ can also observe $a=\bot$, $a=1$, or a later version of $x$. In
-the hierarchy of existing isolation properties, we place RA below
+the hierarchy of existing isolation properties, we place MAV below
 Adya's \textit{PL-2L} (as it does not necessarily enforce transitive
 read-write dependencies) but above Read Committed ($PL-2$). Notably,
-RA requires disallows reading intermediate writes (Adya's $G1b$):
+MAV requires disallows reading intermediate writes (Adya's $G1b$):
 observing all effects of a transaction implicitly requires observing
@@
 
-Perplexingly, discussions of RA are absent from existing treatments of
+Perplexingly, discussions of MAV are absent from existing treatments of
 weak isolation. This is perhaps again due to the single-node context
 in which prior work was developed: on a single server (or a fully
-replicated database), RA is achievable via lightweight locking and/or
+replicated database), MAV is achievable via lightweight locking and/or
 local concurrency control over data items~\cite{gstore,
-  kemme-thesis}. In contrast, in a distributed environment, RA over
+  kemme-thesis}. In contrast, in a distributed environment, MAV over
 arbitrary groups of non-colocated items is considerably more difficult
@@
 
-% RA can be achieved via multi-versioning. If clients attach vector
+% MAV can be achieved via multi-versioning. If clients attach vector
 % clocks to every write (incrementing their own position for each
@@
 % causal consistency and are subsequently sticky available (each is
-% implemented via client-side caching). However, RA is achievable
+% implemented via client-side caching). However, MAV is achievable
 % without stickiness via an alternate algorithm: if servers wait to
@@
 chosen bound. If the lower bound is advanced along transactional
-boundaries, clients will observe RA. This algorithm has several
+boundaries, clients will observe MAV. This algorithm has several
 variants in the literature~\cite{readonly, swift}: bounds can be
@@
 
-We have developed a more efficient RA algorithm, which we sketch here
+We have developed a more efficient MAV algorithm, which we sketch here
 and provide greater detail in
@@
 risk of invalidating application-level integrity constraints and can
-maintain limited criteria like foreign key constraints (via RA). We do
+maintain limited criteria like foreign key constraints (via MAV). We do
 not describe the entire space of application-level consistency
@@
 reads (i.e., Causal Transactional Predicate Cut Isolation). If we
-combine RA and P-CI, we have transactional snapshot reads. We can
+combine MAV and P-CI, we have transactional snapshot reads. We can
 achieve RC, MR, and RYW by simply sticking clients to servers. We can
@@
 \begin{tabular}{| c | p{6cm} | }\hline
-HA & Read Uncommitted (RU), Read Committed (RC), Read
-Atomic (RA), Item Cut Isolation (I-CI), Predicate Cut Isolation
-(P-CI), Writes Follow Reads (WFR), Monotonic Reads (MR), Monotonic
-Writes (MW)\\\hline Sticky & Read Your Writes (RYW), PRAM,
-Causal\\\hline Unavailable & Cursor Stability (CS)\lostupdate,
-Snapshot Isolation (SI)\lostupdate, Repeatable Read
-(RR)\lostupdate\rwskew, One-Copy Serializability
-(1SR)\lostupdate\rwskew, Recency\linearizable, Safe\linearizable,
-Regular\linearizable, Linearizability\linearizable, Strong
-1SR\lostupdate\rwskew\linearizable \\\hline
+HA & Read Uncommitted (RU), Read Committed (RC), Monotonic Atomic View
+(MAV), Item Cut Isolation (I-CI), Predicate Cut Isolation (P-CI),
+Writes Follow Reads (WFR), Monotonic Reads (MR), Monotonic Writes
+(MW)\\\hline Sticky & Read Your Writes (RYW), PRAM, Causal\\\hline
+Unavailable & Cursor Stability (CS)\lostupdate, Snapshot Isolation
+(SI)\lostupdate, Repeatable Read (RR)\lostupdate\rwskew, One-Copy
+Serializability (1SR)\lostupdate\rwskew, Recency\linearizable,
+Safe\linearizable, Regular\linearizable, Linearizability\linearizable,
+Strong 1SR\lostupdate\rwskew\linearizable \\\hline
 \end{tabular}
@@
   update\lostupdate, preventing write skew\rwskew, and requiring
-  recency guarantees\linearizable.}\vspace{-2em}
+  recency guarantees\linearizable.}
 \label{table:hatcompared}
@@
 
- \node[draw=none,fill=none] (ra) at (0, 1.475) {RA};
+ \node[draw=none,fill=none] (ra) at (0, 1.475) {MAV};
 
@@
 \end{figure}
-
-\subsection{Additional Discussion}
-\label{sec:discussion}
-
-In this section, we discuss several subtleties in our results,
-specifically addressing model composition, read atomicity
-versus linearizability, and benefits of stickiness.
-
-\vspace{.5em}\noindent\textbf{Model Composition} Choosing between
-combinations of compatible guarantees requires care. Consider the
-following transactions:
-\begin{align*}
-\small\vspace{-1em}
-T_1 &: w_x(1)~w_y(1)
-\\T_2 &: w_x(2)~w_y(2)
-\\T_3 &: r_x(a)~r_y(b)\vspace{-1em}
-\end{align*}
-If we want to guarantee both cut isolation and RA isolation and the
-system only executes $T_1$, $T_2$, and $T_3$, then $T_3$ needs to read
-$a=b=\bot$, $a=b=1$, or $a=b=2$. This means that either the
-implementation should frequently return $\bot$ (definitely undesirable
-and possibly non-convergent), keep multiple versions of each data item
-(necessitating potentially complicated distributed garbage
-collection), or use pre-declared read sets to fetch a consistent cut
-of keys before each transaction begins to execute. Using client-side
-caching can alleviate some of these challenges~\cite{bolton, swift} at
-the expense of requiring sticky availability. Study of the
-architectural interaction and overheads of various combinations is a
-promising area for future work.
-
-\vspace{.5em}\noindent\textbf{Linearizability and Read
-  Atomicity} The relationship between linearizability and
-RA is non-obvious. RA dictates that writes to
-multiple keys across multiple servers are made visible to readers all
-at once, while linearizability dictates that writes to a single key on
-multiple servers are made visible to all readers at once---what is
-different? First, in linearizable (and safe and regular) systems,
-writes are made visible to clients \textit{immediately} after they
-finish. With RA, there is no recency
-guarantee. Second, in linearizable systems, all clients see all writes
-at the same time. With RA as defined here, clients may see writes at
-different times depending on which replicas they contact. We are not
-aware of an analogous model in the distributed systems
-literature. Accordingly, despite apparent similarities, RA is
-incomparable with and less coordination intensive (by availability
-measures) than linearizability.
-
-\vspace{.5em}\noindent\textbf{Visibility and Stickiness} Sticky
-availability can improve write \textit{visibility}: clients will be
-able to safely read writes more quickly in a sticky available
-system. In the model we discussed, it is possible to achieve several
-properties like monotonic reads in a highly available system by
-waiting to reveal a write until all servers have seen it and its
-relevant dependencies. However, this incurs severe visibility
-penalties---new writes will not become visible to clients in the
-presence of partitions. A client that does not want to guarantee
-read-your-writes (due to the sticky availability requirement) may
-still wish to read other clients' writes with timeliness.

### 2013-10-01 writeup/hat-vldb.tex
Editing pass

@@
 suffer unavailability during system partitions or incur high network
-latency.  We introduce a new taxonomy of highly available systems and
+latency.  We introduce a taxonomy of highly available systems and
 analyze existing ACID and distributed consistency guarantees to
@@
 quantify the availability and performance benefits of HATs---often two
-to three orders of magnitude over wide-area networks---as well as
+to three orders of magnitude over wide-area networks---and discuss
 their necessary semantic compromises.
@@
 \input{conclusion.tex}
+
+\section*{Acknowledgments}
+We would like to thank Peter Alvaro, Neil Conway, Evan Jones, Adam
+Oliner, Aurojit Panda, and Shivaram Venkataraman for their helpful
+feedback on this work. This research was supported in part by the Air
+Force Office of Scientiﬁc Research (grant FA95500810352), DARPA XData
+Award FA8750-12-2-0331, the Natural Sciences and Engineering Research
+Council of Canada, the National Science Foundation (grants
+CNS-0722077, IIS-0713661, and IIS-0803690), NSF CISE Expeditions award
+CCF-1139158, the National Science Foundation Graduate Research
+Fellowship (grant DGE-1106400), and gifts from Amazon Web Services,
+Google, SAP, Cisco, Clearstory Data, Cloudera, Ericsson, Facebook,
+FitWave, General Electric, Hortonworks, Huawei, Intel, Microsoft,
+NetApp, Oracle, Samsung, Splunk, VMware, WANdisco and Yahoo!.
 

### 2013-10-01 writeup/intro.tex
Editing pass

@@
 
-\item For correct behavior, most applications will require a
-  combination of HAT and (ideally sparing use of) non-HAT isolation levels;
-  future database designers should act accordingly.
+\item For correct behavior, applications may require a combination of
+  HAT and (ideally sparing use of) non-HAT isolation levels; future
+  database designers should act accordingly.
 \end{introenumerate}

### 2013-10-02 writeup/conclusion.tex
More editing

@@
 of high availability will remain a critical aspect of many future data
-storage systems. Accordingly, we expose a broad design space of Highly
-Available Transactions (HATs), which can offer the key benefits of
-highly available distributed systems---``always on'' operation during
-partitions and low-latency operations---while also providing a family
-of transactional isolation levels and replica semantics that have been
-adopted in practice.  In taxonomizing this design space, we show that
-many transactional guarantees are achievable with high availability,
-while a few, like preventing Lost Update and Write Skew, are
-not. Surprisingly, many of the default (and sometimes strongest)
-semantics provided by today's traditional database systems are
-achievable as HATs, hinting that distributed databases need not
-compromise availability in order to serve many existing applications.
+storage systems. We expose a broad design space of Highly Available
+Transactions (HATs), which can offer the key benefits of highly
+available distributed systems---``always on'' operation during
+partitions and low-latency operations (often orders of magnitude lower
+than non-compliant)---while also providing a family of transactional
+isolation levels and replica semantics that have been adopted in
+practice.  We also identify many semantic guarantees that are
+unachievable with high availability, including Lost Update and Write
+Skew anomaly prevention, concurrent update prevention, and bounds on
+data recency. Despite these limitations, and somewhat surprisingly,
+many of the default (and sometimes strongest) semantics provided by
+today's traditional database systems are achievable as HATs, hinting
+that distributed databases need not compromise availability, low
+latency, or scalability in order to serve many existing applications.
 
-HATs raise several challenges for future study. In this paper, we have
-largely focused on previously defined isolation and data consistency
-models from the consistency from the database and distributed systems
-communities. Their previous definitions and, in many cases, widespread
-adoption hints at their utility to end-users. However, we believe
-there is considerable work to be done to improve the programmability
-of highly available systems. Isolation and data consistency are means
-by which \textit{application-level} consistency is achieved but are
-typically not end goals for end-user applications. Our results hint
-that a mix of HAT and non-HAT semantics (the latter used sparingly) is
-required for practical applications, but the decision to employ each
-and the system architectures for a hybrid approach remain open
-problems. While we have studied the analytical and experiment behavior
-of several HAT models, there is substantial work in further
-understanding the performance and design of systems within the large
-set of HAT models. Weakened failure assumptions as in escrow or in the
-form of bounded network asynchrony could enable richer HAT semantics
-at the cost of general-purpose availability. Alternatively, there is a
-range of possible solutions providing strong semantics during
-partition-free periods and weakened semantics during partitions. Based
-on our understanding of what is desired in the field and newfound
-knowledge of what is possible to achieve, we believe HATs provide a
-large and useful design space for exploration.
+In this paper, we have largely focused on previously defined isolation
+and data consistency models from the consistency from the database and
+distributed systems communities. Their previous definitions and, in
+many cases, widespread adoption hints at their utility to
+end-users. However, we believe there is considerable work to be done
+to improve the programmability of highly available systems. Isolation
+and data consistency are means by which \textit{application-level}
+consistency is achieved but are typically not end goals for end-user
+applications. Our results hint that a mix of HAT and non-HAT semantics
+(the latter used sparingly) is required for practical applications,
+but the decision to employ each and the system architectures for a
+hybrid approach remain open problems. While we have studied the
+analytical and experiment behavior of several HAT models, there is
+substantial work in further understanding the performance and design
+of systems within the large set of HAT models. Weakened failure
+assumptions as in escrow or in the form of bounded network asynchrony
+could enable richer HAT semantics at the cost of general-purpose
+availability. Alternatively, there is a range of possible solutions
+providing strong semantics during partition-free periods and weakened
+semantics during partitions. Based on our understanding of what is
+desired in the field and newfound knowledge of what is possible to
+achieve, we believe HATs represent a large and useful design space for
+exploration.
 

### 2013-10-02 writeup/evaluation.tex
More editing

@@
 infrastructure. Specifically:
-
-\begin{myenumerate}\vspace{-.5em}
+\begin{myenumerate}
 \item We revisit traditional database concurrency control with a focus
@@
 
-\subsection{Existing Algorithms}
+\subsection{HA and Existing Algorithms}
 \label{sec:eval-existing}
 
-Most existing database transaction and concurrency control algorithms
-are not designed for high availability and often presume a
-single-server deployment or a requirement for serializability. In this
-section, we briefly discuss design decisions and algorithmic details
-that preclude high availability.
+While we have shown that many database isolation levels are achievable
+as HATs, many traditional concurrency control mechanisms do not
+provide high availability---even for HAT-compliant isolation
+levels. Existing mechanisms often presume (or are adapted from)
+single-server non-partitioned deployments or otherwise focus on
+serializability as a primary use case. In this section, we briefly
+discuss design decisions and algorithmic details that preclude high
+availability.
 
@@
 read-only transactions have item-cut isolation with causal consistency
-and MAV (session \textit{PL-2L}~\cite{adya}) but are unavailable in the
-presence of coordinator failure and assume serializable update
+and MAV (session \textit{PL-2L}~\cite{adya}) but are unavailable in
+the presence of coordinator failure and assume serializable update
 transactions~\cite{readonly}; this is similar to read-only and
-write-only transactions more recently proposed by
-Eiger~\cite{eiger}.  Brantner's S3 database~\cite{kraska-s3}
-and Bayou~\cite{sessionguarantees} can all provide variants of session
+write-only transactions more recently proposed by Eiger~\cite{eiger}.
+Brantner's S3 database~\cite{kraska-s3} and
+Bayou~\cite{sessionguarantees} can all provide variants of session
 \textit{PL-2L} with high availability, but none provide this HAT
 functionality without substantial modification. Swift~\cite{swift} and
-bolt-on causal consistency~\cite{bolton} are closest to providing
-maximal sticky HAT semantics. Accordingly, it is possible to implement
-many guarantees weaker than serializability---including guarantees
-that are achievable with high availability---and still not achieve
-high availability.
+bolt-on causal consistency~\cite{bolton} are closest to providing the
+maximal sticky HAT semantics we have considered. Accordingly, it is
+possible to implement many guarantees weaker than
+serializability---including guarantees that are achievable with high
+availability---and still not achieve high availability. We view high
+availability as a core design consideration in future concurrency
+control designs.
 
@@
 latency comes in the inability to prevent Lost Update, Write Skew, and
-provide recency bounds. In this section, we attempt to understand when
-these guarantees matter in a representative transactional application
-based on the TPC-C benchmark~\cite{tpcc}.
-
-\vspace{.5em}\noindent\textbf{TPC-C} To better understand the impact
-of HAT-compliance in an application context, we consider a concrete
+provide recency bounds. To better understand the impact of
+HAT-compliance in an application context, we consider a concrete
 application: the TPC-C benchmark. In brief, we find that four of five
-transactions can be executed with HATs, while the fifth may require
+transactions can be executed via HATs, while the fifth requires
 unavailability.
@@
 in a real-world environment, we implemented a HAT database
-prototype. Across a range of deployments on public cloud
-infrastructures, we found that, as Section~\ref{sec:latency}'s
+prototype. Here, we verify that, as Section~\ref{sec:latency}'s
 measurements suggested, ``strongly consistent'' algorithms incur
-substantial latency penalties: over WAN, 10 to 100 times higher than
-their HAT counterparts. We find that HAT
-properties---while more resource-intensive than basic eventual
-consistency---scale well to many servers.
-
+substantial latency penalties (over WAN, 10 to 100 times higher than
+their HAT counterparts) compared to HAT-compliant algorithms, which
+scale linearly.
+x
 \vspace{.5em}\noindent\textit{Implementation.} Our prototype database
-is a partially replicated (hash-based, partitioned) key-value backed
-by LevelDB and implemented in Java using Apache Thrift. It currently
+is a partially replicated (hash-based partitioned) key-value backed by
+LevelDB and implemented in Java using Apache Thrift. It currently
 supports eventual consistency (hereafter, \texttt{eventual};
@@
 replicas) and the efficient HAT MAV algorithm as sketched in
-Section~\ref{sec:ta}. (hereafter, \texttt{MAV}). It also supports
-non-HAT operation whereby all operations for a given key are routed to
-a (randomly) designated \texttt{master} replica for each key
+Section~\ref{sec:ta}. (hereafter, \texttt{MAV}). We support non-HAT
+operation whereby all operations for a given key are routed to a
+(randomly) designated \texttt{master} replica for each key
 (guaranteeing single-key linearizability, as in Gilbert and Lynch's
 CAP Theorem proof~\cite{gilbert-cap} and in PNUTS~\cite{pnuts}'s
-\texttt{read-latest} operation; hereafter, \texttt{master}) as well as
+``read latest'' operation; hereafter, \texttt{master}) as well as
 distributed two-phase locking. Servers are durable: they synchronously
 write to LevelDB before responding to client requests, while new
-writes in MAV are sent to a write-ahead log, which is synchronously
-flushed to disk.
+writes in MAV are synchronously flushed to a disk-resident write-ahead
+log.
 
@@
 contain a single, fully replicated copy of the data---typically across
-datacenters, sticking all clients within a datacenter to their
+datacenters and stick all clients within a datacenter to their
 respective cluster (trivially providing read-your-writes and monotonic
-reads guarantees). By default, we use 5 Amazon EC2 \texttt{m1.xlarge}
-instances as servers in each cluster. For our workload, we link our
-client library to the YCSB benchmark~\cite{ycsb}, which is well-suited
-to LevelDB's key-value schema, grouping every eight YCSB operations
-from the default workload (50\% reads, 50\% writes) to form a
-transaction. We increase the number of keys in the workload from the
-default 1,000 to 100,000 with uniform random key access, keeping
-the default value size of $1KB$, and running YCSB for 180 seconds per
-configuration.
+reads guarantees). By default, we deploy 5 Amazon EC2
+\texttt{m1.xlarge} instances as servers in each cluster. For our
+workload, we link our client library to the YCSB
+benchmark~\cite{ycsb}, which is well-suited to LevelDB's key-value
+schema, grouping every eight YCSB operations from the default workload
+(50\% reads, 50\% writes) to form a transaction. We increase the
+number of keys in the workload from the default 1,000 to 100,000 with
+uniform random key access, keeping the default value size of $1KB$,
+and running YCSB for 180 seconds per configuration.
 
@@
 transaction, \texttt{MAV} throughput is within $60\%$ ($1898$ bytes
-overhead). This reflects our \texttt{MAV} algorithm's metadata requirements,
-which are proportional to transaction length and consume IOPS and
-network bandwidth.
+overhead). This reflects our \texttt{MAV} algorithm's metadata
+requirements, which are proportional to transaction length and consume
+IOPS and network bandwidth. We are currently investigating alternative
+HAT algorithms that do not incur this overhead.
 
@@
 performance heterogeneity across servers (Calvin's authors report
-similar heterogeneity on EC2~\cite{calvin}). Nevertheless, this linear
-scalability is not typical of traditional database system
-designs~\cite{gray-isolation}.
+similar heterogeneity on EC2~\cite{calvin}). Our early experiments
+with a newer database prototype with more efficient (non-Thrift) RPC
+and connection pooling suggest that this scalability can be
+substantially reduced. Nevertheless, this linear scalability is not
+typical of traditional database system designs~\cite{gray-isolation}.
 

### 2013-10-02 writeup/hat-analysis.tex
More editing

@@
 
+In light the of current practice of deploying weak isolation levels
+(Section~\ref{sec:modernacid}), it is perhaps surprising that so many
+weak isolation levels are achievable as HATs. Indeed, isolation levels
+such as Read Committed expose and are defined in terms of end-user
+anomalies that could not arise during serializable execution. However,
+the prevalence of these models suggests that, in many cases,
+applications can tolerate these their associated anomalies. Given our
+HAT-compliance results, this in turn hints that--despite
+idiosyncrasies relating to concurrent updates and data recency--highly
+available database systems can provide sufficiently strong semantics
+for many applications. Indeed, HAT databases may expose more anomalies
+than a single-site database operating under weak isolation
+(particularly during network partitions). However, for a fixed
+isolation level (which, in practice, can vary across databases and may
+differ from implementation-agnostic definitions in the literature),
+users of single-site database are subject to the same (worst-case)
+application-level anomalies as a HAT implementation. The necessary
+(indefinite) visibility penalties (i.e., the right side of
+Figure~\ref{fig:hat-order}) and lack of support for preventing
+concurrent updates (via the upper left half of
+Figure~\ref{fig:hat-order}) mean HATs are \textit{not} well-suited for
+all applications (see Section~\ref{sec:evaluation}): these limitations
+are fundamental. However, common practices such as ad-hoc, user-level
+compensation and per-statement isolation ``upgrades'' (e.g.,
+\texttt{SELECT FOR UPDATE} under weak isolation)--commonly used to
+augment weak isolation--are also applicable in HAT systems (although
+they may in turn compromise availability).
+
  \newcommand{\lostupdate}{$^\dagger$}

### 2013-10-02 writeup/hat-vldb.tex
More editing

--- 2013-10-01 writeup/hat-vldb.tex
@@
 
-   \vspace{0mm}
+   \vspace{-.5em}
    \newcounter{qcounter}
@@
 suffer unavailability during system partitions or incur high network
-latency.  We introduce a new taxonomy of highly available systems and
-analyze existing ACID and distributed consistency guarantees to
-identify which can and cannot be achieved in HAT systems. This unifies
-the literature on weak transactional isolation, replica consistency,
-and highly available systems. We analytically and experimentally
-quantify the availability and performance benefits of HATs---often two
-to three orders of magnitude over wide-area networks---as well as
-their necessary semantic compromises.
+latency.  We introduce a taxonomy of highly available systems and
+analyze existing ACID isolation and distributed data consistency
+guarantees to identify which can and cannot be achieved in HAT
+systems. This unifies the literature on weak transactional isolation,
+replica consistency, and highly available systems. We analytically and
+experimentally quantify the availability and performance benefits of
+HATs---often two to three orders of magnitude over wide-area
+networks---and discuss their necessary semantic compromises.
 \end{abstract}
@@
 \input{conclusion.tex}
+
+\section*{Acknowledgments}
+We would like to thank Peter Alvaro, Neil Conway, Evan Jones, Adam
+Oliner, Aurojit Panda, and Shivaram Venkataraman for their helpful
+feedback on this work. This research was supported in part by the Air
+Force Office of Scientiﬁc Research (grant FA95500810352), DARPA XData
+Award FA8750-12-2-0331, the Natural Sciences and Engineering Research
+Council of Canada, the National Science Foundation (grants
+CNS-0722077, IIS-0713661, and IIS-0803690), NSF CISE Expeditions award
+CCF-1139158, the National Science Foundation Graduate Research
+Fellowship (grant DGE-1106400), and gifts from Amazon Web Services,
+Google, SAP, Cisco, Clearstory Data, Cloudera, Ericsson, Facebook,
+FitWave, General Electric, Hortonworks, Huawei, Intel, Microsoft,
+NetApp, Oracle, Samsung, Splunk, VMware, WANdisco, and Yahoo!.
 

### 2013-10-02 writeup/intro.tex
More editing

@@
 consistency, transactional isolation, and high availability. The
-recent resurgence of transactional systems like Spanner~\cite{spanner}
-suggests that programmers value transactional semantics, but,
-unfortunately, existing transactional data stores do not provide
-availability in the presence of partitions~\cite{orleans,foundation-article,
-  hstore,spanner,eiger, walter,calvin}.
+recent resurgence of transactional systems suggests that programmers
+value transactional semantics, but, unfortunately, existing
+transactional data stores do not provide availability in the presence
+of partitions~\cite{orleans,foundation-article, hstore,spanner,eiger,
+  walter,calvin}.
 
@@
   hstore, calvin} or snapshot
-isolation~\cite{daudjee-snapshot, walter}. Additionally, the distributed computing and parallel
+isolation~\cite{kemme-thesis,walter}. Additionally, the distributed computing and parallel
 hardware literature contains many consistency models for single

### 2013-10-02 writeup/motivation.tex
More editing

@@
 \textit{latency}. In this section, we draw on a range of evidence that
-indicates that partitions occur relatively frequently in real-world
-deployments and best-case latencies between datacenters are
-substantial, often on the order of several hundreds of milliseconds.
+indicates that partitions occur with frequency in real-world
+deployments and latencies between datacenters are substantial, often
+on the order of several hundreds of milliseconds.
 
@@
 light. In the best case, two servers on opposite sides of the
-Earth---communicating via a hypothetical link along the planet's
-diameter---require a minimum 85.1ms round-trip time (RTT; 133.7ms if
-sent at surface level). As services are replicated to multiple,
+Earth---communicating via a hypothetical link through the planet's
+core---require a minimum 85.1ms round-trip time (RTT; 133.7ms if sent
+at surface level). As services are replicated to multiple,
 geographically distinct sites, this cost of communication increases.

### 2013-10-02 writeup/relatedwork.tex
More editing

@@
 relies on two-phase commit and two-phase locking for read/write
-transactions~\cite{spanner}. As we have discussed, the latency
-penalties associated with this design choice are fundamental to
-serializability. For users willing to tolerate these costs, Spanner,
-or similar strongly consistent, unavailable
+transactions~\cite{spanner}. As we have discussed, the penalties
+associated with this design are fundamental to serializability. For
+users willing to tolerate unavailability and increased latency,
+Spanner, or similar ``strongly consistent''
 systems~\cite{kemme-classification}---including Calvin~\cite{calvin},
@@
 Walter~\cite{walter}, and a range of snapshot isolation
-techniques~\cite{daudjee-session, daudjee-snapshot}---are a reasonable
-choice.
+techniques~\cite{daudjee-session}---reasonable
+choices.
 

### 2013-10-04 writeup/evaluation.tex
more edits

@@
 
-To further understand the performance implications of HAT guarantees
-in a real-world environment, we implemented a HAT database
-prototype. Here, we verify that, as Section~\ref{sec:latency}'s
-measurements suggested, ``strongly consistent'' algorithms incur
-substantial latency penalties (over WAN, 10 to 100 times higher than
-their HAT counterparts) compared to HAT-compliant algorithms, which
-scale linearly.
-x
+To demonstrate the performance implications of HAT guarantees in a
+real-world environment, we implemented a HAT database prototype. We
+verify that, as Section~\ref{sec:latency}'s measurements suggested,
+``strongly consistent'' algorithms incur substantial latency penalties
+(over WAN, 10 to 100 times higher than their HAT counterparts)
+compared to HAT-compliant algorithms, which scale linearly. Our goal
+is \textit{not} a complete performance analysis of HAT semantics but
+instead a proof-of-concept demonstration of a small subset of the HAT
+space on real-world infrastructure.
+
 \vspace{.5em}\noindent\textit{Implementation.} Our prototype database

### 2013-10-04 writeup/hat-analysis.tex
more edits

@@
 It is possible to satisfy Item Cut Isolation with high availability by
-having transactions cache their reads at the client such that the
-values that they read for each item never changes unless they
-overwrite it themselves. The cache can be cleared at the end of each
-transaction and can alternatively be accomplished on (sticky) replicas
-via multi-versioning. Predicate Cut Isolation is also achievable in
-HAT systems via similar caching or multi-versioning that track entire
-logical ranges of predicates in addition to item-based reads.
+having transactions store a copy of any read data at the client such
+that the values that they read for each item never changes unless they
+overwrite it themselves. These stored values can be discarded at the
+end of each transaction and can alternatively be accomplished on
+(sticky) replicas via multi-versioning. Predicate Cut Isolation is
+also achievable in HAT systems via similar caching middleware or
+multi-versioning that track entire logical ranges of predicates in
+addition to item-based reads.
 
@@
 (indefinite) visibility penalties (i.e., the right side of
-Figure~\ref{fig:hat-order}) and lack of support for preventing
+Figure~\ref{fig:hatcompared}) and lack of support for preventing
 concurrent updates (via the upper left half of
-Figure~\ref{fig:hat-order}) mean HATs are \textit{not} well-suited for
-all applications (see Section~\ref{sec:evaluation}): these limitations
-are fundamental. However, common practices such as ad-hoc, user-level
-compensation and per-statement isolation ``upgrades'' (e.g.,
-\texttt{SELECT FOR UPDATE} under weak isolation)--commonly used to
-augment weak isolation--are also applicable in HAT systems (although
-they may in turn compromise availability).
+Figure~\ref{fig:hatcompared}) mean HATs are \textit{not} well-suited
+for all applications (see Section~\ref{sec:evaluation}): these
+limitations are fundamental. However, common practices such as ad-hoc,
+user-level compensation and per-statement isolation ``upgrades''
+(e.g., \texttt{SELECT FOR UPDATE} under weak isolation)--commonly used
+to augment weak isolation--are also applicable in HAT systems
+(although they may in turn compromise availability).
 

### 2013-10-04 writeup/hat-appendix.tex
more edits

@@
 
-\begin{definition}[Read Atomic (RA)]
-A system that provides Read Atomic isolation prohibits phenomenon
-OTV in addition to providing Read Committed isolation.
+\begin{definition}[Monotonic Atomic View (MAV)]
+A system that provides Monotonic Atomic View isolation prohibits
+phenomenon OTV in addition to providing Read Committed isolation.
 \end{definition}
@@
 
-\section{Read Atomic Isolation}
+\section{Monotonic Atomic View Isolation}
 
 In this section, we provide additional information on our two-phase
-algorithm for providing Read Atomic isolation.
+algorithm for providing Monotonic Atomic View isolation. This section
+is effectively a work-in-progress; we are actively investgating
+alternative algorithms (particularly with respect to metadata
+overheads) and stronger (but still HAT) semantics that complement the
+techniques presented here. The interested reader (or would-be systems
+implementer) should contact Peter Bailis for more details.
 
@@
 \end{align*}
-For RA, the system must ensure that $a=1$. Consider a system with one
+For MAV, the system must ensure that $a=1$. Consider a system with one
 replica for each of $x$ and $y$ (denoted $R_x$ and $R_y$). Say client
@@
 
-\vspace{1em}\noindent\textbf{Overheads.} RA requires overhead both on client and server.
+\vspace{1em}\noindent\textbf{Overheads.} MAV requires overhead both on client and server.
 First, clients must store a vector (\texttt{required}) for the duration of a transaction in order to determine what values to read.
@@
 We have also not experimentally determined the visibility effect of the side-channel pending stable notifications as in the first optimization.
-
-\vspace{1em}\noindent\textbf{Algorithmic improvements.} We are actively investigating more efficient alternatives to the algorithm presented here.
-In particular, a lower-overhead (but still highly available) implementation that used less metadata would be appealing.
-A system using server-side vector clocks is feasible but does not necessarily reduce overhead because the number of items written per transaction is likely smaller than the number of servers in a system.
-Similarly, we have observed an algorithm trade-off between the size of $ts_{required}$ and write metadata.
-While it is certainly possible to reduce overheads by reducing availability~\cite{eiger}, we believe high availability is a worthwhile goal for future exploration.
 

### 2013-10-04 writeup/model.tex
more edits

@@
 replica, distributed algorithms often assume a model in which clients
-always contact the same logical (sets of) replica(s) across subsequent
-operations. As we will discuss in Section~\ref{sec:hats}, clients can
-ensure continuity between operations (e.g., reading their prior
-updates to a data item) by maintaining affinity or ``stickiness'' with
-a server or set of servers~\cite{vogels-defs}. In a fully-replicated
-system, where all servers are replicas for all data items, stickiness
-is simple: a client can maintain stickiness by contacting the same
-server for each of its requests. However, to stay ``sticky'' in a
-partially-replicated system, where servers are replicas for subsets of
-the set of data items (which we consider in this paper), a client must
-maintain stickiness with a single \textit{logical} server that may
-consist of multiple physical servers. We say that a system provides
-\textbf{sticky availability} if, whenever a client's operation on an
-item is executed against a logical server that has observed all of its
-prior operations, it eventually receives a response, even in the
-presence of indefinitely long partitions. A client may choose to
-become sticky available by acting as a server itself; for example, a
-client might cache its reads and writes~\cite{bolton,
-  sessionguarantees, swift}. Any guarantee achievable in a highly
-available system is achievable in a sticky high availability system
-but not vice-versa.
+always contact the same logical replica(s) across subsequent
+operations, whereby each of the client's prior operations (but not
+necessarily other clients' operations) are reflected in the database
+state that they observe. As we will discuss in Section~\ref{sec:hats},
+clients can ensure continuity between operations (e.g., reading their
+prior updates to a data item) by maintaining affinity or
+``stickiness'' with a server or set of servers~\cite{vogels-defs}. In
+a fully-replicated system, where all servers are replicas for all data
+items, stickiness is simple: a client can maintain stickiness by
+contacting the same server for each of its requests. However, to stay
+``sticky'' in a partially-replicated system, where servers are
+replicas for subsets of the set of data items (which we consider in
+this paper), a client must maintain stickiness with a single
+\textit{logical} copy of the database, which may consist of multiple
+physical servers. We say that a system provides \textbf{sticky
+  availability} if, whenever a client's transactions is executed
+against a copy of database state that reflects all of the client's prior
+transactions, it eventually receives a response, even in the presence
+of indefinitely long partitions. A client may choose to become sticky
+available by acting as a server itself; for example, a client might
+cache its reads and writes~\cite{bolton, sessionguarantees,
+  swift}. Any guarantee achievable in a highly available system is
+achievable in a sticky high availability system but not vice-versa.
 

### 2013-10-04 writeup/evaluation.tex
More edits'

@@
 \textit{PL-2L} with high availability, but none provide this HAT
-functionality without substantial modification. Swift~\cite{swift} and
-bolt-on causal consistency~\cite{bolton} are closest to providing the
-maximal sticky HAT semantics we have considered. Accordingly, it is
+functionality without substantial modification. Accordingly, it is
 possible to implement many guarantees weaker than
-serializability---including guarantees that are achievable with high
-availability---and still not achieve high availability. We view high
-availability as a core design consideration in future concurrency
-control designs.
+serializability---including HAT semantics---and still not achieve high
+availability. We view high availability as a core design consideration
+in future concurrency control designs.
 

### 2013-10-04 writeup/hat-analysis.tex
More edits'

@@
 observed and construct a lower bound on the versions that can be found
-on each replica. At the start of each transaction, clients choose a
-bound that is guaranteed to be lower than or equal to the this global
-lower bound, and, during transaction execution, replicas return the
-latest version of each item that is not greater than the client's
-chosen bound. If the lower bound is advanced along transactional
-boundaries, clients will observe MAV. This algorithm has several
-variants in the literature~\cite{readonly, swift}: bounds can be
-represented by vector clocks, and older versions can be asynchronously
-garbage collected.
+on every replica (which can be represented by either a list of
+versions, or, more realistically, a vector clock). At the start of
+each transaction, clients can choose a \textit{read timestamp} that is
+lower than or equal to the this global lower bound, and, during
+transaction execution, replicas return the latest version of each item
+that is not greater than the client's chosen timestamp. If this lower
+bound is advanced along transactional boundaries, clients will observe
+MAV. This algorithm has several variants in the
+literature~\cite{readonly, swift}: bounds can be represented by vector
+clocks, and older versions can be asynchronously garbage collected.
 
@@
 \begin{align*}
-\small\vspace{-1em}
+\small
 T_1 &: w_x(1)
-\\T_2 &: r_x(a)\vspace{-1em}
+\\T_2 &: r_x(a)
 \end{align*}
@@
 
-\vspace{.5em}\noindent{\textbf{Convergence}} To require that the
-system propagates writes between replicas, we can require convergence,
-or \textit{eventual consistency}: in the absence of new mutations to a
-data item, in the absence of partitions, all servers should eventually
-agree on the value for each item~\cite{cac, vogels-defs}. This is
-typically accomplished by any number of anti-entropy protocols, which
-periodically update neighboring servers with the latest value for each
-data item~\cite{antientropy}. Establishing a final convergent value is
+\vspace{.5em}\noindent{\textbf{Convergence} Under arbitrary (but not
+  infinite delays), HAT systems can ensure convergence, or
+  \textit{eventual consistency}: in the absence of new mutations to a
+  data item, all servers should eventually agree on the value for each
+  item~\cite{cac, vogels-defs}. This is typically accomplished by any
+  number of anti-entropy protocols, which periodically update
+  neighboring servers with the latest value for each data
+  item~\cite{antientropy}. Establishing a final convergent value is
   related to determining a total order on transaction updates to each

### 2013-10-04 writeup/hat-vldb.tex
More edits'

--- 2013-10-01 writeup/hat-vldb.tex
@@
 \let\endproof\relax
-\usepackage{draftwatermarkbottom, draftwatermarktop}
+%\usepackage{draftwatermarkbottom, draftwatermarktop}
 \usepackage{scrextend, MnSymbol}
@@
 
+\setlength{\abovedisplayskip}{4pt}
+\setlength{\abovedisplayshortskip}{4pt}
+\setlength{\belowdisplayskip}{4pt}
+\setlength{\belowdisplayshortskip}{4pt}
+
 % ****************** TITLE ****************************************
@@
 \input{conclusion.tex}
+
+\section*{Acknowledgments}
+We would like to thank Peter Alvaro, Neil Conway, Evan Jones, Adam
+Oliner, Aurojit Panda, Shivaram Venkataraman, and the HotOS and VLDB
+reviewers for their helpful feedback on earlier revisions of this
+work. This research was supported in part by the Air Force Office of
+Scientiﬁc Research (grant FA95500810352), DARPA XData Award
+FA8750-12-2-0331, the National Science Foundation (grants CNS-0722077,
+IIS-0713661, and IIS-0803690), NSF CISE Expeditions award CCF-1139158,
+the National Science Foundation Graduate Research Fellowship (grant
+DGE-1106400), and by gifts from Amazon Web Services, Google, SAP,
+Cisco, Clearstory Data, Cloudera, Ericsson, Facebook, FitWave, General
+Electric, Hortonworks, Huawei, Intel, Microsoft, NetApp, Oracle,
+Samsung, Splunk, VMware, WANdisco, and Yahoo!.
 

### 2013-10-04 writeup/intro.tex
More edits'

@@
 unavailable. However, Read Committed isolation, transactional
-atomicity (Read Atomic isolation), and many of the weak consistency
+atomicity (``Monotonic Atomic View'' isolation), and many other consistency
 models from database and distributed systems are achievable via
@@
 distributed consistency models (and therefore in our taxonomization)
-can be confusing and that the subtle distinctions between them may
-appear to be primarily of academic concern. Accordingly, we offer the
-following pragmatic takeaways:
+can be confusing and that the subtle distinctions between them can
+often appear to be primarily of academic concern. Accordingly, we
+offer the following pragmatic takeaways:
 \begin{introenumerate}
@@
 \item In addition to providing a guaranteed response and horizontal
-  scale-out, these HAT models allow for one to three order of
-  magnitude lower latencies on current infrastructure.
+  scale-out, these highly available HAT models allow one to three
+  order of magnitude lower latencies on current infrastructure.
 

### 2013-10-04 writeup/relatedwork.tex
More edits'

@@
 stickiness~\cite{sessionguarantees, vogels-defs}---which we formalize
-in Section~\ref{sec:sticky}).  However, as we have discussed, the
-connection between weak, multi-object consistency models in a
-distributed setting and high availability is relatively
-unexplored. With the exception of our preliminary workshop paper
+in Section~\ref{sec:sticky}).  Aside from our earlier workshop paper
 discussing transactional availability, real-world ACID, and HAT RC and
-I-CI~\cite{hat-hotos}---which this work expands, providing additional
-algorithms and impossibility results, a full taxonomization,
-experimental and application analysis, and a greater discussion of
-implications and related work---we believe this paper is the first to
-explore this area.
+I-CI~\cite{hat-hotos}---which this work expands with additional
+semantics, algorithms, and analysis---we believe this paper is the
+first to explore the connections between transactional semantics, data
+consistency, and (Gilbert and Lynch~\cite{gilbert-cap}) availability.
 

### 2013-10-04 writeup/evaluation.tex
More tweaks

@@
 general-purpose read-write transactions in a distributed
-setting~\cite{bernstein-book, davidson-survey}
-require at least one RTT before committing. As an
-example, traditional two-phase locking for a transaction of length $T$
-may require $T$ \texttt{lock} operations and will require at least one
-\texttt{lock} and one \texttt{unlock} operation.  In a distributed
-environment, each of these lock operations requires coordination,
-either with other database servers or with a lock service. If this
-coordination mechanism is unavailable, transactions cannot safely
+setting~\cite{bernstein-book, davidson-survey} require at least one
+RTT before committing. As an example, traditional two-phase locking
+for a transaction of length $T$ may require $T$ \texttt{lock}
+operations and will require at least one \texttt{lock} and one
+\texttt{unlock} operation.  In a distributed environment, each of
+these lock operations requires coordination, either with other
+database servers or with a lock service. If this coordination
+mechanism is unavailable, transactions cannot safely
 commit. Similarly, optimistic concurrency control requires
@@
 to a designated master or coordination service for each of these
-classic algorithms.  The cost of round trips will be determined by the
-deployment environment, as we saw in Section~\ref{sec:motivation}; we
-will demonstrate this cost on public cloud infrastructure in
-Section~\ref{sec:prototype}.
+classic algorithms.  As we saw in Section~\ref{sec:motivation}, is
+will be determined by the deployment environment; we will further
+demonstrate this in Section~\ref{sec:prototype}.
 
@@
 and connection pooling suggest that this scalability can be
-substantially reduced. Nevertheless, this linear scalability is not
-typical of traditional database system designs~\cite{gray-isolation}.
+substantially reduced.
 

### 2013-10-04 writeup/hat-analysis.tex
More tweaks

@@
 We have developed a more efficient MAV algorithm, which we sketch here
-and provide greater detail in
-\iftechreport
- Appendix B.
-\else
- our extended Technical Report~\cite{hat-tr}.
-\fi
-We begin with our Read
+and provide greater detail in \iftechreport Appendix B.  \else our
+extended Technical Report~\cite{hat-tr}.  \fi We begin with our Read
 Committed algorithm, but replicas wait to reveal new writes to readers
@@
 stable. This is entirely master-less and operations never block due to
-replica coordination. Moreover, there are several possible
-optimizations that, for brevity, we do not describe here.
+replica coordination.
 
@@
 
-The solution to this conundrum is to give up high availability and
-settle for sticky availability. Sticky availability permits three
-additional guarantees, which we first define and then prove are
-unachievable in a generic highly available system:
+A solution to this conundrum is to forgo high availability and settle
+for sticky availability. Sticky availability permits three additional
+guarantees, which we first define and then prove are unachievable in a
+generic highly available system:
 
@@
 guarantee algorithms we presented earlier: only reveal new writes to
-clients when their dependencies have been revealed.
+clients when their (respective, model-specific) dependencies have been revealed.
 

### 2013-10-04 writeup/hat-vldb.tex
More tweaks

@@
 
-\section*{Acknowledgments}
-We would like to thank Peter Alvaro, Neil Conway, Evan Jones, Adam
-Oliner, Aurojit Panda, Shivaram Venkataraman, and the HotOS and VLDB
-reviewers for their helpful feedback on earlier revisions of this
-work. This research was supported in part by the Air Force Office of
-Scientiﬁc Research (grant FA95500810352), DARPA XData Award
-FA8750-12-2-0331, the National Science Foundation (grants CNS-0722077,
-IIS-0713661, and IIS-0803690), NSF CISE Expeditions award CCF-1139158,
-the National Science Foundation Graduate Research Fellowship (grant
-DGE-1106400), and by gifts from Amazon Web Services, Google, SAP,
-Cisco, Clearstory Data, Cloudera, Ericsson, Facebook, FitWave, General
-Electric, Hortonworks, Huawei, Intel, Microsoft, NetApp, Oracle,
-Samsung, Splunk, VMware, WANdisco, and Yahoo!.
+\vspace{.5em}\noindent\textbf{Acknowledgments} We would like to thank
+Peter Alvaro, Neil Conway, Evan Jones, Adam Oliner, Aurojit Panda,
+Shivaram Venkataraman, and the HotOS and VLDB reviewers for their
+helpful feedback on this work. This research was supported in part by
+the Air Force Office of Scientiﬁc Research (grant FA95500810352),
+DARPA XData Award FA8750-12-2-0331, the National Science Foundation
+(grants CNS-0722077, IIS-0713661, and IIS-0803690), NSF CISE
+Expeditions award CCF-1139158, the National Science Foundation
+Graduate Research Fellowship (grant DGE-1106400), and by gifts from
+Amazon Web Services, Google, SAP, Cisco, Clearstory Data, Cloudera,
+Ericsson, Facebook, FitWave, General Electric, Hortonworks, Huawei,
+Intel, Microsoft, NetApp, Oracle, Samsung, Splunk, VMware, WANdisco,
+and Yahoo!.
 

### 2013-10-04 writeup/model.tex
More tweaks

@@
     replicated} system, in which all servers are replicas for all data
-  items, and a \textit{partially replicated} system, in which at least
-  one server acts as a replica for only a subset of all data
-  items. For generality, and, for applicability to many modern
-  ``sharded'' or ``partitioned'' data storage systems~\cite{ bigtable,
-    pnuts, spanner, dynamo, hstore}, we consider partially replicated
-  systems.}
+  items and a \textit{partially replicated} system, in which at least
+  one server acts as a replica for a proper subset of all data
+  items. For generality, and, given the prevalence of these
+  ``sharded'' or ``partitioned'' systems~\cite{ bigtable, pnuts,
+    spanner, dynamo, hstore}, we consider partial replication here.}
 

### 2013-10-04 writeup/motivation.tex
More tweaks

@@
 to full site disruptions for popular web services like Reddit,
-Foursquare, and Heroku~\cite{ec2-downsites}. At global scale,
-hardware failures---like the 2011 outages in Internet backbones in
-North America and Europe due a router
-bug~\cite{juniper-partition}---and misconfigurations like the BGP
-faults in 2008~\cite{pakistan-youtube} and
-2010~\cite{research-experiment-partition} can cause widespread
-partitions.
+Foursquare, and Heroku~\cite{ec2-downsites}. At global scale, hardware
+failures---like the 2011 outages in Internet backbones in North
+America and Europe due a router bug~\cite{juniper-partition}---and
+misconfigurations like the BGP faults in 2008~\cite{pakistan-youtube}
+and 2010~\cite{research-experiment-partition} can cause widespread
+partitioning behavior.
 

### 2013-10-04 writeup/intro.tex
More tweaks
more writing

@@
 constraints for multi-item updates and, in some cases, provide limited
-uniqueness guarantees. However, particularly for programs with
-non-monotonic logic or that are otherwise concurrency-sensitive, HATs
-can fall short, requiring non-HA coordination.
+uniqueness guarantees. However, HATs can fall short for applications
+with concurrency-sensitive operations, requiring non-HA coordination.
 
-We recognize that the large variety of ACID isolation levels and
-distributed consistency models (and therefore in our taxonomization)
-can be confusing and that the subtle distinctions between them can
-often appear to be primarily of academic concern. Accordingly, we
+Finally, we recognize that the large variety of ACID isolation levels
+and distributed consistency models (and therefore those in our
+taxonomization) can be confusing; the subtle distinctions between
+models can appear to be of primarily academic concern. Accordingly, we
 offer the following pragmatic takeaways:
@@
 
-\item Most of these ``weak isolation'' models are achievable without
+\item Many of these ``weak isolation'' models are achievable without
   sacrificing high availability if implemented correctly. However,

### 2013-10-04 writeup/motivation.tex
more writing

@@
 
-These results---which do not consider server failures---indicate
-that network partitions \textit{do} occur within and across modern
-datacenters. We observe that, as a general trend, partitions exist and
-must accordingly be met with either unavailability at some servers or,
-as we will discuss, relaxed semantic guarantees.
+These studies indicate that network partitions \textit{do} occur
+within and across modern datacenters. We observe that these partitions
+must be met with either unavailability at some servers or, as we will
+discuss, relaxed semantic guarantees.
 
@@
 
-\caption{Mean RTT times on EC2 (min and max highlighted)}\vspace{-.5em}
+\caption{Mean RTT times on EC2 (min and max highlighted)}\vspace{-1em}
 \label{table:rtt}

### 2013-10-05 writeup/intro.tex
More editing

@@
 many previously proposed isolation and consistency models to the goal
-high availability, which guarantees a response from every data replica
-in the presence of arbitrary partitions between them.  We classify
-which among the wide array of models are achievable with high
+of high availability, which guarantees a response from each non-failed
+data replica in the presence of arbitrary partitions between them.  We
+classify which among the wide array of models are achievable with high
 availability, denoting them as {\em Highly Available Transactions}
@@
 unavailable. However, Read Committed isolation, transactional
-atomicity (``Monotonic Atomic View'' isolation), and many other consistency
-models from database and distributed systems are achievable via
-algorithms that rely on multi-versioning and limited client-side
-caching. For several guarantees, such as causal consistency with
-phantom prevention and ANSI Repeatable Read, we consider a modified
-form of high availability in which clients ``stick to'' (i.e., have
-affinity with) at least one server---a property which is often
-implicit in the distributed systems literature~\cite{herlihy-art,
-  eiger, cac} but which requires explicit consideration in a
-client-server replicated database context. This sticky availability is
-widely employed~\cite{eiger, vogels-defs} but is a less restrictive
-model (and therefore more easily achievable) than traditional high
+atomicity (Section~\ref{sec:ta}), and many other consistency models
+from database and distributed systems are achievable via algorithms
+that rely on multi-versioning and limited client-side caching. For
+several guarantees, such as causal consistency with phantom prevention
+and ANSI Repeatable Read, we consider a modified form of high
+availability in which clients ``stick to'' (i.e., have affinity with)
+at least one server---a property which is often implicit in the
+distributed systems literature~\cite{herlihy-art, eiger, cac} but
+which requires explicit consideration in a client-server replicated
+database context. This sticky availability is widely
+employed~\cite{eiger, vogels-defs} but is a less restrictive model
+(and therefore more easily achievable) than traditional high
 availability.
@@
 taxonomization) can be confusing; the subtle distinctions between
-models can appear to be of primarily academic concern. Accordingly, we
-offer the following pragmatic takeaways:
+models may appear to be of academic concern. Accordingly, we offer the
+following pragmatic takeaways:
 \begin{introenumerate}

### 2013-10-05 writeup/model.tex
More editing

@@
 especially given the possibility of aborts.
-
-\subsection{Replica Availability}
 
@@
 
-If a client wishes to execute a transaction that performs operations
-on multiple data items, then the client must be able to contact and
-receive a response from at least one replica for each data item. We
-say that a transaction has \textbf{replica availability} if it can
-contact at least one replica for every item it attempts to
-access. This may result in ``lower availability'' than a
-non-transactional requirement. Additionally, given the possibility of
-system-initiated aborts, we need to ensure useful forward progress: a
-system can trivially guarantee clients a response by always aborting
-all transactions. However, this is an unsatisfactory system because
+We say that a transaction has \textbf{replica availability} if it can
+contact at least one replica for every item it attempts to access;
+this may result in ``lower availability'' than a non-transactional
+availability requirement (e.g., single-item
+availability). Additionally, given the possibility of system-initiated
+aborts, we need to ensure useful forward progress: a system can
+trivially guarantee clients a response by always aborting all
+transactions. However, this is an unsatisfactory system because
 nothing good (transaction commit) ever happens; we should require a
-\textit{liveness} property~\cite{transaction-liveness}. We cannot
-guarantee that every transaction will commit---transactions may choose
-to abort themselves---but we need to make sure that the system will
-not indefinitely abort transactions on its own volition. We call a
-transaction abort due to a transaction's own choosing (e.g., as an
-operation of the transaction itself or due to a would-be violation of a declared integrity constraint) an \textit{internal abort} and an
-abort due to system implementation or operation an \textit{external
-  abort}. We say that a system provides \textbf{transactional
-  availability} if, given replica availability for every data item in
-a transaction, the transaction eventually commits (possibly after
-multiple client retries) or internally aborts~\cite{hat-hotos}.
+\textit{liveness} property~\cite{transaction-liveness}.
+
+A system cannot guarantee that every transaction will
+commit---transactions may choose to abort themselves---but we need to
+make sure that the system will not indefinitely abort transactions on
+its own volition. We call a transaction abort due to a transaction's
+own choosing (e.g., as an operation of the transaction itself or due
+to a would-be violation of a declared integrity constraint) an
+\textit{internal abort} and an abort due to system implementation or
+operation an \textit{external abort}. We say that a system provides
+\textbf{transactional availability} if, given replica availability for
+every data item in a transaction, the transaction eventually commits
+(possibly after multiple client retries) or internally
+aborts~\cite{hat-hotos}. A system provides \textbf{sticky
+  transactional availability} if, given sticky availability, a
+transaction eventually commits or internally aborts.

### 2013-10-05 writeup/evaluation.tex
Grammar

@@
 \vspace{.5em}\noindent\textit{Summary.} Many---but not all---TPC-C
-transactions are well-served by HATs. The two problematic
+transactions are well served by HATs. The two problematic
 transactions, New-Order and Payment, rely on non-monotonic state
@@
 workload, we link our client library to the YCSB
-benchmark~\cite{ycsb}, which is well-suited to LevelDB's key-value
+benchmark~\cite{ycsb}, which is well suited to LevelDB's key-value
 schema, grouping every eight YCSB operations from the default workload
@@
 across two datacenters, the performance of eventual, RC, and MAV are
-near-identical to a single-datacenter deployment.
+near identical to a single-datacenter deployment.
 
@@
 for \texttt{eventual}, \texttt{RC}, and \texttt{MAV}. \texttt{RC} and
-\texttt{eventual} scale perfectly with an approximately $5$x
-throughput increase moving from $5$ to $25$ servers per cluster. For
-the same configuration, \texttt{MAV} scales by $3.8$x, achieving over
-$260,000$ operations per second. \texttt{MAV} performance is due to
-resource contention---with a memory-backed database (instead of
-LevelDB), \texttt{MAV} scales by $4.25$x (not shown)---and MAV-related
+\texttt{eventual} scale linearly: increasing the number of servers per
+cluster from $5$ to $25$ yields an approximately $5$x throughput
+increase. For the same configurations, \texttt{MAV} scales by $3.8$x,
+achieving over $260,000$ operations per second. \texttt{MAV} suffers
+from contention in LevelDB---with a memory-backed database,
+\texttt{MAV} scales by $4.25$x (not shown)---and MAV-related
 performance heterogeneity across servers (Calvin's authors report
-similar heterogeneity on EC2~\cite{calvin}). Our early experiments
-with a newer database prototype with more efficient (non-Thrift) RPC
-and connection pooling suggest that this scalability can be
-substantially reduced.
+similar heterogeneity on EC2~\cite{calvin}). Initial experiments with
+a newer prototype including more efficient (non-Thrift) RPC and
+connection pooling suggest that this scalability can be substantially
+improved.
 

### 2013-10-05 writeup/hat-analysis.tex
Grammar

@@
 
-To begin, \textbf{Read Uncommitted} isolation is captured by Adya as
+To begin, Adya captures \textbf{Read Uncommitted} isolation as
 \textit{PL-1}. In this model, writes to each object are totally
@@
 $T_2$'s $w_y(2)$ before $T_1$'s $w_y(1)$. Read Uncommitted is easily
-achieved by marking each of a transactions' writes with the same
+achieved by marking each of a transaction's writes with the same
 timestamp (unique across transactions; e.g., combining a client's ID
@@
 \textbf{Read Committed} isolation is particularly important in
-practice as it is the default of many DBMSs (as in
-Section~\ref{sec:modernacid}). Centralized implementations differ,
+practice as it is the default isolation level of many DBMSs
+(Section~\ref{sec:modernacid}). Centralized implementations differ,
 with some based on long-duration exclusive locks and short-duration
@@
 properties beyond what is implied by the name ``Read Committed'' and
-what is captured by the implementation-agnostic definition: under
-Read Committed, transactions should not access uncommitted or
-intermediate versions of data items. This prohibits both ``Dirty
-Writes'', as above, and also ``Dirty Reads'' phenomena.  This
-isolation is Adya's \textit{PL-2} and is formalized by prohibiting
-Adya's \textit{G1\{a-c\}} (or ANSI's $P1$, or ``broad'' $P1$ [2.2]
-from Berenson et al.). For instance, in the example below, $T_3$
-should never see $a=1$, and, if $T_2$ aborts, $T_3$ should not read
-$a=3$:
+what is captured by the implementation-agnostic definition: under Read
+Committed, transactions should not access uncommitted or intermediate
+versions of data items. This prohibits both ``Dirty Writes'', as
+above, and also ``Dirty Reads'' phenomena.  This isolation is Adya's
+\textit{PL-2} and is formalized by prohibiting Adya's
+\textit{G1\{a-c\}} (or ANSI's $P1$, or ``broad'' $P1$ [2.2] from
+Berenson et al.). For instance, in the example below, $T_3$ should
+never see $a=1$, and, if $T_2$ aborts, $T_3$ should not read $a=3$:
 \begin{align*}
@@
 multi-versioning that track entire logical ranges of predicates in
-addition to item-based reads.
+addition to item based reads.
 
@@
   kemme-thesis}. In contrast, in a distributed environment, MAV over
-arbitrary groups of non-colocated items is considerably more difficult
+arbitrary groups of non-co-located items is considerably more difficult
 to achieve with high availability.
@@
 
-As a straw-man, replicas can store all versions ever written to each
+As a straw man, replicas can store all versions ever written to each
 data item. Replicas can gossip information about versions they have
@@
 MAV. This algorithm has several variants in the
-literature~\cite{readonly, swift}: bounds can be represented by vector
-clocks, and older versions can be asynchronously garbage collected.
+literature~\cite{readonly, swift}, and older versions can be
+asynchronously garbage collected.
 
@@
 \vspace{.5em}\noindent\textbf{{Monotonic writes}} requires that each
-session's writes become visible in the order they were submitted by
-the client. Any order on transactions should also be consistent with
-any precedence that a global observer would see.
+session's writes become visible in the order they were submitted. Any
+order on transactions should also be consistent with any precedence
+that a global observer would see.
 
@@
 from a globally agreed upon lower bound on the versions written. This
-is highly available as a client will never block due to inability to
-find a server with a sufficiently up-to-date version of a data
+is highly available because a client will never block due to inability
+to find a server with a sufficiently up-to-date version of a data
 item. However, it does not imply that transactions will read their own

### 2013-10-05 writeup/intro.tex
Grammar

@@
 replica, low latency, and a range of semantic guarantees including
-several whose usefulness is widely acepted such as Read
+several whose usefulness is widely accepted such as Read
 Committed. However, highly available systems are fundamentally unable
@@
 and distributed consistency models (and therefore those in our
-taxonomization) can be confusing; the subtle distinctions between
-models may appear to be of academic concern. Accordingly, we offer the
-following pragmatic takeaways:
+taxonomy) can be confusing; the subtle distinctions between models may
+appear to be of academic concern. Accordingly, we offer the following
+pragmatic takeaways:
 \begin{introenumerate}
 \item The default (and sometimes strongest) configurations of most
-  widely-deployed database systems expose a range of anomalies that
+  widely deployed database systems expose a range of anomalies that
   can compromise application-level consistency.
@@
   sacrificing high availability if implemented correctly. However,
-  none of these models prevents concurrent modifications.
+  none of the achievable models prevents concurrent modifications.
 

### 2013-10-05 writeup/model.tex
Grammar

@@
 distributed database, designated servers might perform operations for
-different data items; a server that can handle an operation for a
+different data items. A server that can handle an operation for a
 given data item is called a \textit{replica} for that
@@
 ``stickiness'' with a server or set of servers~\cite{vogels-defs}. In
-a fully-replicated system, where all servers are replicas for all data
+a fully replicated system, where all servers are replicas for all data
 items, stickiness is simple: a client can maintain stickiness by

### 2013-10-05 writeup/relatedwork.tex
Grammar

@@
 ``optimistic replication'' strategies~\cite{optimistic} and several
-recent efforts at further taxonomizing these strategies in light of
+recent efforts at further understanding these strategies in light of
 current practice and the proliferation of ``eventually consistent''
-stores~\cite{bailis-ec, bernstein-survey} (notably, Bernstein and
+stores~\cite{bailis-ec, bernstein-survey}. Notably, Bernstein and
 Das~\cite{bernstein-survey} specifically highlight the importance of
 stickiness~\cite{sessionguarantees, vogels-defs}---which we formalize
-in Section~\ref{sec:sticky}).  Aside from our earlier workshop paper
+in Section~\ref{sec:sticky}.  Aside from our earlier workshop paper
 discussing transactional availability, real-world ACID, and HAT RC and

### 2013-10-07 writeup/hat-analysis.tex
Final edits

@@
 
-HAT systems provide transactions with transactional availability and
-either high availability or sticky availability. They offer latency
-and availability benefits over traditional distributed databases, yet
-they cannot achieve all possible semantics. In this section, we
-delineate ACID, distributed replica consistency, and session
-consistency levels which can be achieved with high availability (Read
-Committed isolation, variants of Repeatable Read, atomic reads, and
-many session guarantees), those with sticky availability (read your
-writes, PRAM and causal consistency), and properties that cannot be
+HAT systems provide transactions with transactional availability or
+sticky transactional availability. They offer latency and availability
+benefits over traditional distributed databases, yet they cannot
+achieve all possible semantics. In this section, we describe ACID,
+distributed replica consistency, and session consistency levels which
+can be achieved with high availability (Read Committed isolation,
+variants of Repeatable Read, atomic reads, and many session
+guarantees), those with sticky availability (read your writes, PRAM
+and causal consistency). We also discuss properties that cannot be
 provided in a HAT system (those preventing Lost Update and Write Skew
@@
 session's writes become visible in the order they were submitted. Any
-order on transactions should also be consistent with any precedence
-that a global observer would see.
+order on transactions (as in Read Uncommitted isolation) should also
+be consistent with any precedence that a global observer would see.
 
@@
 ``happens-before'' relation~\cite{lamportclocks}.  Any order on
-transactions should respect the transitive reads-from
-order.\vspace{.5em}
+transactions should respect this transitive order.\vspace{.5em}
 
@@
 achievable, but we find several particularly compelling. If we combine
-all HAT and sticky guarantees, we have transactional, causal snapshot
-reads (i.e., Causal Transactional Predicate Cut Isolation). If we
-combine MAV and P-CI, we have transactional snapshot reads. We can
-achieve RC, MR, and RYW by simply sticking clients to servers. We can
-also combine unavailable models---for example, an unavailable system
-might provide PRAM and One-Copy
+all HAT and sticky guarantees, we have transactional, causally
+consistent snapshot reads (i.e., Causal Transactional Predicate Cut
+Isolation). If we combine MAV and P-CI, we have transactional snapshot
+reads. We can achieve RC, MR, and RYW by simply sticking clients to
+servers. We can also combine unavailable models---for example, an
+unavailable system might provide PRAM and One-Copy
 Serializability~\cite{daudjee-session}.
@@
 To the best of our knowledge, this is the first unification of
-database ACID transactions, distributed consistency, and session
+transactional isolation, distributed consistency, and session
 guarantee models. Interestingly, strong one-copy serializability
@@
 user-level compensation and per-statement isolation ``upgrades''
-(e.g., \texttt{SELECT FOR UPDATE} under weak isolation)--commonly used
-to augment weak isolation--are also applicable in HAT systems
+(e.g., \texttt{SELECT FOR UPDATE} under weak isolation)---commonly used
+to augment weak isolation---are also applicable in HAT systems
 (although they may in turn compromise availability).

### 2013-10-07 writeup/hat-vldb.tex
more writing
Final edits

@@
 \newif\iftechreport
-%\techreporttrue
-\techreportfalse
+\techreporttrue
+%\techreportfalse
 
@@
 
+\pagestyle{empty}
+
 % ****************** TITLE ****************************************
@@
 {\author{Peter Bailis, Aaron Davidson, Alan Fekete{\fontsize{12}{14}$^\dagger$}, Ali Ghodsi, Joseph M.
-Hellerstein, Ion Stoica \\{\affaddr{UC Berkeley and {\fontsize{12}{14}$^\dagger$}University of Sydney}}}}
+Hellerstein, Ion Stoica \\[1mm]{\affaddr{UC Berkeley and {\fontsize{12}{14}$^\dagger$}University of Sydney}}}}
 \maketitle
@@
 Intel, Microsoft, NetApp, Oracle, Samsung, Splunk, VMware, WANdisco,
-and Yahoo!.
+and Yahoo!.\vspace{.5em}
 

### 2013-10-07 writeup/intro.tex
Final edits

@@
 recent resurgence of transactional systems suggests that programmers
-value transactional semantics, but, unfortunately, existing
-transactional data stores do not provide availability in the presence
-of partitions~\cite{orleans,foundation-article, hstore,spanner,eiger,
+value transactional semantics, but most existing transactional data
+stores do not provide availability in the presence of
+partitions~\cite{orleans,foundation-article, hstore,spanner,eiger,
   walter,calvin}.
@@
 ACID databases---are not achievable with high availability in the
-presence of partitions~\cite{davidson-survey}. However, database
+presence of network partitions~\cite{davidson-survey}. However, database
 systems have a long tradition of providing weaker isolation and
@@
 Our main contributions in this paper are as follows. We relate the
-many previously proposed isolation and consistency models to the goal
-of high availability, which guarantees a response from each non-failed
-data replica in the presence of arbitrary partitions between them.  We
-classify which among the wide array of models are achievable with high
-availability, denoting them as {\em Highly Available Transactions}
-(HATs). In doing so, we demonstrate that although many implementations
-of HAT semantics are not highly available, this is an artifact of the
-implementations rather than an inherent property of the semantics. Our
-investigation shows that, besides serializability, Snapshot Isolation
-and Repeatable Read isolation are not HAT-compliant, while most other
-isolation levels are achievable with high availability. We also
-demonstrate that many weak replica consistency models from distributed
-systems are both HAT-compliant and simultaneously achievable with
-several ACID properties.
+many previously proposed database isolation and data consistency
+models to the goal of high availability, which guarantees a response
+from each non-failing server in the presence of arbitrary network
+partitions between them.  We classify which among the wide array of
+models are achievable with high availability, denoting them as {\em
+  Highly Available Transactions} (HATs). In doing so, we demonstrate
+that although many implementations of HAT semantics are not highly
+available, this is an artifact of the implementations rather than an
+inherent property of the semantics. Our investigation shows that,
+besides serializability, Snapshot Isolation and Repeatable Read
+isolation are not HAT-compliant, while most other isolation levels are
+achievable with high availability. We also demonstrate that many weak
+replica consistency models from distributed systems are both
+HAT-compliant and simultaneously achievable with several ACID
+properties.
 
@@
 uniqueness guarantees. However, HATs can fall short for applications
-with concurrency-sensitive operations, requiring non-HA coordination.
+with concurrency-sensitive operations, requiring unavailable,
+synchronous coordination.
 
@@
   HAT and (ideally sparing use of) non-HAT isolation levels; future
-  database designers should act accordingly.
+  database designers should plan accordingly.
 \end{introenumerate}

### 2013-10-07 writeup/model.tex
Final edits

@@
   availability} if, whenever a client's transactions is executed
-against a copy of database state that reflects all of the client's prior
-transactions, it eventually receives a response, even in the presence
-of indefinitely long partitions. A client may choose to become sticky
+against a copy of database state that reflects all of the client's
+prior operations, it eventually receives a response, even in the
+presence of indefinitely long partitions (where ``reflects'' is
+dependent on semantics). A client may choose to become sticky
 available by acting as a server itself; for example, a client might

### 2013-10-07 writeup/motivation.tex
Final edits

@@
 network is reliable. \textit{2.)} Latency is
-zero''~\cite{fallacies-deutsch}; in a distributed setting, network
+zero''~\cite{fallacies-deutsch}. In a distributed setting, network
 failures may prevent database servers from communicating, and, in the
@@
 distance, network congestion, and routing. As we will see
-(Section~\ref{sec:availability}), highly available system design
-mitigates the effects of network \textit{partitions} and
+(Section~\ref{sec:availability}), highly available system designs
+mitigate the effects of network \textit{partitions} and
 \textit{latency}. In this section, we draw on a range of evidence that
@@
 and 1000 seconds~\cite{ip-backbone-failures} as well as frequent path
-routing failures on the Internet~\cite{labovitz-failures}. Finally,
-Kingsbury and Bailis have recently collected an informal survey of
-network failures~\cite{aphyr-post}. Isolating, quantifying, and
-accounting for these network failures is an area of active research in
-networking community~\cite{uw-failure-networks}.
+routing failures on the Internet~\cite{labovitz-failures}. A recent,
+informal report by Kingsbury and Bailis catalogs a host of additional
+practitioner reports~\cite{aphyr-post}. Not surprisingly, isolating,
+quantifying, and accounting for these network failures is an area of
+active research in networking community~\cite{uw-failure-networks}.
 
@@
 single ``availability zone'' (datacenter), at a granularity of
-1s\footnote{Data available at
-  \url{http://bailis.org/blog/communication-costs-in-real-world-networks/}}. We
-summarize the results of our network measurement study in
+1s\footnote{Data available at {\scriptsize
+    \url{http://bailis.org/blog/communication-costs-in-real-world-networks/},
+    \url{https://github.com/pbailis/aws-ping-traces}}}. We summarize
+the results of our network measurement study in
 Table~\ref{table:rtt}. On average, intra-datacenter communication
@@
 distribution of latencies varies between links, but the trend is
-clear: remote communication has a substantial cost. As with network
-failures, quantifying and minimizing communication delays is an active
-area of research in the networking community~\cite{bobtail}.
+clear: remote communication has a substantial cost. Quantifying and
+minimizing communication delays is also an active area of research in
+the networking community~\cite{bobtail}.
 
@@
 single-node database, the coordination penalties associated with
-serializability can be severe, manifesting themselves in the form of
+serializability can be severe and are manifested in the form of
 decreased concurrency (and, subsequently, performance degradation,
-scalability limitations, and, often, external aborts like
-deadlocks)~\cite{gray-isolation}. Accordingly, to increase
+scalability limitations, and, often, aborts due to deadlock or
+contention)~\cite{gray-isolation}. Accordingly, to increase
 concurrency, database systems offer a range of ACID properties weaker
@@
 are allowable by the system~\cite{adya, ansi-sql, ansicritique}. None
-of these weak isolation models guarantees serializability, but, as we see below,
-their benefits are often considered to outweigh costs of possible
-consistency anomalies that might arise from their use.
+of these weak isolation models guarantees serializability, but, as we
+see below, their benefits are often considered to outweigh costs of
+possible consistency anomalies that might arise from their use.
 

### 2013-11-27 writeup/conclusion.tex
Fix typo

@@
 In this paper, we have largely focused on previously defined isolation
-and data consistency models from the consistency from the database and
+and data consistency models from the database and
 distributed systems communities. Their previous definitions and, in
